### 2009-02-24 pkg/inst/slides/partykit_useR2009.tex
+ useR 2009 slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@125 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[10pt]{article}

\usepackage{url}
\usepackage{natbib}

\renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
\newcommand{\affiliations}{\footnotesize}
\newcommand{\keywords}{\paragraph{Keywords:}}

\setlength{\oddsidemargin}{0cm} \setlength{\evensidemargin}{0cm}
\setlength{\textwidth}{16.5cm} \setlength{\topmargin}{-1cm}
\setlength{\textheight}{24.5cm}

\begin{document}
\pagestyle{empty}

\title{\textbf{partykit}: A Toolbox for Recursive Partytioning}

\begin{center}
  {\bf Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}
\end{center}

\begin{affiliations}
1. Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen \par
2. Department of Statistics and Mathematics, Wirtschaftsuniversit\"at Wien \par
\texttt{\{Torsten.Hothorn, Achim.Zeileis\}@R-project.org}
\end{affiliations}

\keywords Regression Trees, Visualization, CTree, CHAID, Weka, PMML.

\vskip 0.8cm

Recursive Partitioning methods, or simply `trees', are simple yet powerful
methods to estimate regression relationships. Since the publication of the
Automated Interaction Detection (AID) algorithm in 1964, many extentions,
modifications and new approaches have been suggested in both the statistics
and machine learning communities. Most of the algorithms in use today are
available to the \textsf{R} user, for example from packages \textbf{rpart}
\citep{PKG:rpart},
\textbf{RWeka} \citep{PKG:RWeka}, \textbf{party} \citep{PKG:party}, or \textbf{mvpart}
\citep{PKG:mvpart}.

Computations on fitted trees or the implementation of a new or modified
recursive partitioning method, however, are relatively difficult tasks.
As an example, one might want to predict a numeric response by the median
instead of the mean based on an \textit{rpart} object. Or, instead of
binary splits, one is going to implement an algorithm which potentially
uses multiway splits, maybe containing a non-trivial model in each terminal
node. These, and other, tasks require extensive programming,
mainly because each package dealing with (some specialized form of)
recursive partitioning uses a different internal representation of a tree. Thus, no
standard user-interface exists, and methods for predictions, graphical or textual
visualizations, or other computations heavily rely on these internals.

To overcome these difficulties, the \textbf{partykit} package \citep{PKG:partykit} offers a 
unified representation of tree objects along with \texttt{predict}, 
\texttt{print}, and \texttt{plot} methods. Trees are represented through 
a class \textit{party}, with building blocks \textit{partynode} and 
\textit{partysplit}. The design of these classes covers splits in arbitrary
functions of (one or more) variables, including multiway splits. In
addition, non-constant fits in the terminal nodes are supported as well.
Conversion methods exist for the most important sources of recursive
partitioning objects, currently \textit{rpart}, \textit{J48} and
\textit{pmmlTreeModel} objects can easily be converted into a \textit{party}
object to compute with. The \textbf{partykit} also offers a re-implementation
of conditional inference trees \cite{Hothorn:2006:JCGS}, illustrating the usefulness of the generic
infrastructure.

In our presentation, we will only sketch details of these classes and 
corresponding methods and focus
on applications of the toolkit, for example for producing nice and
non-standard plots
for \textit{rpart} or \textit{J48} objects, for computing predictions
on millions of new observations really fast, and for implementing CHAID, 
a popular classification tree algorithm.

\bibliographystyle{plainnat}
\bibliography{partykit}

\end{document}

### 2009-02-25 pkg/inst/slides/partykit_useR2009.tex
some re-phrasing and style touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@126 255f4249-2263-4417-b1ea-8498779e27a7

@@
-\documentclass[10pt]{article}
+\documentclass[10pt,a4paper]{article}
 
-\usepackage{url}
-\usepackage{natbib}
+\usepackage{url,a4wide}
+\usepackage[authoryear,round]{natbib}
 
 \renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
-\newcommand{\affiliations}{\footnotesize}
-\newcommand{\keywords}{\paragraph{Keywords:}}
+\renewcommand{\author}[1]{\begin{center}{\bf #1}\end{center}}
+\newenvironment{affiliations}{\footnotesize \begin{enumerate}}{\end{enumerate}}
+\newenvironment{keywords}{\paragraph{Keywords:}}{\bigskip}
 
-\setlength{\oddsidemargin}{0cm} \setlength{\evensidemargin}{0cm}
-\setlength{\textwidth}{16.5cm} \setlength{\topmargin}{-1cm}
-\setlength{\textheight}{24.5cm}
+\let\code=\texttt
+\let\pkg=\textbf
+\let\proglang=\textsf
+\newcommand{\class}[1]{`\code{#1}'}
 
@@
 
-\title{\textbf{partykit}: A Toolbox for Recursive Partytioning}
-
-\begin{center}
-  {\bf Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}
-\end{center}
+\title{\pkg{partykit}: A Toolbox for Recursive Partytioning}
+\author{Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}
 
 \begin{affiliations}
-1. Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen \par
-2. Department of Statistics and Mathematics, Wirtschaftsuniversit\"at Wien \par
-\texttt{\{Torsten.Hothorn, Achim.Zeileis\}@R-project.org}
+  \item Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen, Germany\\
+        \texttt{Torsten.Hothorn@R-project.org}
+  \item Department of Statistics and Mathematics, Wirtschaftsuniversit\"at Wien, Austria\\
+        \texttt{Achim.Zeileis@R-project.org}
 \end{affiliations}
 
-\keywords Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
+\begin{keywords}
+Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
+\end{keywords}
 
-\vskip 0.8cm
+Recursive partitioning methods, or simply ``trees'', are simple yet powerful
+methods for capturing regression relationships. Since the publication of the
+automated interaction detection (AID) algorithm in 1964, many extentions,
+modifications, and new approaches have been suggested in both the statistics
+and machine learning communities. Most of the standard algorithms are
+available to the \proglang{R} user, e.g., through packages
+\pkg{rpart} \citep{PKG:rpart},
+\pkg{RWeka} \citep{PKG:RWeka},
+\pkg{party} \citep{PKG:party}, or
+\pkg{mvpart} \citep{PKG:mvpart}.
 
-Recursive Partitioning methods, or simply `trees', are simple yet powerful
-methods to estimate regression relationships. Since the publication of the
-Automated Interaction Detection (AID) algorithm in 1964, many extentions,
-modifications and new approaches have been suggested in both the statistics
-and machine learning communities. Most of the algorithms in use today are
-available to the \textsf{R} user, for example from packages \textbf{rpart}
-\citep{PKG:rpart},
-\textbf{RWeka} \citep{PKG:RWeka}, \textbf{party} \citep{PKG:party}, or \textbf{mvpart}
-\citep{PKG:mvpart}.
+However, no common infrastructure is available for representing trees
+fitted by different packages. Consequently, the capabilities for extraction of
+information---such as predictions, printed summaries, or visualizations---vary
+between packages and come with somewhat different user interfaces.
+Furthermore, extensions or modifications often require considerable
+programming effort, e.g., if the median instead of the mean of a numerical
+response should be predicted in each leaf of an \class{rpart} tree.
+Similarly, implementations of new tree algorithms might also require new
+infrastructure if they have features not available in the above-mentioned
+packages, e.g., multi-way splits or more complex models in the leafs.
 
-Computations on fitted trees or the implementation of a new or modified
-recursive partitioning method, however, are relatively difficult tasks.
-As an example, one might want to predict a numeric response by the median
-instead of the mean based on an \textit{rpart} object. Or, instead of
-binary splits, one is going to implement an algorithm which potentially
-uses multiway splits, maybe containing a non-trivial model in each terminal
-node. These, and other, tasks require extensive programming,
-mainly because each package dealing with (some specialized form of)
-recursive partitioning uses a different internal representation of a tree. Thus, no
-standard user-interface exists, and methods for predictions, graphical or textual
-visualizations, or other computations heavily rely on these internals.
-
-To overcome these difficulties, the \textbf{partykit} package \citep{PKG:partykit} offers a 
-unified representation of tree objects along with \texttt{predict}, 
-\texttt{print}, and \texttt{plot} methods. Trees are represented through 
-a class \textit{party}, with building blocks \textit{partynode} and 
-\textit{partysplit}. The design of these classes covers splits in arbitrary
-functions of (one or more) variables, including multiway splits. In
-addition, non-constant fits in the terminal nodes are supported as well.
-Conversion methods exist for the most important sources of recursive
-partitioning objects, currently \textit{rpart}, \textit{J48} and
-\textit{pmmlTreeModel} objects can easily be converted into a \textit{party}
-object to compute with. The \textbf{partykit} also offers a re-implementation
-of conditional inference trees \cite{Hothorn:2006:JCGS}, illustrating the usefulness of the generic
-infrastructure.
+To overcome these difficulties, the \pkg{partykit} package \citep{PKG:partykit} offers a 
+unified representation of tree objects along with \code{predict()}, 
+\code{print()}, and \code{plot()} methods. Trees are represented through 
+a new flexible class \class{party} which can, in principle, capture
+all trees mentioned above but can also accomodate multi-way or functional
+splits, as well as complex models in (leaf) nodes. The package is currently
+under development at R-Forge but already provides conversion methods for
+trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
+as well as a re-implementation of conditional inference trees \citep{Hothorn:2006:JCGS}.
 
 In our presentation, we will only sketch details of these classes and 
-corresponding methods and focus
-on applications of the toolkit, for example for producing nice and
-non-standard plots
-for \textit{rpart} or \textit{J48} objects, for computing predictions
-on millions of new observations really fast, and for implementing CHAID, 
-a popular classification tree algorithm.
+corresponding methods and focus on applications of the toolkit including
+extended visualizations for \class{rpart} or \class{J48} objects, fast predictions
+on millions of new observations, and a new implementation of the classical CHAID
+algorithm.
 
-\bibliographystyle{plainnat}
+\bibliographystyle{jss}
 \bibliography{partykit}

### 2009-02-25 pkg/inst/slides/useR-2009_abstract.tex
renamed and added slightly modified abstract for DSC 2009
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@127 255f4249-2263-4417-b1ea-8498779e27a7

--- 2009-02-25 pkg/inst/slides/partykit_useR2009.tex
@@
 available to the \proglang{R} user, e.g., through packages
-\pkg{rpart} \citep{PKG:rpart},
-\pkg{RWeka} \citep{PKG:RWeka},
-\pkg{party} \citep{PKG:party}, or
-\pkg{mvpart} \citep{PKG:mvpart}.
+\pkg{rpart} \citep{partykit:rpart},
+\pkg{RWeka} \citep{partykit:RWeka},
+\pkg{party} \citep{partykit:party}, or
+\pkg{mvpart} \citep{partykit:mvpart}.
 
@@
 
-To overcome these difficulties, the \pkg{partykit} package \citep{PKG:partykit} offers a 
+To overcome these difficulties, the \pkg{partykit} package \citep{partykit:partykit} offers a 
 unified representation of tree objects along with \code{predict()}, 
@@
 trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
-as well as a re-implementation of conditional inference trees \citep{Hothorn:2006:JCGS}.
+as well as a re-implementation of conditional inference trees \citep{partykit:Hothorn+Hornik+Zeileis:2006}.
 

### 2009-02-25 pkg/inst/slides/DSC-2009_abstract.tex
new DSC abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@128 255f4249-2263-4417-b1ea-8498779e27a7

--- 2009-02-25 pkg/inst/slides/useR-2009_abstract.tex
@@
 \usepackage[authoryear,round]{natbib}
+\topmargin-1.5cm
+\textheight28cm
 
@@
 \begin{keywords}
-Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
+Regression Trees, Visualization, CTree, CHAID, MOB, Weka, PMML.
 \end{keywords}
@@
 \code{print()}, and \code{plot()} methods. Trees are represented through 
-a new flexible class \class{party} which can, in principle, capture
+a new flexible class \class{party} comprising essentially a recursive list of \class{partynode}
+objects along with \class{partysplit} information. Additionally, further
+information about the underlying data as well as the associated model fit
+can be stored and accessed. Therefore, the \class{party} class can capture
 all trees mentioned above but can also accomodate multi-way or functional
 splits, as well as complex models in (leaf) nodes. The package is currently
-under development at R-Forge but already provides conversion methods for
+under development at \proglang{R}-Forge but already provides conversion methods for
 trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
@@
 
-In our presentation, we will only sketch details of these classes and 
-corresponding methods and focus on applications of the toolkit including
-extended visualizations for \class{rpart} or \class{J48} objects, fast predictions
-on millions of new observations, and a new implementation of the classical CHAID
-algorithm.
+In our presentation, we discuss the design of the new \class{party} class
+and its methods and illustrate how it can be utilized to interface other classes such
+as \class{rpart} or \class{J48}, obtain new implementations of existing algorithms
+such as CHAID or re-implement techniques such as model-based recursive partitioning
+\citep[MOB,][]{partykit:Zeileis+Hothorn+Hornik:2008}.
 

### 2009-07-01 pkg/inst/slides/useR-2009_slides.Rnw
slides Rennes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@136 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[landscape]{slides}

%%\usepackage{Sweave}
\newenvironment{thebibliography}{}

\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<setup, echo = FALSE, results = hide>>=
set.seed(290875)
options(prompt = "R> ")
options(SweaveHooks = list(mai = function() par(mai = par("mai") * c(1, 2.4,
1, 1)),
                           mai3 = function() par(mai = par("mai") * c(1, 2,
1, 1))))
library("rpart")
library("ipred")
library("partykit")
library("RWeka")
library("CHAID")
library("rJava")
@

\usepackage[coloremph,colorhighlight,LMU,printout]{RSlides}
\usepackage{url,listings}
\usepackage{SweaveSlides}
\usepackage{natbib}
\usepackage{amssymb}

\usepackage[utf8]{inputenc}

\usepackage{hyperref}

\newcommand{\pkg}[1]{\textbf{#1}}

%%\input{header}
\usepackage{amstext}

\setkeys{Gin}{height = 0.75\textheight}

\title{\pkg{partykit}: \\ A Toolbox for Recursive Partytioning}

\author{Torsten Hothorn \\ Ludwig-Maximilians-Universit\"at M\"unchen \vspace*{1cm} \\ 
        Achim Zeileis \\ Wirtschaftsuniversit\"at Wien}
\fancyfoot[L]{\tiny 2009-07-09} %%\copyright{} 2007 Friedrich Leisch
\fancyfoot[R]{\arabic{page}}

\sloppy

\begin{document}

\maketitle

\newslide{Trees in R}

The Machine Learning task view lists the following tree-related packages
\begin{compactitem}
\item \pkg{rpart} (CART)
\item \pkg{tree} (CART)
\item \pkg{mvpart} (multivariate CART)
\item \pkg{knnTree} (nearest neighbor based trees)
\item \pkg{RWeka} (J48, M5$^\prime$, LMT)
\item \pkg{LogicReg} (Logic Regression)
\item \pkg{BayesTree} (with Bayesian flavor)
\item \pkg{TWIX} (with extra splits)
\item \pkg{party} (conditional inference trees, model-based trees)
\item ...
\end{compactitem}

%% \newslide{}

In addition, some packages implementing ensemble methods 
are based on trees as well, for example
\begin{compactitem}
\item \pkg{randomForest} (CART-based random forests)
\item \pkg{randomSurvivalForest} (for censored responses)
\item \pkg{party} (conditional random forests)
\item \pkg{gbm} (tree-based gradient boosting)
\item \pkg{mboost} (model-based and tree-based gradient boosting)
\item ...
\end{compactitem}

These packages deal with very similar problems.
However, similar or even the same tasks are implemented
differently by each package. No code reuse takes place.

\newslide{Missing in Action}

Moreover, important methods in this field, mostly from the statistics literature,
are missing from the above list, for example
\begin{compactitem}
\item CHAID
\item FACT
\item QUEST
\item GUIDE
\item LOTUS
\item C4.5
\item ...
\end{compactitem}

\newslide{Adding more Trees?}

It is hard to come up with implementations of new tree-based
algorithms, because one needs to take care not only
of fitting the tree but also of
\begin{itemize}
\item representing fitted trees,
\item printing trees,
\item plotting trees, and
\item computing predictions from trees.
\end{itemize}

\newslide{Wouldn't it be nice to...}

... have a common infrastructure for
\begin{itemize}
\item representing fitted trees,
\item printing trees,
\item plotting trees, and
\item computing predictions from trees?
\end{itemize}

\newslide{R-Forge Package \pkg{partykit}}

offers a unified representation of trees with methods for
\begin{itemize}
\item printing,
\item plotting,
\item computing predictions and
\item coercing from/to different formats.
\end{itemize}

\newslide{Unified Representation of Trees}

The most important classes in \pkg{partykit} are 
\begin{description}
\item[\texttt{partysplit}:] binary, multiway and functional splits in ordered and unordered variables,
\item[\texttt{partynode}:] inner nodes containing splits, surrogate splits, kid nodes, and 
                           terminal nodes containing models or constant fits, 
\item[\texttt{party}:] trees (nodes and data).
\end{description}

The API essentially consists of the corresponding constructors.

\newslide{Methods}

Coercing: \\
\texttt{as.party.rpart}, \texttt{as.party.J48}, \texttt{as.party.pmmlTreeModel}

Inspecting: \\
\texttt{print.party}, \texttt{plot.party}, both are extensible

Computing: \\
\texttt{predict.party}, subset methods

\newslide{Example: Pretty rpart Trees}

<<ex-rpart, echo = TRUE>>=
data("GlaucomaM", package = "ipred")
g_rpart <- rpart(Class ~ ., data = GlaucomaM)
print(g_rpart)
@

\newpage

\begin{center}
<<ex-rpart-plot, echo = TRUE, fig = TRUE>>=
plot(g_rpart)
text(g_rpart)
@

\newpage

{\small
<<ex-rpart-partykit, echo = TRUE>>=
g_party <- as.party(g_rpart)
print(g_party, header = FALSE)
@
}

\newpage

<<ex-rpart-partykit-plot1, echo = TRUE, fig = TRUE, width = 8>>=
plot(g_party, type = "simple")
@

\newpage
<<ex-rpart-partykit-plot2, echo = TRUE, fig = TRUE, width = 10, height = 8>>=
plot(g_party, type = "extended")
@
\end{center}

\newslide{Example: Fast Predictions}

<<ex-J48-partykit, echo = TRUE>>=
g_J48 <- J48(Class ~ ., data = GlaucomaM)
g_party <- as.party(g_J48)
nd <- GlaucomaM[rep(1:nrow(GlaucomaM), 100),]
system.time(p1 <- predict(g_J48, newdata = nd))
system.time(p2 <- predict(g_party, newdata = nd))
table(p1, p2)
@

\newslide{Example: CHAID}

\pkg{partykit} makes it `easy' to implement new or old tree algorithms.

My students implemented CHAID as a software project last winter. I wanted
them to focus on the algorithm, not on technical details.

We only had to implement variable selection, splitting and stopping rules,
\pkg{partykit} takes care of the rest.

The resulting package \pkg{CHAID} is available from R-forge and
consists of only 362 lines of R code and was validated against SPSS 15.

\newpage

Fit a CHAID tree to BreastCancer data from package \pkg{mlbench}:

\small{
<<ex-CHAID, echo = TRUE>>=
data("BreastCancer", package = "mlbench")
b_chaid <- chaid(Class ~ Cl.thickness + Cell.size + Cell.shape + 
                  Marg.adhesion + Epith.c.size + Bare.nuclei + 
                  Bl.cromatin + Normal.nucleoli + Mitoses, 
                  data = BreastCancer)
@
}

\begin{center}
<<ex-CHAID-plot, echo = TRUE, fig = TRUE, width = 17, height = 8>>=               
plot(b_chaid)
@
\end{center}

\newpage

Benchmark experiment CHAID vs. CTREE\\
(Bachelor thesis by Stefanie Thiemichen)
\begin{center}
\includegraphics{ratiobox}
\end{center}

\newslide{Example: PMML \& SPSS}

Toy PMML example:
<<ex-xml, echo = FALSE>>=
writeLines(c(readLines("pmml.xml")[1:16], "...\n"))
@

\begin{center}
<<ex-pmml, echo = TRUE, fig = TRUE>>=
pm <- pmmlTreeModel("pmml.xml")
plot(pm)
@
\end{center}

\newpage

<<ex-pmml-predict, echo = TRUE>>=
nd <- data.frame(nutzung = gl(3, 10), beruf = sample(gl(3, 10)))
predict(pm, newdata = nd)
@

\newslide{What's next?}

Once a stable version of \pkg{partykit} is available from CRAN, we hope that
useRs will start
\begin{itemize}
\item adding new tree algorithms,
\item adding coercing methods from/to other packages,
\item adding new plot functionality, and generally
\item will use our kit to party harder.
\end{itemize}

\end{document}

### 2009-07-02 pkg/inst/slides/useR-2009_slides.Rnw
Z comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@137 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \author{Torsten Hothorn \\ Ludwig-Maximilians-Universit\"at M\"unchen \vspace*{1cm} \\ 
-        Achim Zeileis \\ Wirtschaftsuniversit\"at Wien}
+        Achim Zeileis \\ WU Wirtschaftsuniversit\"at Wien}
 \fancyfoot[L]{\tiny 2009-07-09} %%\copyright{} 2007 Friedrich Leisch
@@
 \item \pkg{mvpart} (multivariate CART)
-\item \pkg{knnTree} (nearest neighbor based trees)
-\item \pkg{RWeka} (J48, M5$^\prime$, LMT)
+\item \pkg{knnTree} (nearest-neighbor-based trees)
+\item \pkg{RWeka} (J4.8, M5$^\prime$, LMT)
 \item \pkg{LogicReg} (Logic Regression)
@@
 \item LOTUS
-\item C4.5
+\item CRUISE
 \item ...
@@
 
+{\color{white} hansi \\}
 ... have a common infrastructure for
@@
 
-offers a unified representation of trees with methods for
+{\color{white} hansi \\}
+offers a unified approach to 
 \begin{itemize}
-\item printing,
-\item plotting,
-\item computing predictions and
-\item coercing from/to different formats.
+\item representing fitted trees,
+\item printing trees,
+\item plotting trees, and
+\item computing predictions from trees.
 \end{itemize}
@@
 data("GlaucomaM", package = "ipred")
+library("rpart")
 g_rpart <- rpart(Class ~ ., data = GlaucomaM)
@@
 <<ex-rpart-partykit, echo = TRUE>>=
+library("partykit")
 g_party <- as.party(g_rpart)
@@
 
-The resulting package \pkg{CHAID} is available from R-forge and
+The resulting package \pkg{CHAID} is available from R-Forge and
 consists of only 362 lines of R code and was validated against SPSS 15.
@@
 data("BreastCancer", package = "mlbench")
+library("CHAID")
 b_chaid <- chaid(Class ~ Cl.thickness + Cell.size + Cell.shape + 
@@
 
-Benchmark experiment CHAID vs. CTREE\\
+Benchmark experiment misclassification error CHAID vs.~CTREE\\
 (Bachelor thesis by Stefanie Thiemichen)
@@
 <<ex-pmml-predict, echo = TRUE>>=
-nd <- data.frame(nutzung = gl(3, 10), beruf = sample(gl(3, 10)))
+nd <- data.frame(X1 = gl(3, 10), X2 = sample(gl(3, 10)))
 predict(pm, newdata = nd)
@@
 \item adding new tree algorithms,
+\item adding new ensemble methods,
 \item adding coercing methods from/to other packages,

### 2010-04-21 pkg/R/ctree.R
mvtnorm is only suggested
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@161 255f4249-2263-4417-b1ea-8498779e27a7

@@ -211,6 +211,7 @@ ctree_control <- function(teststat = c("quad", "max"),
     multiway = FALSE, splittry = 2L) {
 
     teststat <- match.arg(teststat)
+    if (teststat == "max") stopifnot(require("mvtnorm"))
     testtype <- match.arg(testtype)
     list(teststat = teststat,
          testtype = testtype, mincriterion = log(mincriterion),

### 2010-04-21 pkg/inst/doc/introduction.Rnw
start vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@162 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}

%\VignetteIndexEntry{partykit: A Toolkit for Recursive Partytioning}
%\VignetteDepends{partykit}
%\VignetteKeywords{recursive partitioning}
%\VignettePackage{partykit}

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
%% need no \usepackage{Sweave}

%% commands
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

\hyphenation{Qua-dra-tic}

\title{\pkg{partykit}: A Toolkit for Recursive Partytioning}
\Plaintitle{partykit: A Toolkit for Recursive Partytioning}

\author{Torsten Hothorn\\Ludwig-Maximilians-\\Universit\"at M\"unchen
   \And Achim Zeileis\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

\Abstract{
  The \pkg{partykit} package 
}
\Keywords{recursive partitioning}

\Address{
  Torsten Hothorn\\
  Institut f\"ur Statistik\\
  Ludwig-Maximilians-Universit\"at M\"unchen\\
  Ludwigstra{\ss}e 33\\
  80539 M\"unchen, Germany\\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://www.stat.uni-muenchen.de/~hothorn/}\\

  Achim Zeileis\\
  Institut für Statistik \\
  Fakultät für Volkswirtschaft und Statistik \\
  Leopold-Franzens-Universität Innsbruck \\
  Universitätsstraße 15 \\
  6020 Innsbruck, Österreich
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://statmath.wu.ac.at/~zeileis/}
}

\begin{document}

<<setup, echo = FALSE, results = hide>>=
options(width = 70, SweaveHooks = list(leftpar = 
    function() par(mai = par("mai") * c(1, 1.1, 1, 1))))
require("partykit")
set.seed(290875)
data("iris")
@

\setkeys{Gin}{width=\textwidth}

\section{Introduction}

\section{Splits}

A split is basically a function that maps data, 
more specifically a partitioning variable, to child nodes. 
Objects of class \code{partysplit} are designed to represent 
such functions and are set-up by the \code{partysplit()} 
constructor:
<<partysplit-1, echo = TRUE>>=
## binary split in numeric variable `Sepal.Length'
sl5 <- partysplit(which(names(iris) == "Sepal.Length"),
    breaks = 5)
class(sl5)
@
The internal structure of class \code{partysplit} contains
information about the partitioning variable, the split-points,
the handling of split-points, the treatment of observations with
missing values and the daughter nodes to send observations to:
<<partysplit-2, echo = TRUE>>=
unclass(sl5)
@
Here, the split is defined in the first variable (corresponds to
\code{Sepal.Length} in data frame \code{iris}) and the splitting rule
is \code{Sepal.Length} $\le 5$:
<<partysplit-3, echo = TRUE>>=
character_split(sl5, data = iris)
@
This representation of splits is completely abstract and, most importantly,
independent of any data. Now, data comes into play when we actually
want to perform splits:
<<partysplit-4, echo = TRUE>>= 
kidids_split(sl5, data = iris)
@
For each observation in \code{iris} the split is performed and the 
number of the daughter node to send this observation to
is returned. Of course, this is a very complicated way of saying
<<partysplit-5, echo = TRUE>>=
(!with(iris, Sepal.Length <= 5)) + 1
@

However, splits may become more complex. Consider a split in a categorical
variable at three levels where the first two levels go to the left
daugther node and the third one to the right daugther node:
<<partysplit-6, echo = TRUE>>=
## binary split in factor `Species'
sp <- partysplit(which(names(iris) == "Species"),
    index = c(1L, 1L, 2L))
character_split(sp, data = iris)
table(kidids_split(sp, data = iris), iris$Species)
@
The internal structure of this object now contains an index slot
<<partysplit-6, echo = TRUE>>=
unclass(sp)
@
that maps levels to daughter nodes. This mapping is also
useful with splits in ordered variables, for example when 
representing multiway splits:
<<partysplit-7, echo = TRUE>>=
## multiway split in numeric variable `Sepal.Width',    
## higher values go to the first kid, smallest values
## to the last kid
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),       
    breaks = c(3, 3.5), index = 3:1)    
character_split(sw23, data = iris)
table(kidids_split(sw23, data = iris),
    cut(iris$Sepal.Width, breaks = c(-Inf, 2, 3, Inf)))
@
The mapping of classes of the categorized numeric variable
to daugther nodes can be changed by modifying \code{index}:
<<partysplit-8, echo = TRUE>>=
sw23 <- partysplit(which(names(iris) == "Sepal.Width"),
    breaks = c(3, 3.5), index = c(1L, 3L, 2L))
character_split(sw23, data = iris)
@

The additional argument \code{prop} is used to specify a 
discrete probability distribution over the daugther nodes that
is used to map observations with missing values to daugther nodes.
Furthermore, the \code{info} argument and slot takes arbitrary objects
to be stored with the split (for example split statistics) but is not
structured at the moment.

The slots of \code{partysplit} objects shall be accessed by the corresponding
accessor functions.

\section{Nodes}

Inner and terminal nodes are represented by objects of class \code{partynode}.
Each node has a unique identifier \code{id}. A node consisting only
of such an identifier (and possibly additional information  
in \code{info}) is a terminal node:
<<partynode-1, echo = TRUE>>=
n1 <- partynode(id = 1L)
is.terminal(n1)
print(n1)
@
Inner nodes have to have a primary split \code{split} and at least two
daugther nodes. The daugther nodes are objects of class \code{partynode}
itself and thus represent the recursive nature of this data structure.
The daugther nodes are pooled in a list \code{kids}. 
In addition, a list of \code{partysplit} objects offering 
surrogate splits can be supplied; a list of \code{partysplit} objects
in slot \code{surrogates} defines such additional splits (mostly used 
for handling missing values). Note that \code{partynode} objects aren't 
connected to the actual data.

Based on the binary split \code{sl5} defined in the previous section, 
we set-up an inner node with two terminal daugther nodes and
print this stump (the data is needed because neither split nor nodes
contain information about variable names or levels):
<<partynode-2, echo = TRUE>>=
n1 <- partynode(id = 1L, split = sl5, kids = sapply(2:3, partynode))
print(n1, data = iris)
@
Now that we have defined our first simple tree, we want to assign
observations to terminal nodes:
<<partynode-3, echo = TRUE>>=
fitted_node(n1, data = iris)
@
Here, the \code{id}s of the terminal node each observations falls into
are returned. Alternatively, we could compute the position of these
daugther nodes in the list \code{kids}:
<<partynode-4, echo = TRUE>>=
kidids_node(n1, data = iris)
@
Furthermore, the \code{info} argument and slot takes arbitrary objects
to be stored with the node (predictions, for example, but we will handle
this issue later). The slots can be extracted by means of the corresponding
accessor functions.

A number of methods are defined for \code{partynode} objects.
\code{is.partynode} checks if the argument is a valid \code{partynode}
object. \code{is.terminal} is \code{TRUE} for terminal nodes
and \code{FALSE} for inner nodes. The subset methods
return the \code{partynode} object corresponding to the \code{i}th
kid:
<<partynode-5, echo = TRUE>>=
n1[2]
@

The \code{as.partynode} and \code{as.list} methods can be used
to convert flat list structures into recursive \code{partynode}
objects and vice versa. \code{as.partynode} applied to
\code{partynode} objects renumbers the recursive nodes
starting with root node identifier \code{from}.

\code{length} gives the number of kid nodes of the root node,
\code{depth} the depth of the tree and \code{width}
the number of terminal nodes.

\section{Trees}

Although tree structures can be represented by \code{partynode} objects,
a tree is more than a number of nodes and splits. More
information about (parts of the) corresponding data is necessary for 
high-level computations on trees.

Objects of class \code{party} basically consist of a \code{partynode}
object representing the tree structure in a recursive way and 
data. The \code{data} argument takes a \code{data.frame} which, however,
might have zero columns. Optionally, a \code{data.frame} with at least one
variable \code{(fitted)} containing the terminal node numbers of
data used for fitting the tree may be specified along with a   
\code{terms} object or any additional (currently unstructured)
information as \code{info}. Argument \code{names} defines names
for all nodes in \code{node}.

<<party-1, echo = TRUE>>=
t1 <- party(n1, 
   data = data.frame(iris, 
       "(fitted)" = fitted_node(n1, data = iris),
       "(response)" = iris$Species))
@

\end{document}

### 2010-04-22 pkg/inst/doc/introduction.Rnw
tree example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@163 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\author{Torsten Hothorn\\Ludwig-Maximilians-\\Universit\"at M\"unchen
-   \And Achim Zeileis\\Wirtschaftsuniversit\"at Wien}
+\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at M\"unchen
+   \And Achim Zeileis\\Leopold-Franzens-Universität Innsbruck}
 \Plainauthor{Torsten Hothorn, Achim Zeileis}
@@
   Achim Zeileis\\
-  Institut für Statistik \\
-  Fakultät für Volkswirtschaft und Statistik \\
-  Leopold-Franzens-Universität Innsbruck \\
-  Universitätsstraße 15 \\
-  6020 Innsbruck, Österreich
+  Institut f\"ur Statistik \\
+  Fakult\"at f\"ur Volkswirtschaft und Statistik \\
+  Leopold-Franzens-Universit\"at Innsbruck \\
+  Universit\"atsstra{\ss}e 15 \\
+  6020 Innsbruck, \"Osterreich \\
   E-mail: \email{Achim.Zeileis@R-project.org}\\
@@
 \begin{document}
+
+\SweaveOpts{eps=FALSE, keep.source=TRUE, eval = TRUE}
 
@@
 
+\section{My first Tree}
+
+Package \pkg{partykit} does not offer unified infrastructure for growing trees. However,
+once you know how to estimate splits from data, it is fairly straightforward to
+implement trees. Consider a very simple tree algorithm. We assume that both response
+and features are numeric. We search for the binary best split by means of $t$-test 
+$p$-values, i.e., we cycle through all variables and potential split points and assess
+the quality of the split by comparing the distributions of the response in the so-defined
+two groups. We select the feature/split point combination with lowest two-sided $p$-value,
+however only if this result is significant at level $\alpha = 0.05$.
+
+This strategy can be implemented based on the data (response and features) and 
+some case weights as follows:
+<<mytree-1, echo = TRUE>>=
+findsplit <- function(response, data, weights) {
+
+    ### extract response values from data
+    y <- data[[response]]
+
+    logpmin <- 0
+    xselect <- NULL
+
+    ### cycle through all features
+    for (i in which(names(data) != response)) {
+
+        ### expand data
+        x <- data[[i]]
+        xt <- rep(x, weights)
+        yt <- rep(y, weights)
+
+        ### potential split points (not too many)
+        qx <- unique(quantile(xt, 
+                     prob = seq(from = 0.1, to = 0.9, by = 0.05)))
+
+        ### assess all potential splits by their t-test
+        ### log-p-value
+        logp <- sapply(qx, function(q) {
+            tt <- t.test(yt[xt <= q], yt[xt > q])
+            pt(-abs(tt$statistic), tt$parameter, log = TRUE) + log(2)
+        })
+
+        ### if the best split in variable i significant AND
+        ### better than what we already had store this information
+        if (min(logp) < logpmin & min(logp) < log(0.05)) {
+            logpmin <- min(logp)
+            xselect <- i
+            splitpoint <- qx[which.min(logp)]
+        }
+    }
+
+    ### no significant split found, give up
+    if (is.null(xselect)) return(NULL)
+
+    ### return split as partysplit object
+    return(partysplit(
+        varid = as.integer(xselect),       ### which variable?
+        breaks = as.numeric(splitpoint),   ### which split point?
+        info = list(pvalue = exp(logpmin)  ### save p-value in addition
+    )))
+}
+@
+
+Now we have to set-up the recursion for growing a recursive 
+\code{partynode} structure:
+<<mytree-2, echo = TRUE>>=
+growtree <- function(id = 1L, response, data, weights) {
+
+    ### for less than 30 obs. stop here
+    if (sum(weights) < 30) return(partynode(id = id))
+
+    ### find best split
+    sp <- findsplit(response, data, weights)
+    ### no split found, stop here
+    if (is.null(sp)) return(partynode(id = id))
+
+    ### actually split the data
+    kidids <- kidids_split(sp, data = data)
+
+    ### set-up all daugther nodes
+    kids <- vector(mode = "list", length = max(kidids))
+    for (kidid in 1:max(kidids)) {
+          ### select obs for current node
+          w <- weights
+          w[kidids != kidid] <- 0
+          ### get next node id
+          if (kidid > 1) {
+              myid <- max(nodeids(kids[[kidid - 1]]))
+          } else {
+              myid <- id
+          }
+          ### start recursion on this daugther node
+          kids[[kidid]] <- growtree(id = as.integer(myid + 1), 
+              response, data, w)
+    }
+
+    ### return nodes
+    return(partynode(id = as.integer(id), 
+                     split = sp, kids = kids))
+}
+@
+
+<<mytree-3, echo = TRUE>>=
+mytree <- function(formula, data, weights = NULL) {
+
+    ### name of the response variable
+    response <- all.vars(formula)[1]
+    ### data without missing values, response comes last
+    data <- data[complete.cases(data), 
+                 c(all.vars(formula)[-1], response)]
+    if (is.null(weights)) weights <- rep(1, nrow(data))
+
+    ### grow tree
+    nodes <- growtree(id = 1L, response, data, weights)
+
+    ### compute terminal node number for each obs.
+    fitted <- fitted_node(nodes, data = data)
+    ### return rich object
+    ret <- party(nodes, 
+        data = data,
+        fitted = data.frame("(fitted)" = fitted,
+                            "(response)" = data[[response]],
+                            "(weights)" = weights,
+                            check.names = FALSE),
+        terms = terms(formula))
+    as.constparty(ret)
+}
+@
+
+<<mytree-4, echo = TRUE>>=
+aqt <- mytree(Ozone ~ Solar.R + Wind + Temp, data = airquality)
+aqt
+@
+
+\begin{figure}
+\begin{center}
+<<mytree-5, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
+plot(aqt)
+@
+\caption{Tree}
+\end{center}
+\end{figure}
+
+<<mytree-6, echo = TRUE>>=
+predict(aqt, newdata = airquality[1:10,])
+@
+
+<<mytree-7, echo = TRUE>>=
+aqt4 <- aqt[4]
+aqt4
+@
+
+\begin{figure}
+\begin{center}
+<<mytree-8, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
+plot(aqt4)
+@
+\caption{Subtree}
+\end{center}
+\end{figure}
+
+<<mytree-9, echo = TRUE>>=
+info_split(split_node(node_party(aqt4)))
+@
+
 \end{document}

### 2010-04-22 pkg/inst/doc/introduction.Rnw
playing with trees
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@164 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+A very rough sketch of formula-based user-interface needs
+to set-up the data and call \code{growtree()}:
 <<mytree-3, echo = TRUE>>=
@@
 
+We now can fit this tree, for example to the airquality data; the
+\code{print()} method provides us with a first overview on the 
+resulting model
 <<mytree-4, echo = TRUE>>=
@@
 @
-\caption{Tree}
+\caption{Tree. \label{plottree}}
 \end{center}
@@
 
+We depict the model graphically using \code{plot()} (see Figure~\ref{plottree})
+and compute predictions using
 <<mytree-6, echo = TRUE>>=
@@
 
+An interesting feature is the ability to extract subsets of trees:
 <<mytree-7, echo = TRUE>>=
@@
 @
+which again are objects inheriting from \code{party} and thus
+can be plotted easily (see Figure~\ref{subtree}).
 
@@
 @
-\caption{Subtree}
+\caption{Subtree. \label{subtree}}
 \end{center}
@@
 
-<<mytree-9, echo = TRUE>>=
-info_split(split_node(node_party(aqt4)))
+We also might be interested in extracting the $p$-values in the
+inner nodes in a nicely formatted way:
+<<mytree-10, echo = TRUE>>=
+fun <- function(x) 
+    format.pval(info_split(split_node(x))$pvalue, 
+                digits = 3, eps = 0.001)
+nid <- nodeids(aqt)
+iid <- nid[!(nid %in% nodeids(aqt, terminal = TRUE))]
+unlist(nodeapply(aqt, ids = iid, FUN = fun))
 @

### 2010-04-22 pkg/inst/doc/introduction.Rnw
some checks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@165 255f4249-2263-4417-b1ea-8498779e27a7

@@
 t1 <- party(n1, 
-   data = data.frame(iris, 
-       "(fitted)" = fitted_node(n1, data = iris),
-       "(response)" = iris$Species))
+   data = iris,
+   fitted = data.frame("(fitted)" = fitted_node(n1, data = iris),
+                       "(response)" = iris$Species,
+                       check.names = FALSE))
+t1
 @
@@
 This strategy can be implemented based on the data (response and features) and 
-some case weights as follows:
+some case weights as follows (\code{response} is just the name of the response
+and \code{data} is a data frame with all variables):
 <<mytree-1, echo = TRUE>>=
@@
 
-Now we have to set-up the recursion for growing a recursive 
+In order to actually grow a tree on data, 
+we have to set-up the recursion for growing a recursive 
 \code{partynode} structure:
@@
                  c(all.vars(formula)[-1], response)]
+    ### data is numeric
+    stopifnot(all(sapply(data, is.numeric)))
+
     if (is.null(weights)) weights <- rep(1, nrow(data))
+    ### weights are case weights, i.e., integers
+    stopifnot(length(weights) == nrow(data) &
+              max(abs(weights - floor(weights))) < .Machine$double.eps)
 

### 2010-04-23 pkg/inst/doc/introduction.Rnw
splits more formally
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@166 255f4249-2263-4417-b1ea-8498779e27a7

@@
 accessor functions.
+
+Formally, a split is a function $f$ mapping an element $x = (x_1, \dots, x_p)$ of a $p$-dimensional 
+sample space $\mathcal{X}$ into a set of $k$ daugther nodes $\mathcal{D} = \{d_1, \dots, d_k\}$.
+This mapping is defined as a composition $f = h \circ g$ of two functions $g: \mathcal{X} \rightarrow \mathcal{I}$
+and $h: \mathcal{I} \rightarrow \mathcal{D}$ with index set $\mathcal{I} = \{1, \dots, l\}, l \ge k$.
+
+Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
+($(\mu_1, \dots, \mu_{l - 1})$ = \code{break}). We are interested to split according to the 
+information contained in the $i$th element of $x$ ($i$ = \code{varid}).
+For numeric $x_i$, the split points are also numeric. If $x_i$ is a factor at
+levels $1, \dots, K$, the default split points are $\mu = (-\infty, 1, \dots, K - 1, \infty)$.
+
+The function $g$ essentially determines, which of the intervals (defined by $\mu$) the value
+$x_i$ is contained in ($I$ denotes the indicator function here):
+\begin{eqnarray*}
+x \mapsto g(x) = \sum_{j = 1}^l j I_{A(j)}(x_i)
+\end{eqnarray*}
+where $A(j) = (\mu_{j - 1}, \mu_j]$ for \code{right = TRUE} except $A(l) = (\mu_{l - 1}, \infty)$.
+If \code{right = FALSE}, then $A(j) = [\mu_{j - 1}, \mu_j)$ except 
+$A(1) = (-\infty, \mu_1)$. Note that with some categorical variable $x_i$ and default
+split points, $g$ is the identity.
+
+Now, $h$ maps from the index set $\mathcal{I}$ into the set of daugther nodes:
+\begin{eqnarray*}
+f(x) = h(g(x)) = d_{\sigma_{g(x)}}
+\end{eqnarray*}
+where $\sigma = (\sigma_1, \dots, \sigma_l) \in \{1, \dots, k\}^l$ (\code{index}). 
+By default, $\sigma = (1, \dots, l)$ and $k = l$.
+
+If $x_i$ is missing, then $f(x)$ is randomly drawn with $\mathbb{P}(f(x) = d_j) = p_j, j = 1, \dots, k$
+for a discrete probability distribution $p = (p_1, \dots, p_k)$ over the $k$ daugther nodes (\code{prob}).
+
+In the simplest case of a binary split in a numeric variable $x_i$, there is only
+one split point $\mu_1$ and, with $\sigma = (1, 2)$, observations with $x_i \le \mu_1$ are sent
+to daugther node $d_1$ and observations with $x_i > \mu_1$ to $d_2$.
 

### 2010-04-23 pkg/inst/doc/introduction.Rnw
splits more formally
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@167 255f4249-2263-4417-b1ea-8498779e27a7
cosmetics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@168 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at M\"unchen
-   \And Achim Zeileis\\Leopold-Franzens-Universität Innsbruck}
+\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
+   \And Achim Zeileis\\Leopold-Franzens-Universit\"at \\ Innsbruck}
 \Plainauthor{Torsten Hothorn, Achim Zeileis}
@@
 A split is basically a function that maps data, 
-more specifically a partitioning variable, to child nodes. 
+more specifically a partitioning variable, to daugther nodes. 
 Objects of class \code{partysplit} are designed to represent 
@@
 
-However, splits may become more complex. Consider a split in a categorical
+Formally, a split is a function $f$ mapping an element $x = (x_1, \dots, x_p)$ of a $p$-dimensional 
+sample space $\mathcal{X}$ into a set of $k$ daugther nodes $\mathcal{D} = \{d_1, \dots, d_k\}$.
+This mapping is defined as a composition $f = h \circ g$ of two functions $g: \mathcal{X} \rightarrow \mathcal{I}$
+and $h: \mathcal{I} \rightarrow \mathcal{D}$ with index set $\mathcal{I} = \{1, \dots, l\}, l \ge k$.
+
+Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
+($(\mu_1, \dots, \mu_{l - 1})$ = \code{break}). We are interested to split according to the 
+information contained in the $i$th element of $x$ ($i$ = \code{varid}).
+For numeric $x_i$, the split points are also numeric. If $x_i$ is a factor at
+levels $1, \dots, K$, the default split points are $\mu = (-\infty, 1, \dots, K - 1, \infty)$.
+
+The function $g$ essentially determines, which of the intervals (defined by $\mu$) the value
+$x_i$ is contained in ($I$ denotes the indicator function here):
+\begin{eqnarray*}
+x \mapsto g(x) = \sum_{j = 1}^l j I_{\mathcal{A}(j)}(x_i)
+\end{eqnarray*}
+where $\mathcal{A}(j) = (\mu_{j - 1}, \mu_j]$ for \code{right = TRUE} except 
+$\mathcal{A}(l) = (\mu_{l - 1}, \infty)$.
+If \code{right = FALSE}, then $\mathcal{A}(j) = [\mu_{j - 1}, \mu_j)$ except 
+$\mathcal{A}(1) = (-\infty, \mu_1)$. Note that with some categorical variable $x_i$ and default
+split points, $g$ is the identity.
+
+Now, $h$ maps from the index set $\mathcal{I}$ into the set of daugther nodes:
+\begin{eqnarray*}
+f(x) = h(g(x)) = d_{\sigma_{g(x)}}
+\end{eqnarray*}
+where $\sigma = (\sigma_1, \dots, \sigma_l) \in \{1, \dots, k\}^l$ (\code{index}). 
+By default, $\sigma = (1, \dots, l)$ and $k = l$.
+
+If $x_i$ is missing, then $f(x)$ is randomly drawn with $\mathbb{P}(f(x) = d_j) = p_j, j = 1, \dots, k$
+for a discrete probability distribution $p = (p_1, \dots, p_k)$ over the $k$ daugther nodes (\code{prob}).
+
+In the simplest case of a binary split in a numeric variable $x_i$, there is only
+one split point $\mu_1$ and, with $\sigma = (1, 2)$, observations with $x_i \le \mu_1$ are sent
+to daugther node $d_1$ and observations with $x_i > \mu_1$ to $d_2$. However,
+this representation of splits is general enough to deal with more complicated
+set-ups like surrogate splits, where typically the index needs modification, for example
+$\sigma = (2, 1)$, categorical splits, i.e., there is one data structure for both
+ordered and unordered splits, multiway splits, and functional splits. The latter
+can be implemented by defining a new artificial splitting variable $x_{p + 1}$ by
+means of a potentially very complex function of $x$ later used for splitting.
+
+As an example, consider a split in a categorical
 variable at three levels where the first two levels go to the left
@@
 @
-The internal structure of this object now contains an index slot
+The internal structure of this object contains the \code{index} slot
 <<partysplit-6, echo = TRUE>>=
@@
 accessor functions.
-
-Formally, a split is a function $f$ mapping an element $x = (x_1, \dots, x_p)$ of a $p$-dimensional 
-sample space $\mathcal{X}$ into a set of $k$ daugther nodes $\mathcal{D} = \{d_1, \dots, d_k\}$.
-This mapping is defined as a composition $f = h \circ g$ of two functions $g: \mathcal{X} \rightarrow \mathcal{I}$
-and $h: \mathcal{I} \rightarrow \mathcal{D}$ with index set $\mathcal{I} = \{1, \dots, l\}, l \ge k$.
-
-Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
-($(\mu_1, \dots, \mu_{l - 1})$ = \code{break}). We are interested to split according to the 
-information contained in the $i$th element of $x$ ($i$ = \code{varid}).
-For numeric $x_i$, the split points are also numeric. If $x_i$ is a factor at
-levels $1, \dots, K$, the default split points are $\mu = (-\infty, 1, \dots, K - 1, \infty)$.
-
-The function $g$ essentially determines, which of the intervals (defined by $\mu$) the value
-$x_i$ is contained in ($I$ denotes the indicator function here):
-\begin{eqnarray*}
-x \mapsto g(x) = \sum_{j = 1}^l j I_{A(j)}(x_i)
-\end{eqnarray*}
-where $A(j) = (\mu_{j - 1}, \mu_j]$ for \code{right = TRUE} except $A(l) = (\mu_{l - 1}, \infty)$.
-If \code{right = FALSE}, then $A(j) = [\mu_{j - 1}, \mu_j)$ except 
-$A(1) = (-\infty, \mu_1)$. Note that with some categorical variable $x_i$ and default
-split points, $g$ is the identity.
-
-Now, $h$ maps from the index set $\mathcal{I}$ into the set of daugther nodes:
-\begin{eqnarray*}
-f(x) = h(g(x)) = d_{\sigma_{g(x)}}
-\end{eqnarray*}
-where $\sigma = (\sigma_1, \dots, \sigma_l) \in \{1, \dots, k\}^l$ (\code{index}). 
-By default, $\sigma = (1, \dots, l)$ and $k = l$.
-
-If $x_i$ is missing, then $f(x)$ is randomly drawn with $\mathbb{P}(f(x) = d_j) = p_j, j = 1, \dots, k$
-for a discrete probability distribution $p = (p_1, \dots, p_k)$ over the $k$ daugther nodes (\code{prob}).
-
-In the simplest case of a binary split in a numeric variable $x_i$, there is only
-one split point $\mu_1$ and, with $\sigma = (1, 2)$, observations with $x_i \le \mu_1$ are sent
-to daugther node $d_1$ and observations with $x_i > \mu_1$ to $d_2$.
 

### 2010-04-30 pkg/inst/doc/introduction.Rnw
Z comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@170 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \section{Introduction}
+
+### Aufbau: erstmal Baum zusammenbauen, den plotten (ohne code) und diesen
+### Baum dann Schritt fuer Schritt zusammensetzen vielleicht:
+
+<<weather-data, echo = FALSE>>=
+weather <-
+structure(list(outlook = structure(c(1L, 1L, 2L, 3L, 3L, 3L, 
+2L, 1L, 1L, 3L, 1L, 2L, 2L, 3L), .Label = c("sunny", "overcast", 
+"rainy"), class = "factor"), temperature = c(85, 80, 83, 70, 
+68, 65, 64, 72, 69, 75, 75, 72, 81, 71), humidity = c(85, 90, 
+86, 96, 80, 70, 65, 95, 70, 80, 70, 90, 75, 91), windy = c(FALSE, 
+TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE, TRUE, 
+TRUE, FALSE, TRUE), play = structure(c(2L, 2L, 1L, 1L, 1L, 2L, 
+1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L), .Label = c("yes", "no"), class = "factor")), .Names = c("outlook", 
+"temperature", "humidity", "windy", "play"), row.names = c(NA, 
+-14L), class = "data.frame")
+@
+### response: play siehe Witten/Frank im Weka Buch
 
@@
 \section{Trees}
+
+### code in eine Tabelle
 

### 2010-05-26 pkg/inst/doc/introduction.Rnw
LaTeX comments %% have to be used in LaTeX mode (not R comments ##)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@171 255f4249-2263-4417-b1ea-8498779e27a7

--- 2010-04-23 pkg/inst/doc/introduction.Rnw
@@
 \section{Introduction}
+
+%%% Aufbau: erstmal Baum zusammenbauen, den plotten (ohne code) und diesen
+%%% Baum dann Schritt fuer Schritt zusammensetzen vielleicht:
+
+<<weather-data, echo = FALSE>>=
+weather <-
+structure(list(outlook = structure(c(1L, 1L, 2L, 3L, 3L, 3L, 
+2L, 1L, 1L, 3L, 1L, 2L, 2L, 3L), .Label = c("sunny", "overcast", 
+"rainy"), class = "factor"), temperature = c(85, 80, 83, 70, 
+68, 65, 64, 72, 69, 75, 75, 72, 81, 71), humidity = c(85, 90, 
+86, 96, 80, 70, 65, 95, 70, 80, 70, 90, 75, 91), windy = c(FALSE, 
+TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE, TRUE, 
+TRUE, FALSE, TRUE), play = structure(c(2L, 2L, 1L, 1L, 1L, 2L, 
+1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L), .Label = c("yes", "no"), class = "factor")), .Names = c("outlook", 
+"temperature", "humidity", "windy", "play"), row.names = c(NA, 
+-14L), class = "data.frame")
+@
+%%% response: play siehe Witten/Frank im Weka Buch
 
@@
 \section{Trees}
+
+%%% code in eine Tabelle
 

### 2011-07-07 pkg/evtree/inst/doc/article.Rnw
doc added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@193 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}
\usepackage{amssymb, amsmath, amsthm} 
%\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\raggedbottom

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Thomas Grubinger\\ Innsbruck Medical University 
\And Achim Zeileis\\ Universit\"at Innsbruck 
\And Karl-Peter Pfeiffer\\ Innsbruck Medical University
}

\title{\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}}
%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Thomas Grubinger, Achim Zeileis and Karl-Peter Pfeiffer} %% comma-separated
\Plaintitle{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% without formatting
\Shorttitle{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Commonly used classification and regression tree methods like the CART algorithm are recursive partitioning methods that build the model in a forward stepwise search. Although this approach is known to be an efficient heuristic, the results of recursive tree methods are only locally optimal, as splits are chosen to maximize homogeneity at the next step only. An alternative way to search over the parameter space of trees is to use global optimization methods like evolutionary algorithms. This paper describes the \pkg{evtree} package, which implements an evolutionary algorithm for learning globally optimal classification and regression trees in \proglang{R}. CPU and memory-intensive tasks are fully computed in \proglang{C++} while the \pkg{partykit} \citep{hothorn2009partykit} package is leveraged to represent the resulting trees in \proglang{R}, providing unified infrastructure for summaries, visualizations, and predictions. \code{evtree} is compared to \code{rpart} \citep{therneau1997introduction}, the open-source CART implementation, and \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20 evaluated data-sets, including several benchmark problems from the UCI machine learning repository, \code{evtree} models offered at least similar and most of the time increased predictive accuracy.  
}

\Keywords{machine learning, classification trees, regression trees, evolutionary algorithms, \proglang{R}}
\Plainkeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R} %% without formatting
%% at least one keyword must be supplied

%% publication informatione
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Thomas Grubinger, Karl-Peter Pfeiffer\\
  Innsbruck Medical University\\
  Department for Medical Statistics, Informatics and Health Economics\\
  6020, Innsbruck\\
  E-mail: \email{thomas.grubinger@i-med.ac.at}\\
  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger}\\

  Achim Zeileis \\
  Universit\"at Innsbruck\\
  Department of Statistics\\
  6020, Innsbruck\\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Introduction}

Classification and regression trees are commonly applied for exploration and modeling of complex data. They are able to handle strongly nonlinear relationships with high order interactions and different variable types. The resulting model can be interpreted as a tree structure, which gives a compact and intuitive representation. Commonly used classification and regression tree algorithms, including \textit{CART} \citep{breiman1984cart} and \textit{C4.5} \citep{quinlan1993}, use a greedy heuristic, where split rules are selected in a forward stepwise search. The data is recursively split into two groups. The split rule at each internal node is selected to maximize the homogeneity of its two child nodes, without consideration of nodes further down in the tree. Split rules selected by this top-down heuristic are only locally optimal, as splits are chosen to maximize homogeneity at the next step only. Nonetheless, the greedy heuristic is computationally efficient and often yields reasonably good results \citep{murthy1995dti}. However, for some problems, greedily induced trees can be far from the optimal solution, and a global search over the tree's parameter space can lead to much more compact and accurate models.

One possibility to build globally optimal trees is to use stochastic optimization methods like evolutionary algorithms. Evolutionary algorithms are methods that are inspired by natural Darwinian evolution. Important concepts include inheritance, mutation, and natural selection. They are population based, i.e., a whole collection of candidate solutions are processed simultaneously. The initial population is either initialized randomly, or by a problem dependent algorithm. The individuals in the populations are then iteratively modified by \textit{variation operators}. For tree algorithms variation operators can be applied to modify the tree structure and the tests that are applied in the internal nodes. Variation operators that apply to single individuals are called \textit{mutation}, while operators that merge information from different solutions are called \textit{crossover}. The survivor selection process favors individuals that perform well according to some quality criteria, which is called \textit{fitness function} or \textit{evaluation function}. In this evolutionary process the mean quality of the population increases over time \citep{baeck,eiben2003iec}.

In practice, evolutionary algorithm and other stochastic optimization methods are rarely used in decision tree induction. One reason is that they are computationally much more demanding than a forward stepwise search. Perhaps another important reason is their lack of availability in major software packages. \proglang{R} \citep{team2011r} provides several packages with tree-based methods. \pkg{rpart} \citep{therneau1997introduction} is an \proglang{R} implementations of CART. Package \pkg{party} provides two tree-based algorithms with unbiased variable selection and statistical stopping criteria. Several tree algorithms are available in \proglang{Weka} \citep{weka}. Package \pkg{RWeka} \citep{rweka} provides an interface to \proglang{Weka} for \proglang{R} users. Decision tree methods in \proglang{Weka} include implementations of C4.5 and \textit{M5} \citep{quinlan1992lcc}. However, all of these methods are based on a top-down heuristic. An exception is the \pkg{LogicReg} package, which implements logic regression. This method is applicable for regression problems with binary covariates. The resulting models of logic regression can be visualized as logic trees.

In this article, we present \pkg{evtree}, a software package that uses evolutionary methods for learning globally optimal classification and regression trees in \proglang{R}. The package is available from the Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/package=evtree}. An important difference to comparable algorithms is the survivor selection mechanism, where we use a simple $(1+1)$ selection strategy. That is, one parent solution competes with one offspring for a place in the subsequent population. As further discussed in Section~\ref{evo}, this design offers computational advantages for the application of classification and regression trees. Furthermore, we argue that the chosen selection mechanism helps to avoid premature convergence.

The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree} is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and conditional inference trees in \proglang{R}. The comparison comprises 14 benchmark data-sets, 3 real-world data-sets and 3 simulated data-sets. It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods. Finally, in Section~\ref{conc}, we will summarize and draw conclusions on the implementation and the performance of our algorithm.

\section{Globally and locally optimal decision trees}
\label{got}

The aim of classification and regression tree analysis can be stated by explaining a response variable $Y$ by a vector of $P$ predictor variables $X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is qualitative and for regression trees $Y$ is quantitative. Tree-based methods first partition the input space $X$ into a set of $M$ rectangular regions $R_m$ where $m = 1,...,M$. In the second step a model is fit to each region. The model fitted to the distribution of $\{Y \in R_m\}$ can be any proper measure of the location, e.g. the median, or dispersion, e.g. the variance, of a random variable. Typically the mode is used for classification trees and the arithmetic mean is applied for regression trees.

In this work only binary tree models are considered, where regions are split into two halves at a time. However, this is not a big issue in global tree induction. As binary splits are searched globally and any multiway split can be represented by a series of binary splits. Based on a set of split rules, binary classification and regression trees recursively partition the input space $X$ into two regions. Samples that fulfill the criterion of a split rule propagate down into the descendant left node and the other instances into the right node. Split points of a tree are termed \textit{internal nodes} and nodes without successors are called \textit{terminal nodes}. A binary tree with $M$ terminal nodes has $M-1$ internal nodes. Generally, the maximum number of terminal nodes $M_{max}=\infty$. In practice, $M_{max}$ is restricted by the maximum tree depth $D_{max}$ as $M_{max} = 2^{D_{max}}-1$. When the trees are constrained to have a minimum number of observations in the internal and/or terminal nodes, $M_{max}$ is also restricted by the finiteness of the training data.

A binary tree model can be uniquely defined by the positions of the internal nodes, being a subset of $\{1,...,M_{max}-1\}$, and the corresponding split rules. Let the positions of the internal nodes of a tree with $M-1$ terminal nodes be indicated by $n_1,...,n_{M-1}$. The positions $\{n_1,...,n_{M-1}\} \in \{1,...,M_{max}-1\}$ and $n_i < n_j \Leftrightarrow i < j$. Let the variable selected for splitting, at a new position $r \in n_1,...,n_{M_{max}-1}$, be denoted as $v_r$, where $v_r=1,...,P$. Depending on the domain of $X_{v_r}$, a split point or a set of categories has to be selected, which we denote as $s_r$. If $X_{v_r}$ is numerical or ordinal, a binary split rule is defined as $X_{v_r} < s_r$. If $X_{v_r}$ has $u$ unique values, then there are $u-1$ possible split points and $s_r=1,...,u-1$. In the case of a nominal split variable $X_{v_r}$, a split rule is defined as $X_{v_r} \in s_r$, where $s_r \subset \{1,...,K\}$ and $s_r \neq \{\emptyset\}$. If $X_{v_r}$ has $K$ categories there are $2^{K-1}-1$ possible splits.

\subsection{The parameter space of globally optimal decision trees}

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a function of the number of terminal nodes, without further considering the depth or the shape of trees. Then the overall parameter space $\Theta$ can be written as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic element of $\Theta$, then $\theta$ is fully defined by:

\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}

The overall goal of classification and regression tree analysis can be defined as finding a set of parameters $\hat{\theta}$, such that

\begin{equation}
	\hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; loss(Y, f(X,\theta)) + comp(\theta).
	\label{generalcostfunction}
\end{equation}

Based on the input data $X$ and the paramters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$. $loss$ is a suitable loss function for the domain of $Y$. Commonly, the misclassification error for classification problems, or the squared error for regression problems, is used. $comp$ quantifies the complexity of trees and penalizes for more complex models in the tree selection process. In CART, the complexity is a non-negative function of the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all $\Theta_m$.

The parameter space $\Theta$ becomes large for already medium sized problems and a complete search for larger problems is computationally intractable. In fact, \cite{hyafil1976cob} showed that building optimal binary decision trees, such that the expected number of tests required to classify an unknown sample is minimized, is NP-complete. \cite{zantema2000finding} proved that finding a decision tree of minimal size that is decision equivalent to a given decision tree is an NP-hard problem. As a consequence the search space is usually limited by heuristics.

\subsection{The parameter space of locally optimal decision trees}

Instead of searching all combination of $\Theta$ simultaneously, recursive partitioning algorithms only consider one split at a time. At each internal node $r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding split point $s_r$ is selected such that it minimizes the impurity in the two resulting terminal nodes. Let the parameters of a tree, constructed in $i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} = (\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the position of the new internal node $r$, as well as the corresponding split variable $v_r$ and split point $s_r$, are chosen such that:

\begin{equation}
		\min_{v_r, s_r} \; loss(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}

CART's standard measures for node impurity are the Gini index, for classification problems and the mean squared error, for regression problems. After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by $\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This process of recursively selecting and executing splits continues until the homogeneity of the nodes cannot be further improved. Additionally, often additional stopping criteria are defined, e.g. minimum sample sizes in the internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is followed by a pruning step. The idea behind pruning is to first grow a large tree without or with little restriction on the magnitude of impurity reduction. Starting from the bottom of the tree pruning then removes worthless splits. The method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning iteratively prunes the internal node that:

\begin{equation}
		\min_{r=1,...,n_{M-1}} \; loss(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + comp(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}

The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus (\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

For nontrivial problems, recursive partitioning methods only search a small fraction of the global search space $(v_1, s_1, ..., v_{M_{max}-1}, s_{M_{max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of the other split rules. The selection of $v_r$ and $s_r$ at each step $i$ requires a search of the locally optimal split over $M=i$ terminal nodes. The pruning step requires $j$ selections of an internal node $r$.

In contrast to CART and C4.5, conditional inference trees utilize an unbiased variable selection approach that is based on permutation testing. Bias in the split variable selection process arises from a preference towards variables with many potential split points. The conditional inference framework overcomes this problem by separating the variable and split point selection. Although this distinction influences the selection of split rules, conditional inference tree solve the same minimization problem (Equation \ref{eq:whichsplit}). The \code{ctree} implementation of conditional inference trees does not employ pruning, but uses a statistical stopping criterion.

\subsection{An illustration showing the limitations of locally optimal decision trees}

A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}. The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions. As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split. For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem. One solution, which is the tree constructed by \code{evtree} is given by:

<< echo = FALSE, split= TRUE>>=
library("evtree")
X1 <- c(rep(0.25, 4), rep(0.75, 4), rep(1.25 ,4),  rep(1.75, 4))
X2 <- c(rep(seq(0.25, 1.75 ,0.5),4))
Y <- array(1,16)
Y[X1 < 1 & X2 < 1]=2
Y[X1 > 1 & X2 > 1]=2
chess22 <- as.data.frame(cbind(X1, X2, Y))
chess22[,3] <- as.factor( chess22[,3] )
levels(chess22[,3]) = c("class: o","class: x")
print(evtree(Y~.,data=chess22, minbucket = 1, minsplit=2))
@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
plot(chess22[chess22[,3]=="class: o",1], chess22[chess22[,3]=="class: o",2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", xlim=c(0,2), ylim=c(0,2), cex=2)
points(chess22[chess22[,3]=="class: x",1], chess22[chess22[,3]=="class: x",2], pch= c("o"), cex=2)
par(cex.axis=1.5, cex.lab=1.5)
@
\caption{Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\label{fig:illustration}
\end{figure}

All instances all are classified correctly. Each of the terminal nodes 3 and 7 contain four instances of the class $(x)$. Four instances of class $(o)$ are assigned to each of the terminal nodes 4 and 6.

\subsection{Approaches for learning globally optimal decision trees}

When compared with the described forward stepwise search, a less greedy approach is to calculate the effects of the choice of the split rule deeper down in the tree. In this way optimal trees can be found for simple problems. However, feature selection at a node of greedily induced trees has complexity  $O(PN)$. Through a search of the optimal split up to $d-1$ levels the complexity increases to $O(P^dN^d)$ \citep{papagelis2001bdt}. Therefore, for more difficult problems the search space at each node must be limited by the depth of the search. Additionally, the search space can be reduced, limiting the number of possible split points of the selected variables and/or other heuristics. These methods are known as \textit{look-ahead algorithms} \citep{esmeir2007ald}. Another class of algorithms is given by stochastic optimization methods. Given an initial tree, these methods seek improved solutions through stochastic changes to the tree structure. Besides evolutionary algorithms \citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and \textit{simulated annealing} \citep{sutton1992ict} were used successfully to solve difficult classification and regression tree problems. \cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees. \cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several data-sets from the UCI machine learning repository \citep{Frank+Asuncion:2010}. A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.  \cite{cantu2003inducing} used an evolutionary algorithm to induce oblique classification trees. \cite{gray2008cta} introduced a classification tree algorithm that allows univariate splits, but also splits defined by linear combinations of two or three variables.

\section[The evtree algorithm]{The \code{evtree} algorithm}
\label{desc}

The general framework of evolutionary algorithms emerged from different representatives. \cite{Holland} called his method \textit{genetic algorithms}, \cite{rechenberg} invented \textit{evolution strategies}, and \cite{Fogel} introduced \textit{evolutionary programming}. More recently, \cite{Koza} introduced a fourth stream and called it \textit{genetic programming}. All four representatives only differ in the technical details, for example the encoding of the individual solutions, but follow the same general outline \citep{eiben2003iec}. Evolutionary algorithms are being increasingly widely applied to a variety of optimization and search problems. Common areas of application include data mining \citep{freitas2002sea,cano2003uea}, statistics \citep{de34glmulti}, signal and image processing \citep{man1997gac}, and planning and scheduling \citep{jensen2001raf}.

The pseudocode for the general scheme of an evolutionary algorithm is depicted in Table~\ref{pc:genericEA}. With the description of these components together with the definition of individuals, the most important aspects of an evolutionary algorithm are specified. The representation of an individual is given by a binary tree structure, which is fully defined with the definition of $\theta$ in Equation~\ref{treeRep}. This section continues with the description of the remaining components following the general outline displayed in Table~\ref{pc:genericEA}.

\begin{table}[tbh!]
\begin{center}
\begin{tabular}{l}
\toprule
1. initialize the population \\
2. evaluate each individual  \\
3. while(termination condition is not satisfied) do  \\
\hspace{8mm} a. select parents  \\
\hspace{8mm} b. alter selected individuals via variation operators \\
\hspace{8mm} c. evaluate new solutions  \\
\hspace{8mm} d. select survivors for the next generation  \\
\hspace{3.8mm} od  \\
\bottomrule
\end{tabular}
\end{center}
\caption{Pseudocode of the general schema of an evolutionary algorithm.} 
\label{pc:genericEA}
\end{table}

This section describes the general functioning of the \code{evtree} algorithm. Implementation details as user-specified parameters, including their default values and their possible ranges, are discussed in the subsequent section. These implementation details also include parameters that specify a minimum number of observation that are required in each internal node and  terminal node of a tree. In the following, the term \textit{invalid node} refers to a node that do not contain enough observations.

\subsection{Initialization}
\label{init}

Each tree of the population is initialized with a valid, randomly generated, split rule in the root node. First, $v_1$ is selected with uniform probability from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is nominal and has $K$ categories, each $k=1,...,K$ has a probability of $50\%$ to be assigned to the left or the right terminal node. In cases where all $k$ are allocated to the same terminal node, the assignment of one category is flipped to the other terminal node. If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the initialized tree is fully described by $\tilde{\theta}_{1}= (v_1, s_1)$.

\subsection{Parent selection}
\label{parent}

In every iteration, each tree is selected once to be modified by one of the variation operators. In cases where the crossover operator is applied, the second parent is selected randomly from the remaining population. In this way, some trees are selected more than once in each iteration.

\subsection{Variation operators}
\label{vari}

Four types of mutation operators and one crossover operator are utilized by our algorithm. In each modification step, one of the variation operators is randomly selected for each tree. The mutation and crossover operators are described below.

\subsubsection{Split}

\textit{Split} selects a random terminal-node and assigns a valid, randomly generated, split rule to it. As a consequence, the selected terminal node becomes an internal node $r$ and two new terminal nodes are generated.

The split rule of the new internal node is chosen as follows. First, $v_r$ is selected with uniform probability from $1,...,P$. Second, a split point $s_r$ is selected. If $X_{v_r}$ is numeric or ordinal, a split point $s_r$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_r}$. If $X_{v_r}$ is nominal, each category of $X_{v_r}$, has a probability of $50\%$ to be assigned to the left or the right terminal node. In cases where all $k$ are allocated to the same terminal node, one of them is changed to the other terminal node. If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated for a maximum of $P$ iterations.

In cases where no valid split rule can be assigned to the internal node at position $r$, the search for a valid split rule is carried out on another randomly selected terminal node. If, after $10$ attempts, no valid split rule can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$. Otherwise, the set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}= \tilde{\theta}_i \cup (v_{r},s_{r})$.

\subsubsection{Prune}
\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two terminal nodes as successors and prunes it into a terminal node. The tree's parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i} \setminus (v_{r},s_{r})$. If $\tilde{\theta}_{i}$ only comprises one internal node, i.e., the root node, than $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.

\subsubsection{Major split rule mutation}
\textit{Major split rule mutation} selects a random internal node $r$ and changes the split rule, defined by the corresponding split variable $v_r$, and the split point $s_r$. With a probability of $50\%$, a value from the range $1,...,P$ is assigned to $v_r$. Otherwise $v_r$ remains unchanged and only $s_r$ is modified. Again, depending on the domain of $X_{v_r}$, either a random split point from the range of possible values of $X_{v_r}$ is selected, or a non-empty set of categories is assigned to each of the two terminal nodes. If the split rule at $r$ becomes invalid, the mutation operation is reversed and the procedure, starting with the selection of $r$, is repeated for a maximum of $3$ attempts. Subsequent nodes that become invalid are pruned.

If no pruning occurs, $\tilde{\theta}_{i}$ and $\tilde{\theta}_{i+1}$ contain the same set of parameters. Otherwise, the set of parameters $(v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$, corresponding to invalid nodes, is removed from $\tilde{\theta}_i$. Thus, $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.

\subsubsection{Minor split rule mutation}

\textit{minor split rule mutation} is similar to the \textit{major split rule mutation} operator, but it does not alter $v_r$ and only changes the split point $s_r$ by a minor degree.
If $X_{v_r}$ is numerical or ordinal the split point $s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases where $X_{v_r}$ has less then $20$ unique values, the split point is change to to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise, $s_r$ is randomly shifted by a number of unique values that is not larger than $10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal variable, with less than $20$ categories, one of the categories is randomly modified. Otherwise, at least one and at most $10\%$ of the variable's categories are changed. In cases where subsequent nodes become invalid, further split points are searched that preserve the tree's topology. After five non-successful attempts at finding a topology preserving split point, the non-valid nodes are pruned.

Equivalently to the \textit{major split rule mutation} operator the subsequent solution $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.

\subsubsection{Crossover}

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees that are chosen for crossover. First, an internal node is selected randomly from each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as $n_k^1$ and $n_k^2$. Let $sub1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees rooted by $n_k^j$, where $j=1,2$. Then, $sub1(\tilde{\theta}_i^j)$ contains the parameters $(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, ...,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set of descendent node positions of $n^j_k$. Now, the two selected subtrees can be written as $sub1(\tilde{\theta}_{i}^1, n^1_k) $ and $sub1(\tilde{\theta}_{i}^2, n^2_k)$. Furthermore, $sub2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1 \setminus sub1(\tilde{\theta}_{i}^1, n^1_k)$ and $sub2(\tilde{\theta}_{i}^2, n^2_k) = \tilde{\theta}_{i}^2 \setminus sub1(\tilde{\theta}_{i}^2, n^2_k)$.

After the application of the \textit{crossover} operator, the trees' parameter spaces are given by $\tilde{\theta}_{i+1}^1= sub2(\tilde{\theta}_{i}^1, n^1_k) \cup sub1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2= sub2(\tilde{\theta}_{i}^2, n^2_k) \cup sub1(\tilde{\theta}_{i}^1, n^1_k) \setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here, $(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset sub1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}}) \subset sub1(\tilde{\theta}_i^2, n^2_k)$ are the parameters of the invalid nodes.

\subsection{Evaluation function}
\label{eval}

The evaluation function represents the requirements the population should adapt to. In general, these requirements are formulated by Equation~\ref{generalcostfunction}. A suitable evaluation function for classification and regression trees minimizes the models' accuracy on the training data, and the models' complexity. This subsection describes the currently implemented choices of evaluation functions for classification and for regression.

\subsubsection{Classification}

The quality of a classification tree is most commonly measured as a function of its misclassification loss $MC(Y, f(X, \theta)) = \sum_{n=1}^N I(Y_n \neq \hat{f}(X_{\cdot n}, \theta))$, and the complexity of a tree by a function of the number of its terminal nodes $M$. \code{evtree} uses $2 \cdot MC(Y, f(X, \theta))$ as a loss function. The number of terminal nodes, weighted by $\ln N$ and a user-specified parameter $\alpha$, measures the complexity of trees. With these particular choices, the general definition of the optimization problem in Equation~\ref{generalcostfunction} can be written as:

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
\label{eq:classification}
\end{equation}

Equation~\ref{eq:classification} seeks parameters $\hat{\theta}$ that minimize the misclassification loss and the number of terminal nodes. Increasing values of $\alpha$ encourage decreasing tree sizes.

Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian Information Criteria (BIC)} \citep{gray2008cta} and \textit{Minimum Description Length (MDL)} \citep{quinlan1989inferring}. For both evaluation functions deviance is used for accuracy estimation. Deviance is usually preferred over the misclassification rate in recursive partitioning methods, as it is more sensitive to changes in the node probabilities \citep[pp.~308--310]{hastie}. However this is not necessarily an advantage for global tree building methods like evolutionary algorithms.

\subsubsection{Regression}

For regression trees, accuracy is usually measured by the mean squared error $MSE(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$. Using $N \cdot \ln MSE(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln MSE(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
\label{eq:regression}
\end{equation}

Here, $M+1$ is the effective number of estimated parameters, taking into account the estimates of a mean parameter in each of the terminal nodes and the constant error variance term. With $\alpha = 0.25$ the criteria is, up to a constant, equivalent to the BIC used by \cite{fan2005rta}. However, the effective number of parameters estimated for is actually much higher than $M+1$ due to the selection of parameters in the split variable and the selection of the variable itself. It is however unclear how these should be counted \citep[p.~222]{gray2008cta, ripley}. Therefore, a more conservative default value of $\alpha=1$ is assumed.

\subsection{Survivor selection}
\label{evo}

The population size stays constant during the evolution and only a fixed subset of the candidate solution can be kept in memory. A common strategy is the $(\mu + \lambda)$ selection, where $\mu$ survivors for the next generation are selected from the union of $\mu$ parents and $\lambda$ offsprings. An alternative approach is the $(\mu,\lambda)$ strategy. In this schema $\mu$ survivors for the next generation are selected from $\lambda$ offsprings.

Our algorithm uses a $(1+1)$ selection schema, where one parent solution competes with one offspring for a place in the population. In the case of a mutation operator either the solution before $\tilde{\theta}_i$ or after modification $\tilde{\theta}_{i+1}$ is kept in memory. In the case of the crossover operator, the initial solutions of $\tilde{\theta}_i^1$ competes with its subsequent solutions $\tilde{\theta}_{i+1}^1$. Correspondingly, one of the two solutions $\tilde{\theta}_i^2$ and $\tilde{\theta}_{i+1}^2$ is rejected. The survivor selection is done deterministically. The tree with lower fitness, according to the evaluation function, is rejected. Note that, due to the definition of the crossover operator, that some trees are selected more than once in an iteration. Correspondingly, these trees undergo the survival selection process more than once in an iteration.

As in classification and regression tree analysis the individual solutions are represented by trees. This design offers computational advantages over $(\mu + \lambda)$, with $\mu>1 \;  \& \; \lambda>1$, and $(\mu,\lambda)$ strategies. In particular, for the application of mutation operators no new trees have to be constructed. The tree after modification is simply accepted or reversed to the previous solution.

There are two important issues in the evolution process of an evolutionary algorithm: population diversity and selective pressure \citep{michalewicz1994genetic}. These factors are related, as with increasing selective pressure the search is focused more around the currently best solutions. An overly strong selective pressure can cause the algorithm to converge early in a local optima. On the other hand, an overly weak selective pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in situations as follows. Suppose the $b^{th}$ tree of the population is one of the fittest trees in iteration $i$, and in iteration $i$  one split rule of the $b^{th}$ tree is changed only by a minor degree. Then very few instances are classified differently and the overall misclassification might not even change. However, as they represent one of the best solutions in iteration $i$, they are both selected for the subsequent population. This situation can occur frequently, especially when a fine-tuning operator like the \textit{minor split rule mutation} is used. Then, the diversity of different trees is lost quickly and the algorithm likely terminates in a local optimum. The $(1+1)$ selection mechanism clearly avoids these situations, as only the parent or the offspring can be part of the subsequent population.

\subsection{Termination}
\label{term}

The algorithm terminates when the quality of the best $5\%$ of trees stabilizes for $100$ iterations, but not before $1000$ iterations. If the run does not converge the algorithm terminates after a user-specified number of iterations. In cases where the algorithm does not converge, a warning message is written to the command line. The tree with the highest quality according to the evaluation function is returned.

\section{Implementation}
\label{impl}

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are

\begin{verbatim}
evtree(formula, data = list(), weights = NULL, subset = NULL, control = 
evtree.control(...), ...)
\end{verbatim}

where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters

\begin{verbatim}
evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
niterations = 10000L, ntrees = 100L, alpha = 1, operatorprob = list(
pmutatemajor = 0.2, pmutateminor = 0.2, pcrossover = 0.2, psplit = 0.2, 
pprune = 0.2), seed = NULL, ...)
\end{verbatim}

where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.

<<>>=
library("evtree")
data("iris")
model <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
The fitted model can be examined using the \code{print()} and \code{plot()} functions.

<<>>=
print(model)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[tbh!]
    \begin{center}
<<echo= FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
<<fig=TRUE, echo=TRUE, width = 13.5, height= 8.5>>=
plot(model)
@
     \end{center}
     \caption{Example of a tree plotted by the \code{plot()} function}
     \label{fig:example}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.

<<>>=
predict(model, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

\section{Performance comparison}
\label{comp}

In this Section we compare \code{evtree} with \code{rpart} and \code{ctree}. All models are constrained to a minimum number of 7 observations per terminal node, 20 observations per internal node and a maximum tree depth of 9. Apart from that, the default settings of the algorithms are used.

In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark data-sets that are publicly available and $3$ real-world data-sets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}. The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 data-sets. The misclassification rate on the \textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of predictive accuracy. Furthermore, the complexity is estimated by the number of terminal nodes. Missing values are currently not supported by \code{evtree}. The only data-set with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis. Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual data-sets .

In the second part (Section~\ref{comp2}) the algorithms' performances are assessed on an artificial problem that is simulated with different noise levels. The estimation of predictive accuracy and the number of terminal nodes is based on $250$ realizations for each simulation.

\subsection{Benchmark and real-world problems}
\label{comp1}

In Table~\ref{tab:dataDescription} the benchmark and real-world data-sets from the Austrian DRG-system are described. In the Austrian DRG-system, resources are allocated to hospitals by simple rules regarding diagnosis, selected individual services, and patient characteristics. Within this system regression tree analysis is performed to classify hospital patients into clinically meaningful and comprehensible groups that consume similar hospital resources. The data-sets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure. The task of the regression tree analysis is to further divide these groups with the intention of finding groups with similar resource consumption. Possible split-variables are the patients' diagnoses, procedures, age, and gender. The dependent variable is the patients' length of stay, which has a good relationship with the total costs and is available \citep{theurl2007ihf}. Permission to use the data was granted by the \citet{BMG2009}. A more detailed description of the application can be found in \citet{grubinger2010regression}.

\begin{table}
\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
\textsc{Data-set} & \textsc{Instances} &  \multicolumn{5}{l}{\textsc{Attributes}}   \\ 
                  &                    &  binary              & nominal          &  ordered      &  metric     & classes \\ 
\midrule
Glass Identification$^\#$  &   214   &  - &  - &  - & 9 &  6      	 \\
Statlog Heart*         &   270   &  3 &   3 & 1 & 6 &   2     	 \\
Ionosphere$^\#$ &   351   &  2 &  - &  - & 32 & 2  	             \\
Musk$^+$  &  476  & - &  - &  - & 166 & 2                     	 \\
Breast Cancer Database$^\#$ &   685   &  - &  4 & 5 &  - &  2  	 \\
Pima Indians Diabetes$^\#$ &   768   &  - &   - & - & 8 &  2      	 \\
Vowel$^\#$  &  990  & - &  1 &  - & 9 & 11     	                     \\
Statlog German Credit* &  1000   &  2 &  10 & 1 & 7 &  2      	 \\
Contraceptive Method* &   1437   & 3 &  - &  4 &  2 &  3  	     \\
DNA$^\#$  &  3186  & 180 &  - &  - & - & 3   	                       \\
Spam$^+$  &  4601  & - &  - &  - & 57 & 2                      	 \\
MAGIC Gamma Telescope*  &  19020  & - &  - &  - & 10 & 2     	   \\
Servo$^\#$ & 167 &  - & 4 & - &  -  &  -											   \\
Boston Housing$^\#$ & 506 & 1 & - & - &  12  & -                 \\
MEL0101$^\diamond$ &   875 &  1 & 4 & 1 &  108  & -              \\
HDG0202$^\diamond$ &  3933 &  1 & 7 & 1 &   46  & -              \\
HDG0502$^\diamond$ & 8251 &  1 & 7 & 1 &   91  & -               \\
\bottomrule
\end{tabular}
\end{center}
\caption{Description of the evaluated benchmark data-sets. The data-sets marked with a $(^*)$ are from the UCI machine learning repository \citep{Frank+Asuncion:2010}. Data-sets marked with a $(^\#)$ and $(^{+})$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and \pkg{kernlab} \citep{kernlab}. The $3$ real-world data-sets from the Austrian DRG system are marked with a $(^\diamond)$. } 
\label{tab:dataDescription}
\end{table}

The relative performance of \code{evtree} and \code{rpart} is summarized in Figure~\ref{fig:crpart}. Performance differences are displayed relative to \code{evtree}'s performance. For example, on the \textit{Glass} data-set, the average misclassification rate of \code{rpart} is $2.8\%$ higher than the misclassification rate of \code{evtree}. It can be observed that on 13 out of 17 data-sets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy. Only on the \textit{Contraceptive Method} data-set \code{evtree} performs slightly worse.  In terms of complexity, \code{evtree} models are significantly more complex on 9 and less complex on 7 data-sets.

Figure~\ref{fig:cptree} summarizes the relative performance of \code{evtree} and \code{ctree}. For 14 out of 17 data-sets \code{evtree} shows a better predictive performance. The algorithms' performances on the 3 other data-sets is similar. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 data-set.

%% vorberechnungen
<<label=figciplots, echo = FALSE>>=
rm(list=ls())
library("plotrix")
library("lme4")
library("multcomp")
load(file="./results/breastcancer.RData")
load(file="./results/glass.RData")
load(file="./results/credit.RData")
load(file="./results/heart.RData")
load(file="./results/pima.RData")
load(file="./results/contraceptive.RData")
load(file="./results/dna.RData")
load(file="./results/musk.RData")
load(file="./results/spam.RData")
load(file="./results/vowel.RData")
load(file="./results/ionosphere.RData")
load(file="./results/magicgamma.RData")
load(file="./results/hdg0202.RData")
load(file="./results/hdg0502.RData")
load(file="./results/mel0101.RData")
load(file="./results/bostonhousing.RData")
load(file="./results/servo.RData")

preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
    if(isclassification == TRUE)
        d[,1:3] <- 1-d[,1:3]
    d <- as.data.frame(d)
    colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")
    d[,2] <- d[,2] / d[,1] *100
    d[,3] <- d[,3] / d[,1] *100
    d[,1] <- d[,1] / d[,1] *100
    d[,5] <- d[,5] / d[,4] *100
    d[,6] <- d[,6] / d[,4] *100
    d[,4] <- d[,4] / d[,4] *100
    x <- d[ ,1:3]
    y <- d[ ,4:6]
    x <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
    y <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")
    x$ds <- dname
    y$ds <- dname
    return( cbind(x,y) )
}

r <- preprocess(d = rglass, dname = "Glass Identification", isclassification = T)
r <- rbind(r, preprocess(d = rheart, dname = "Statlog Heart", isclassification = T))
r <- rbind(r, preprocess(d = rionosphere, dname = "Ionosphere", isclassification = T))
r <- rbind(r, preprocess(d = rmusk, dname = "Musk", isclassification = T))
r <- rbind(r, preprocess(d = rbreastcancer, dname = "Breast Cancer Database", isclassification = T))
r <- rbind(r, preprocess(d = rpima, dname = "Pima Indians Diabetes", isclassification = T))
r <- rbind(r, preprocess(d = rvowel, dname = "Vowel", isclassification = T))
r <- rbind(r, preprocess(d = rcredit, dname = "Statlog German Credit", isclassification = T))
r <- rbind(r, preprocess(d = rcontraceptive, dname = "Contraceptive Method", isclassification = T))
r <- rbind(r, preprocess(d = rdna, dname = "DNA", isclassification = T))
r <- rbind(r, preprocess(d = rspam, dname = "Spam", isclassification = T))
r <- rbind(r, preprocess(d = rmagic, dname = "Magic Gamma Telescope", isclassification = T))
r <- rbind(r, preprocess(d = rservo, dname = "Servo", isclassification = F))
r <- rbind(r, preprocess(d = rbostonhousing, dname = "Boston Housing", isclassification = F))
r <- rbind(r, preprocess(d = mel0101, dname = "MEL0101", isclassification = F))
r <- rbind(r, preprocess(d = hdg0202, dname = "HDG0202", isclassification = F))
r <- rbind(r, preprocess(d = rhdg0502, dname = "HDG0502", isclassification = F))

ra <- r[,1:4]
rc <- r[,5:8]

tf <- function(x){
    names(x)[[2]] <- "value"
    x$samp <- factor(x$samp) 
    x$ds <- factor(x$ds)
    x$dssamp <- x$ds:x$samp # unique levels of dataset and bootstrap sample
    x$alg <- factor(x$alg)
    x$alg <- relevel(x$alg, "evtree") # change order of algorithms
    x
}
ra <- tf(ra)
rc <- tf(rc)

uniquedatasetlabels <- unique(ra$ds)[length(unique(ra$ds)):1]

cstats <- function(x= ra, alg= "rpart"){
    mean  <- array(0,17)
    lower <- array(0,17)
    upper <- array(0,17)
    for(i in 1:length(uniquedatasetlabels)){
        mod1 <- lm(value ~ alg, data = subset(x, x$ds == uniquedatasetlabels[i]))
        pt <- glht(mod1, linfct = mcp(alg = "Dunnett"))
        w <- confint(pt)$confint
        if(alg == "rpart")
            d <- 1
        else
            d <- 0
        mean[i]  <-  w[1+d]
        lower[i] <-  w[3+d]
        upper[i] <-  w[5+d]
    }
    return(as.data.frame(cbind(mean,lower,upper)))
}

plotd <- function(alg){
    par(las=1, cex=1.5, cex.lab=1.3, mfrow= c(1,2), omd=c(0.16,1,0,1), mar=c(5, 0, 1, 2) + 0.1)
    dummy <- c(1:17)
    names(dummy) <- uniquedatasetlabels
    stats <- cstats(ra, alg)
    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-12,45), axes=F, ann=T, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
    axis(side = 2, at = c(1:length(dummy)), labels= names(dummy), lwd=0.1)
    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-10,-5,0,5,10,15,20,25,30,35,40))
    box(lwd=1)
    abline(v=0, lty=2)
    abline(h=5.5, lty=1, lwd=0.5)
    stats <- cstats(rc, alg)
    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-160,210), axes=F, ann=T, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-150,-100,-50,0,50,100,150,200))
    box(lwd=1)
    abline(v=0, lty=2)
    abline(h=5.5, lty=1, lwd=0.5)
}
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("rpart")
@
\caption{Performance comparison of \code{evtree} and \code{rpart}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
 \label{fig:crpart}
\end{figure}

\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("ctree")
@
\caption{Performance comparison of \code{evtree} and \code{ctree}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
\label{fig:cptree}
\end{figure}

The disadvantages of our algorithm are computation time and memory requirements. While the smallest of the analyzed data-sets, \textit{Glass Identification}, only needed approximately 4--6 seconds to fit, larger data-sets demanded several minutes. The fit of a model from the largest data-set, \textit{MAGIC Gamma Telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit. The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.

\subsection{Artificial Problem}
\label{comp2}

In this section we demonstrate the ability of \code{evtree} to solve an artificial problem that is difficult to solve for most recursive classification tree algorithms \citep{loh2009improving}. The data was simulated with $2000$ instances for both the training-set and the test-set. Predictor variables $X_1$ and $X_2$ are simulated to be uniformly distributed in the interval $[0,4]$. The classes are distributed in alternating squares forming a $4 \times 4$ chessboard in the $(X_1, X_2)$-plane. One realization of the simulated data is shown in Figure~\ref{fig:chess}. Furthermore, variables $X_3-X_8$ are noise variables that are uniformly distributed on the interval [0, 1]. The ideal model for this problem only uses variables $X_1$ and $X_2$ and has 16 terminal nodes, whereas each terminal node comprises the observations that are in the region of one square. Two further simulations are done in the same way, but $5\%$ and $10\%$ percent of the class labels are randomly changed to the other class.

<<echo = FALSE>>=
chessboard44 <- function(totalpoints = 4000, noisevariables = 6, noise = 0){
    chess44 <- array(0,c(totalpoints,noisevariables+3))
    for(i in 1:(noisevariables+2))
        chess44[,i] <- as.numeric(runif(dim(chess44)[1]))*4

     x <- chess44[,1]
     y <- chess44[,2]
     chess44[,ncol(chess44)] <- 0
     for(k in 1:4)  
        chess44[(x <= k & x > k-1 & y <= k & y > k-1), ncol(chess44)] <- 1
     for(k in 1:2)  
        chess44[(x <= k & x > k-1 & y <= k+2 & y > k+1), ncol(chess44)] <- 1
     for(k in 1:2)  
        chess44[(y <= k & y > k-1 & x <= k+2 & x > k+1), ncol(chess44)] <- 1

     if( noise > 0){
        flipclasslist <- sample(totalpoints, totalpoints * (noise / 100), replace = F)

        for(i in 1:length(flipclasslist)){
            if(chess44[flipclasslist[i], ncol(chess44)] == 1)
                chess44[flipclasslist[i], ncol(chess44)] = 0
            else if(chess44[flipclasslist[i], ncol(chess44)] == 0)
                chess44[flipclasslist[i], ncol(chess44)] = 1
        }
    }

    chess44 <- as.data.frame(chess44)
    chess44[,ncol(chess44)] <- as.factor(chess44[,ncol(chess44)])
    names(chess44)[9] <- "class"
    chess44
}
@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
ch <- chessboard44(totalpoints= 2000)
par(cex.axis=1.5, cex.lab=1.5)
plot(ch[ch[,9]==1,1], ch[ch[,9]==1,2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", cex=1)
points(ch[ch[,9]==0,1], ch[ch[,9]==0,2], pch= c("o"))
@
\caption{Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\label{fig:chess}
\end{figure}

The results are summarized in Table~\ref{tab:resultsChessboard}. It can be seen that, in the absence of noise, \code{rpart} models on average classify $67.2\%$ of the data points correctly and had $13.0$ terminal nodes. An average \code{ctree} model only has $1.1$ terminal nodes and a classification accuracy of $50.2\%$. In contrast, \code{evtree} classifies $92.8\%$ of the instances correctly and requires 14.0 terminal nodes. In the presence of $5\%$ and $10\%$ noise, \code{evtree} classifies $88.7\%$ and $84.2\%$ of the data correctly.

<<label =  tab:resultsChessboard,  echo = FALSE, results=tex>>=
library(xtable)
load(file = "./results/chessboard44_0.RData")
load(file = "./results/chessboard44_5.RData")
load(file = "./results/chessboard44_10.RData")

chesstable_means  <- as.data.frame( rbind(apply(rchessboard44_0,2,mean), apply(rchessboard44_5,2,mean) , apply(rchessboard44_10,2,mean) )) 
names(chesstable_means) <-  c("evtree", "rpart", "ctree", "evtree", "rpart", "ctree")
chesstable_means[,1:3] <-  format(chesstable_means[,1:3]*100, digits=1, nsmall=1)
chesstable_means[,4:6] <-  format(chesstable_means[,4:6], digits=1, nsmall=1)

chesstable_sd  <- as.data.frame( rbind(apply(rchessboard44_0,2,sd), apply(rchessboard44_5,2,sd) , apply(rchessboard44_10,2,sd) )) 
names(chesstable_sd) <-  c("evtree", "rpart", "ctree", "evtree", "rpart", "ctree")
chesstable_sd[,1:3] <-  format(chesstable_sd[,1:3]*100, digits=1, nsmall=1)
chesstable_sd[,4:6] <-  format(chesstable_sd[,4:6], digits=1, nsmall=1)

chesstable <- chesstable_means
for(j in 1:ncol(chesstable_means)){
	for(i in 1:nrow(chesstable_means)){
		chesstable[i,j] <- paste(chesstable_means[i,j] ,  "(", chesstable_sd[i,j], ")",  sep="")	
	}
}

chesstable <- cbind(as.integer(rbind(0,5,10)), chesstable)
colnames(chesstable)[1] = ""

print(xtable(chesstable,
caption = "Comparison of accuracy on simulated $4 \\times 4$ chess-board examples. Accuracy and the number of terminal nodes are displayed as $mean(std)$.",
caption.placement= "bottom",
label= "tab:resultsChessboard"), 
include.rownames = FALSE, allign= "rllllll", hline.after=NULL,
add.to.row=list(pos=list(-1,-1, 0, 3), command=c(
"\\toprule", 
c("\\multicolumn{1}{l}{Noise [\\%]} & \\multicolumn{3}{l}{Accuracy}  & \\multicolumn{3}{l}{Terminal Nodes}\\\\",
"\\midrule",
"\\bottomrule"
)))
)
@

\section{Conclusions}
\label{conc}

In this paper, we presented the \pkg{evtree} package, which implements classification and regression trees that are evolved by through an evolutionary algorithm. The package uses standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data from the \pkg{partykit} package. As the learning step of our algorithm is computationally demanding, most calculations are conducted in \proglang{C++}. At the moment our algorithm does not support parallelism. However, we intend to extend \pkg{evtree} to parallel computing.

In Section~\ref{comp} we compared \code{evtree} with \code{rpart} and \code{ctree} on 14 benchmark and 3 real-world data-sets. The data-sets varied by sample size, the number of classes, and by the number and types of attributes. \code{evtree} models offered at least comparable and, for the majority of the data-sets, improved predictive accuracy. In an artificial problem of a $4 \times 4$ chessboard we demonstrated \code{evtree}'s performance in the presence of a higher order of interactions and noise.

The goal of this package is not to replace the well-established algorithms like \code{rpart} and \code{ctree}. We rather want to provide \proglang{R} users with an alternative which, given sufficient amounts of time and main memory, is often more accurate than recursive partitioning methods. By the nature of the algorithm it is able to discover patterns which cannot be modeled by a greedy algorithm. As \code{evtree} models can be substantially different to recursively fitted models, it can be beneficial to use both approaches, as this may reveal additional relationships in the data.

\bibliography{article.bib}

\end{document}

### 2011-08-05 pkg/evtree/inst/doc/evtree.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@197 255f4249-2263-4417-b1ea-8498779e27a7
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@198 255f4249-2263-4417-b1ea-8498779e27a7
started some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@205 255f4249-2263-4417-b1ea-8498779e27a7

--- 2011-07-07 pkg/evtree/inst/doc/article.Rnw
@@
 \documentclass[nojss]{jss}
-\usepackage{amssymb, amsmath, amsthm} 
-%\usepackage{longtable}
-\usepackage{booktabs}
-\usepackage{listings}
-\raggedbottom
+\usepackage{amssymb, amsmath, amsthm, booktabs, thumbpdf}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Sweave/vignette information and metadata
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine = R, eps = FALSE, keep.source = TRUE}
+%\VignetteIndexEntry{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
+%\VignetteDepends{stats,evtree,partykit,mlbench,plotrix,lme4,multcomp,rpart,ctree,xtable}
+%\VignetteKeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
+%\VignettePackage{evtree}
 
-%% almost as usual
 \author{Thomas Grubinger\\ Innsbruck Medical University 
@@
 }
+\Plainauthor{Thomas Grubinger, Achim Zeileis, Karl-Peter Pfeiffer}
 
 \title{\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}}
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Thomas Grubinger, Achim Zeileis and Karl-Peter Pfeiffer} %% comma-separated
-\Plaintitle{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% without formatting
-\Shorttitle{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% a short title (if necessary)
+\Plaintitle{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
+\Shorttitle{\pkg{evtree}: Evolutionary Learning of Globally Optimal Trees in \proglang{R}}
 
-%% an abstract and keywords
 \Abstract{
-Commonly used classification and regression tree methods like the CART algorithm are recursive partitioning methods that build the model in a forward stepwise search.
-Although this approach is known to be an efficient heuristic, the results of recursive tree methods are only locally optimal, as splits are chosen to maximize homogeneity at the next step only.
-An alternative way to search over the parameter space of trees is to use global optimization methods like evolutionary algorithms.
-This paper describes the \pkg{evtree} package, which implements an evolutionary algorithm for learning globally optimal classification and regression trees in \proglang{R}.
-CPU and memory-intensive tasks are fully computed in \proglang{C++} while the \pkg{partykit} \citep{hothorn2009partykit} package is leveraged to represent the resulting trees in \proglang{R}, providing unified infrastructure for summaries, visualizations, and predictions.
-\code{evtree} is compared to \code{rpart} \citep{therneau1997introduction}, the open-source CART implementation, and \code{ctree} \citep{hothorn2006urp} (conditional inference trees).
-On 20 evaluated data-sets, including several benchmark problems from the UCI machine learning repository, \code{evtree} models offered at least similar and most of the time increased predictive accuracy.
+  Commonly used classification and regression tree methods like the CART
+  algorithm are recursive partitioning methods that build the model in a forward
+  stepwise search. Although this approach is known to be an efficient heuristic,
+  the results of recursive tree methods are only locally optimal, as splits are
+  chosen to maximize homogeneity at the next step only. An alternative way to
+  search over the parameter space of trees is to use global optimization methods
+  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
+  which implements an evolutionary algorithm for learning globally optimal
+  classification and regression trees in \proglang{R}. CPU and memory-intensive
+  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
+  \citep{hothorn2009partykit} package is leveraged to represent the resulting
+  trees in \proglang{R}, providing unified infrastructure for summaries,
+  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
+  \citep{therneau1997introduction}, the open-source CART implementation, and
+  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
+  evaluated data-sets, including several benchmark problems from the UCI machine
+  learning repository, \code{evtree} models offered at least similar and most of
+  the time increased predictive accuracy.  
 }
@@
 \Keywords{machine learning, classification trees, regression trees, evolutionary algorithms, \proglang{R}}
-\Plainkeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R} %% without formatting
-%% at least one keyword must be supplied
+\Plainkeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
 
-%% publication informatione
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{13}
-%% \Issue{9}
-%% \Month{September}
-%% \Year{2004}
-%% \Submitdate{2004-09-29}
-%% \Acceptdate{2004-09-29}
-
-%% The address of (at least) one author should be given
-%% in the following format:
 \Address{
@@
   Department for Medical Statistics, Informatics and Health Economics\\
-  6020, Innsbruck\\
+  6020, Innsbruck, Austria\\
   E-mail: \email{thomas.grubinger@i-med.ac.at}\\
@@
   Department of Statistics\\
-  6020, Innsbruck\\
+  6020, Innsbruck, Austria\\
   E-mail: \email{Achim.Zeileis@R-project.org}\\
@@
 }
-
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/1/31336-5053
-%% Fax: +43/1/31336-734
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 One solution, which is the tree constructed by \code{evtree} is given by:
-
+%
 << echo = FALSE, split= TRUE>>=
@@
 @
-\caption{Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
-\label{fig:illustration}
+\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
 \end{figure}
@@
 The arguments of  \code{evtree()} are
-
-\begin{verbatim}
-evtree(formula, data = list(), weights = NULL, subset = NULL, control = 
-evtree.control(...), ...)
-\end{verbatim}
-
+%
+\begin{Code}
+  evtree(formula, data = list(), weights = NULL, subset = NULL,
+    control = evtree.control(...), ...)
+\end{Code}
+%
 where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data.
@@
 \code{control} comprises a list of control parameters
-
-\begin{verbatim}
+%
+\begin{Code}
   evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
-niterations = 10000L, ntrees = 100L, alpha = 1, operatorprob = list(
-pmutatemajor = 0.2, pmutateminor = 0.2, pcrossover = 0.2, psplit = 0.2, 
-pprune = 0.2), seed = NULL, ...)
-\end{verbatim}
-
+    niterations = 10000L, ntrees = 100L, alpha = 1,
+    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
+      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
+    seed = NULL, ...)
+\end{Code}
+%
 where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth.
@@
 In this case, iterative calls of \code{evtree} will possibly result in different models.
-
+%
 <<>>=
@@
 @
+%
 The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
-
+%
 <<>>=
@@
 \setkeys{Gin}{width=1\textwidth}
-\begin{figure}[tbh!]
-    \begin{center}
+\begin{figure}[t!]
+\centering
 <<echo= FALSE>>=
@@
 @
-     \end{center}
-     \caption{Example of a tree plotted by the \code{plot()} function}
-     \label{fig:example}
+\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
 \end{figure}
@@
 The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
-
+%
 <<>>=
@@
 <<label=figciplots, echo = FALSE>>=
-rm(list=ls())
+rm(list = ls())
 library("plotrix")
@@
 \setkeys{Gin}{width=1\textwidth}
-\begin{figure}[tbh!]
+\begin{figure}[t!]
 \centering
@@
 @
-\caption{Performance comparison of \code{evtree} and \code{rpart}.
+\caption{\label{fig:crpart} Performance comparison of \code{evtree} and \code{rpart}.
 Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error.
 The complexity is compared by the relative difference of the number of terminal nodes.}
- \label{fig:crpart}
 \end{figure}
@@
 @
-\caption{Performance comparison of \code{evtree} and \code{ctree}.
+\caption{\label{fig:cptree} Performance comparison of \code{evtree} and \code{ctree}.
 Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error.
 The complexity is compared by the relative difference of the number of terminal nodes.}
-\label{fig:cptree}
 \end{figure}
@@
 @
-\caption{Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
-\label{fig:chess}
+\caption{\label{fig:chess} Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
 \end{figure}
@@
 <<label =  tab:resultsChessboard,  echo = FALSE, results=tex>>=
-library(xtable)
+library("xtable")
 load(file = "./results/chessboard44_0.RData")
@@
 
-\bibliography{article.bib}
+\bibliography{evtree}
 

### 2011-08-05 pkg/evtree/inst/doc/evtree.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@206 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Performance differences are displayed relative to \code{evtree}'s performance.
-For example, on the \textit{Glass} data-set, the average misclassification rate of \code{rpart} is $2.8\%$ higher than the misclassification rate of \code{evtree}.
-It can be observed that on 13 out of 17 data-sets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy.
+For example, on the \textit{Glass} data-set, the average misclassification rate of \code{rpart} is $2.7\%$ higher than the misclassification rate of \code{evtree}.
+It can be observed that on $12$ out of $17$ data-sets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy.
 Only on the \textit{Contraceptive Method} data-set \code{evtree} performs slightly worse.
-In terms of complexity, \code{evtree} models are significantly more complex on 9 and less complex on 7 data-sets.
+In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ data-sets.
 
 Figure~\ref{fig:cptree} summarizes the relative performance of \code{evtree} and \code{ctree}.
-For 14 out of 17 data-sets \code{evtree} shows a better predictive performance.
-The algorithms' performances on the 3 other data-sets is similar.
+For 15 out of 17 data-sets \code{evtree} shows a better predictive performance.
+The algorithms' performances  is significantly worse on the \textit{MEL0101} data-set, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower.
+However, on this data-set, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models.
 The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 data-set.
@@
 r <- rbind(r, preprocess(d = rspam, dname = "Spam", isclassification = T))
-r <- rbind(r, preprocess(d = rmagic, dname = "Magic Gamma Telescope", isclassification = T))
+r <- rbind(r, preprocess(d = rmagicgamma, dname = "Magic Gamma Telescope", isclassification = T))
 r <- rbind(r, preprocess(d = rservo, dname = "Servo", isclassification = F))
 r <- rbind(r, preprocess(d = rbostonhousing, dname = "Boston Housing", isclassification = F))
-r <- rbind(r, preprocess(d = mel0101, dname = "MEL0101", isclassification = F))
-r <- rbind(r, preprocess(d = hdg0202, dname = "HDG0202", isclassification = F))
+r <- rbind(r, preprocess(d = rmel0101, dname = "MEL0101", isclassification = F))
+r <- rbind(r, preprocess(d = rhdg0202, dname = "HDG0202", isclassification = F))
 r <- rbind(r, preprocess(d = rhdg0502, dname = "HDG0502", isclassification = F))
@@
     stats <- cstats(ra, alg)
-    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-12,45), axes=F, ann=T, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
+    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-11,46), axes=F, ann=T, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
     axis(side = 2, at = c(1:length(dummy)), labels= names(dummy), lwd=0.1)
-    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-10,-5,0,5,10,15,20,25,30,35,40))
+    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-10,-5,0,5,10,15,20,25,30,35,40,45))
     box(lwd=1)
@@
     stats <- cstats(rc, alg)
-    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-160,210), axes=F, ann=T, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
+    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-110,210), axes=F, ann=T, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
     axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-150,-100,-50,0,50,100,150,200))
@@
 The results are summarized in Table~\ref{tab:resultsChessboard}.
-It can be seen that, in the absence of noise, \code{rpart} models on average classify $67.2\%$ of the data points correctly and had $13.0$ terminal nodes.
-An average \code{ctree} model only has $1.1$ terminal nodes and a classification accuracy of $50.2\%$.
-In contrast, \code{evtree} classifies $92.8\%$ of the instances correctly and requires 14.0 terminal nodes.
-In the presence of $5\%$ and $10\%$ noise, \code{evtree} classifies $88.7\%$ and $84.2\%$ of the data correctly.
+It can be seen that, in the absence of noise, \code{rpart} models on average classify $69.1\%$ of the data points correctly and had $16.6$ terminal nodes.
+An average \code{ctree} model only has $1.1$ terminal nodes and a classification accuracy of $49.9\%$.
+In contrast, \code{evtree} classifies $93.2\%$ of the instances correctly and requires $14.4$ terminal nodes.
+In the presence of $5\%$ and $10\%$ noise, \code{evtree} classifies $89.0\%$ and $84.5\%$ of the data correctly.
 

### 2011-08-05 pkg/evtree/analysis/chessboard44.R

@@ -1,4 +1,4 @@
-chessboard44 <- function(totalpoints = 8000, noisevariables = 6, noise = 0){
+chessboard44 <- function(totalpoints = 4000, noisevariables = 6, noise = 0){
     chess44 <- array(0,c(totalpoints,noisevariables+3))
 
     for(i in 1:(noisevariables+2))
@@ -33,7 +33,7 @@ chessboard44 <- function(totalpoints = 8000, noisevariables = 6, noise = 0){
 
 
 
-benchchessboard44 <- function(totalpoints = 8000, noisevariables = 6, noise = 0, nrealizations = 250, seed = 1000){
+benchchessboard44 <- function(totalpoints = 4000, noisevariables = 6, noise = 0, nrealizations = 250, seed = 1000){
     set.seed(seed)
     accuracy_evtree <- array(-999999, nrealizations)
     accuracy_rpart <- array(-999999, nrealizations)

### 2011-09-07 pkg/partykit/inst/doc/partykit.Rnw
renamed vignette file and added introductory WeatherPlay example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@211 255f4249-2263-4417-b1ea-8498779e27a7

--- 2010-04-23 pkg/inst/doc/introduction.Rnw
@@
 %\VignetteDepends{partykit}
-%\VignetteKeywords{recursive partitioning}
+%\VignetteKeywords{recursive partitioning, regression trees, classification trees, decision trees}
 %\VignettePackage{partykit}
@@
 
-%% commands
+%% additional commands
+\newcommand{\squote}[1]{`{#1}'}
+\newcommand{\dquote}[1]{``{#1}''}
+\newcommand{\fct}[1]{{\texttt{#1()}\index{#1@\texttt{#1()}}}}
+\newcommand{\class}[1]{\dquote{\texttt{#1}}}
+
+%% further commands
 \renewcommand{\Prob}{\mathbb{P} }
@@
 \author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
-   \And Achim Zeileis\\Leopold-Franzens-Universit\"at \\ Innsbruck}
+   \And Achim Zeileis\\Universit\"at \\ Innsbruck}
 \Plainauthor{Torsten Hothorn, Achim Zeileis}
@@
 \Abstract{
-  The \pkg{partykit} package 
+  This vignette is intended to be a short technical introduction to the \pkg{partykit} package.
+  It is still unfinished but hopefully already helpful to some interested users.
+
+  The purpose of the package is to provide a toolkit with infrastructure for representing,
+  summarizing, and visualizing tree-structured regression and classification models.
+  Thus, the focus is not on \emph{inferring} such a tree structure from data but
+  to \emph{represent} a given tree so that print/plotting and computing predictions
+  can be performed in a standardized way. In particular, this unified infrastructure can be
+  used for reading/coercing tree models from different sources
+  (\pkg{rpart}, \pkg{RWeka}, PMML) yielding objects that share functionality for
+  \code{print}/\code{plot}/\code{predict} methods.
 }
-\Keywords{recursive partitioning}
+\Keywords{recursive partitioning, regression trees, classification trees, decision trees}
 
@@
   Ludwig-Maximilians-Universit\"at M\"unchen\\
-  Ludwigstra{\ss}e 33\\
+  Ludwigstr.~33\\
   80539 M\"unchen, Germany\\
@@
   Achim Zeileis\\
-  Institut f\"ur Statistik \\
-  Fakult\"at f\"ur Volkswirtschaft und Statistik \\
-  Leopold-Franzens-Universit\"at Innsbruck \\
-  Universit\"atsstra{\ss}e 15 \\
-  6020 Innsbruck, \"Osterreich \\
+  Department of Statistics \\
+  Universit\"at Innsbruck \\
+  Universit\"atsstr.~15 \\
+  6020 Innsbruck, Austria \\
   E-mail: \email{Achim.Zeileis@R-project.org}\\
-  URL: \url{http://statmath.wu.ac.at/~zeileis/}
+  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
 }
@@
 <<setup, echo = FALSE, results = hide>>=
-options(width = 70, SweaveHooks = list(leftpar = 
-    function() par(mai = par("mai") * c(1, 1.1, 1, 1))))
-require("partykit")
+options(width = 70)
+library("partykit")
 set.seed(290875)
@@
 
+\section{Introduction}
+
+To illustrate how \pkg{partykit} can be used to represent trees,
+a simple artificial data set from Witten \& Frank's book
+\emph{Data Mining: Practical Machine Learning Tools and Techniques}
+is used. It concerns the conditions suitable for playing some
+unspecified game.
+<<weather-data>>=
+data("WeatherPlay", package = "partykit")
+WeatherPlay
+@
+To represent the \code{play} decision based on the corresponding
+weather condition variables one could use the tree displayed in
+Figure~\ref{weather-plot}. For now, it is ignored how this tree
+was \emph{inferred} and it is simply assumed to be given.
+
+\setkeys{Gin}{width=0.8\textwidth}
+\begin{figure}[t!]
+\centering
+<<weather-plot0, echo=FALSE, fig=TRUE, height=5, width=7.5>>=
+py <- party(
+  partynode(1L,
+    split = partysplit(1L, index = 1:3),
+    kids = list(
+      partynode(2L,
+        split = partysplit(3L, breaks = 75),
+        kids = list(
+          partynode(3L, info = "yes"),
+          partynode(4L, info = "no"))),
+      partynode(5L, info = "yes"),
+      partynode(6L,
+        split = partysplit(4L, index = 1:2),
+        kids = list(
+          partynode(7L, info = "yes"),
+          partynode(8L, info = "no"))))),
+  WeatherPlay)
+plot(py)
+@
+\caption{\label{weather-plot} Decision tree for \code{play} decision based on
+  weather conditions in \code{WeatherPlay} data.}
+\end{figure}
 \setkeys{Gin}{width=\textwidth}
 
-\section{Introduction}
+To represent this tree (also known as recursive partition) in \pkg{partykit},
+two basic building blocks are used: splits of class \class{partysplit} and
+nodes of class \class{partynode}. The resulting recursive partition can then
+be associated with a data set in an object of class \class{party}.
+
+\subsubsection*{Splits}
+
+First, we employ the \fct{partysplit} function to creat the three
+splits in the ``weather tree'' from Figure~\ref{weather-plot}. The
+function takes the following arguments
+\begin{Code}
+  partysplit(varid, breaks = NULL, index = NULL, ..., info = NULL)
+\end{Code}
+where \code{varid} is an integer id of the variable used for splitting
+e.g., \code{1L} for \code{outlook}, \code{3L} for \code{humidity},
+\code{4L} for \code{windy} etc. Then, \code{breaks} and \code{index}
+determine which observations are sent to which of the branches, e.g.,
+\code{breaks = 75} for the humidity split. Apart from further arguments
+not shown above (and just comprised under `\code{...}'), some arbitrary
+information can be associated with a \class{partysplit} object by passing
+it to the \code{info} argument. The three splits from Figure~\ref{weather-plot}
+can then be created via
+<<weather-splits>>=
+sp_o <- partysplit(1L, index = 1:3)
+sp_h <- partysplit(3L, breaks = 75)
+sp_w <- partysplit(4L, index = 1:2)
+@
+For the numeric \code{humidity} variable the \code{breaks} are set while
+for the factor variables \code{outlook} and \code{windy} the information
+is supplied which of the levels should be associated with which of the
+branches of the tree.
+
+\subsubsection*{Nodes}
+
+Second, we use these splits in the creation of the whole decision tree.
+In \pkg{partykit} a tree is represented by a \class{partynode} object
+which is recursive in that it may have ``kids'' that are again \class{partynode}
+objects. These can be created with the function
+\begin{Code}
+  partynode(id, split = NULL, kids = NULL, ..., info = NULL)
+\end{Code}
+where \code{id} is an integer identifier of the node number, \code{split}
+is a \class{partysplit} object, and \code{kids} is a list of \class{partynode}
+objects. Again, there are further arguments not shown (\code{...}) and
+arbitrary information can be supplied in \code{info}. The whole tree
+from Figure~\ref{weather-plot} can then be created via
+<<weather-nodes>>=
+pn <- partynode(1L, split = sp_o, kids = list(
+  partynode(2L, split = sp_h, kids = list(
+    partynode(3L, info = "yes"),
+    partynode(4L, info = "no"))),
+  partynode(5L, info = "yes"),
+  partynode(6L, split = sp_w, kids = list(
+    partynode(7L, info = "yes"),
+    partynode(8L, info = "no")))))
+@
+where the previously created \class{partysplit} objects are used as splits
+and the nodes are simply numbered from~1 to~8. For the terminal nodes of the
+tree there are no \code{kids} and the corresponding \code{play} decision
+is stored in the \code{info} argument. Printing the \class{partynode} object
+reflects the recursive structure stored.
+<<weather-nodes-print>>=
+pn
+@
+However, the displayed information is still rather raw as it has not yet
+been associated with the \code{WeatherPlay} data set.
+
+\subsubsection*{Trees (or recursive partitions)}
+
+Hence, in a third step the recursive tree structure stored
+in \code{pn} is coupled with the corresponding data in a \class{party}
+object.
+<<weather-party>>=
+py <- party(pn, WeatherPlay)
+py
+@
+In addition to the \class{partynode} and the \class{data.frame},
+the function \fct{party} takes several further arguments including
+\code{fitted} values, a \code{terms} object, arbitrary additional
+\code{info}, and some further arguments (\code{...}):
+\begin{Code}
+  party(node, data, fitted = NULL, terms = NULL, ..., info = NULL)
+\end{Code}
+
+The tree \code{py} can then be easily visualized by
+<<weahter-plot, eval=FALSE>>=
+plot(py)
+@
+yielding Figure~\ref{weather-plot}.
+
+\subsubsection*{More}
+
+<<weather-party-print>>=
+print(py,
+  terminal_panel = function(node) paste(": play=", node$info, sep = ""))
+@
+
+%foo_node
+%foo_split
+
+%kidids_node(partynode_object)
+%kidids_split(partynode_object)
 
@@
 ## binary split in numeric variable `Sepal.Length'
-sl5 <- partysplit(which(names(iris) == "Sepal.Length"),
-    breaks = 5)
+sl5 <- partysplit(which(names(iris) == "Sepal.Length"), breaks = 5)
 class(sl5)
@@
 ## binary split in factor `Species'
-sp <- partysplit(which(names(iris) == "Species"),
-    index = c(1L, 1L, 2L))
+sp <- partysplit(which(names(iris) == "Species"), index = c(1L, 1L, 2L))
 character_split(sp, data = iris)
@@
 
+%%% code in eine Tabelle
+
 Although tree structures can be represented by \code{partynode} objects,
@@
   data = iris,
-   fitted = data.frame("(fitted)" = fitted_node(n1, data = iris),
+  fitted = data.frame(
+    "(fitted)" = fitted_node(n1, data = iris),
     "(response)" = iris$Species,
-                       check.names = FALSE))
+    check.names = FALSE)
+)
 t1
@@
 
-\section{My first Tree}
+\section{My first tree}
 
@@
   ### start recursion on this daugther node
-          kids[[kidid]] <- growtree(id = as.integer(myid + 1), 
-              response, data, w)
+  kids[[kidid]] <- growtree(id = as.integer(myid + 1), response, data, w)
   }
@@
   ### return nodes
-    return(partynode(id = as.integer(id), 
-                     split = sp, kids = kids))
+  return(partynode(id = as.integer(id), split = sp, kids = kids))
 }
@@
   ### data without missing values, response comes last
-    data <- data[complete.cases(data), 
-                 c(all.vars(formula)[-1], response)]
+  data <- data[complete.cases(data), c(all.vars(formula)[-1], response)]
   ### data is numeric
@@
     data = data,
-        fitted = data.frame("(fitted)" = fitted,
+    fitted = data.frame(
+      "(fitted)" = fitted,
       "(response)" = data[[response]],
@@
 
-\begin{figure}
-\begin{center}
+\begin{figure}[t!]
+\centering
 <<mytree-5, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
@@
 \caption{Tree. \label{plottree}}
-\end{center}
 \end{figure}
@@
 
-\begin{figure}
-\begin{center}
+\begin{figure}[t!]
+\centering
 <<mytree-8, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
@@
 \caption{Subtree. \label{subtree}}
-\end{center}
 \end{figure}
@@
 <<mytree-10, echo = TRUE>>=
-fun <- function(x) 
-    format.pval(info_split(split_node(x))$pvalue, 
+fun <- function(x) format.pval(info_split(split_node(x))$pvalue,
   digits = 3, eps = 0.001)

### 2011-09-07 pkg/partykit/R/ctree.R
fixed(?) small problems, see comments with ## Z
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@212 255f4249-2263-4417-b1ea-8498779e27a7

@@ -30,7 +30,10 @@
         X2 <- crossprod(tmp, Xplus$Xplus) %*% tmp
         df <- Xplus$rank
     }
-    if (pval) 
+    
+    X2 <- pmax(0, X2) ## Z: X2 may be slightly negative...
+    
+    if (pval)
         return(c(log(X2), pchisq(X2, df = df, lower.tail = TRUE, 
                                  log.p = TRUE)))
     return(c(log(X2), NA))
@@ -191,7 +194,7 @@
         kidids[xna] <- kidids_node(ret, data, obs = xna)
     }
 
-    kids <- vector(mode = "list", length = 1:max(kidids))
+    kids <- vector(mode = "list", length = max(kidids)) ## Z: was 1:max(kidids)
     nextid <- id + 1
     for (k in 1:max(kidids)) {
         w <- weights

### 2011-09-07 pkg/partykit/R/node.R

@@ -134,6 +134,20 @@ info_node <- function(node) {
     node$info
 }
 
+characterinfo_node <- function(node, default = "", prefix = NULL, ...) {
+    stopifnot(inherits(node, "partynode"))
+    info <- node$info
+    if(is.null(info)) info <- default
+    if(is.atomic(info)) info <- as.character(info) ## has class?
+    if(!is.character(info)) info <- capture.output(print(info), ...)
+    ## FIXME: check re-using of print method
+    if(!is.null(prefix)) {
+      info <- if(length(info) > 1L) c(prefix, info) else paste(prefix, info, sep = "")
+    }
+    info
+}
+
+
 ### FIXME: permutation and surrogate splits: is only the primary
 ### variable permuted?
 kidids_node <- function(node, data, vmatch = 1:ncol(data), obs = NULL, 

### 2011-09-07 pkg/partykit/R/plot.R

@@ -83,10 +83,7 @@ node_terminal <- function(obj,
 {
   nam <- names(obj)
 
-  extract_label <- function(node) {
-    return(c("terminal", "node"))
-    ## FIXME: re-use print method
-  }
+  extract_label <- function(node) characterinfo_node(node, default = c("terminal", "panel"))
 
   maxstr <- function(node) {
       lab <- extract_label(node)
@@ -102,20 +99,24 @@ node_terminal <- function(obj,
   rval <- function(node) {
     fill <- rep(fill, length.out = 2)	    
 
+    lab <- extract_label(node)
+
     node_vp <- viewport(x = unit(0.5, "npc"),	
       y = unit(0.5, "npc"),   
       width = unit(1, "strwidth", nstr) * 1.1,
-      height = unit(3, "lines"),
+      height = unit(length(lab) + 1, "lines"),
       name = paste("node_terminal", id_node(node), sep = ""),
       gp = gp
     )
     pushViewport(node_vp)
 
-    lab <- extract_label(node)
-
     grid.rect(gp = gpar(fill = fill[1]))
-    grid.text(y = unit(2, "lines"), lab[1])
-    grid.text(y = unit(1, "lines"), lab[2])
+    center <- TRUE ## FIXME: justification?
+    if(center) {    
+      for(i in seq_along(lab)) grid.text(y = unit(length(lab) - i + 1, "lines"), lab[i])
+    } else {
+      for(i in seq_along(lab)) grid.text(x = unit(1, "strwidth", "a"), y = unit(length(lab) - i + 1, "lines"), lab[i], just = "left")    
+    }    
 
     if(id) {
       nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, "npc"),

### 2011-09-07 pkg/partykit/R/print.R

@@ -8,6 +8,9 @@ print.partynode <- function(x, data = NULL, names = NULL,
   }
 
   if (length(x) > 0) {
+    ## add indentation
+    nextprefix <- paste(prefix, "|   ", sep = "")
+  
     ## split labels
     slabs <- character_split(split_node(x), data = data, digits = digits, ...)
     slabs <- ifelse(substr(slabs$levels, 1, 1) %in% c("<", ">"),
@@ -18,8 +21,18 @@ print.partynode <- function(x, data = NULL, names = NULL,
     knodes <- kids_node(x)    
     knam <- sapply(knodes, function(z) names[id_node(z)])
     klabs <- sapply(knodes, function(z)
-      if(is.terminal(z)) paste(terminal_panel(z), collapse = "\n")
-        else paste(inner_panel(z), collapse = "\n"))
+      if(is.terminal(z)) {
+        char <- terminal_panel(z)
+        if(length(char) > 1L) {
+	  paste(char[1L], "\n",
+	    paste(nextprefix, "    ", char[-1L], sep = "", collapse = "\n"),
+	    sep = "")
+	} else {
+	  char
+	}
+      } else {
+        paste(inner_panel(z), collapse = "\n")
+      })
 
     ## merge, cat, and call recursively
     labs <- paste("|   ", prefix, "[", knam, "] ", slabs, klabs, "\n", sep = "")          
@@ -27,13 +40,13 @@ print.partynode <- function(x, data = NULL, names = NULL,
       cat(labs[i])
       print.partynode(x[i], data = data, names = names,
         inner_panel = inner_panel, terminal_panel = terminal_panel,
-        prefix = paste(prefix, "|   ", sep = ""),  first = FALSE, digits = digits, ...)
+        prefix = nextprefix,  first = FALSE, digits = digits, ...)
     }
   }
 }
 
 print.party <- function(x,
-  terminal_panel = function(node) " *", tp_args = list(),
+  terminal_panel = function(node) characterinfo_node(node, default = "*", prefix = ": "), tp_args = list(),
   inner_panel = function(node) "", ip_args = list(),
   header_panel = function(party) "",
   footer_panel = function(party) "",

### 2011-09-07 pkg/partykit/inst/doc/partykit.Rnw
further improvements to vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@220 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
-   \And Achim Zeileis\\Universit\"at \\ Innsbruck}
+   \And Achim Zeileis\\Universit\"at Innsbruck}
 \Plainauthor{Torsten Hothorn, Achim Zeileis}
@@
   (\pkg{rpart}, \pkg{RWeka}, PMML) yielding objects that share functionality for
-  \code{print}/\code{plot}/\code{predict} methods.
+  \fct{print}/\fct{plot}/\fct{predict} methods.
 }
@@
 
-\section{Introduction}
+\section{Motivating example and overview} \label{sec:intro}
 
@@
 @
+And Figure~\ref{weather-plot} can easily be created by
+<<weahter-plot, eval=FALSE>>=
+plot(py)
+@
 In addition to the \class{partynode} and the \class{data.frame},
-the function \fct{party} takes several further arguments including
-\code{fitted} values, a \code{terms} object, arbitrary additional
-\code{info}, and some further arguments (\code{...}):
+the function \fct{party} takes several further arguments 
 \begin{Code}
@@
 \end{Code}
+i.e., \code{fitted} values, a \code{terms} object, arbitrary additional
+\code{info}, and again some further arguments comprised in \code{...}.
 
-The tree \code{py} can then be easily visualized by
-<<weahter-plot, eval=FALSE>>=
-plot(py)
-@
-yielding Figure~\ref{weather-plot}.
+\subsubsection*{Further information}
 
-\subsubsection*{More}
+More detailed technical information is provided in the subsequent sections.
+Section~\ref{sec:splits}, \ref{sec:nodes}, and~\ref{sec:trees} discuss
+the \fct{partysplit}, \fct{partynode}, and \fct{party} functions, respectively.
+Section~\ref{sec:application} illustrates how this infrastructure can
+be employed in a function that recursively \emph{infers} a tree.
 
+\subsubsection{Design principles}
+
+To facilitate reading of the subsequent sections, two design principles
+employed in the creation of \pkg{partykit} are briefly explained.
+
+(1)~Many helper utilities are encapsulated in functions that follow
+a simple naming convention. To extract/compute some information \emph{foo}
+from splits, nodes, or trees, \pkg{partykit} provides
+\emph{foo}\code{_split}, \emph{foo}\code{_node}, \emph{foo}\code{_party}
+functions (that are applicable to \class{partysplit}, \class{partynode},
+and \class{party} objects, repectively). An example for the information
+\emph{foo} might be \code{kidids}. Such functions are typically
+not to be called by the end-user but potentially by package designers that
+want to build functionality on top of \pkg{partykit}.
+
+(2)~Printing and plotting relies on \emph{panel functions} that visualize
+and/or format certain aspects of the resulting display, e.g., that of inner nodes,
+terminal nodes, headers, footers, etc. A simple example would be printing
+with a custom panel function for formatting the terminal node:
 <<weather-party-print>>=
@@
 @
+Furthermore, arguments like \code{terminal_panel} can also take
+\emph{panel-generating functions}, i.e., functions that produce a panel
+function when applied to the \class{party} object.
 
-%foo_node
-%foo_split
-
-%kidids_node(partynode_object)
-%kidids_split(partynode_object)
-
-\section{Splits}
+\section{Splits} \label{sec:splits}
 
@@
 more specifically a partitioning variable, to daugther nodes. 
-Objects of class \code{partysplit} are designed to represent 
-such functions and are set-up by the \code{partysplit()} 
+Objects of class \class{partysplit} are designed to represent 
+such functions and are set-up by the \fct{partysplit} 
 constructor:
@@
 @
-The internal structure of class \code{partysplit} contains
+The internal structure of class \class{partysplit} contains
 information about the partitioning variable, the split-points,
@@
 Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
-($(\mu_1, \dots, \mu_{l - 1})$ = \code{break}). We are interested to split according to the 
+($(\mu_1, \dots, \mu_{l - 1})$ = \code{breaks}). We are interested to split according to the 
 information contained in the $i$th element of $x$ ($i$ = \code{varid}).
@@
 
-The slots of \code{partysplit} objects shall be accessed by the corresponding
+The slots of \class{partysplit} objects shall be accessed by the corresponding
 accessor functions.
 
-\section{Nodes}
+\section{Nodes} \label{sec:nodes}
 
-Inner and terminal nodes are represented by objects of class \code{partynode}.
+Inner and terminal nodes are represented by objects of class \class{partynode}.
 Each node has a unique identifier \code{id}. A node consisting only
@@
 Inner nodes have to have a primary split \code{split} and at least two
-daugther nodes. The daugther nodes are objects of class \code{partynode}
+daugther nodes. The daugther nodes are objects of class \class{partynode}
 itself and thus represent the recursive nature of this data structure.
 The daugther nodes are pooled in a list \code{kids}. 
-In addition, a list of \code{partysplit} objects offering 
-surrogate splits can be supplied; a list of \code{partysplit} objects
+In addition, a list of \class{partysplit} objects offering 
+surrogate splits can be supplied; a list of \class{partysplit} objects
 in slot \code{surrogates} defines such additional splits (mostly used 
-for handling missing values). Note that \code{partynode} objects aren't 
+for handling missing values). Note that \class{partynode} objects aren't 
 connected to the actual data.
@@
 
-A number of methods are defined for \code{partynode} objects.
-\code{is.partynode} checks if the argument is a valid \code{partynode}
-object. \code{is.terminal} is \code{TRUE} for terminal nodes
+A number of methods are defined for \class{partynode} objects.
+\fct{is.partynode} checks if the argument is a valid \class{partynode}
+object. \fct{is.terminal} is \code{TRUE} for terminal nodes
 and \code{FALSE} for inner nodes. The subset methods
-return the \code{partynode} object corresponding to the \code{i}th
+return the \class{partynode} object corresponding to the \code{i}th
 kid:
@@
 
-The \code{as.partynode} and \code{as.list} methods can be used
-to convert flat list structures into recursive \code{partynode}
-objects and vice versa. \code{as.partynode} applied to
-\code{partynode} objects renumbers the recursive nodes
+The \fct{as.partynode} and \fct{as.list} methods can be used
+to convert flat list structures into recursive \class{partynode}
+objects and vice versa. \fct{as.partynode} applied to
+\class{partynode} objects renumbers the recursive nodes
 starting with root node identifier \code{from}.
 
-\code{length} gives the number of kid nodes of the root node,
-\code{depth} the depth of the tree and \code{width}
+\fct{length} gives the number of kid nodes of the root node,
+\fct{depth} the depth of the tree and \fct{width}
 the number of terminal nodes.
 
-\section{Trees}
+\section{Trees} \label{sec:trees}
 
@@
 
-Although tree structures can be represented by \code{partynode} objects,
+Although tree structures can be represented by \class{partynode} objects,
 a tree is more than a number of nodes and splits. More
@@
 
-Objects of class \code{party} basically consist of a \code{partynode}
+Objects of class \class{party} basically consist of a \class{partynode}
 object representing the tree structure in a recursive way and 
-data. The \code{data} argument takes a \code{data.frame} which, however,
-might have zero columns. Optionally, a \code{data.frame} with at least one
+data. The \code{data} argument takes a \class{data.frame} which, however,
+might have zero columns. Optionally, a \class{data.frame} with at least one
 variable \code{(fitted)} containing the terminal node numbers of
@@
 
-\section{My first tree}
+\section{My first tree} \label{sec:application}
 
@@
 we have to set-up the recursion for growing a recursive 
-\code{partynode} structure:
+\class{partynode} structure:
 <<mytree-2, echo = TRUE>>=
@@
 A very rough sketch of formula-based user-interface needs
-to set-up the data and call \code{growtree()}:
+to set-up the data and call \fct{growtree}:
 <<mytree-3, echo = TRUE>>=
@@
 We now can fit this tree, for example to the airquality data; the
-\code{print()} method provides us with a first overview on the 
+\fct{print} method provides us with a first overview on the 
 resulting model
@@
 
-We depict the model graphically using \code{plot()} (see Figure~\ref{plottree})
+We depict the model graphically using \fct{plot} (see Figure~\ref{plottree})
 and compute predictions using
@@
 @
-which again are objects inheriting from \code{party} and thus
+which again are objects inheriting from \class{party} and thus
 can be plotted easily (see Figure~\ref{subtree}).

### 2011-09-08 pkg/evtree/inst/doc/evtree.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@208 255f4249-2263-4417-b1ea-8498779e27a7
Avoid using T/F, use TRUE/FALSE instead.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@222 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \caption{Description of the evaluated benchmark data-sets.
-The data-sets marked with a $(^*)$ are from the UCI machine learning repository \citep{Frank+Asuncion:2010}.
+The data-sets marked with a $(^*)$ originate from the UCI machine learning repository \citep{Frank+Asuncion:2010} and are made available in the \pkg{evtree} package.
 Data-sets marked with a $(^\#)$ and $(^{+})$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and \pkg{kernlab} \citep{kernlab}.
@@
 preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
-    if(isclassification == TRUE)
+    if(isclassification)
         d[,1:3] <- 1-d[,1:3]
@@
 
-r <- preprocess(d = rglass, dname = "Glass Identification", isclassification = T)
-r <- rbind(r, preprocess(d = rheart, dname = "Statlog Heart", isclassification = T))
-r <- rbind(r, preprocess(d = rionosphere, dname = "Ionosphere", isclassification = T))
-r <- rbind(r, preprocess(d = rmusk, dname = "Musk", isclassification = T))
-r <- rbind(r, preprocess(d = rbreastcancer, dname = "Breast Cancer Database", isclassification = T))
-r <- rbind(r, preprocess(d = rpima, dname = "Pima Indians Diabetes", isclassification = T))
-r <- rbind(r, preprocess(d = rvowel, dname = "Vowel", isclassification = T))
-r <- rbind(r, preprocess(d = rcredit, dname = "Statlog German Credit", isclassification = T))
-r <- rbind(r, preprocess(d = rcontraceptive, dname = "Contraceptive Method", isclassification = T))
-r <- rbind(r, preprocess(d = rdna, dname = "DNA", isclassification = T))
-r <- rbind(r, preprocess(d = rspam, dname = "Spam", isclassification = T))
-r <- rbind(r, preprocess(d = rmagicgamma, dname = "Magic Gamma Telescope", isclassification = T))
-r <- rbind(r, preprocess(d = rservo, dname = "Servo", isclassification = F))
-r <- rbind(r, preprocess(d = rbostonhousing, dname = "Boston Housing", isclassification = F))
-r <- rbind(r, preprocess(d = rmel0101, dname = "MEL0101", isclassification = F))
-r <- rbind(r, preprocess(d = rhdg0202, dname = "HDG0202", isclassification = F))
-r <- rbind(r, preprocess(d = rhdg0502, dname = "HDG0502", isclassification = F))
+r <- preprocess(d = rglass, dname = "Glass Identification", isclassification = TRUE)
+r <- rbind(r, preprocess(d = rheart, dname = "Statlog Heart", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rionosphere, dname = "Ionosphere", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rmusk, dname = "Musk", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rbreastcancer, dname = "Breast Cancer Database", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rpima, dname = "Pima Indians Diabetes", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rvowel, dname = "Vowel", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rcredit, dname = "Statlog German Credit", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rcontraceptive, dname = "Contraceptive Method", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rdna, dname = "DNA", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rspam, dname = "Spam", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rmagicgamma, dname = "Magic Gamma Telescope", isclassification = TRUE))
+r <- rbind(r, preprocess(d = rservo, dname = "Servo", isclassification = FALSE))
+r <- rbind(r, preprocess(d = rbostonhousing, dname = "Boston Housing", isclassification = FALSE))
+r <- rbind(r, preprocess(d = rmel0101, dname = "MEL0101", isclassification = FALSE))
+r <- rbind(r, preprocess(d = rhdg0202, dname = "HDG0202", isclassification = FALSE))
+r <- rbind(r, preprocess(d = rhdg0502, dname = "HDG0502", isclassification = FALSE))
 
@@
     stats <- cstats(ra, alg)
-    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-11,46), axes=F, ann=T, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
+    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-11,46), axes=FALSE, ann=TRUE, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
     axis(side = 2, at = c(1:length(dummy)), labels= names(dummy), lwd=0.1)
@@
     stats <- cstats(rc, alg)
-    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-110,210), axes=F, ann=T, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
+    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-110,210), axes=FALSE, ann=TRUE, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
     axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-150,-100,-50,0,50,100,150,200))
@@
      if( noise > 0){
-        flipclasslist <- sample(totalpoints, totalpoints * (noise / 100), replace = F)
+        flipclasslist <- sample(totalpoints, totalpoints * (noise / 100), replace = FALSE)
 

### 2011-09-08 pkg/evtree/inst/doc/evtree.Rnw
cleanup of R code, figures, tables
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@225 255f4249-2263-4417-b1ea-8498779e27a7

--- 2011-08-05 pkg/evtree/inst/doc/evtree.Rnw
@@
 %\VignetteIndexEntry{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
-%\VignetteDepends{stats,evtree,partykit,mlbench,plotrix,lme4,multcomp,rpart,ctree,xtable}
+%\VignetteDepends{stats,evtree,partykit,mlbench,multcomp,rpart,party,xtable}
 %\VignetteKeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
@@
   \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
-  evaluated data-sets, including several benchmark problems from the UCI machine
+  evaluated datasets, including several benchmark problems from the UCI machine
   learning repository, \code{evtree} models offered at least similar and most of
@@
   6020, Innsbruck, Austria\\
-  E-mail: \email{thomas.grubinger@i-med.ac.at}\\
-  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger}\\
+  E-mail: \email{Thomas.Grubinger@i-med.ac.at}, \email{Karl-Peter.Pfeiffer@i-med.ac.at}\\
+  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger/}\\
+  \phantom{URL: }\url{http://www.i-med.ac.at/msig/mitarbeiter/pfeiffer/}\\
 
@@
 }
+
+<<setup, echo=FALSE>>=
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+library("evtree")
+@
 
@@
 
-The remainder of this paper is structured as follows.
-Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms.
-In Section~\ref{desc} the \code{evtree} algorithm will be introduced.
-Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions.
-In Section~\ref{comp} \code{evtree} is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and conditional inference trees in \proglang{R}.
-The comparison comprises 14 benchmark data-sets, 3 real-world data-sets and 3 simulated data-sets.
+The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it
+to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will
+be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree}
+is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and
+conditional inference trees in \proglang{R}.
+The comparison comprises 14~benchmark datasets, 3~real-world datasets and 3~simulated datasets.
 It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods.
@@
 %
-<< echo = FALSE, split= TRUE>>=
-library("evtree")
-X1 <- c(rep(0.25, 4), rep(0.75, 4), rep(1.25 ,4),  rep(1.75, 4))
-X2 <- c(rep(seq(0.25, 1.75 ,0.5),4))
-Y <- array(1,16)
-Y[X1 < 1 & X2 < 1]=2
-Y[X1 > 1 & X2 > 1]=2
-chess22 <- as.data.frame(cbind(X1, X2, Y))
-chess22[,3] <- as.factor( chess22[,3] )
-levels(chess22[,3]) = c("class: o","class: x")
-print(evtree(Y~.,data=chess22, minbucket = 1, minsplit=2))
+<<chess22, echo=FALSE>>=
+X1 <- rep(seq(0.25, 1.75, 0.5), each = 4)
+X2 <- rep(seq(0.25, 1.75, 0.5), 4)
+Y <- rep(1, 16)
+Y[(X1 < 1 & X2 < 1) | (X1 > 1 & X2 > 1)] <- 2
+Y <- factor(Y, labels = c("O", "X"))
+chess22 <- data.frame(Y, X1, X2)
+print(evtree(Y ~ ., data = chess22, minbucket = 1, minsplit = 2))
 @
 
-\setkeys{Gin}{width=0.4\textwidth}
-\begin{figure}[tbh!]
+\setkeys{Gin}{width=0.5\textwidth}
+\begin{figure}[t!]
 \centering
-<<fig = TRUE, echo = FALSE>>=
-par(cex.axis=1.5, cex.lab=1.5)
-plot(chess22[chess22[,3]=="class: o",1], chess22[chess22[,3]=="class: o",2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", xlim=c(0,2), ylim=c(0,2), cex=2)
-points(chess22[chess22[,3]=="class: x",1], chess22[chess22[,3]=="class: x",2], pch= c("o"), cex=2)
-par(cex.axis=1.5, cex.lab=1.5)
+<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
+plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
 @
-\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
+\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
 \end{figure}
 
-All instances all are classified correctly.
-Each of the terminal nodes 3 and 7 contain four instances of the class $(x)$.
-Four instances of class $(o)$ are assigned to each of the terminal nodes 4 and 6.
+All instances all are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
+\code{O} are assigned to each of the terminal nodes~4 and~6.
 
@@
 \cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees.
-\cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several data-sets from the UCI machine learning repository \citep{Frank+Asuncion:2010}.
+\cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several datasets from the UCI machine learning repository \citep{Frank+Asuncion:2010}.
 A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.
@@
 
-\begin{table}[tbh!]
-\begin{center}
-\begin{tabular}{l}
-\toprule
-1. initialize the population \\
-2. evaluate each individual  \\
-3. while(termination condition is not satisfied) do  \\
-\hspace{8mm} a. select parents  \\
-\hspace{8mm} b. alter selected individuals via variation operators \\
-\hspace{8mm} c. evaluate new solutions  \\
-\hspace{8mm} d. select survivors for the next generation  \\
-\hspace{3.8mm} od  \\
-\bottomrule
-\end{tabular}
-\end{center}
+\begin{table}[t!]
+\centering
+\hrulefill
+\begin{enumerate}
+  \item Initialize the population.
+  \item Evaluate each individual.
+  \item While(termination condition is not satisfied) do:
+  \begin{itemize}
+    \item[a.] Select parents.
+    \item[b.] Alter selected individuals via variation operators.
+    \item[c.] Evaluate new solutions.
+    \item[d.] Select survivors for the next generation.
+  \end{itemize}
+\end{enumerate}
+\hrulefill
 \caption{Pseudocode of the general schema of an evolutionary algorithm.} 
@@
 
-Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian Information Criteria (BIC)} \citep{gray2008cta} and \textit{Minimum Description Length (MDL)} \citep{quinlan1989inferring}.
+Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian information criterion} \citep[BIC, as in][]{gray2008cta} and
+\textit{minimum description length} \citep[MDL, as in][]{quinlan1989inferring}.
 For both evaluation functions deviance is used for accuracy estimation.
@@
 
-For regression trees, accuracy is usually measured by the mean squared error $MSE(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$.
-Using $N \cdot \ln MSE(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:
+For regression trees, accuracy is usually measured by the mean squared error $\mathit{MSE}(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$.
+Using $N \cdot \ln \mathit{MSE}(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:
 
 \begin{equation}
-	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln MSE(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
+	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
 \label{eq:regression}
@@
 %
-<<>>=
+<<iris-tree>>=
 library("evtree")
-data("iris")
-model <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
+data("iris", package = "datasets")
+ev <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
 @
@@
 %
-<<>>=
-print(model)
+<<iris-print>>=
+ev
+plot(ev)
 @
@@
 \centering
-<<echo= FALSE>>=
-par(cex.axis=1.5, cex.lab=1.5)
-<<fig=TRUE, echo=TRUE, width = 13.5, height= 8.5>>=
-plot(model)
+<<iris-plot, fig=TRUE, echo=FALSE, width=13.2, height=8>>=
+plot(ev)
 @
@@
 %
-<<>>=
-predict(model, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
+<<iris-predict>>=
+predict(ev, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
 @
@@
 
-In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark data-sets that are publicly available and $3$ real-world data-sets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}.
-The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 data-sets.
+In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark datasets that are publicly available and $3$ real-world datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}.
+The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 datasets.
 The misclassification rate on the \textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of predictive accuracy.
@@
 Missing values are currently not supported by \code{evtree}.
-The only data-set with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis.
-Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual data-sets .
+The only dataset with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis.
+Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual datasets .
 
@@
 
-In Table~\ref{tab:dataDescription} the benchmark and real-world data-sets from the Austrian DRG-system are described.
+In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from the Austrian DRG-system are described.
 In the Austrian DRG-system, resources are allocated to hospitals by simple rules regarding diagnosis, selected individual services, and patient characteristics.
 Within this system regression tree analysis is performed to classify hospital patients into clinically meaningful and comprehensible groups that consume similar hospital resources.
-The data-sets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure.
+The datasets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure.
 The task of the regression tree analysis is to further divide these groups with the intention of finding groups with similar resource consumption.
@@
 
-\begin{table}
-\begin{center}
+\begin{table}[t!]
+\centering
 \begin{tabular}{lrrrrrr}
 \toprule
-\textsc{Data-set} & \textsc{Instances} &  \multicolumn{5}{l}{\textsc{Attributes}}   \\ 
-                  &                    &  binary              & nominal          &  ordered      &  metric     & classes \\ 
+{Dataset} & {Instances} &  \multicolumn{5}{l}{{Attributes}}   \\ 
+           &             &  Binary              & Nominal          &  Ordered      &  Metric     & Classes \\ 
 \midrule
-Glass Identification$^\#$  &   214   &  - &  - &  - & 9 &  6      	 \\
-Statlog Heart*         &   270   &  3 &   3 & 1 & 6 &   2     	 \\
+Glass identification$^\#$  &   214   &  - &  - &  - & 9 &  6      	 \\
+Statlog heart*         &   270   &  3 &   3 & 1 & 6 &   2     	 \\
 Ionosphere$^\#$ &   351   &  2 &  - &  - & 32 & 2  	             \\
 Musk$^+$  &  476  & - &  - &  - & 166 & 2                     	 \\
-Breast Cancer Database$^\#$ &   685   &  - &  4 & 5 &  - &  2  	 \\
-Pima Indians Diabetes$^\#$ &   768   &  - &   - & - & 8 &  2      	 \\
+Breast cancer database$^\#$ &   685   &  - &  4 & 5 &  - &  2  	 \\
+Pima Indians diabetes$^\#$ &   768   &  - &   - & - & 8 &  2      	 \\
 Vowel$^\#$  &  990  & - &  1 &  - & 9 & 11     	                     \\
-Statlog German Credit* &  1000   &  2 &  10 & 1 & 7 &  2      	 \\
-Contraceptive Method* &   1437   & 3 &  - &  4 &  2 &  3  	     \\
+Statlog German credit* &  1000   &  2 &  10 & 1 & 7 &  2      	 \\
+Contraceptive method* &   1437   & 3 &  - &  4 &  2 &  3  	     \\
 DNA$^\#$  &  3186  & 180 &  - &  - & - & 3   	                       \\
 Spam$^+$  &  4601  & - &  - &  - & 57 & 2                      	 \\
-MAGIC Gamma Telescope*  &  19020  & - &  - &  - & 10 & 2     	   \\
+MAGIC gamma telescope*  &  19020  & - &  - &  - & 10 & 2     	   \\
 Servo$^\#$ & 167 &  - & 4 & - &  -  &  -											   \\
-Boston Housing$^\#$ & 506 & 1 & - & - &  12  & -                 \\
+Boston housing$^\#$ & 506 & 1 & - & - &  12  & -                 \\
 MEL0101$^\diamond$ &   875 &  1 & 4 & 1 &  108  & -              \\
@@
 \end{tabular}
-\end{center}
-\caption{Description of the evaluated benchmark data-sets.
-The data-sets marked with a $(^*)$ are from the UCI machine learning repository \citep{Frank+Asuncion:2010}.
-Data-sets marked with a $(^\#)$ and $(^{+})$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and \pkg{kernlab} \citep{kernlab}.
-The $3$ real-world data-sets from the Austrian DRG system are marked with a $(^\diamond)$.
-}
-\label{tab:dataDescription}
+\caption{\label{tab:dataDescription} Description of the evaluated benchmark datasets. The datasets marked with
+  $*$ originate from the UCI machine learning repository \citep{Frank+Asuncion:2010} and are made available in the \pkg{evtree} package.
+  Datasets marked with $\#$ and $+$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and
+  \pkg{kernlab} \citep{kernlab}, respectively. The three real-world datasets from the Austrian DRG system are marked with $\diamond$.} 
 \end{table}
 
-The relative performance of \code{evtree} and \code{rpart} is summarized in Figure~\ref{fig:crpart}.
+The relative performance of \code{evtree} and \code{rpart} is summarized in Figure~\ref{fig:benchmark} (upper panels).
 Performance differences are displayed relative to \code{evtree}'s performance.
-For example, on the \textit{Glass} data-set, the average misclassification rate of \code{rpart} is $2.7\%$ higher than the misclassification rate of \code{evtree}.
-It can be observed that on $12$ out of $17$ data-sets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy.
-Only on the \textit{Contraceptive Method} data-set \code{evtree} performs slightly worse.
-In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ data-sets.
+For example, on the \textit{Glass} dataset, the average misclassification rate of \code{rpart} is $2.7\%$ higher than the misclassification rate of \code{evtree}.
+It can be observed that on $12$ out of $17$ datasets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy.
+Only on the \textit{Contraceptive Method} dataset \code{evtree} performs slightly worse.
+In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ datasets.
 
-Figure~\ref{fig:cptree} summarizes the relative performance of \code{evtree} and \code{ctree}.
-For 15 out of 17 data-sets \code{evtree} shows a better predictive performance.
-The algorithms' performances  is significantly worse on the \textit{MEL0101} data-set, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower.
-However, on this data-set, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models.
-The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 data-set.
+Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}.
+For 15 out of 17 datasets \code{evtree} shows a better predictive performance.
+The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower.
+However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models.
+The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.
 
-%% vorberechnungen
-<<label=figciplots, echo = FALSE>>=
-rm(list = ls())
-library("plotrix")
-library("lme4")
-library("multcomp")
-load(file="./results/breastcancer.RData")
-load(file="./results/glass.RData")
-load(file="./results/credit.RData")
-load(file="./results/heart.RData")
-load(file="./results/pima.RData")
-load(file="./results/contraceptive.RData")
-load(file="./results/dna.RData")
-load(file="./results/musk.RData")
-load(file="./results/spam.RData")
-load(file="./results/vowel.RData")
-load(file="./results/ionosphere.RData")
-load(file="./results/magicgamma.RData")
-load(file="./results/hdg0202.RData")
-load(file="./results/hdg0502.RData")
-load(file="./results/mel0101.RData")
-load(file="./results/bostonhousing.RData")
-load(file="./results/servo.RData")
+<<benchmark-results, echo=FALSE>>=
+## load results
+for(i in Sys.glob("results/*.RData")) load(i)
 
+## preprocess for reference evtree
 preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
-    if(isclassification == TRUE)
-        d[,1:3] <- 1-d[,1:3]
+    if(isclassification) d[, 1:3] <- 1 - d[ ,1:3]
     d <- as.data.frame(d)
     colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")
-    d[,2] <- d[,2] / d[,1] *100
-    d[,3] <- d[,3] / d[,1] *100
-    d[,1] <- d[,1] / d[,1] *100
-    d[,5] <- d[,5] / d[,4] *100
-    d[,6] <- d[,6] / d[,4] *100
-    d[,4] <- d[,4] / d[,4] *100
-    x <- d[ ,1:3]
-    y <- d[ ,4:6]
-    x <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
-    y <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")
-    x$ds <- dname
-    y$ds <- dname
-    return( cbind(x,y) )
+    for(i in 3:1) d[, i] <- d[, i] / d[, 1] * 100
+    for(i in 6:4) d[, i] <- d[, i] / d[, 4] * 100
+    x <- d[, 1:3]
+    y <- d[, 4:6]
+    rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
+    names(rval)[2] <- "accuracy"
+    rval$complexity <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")[,2]
+    rval$alg <- factor(rval$alg, levels = c("evtree", "ctree", "rpart"))
+    rval$ds <- dname
+    rval
 }
 
-r <- preprocess(d = rglass, dname = "Glass Identification", isclassification = T)
-r <- rbind(r, preprocess(d = rheart, dname = "Statlog Heart", isclassification = T))
-r <- rbind(r, preprocess(d = rionosphere, dname = "Ionosphere", isclassification = T))
-r <- rbind(r, preprocess(d = rmusk, dname = "Musk", isclassification = T))
-r <- rbind(r, preprocess(d = rbreastcancer, dname = "Breast Cancer Database", isclassification = T))
-r <- rbind(r, preprocess(d = rpima, dname = "Pima Indians Diabetes", isclassification = T))
-r <- rbind(r, preprocess(d = rvowel, dname = "Vowel", isclassification = T))
-r <- rbind(r, preprocess(d = rcredit, dname = "Statlog German Credit", isclassification = T))
-r <- rbind(r, preprocess(d = rcontraceptive, dname = "Contraceptive Method", isclassification = T))
-r <- rbind(r, preprocess(d = rdna, dname = "DNA", isclassification = T))
-r <- rbind(r, preprocess(d = rspam, dname = "Spam", isclassification = T))
-r <- rbind(r, preprocess(d = rmagicgamma, dname = "Magic Gamma Telescope", isclassification = T))
-r <- rbind(r, preprocess(d = rservo, dname = "Servo", isclassification = F))
-r <- rbind(r, preprocess(d = rbostonhousing, dname = "Boston Housing", isclassification = F))
-r <- rbind(r, preprocess(d = rmel0101, dname = "MEL0101", isclassification = F))
-r <- rbind(r, preprocess(d = rhdg0202, dname = "HDG0202", isclassification = F))
-r <- rbind(r, preprocess(d = rhdg0502, dname = "HDG0502", isclassification = F))
+## collect results for all datasets
+r <- rbind(
+  preprocess(d = rglass, dname = "Glass identification", isclassification = TRUE),
+  preprocess(d = rheart, dname = "Statlog heart", isclassification = TRUE),
+  preprocess(d = rionosphere, dname = "Ionosphere", isclassification = TRUE),
+  preprocess(d = rmusk, dname = "Musk", isclassification = TRUE),
+  preprocess(d = rbreastcancer, dname = "Breast cancer database", isclassification = TRUE),
+  preprocess(d = rpima, dname = "Pima Indians diabetes", isclassification = TRUE),
+  preprocess(d = rvowel, dname = "Vowel", isclassification = TRUE),
+  preprocess(d = rcredit, dname = "Statlog German credit", isclassification = TRUE),
+  preprocess(d = rcontraceptive, dname = "Contraceptive method", isclassification = TRUE),
+  preprocess(d = rdna, dname = "DNA", isclassification = TRUE),
+  preprocess(d = rspam, dname = "Spam", isclassification = TRUE),
+  preprocess(d = rmagicgamma, dname = "Magic gamma telescope", isclassification = TRUE),
+  preprocess(d = rservo, dname = "Servo", isclassification = FALSE),
+  preprocess(d = rbostonhousing, dname = "Boston housing", isclassification = FALSE),
+  preprocess(d = rmel0101, dname = "MEL0101", isclassification = FALSE),
+  preprocess(d = rhdg0202, dname = "HDG0202", isclassification = FALSE),
+  preprocess(d = rhdg0502, dname = "HDG0502", isclassification = FALSE)
+)
+r$ds <- factor(r$ds)
+r$samp <- factor(r$samp)
+r$dssamp <- r$ds:r$samp
 
-ra <- r[,1:4]
-rc <- r[,5:8]
-
-tf <- function(x){
-    names(x)[[2]] <- "value"
-    x$samp <- factor(x$samp) 
-    x$ds <- factor(x$ds)
-    x$dssamp <- x$ds:x$samp # unique levels of dataset and bootstrap sample
-    x$alg <- factor(x$alg)
-    x$alg <- relevel(x$alg, "evtree") # change order of algorithms
-    x
-}
-ra <- tf(ra)
-rc <- tf(rc)
-
-uniquedatasetlabels <- unique(ra$ds)[length(unique(ra$ds)):1]
-
-cstats <- function(x= ra, alg= "rpart"){
-    mean  <- array(0,17)
-    lower <- array(0,17)
-    upper <- array(0,17)
-    for(i in 1:length(uniquedatasetlabels)){
-        mod1 <- lm(value ~ alg, data = subset(x, x$ds == uniquedatasetlabels[i]))
+## compute multiple comparisons
+library("multcomp")
+cstats <- function(alg = "rpart", value = "accuracy", data = r) {
+  dlab <- rev(unique(data$ds))
+  k <- length(dlab)
+  mean  <- numeric(k)
+  lower <- numeric(k)
+  upper <- numeric(k)
+  names(data)[names(data) == value] <- "value"
+  for(i in 1:k) {
+    mod1 <- lm(value ~ alg, data = subset(data, ds == dlab[i]))
     pt <- glht(mod1, linfct = mcp(alg = "Dunnett"))
     w <- confint(pt)$confint
-        if(alg == "rpart")
-            d <- 1
-        else
-            d <- 0
+    d <- which(levels(r$alg) == alg) - 2
     mean[i]  <-  w[1+d]
@@
   }
-    return(as.data.frame(cbind(mean,lower,upper)))
+  rval <- data.frame(mean, lower, upper)
+  rownames(rval) <- dlab
+  return(rval)
 }
 
-plotd <- function(alg){
-    par(las=1, cex=1.5, cex.lab=1.3, mfrow= c(1,2), omd=c(0.16,1,0,1), mar=c(5, 0, 1, 2) + 0.1)
-    dummy <- c(1:17)
-    names(dummy) <- uniquedatasetlabels
-    stats <- cstats(ra, alg)
-    plotCI(stats$mean, dummy, li=stats$lower, ui=stats$upper, err="x", xlim=c(-11,46), axes=F, ann=T, ylab= "", xlab= "difference in predictive accuracy [%]", lwd=1.5 )
-    axis(side = 2, at = c(1:length(dummy)), labels= names(dummy), lwd=0.1)
-    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-10,-5,0,5,10,15,20,25,30,35,40,45))
-    box(lwd=1)
-    abline(v=0, lty=2)
-    abline(h=5.5, lty=1, lwd=0.5)
-    stats <- cstats(rc, alg)
-    plotCI(stats$mean, dummy, li= stats$lower, ui= stats$upper, err="x", xlim=c(-110,210), axes=F, ann=T, ylab= "", xlab= "difference in complexity [%]", lwd=1.5 )
-    axis(side = 1, lwd=0.1, lwd.ticks=1, at=c(-150,-100,-50,0,50,100,150,200))
-    box(lwd=1)
-    abline(v=0, lty=2)
-    abline(h=5.5, lty=1, lwd=0.5)
+acc_rpart <- cstats("rpart", "accuracy")
+com_rpart <- cstats("rpart", "complexity")
+acc_ctree <- cstats("ctree", "accuracy")
+com_ctree <- cstats("ctree", "complexity")
+
+## function for visualization
+ciplot <- function(x, xlim = NULL, main = "", xlab = "", ylab = TRUE) {
+  nam <- rownames(x)
+  k <- length(nam)
+  plot(x$mean, 1:k, xlim = xlim, axes = FALSE, xlab = "", ylab = "", pch = 19)
+  arrows(x$lower, 1:k, x$upper, 1:k, angle = 90, length = 0.05, code = 3)
+  if(xlab == "") axis(1, labels = FALSE) else axis(1)
+  if(ylab) ylab <- nam
+  axis(2, at = 1:k, labels = ylab, las = 1, cex = 0.8)  
+  axis(2, at = k + 1.5, labels = main, tick = FALSE, las = 1, outer = TRUE, cex.axis = 1.5, xpd = TRUE)
+  mtext(xlab, side = 1, line = 3, xpd = TRUE)
+  abline(h = 5.5)
+  abline(v = 0, lty = 2)  
+  box()
 }
@@
 \centering
-<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
-plotd("rpart")
+<<benchmark-plot, fig = TRUE, echo = FALSE, width=10, height=10>>=
+par(mfrow = c(2, 2), oma = c(5, 10, 2, 0), mar = c(1, 1, 2, 1))
+
+xlim1 <- range(cbind(acc_rpart, acc_ctree))
+xlim2 <- range(cbind(com_rpart, com_ctree))
+
+ciplot(acc_rpart, xlim = xlim1, main = "rpart", ylab = TRUE, xlab = "")
+ciplot(com_rpart, xlim = xlim2, main = "",      ylab = FALSE, xlab = "")
+ciplot(acc_ctree, xlim = xlim1, main = "ctree", ylab = TRUE,
+  xlab = "relative difference in predictive accuracy (%)")
+ciplot(com_ctree, xlim = xlim2, main = "",      ylab = FALSE,
+  xlab = "relative difference in complexity (%)")
 @
-\caption{\label{fig:crpart} Performance comparison of \code{evtree} and \code{rpart}.
-Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error.
-The complexity is compared by the relative difference of the number of terminal nodes.}
+\caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{ctree} (lower panels).
+  Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean-squared error.
+  The complexity (right panels) is compared by the relative difference of the number of terminal nodes.}
 \end{figure}
 
-\begin{figure}[tbh!]
-\centering
-<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
-plotd("ctree")
-@
-\caption{\label{fig:cptree} Performance comparison of \code{evtree} and \code{ctree}.
-Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error.
-The complexity is compared by the relative difference of the number of terminal nodes.}
-\end{figure}
-
-The disadvantages of our algorithm are computation time and memory requirements.
-While the smallest of the analyzed data-sets, \textit{Glass Identification}, only needed approximately 4--6 seconds to fit, larger data-sets demanded several minutes.
-The fit of a model from the largest data-set, \textit{MAGIC Gamma Telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit.
+The disadvantages of our algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification},
+only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes.
+The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit.
 The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.
@@
 
-<<echo = FALSE>>=
-chessboard44 <- function(totalpoints = 4000, noisevariables = 6, noise = 0){
-    chess44 <- array(0,c(totalpoints,noisevariables+3))
+<<chessboard, echo=FALSE>>=
+chessboard44 <- function(n = 4000, noisevariables = 6, noise = 0) {
+  chess44 <- array(0,c(n,noisevariables+3))
   for(i in 1:(noisevariables+2))
@@
    y <- chess44[,2]
-     chess44[,ncol(chess44)] <- 0
+   chess44[, ncol(chess44)] <- 0
    for(k in 1:4)  
@@
 
-     if( noise > 0){
-        flipclasslist <- sample(totalpoints, totalpoints * (noise / 100), replace = F)
+   if(noise > 0) {
+      flipclasslist <- sample(n, n * (noise / 100), replace = FALSE)
 
@@
   chess44[,ncol(chess44)] <- as.factor(chess44[,ncol(chess44)])
-    names(chess44)[9] <- "class"
+  names(chess44) <- c(paste("X", 1:8, sep = ""), "Y")
   chess44
@@
 
-\setkeys{Gin}{width=0.4\textwidth}
-\begin{figure}[tbh!]
+\setkeys{Gin}{width=0.5\textwidth}
+\begin{figure}[t!]
 \centering
-<<fig = TRUE, echo = FALSE>>=
-par(cex.axis=1.5, cex.lab=1.5)
-ch <- chessboard44(totalpoints= 2000)
-par(cex.axis=1.5, cex.lab=1.5)
-plot(ch[ch[,9]==1,1], ch[ch[,9]==1,2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", cex=1)
-points(ch[ch[,9]==0,1], ch[ch[,9]==0,2], pch= c("o"))
+<<chessboard-plot, height=4.5, width=4.5, fig = TRUE, echo = FALSE>>=
+chess44 <- chessboard44(2000)
+plot(X2 ~ X1, data = chess44, xlim = c(0, 4), ylim = c(0, 4), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
 @
-\caption{\label{fig:chess} Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
+\caption{\label{fig:chess} Class distribution of the simulated $4 \times 4$ chessboard problem with zero noise,
+  plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses, respectively.}
 \end{figure}
@@
 
-<<label =  tab:resultsChessboard,  echo = FALSE, results=tex>>=
+<<chessboard-table, echo=FALSE, results=tex>>=
 library("xtable")
-load(file = "./results/chessboard44_0.RData")
-load(file = "./results/chessboard44_5.RData")
-load(file = "./results/chessboard44_10.RData")
+load("./results/chessboard44_0.RData")
+load("./results/chessboard44_5.RData")
+load("./results/chessboard44_10.RData")
 
 chesstable_means  <- as.data.frame( rbind(apply(rchessboard44_0,2,mean), apply(rchessboard44_5,2,mean) , apply(rchessboard44_10,2,mean) )) 
-names(chesstable_means) <-  c("evtree", "rpart", "ctree", "evtree", "rpart", "ctree")
+names(chesstable_means) <-  c("\\code{evtree}", "\\code{rpart}", "\\code{ctree}", "\\code{evtree}", "\\code{rpart}", "\\code{ctree}")
 chesstable_means[,1:3] <-  format(chesstable_means[,1:3]*100, digits=1, nsmall=1)
@@
 chesstable_sd  <- as.data.frame( rbind(apply(rchessboard44_0,2,sd), apply(rchessboard44_5,2,sd) , apply(rchessboard44_10,2,sd) )) 
-names(chesstable_sd) <-  c("evtree", "rpart", "ctree", "evtree", "rpart", "ctree")
+names(chesstable_sd) <-  c("\\code{evtree}", "\\code{rpart}", "\\code{ctree}", "\\code{evtree}", "\\code{rpart}", "\\code{ctree}")
 chesstable_sd[,1:3] <-  format(chesstable_sd[,1:3]*100, digits=1, nsmall=1)
@@
 print(xtable(chesstable,
-caption = "Comparison of accuracy on simulated $4 \\times 4$ chess-board examples. Accuracy and the number of terminal nodes are displayed as $mean(std)$.",
+caption = "Mean (and standard deviation) of accuracy and number of terminal nodes for simulated $4 \\times 4$ chessboard examples.",
 caption.placement= "bottom",
@@
 include.rownames = FALSE, allign= "rllllll", hline.after=NULL,
+sanitize.text.function = identity,
 add.to.row=list(pos=list(-1,-1, 0, 3), command=c(
 "\\toprule", 
-c("\\multicolumn{1}{l}{Noise [\\%]} & \\multicolumn{3}{l}{Accuracy}  & \\multicolumn{3}{l}{Terminal Nodes}\\\\",
+c("\\multicolumn{1}{l}{Noise (\\%)} & \\multicolumn{3}{l}{Accuracy}  & \\multicolumn{3}{l}{Terminal nodes}\\\\",
 "\\midrule",
@@
 
-In Section~\ref{comp} we compared \code{evtree} with \code{rpart} and \code{ctree} on 14 benchmark and 3 real-world data-sets.
-The data-sets varied by sample size, the number of classes, and by the number and types of attributes.
-\code{evtree} models offered at least comparable and, for the majority of the data-sets, improved predictive accuracy.
+In Section~\ref{comp} we compared \code{evtree} with \code{rpart} and \code{ctree} on 14 benchmark and 3 real-world datasets.
+The datasets varied by sample size, the number of classes, and by the number and types of attributes.
+\code{evtree} models offered at least comparable and, for the majority of the datasets, improved predictive accuracy.
 In an artificial problem of a $4 \times 4$ chessboard we demonstrated \code{evtree}'s performance in the presence of a higher order of interactions and noise.

### 2011-09-18 pkg/evtree/inst/doc/evtree.Rnw
started new 'implementation and application in practice section'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@241 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Based on the input data $X$ and the paramters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$.
+Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$.
 $loss$ is a suitable loss function for the domain of $Y$.
@@
 
-\section{Implementation}
+\section{Implementation and application in practice}
 \label{impl}
@@
 In this case, iterative calls of \code{evtree} will possibly result in different models.
+
+An interesting application for classification tree analysis is target marketing, where limited resources are aimed at a distinct group of potential customers.
+An example of a marketing case study is provided by the \textit{Bookbinder's Book Club} \citep{lilien2004} dataset.
+The data is a marketing case study about a (fictitious) American book club.
+In this case study, a brochure of the book \textit{The Art History of Florence} was sent to 20,000 customers and 1,806 of which bought the book.
+The dataset contains a subsample of 1,300 customers for building a predictive model for the customers choice of buying the advertised book.
+Besides predictive accuracy, model complexity is a crucial issue in this application.
+More simple trees are easier to interpret and communicable to marketing experts and management professionals.
+
+In the following analysis of the \textit{Bookbinder's Book Club} data \code{evtree} is used in conjunction with \code{rpart} and \code{ctree}.
+All trees are constrained to have a minimum number of 26 observations per terminal node ($2\%$ of the total number of observations).
+To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$.
+To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}.
+The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
 %
-<<iris-tree>>=
-library("evtree")
-data("iris", package = "datasets")
-ev <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
+<<recursive-trees>>=
+data("BBBClub", package = "evtree")
+library("rpart")
+set.seed(1090)
+rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26))
+ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26)
+plot(rp)
+plot(ct)
 @
 %
-The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
+The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
 %
-<<iris-print>>=
-ev
-plot(ev)
-@
-
 \setkeys{Gin}{width=1\textwidth}
@@
 \centering
-<<iris-plot, fig=TRUE, echo=FALSE, width=13.2, height=8>>=
+\begin{minipage}{1\textwidth}
+\begin{flushleft} \code{rpart} \end{flushleft}
+\centering
+<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
+plot(rp)
+@
+\end{minipage}
+\begin{minipage}{1\textwidth}
+\begin{flushleft} \code{ctree} \end{flushleft}
+\centering
+<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
+plot(ct)
+@
+\end{minipage}
+\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel).
+Target variable is the customers choice of buying the book.
+The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the BBB Club; \code{gender} the costumers gender.
+}
+\end{figure}
+%
+With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$.
+The resulting tree is displayed in Figure~\ref{fig:example_evtree}.
+Alternatively the tree can be examined by the textual output provided by the \code{print()} function.
+%
+<<evtree-plot>>=
+ev <- evtree(choice ~ ., data = BBBClub, minbucket = 26, maxdepth = 2)
+plot(ev)
+print(ev)
+@
+%
+\setkeys{Gin}{width=1\textwidth}
+\begin{figure}[t!]
+\begin{flushleft} \code{evtree} \end{flushleft}
+\centering
+<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
 plot(ev)
 @
-\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
+\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}.
+Target variable is the customers choice of buying the book.
+The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase}
 \end{figure}
+%
+While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$.
+With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler.
+%
+<<evtree-performance>>=
+## accuracy: misclassification rate
+mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
+c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
 
-The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
+## complexity: number of terminal nodes
+c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
+@
+The accuracy of the three models are similar.
+With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.9\%$).
 %
-<<iris-predict>>=
-predict(ev, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
+<<evtree-structure>>=
+## compare structure of predictions
+ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
+  ctree  = predict(ct), observed = BBBClub$choice))
+
+## compare customer predictions only (absolute, proportion correct)
+sapply(c("evtree", "rpart", "ctree"), function(nam) {
+  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
+  c(abs = as.vector(rowSums(mt))[2],
+    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
+})
 @
+%
+Although the constructed trees are considerably different, the structure of the predictions are relatively similar.
+In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($278$).
+Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $63.7\%$ (\code{ctree}) actually bought the book.
+%
+<<evtree-performance>>=
+rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26, 
+maxdepth = 2))
+ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26, 
+maxdepth = 2)
+## accuracy: misclassification rate
+mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
+c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
+## complexity: number of terminal nodes
+c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
+@
+%
+\code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previous \code{evtree} model.
+With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$).
+This shows an particular advantage of \code{evtree}, which is its ability to built the best tree for a given tree size.
 

### 2011-09-19 pkg/evtree/inst/doc/evtree.Rnw
BBBClub example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@242 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-In the following analysis of the \textit{Bookbinder's Book Club} data \code{evtree} is used in conjunction with \code{rpart} and \code{ctree}.
-All trees are constrained to have a minimum number of 26 observations per terminal node ($2\%$ of the total number of observations).
+This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyse the \textit{Bookbinder's Book Club} data.
+In this session, all trees are constrained to have a minimum number of $26$ observations per terminal node ($2\%$ of the total number of observations).
 To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$.
@@
 %
-<<recursive-trees>>=
+<<rpart-ctree>>=
 data("BBBClub", package = "evtree")
@@
 \caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel).
-Target variable is the customers choice of buying the book.
-The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the BBB Club; \code{gender} the costumers gender.
+The target variable is the customers choice of buying the book.
+The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
 }
@@
 The resulting tree is displayed in Figure~\ref{fig:example_evtree}.
-Alternatively the tree can be examined by the textual output provided by the \code{print()} function.
+Alternatively, the tree can be examined by the textual output provided by the \code{print()} function.
 %
@@
 \caption{\label{fig:example_evtree} Tree constructed by \code{evtree}.
-Target variable is the customers choice of buying the book.
-The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase}
+The target variable is the customers choice of buying the book.
+The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
 \end{figure}
@@
 Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $63.7\%$ (\code{ctree}) actually bought the book.
+In a final step \code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previously fitted \code{evtree} model.
 %
@@
 %
-\code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previous \code{evtree} model.
 With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$).
-This shows an particular advantage of \code{evtree}, which is its ability to built the best tree for a given tree size.
+Due to its unbalanced structure, the \code{rpart} model has only $3$ terminal nodes.
+Hence, it does not use the maximum tree depth efficiently.
+In contrast, a global optimization algorithm like \code{evtree} can be used to search the best tree under given constraints.
 

### 2011-09-20 pkg/evtree/inst/doc/evtree.Rnw
BBBClub example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@243 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyse the \textit{Bookbinder's Book Club} data.
-In this session, all trees are constrained to have a minimum number of $26$ observations per terminal node ($2\%$ of the total number of observations).
+This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyze the \textit{Bookbinder's Book Club} data.
+In this session, all trees are constrained to have a minimum number of $10$ observations per terminal node.
 To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$.
@@
 set.seed(1090)
-rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26))
-ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26)
+rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
+ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10)
 plot(rp)
@@
 <<evtree-plot>>=
-ev <- evtree(choice ~ ., data = BBBClub, minbucket = 26, maxdepth = 2)
+ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
 plot(ev)
@@
 <<evtree-performance>>=
-## accuracy: misclassification rate
 mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
 c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
-
-## complexity: number of terminal nodes
-c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
 @
 The accuracy of the three models are similar.
-With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.9\%$).
+With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.8\%$).
 %
 <<evtree-structure>>=
-## compare structure of predictions
 ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
@@
 
-## compare customer predictions only (absolute, proportion correct)
 sapply(c("evtree", "rpart", "ctree"), function(nam) {
@@
 Although the constructed trees are considerably different, the structure of the predictions are relatively similar.
-In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($278$).
-Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $63.7\%$ (\code{ctree}) actually bought the book.
+In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$).
+Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $66.4\%$ (\code{ctree}) actually bought the book.
 In a final step \code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previously fitted \code{evtree} model.
@@
 <<evtree-performance>>=
-rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26, 
+rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10, 
 maxdepth = 2))
-ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26, 
+ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10, 
 maxdepth = 2)
-## accuracy: misclassification rate
+
 mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
 c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
-## complexity: number of terminal nodes
+
 c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))

### 2011-09-20 pkg/evtree/inst/doc/evtree.Rnw
BBBClub.Rd
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@244 255f4249-2263-4417-b1ea-8498779e27a7

@@
 rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
-ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10)
+ct <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
 plot(rp)
@@
 maxdepth = 2))
-ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10, 
+ct <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99, 
 maxdepth = 2)

### 2011-09-20 pkg/evtree/inst/doc/evtree.Rnw
improved intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@245 255f4249-2263-4417-b1ea-8498779e27a7

@@
   which implements an evolutionary algorithm for learning globally optimal
-  classification and regression trees in \proglang{R}. CPU and memory-intensive
+  classification and regression trees in \proglang{R}. Computationally intensive
   tasks are fully computed in \proglang{C++} while the \pkg{partykit}
-  \citep{hothorn2009partykit} package is leveraged to represent the resulting
+  \citep{hothorn2009partykit} package is leveraged for representing the resulting
   trees in \proglang{R}, providing unified infrastructure for summaries,
@@
   \citep{therneau1997introduction}, the open-source CART implementation, and
-  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
-  evaluated datasets, including several benchmark problems from the UCI machine
-  learning repository, \code{evtree} models offered at least similar and most of
-  the time increased predictive accuracy.  
+  conditional inference trees \citep[\code{ctree},][]{hothorn2006urp}. The usefulness
+  of \code{evtree} is illustrated in a textbook customer classification task
+  and a benchmark study of predictive accuracy in which \code{evtree}
+  achieved at least similar and most of the time better results compared to
+  the recursive algorithms \code{rpart} and \code{ctree}.  
 }
@@
 options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+library("rpart")
 library("evtree")
+data("BBBClub", package = "evtree")
+cache <- TRUE
 @
@@
 
-Classification and regression trees are commonly applied for exploration and modeling of complex data.
-They are able to handle strongly nonlinear relationships with high order interactions and different variable types.
-The resulting model can be interpreted as a tree structure, which gives a compact and intuitive representation.
-Commonly used classification and regression tree algorithms, including \textit{CART} \citep{breiman1984cart} and \textit{C4.5} \citep{quinlan1993}, use a greedy heuristic, where split rules are selected in a forward stepwise search.
-The data is recursively split into two groups.
-The split rule at each internal node is selected to maximize the homogeneity of its two child nodes, without consideration of nodes further down in the tree.
-Split rules selected by this top-down heuristic are only locally optimal, as splits are chosen to maximize homogeneity at the next step only.
-Nonetheless, the greedy heuristic is computationally efficient and often yields reasonably good results \citep{murthy1995dti}.
-However, for some problems, greedily induced trees can be far from the optimal solution, and a global search over the tree's parameter space can lead to much more compact and accurate models.
-
-One possibility to build globally optimal trees is to use stochastic optimization methods like evolutionary algorithms.
-Evolutionary algorithms are methods that are inspired by natural Darwinian evolution.
-Important concepts include inheritance, mutation, and natural selection.
-They are population based, i.e., a whole collection of candidate solutions are processed simultaneously.
-The initial population is either initialized randomly, or by a problem dependent algorithm.
-The individuals in the populations are then iteratively modified by \textit{variation operators}.
-For tree algorithms variation operators can be applied to modify the tree structure and the tests that are applied in the internal nodes.
-Variation operators that apply to single individuals are called \textit{mutation}, while operators that merge information from different solutions are called \textit{crossover}.
-The survivor selection process favors individuals that perform well according to some quality criteria, which is called \textit{fitness function} or \textit{evaluation function}.
-In this evolutionary process the mean quality of the population increases over time \citep{baeck,eiben2003iec}.
-
-In practice, evolutionary algorithm and other stochastic optimization methods are rarely used in decision tree induction.
-One reason is that they are computationally much more demanding than a forward stepwise search.
-Perhaps another important reason is their lack of availability in major software packages.
-\proglang{R} \citep{team2011r} provides several packages with tree-based methods.
-\pkg{rpart} \citep{therneau1997introduction} is an \proglang{R} implementations of CART.
-Package \pkg{party} provides two tree-based algorithms with unbiased variable selection and statistical stopping criteria.
-Several tree algorithms are available in \proglang{Weka} \citep{weka}.
-Package \pkg{RWeka} \citep{rweka} provides an interface to \proglang{Weka} for \proglang{R} users.
-Decision tree methods in \proglang{Weka} include implementations of C4.5 and \textit{M5} \citep{quinlan1992lcc}.
-However, all of these methods are based on a top-down heuristic.
-An exception is the \pkg{LogicReg} package, which implements logic regression.
-This method is applicable for regression problems with binary covariates.
-The resulting models of logic regression can be visualized as logic trees.
-
-In this article, we present \pkg{evtree}, a software package that uses evolutionary methods for learning globally optimal classification and regression trees in \proglang{R}.
-The package is available from the Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/package=evtree}.
-An important difference to comparable algorithms is the survivor selection mechanism, where we use a simple $(1+1)$ selection strategy.
-That is, one parent solution competes with one offspring for a place in the subsequent population.
-As further discussed in Section~\ref{evo}, this design offers computational advantages for the application of classification and regression trees.
-Furthermore, we argue that the chosen selection mechanism helps to avoid premature convergence.
-
-The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it
-to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will
-be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree}
-is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and
-conditional inference trees in \proglang{R}.
-The comparison comprises 14~benchmark datasets, 3~real-world datasets and 3~simulated datasets.
-It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods.
-Finally, in Section~\ref{conc}, we will summarize and draw conclusions on the implementation and the performance of our algorithm.
+Classification and regression trees are commonly applied for exploration and
+modeling of complex data. They are able to handle strongly nonlinear
+relationships with high order interactions and different variable types. The
+resulting model can be interpreted as a tree structure providing a compact
+and intuitive representation. Commonly used classification and regression tree
+algorithms, including \textit{CART} \citep{breiman1984cart} and \textit{C4.5}
+\citep{quinlan1993}, use a greedy heuristic, where split rules are selected in a
+forward stepwise search for recursively partitioning the data into groups. The
+split rule at each internal node is selected to maximize the homogeneity of its
+child nodes, without consideration of nodes further down the tree, hence yielding
+only locally optimal trees. Nonetheless, the
+greedy heuristic is computationally efficient and often yields reasonably good
+results \citep{murthy1995dti}. However, for some problems, greedily induced
+trees can be far from the optimal solution, and a global search over the tree's
+parameter space can lead to much more compact and accurate models.
+
+The main challenge in growing globally optimal trees is that the search
+space is typically huge rendering full-grid searches computationally infeasible.
+One possibility to solve this problem is to use stochastic
+optimization methods like evolutionary algorithms. 
+In practice, however, such stochastic methods
+are rarely used in decision tree induction. One reason is probably that they are
+computationally much more demanding than a recursive forward search but another one is
+likely to be the lack of availability in major software packages.
+In particular, while there are several packages for \proglang{R} \citep{team2011r} providing
+forward-search tree algorithms, there is only little support for globally
+optimal trees. The former group of packages includes (among others)
+\pkg{rpart} \citep{therneau1997introduction}, the open-source
+implementation of the CART algorithm; \pkg{party}, containing two tree algorithms
+with unbiased variable selection and statistical stopping criteria
+\citep{hothorn2006urp,zeileis2008mob}; and \pkg{RWeka} \citep{rweka}, the
+\proglang{R} interface to \pkg{Weka} \citep{weka} with open-source implementations
+of tree algorithms such as C4.5 or \textit{M5} \citep{quinlan1992lcc}.
+A notable exception is the
+\pkg{LogicReg} package \citep{logicreg} for logic regression, an algorithm for globally optimal
+trees based on binary covariates only  and using simulated annealing. See
+\cite{hothorn2011ctv} for an overview of further recursive partitioning packages
+for \proglang{R}.
+
+To fill this gap, we introduce a new \proglang{R} package \pkg{evtree}, 
+available from the Comprehensive \proglang{R} Archive Network at
+\url{http://CRAN.R-project.org/package=evtree}, providing evolutionary
+methods for learning globally optimal classification and regression trees.
+Generally speaking, evolutionary algorithms are inspired by natural Darwinian
+evolution employing concepts such as inheritance, mutation, and natural selection.
+They are population-based, i.e., a whole collection of candidate solutions -- trees in this application -- is
+processed simultaneously and iteratively modified by \textit{variation operators}
+called \textit{mutation} (applied to single solutions) and \textit{crossover}
+(merging different solutions). Finally, a survivor selection process favors
+solutions that perform well according to some quality criterion, usually
+called \textit{fitness function} or \textit{evaluation function}. In this
+evolutionary process the mean quality of the population increases over time
+\citep{baeck,eiben2003iec}.  
+In the case of learning decision trees, this means that the variation
+operators can be applied to modify the tree structure (e.g., number of splits, splitting
+variables, and corresponding splitpoints etc.) in order to optimize a
+fitness functions such as the misclassificaion or error rate penalized by
+the complexity of the tree.
+A notable difference to comparable algorithms is the survivor selection
+mechanism where it is important to avoid premature convergence. In the
+following, we use a simple $(1+1)$ selection strategy (i.e., one parent
+solution competes with one offspring for a place in the population)
+which can be argued to offer computational
+advantages for the application to classification and regression trees.
+
+The remainder of this paper is structured as follows: Section~\ref{got}
+describes the problem of learning globally optimal decision trees and contrasts
+it to the locally optimal top-down heuristic that is utilized by recursive
+partitioning algorithms. Section~\ref{desc} introduces the \code{evtree} algorithm
+before Section~\ref{impl} addresses implementation details along with an
+overview of the implemented functions. A benchmark comparison -- comprising
+14~benchmark datasets, 3 real-world datasets, and 3 simulated scenarios -- is carried out
+in Section~\ref{comp}, showing that the predictive performance of \code{evtree}
+is often significantly better compared to commonly used algorithms \code{rpart}
+and \code{ctree} (from \pkg{party} package). Finally, Section~\ref{conc} gives
+concluding remarks about the implementation and the performance of the new
+algorithm.
 
@@
 
-The aim of classification and regression tree analysis can be stated by explaining a response variable $Y$ by a vector of $P$ predictor variables $X=(X_1,...,X_P)$.
-For classification trees the output variable $Y$ is qualitative and for regression trees $Y$ is quantitative.
-Tree-based methods first partition the input space $X$ into a set of $M$ rectangular regions $R_m$ where $m = 1,...,M$.
-In the second step a model is fit to each region.
-The model fitted to the distribution of $\{Y \in R_m\}$ can be any proper measure of the location, e.g.
-the median, or dispersion, e.g.
-the variance, of a random variable.
-Typically the mode is used for classification trees and the arithmetic mean is applied for regression trees.
-
-In this work only binary tree models are considered, where regions are split into two halves at a time.
-However, this is not a big issue in global tree induction.
-As binary splits are searched globally and any multiway split can be represented by a series of binary splits.
-Based on a set of split rules, binary classification and regression trees recursively partition the input space $X$ into two regions.
-Samples that fulfill the criterion of a split rule propagate down into the descendant left node and the other instances into the right node.
-Split points of a tree are termed \textit{internal nodes} and nodes without successors are called \textit{terminal nodes}.
-A binary tree with $M$ terminal nodes has $M-1$ internal nodes.
-Generally, the maximum number of terminal nodes $M_{max}=\infty$.
-In practice, $M_{max}$ is restricted by the maximum tree depth $D_{max}$ as $M_{max} = 2^{D_{max}}-1$.
-When the trees are constrained to have a minimum number of observations in the internal and/or terminal nodes, $M_{max}$ is also restricted by the finiteness of the training data.
-
-A binary tree model can be uniquely defined by the positions of the internal nodes, being a subset of $\{1,...,M_{max}-1\}$, and the corresponding split rules.
-Let the positions of the internal nodes of a tree with $M-1$ terminal nodes be indicated by $n_1,...,n_{M-1}$.
-The positions $\{n_1,...,n_{M-1}\} \in \{1,...,M_{max}-1\}$ and $n_i < n_j \Leftrightarrow i < j$.
-Let the variable selected for splitting, at a new position $r \in n_1,...,n_{M_{max}-1}$, be denoted as $v_r$, where $v_r=1,...,P$.
-Depending on the domain of $X_{v_r}$, a split point or a set of categories has to be selected, which we denote as $s_r$.
-If $X_{v_r}$ is numerical or ordinal, a binary split rule is defined as $X_{v_r} < s_r$.
-If $X_{v_r}$ has $u$ unique values, then there are $u-1$ possible split points and $s_r=1,...,u-1$.
-In the case of a nominal split variable $X_{v_r}$, a split rule is defined as $X_{v_r} \in s_r$, where $s_r \subset \{1,...,K\}$ and $s_r \neq \{\emptyset\}$.
-If $X_{v_r}$ has $K$ categories there are $2^{K-1}-1$ possible splits.
+The aim of classification and regression tree analysis can be stated by
+explaining a response variable $Y$ by a vector of $P$ predictor variables
+$X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
+qualitative and for regression trees $Y$ is quantitative. Tree-based methods
+first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
+where $m = 1,...,M$. In the second step a model is fit to each region. The model
+fitted to the distribution of $\{Y \in R_m\}$ can be any proper measure of the
+location, e.g. the median, or dispersion, e.g. the variance, of a random
+variable. Typically the mode is used for classification trees and the arithmetic
+mean is applied for regression trees.
+
+In this work only binary tree models are considered, where regions are split
+into two halves at a time. However, this is not a big issue in global tree
+induction. As binary splits are searched globally and any multiway split can be
+represented by a series of binary splits. Based on a set of split rules, binary
+classification and regression trees recursively partition the input space $X$
+into two regions. Samples that fulfill the criterion of a split rule propagate
+down into the descendant left node and the other instances into the right node.
+Split points of a tree are termed \textit{internal nodes} and nodes without
+successors are called \textit{terminal nodes}. A binary tree with $M$ terminal
+nodes has $M-1$ internal nodes. Generally, the maximum number of terminal nodes
+$M_{max}=\infty$. In practice, $M_{max}$ is restricted by the maximum tree depth
+$D_{max}$ as $M_{max} = 2^{D_{max}}-1$. When the trees are constrained to have a
+minimum number of observations in the internal and/or terminal nodes, $M_{max}$
+is also restricted by the finiteness of the training data.
+
+A binary tree model can be uniquely defined by the positions of the internal
+nodes, being a subset of $\{1,...,M_{max}-1\}$, and the corresponding split
+rules. Let the positions of the internal nodes of a tree with $M-1$ terminal
+nodes be indicated by $n_1,...,n_{M-1}$. The positions $\{n_1,...,n_{M-1}\} \in
+\{1,...,M_{max}-1\}$ and $n_i < n_j \Leftrightarrow i < j$. Let the variable
+selected for splitting, at a new position $r \in n_1,...,n_{M_{max}-1}$, be
+denoted as $v_r$, where $v_r=1,...,P$. Depending on the domain of $X_{v_r}$, a
+split point or a set of categories has to be selected, which we denote as $s_r$.
+If $X_{v_r}$ is numerical or ordinal, a binary split rule is defined as $X_{v_r}
+< s_r$. If $X_{v_r}$ has $u$ unique values, then there are $u-1$ possible split
+points and $s_r=1,...,u-1$. In the case of a nominal split variable $X_{v_r}$, a
+split rule is defined as $X_{v_r} \in s_r$, where $s_r \subset \{1,...,K\}$ and
+$s_r \neq \{\emptyset\}$. If $X_{v_r}$ has $K$ categories there are $2^{K-1}-1$
+possible splits.
 
@@
 
-As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a function of the number of terminal nodes, without further considering the depth or the shape of trees.
-Then the overall parameter space $\Theta$ can be written as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the parameter space of models with $M$ terminal nodes.
-Let $\theta$ be a generic element of $\Theta$, then $\theta$ is fully defined by:
-
+As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
+function of the number of terminal nodes, without further considering the depth
+or the shape of trees. Then the overall parameter space $\Theta$ can be written
+as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
+parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
+element of $\Theta$, then $\theta$ is fully defined by:
+%
 \begin{equation}
@@
 \end{equation}
-
-The overall goal of classification and regression tree analysis can be defined as finding a set of parameters $\hat{\theta}$, such that
-
+%
+The overall goal of classification and regression tree analysis can be defined
+as finding a set of parameters $\hat{\theta}$, such that
+%
 \begin{equation}
@@
 \end{equation}
-
-Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$.
-$loss$ is a suitable loss function for the domain of $Y$.
-Commonly, the misclassification error for classification problems, or the squared error for regression problems, is used.
-$comp$ quantifies the complexity of trees and penalizes for more complex models in the tree selection process.
-In CART, the complexity is a non-negative function of the terminal-nodes.
-Note that finding $\hat{\theta}$ requires a search over all $\Theta_m$.
-
-The parameter space $\Theta$ becomes large for already medium sized problems and a complete search for larger problems is computationally intractable.
-In fact, \cite{hyafil1976cob} showed that building optimal binary decision trees, such that the expected number of tests required to classify an unknown sample is minimized, is NP-complete.
-\cite{zantema2000finding} proved that finding a decision tree of minimal size that is decision equivalent to a given decision tree is an NP-hard problem.
-As a consequence the search space is usually limited by heuristics.
+%
+Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
+predictions for the output $Y$. $loss$ is a suitable loss function for the
+domain of $Y$. Commonly, the misclassification error for classification
+problems, or the squared error for regression problems, is used. $comp$
+quantifies the complexity of trees and penalizes for more complex models in the
+tree selection process. In CART, the complexity is a non-negative function of
+the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
+$\Theta_m$.
+
+The parameter space $\Theta$ becomes large for already medium sized problems and
+a complete search for larger problems is computationally intractable. In fact,
+\cite{hyafil1976cob} showed that building optimal binary decision trees, such
+that the expected number of tests required to classify an unknown sample is
+minimized, is NP-complete. \cite{zantema2000finding} proved that finding a
+decision tree of minimal size that is decision equivalent to a given decision
+tree is an NP-hard problem. As a consequence the search space is usually limited
+by heuristics.
 
@@
 
-Instead of searching all combination of $\Theta$ simultaneously, recursive partitioning algorithms only consider one split at a time.
-At each internal node $r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding split point $s_r$ is selected such that it minimizes the impurity in the two resulting terminal nodes.
-Let the parameters of a tree, constructed in $i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} = (\emptyset)$.
-Then, for a tree with parameters $\tilde{\theta}_{i}$, the position of the new internal node $r$, as well as the corresponding split variable $v_r$ and split point $s_r$, are chosen such that:
-
+Instead of searching all combination of $\Theta$ simultaneously, recursive
+partitioning algorithms only consider one split at a time. At each internal node
+$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
+split point $s_r$ is selected such that it minimizes the impurity in the two
+resulting terminal nodes. Let the parameters of a tree, constructed in
+$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
+(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
+position of the new internal node $r$, as well as the corresponding split
+variable $v_r$ and split point $s_r$, are chosen such that:
+%
 \begin{equation}
@@
 \end{equation}
-
-CART's standard measures for node impurity are the Gini index, for classification problems and the mean squared error, for regression problems.
-After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by $\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$.
-This process of recursively selecting and executing splits continues until the homogeneity of the nodes cannot be further improved.
-Additionally, often additional stopping criteria are defined, e.g.
-minimum sample sizes in the internal nodes and terminal nodes.
-
-In CART, and many other recursive tree methods, the tree growing step is followed by a pruning step.
-The idea behind pruning is to first grow a large tree without or with little restriction on the magnitude of impurity reduction.
-Starting from the bottom of the tree pruning then removes worthless splits.
-The method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and used by CART is called \textit{cost-complexity pruning}.
-Cost-complexity pruning iteratively prunes the internal node that:
-
+%
+CART's standard measures for node impurity are the Gini index, for
+classification problems and the mean squared error, for regression problems.
+After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
+$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
+process of recursively selecting and executing splits continues until the
+homogeneity of the nodes cannot be further improved. Additionally, often
+additional stopping criteria are defined, e.g. minimum sample sizes in the
+internal nodes and terminal nodes.
+
+In CART, and many other recursive tree methods, the tree growing step is
+followed by a pruning step. The idea behind pruning is to first grow a large
+tree without or with little restriction on the magnitude of impurity reduction.
+Starting from the bottom of the tree pruning then removes worthless splits. The
+method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
+used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
+iteratively prunes the internal node that:
+%
 \begin{equation}
@@
 \end{equation}
-
-The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus (\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.
-
-For nontrivial problems, recursive partitioning methods only search a small fraction of the global search space $(v_1, s_1, ..., v_{M_{max}-1}, s_{M_{max}-1} )$.
-They only search each $(v_r, s_r)$ once, and independently of the other split rules.
-The selection of $v_r$ and $s_r$ at each step $i$ requires a search of the locally optimal split over $M=i$ terminal nodes.
-The pruning step requires $j$ selections of an internal node $r$.
-
-In contrast to CART and C4.5, conditional inference trees utilize an unbiased variable selection approach that is based on permutation testing.
-Bias in the split variable selection process arises from a preference towards variables with many potential split points.
-The conditional inference framework overcomes this problem by separating the variable and split point selection.
-Although this distinction influences the selection of split rules, conditional inference tree solve the same minimization problem (Equation \ref{eq:whichsplit}).
-The \code{ctree} implementation of conditional inference trees does not employ pruning, but uses a statistical stopping criterion.
-
-\subsection{An illustration showing the limitations of locally optimal decision trees}
-
-A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}.
-The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions.
-As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split.
-For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem.
-One solution, which is the tree constructed by \code{evtree} is given by:
+%
+The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
+(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.
+
+For nontrivial problems, recursive partitioning methods only search a small
+fraction of the global search space $(v_1, s_1, ..., v_{M_{max}-1},
+s_{M_{max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
+the other split rules. The selection of $v_r$ and $s_r$ at each step $i$
+requires a search of the locally optimal split over $M=i$ terminal nodes. The
+pruning step requires $j$ selections of an internal node $r$.
+
+In contrast to CART and C4.5, conditional inference trees utilize an unbiased
+variable selection approach that is based on permutation testing. Bias in the
+split variable selection process arises from a preference towards variables with
+many potential split points. The conditional inference framework overcomes this
+problem by separating the variable and split point selection. Although this
+distinction influences the selection of split rules, conditional inference tree
+solve the same minimization problem (Equation~\ref{eq:whichsplit}). The
+\code{ctree} implementation of conditional inference trees does not employ
+pruning, but uses a statistical stopping criterion.
+
+\subsection{An illustration of the limitations of locally optimal decision trees}
+
+A very simple example that illustrates the limitation of recursive partitioning
+methods is depicted in Figure~\ref{fig:illustration}. The example only contains
+two independent variables and can be solved with only three splits that
+partition the input space into four regions. As expected the recursive
+partitioning methods \code{rpart} and \code{ctree} fail to find any split at
+all, as the impurity of the resulting subsets cannot be reduced by the first
+split. For methods that explore $\Theta$ in a more global fashion it is
+straightforward to find an optimal solution to this problem. One solution, which
+is the tree constructed by \code{evtree} is given by: 
 %
@@
 %
-<<rpart-ctree>>=
+<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
 data("BBBClub", package = "evtree")
 library("rpart")
-set.seed(1090)
 rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
@@
 %
+With the objective of building a smaller, but at least equally accurate tree,
+\code{evtree} is constrained to a maximum tree depth of $2$.
+%
+<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
+set.seed(1090)
+ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
+@
+%
+<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
+if(cache & file.exists("BBBClub-trees.rda")) {
+load("BBBClub-trees.rda")
+} else {
+<<BBBClub-rpart-ctree>>
+<<BBBClub-evtree>>
+if(cache) {
+  save(rp, ct, ev, file = "BBBClub-trees.rda")
+} else {
+  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
+}
+}
+@
+%
 \setkeys{Gin}{width=1\textwidth}
-\begin{figure}[t!]
+\begin{figure}[p!]
 \centering
@@
 \centering
-<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
+\setkeys{Gin}{width=0.75\textwidth}
+<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
 plot(rp)
@@
 \centering
-<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
+\setkeys{Gin}{width=1\textwidth}
+<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
 plot(ct)
@@
 %
-With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$.
-The resulting tree is displayed in Figure~\ref{fig:example_evtree}.
-Alternatively, the tree can be examined by the textual output provided by the \code{print()} function.
-%
-<<evtree-plot>>=
-ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
+The resulting tree is displayed in the printed output below and visualized in
+Figure~\ref{fig:example_evtree}.
+%
+<<BBBClub-evtree-display>>=
 plot(ev)
-print(ev)
+ev
 @
@@
 \centering
-<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
+\setkeys{Gin}{width=0.7\textwidth}
+<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
 plot(ev)
@@
 %
-While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$.
-With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler.
+While the \code{rpart} model has \Sexpr{width(rp)}~terminal nodes and a tree depth
+of~\Sexpr{depth(rp)}, the tree constructed by \code{ctree} has
+\Sexpr{width(ct)}~terminal nodes and a tree depth of~\Sexpr{depth(ct)}.
+With a tree depth of~\Sexpr{depth(ev)} and \Sexpr{width(ev)}~terminal nodes
+the \code{evtree} model is much simpler. 
 %
@@
 @
-The accuracy of the three models are similar.
-With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.8\%$).
+The accuracy of the three models are similar. With a misclassification rate of
+$24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model
+($23.8\%$) and better than the \code{ctree} model ($24.8\%$). 
 %
@@
 @
-\caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{ctree} (lower panels).
+\caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{evtree} vs.\ \code{ctree} (lower panels).
   Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean-squared error.

### 2011-09-21 pkg/evtree/inst/doc/evtree.Rnw
started with Section 2, more work to do
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@246 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{amssymb, amsmath, amsthm, booktabs, thumbpdf}
+\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
 
@@
 parameter space can lead to much more compact and accurate models.
+
+\pagebreak
 
@@
 describes the problem of learning globally optimal decision trees and contrasts
-it to the locally optimal top-down heuristic that is utilized by recursive
+it to the locally optimal forward-search heuristic that is utilized by recursive
 partitioning algorithms. Section~\ref{desc} introduces the \code{evtree} algorithm
@@
 overview of the implemented functions. A benchmark comparison -- comprising
-14~benchmark datasets, 3 real-world datasets, and 3 simulated scenarios -- is carried out
+14~benchmark datasets, 3~real-world datasets, and 3~simulated scenarios -- is carried out
 in Section~\ref{comp}, showing that the predictive performance of \code{evtree}
-is often significantly better compared to commonly used algorithms \code{rpart}
-and \code{ctree} (from \pkg{party} package). Finally, Section~\ref{conc} gives
+is often significantly better compared to the commonly used algorithms \code{rpart}
+and \code{ctree} (from the \pkg{party} package). Finally, Section~\ref{conc} gives
 concluding remarks about the implementation and the performance of the new
@@
 
-The aim of classification and regression tree analysis can be stated by
-explaining a response variable $Y$ by a vector of $P$ predictor variables
+Classification and regression tree analysis aims at
+modeling a response variable $Y$ by a vector of $P$ predictor variables
 $X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
@@
 first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
-where $m = 1,...,M$. In the second step a model is fit to each region. The model
-fitted to the distribution of $\{Y \in R_m\}$ can be any proper measure of the
-location, e.g. the median, or dispersion, e.g. the variance, of a random
-variable. Typically the mode is used for classification trees and the arithmetic
-mean is applied for regression trees.
+($m = 1,...,M$) then fit a (typically simple) model within each region $\{Y | X \in R_m\}$,
+e.g., the mean, median, or variance etc. Typically, the mode is used for
+classification trees and the arithmetic mean is applied for regression trees.
 
@@
 \begin{equation}
-	\hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; loss(Y, f(X,\theta)) + comp(\theta).
+	\hat{\theta} = \argmin_{\theta \in \Theta} \; \mathit{loss}(Y, f(X,\theta)) + \mathit{comp}(\theta).
 	\label{generalcostfunction}
@@
 \begin{equation}
-		\min_{v_r, s_r} \; loss(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
+		\min_{v_r, s_r} \; \mathit{loss}(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
 	\label{eq:whichsplit}
@@
 \begin{equation}
-		\min_{r=1,...,n_{M-1}} \; loss(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + comp(\tilde{\theta}_i \setminus (v_r, s_r)).
+		\min_{r=1,...,n_{M-1}} \; \mathit{loss}(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + \mathit{comp}(\tilde{\theta}_i \setminus (v_r, s_r)).
 	\label{eq:pruning}
@@
 \begin{equation}
-	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
+	 \hat{\theta} = \argmin_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
 \label{eq:classification}
@@
 \begin{equation}
-	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
+	 \hat{\theta} = \argmin_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
 \label{eq:regression}

### 2011-09-26 pkg/evtree/inst/doc/evtree.Rnw
Improved almost all sections of the paper, a few FIXMEs remain.
Compilation requires deleting the cached .rda first.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@249 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \newcommand{\argmin}{\operatorname{argmin}\displaylimits}
+
+%% for internal use
+\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
+\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}
 
@@
 modeling a response variable $Y$ by a vector of $P$ predictor variables
-$X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
+$X=(X_1,...,X_P)$ where for classification trees $Y$ is
 qualitative and for regression trees $Y$ is quantitative. Tree-based methods
@@
 
-In this work only binary tree models are considered, where regions are split
-into two halves at a time. However, this is not a big issue in global tree
-induction. As binary splits are searched globally and any multiway split can be
-represented by a series of binary splits. Based on a set of split rules, binary
-classification and regression trees recursively partition the input space $X$
-into two regions. Samples that fulfill the criterion of a split rule propagate
-down into the descendant left node and the other instances into the right node.
-Split points of a tree are termed \textit{internal nodes} and nodes without
-successors are called \textit{terminal nodes}. A binary tree with $M$ terminal
-nodes has $M-1$ internal nodes. Generally, the maximum number of terminal nodes
-$M_{max}=\infty$. In practice, $M_{max}$ is restricted by the maximum tree depth
-$D_{max}$ as $M_{max} = 2^{D_{max}}-1$. When the trees are constrained to have a
-minimum number of observations in the internal and/or terminal nodes, $M_{max}$
-is also restricted by the finiteness of the training data.
-
-A binary tree model can be uniquely defined by the positions of the internal
-nodes, being a subset of $\{1,...,M_{max}-1\}$, and the corresponding split
-rules. Let the positions of the internal nodes of a tree with $M-1$ terminal
-nodes be indicated by $n_1,...,n_{M-1}$. The positions $\{n_1,...,n_{M-1}\} \in
-\{1,...,M_{max}-1\}$ and $n_i < n_j \Leftrightarrow i < j$. Let the variable
-selected for splitting, at a new position $r \in n_1,...,n_{M_{max}-1}$, be
-denoted as $v_r$, where $v_r=1,...,P$. Depending on the domain of $X_{v_r}$, a
-split point or a set of categories has to be selected, which we denote as $s_r$.
-If $X_{v_r}$ is numerical or ordinal, a binary split rule is defined as $X_{v_r}
-< s_r$. If $X_{v_r}$ has $u$ unique values, then there are $u-1$ possible split
-points and $s_r=1,...,u-1$. In the case of a nominal split variable $X_{v_r}$, a
-split rule is defined as $X_{v_r} \in s_r$, where $s_r \subset \{1,...,K\}$ and
-$s_r \neq \{\emptyset\}$. If $X_{v_r}$ has $K$ categories there are $2^{K-1}-1$
-possible splits.
+To show why forward-search recursive partitioning algorithms typically
+lead to globablly suboptimal solutions, their parameter spaces and optimization
+problems are presented and contrasted in a unified notation. Although all
+arguments hold more generally, only binary tree models with some maximum number
+of splits $M_{\max}$ are considered. Both restrictions make the notation somewhat
+simpler while not really restricting the problem: (a)~Multiway splits are equivalent
+to a sequence of binary splits in predictions and number of resulting subsamples.
+(b)~The maximal size of the tree is always limited by the number of observations
+in the learning sample.
+
+In the following, a binary tree model with $M$ terminal nodes (which consequently has
+$M-1$ internal splits) is denoted by
+%
+\begin{equation}
+  \theta ~=~ (v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ), \label{eq:tree}
+\end{equation}
+%
+where the $n_r \in \{1,...,M_{\max}-1\}$ are the positions of the
+internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
+\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
+Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
+(for ordered and numeric variables) or a a nonempty subset of $\{1, \dots, c\}$
+(for a categorical variable with $c$~levels), determining which observations are
+sent to the first or second subsample. In the former case, there are $u-1$ possible
+split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
+are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
+forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
+trees with $M$ terminal nodes. The overall parameter space is then 
+$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
+by exluding elements $\theta$ resulting in too small subsamples etc.).
+
+Finally, $f(X,\theta)$ denotes the prediction function based on all explanatory
+variables $X$ and the chosen tree structure $\theta$ from Equation~\ref{eq:tree}.
+As pointed out above, this is typically constructed using the means or modes in the respective
+partitions of the learning sample.
 
@@
 function of the number of terminal nodes, without further considering the depth
-or the shape of trees. Then the overall parameter space $\Theta$ can be written
-as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
-parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
-element of $\Theta$, then $\theta$ is fully defined by:
-%
-\begin{equation}
-	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
-	\label{treeRep}
+or the shape of trees. The goal is then to find that classification and regression tree
+which optimizes some tradeoff between predicition performance and complexity:
+%
+\begin{equation} \label{eq:globally}
+  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
 \end{equation}
 %
-The overall goal of classification and regression tree analysis can be defined
-as finding a set of parameters $\hat{\theta}$, such that
-%
-\begin{equation}
-	\hat{\theta} = \argmin_{\theta \in \Theta} \; \mathit{loss}(Y, f(X,\theta)) + \mathit{comp}(\theta).
-	\label{generalcostfunction}
-\end{equation}
-%
-Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
-predictions for the output $Y$. $loss$ is a suitable loss function for the
-domain of $Y$. Commonly, the misclassification error for classification
-problems, or the squared error for regression problems, is used. $comp$
-quantifies the complexity of trees and penalizes for more complex models in the
-tree selection process. In CART, the complexity is a non-negative function of
-the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
-$\Theta_m$.
+where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
+domain of $Y$; typically, the misclassification rate MC 
+and the mean squared error MSE are employed for classification and regression,
+respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
+non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
+more complex models in the tree selection selection process.
+Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.
 
@@
 \cite{hyafil1976cob} showed that building optimal binary decision trees, such
-that the expected number of tests required to classify an unknown sample is
+that the expected number of splits required to classify an unknown sample is
 minimized, is NP-complete. \cite{zantema2000finding} proved that finding a
-decision tree of minimal size that is decision equivalent to a given decision
-tree is an NP-hard problem. As a consequence the search space is usually limited
+decision tree of minimal size that is decision-equivalent to a given decision
+tree is also NP-hard. As a consequence the search space is usually limited
 by heuristics.
@@
 
-Instead of searching all combination of $\Theta$ simultaneously, recursive
+Instead of searching all combinations in $\Theta$ simultaneously, recursive
 partitioning algorithms only consider one split at a time. At each internal node
-$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
-split point $s_r$ is selected such that it minimizes the impurity in the two
-resulting terminal nodes. Let the parameters of a tree, constructed in
-$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
-(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
-position of the new internal node $r$, as well as the corresponding split
-variable $v_r$ and split point $s_r$, are chosen such that:
-%
-\begin{equation}
-		\min_{v_r, s_r} \; \mathit{loss}(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
-	\label{eq:whichsplit}
-\end{equation}
-%
-CART's standard measures for node impurity are the Gini index, for
-classification problems and the mean squared error, for regression problems.
-After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
-$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
-process of recursively selecting and executing splits continues until the
-homogeneity of the nodes cannot be further improved. Additionally, often
-additional stopping criteria are defined, e.g. minimum sample sizes in the
-internal nodes and terminal nodes.
-
-In CART, and many other recursive tree methods, the tree growing step is
-followed by a pruning step. The idea behind pruning is to first grow a large
-tree without or with little restriction on the magnitude of impurity reduction.
-Starting from the bottom of the tree pruning then removes worthless splits. The
-method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
-used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
-iteratively prunes the internal node that:
-%
-\begin{equation}
-		\min_{r=1,...,n_{M-1}} \; \mathit{loss}(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + \mathit{comp}(\tilde{\theta}_i \setminus (v_r, s_r)).
+$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
+split point $s_r$ are selected to locally minimize the loss function.
+Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
+recursively for and subsequently \emph{pruned} to satisfy the complexity tradeoff:
+%
+\begin{eqnarray} \label{eq:locally}
+  \bar \theta_r
+    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
+    \qquad (r = 1, \dots, M_{\max} -1), \\
+  \bar \theta
+    & = & \argmin_{\bar \theta_r} \; \mathrm{loss}\{Y, f(X, \bar \theta_r)\} ~+~ \mathrm{comp}(\bar \theta_r).
   \label{eq:pruning}
-\end{equation}
-%
-The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
-(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.
-
-For nontrivial problems, recursive partitioning methods only search a small
-fraction of the global search space $(v_1, s_1, ..., v_{M_{max}-1},
-s_{M_{max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
-the other split rules. The selection of $v_r$ and $s_r$ at each step $i$
-requires a search of the locally optimal split over $M=i$ terminal nodes. The
-pruning step requires $j$ selections of an internal node $r$.
-
-In contrast to CART and C4.5, conditional inference trees utilize an unbiased
-variable selection approach that is based on permutation testing. Bias in the
-split variable selection process arises from a preference towards variables with
-many potential split points. The conditional inference framework overcomes this
-problem by separating the variable and split point selection. Although this
-distinction influences the selection of split rules, conditional inference tree
-solve the same minimization problem (Equation~\ref{eq:whichsplit}). The
-\code{ctree} implementation of conditional inference trees does not employ
-pruning, but uses a statistical stopping criterion.
+\end{eqnarray}
+%
+For nontrivial problems, forward-search recursive partitioning methods only search a small
+fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
+s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
+the subsequent split rules, hence typically leading to a globally suboptimal solution
+$\bar \theta$.
+
+Note that the notation above uses an exhaustive search for the $r$-th split,
+jointly over $(v_r, s_r)$, as is employed in CART or C4.5. So-called \emph{unbiased}
+recursive partitioning techniques modify this search by first seleting the variable $v_r$ using
+statistical significance tests and subsequently selecting the optimal split $s_r$ for that
+particular variable. This approach is used in conditional inference trees
+\citep[see][for references to other algorithms]{hothorn2006urp} and avoids selecting
+variables with many potential splits more often than those with fewer potential
+splits.
 
@@
 
-A very simple example that illustrates the limitation of recursive partitioning
+A very simple example that illustrates the limitation of forward-search recursive partitioning
 methods is depicted in Figure~\ref{fig:illustration}. The example only contains
-two independent variables and can be solved with only three splits that
+two independent variables and can be solved with three splits that
 partition the input space into four regions. As expected the recursive
 partitioning methods \code{rpart} and \code{ctree} fail to find any split at
-all, as the impurity of the resulting subsets cannot be reduced by the first
+all, as the loss function on the resulting subsets cannot be reduced by the first
 split. For methods that explore $\Theta$ in a more global fashion it is
-straightforward to find an optimal solution to this problem. One solution, which
-is the tree constructed by \code{evtree} is given by: 
+straightforward to find an optimal solution to this problem. One solution
+is the tree constructed by \code{evtree}: 
 %
@@
 
-All instances all are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
+All instances are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
 \code{O} are assigned to each of the terminal nodes~4 and~6.
@@
 
-When compared with the described forward stepwise search, a less greedy approach is to calculate the effects of the choice of the split rule deeper down in the tree.
-In this way optimal trees can be found for simple problems.
-However, feature selection at a node of greedily induced trees has complexity  $O(PN)$.
-Through a search of the optimal split up to $d-1$ levels the complexity increases to $O(P^dN^d)$ \citep{papagelis2001bdt}.
-Therefore, for more difficult problems the search space at each node must be limited by the depth of the search.
-Additionally, the search space can be reduced, limiting the number of possible split points of the selected variables and/or other heuristics.
-These methods are known as \textit{look-ahead algorithms} \citep{esmeir2007ald}.
-Another class of algorithms is given by stochastic optimization methods.
-Given an initial tree, these methods seek improved solutions through stochastic changes to the tree structure.
-Besides evolutionary algorithms \citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and \textit{simulated annealing} \citep{sutton1992ict} were used successfully to solve difficult classification and regression tree problems.
-\cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees.
-\cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several datasets from the UCI machine learning repository \citep{Frank+Asuncion:2010}.
-A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.
-\cite{cantu2003inducing} used an evolutionary algorithm to induce oblique classification trees.
-\cite{gray2008cta} introduced a classification tree algorithm that allows univariate splits, but also splits defined by linear combinations of two or three variables.
+When compared with the described forward stepwise search, a less greedy approach
+is to calculate the effects of the split rules deeper down in the
+tree. In this way optimal trees can be found for simple problems. However,
+split selection at a given node in Equation~\ref{eq:locally} has complexity  $O(PN)$
+(if all $P$ variables are numeric/ordered with $N$ distinct values).
+Through a global search up to $M-1$ levels (\ref{eq:globally}) the complexity
+increases to $O(P^M N^M)$ \citep{papagelis2001bdt}. One conceivable compromise
+between these two extremes is to look ahead $d$~steps with $1 < d < M$
+\citep[see e.g.,][]{esmeir2007ald}, also yielding a locally optimal tree but
+less constrained than that from a 1-step-ahead search.
+Another class of algorithms is given by stochastic optimization methods that,
+given an initial tree, seek improved solutions through stochastic changes to
+the tree structure. Thus, these algorithms try to explore the full parameter
+space $\Theta$ but cannot be guaranteed to find the globally optimal solution
+but only an approximation thereof. Besides evolutionary algorithms
+\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
+\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
+solve difficult classification and regression tree problems.
+\cite{koza1991concept} first formulated the concept of using evolutionary
+algorithms as a stochastic optimization method to build classification and
+regression trees. \cite{papagelis2001bdt} presented a classification tree
+algorithm and provided results on several datasets from the UCI machine learning
+repository \citep{Frank+Asuncion:2010}. Another method for the construction of
+classification and regression trees via evolutionary algorithms was introduced by
+\cite{gray2008cta} and \cite{fan2005rta}, respectively.
+\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
+classification trees.
 
@@
 
-The general framework of evolutionary algorithms emerged from different representatives.
-\cite{Holland} called his method \textit{genetic algorithms}, \cite{rechenberg} invented \textit{evolution strategies}, and \cite{Fogel} introduced \textit{evolutionary programming}.
-More recently, \cite{Koza} introduced a fourth stream and called it \textit{genetic programming}.
-All four representatives only differ in the technical details, for example the encoding of the individual solutions, but follow the same general outline \citep{eiben2003iec}.
-Evolutionary algorithms are being increasingly widely applied to a variety of optimization and search problems.
-Common areas of application include data mining \citep{freitas2002sea,cano2003uea}, statistics \citep{de34glmulti}, signal and image processing \citep{man1997gac}, and planning and scheduling \citep{jensen2001raf}.
-
-The pseudocode for the general scheme of an evolutionary algorithm is depicted in Table~\ref{pc:genericEA}.
-With the description of these components together with the definition of individuals, the most important aspects of an evolutionary algorithm are specified.
-The representation of an individual is given by a binary tree structure, which is fully defined with the definition of $\theta$ in Equation~\ref{treeRep}.
-This section continues with the description of the remaining components following the general outline displayed in Table~\ref{pc:genericEA}.
-
-\begin{table}[t!]
+The general framework of evolutionary algorithms emerged from different
+representatives. \cite{Holland} called his method \textit{genetic algorithms},
+\cite{rechenberg} invented \textit{evolution strategies}, and \cite{Fogel}
+introduced \textit{evolutionary programming}. More recently, \cite{Koza}
+introduced a fourth stream and called it \textit{genetic programming}. All four
+representatives only differ in the technical details, for example the encoding
+of the individual solutions, but follow the same general outline
+\citep{eiben2003iec}. Evolutionary algorithms are being increasingly widely
+applied to a variety of optimization and search problems. Common areas of
+application include data mining \citep{freitas2002sea,cano2003uea}, statistics
+\citep{de34glmulti}, signal and image processing \citep{man1997gac}, and
+planning and scheduling \citep{jensen2001raf}.
+
+\begin{table}[b!]
 \centering
@@
 \hrulefill
-\caption{Pseudocode of the general schema of an evolutionary algorithm.} 
+\caption{Pseudocode of the general evolutionary algorithm.} 
 \label{pc:genericEA}
@@
 
-This section describes the general functioning of the \code{evtree} algorithm.
-Implementation details as user-specified parameters, including their default values and their possible ranges, are discussed in the subsequent section.
-These implementation details also include parameters that specify a minimum number of observation that are required in each internal node and  terminal node of a tree.
-In the following, the term \textit{invalid node} refers to a node that do not contain enough observations.
+The pseudocode for the general evolutionary algorithm is provided
+in Table~\ref{pc:genericEA}. In the context of classification and regression trees,
+all \emph{individuals} from the population (of some given size) are $\theta$s as defined in 
+Equation~\ref{eq:tree}. The details of their evolutionary selection is
+given below following the general outline displayed in Table~\ref{pc:genericEA}.
+
+As pointed out in Section~\ref{got}, some elements $\theta \in \Theta$ are typically
+excluded in practice to satisfy minimal subsample size requirements.
+In the following, the term \emph{invalid node} refers to such excluded cases,
+not meeting sample size restrictions.
 
@@
 
-Each tree of the population is initialized with a valid, randomly generated, split rule in the root node.
-First, $v_1$ is selected with uniform probability from $1,...,P$.
-Second, a split point $s_1$ is selected.
-If $X_{v_1}$ is numeric or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_1}$.
-If $X_{v_1}$ is nominal and has $K$ categories, each $k=1,...,K$ has a probability of $50\%$ to be assigned to the left or the right terminal node.
-In cases where all $k$ are allocated to the same terminal node, the assignment of one category is flipped to the other terminal node.
-If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated.
-With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the initialized tree is fully described by $\tilde{\theta}_{1}= (v_1, s_1)$.
+Each tree of the population is initialized with a valid, randomly generated,
+split rule in the root node. First, $v_1$ is selected with uniform probability
+from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
+or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
+probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is
+nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
+be assigned to the left or the right daughter node. In cases where all $k$ are
+allocated to the same terminal node, the assignment of one category is flipped
+to the other terminal node. If this procedure results in a non-valid split rule,
+the two steps of random split variable selection and split point selection are
+repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the
+initialization is complete and each individual of the population of trees is of
+type $\tilde{\theta}_{1}= (v_1, s_1)$.
 
@@
 
-In every iteration, each tree is selected once to be modified by one of the variation operators.
-In cases where the crossover operator is applied, the second parent is selected randomly from the remaining population.
-In this way, some trees are selected more than once in each iteration.
+In every iteration, each tree is selected once to be modified by one of the
+variation operators. In cases where the crossover operator is applied, the
+second parent is selected randomly from the remaining population. In this way,
+some trees are selected more than once in each iteration.
 
@@
 
-Four types of mutation operators and one crossover operator are utilized by our algorithm.
-In each modification step, one of the variation operators is randomly selected for each tree.
-The mutation and crossover operators are described below.
+Four types of mutation operators and one crossover operator are utilized by our
+algorithm. In each modification step, one of the variation operators is randomly
+selected for each tree. The mutation and crossover operators are described
+below.
 
@@
 
-\textit{Split} selects a random terminal-node and assigns a valid, randomly generated, split rule to it.
-As a consequence, the selected terminal node becomes an internal node $r$ and two new terminal nodes are generated.
-
-The split rule of the new internal node is chosen as follows.
-First, $v_r$ is selected with uniform probability from $1,...,P$.
-Second, a split point $s_r$ is selected.
-If $X_{v_r}$ is numeric or ordinal, a split point $s_r$ is selected with uniform probability from the $u-1$ possible spit points of $X_{v_r}$.
-If $X_{v_r}$ is nominal, each category of $X_{v_r}$, has a probability of $50\%$ to be assigned to the left or the right terminal node.
-In cases where all $k$ are allocated to the same terminal node, one of them is changed to the other terminal node.
-If this procedure results in a non-valid split rule, the two steps of random split variable selection and split point selection are repeated for a maximum of $P$ iterations.
-
-In cases where no valid split rule can be assigned to the internal node at position $r$, the search for a valid split rule is carried out on another randomly selected terminal node.
-If, after $10$ attempts, no valid split rule can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.
-Otherwise, the set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}= \tilde{\theta}_i \cup (v_{r},s_{r})$.
+\textit{Split} selects a random terminal-node and assigns a valid, randomly
+generated, split rule to it. As a
+consequence, the selected terminal node becomes an internal node $r$ and
+two new terminal nodes are generated.
+
+The search for a valid split rule is conducted as in see Section~\ref{init}
+for a maximum of $P$ iterations. 
+In cases where no valid split rule can be assigned to the internal node at
+position $r$, the search for a valid split rule is carried out on another
+randomly selected terminal node. If, after 10~attempts, no valid split rule
+can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$. Otherwise, the
+set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}=
+\tilde{\theta}_i \cup (v_{r},s_{r})$.
 
 \subsubsection{Prune}
-\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two terminal nodes as successors and prunes it into a terminal node.
-The tree's parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i} \setminus (v_{r},s_{r})$.
-If $\tilde{\theta}_{i}$ only comprises one internal node, i.e., the root node, than $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.
+
+\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two
+terminal nodes as successors and prunes it into a terminal node. The tree's
+parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i}
+\setminus (v_{r},s_{r})$. If $\tilde{\theta}_{i}$ only comprises one internal
+node, i.e., the root node, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.
 
 \subsubsection{Major split rule mutation}
-\textit{Major split rule mutation} selects a random internal node $r$ and changes the split rule, defined by the corresponding split variable $v_r$, and the split point $s_r$.
-With a probability of $50\%$, a value from the range $1,...,P$ is assigned to $v_r$.
-Otherwise $v_r$ remains unchanged and only $s_r$ is modified.
-Again, depending on the domain of $X_{v_r}$, either a random split point from the range of possible values of $X_{v_r}$ is selected, or a non-empty set of categories is assigned to each of the two terminal nodes.
-If the split rule at $r$ becomes invalid, the mutation operation is reversed and the procedure, starting with the selection of $r$, is repeated for a maximum of $3$ attempts.
-Subsequent nodes that become invalid are pruned.
-
-If no pruning occurs, $\tilde{\theta}_{i}$ and $\tilde{\theta}_{i+1}$ contain the same set of parameters.
-Otherwise, the set of parameters $(v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$, corresponding to invalid nodes, is removed from $\tilde{\theta}_i$.
-Thus, $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.
+
+\textit{Major split rule mutation} selects a random internal node $r$ and
+changes the split rule, defined by the corresponding split variable $v_r$, and
+the split point $s_r$. With a probability of $50\%$, a value from the range
+$1,...,P$ is assigned to $v_r$. Otherwise $v_r$ remains unchanged and only $s_r$
+is modified. Again, depending on the domain of $X_{v_r}$, either a random split
+point from the range of possible values of $X_{v_r}$ is selected, or a non-empty
+set of categories is assigned to each of the two terminal nodes. If the split
+rule at $r$ becomes invalid, the mutation operation is reversed and the
+procedure, starting with the selection of $r$, is repeated for a maximum of
+3~attempts. Subsequent nodes that become invalid are pruned.
+
+If no pruning occurs, $\tilde{\theta}_{i}$ and $\tilde{\theta}_{i+1}$ contain
+the same set of parameters. Otherwise, the set of parameters $(v_{m_1}, s_{m_1},
+..., v_{m_f}, s_{m_f})$, corresponding to invalid nodes, is removed from
+$\tilde{\theta}_i$. Thus, $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus
+(v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.
 
@@
 
-\textit{minor split rule mutation} is similar to the \textit{major split rule mutation} operator, but it does not alter $v_r$ and only changes the split point $s_r$ by a minor degree.
-If $X_{v_r}$ is numerical or ordinal the split point $s_r$ is changed by a non-zero number of unique values of $X_{v_r}$.
-In cases where $X_{v_r}$ has less then $20$ unique values, the split point is change to to the next larger, or the next lower, unique value of $X_{v_r}$.
-Otherwise, $s_r$ is randomly shifted by a number of unique values that is not larger than $10\%$ of the range of unique values of $X_{v_r}$.
-If $X_{v_r}$ is a nominal variable, with less than $20$ categories, one of the categories is randomly modified.
-Otherwise, at least one and at most $10\%$ of the variable's categories are changed.
-In cases where subsequent nodes become invalid, further split points are searched that preserve the tree's topology.
-After five non-successful attempts at finding a topology preserving split point, the non-valid nodes are pruned.
-
-Equivalently to the \textit{major split rule mutation} operator the subsequent solution $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.
+\textit{Minor split rule mutation} is similar to the \textit{major split rule
+mutation} operator, but it does not alter $v_r$ and only changes the split point
+$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
+$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
+where $X_{v_r}$ has less then $20$ unique values, the split point is change to
+to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
+$s_r$ is randomly shifted by a number of unique values that is not larger than
+$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
+variable, with less than $20$ categories, one of the categories is randomly
+modified. Otherwise, at least one and at most $10\%$ of the variable's
+categories are changed. In cases where subsequent nodes become invalid, further
+split points are searched that preserve the tree's topology. After five
+non-successful attempts at finding a topology preserving split point, the
+non-valid nodes are pruned.
+
+Equivalently to the \textit{major split rule mutation} operator the subsequent
+solution $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1},
+..., v_{m_f}, s_{m_f})$.
 
@@
 
-\textit{Crossover} exchanges, randomly selected, subtrees between two trees.
-Let $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees that are chosen for crossover.
-First, an internal node is selected randomly from each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as $n_k^1$ and $n_k^2$.
-Let $sub1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees rooted by $n_k^j$, where $j=1,2$.
-Then, $sub1(\tilde{\theta}_i^j)$ contains the parameters $(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, ...,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set of descendent node positions of $n^j_k$.
-Now, the two selected subtrees can be written as $sub1(\tilde{\theta}_{i}^1, n^1_k) $ and $sub1(\tilde{\theta}_{i}^2, n^2_k)$.
-Furthermore, $sub2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1 \setminus sub1(\tilde{\theta}_{i}^1, n^1_k)$ and $sub2(\tilde{\theta}_{i}^2, n^2_k) = \tilde{\theta}_{i}^2 \setminus sub1(\tilde{\theta}_{i}^2, n^2_k)$.
-
-After the application of the \textit{crossover} operator, the trees' parameter spaces are given by $\tilde{\theta}_{i+1}^1= sub2(\tilde{\theta}_{i}^1, n^1_k) \cup sub1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2= sub2(\tilde{\theta}_{i}^2, n^2_k) \cup sub1(\tilde{\theta}_{i}^1, n^1_k) \setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$.
-Here, $(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset sub1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ..., v_{m^2_{f}}, s_{m^2_{f}}) \subset sub1(\tilde{\theta}_i^2, n^2_k)$ are the parameters of the invalid nodes.
+\fixme{I really had problems following this section. The sub-operator is useful
+but is it possible to introduce this with fewer sub- and super-scripts? It makes
+it very challenging to read.}
+
+\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
+$\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees
+that are chosen for crossover. First, an internal node is selected randomly from
+each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
+$n_k^1$ and $n_k^2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees
+rooted by $n_k^j$ ($j=1,2$). Then, $\mathrm{sub}1(\tilde{\theta}_i^j) =
+(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, \dots ,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set
+of descendent node positions of $n^j_k$. Now, the two selected subtrees can be
+written as $\mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k) $ and $\mathrm{sub}1(\tilde{\theta}_{i}^2,
+n^2_k)$. Furthermore, $\mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1
+\setminus \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)$ and $\mathrm{sub}2(\tilde{\theta}_{i}^2,
+n^2_k) = \tilde{\theta}_{i}^2 \setminus \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k)$.
+
+After the application of the \textit{crossover} operator, the trees' parameter
+spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k)
+\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ...,
+v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2=
+\mathrm{sub}2(\tilde{\theta}_{i}^2, n^2_k) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)
+\setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here,
+$(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset
+\mathrm{sub}1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ...,
+v_{m^2_{f}}, s_{m^2_{f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, n^2_k)$ are the
+parameters of the invalid nodes.
 
@@
 
-The evaluation function represents the requirements the population should adapt to.
-In general, these requirements are formulated by Equation~\ref{generalcostfunction}.
-A suitable evaluation function for classification and regression trees minimizes the models' accuracy on the training data, and the models' complexity.
-This subsection describes the currently implemented choices of evaluation functions for classification and for regression.
+The evaluation function represents the requirements the population should adapt
+to. In general, these requirements are formulated by Equation~\ref{eq:globally}.
+A suitable evaluation function for classification and regression trees minimizes
+the models' accuracy on the training data, and the models' complexity. This
+subsection describes the currently implemented choices of evaluation functions
+for classification and for regression.
 
@@
 
-The quality of a classification tree is most commonly measured as a function of its misclassification loss $MC(Y, f(X, \theta)) = \sum_{n=1}^N I(Y_n \neq \hat{f}(X_{\cdot n}, \theta))$, and the complexity of a tree by a function of the number of its terminal nodes $M$.
-\code{evtree} uses $2 \cdot MC(Y, f(X, \theta))$ as a loss function.
-The number of terminal nodes, weighted by $\ln N$ and a user-specified parameter $\alpha$, measures the complexity of trees.
-With these particular choices, the general definition of the optimization problem in Equation~\ref{generalcostfunction} can be written as:
-
-\begin{equation}
-	 \hat{\theta} = \argmin_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
-\label{eq:classification}
-\end{equation}
-
-Equation~\ref{eq:classification} seeks parameters $\hat{\theta}$ that minimize the misclassification loss and the number of terminal nodes.
-Increasing values of $\alpha$ encourage decreasing tree sizes.
-
-Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian information criterion} \citep[BIC, as in][]{gray2008cta} and
+The quality of a classification tree is most commonly measured as a function of
+its misclassification MC and the complexity of a tree by a function
+of the number of its terminal nodes $M$. \code{evtree} uses $2 N \cdot
+\mathrm{MC}(Y, f(X, \theta))$ as a loss function. The number of terminal nodes,
+weighted by $\log N$ and a user-specified parameter $\alpha$, measures the
+complexity of trees.
+%
+\begin{eqnarray}
+  \mathrm{loss}(Y, f(X, \theta)) & = & 2 N \cdot \mathrm{MC}(Y, f(X, \theta)) \nonumber \\
+                                 & = & 2 \cdot \sum_{n=1}^N I(Y_n \neq f(X_{\cdot n}, \theta)),  \label{eq:classification} \\
+  \mathrm{comp}(\theta)          & = & \alpha \cdot M \cdot \log N. \nonumber
+\end{eqnarray}
+%
+With these particular choices, Equation~\ref{eq:globally} seeks trees
+$\hat{\theta}$ that minimize the misclassification loss at a BIC-type tradeoff
+with the number of terminal nodes.
+
+Other, existing and commonly used choices of evaluation functions include the
+\textit{Bayesian information criterion} \citep[BIC, as in][]{gray2008cta} and
 \textit{minimum description length} \citep[MDL, as in][]{quinlan1989inferring}.
-For both evaluation functions deviance is used for accuracy estimation.
-Deviance is usually preferred over the misclassification rate in recursive partitioning methods, as it is more sensitive to changes in the node probabilities \citep[pp.~308--310]{hastie}.
-However this is not necessarily an advantage for global tree building methods like evolutionary algorithms.
+For both evaluation functions deviance is used for accuracy estimation. Deviance
+is usually preferred over the misclassification rate in recursive partitioning
+methods, as it is more sensitive to changes in the node probabilities
+\citep[pp.~308--310]{hastie}. However, this is not necessarily an advantage for
+global tree building methods like evolutionary algorithms.
 
@@
 
-For regression trees, accuracy is usually measured by the mean squared error $\mathit{MSE}(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$.
-Using $N \cdot \ln \mathit{MSE}(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:
-
-\begin{equation}
-	 \hat{\theta} = \argmin_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
-\label{eq:regression}
-\end{equation}
-
-Here, $M+1$ is the effective number of estimated parameters, taking into account the estimates of a mean parameter in each of the terminal nodes and the constant error variance term.
-With $\alpha = 0.25$ the criteria is, up to a constant, equivalent to the BIC used by \cite{fan2005rta}.
-However, the effective number of parameters estimated for is actually much higher than $M+1$ due to the selection of parameters in the split variable and the selection of the variable itself.
-It is however unclear how these should be counted \citep[p.~222]{gray2008cta, ripley}.
-Therefore, a more conservative default value of $\alpha=1$ is assumed.
+For regression trees, accuracy is usually measured by the mean squared error MSE.
+Here, it is again coupled with a BIC-type complexity measure:
+
+Using $N \cdot \log \mathrm{MSE}$ as a loss function and
+$\alpha \cdot 4 \cdot (M+1) \cdot \log N$ as the complexity part, the general
+formulation of the optimization problem in  can be
+rewritten as:
+%
+\begin{eqnarray} 
+  \mathrm{loss}(Y, f(X, \theta)) & = & N \log \mathrm{MSE}(Y, f(X, \theta)) \nonumber \\
+                                 & = & N \log \left\{ \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2) \right\}, \label{eq:regression} \\
+  \mathrm{comp}(\theta)          & = & \alpha \cdot 4 \cdot (M+1) \cdot \log N. \nonumber
+\end{eqnarray}
+%
+Here, $M+1$ is the effective number of estimated parameters, taking into account
+the estimates of a mean parameter in each of the terminal nodes and the constant
+error variance term. With $\alpha = 0.25$ the criteria is, up to a constant,
+equivalent to the BIC used by \cite{fan2005rta}. However, the effective number
+of parameters estimated for is actually much higher than $M+1$ due to the
+selection of parameters in the split variable and the selection of the variable
+itself. It is however unclear how these should be counted
+\citep[p.~222]{gray2008cta, ripley}. Therefore, a more conservative default
+value of $\alpha=1$ is assumed.
 
@@
 
-The population size stays constant during the evolution and only a fixed subset of the candidate solution can be kept in memory.
-A common strategy is the $(\mu + \lambda)$ selection, where $\mu$ survivors for the next generation are selected from the union of $\mu$ parents and $\lambda$ offsprings.
-An alternative approach is the $(\mu,\lambda)$ strategy.
-In this schema $\mu$ survivors for the next generation are selected from $\lambda$ offsprings.
-
-Our algorithm uses a $(1+1)$ selection schema, where one parent solution competes with one offspring for a place in the population.
-In the case of a mutation operator either the solution before $\tilde{\theta}_i$ or after modification $\tilde{\theta}_{i+1}$ is kept in memory.
-In the case of the crossover operator, the initial solutions of $\tilde{\theta}_i^1$ competes with its subsequent solutions $\tilde{\theta}_{i+1}^1$.
-Correspondingly, one of the two solutions $\tilde{\theta}_i^2$ and $\tilde{\theta}_{i+1}^2$ is rejected.
-The survivor selection is done deterministically.
-The tree with lower fitness, according to the evaluation function, is rejected.
-Note that, due to the definition of the crossover operator, that some trees are selected more than once in an iteration.
-Correspondingly, these trees undergo the survival selection process more than once in an iteration.
-
-As in classification and regression tree analysis the individual solutions are represented by trees.
-This design offers computational advantages over $(\mu + \lambda)$, with $\mu>1 \;  \& \; \lambda>1$, and $(\mu,\lambda)$ strategies.
-In particular, for the application of mutation operators no new trees have to be constructed.
-The tree after modification is simply accepted or reversed to the previous solution.
-
-There are two important issues in the evolution process of an evolutionary algorithm: population diversity and selective pressure \citep{michalewicz1994genetic}.
-These factors are related, as with increasing selective pressure the search is focused more around the currently best solutions.
-An overly strong selective pressure can cause the algorithm to converge early in a local optima.
-On the other hand, an overly weak selective pressure can make the search ineffective.
-Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in situations as follows.
-Suppose the $b^{th}$ tree of the population is one of the fittest trees in iteration $i$, and in iteration $i$  one split rule of the $b^{th}$ tree is changed only by a minor degree.
-Then very few instances are classified differently and the overall misclassification might not even change.
-However, as they represent one of the best solutions in iteration $i$, they are both selected for the subsequent population.
-This situation can occur frequently, especially when a fine-tuning operator like the \textit{minor split rule mutation} is used.
-Then, the diversity of different trees is lost quickly and the algorithm likely terminates in a local optimum.
-The $(1+1)$ selection mechanism clearly avoids these situations, as only the parent or the offspring can be part of the subsequent population.
+The population size stays constant during the evolution and only a fixed subset
+of the candidate solutions can be kept in memory. A common strategy is the $(\mu
++ \lambda)$ selection, where $\mu$ survivors for the next generation are
+selected from the union of $\mu$ parents and $\lambda$ offsprings. An
+alternative approach is the $(\mu,\lambda)$ strategy where $\mu$
+survivors for the next generation are selected from $\lambda$ offsprings.
+
+Our algorithm uses $(1+1)$ selection, where one parent solution
+competes with one offspring for a place in the population. In the case of a
+mutation operator either the solution before $\tilde{\theta}_i$ or after
+modification $\tilde{\theta}_{i+1}$ is kept in memory. In the case of the
+crossover operator, the initial solutions of $\tilde{\theta}_i^1$ competes with
+its subsequent solutions $\tilde{\theta}_{i+1}^1$. Correspondingly, one of the
+two solutions $\tilde{\theta}_i^2$ and $\tilde{\theta}_{i+1}^2$ is rejected. The
+survivor selection is done deterministically. The tree with lower fitness,
+according to the evaluation function, is rejected. Note that, due to the
+definition of the crossover operator, some trees are selected more than
+once in an iteration. Correspondingly, these trees undergo the survival
+selection process more than once in an iteration.
+
+As in classification and regression tree analysis the individual solutions are
+represented by trees. This design offers computational advantages over $(\mu +
+\lambda)$, with $\mu>1$ and $\lambda>1$, and $(\mu,\lambda)$ strategies. In
+particular, for the application of mutation operators no new trees have to be
+constructed. The tree after modification is simply accepted or reversed to the
+previous solution.
+
+There are two important issues in the evolution process of an evolutionary
+algorithm: population diversity and selective pressure
+\citep{michalewicz1994genetic}. These factors are related, as with increasing
+selective pressure the search is focused more around the currently best
+solutions. An overly strong selective pressure can cause the algorithm to
+converge early in local optima. On the other hand, an overly weak selective
+pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy, a
+strong selective pressure can occur in situations as follows. Suppose the $b$-th
+tree of the population is one of the fittest trees in iteration $i$, and in
+iteration $i$  one split rule of the $b$-th tree is changed only by a minor
+degree. Then very few instances are classified differently and the overall
+misclassification might not even change. However, as they represent one of the
+best solutions in iteration $i$, they are both selected for the subsequent
+population. This situation can occur frequently, especially when a fine-tuning
+operator like the \textit{minor split rule mutation} is used. Then, the
+diversity of different trees is lost quickly and the algorithm likely terminates
+in a local optimum. The $(1+1)$ selection mechanism clearly avoids these
+situations, as only the parent or the offspring can be part of the subsequent
+population.
 
@@
 
-The algorithm terminates when the quality of the best $5\%$ of trees stabilizes for $100$ iterations, but not before $1000$ iterations.
-If the run does not converge the algorithm terminates after a user-specified number of iterations.
-In cases where the algorithm does not converge, a warning message is written to the command line.
-The tree with the highest quality according to the evaluation function is returned.
+Using the default parameters,
+the algorithm terminates when the quality of the best $5\%$ of trees stabilizes
+for $100$ iterations, but not before $1000$ iterations. If the run does not
+converge the algorithm terminates after a user-specified number of iterations.
+In cases where the algorithm does not converge, a warning message is written to
+the command line. The tree with the highest quality according to the evaluation
+function is returned.
 
@@
 
-\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}.
-CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}.
-The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages.
-\code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models.
-The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data.
-The arguments of  \code{evtree()} are
+Package \pkg{evtree} provides an efficient implementation of an evolutionary
+algorithm that builds classification trees in \proglang{R}. CPU- and memory-
+intensive tasks are fully computed in \proglang{C++}, while the user interfaces
+and plot functions are written in \proglang{R}. The \code{.C()} interface
+\citep{chambers2008software} was used to pass arguments between the two
+languages. \pkg{evtree} depends on the \pkg{partykit} package 
+\citep{hothorn2009partykit}, which provides an infrastructure for representing,
+summarizing, and visualizing tree-structured models.
+
+\subsection{User interface}
+
+The principal function of the \pkg{evtree} package is the eponymous function
+\code{evtree()} taking arguments
 %
@@
 %
-where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data.
-\code{weights} is an optional vector specifying weights to be used in the fitting process.
-The option vector subset specifies a subset of observations to be used in the fitting process.
-\code{control} comprises a list of control parameters
+where \code{formula}, \code{data}, \code{weights}, and \code{subset} specify
+the data in the usual way, e.g., via \code{formula = y ~ x1 + x2}.
+Additionally, \code{control} comprises a list of control parameters
 %
@@
 %
-where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth.
-Note that the memory requirements increase by the square of the maximum tree depth.
-Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees.
-\code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population.
-With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined.
-Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed.
-If not specified, the random number generator is initialized by the system time.
-In this case, iterative calls of \code{evtree} will possibly result in different models.
-
-An interesting application for classification tree analysis is target marketing, where limited resources are aimed at a distinct group of potential customers.
-An example of a marketing case study is provided by the \textit{Bookbinder's Book Club} \citep{lilien2004} dataset.
-The data is a marketing case study about a (fictitious) American book club.
-In this case study, a brochure of the book \textit{The Art History of Florence} was sent to 20,000 customers and 1,806 of which bought the book.
-The dataset contains a subsample of 1,300 customers for building a predictive model for the customers choice of buying the advertised book.
-Besides predictive accuracy, model complexity is a crucial issue in this application.
-More simple trees are easier to interpret and communicable to marketing experts and management professionals.
-
-This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyze the \textit{Bookbinder's Book Club} data.
-In this session, all trees are constrained to have a minimum number of $10$ observations per terminal node.
-To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$.
-To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}.
-The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
+where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth}
+constrain the solution to a minimum number of observations in each terminal
+node, a minimum number of observation in each internal node and a maximum tree
+depth. Note that the memory requirements increase by the square of the maximum
+tree depth. Parameter \code{alpha} regulates the complexity parameter
+$\alpha$ in Equation~\ref{eq:classification} and~\ref{eq:regression}, respectively.
+\code{niterations} and \code{ntrees} specify the maximum number of iterations
+and the number of trees in the population, respectively. With the argument
+\code{operatorprob}, user-specified probabilities for the variation operators
+can be defined. For making computations reproducible, argument \code{seed} is
+an optional integer seed for the random number generator (at \proglang{C++} level).
+If not specified, the random number generator is initialized by \verb|as.integer(runif(1, max = 2^16))|
+in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
+If set to \code{-1L}, the seed is initializied by the system time.
+
+The trees computed by \code{evtree} inherit from class `\code{party}' supplied
+by the \pkg{partykit} package. The methods inherited in this way
+include standard \code{print()}, \code{summary()}, and \code{plot()} functions
+to display trees and a \code{predict()} function to compute the fitted response
+or node number etc.
+
+\subsection{Case study: Customer targeting}
+
+An interesting application for classification tree analysis is target marketing,
+where limited resources are aimed at a distinct group of potential customers. An
+example is provided by \cite{lilien2004} in the \textit{Bookbinder's Book Club}
+marketing case study about a (fictitious) American book club. In this case study, a brochure of the book
+``The Art History of Florence'' was sent to 20,000~customers, 1,806~of
+which bought the book. The dataset contains a subsample of 1,300~customers for
+building a predictive model of customer choice.
+
+Besides predictive accuracy,
+model complexity is a crucial issue in this application: Smaller trees are easier to
+interpret and communicable to marketing experts and management professionals. 
+Hence, we use \code{evtree} with a maximal depth of two levels of splits only.
+This is contrasted with \code{rpart} and \code{ctree} with and without such a restriction
+of tree depth to show that the evolutionary search of the global parameter space
+can be much more effective in balancing predictive accuracy and complexity
+compared to forward-search recursive partitioning.
+
+All trees are constrained to have a minimum number of 10~observations per terminal node.
+Additionally, a significance level of $1\%$ is employed in the construction of
+conditional inference trees which is more appropriate than the default 5\%~level for
+$1,300$~observations. To provide uniform visualizations and predictions of the
+fitted models, `\code{party}' objects are used to represent all trees. For, `\code{rpart}'
+trees \pkg{partykit} provides a suitable \code{as.party()} method while a
+reimplementation of \code{ctree()} is provided in \pkg{partykit} (as opposed to
+the original in \pkg{party}) that directly leverages the `\code{party}' infrastructure.
+
+First, the data is loaded and the forward-search trees are grown
+with and without depth restriction, visualizing the unrestricted trees
+in Figure~\ref{fig:example_rpart_ctree}.
 %
@@
 rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
+rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
+  maxdepth = 2))
 ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
+ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
+  maxdepth = 2)
 plot(rp)
@@
 %
-The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
-%
-With the objective of building a smaller, but at least equally accurate tree,
-\code{evtree} is constrained to a maximum tree depth of $2$.
+With the objective of building a smaller, but at still accurate tree,
+\code{evtree} is constrained to a maximum tree depth of $2$,
+see Figure~\ref{fig:example_evtree}.
 %
@@
 if(cache) {
-  save(rp, ct, ev, file = "BBBClub-trees.rda")
+  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
 } else {
@@
 \end{minipage}
-\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel).
-The target variable is the customers choice of buying the book.
-The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
-}
+\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
+  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
+  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
+  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club
+  (\code{freq}) and the customer's \code{gender}.}
 \end{figure}
 %
-The resulting tree is displayed in the printed output below and visualized in
+The resulting evolutionary tree is printed output below and visualized in
 Figure~\ref{fig:example_evtree}.
@@
 @
-\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}.
-The target variable is the customers choice of buying the book.
-The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
+\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
+   The target variable is the customer's choice of buying the book. The variables used for splitting are
+   the number of art books purchased previously (\code{art}) and the number of months since the first purchase
+    \code{first}.}
 \end{figure}
-%
-While the \code{rpart} model has \Sexpr{width(rp)}~terminal nodes and a tree depth
-of~\Sexpr{depth(rp)}, the tree constructed by \code{ctree} has
-\Sexpr{width(ct)}~terminal nodes and a tree depth of~\Sexpr{depth(ct)}.
-With a tree depth of~\Sexpr{depth(ev)} and \Sexpr{width(ev)}~terminal nodes
-the \code{evtree} model is much simpler. 
+
+Not surprisingly, the explanatory variable \code{art} -- the number of art books purchased previously at the
+book club -- plays a key role in all constructed classification trees along with the number
+of months since the first purchase (\code{first}), the frequency of previous purchases
+(\code{freq}) and the customer's \code{gender}. Interestingly, though, the forward-search trees select
+the arguably most important variable in the first split while the evolutionary tree uses \code{first}
+in the first split and \code{art} in both second splits. Thus, the evolutionary tree
+uses a different cutoff in \code{art} for bookclub members that joined in the last year as opposed
+to older customers. While the former are predicted to be buyers if they previously bought
+at least one art book, the latter are predicted to purchase the advertised art book only if they
+previously bought at least two other art books. Certainly, this classification is easy to understand
+and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.
+
+However, we still need to answer the question how well it performs in contrast to the other trees.
+Hence, we set up a function \code{mc()} the computes the misclassification rate as a measure of
+predictive accuracy and a function \code{evalfun()} that computes the evaluation function
+(i.e., penalized by tree complexity) from Equation~\ref{eq:classification}.
 %
@@
 mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
-c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
-@
-The accuracy of the three models are similar. With a misclassification rate of
-$24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model
-($23.8\%$) and better than the \code{ctree} model ($24.8\%$). 
+evalfun <- function(obj) 2 * nrow(BBBClub) * mc(obj) +
+  width(obj) * log(nrow(BBBClub))
+trees <- list("evtree" = ev, "rpart" = rp, "ctree" = ct, "rpart2" = rp2,
+  "ctree2" = ct2)
+round(sapply(trees, function(obj) c("misclassification" = mc(obj),
+  "evaluation function" = evalfun(obj))), digits = 3)
+@
+%
+Not surprisingly the evolutionary tree \code{ev} outperforms the depth-restricted
+trees \code{rp2} and \code{ct2}, both in terms of misclassification and the penalized
+evaluation function. However, it is interesting to see that \code{ev} performs even better
+than the unrestricted conditional inference tree \code{ct} and is comparable in performance
+to the unrestricted CART tree \code{rp}. Hence, the practitioner may choose the evolutionary
+tree \code{ev} as it is the easiest to communicate.
+
+Although the constructed trees are considerably different, the code above shows
+that the predictive accuracy is rather similar. Moreover, below we see that the
+structure of the individual predictions on the dataset are rather similar as well:
 %
@@
   ctree  = predict(ct), observed = BBBClub$choice))
-
 sapply(c("evtree", "rpart", "ctree"), function(nam) {
@@
 %
-Although the constructed trees are considerably different, the structure of the predictions are relatively similar.
-In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$).
-Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $66.4\%$ (\code{ctree}) actually bought the book.
-In a final step \code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previously fitted \code{evtree} model.
-%
-<<evtree-performance>>=
-rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10, 
-maxdepth = 2))
-ct <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99, 
-maxdepth = 2)
-
-mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
-c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
-
-c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
-@
-%
-With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$).
-Due to its unbalanced structure, the \code{rpart} model has only $3$ terminal nodes.
-Hence, it does not use the maximum tree depth efficiently.
-In contrast, a global optimization algorithm like \code{evtree} can be used to search the best tree under given constraints.
+In this case, \code{evtree} classifies less
+customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$). However,
+\code{evtree} achieves the highest proportion of correct classification among
+the declared buyers: $72.6\%$ compared to $70.8\%$
+(\code{rpart}) and $66.4\%$ (\code{ctree}).
+
+In summary, this illustrates how \code{evtree} can be employed to better balance
+predictive accuracy and complexity by searching a larger space of potential trees.
+As a final note, it is worth pointing out that in this setup, several runs of
+\code{evtree()} with the same parameters typically lead to the same tree. However,
+this may not always be the case. Due to the stochastic nature of the search algorithm and
+the vast search space, trees with very different structures but similar evaluation
+function values may be found by subsequent runs of \code{evtree()}. Here, this
+problem is alleviated by restricting the maximal depth of the tree, yielding a
+clear solution.
 
@@
 
-In this Section we compare \code{evtree} with \code{rpart} and \code{ctree}.
-All models are constrained to a minimum number of 7 observations per terminal node, 20 observations per internal node and a maximum tree depth of 9.
-Apart from that, the default settings of the algorithms are used.
-
-In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark datasets that are publicly available and $3$ real-world datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}.
-The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 datasets.
-The misclassification rate on the \textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of predictive accuracy.
-Furthermore, the complexity is estimated by the number of terminal nodes.
-Missing values are currently not supported by \code{evtree}.
-The only dataset with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis.
-Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual datasets .
-
-In the second part (Section~\ref{comp2}) the algorithms' performances are assessed on an artificial problem that is simulated with different noise levels.
-The estimation of predictive accuracy and the number of terminal nodes is based on $250$ realizations for each simulation.
+In this section, we compare \code{evtree} with \code{rpart} and \code{ctree}
+in a more rigorous benchmark comparison.
+
+In the first part of the analysis (Section~\ref{comp1}) the algorithms are
+compared on 14~benchmark datasets that are publicly available and $3$ real-world
+datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system
+\citep{LKF2009}. The analysis is based on the evaluation of 250~bootstrap
+samples for each of the 20~datasets. The misclassification rate on the
+\textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of
+predictive accuracy. Furthermore, the complexity is estimated by the number of
+terminal nodes.
+
+In the second part (Section~\ref{comp2}) the algorithms' performances are
+assessed on an artificial chessboard problem that is simulated with different noise levels.
+The estimation of predictive accuracy and the number of terminal nodes is based
+on $250$ realizations for each simulation.
+
+All models are constrained to a minimum number of 7~observations per terminal node,
+20~observations per internal node and a maximum tree depth of~9. Apart from
+that, the default settings of the algorithms are used. For assessment
+of significant differences in predictive accuracy and complexity, respectively,
+Dunnett's correction from \proglang{R}~package \pkg{multcomp} \citep{multcomp} was used
+for calculating simultaneous $95\%$ confidence intervals on the individual datasets.
+
+As missing values are currently not supported by \code{evtree} (e.g., by surrogate splits),
+the 16~missing values in the \textit{Breast Cancer Database} -- the only dataset in the
+study with missing values -- were removed before analysis.
 
@@
 
-In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from the Austrian DRG-system are described.
-In the Austrian DRG-system, resources are allocated to hospitals by simple rules regarding diagnosis, selected individual services, and patient characteristics.
-Within this system regression tree analysis is performed to classify hospital patients into clinically meaningful and comprehensible groups that consume similar hospital resources.
-The datasets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure.
-The task of the regression tree analysis is to further divide these groups with the intention of finding groups with similar resource consumption.
-Possible split-variables are the patients' diagnoses, procedures, age, and gender.
-The dependent variable is the patients' length of stay, which has a good relationship with the total costs and is available \citep{theurl2007ihf}.
-Permission to use the data was granted by the \citet{BMG2009}.
-A more detailed description of the application can be found in \citet{grubinger2010regression}.
+In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from
+the Austrian DRG-system are described. 
+\fixme{The DRG description is far too long. All readers are waiting for the results
+of the benchmark now. One explanatory sentence -- or maybe two at most -- should be
+used to state the general problem and point to the manual pages.}
+In the Austrian DRG-system, resources are
+allocated to hospitals by simple rules regarding diagnosis, selected individual
+services, and patient characteristics. Within this system regression tree
+analysis is performed to classify hospital patients into clinically meaningful
+and comprehensible groups that consume similar hospital resources. The datasets
+MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502
+(acute affections of the respiratory tract and middle atelectasis) contain
+patient groups with the same main diagnosis or procedure. The task of the
+regression tree analysis is to further divide these groups with the intention of
+finding groups with similar resource consumption. Possible split-variables are
+the patients' diagnoses, procedures, age, and gender. The dependent variable is
+the patients' length of stay, which has a good relationship with the total costs
+and is available \citep{theurl2007ihf}. Permission to use the data was granted
+by the \citet{BMG2009}. A more detailed description of the application can be
+found in \citet{grubinger2010regression}.
 
@@
 
-The disadvantages of our algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification},
+The disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification},
 only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes.
@@
 The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.
-
-\subsection{Artificial Problem}
+\fixme{also briefly discuss the issue of instability of the evolutionary trees even for large populations and number of runs}
+
+\subsection{Artificial problem}
 \label{comp2}
@@
 
-In this paper, we presented the \pkg{evtree} package, which implements classification and regression trees that are evolved by through an evolutionary algorithm.
-The package uses standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data from the \pkg{partykit} package.
-As the learning step of our algorithm is computationally demanding, most calculations are conducted in \proglang{C++}.
-At the moment our algorithm does not support parallelism.
-However, we intend to extend \pkg{evtree} to parallel computing.
-
-In Section~\ref{comp} we compared \code{evtree} with \code{rpart} and \code{ctree} on 14 benchmark and 3 real-world datasets.
-The datasets varied by sample size, the number of classes, and by the number and types of attributes.
-\code{evtree} models offered at least comparable and, for the majority of the datasets, improved predictive accuracy.
-In an artificial problem of a $4 \times 4$ chessboard we demonstrated \code{evtree}'s performance in the presence of a higher order of interactions and noise.
-
-The goal of this package is not to replace the well-established algorithms like \code{rpart} and \code{ctree}.
-We rather want to provide \proglang{R} users with an alternative which, given sufficient amounts of time and main memory, is often more accurate than recursive partitioning methods.
-By the nature of the algorithm it is able to discover patterns which cannot be modeled by a greedy algorithm.
-As \code{evtree} models can be substantially different to recursively fitted models, it can be beneficial to use both approaches, as this may reveal additional relationships in the data.
+In this paper, we presented the \pkg{evtree} package, which implements
+classification and regression trees that are grown by through an evolutionary
+algorithm. The package uses standard \code{print()}, \code{summary()}, and
+\code{plot()} functions to display trees and a \code{predict()} function to
+predict the class labels of new data from the \pkg{partykit} package. As evolutionary
+learning of trees is computationally demanding, most calculations
+are conducted in \proglang{C++}. At the moment our algorithm does not support
+parallelism. However, we intend to extend \pkg{evtree} to parallel computing.
+
+The comparisons with recursive partitioning methods \code{rpart}
+and \code{ctree} in Sections~\ref{impl} and~\ref{comp} shows that \code{evtree}
+performs very well in a wide variety of settings, often balancing predictive
+accuracy and complexity better than the forward-search methods.
+
+However, the goal of \code{evtree} is not to replace the well-established algorithms like
+\code{rpart} and \code{ctree} but rather to complement the tree toolbox with
+an alternative method which may perform better given sufficient amounts of time and main memory.
+By the nature of the algorithm it is able to discover patterns which cannot be modeled by a greedy
+forward-search algorithm. As \code{evtree} models can be substantially different to recursively
+fitted models, it can be beneficial to use both approaches, as this may reveal
+additional relationships in the data.
 

### 2011-09-28 pkg/evtree/inst/doc/evtree.Rnw
fixed FIXMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@250 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
-recursively for and subsequently \emph{pruned} to satisfy the complexity tradeoff:
+recursively and subsequently \emph{pruned} to satisfy the complexity tradeoff:
 %
@@
 (if all $P$ variables are numeric/ordered with $N$ distinct values).
-Through a global search up to $M-1$ levels (\ref{eq:globally}) the complexity
-increases to $O(P^M N^M)$ \citep{papagelis2001bdt}. One conceivable compromise
-between these two extremes is to look ahead $d$~steps with $1 < d < M$
+Through a global search up to $D-1$ levels (\ref{eq:globally}) the complexity
+increases to $O(P^D N^D)$ \citep{papagelis2001bdt}. One conceivable compromise
+between these two extremes is to look ahead $d$~steps with $1 < d < D$
 \citep[see e.g.,][]{esmeir2007ald}, also yielding a locally optimal tree but
@@
 
-\fixme{I really had problems following this section. The sub-operator is useful
-but is it possible to introduce this with fewer sub- and super-scripts? It makes
-it very challenging to read.}
-
 \textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
@@
 each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
-$n_k^1$ and $n_k^2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, n^j_k)$ denote the subtrees
-rooted by $n_k^j$ ($j=1,2$). Then, $\mathrm{sub}1(\tilde{\theta}_i^j) =
-(v_{n^j_k}, s_{n^j_k},  v_{n^j_{k1}}, s_{n^j_{k1}}, \dots ,v_{n^j_{km}}, s_{n^j_{km}})$, where $(n^j_{k1}, ..., n^j_{km})$ are the set
-of descendent node positions of $n^j_k$. Now, the two selected subtrees can be
-written as $\mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k) $ and $\mathrm{sub}1(\tilde{\theta}_{i}^2,
-n^2_k)$. Furthermore, $\mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k) = \tilde{\theta}_{i}^1
-\setminus \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)$ and $\mathrm{sub}2(\tilde{\theta}_{i}^2,
-n^2_k) = \tilde{\theta}_{i}^2 \setminus \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k)$.
+$r_1$ and $r_2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ denote the subtrees
+rooted by $r_j$ ($j=1,2$). Thus, $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ contains $r_j$, 
+and its descendant nodes. Then, the complementary part of $\tilde{\theta}_{i}^j$ can be 
+defined as 
+$\mathrm{sub}2(\tilde{\theta}_{i}^j, r_j) =  \tilde{\theta}_{i}^j \setminus \mathrm{sub}1(\tilde{\theta}_{i}^j, r_j)$.
 
 After the application of the \textit{crossover} operator, the trees' parameter
-spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, n^1_k)
-\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, n^2_k) \setminus (v_{m^2_{1}}, s_{m^2_{1}}, ...,
-v_{m^2_{f}}, s_{m^2_{f}})$ and $\tilde{\theta}_{i+1}^2=
-\mathrm{sub}2(\tilde{\theta}_{i}^2, n^2_k) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, n^1_k)
-\setminus (v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}})$. Here,
-$(v_{m^1_{1}}, s_{m^1_{1}}, ..., v_{m^1_{f}}, s_{m^1_{f}}) \subset
-\mathrm{sub}1(\tilde{\theta}_i^1, n^1_k)$ and $(v_{m^2_{1}}, s_{m^2_{1}}, ...,
-v_{m^2_{f}}, s_{m^2_{f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, n^2_k)$ are the
-parameters of the invalid nodes.
+spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, r_1)
+\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, r_2) \setminus (v_{m_{21}}, s_{m_{21}}, ...,
+v_{m_{2f}}, s_{m_{2f}})$ and $\tilde{\theta}_{i+1}^2=
+\mathrm{sub}2(\tilde{\theta}_{i}^2, r_2) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, r_1)
+\setminus (v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}})$. Here,
+$(v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}}) \subset
+\mathrm{sub}1(\tilde{\theta}_i^1, r_1)$ and   
+$(v_{m_{21}}, s_{m_{21}}, ...,v_{m_{2f}}, s_{m_{2f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, r_2)$ 
+are the parameters of the invalid nodes.
 
@@
 %
-The resulting evolutionary tree is printed output below and visualized in
+The resulting evolutionary tree is printed below and visualized in
 Figure~\ref{fig:example_evtree}.
@@
 datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system
-\citep{LKF2009}. The analysis is based on the evaluation of 250~bootstrap
+\citep{LKF2011}. The analysis is based on the evaluation of 250~bootstrap
 samples for each of the 20~datasets. The misclassification rate on the
@@
 In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from
-the Austrian DRG-system are described. 
-\fixme{The DRG description is far too long. All readers are waiting for the results
-of the benchmark now. One explanatory sentence -- or maybe two at most -- should be
-used to state the general problem and point to the manual pages.}
-In the Austrian DRG-system, resources are
-allocated to hospitals by simple rules regarding diagnosis, selected individual
-services, and patient characteristics. Within this system regression tree
-analysis is performed to classify hospital patients into clinically meaningful
-and comprehensible groups that consume similar hospital resources. The datasets
-MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502
-(acute affections of the respiratory tract and middle atelectasis) contain
-patient groups with the same main diagnosis or procedure. The task of the
-regression tree analysis is to further divide these groups with the intention of
-finding groups with similar resource consumption. Possible split-variables are
-the patients' diagnoses, procedures, age, and gender. The dependent variable is
-the patients' length of stay, which has a good relationship with the total costs
-and is available \citep{theurl2007ihf}. Permission to use the data was granted
-by the \citet{BMG2009}. A more detailed description of the application can be
-found in \citet{grubinger2010regression}.
+the Austrian DRG system are described. In the Austrian DRG system, resources are 
+allocated to hospitals by simple rules mainly regarding the patients' diagnoses, procedures, and age. 
+Regression tree analysis is performed to model patient groups with similar resource consumption. 
+A more detailed description of the datasets and the application can 
+be found in \citet{grubinger2010regression}.
 
@@
 
-The disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification},
-only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes.
+Disadvantages of the \code{evtree} algorithm are computation time and memory requirements.
+While the smallest of the analyzed datasets, \textit{Glass identification}, only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes.
 The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit.
 The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.
-\fixme{also briefly discuss the issue of instability of the evolutionary trees even for large populations and number of runs}
+
+Another important issue to be considered is the random nature of evolutionary algorithms.
+For larger datasets, frequently, considerable different solutions exist that yield a similar or even the same evaluation function value.
+Therefore, subsequent runs of \code{evtree} can result in very different tree structures.
+This is not a problem if the tree is intended only for predictive purposes, and it is also not a big issue for many decision and prognosis tasks.
+Typically, in such applications, the resulting model has to be accurate, compact, and meaningful in its interpretation, but the particular tree structure is of secondary importance.
+Examples of such applications include the presented marketing case study and the Austrian DRG system.
+In cases where a model is not meaningful in its interpretation, the possibility of constructing different trees can even be beneficial.
+However, if the primary goal is to interpret relationships in the data, based on the selected splits, the random nature of the algorithm has to be considered.
 
@@
 In this paper, we presented the \pkg{evtree} package, which implements
-classification and regression trees that are grown by through an evolutionary
+classification and regression trees that are grown by an evolutionary
 algorithm. The package uses standard \code{print()}, \code{summary()}, and

### 2011-09-28 pkg/evtree/R/evtree.R

@@ -156,9 +156,11 @@ evtree <- function(formula, data, subset, na.action, weights, control = evtree.c
     out <- tree(nInstances, nVariables, varType, ndata, weights, prediction, splitV, splitP, csplit,
             maxNode, control)
  
-   	 if(out[[7]][1] < 0L) #trees could not be initialized
+   	if(out[[7]][1] < 0L) #trees could not be initialized
     	    stop("no split could be found")
-                            
+        if(out[[14]] == control$niterations)
+            warning("algorithm did not converge, computations may be unreliable")
+        	                    
         mtree = list()
         mtree$varType <-varType
         mtree$splitV <- out[[7]]        

### 2011-09-28 pkg/evtree/inst/doc/evtree.Rnw
some touch ups in notation
shortened/simplified crossover section further

two additional \pagebreak to prettify layout



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@251 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\pagebreak
+\pagebreak %%FIXME
 
@@
 Classification and regression tree analysis aims at
-modeling a response variable $Y$ by a vector of $P$ predictor variables
+modeling a response variable $Y$ by a vector of $P$~predictor variables
 $X=(X_1,...,X_P)$ where for classification trees $Y$ is
@@
 \begin{eqnarray} \label{eq:locally}
-  \bar \theta_r
+  \tilde \theta_r
     & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
     \qquad (r = 1, \dots, M_{\max} -1), \\
-  \bar \theta
-    & = & \argmin_{\bar \theta_r} \; \mathrm{loss}\{Y, f(X, \bar \theta_r)\} ~+~ \mathrm{comp}(\bar \theta_r).
+  \tilde \theta
+    & = & \argmin_{\tilde \theta_r} \; \mathrm{loss}\{Y, f(X, \tilde \theta_r)\} ~+~ \mathrm{comp}(\tilde \theta_r).
   \label{eq:pruning}
@@
 the subsequent split rules, hence typically leading to a globally suboptimal solution
-$\bar \theta$.
+$\tilde \theta$.
 
@@
 split selection at a given node in Equation~\ref{eq:locally} has complexity  $O(PN)$
-(if all $P$ variables are numeric/ordered with $N$ distinct values).
+(if all $P$~variables are numeric/ordered with $N$ distinct values).
 Through a global search up to $D-1$ levels (\ref{eq:globally}) the complexity
@@
 initialization is complete and each individual of the population of trees is of
-type $\tilde{\theta}_{1}= (v_1, s_1)$.
+type ${\theta}_{1}= (v_1, s_1)$.
 
@@
 The search for a valid split rule is conducted as in see Section~\ref{init}
-for a maximum of $P$ iterations. 
+for a maximum of $P$~iterations. 
 In cases where no valid split rule can be assigned to the internal node at
-position $r$, the search for a valid split rule is carried out on another
+position~$r$, the search for a valid split rule is carried out on another
 randomly selected terminal node. If, after 10~attempts, no valid split rule
-can be found, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$. Otherwise, the
-set of parameters in iteration $i+1$ are given by $\tilde{\theta}_{i+1}=
-\tilde{\theta}_i \cup (v_{r},s_{r})$.
+can be found, then ${\theta}_{i+1} = {\theta}_{i}$. Otherwise, the
+set of parameters in iteration $i+1$ are given by ${\theta}_{i+1}=
+{\theta}_i \cup (v_{r},s_{r})$.
 
@@
 terminal nodes as successors and prunes it into a terminal node. The tree's
-parameters at iteration $i+1$ are $\tilde{\theta}_{i+1}=\tilde{\theta}_{i}
-\setminus (v_{r},s_{r})$. If $\tilde{\theta}_{i}$ only comprises one internal
-node, i.e., the root node, then $\tilde{\theta}_{i+1} = \tilde{\theta}_{i}$.
+parameters at iteration $i+1$ are ${\theta}_{i+1}={\theta}_{i}
+\setminus (v_{r},s_{r})$. If ${\theta}_{i}$ only comprises one internal
+node, i.e., the root node, then ${\theta}_{i+1} = {\theta}_{i}$.
 
@@
 
-If no pruning occurs, $\tilde{\theta}_{i}$ and $\tilde{\theta}_{i+1}$ contain
+If no pruning occurs, ${\theta}_{i}$ and ${\theta}_{i+1}$ contain
 the same set of parameters. Otherwise, the set of parameters $(v_{m_1}, s_{m_1},
 ..., v_{m_f}, s_{m_f})$, corresponding to invalid nodes, is removed from
-$\tilde{\theta}_i$. Thus, $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus
+${\theta}_i$. Thus, ${\theta}_{i+1} = {\theta}_{i} \setminus
 (v_{m_1}, s_{m_1}, ..., v_{m_f}, s_{m_f})$.
@@
 Equivalently to the \textit{major split rule mutation} operator the subsequent
-solution $\tilde{\theta}_{i+1} = \tilde{\theta}_{i} \setminus (v_{m_1}, s_{m_1},
+solution ${\theta}_{i+1} = {\theta}_{i} \setminus (v_{m_1}, s_{m_1},
 ..., v_{m_f}, s_{m_f})$.
@@
 \textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
-$\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$ be the parameters of the trees
-that are chosen for crossover. First, an internal node is selected randomly from
-each of $\tilde{\theta}_{i}^1$ and $\tilde{\theta}_{i}^2$, which we denote as
-$r_1$ and $r_2$. Let $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ denote the subtrees
-rooted by $r_j$ ($j=1,2$). Thus, $\mathrm{sub}1(\tilde{\theta}_i^j, r_j)$ contains $r_j$, 
-and its descendant nodes. Then, the complementary part of $\tilde{\theta}_{i}^j$ can be 
-defined as 
-$\mathrm{sub}2(\tilde{\theta}_{i}^j, r_j) =  \tilde{\theta}_{i}^j \setminus \mathrm{sub}1(\tilde{\theta}_{i}^j, r_j)$.
-
-After the application of the \textit{crossover} operator, the trees' parameter
-spaces are given by $\tilde{\theta}_{i+1}^1= \mathrm{sub}2(\tilde{\theta}_{i}^1, r_1)
-\cup \mathrm{sub}1(\tilde{\theta}_{i}^2, r_2) \setminus (v_{m_{21}}, s_{m_{21}}, ...,
-v_{m_{2f}}, s_{m_{2f}})$ and $\tilde{\theta}_{i+1}^2=
-\mathrm{sub}2(\tilde{\theta}_{i}^2, r_2) \cup \mathrm{sub}1(\tilde{\theta}_{i}^1, r_1)
-\setminus (v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}})$. Here,
-$(v_{m_{11}}, s_{m_{11}}, ..., v_{m_{1f}}, s_{m_{1f}}) \subset
-\mathrm{sub}1(\tilde{\theta}_i^1, r_1)$ and   
-$(v_{m_{21}}, s_{m_{21}}, ...,v_{m_{2f}}, s_{m_{2f}}) \subset \mathrm{sub}1(\tilde{\theta}_i^2, r_2)$ 
-are the parameters of the invalid nodes.
+${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
+chosen from the population for crossover. First, an internal nodes $r_1$ and $r_2$
+are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
+Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_j$ rooted
+by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.
+Then, the complementary part of ${\theta}_{i}^j$ can be defined as 
+$\mathrm{sub}2({\theta}_{i}^j, r_j) = {\theta}_{i}^j \setminus \mathrm{sub}1({\theta}_{i}^j, r_j)$. 
+The crossover operator creates two new trees
+${\theta}_{i+1}^1 = \mathrm{sub}2({\theta}_{i}^1, r_1) \cup \mathrm{sub}1({\theta}_{i}^2, r_2)$ and
+${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
+If the crossover creates some invalid nods in either one of the new trees ${\theta}_{i+1}^1$
+or ${\theta}_{i+1}^2$, these are omitted.
 
@@
 competes with one offspring for a place in the population. In the case of a
-mutation operator either the solution before $\tilde{\theta}_i$ or after
-modification $\tilde{\theta}_{i+1}$ is kept in memory. In the case of the
-crossover operator, the initial solutions of $\tilde{\theta}_i^1$ competes with
-its subsequent solutions $\tilde{\theta}_{i+1}^1$. Correspondingly, one of the
-two solutions $\tilde{\theta}_i^2$ and $\tilde{\theta}_{i+1}^2$ is rejected. The
+mutation operator either the solution before ${\theta}_i$ or after
+modification ${\theta}_{i+1}$ is kept in memory. In the case of the
+crossover operator, the initial solutions of ${\theta}_i^1$ competes with
+its subsequent solutions ${\theta}_{i+1}^1$. Correspondingly, one of the
+two solutions ${\theta}_i^2$ and ${\theta}_{i+1}^2$ is rejected. The
 survivor selection is done deterministically. The tree with lower fitness,
@@
 \centering
-\setkeys{Gin}{width=0.7\textwidth}
+\setkeys{Gin}{width=0.75\textwidth} %FIXME: 0.75 just to avoid \pagebreak, should be 0.7
 <<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
@@
 
+\pagebreak %%FIXME
+
 \subsection{Artificial problem}

### 2011-09-28 pkg/evtree/inst/doc/evtree.Rnw
added NEWS file, touch-ups to D/M notation in vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@252 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
+\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
 \end{figure}
@@
 (if all $P$~variables are numeric/ordered with $N$ distinct values).
-Through a global search up to $D-1$ levels (\ref{eq:globally}) the complexity
+Through a global search up to $D$ levels -- i.e., corresponding to a full binary
+tree with $M = 2^D$ terminal nodes -- the complexity
 increases to $O(P^D N^D)$ \citep{papagelis2001bdt}. One conceivable compromise
@@
 less constrained than that from a 1-step-ahead search.
+
 Another class of algorithms is given by stochastic optimization methods that,

### 2011-11-07 pkg/evtree/inst/doc/evtree.Rnw
fixed typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@253 255f4249-2263-4417-b1ea-8498779e27a7
fixed a few typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@269 255f4249-2263-4417-b1ea-8498779e27a7

@@
 data("BBBClub", package = "evtree")
-cache <- TRUE
+cache <- FALSE
 @
@@
 arguments hold more generally, only binary tree models with some maximum number
-of splits $M_{\max}$ are considered. Both restrictions make the notation somewhat
+of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
 simpler while not really restricting the problem: (a)~Multiway splits are equivalent
@@
 
-The search for a valid split rule is conducted as in see Section~\ref{init}
+The search for a valid split rule is conducted as in Section~\ref{init}
 for a maximum of $P$~iterations. 
@@
 $s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
-where $X_{v_r}$ has less then $20$ unique values, the split point is change to
-to the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
+where $X_{v_r}$ has less then $20$ unique values, the split point is changed to
+the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
 $s_r$ is randomly shifted by a number of unique values that is not larger than
@@
 ${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
-chosen from the population for crossover. First, an internal nodes $r_1$ and $r_2$
+chosen from the population for crossover. First, two internal nodes $r_1$ and $r_2$
 are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
@@
 For 15 out of 17 datasets \code{evtree} shows a better predictive performance.
-The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower.
+The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squarred error of \code{ctree} is $5.6\%$ lower.
 However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models.

### 2011-11-11 pkg/evtree/inst/doc/evtree.Rnw
another one
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@270 255f4249-2263-4417-b1ea-8498779e27a7

@@
 are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
-Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_j$ rooted
+Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_i^j$ rooted
 by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.

### 2011-11-11 pkg/evtree/inst/doc/evtree.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@271 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
-(for ordered and numeric variables) or a a nonempty subset of $\{1, \dots, c\}$
+(for ordered and numeric variables) or a nonempty subset of $\{1, \dots, c\}$
 (for a categorical variable with $c$~levels), determining which observations are

### 2012-03-29 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex
almost finished the first revision of the paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@285 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[DIN, pagenumber=false, parskip=half,% 
               fromalign=left, fromphone=true,%  
               fromemail=true, fromurl=false, %
               fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}

\usepackage[latin1]{inputenc}

\usepackage{graphicx}
\usepackage{amsmath}

\setkomavar{fromname}{Thomas Grubinger\\ Department of Medical Statistics, Informatics and Health Economics \\ Innsbruck Medical University}
\setkomavar{fromaddress}{Sch\"opfstrasse 41/1, 6020 Innsbruck, Austria}
\setkomavar{backaddress}{Dept. of Med. Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
\setkomavar{fromphone}{+43 512 9003 70906}
\setkomavar{fromemail}{thomas.grubinger@i-med.ac.at}
\setkomavar{signature}{Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer}
\setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
\setkomavar{subject}[]{Author's response to the review of ``evtree: 
Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''}

\begin{document}

\begin{letter}{The Editor \\ Journal of Statistical Software}

\opening{Dear Editor}

we would like to thank the reviewer for the careful review and the constructive comments on our paper. We have revised our manuscript based on the reviewer's comments and respond on them below. 
\\ \\
\textbf{1. Overall Thoughts}

\textit{The paper did a good job of introducing a new tree-based algorithm overall. A lot of details pertaining to traditional tree models and the concept of evolutionary algorithms were covered, some of which could be clarified further. A clean-up in notation and wording would be greatly beneficial. The code appears easy-to-read and uses parameters consistent with other tree packages.}

We cleaned-up the notation and wording according to the reviewer's recommendations.
\\ \\

\textit{In the end, the authors gave honest comparisons of their methods to the traditional rpart and ctree methods, but it is unclear as to when and why evtree outperforms its competitors apart from the specific chessboard example given in Section 5.2.}

We added the following paragraph to the ``Conclusions'', as an explanation as to when and why evtree outperforms its competitors:

No obvious relationship between the data characteristics and evtree's 
relative performance could be observed. However, one reason is that for some datasets,
the recursive partitioning trees are already (nearly) optimal. For other datasets, 
 evtree's global search strategy can find more complex interactions, 
resulting in better predictive accuracy and/or sparser trees.
\\ \\
\textbf{2 Spelling/Grammatical Errors}

\textit{
\begin{enumerate}
\item (p.3) Above equation (2), ``predicition" should be ``prediction".
\item (p.7) First paragraph, midway, ``spit points" should be ``split points"
\item (p.8) Minor split rule mutation paragraph: ``less then" should be ``less than", and ``the split
point is change to to..." should be ``the split point is changed to".
\item (p.8) Crossover paragraph, last sentence: ``nods" should be ``nodes".
\item (p.8) Evaluation function first paragraph: should be ``The evaluation function represents the
requirements to which the population should adapt."
\item (p.14) Middle of first pararaph: ``bookclub" should be ``book club".
\end{enumerate}
}

We changed all spelling- and grammatical errors as suggested.
\\  \\
\textbf{3 Notation}

\textit{In Section 2, the authors introduce some notation for the tree setup, but the notation seems inconsistent in the paper. The authors write $\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in \{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where $r=1,\dots,M-1$. Further, in Section 2.2, the authors talk about $s_r$ and $v_r$ where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the indexing given
in equation (3). The authors should use a consistent notation for indexing the $v$ and $s$.}

We cleaned up the notation and made it consistent. A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$. Here, $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$). Now, in order to simplify the notation, the positions of the internal nodes are no longer made explicit. 
\\  \\
\textbf{4 Case Study}

\textit{
The authors have provided very minimal information on the dataset. (p.14) The authors write
``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers." The text does not explain which variable corresponds to ``join date", but based on Figure 3, one can indirectly infer that this variable is first. The authors have described te variable first as the number of months since the first purchase, but do not say that members join the book club upon making their first purchase.}

We added a Table, which shortly describes the variables of the Bookbinder's Book Club dataset. We assumed the join date to be equal to the data of the first purchase. To avoid confusions we changed the sentence:

``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers."

to:

``Thus, the evolutionary tree uses a different cutoff in art for book club members that bought there first book in the last year as opposed to older customers.".
\\ \\

\textit{The table comparing the misclassification and evaluation function values for the different methods identify the methods as evtree, rpart, ctree, rpart2, and ctree2. The paragraph which describes this table instead identifies the methods as ev, rp, ct, rp2, and ct2. It would be better to make sure the two are consistent.}

We changed the identifiers in the describing paragraph to evtree, rpart, ctree, rpart2, and ctree2. The notation is now consistent.
\\  \\
\textbf{5 Miscellaneous}

In addition to the above, I have the following suggestions.

\begin{enumerate}

\item \textit{(p.7) Section 3.1, ``the assignment of one category is flipped to the other terminal node" might be adjusted to ``one of the c categories is allocated to the other terminal node, to have the effect of ensuring both terminal nodes are nonempty".}

We changed the sentence according to the recommendations.

\textit{\item (p.7) Split operator paragraphs use r again in 2 different ways: ``internal node r" vs. ``internal node at position r". As with the remainder of the paper, r is the internal node, not the position.}

We changed the notation accordingly.

\textit{\item (p.7) Prune paragraph: We might append to the end of the sentence for clarity: ``...and no pruning occurs."}

We changed the sentence according to the recommendations.

\textit{\item (p.8) Minor split rule mutation paragraph: The description of the mutation rules is rather complex, with really 4 different cases to consider (nominal/$<$ 20 values, nominal/$\ge$ 20 values, continuous/$<$ 20 values, continuous/$\ge$ 20 values). It might be good to present this set of rules in a tabular format.}

The rules are now presented in tabular format and are now easier to read.

\textit{\item (p.8) Crossover paragraph, last sentence: ``these are omitted" might be better written as ``they are omitted" to be clear that ``these" refers to the invalid nodes.}

We changed the sentence according to the recommendations.

\textit{\item Evaluation function paragraph: ``A suitable evaluation function... minimizes the models' accuracy...and the models' complexity." It appears that ``minimizes" should be ``optimizes".}

We change the sentence to: ``A suitable evaluation function for classification and regression trees maximizes the models' accuracy on the training data, and minimizes the models' complexity."

\textit{\item (p.8) Classification paragraph: ``misclassification MC" should be ``misclassification rate MC". }

We changed the sentence according to the recommendations.

\textit{\item (p.9) Survivor selection paragraph, last sentence on the page: This sentence is worded awkwardly. Perhaps the authors meant ``In the case of a mutation operator, either the solution before modification, $\theta_i$, or after modification, $\theta_{i+1}$, is kept in memory." Without the commas, the sentence is unclear.}

We changed the sentence according to the recommendations.

\textit{\item (p.10) Third paragraph: Ambiguous ``they" in ``However, as they represent one of the best..., they are both...". Does ``they" refer to the parent and offspring trees? Does this mean that despite the $(1+1)$ rule, both the parent and offspring are both selected to the next generation? This seems unclear.}

Yes, ``they'' refer to the parent and the offspring. However, the sentence explains the strong selective pressure of a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$. With a $(1+1)$ strategy, the parent and the offspring can never be selected for the next round (as it is stated in the last sentence of the paragraph).

For better clarification, we replaced ``they'' with ``the parent and the offspring" and changed the sentence:

"Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in situations as follows."

with:

"Using a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$, a strong selective pressure can occur in situations as follows.".

\textit{\item (p.17) First paragraph: The authors talked about the demanding memory requirements, for which ``400 Mbit" seems very low. Perhaps the authors meant ``400 megabytes"?}

It should be ``400 MB''. We used the wrong abbreviation.
\end{enumerate}
\vspace{0.5cm}

Sincerely,

\vspace{0.75cm}
Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer

\end{letter}

\end{document}

### 2012-03-29 pkg/evtree/inst/doc/evtree.Rnw
almost finished the first revision of the paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@285 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \begin{equation}
-  \theta ~=~ (v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ), \label{eq:tree}
+  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
 \end{equation}
 %
-where the $n_r \in \{1,...,M_{\max}-1\}$ are the positions of the
-internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
+where $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
 \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
@@
 or the shape of trees. The goal is then to find that classification and regression tree
-which optimizes some tradeoff between predicition performance and complexity:
+which optimizes some tradeoff between prediction performance and complexity:
 %
@@
 partitioning algorithms only consider one split at a time. At each internal node
-$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
+$r \in \{1,...,M-1\}$, the split variable $v_r$ and the corresponding
 split point $s_r$ are selected to locally minimize the loss function.
@@
 \caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
-\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
+%%\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
 \end{figure}
@@
 or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
-probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is
+probability from the $u-1$ possible split points of $X_{v_1}$. If $X_{v_1}$ is
 nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
 be assigned to the left or the right daughter node. In cases where all $k$ are
-allocated to the same terminal node, the assignment of one category is flipped
-to the other terminal node. If this procedure results in a non-valid split rule,
-the two steps of random split variable selection and split point selection are
-repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the
-initialization is complete and each individual of the population of trees is of
-type ${\theta}_{1}= (v_1, s_1)$.
+allocated to the same terminal node, one of the $c$ categories is allocated to 
+the other terminal node, to have the effect of ensuring both terminal nodes are nonempty.
+If this procedure results in a non-valid split rule, the two steps of random split variable 
+selection and split point selection are repeated. With the definition of $r=1$ and the 
+selection of $v_1$ and $s_1$, the initialization is complete and each individual of
+the population of trees is of type ${\theta}_{1}= (v_1, s_1)$.
 
@@
 for a maximum of $P$~iterations. 
-In cases where no valid split rule can be assigned to the internal node at
-position~$r$, the search for a valid split rule is carried out on another
+In cases where no valid split rule can be assigned to internal node~$r$,
+the search for a valid split rule is carried out on another
 randomly selected terminal node. If, after 10~attempts, no valid split rule
@@
 
-\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two
+\textit{Prune} chooses a random internal node $r$, where $r > 1$, which has two
 terminal nodes as successors and prunes it into a terminal node. The tree's
@@
 \setminus (v_{r},s_{r})$. If ${\theta}_{i}$ only comprises one internal
-node, i.e., the root node, then ${\theta}_{i+1} = {\theta}_{i}$.
+node, i.e., the root node, then ${\theta}_{i+1} = {\theta}_{i}$ and no pruning occurs.
 
@@
 \textit{Minor split rule mutation} is similar to the \textit{major split rule
-mutation} operator, but it does not alter $v_r$ and only changes the split point
-$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
-$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
-where $X_{v_r}$ has less then $20$ unique values, the split point is changed to
-the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
-$s_r$ is randomly shifted by a number of unique values that is not larger than
-$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
-variable, with less than $20$ categories, one of the categories is randomly
-modified. Otherwise, at least one and at most $10\%$ of the variable's
-categories are changed. In cases where subsequent nodes become invalid, further
+mutation} operator. However, it does not alter $v_r$ and only changes the split point
+$s_r$ by a minor degree, which is defined by one of the following 4 cases:
+\begin{itemize}
+\item {$X_{v_r}$ is numerical or ordinal and has at least $20$ unique values:}
+The split point $s_r$ is randomly shifted by a non-zero number of unique values of $X_{v_r}$ 
+that is not larger than $10\%$ of the range of unique values.
+\item {$X_{v_r}$ is numerical or ordinal and has less than $20$ unique values:}
+The split point is changed to the next larger, or the next lower, unique value of $X_{v_r}$.
+\item {$X_{v_r}$ is nominal and has at least $20$ categories:}  at least one and at most $10\%$ of the variable's
+categories are changed.
+\item {$X_{v_r}$ is nominal and has less than $20$ categories:} One of the categories is randomly
+modified.
+\end{itemize}
+In cases where subsequent nodes become invalid, further
 split points are searched that preserve the tree's topology. After five
@@
 ${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
-If the crossover creates some invalid nods in either one of the new trees ${\theta}_{i+1}^1$
-or ${\theta}_{i+1}^2$, these are omitted.
+If the crossover creates some invalid nodes in either one of the new trees ${\theta}_{i+1}^1$
+or ${\theta}_{i+1}^2$, they are omitted.
 
@@
 
-The evaluation function represents the requirements the population should adapt
-to. In general, these requirements are formulated by Equation~\ref{eq:globally}.
-A suitable evaluation function for classification and regression trees minimizes
-the models' accuracy on the training data, and the models' complexity. This
+The evaluation function represents the requirements to which the population should adapt. 
+In general, these requirements are formulated by Equation~\ref{eq:globally}.
+A suitable evaluation function for classification and regression trees maximizes
+the models' accuracy on the training data, and minimizes the models' complexity. This
 subsection describes the currently implemented choices of evaluation functions
@@
 The quality of a classification tree is most commonly measured as a function of
-its misclassification MC and the complexity of a tree by a function
+its misclassification rate MC and the complexity of a tree by a function
 of the number of its terminal nodes $M$. \code{evtree} uses $2 N \cdot
@@
 competes with one offspring for a place in the population. In the case of a
-mutation operator either the solution before ${\theta}_i$ or after
-modification ${\theta}_{i+1}$ is kept in memory. In the case of the
+mutation operator, either the solution before modification, ${\theta}_i$, or after
+modification, ${\theta}_{i+1}$, is kept in memory. In the case of the
 crossover operator, the initial solutions of ${\theta}_i^1$ competes with
@@
 converge early in local optima. On the other hand, an overly weak selective
-pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy, a
-strong selective pressure can occur in situations as follows. Suppose the $b$-th
+pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy 
+with $\mu>1$ and $\lambda>1$, a strong selective pressure can occur in 
+situations as follows. Suppose the $b$-th
 tree of the population is one of the fittest trees in iteration $i$, and in
@@
 degree. Then very few instances are classified differently and the overall
-misclassification might not even change. However, as they represent one of the
-best solutions in iteration $i$, they are both selected for the subsequent
+misclassification might not even change. However, as the parent and the offspring
+represent one of the best solutions in iteration $i$, they are both selected for the subsequent
 population. This situation can occur frequently, especially when a fine-tuning
@@
 ``The Art History of Florence'' was sent to 20,000~customers, 1,806~of
-which bought the book. The dataset contains a subsample of 1,300~customers for
-building a predictive model of customer choice.
+which bought the book. The dataset contains a subsample of 1,300~customers with 10~exploratory variables (see Table~\ref{tab:BBBClubDescription}) for building a predictive model of customer choice.
+
+\begin{table}[t!]
+\centering
+\begin{tabular}{ll}
+\toprule	
+     \textbf{Variable} & Description\\
+\midrule	
+     \textbf{amount} & total amount of money spent at the BBB Club. \\
+     \textbf{art} & number of art books purchased. \\
+     \textbf{child} & number of children's books purchased. \\
+     \textbf{cook} & number of cookbooks purchased. \\
+     \textbf{diy} & number of do-it-yourself books purchased. \\    
+     \textbf{first} & number of months since the first purchase. \\   
+     \textbf{freq} & number of books purchased at the BBB Club. \\          
+     \textbf{gender} & factor indicating gender. \\
+     \textbf{last} & number of months since the last purchase. \\         
+     \textbf{youth} & number of youth books purchased. \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:BBBClubDescription} This table describes the predictor variables of the \textit{Bookbinder's Book Club} dataset.} 
+\end{table}
 
@@
 \centering
-\setkeys{Gin}{width=0.75\textwidth} %FIXME: 0.75 just to avoid \pagebreak, should be 0.7
+\setkeys{Gin}{width=0.75\textwidth} 
 <<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
@@
 the arguably most important variable in the first split while the evolutionary tree uses \code{first}
-in the first split and \code{art} in both second splits. Thus, the evolutionary tree
-uses a different cutoff in \code{art} for bookclub members that joined in the last year as opposed
-to older customers. While the former are predicted to be buyers if they previously bought
+in the first split and \code{art} in both second splits.
+Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that bought there
+first book in the last year as opposed to older customers.
+While the former are predicted to be buyers if they previously bought
 at least one art book, the latter are predicted to purchase the advertised art book only if they
@@
 %
-Not surprisingly the evolutionary tree \code{ev} outperforms the depth-restricted
-trees \code{rp2} and \code{ct2}, both in terms of misclassification and the penalized
-evaluation function. However, it is interesting to see that \code{ev} performs even better
-than the unrestricted conditional inference tree \code{ct} and is comparable in performance
-to the unrestricted CART tree \code{rp}. Hence, the practitioner may choose the evolutionary
-tree \code{ev} as it is the easiest to communicate.
+Not surprisingly the evolutionary tree \code{evtree} outperforms the depth-restricted
+trees \code{rpart2} and \code{ctree2}, both in terms of misclassification and the penalized
+evaluation function. However, it is interesting to see that \code{evtree} performs even better
+than the unrestricted conditional inference tree \code{ctree} and is comparable in performance
+to the unrestricted CART tree \code{rpart}. Hence, the practitioner may choose the evolutionary
+tree \code{evtree} as it is the easiest to communicate.
 
@@
 While the smallest of the analyzed datasets, \textit{Glass identification}, only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes.
-The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit.
+The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 MB.
 The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.
@@
 
-\pagebreak %%FIXME
+%%\pagebreak %%FIXME
 
@@
 
+No obvious relationship between the data characteristics and \code{evtree}'s 
+relative performance could be observed. However, one reason is that for some datasets,
+the recursive partitioning trees are already (nearly) optimal. For other datasets, 
+ \code{evtree}'s global search strategy can find more complex interactions, 
+resulting in better predictive accuracy and/or sparser trees.
+
 However, the goal of \code{evtree} is not to replace the well-established algorithms like

### 2012-03-30 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex
continued revision; reply is added to new folder 'Revisions'iin folder 'Submission_jss'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@286 255f4249-2263-4417-b1ea-8498779e27a7

@@
 No obvious relationship between the data characteristics and evtree's
-relative performance could be observed. However, one reason is that for some datasets,
-the recursive partitioning trees are already (nearly) optimal. For other datasets, 
- evtree's global search strategy can find more complex interactions, 
-resulting in better predictive accuracy and/or sparser trees.
+relative performance could be observed. Probably, one explanation is that for some datasets,
+the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
+improvements. For other datasets, evtree's global search strategy can find more
+complex interactions, resulting in better predictive accuracy and/or sparser trees.
 \\ \\
@@
 We cleaned up the notation and made it consistent.
-A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$.
-Here, $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
+A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$, were $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
 Now, in order to simplify the notation, the positions of the internal nodes are no longer made explicit.
@@
 
-\textit{
-The authors have provided very minimal information on the dataset. (p.14) The authors write
+\textit{The authors have provided very minimal information on the dataset. (p.14) The authors write
 ``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers." The text does not explain which variable corresponds to ``join date", but based on Figure 3, one can indirectly infer that this variable is first.
@@
 
-We added a Table, which shortly describes the variables of the Bookbinder's Book Club dataset.
+We added a table (Table 2), which shortly describes the variables of the Bookbinder's Book Club dataset.
 We assumed the join date to be equal to the data of the first purchase.
@@
 
-The rules are now presented in tabular format and are now easier to read.
+The rules are now presented in an easy to read tabular format.
 

### 2012-03-30 pkg/evtree/inst/doc/evtree.Rnw
continued revision; reply is added to new folder 'Revisions'iin folder 'Submission_jss'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@286 255f4249-2263-4417-b1ea-8498779e27a7

@@
 The main challenge in growing globally optimal trees is that the search
-space is typically huge rendering full-grid searches computationally infeasible.
+space is typically huge, rendering full-grid searches computationally infeasible.
 One possibility to solve this problem is to use stochastic
@@
 simpler while not really restricting the problem: (a)~Multiway splits are equivalent
-to a sequence of binary splits in predictions and number of resulting subsamples.
-(b)~The maximal size of the tree is always limited by the number of observations
+to a sequence of binary splits in predictions and number of resulting subsamples;
+(b)~the maximal size of the tree is always limited by the number of observations
 in the learning sample.
@@
 %
-where $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
+where $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
 \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
 Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
-(for ordered and numeric variables) or a nonempty subset of $\{1, \dots, c\}$
+(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
 (for a categorical variable with $c$~levels), determining which observations are
@@
 non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
-more complex models in the tree selection selection process.
+more complex models in the tree selection process.
 Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.
@@
 jointly over $(v_r, s_r)$, as is employed in CART or C4.5. So-called \emph{unbiased}
-recursive partitioning techniques modify this search by first seleting the variable $v_r$ using
+recursive partitioning techniques modify this search by first selecting the variable $v_r$ using
 statistical significance tests and subsequently selecting the optimal split $s_r$ for that
@@
 but only an approximation thereof. Besides evolutionary algorithms
-\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
+\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca}, and
 \textit{simulated annealing} \citep{sutton1992ict} were used successfully to
@@
 If this procedure results in a non-valid split rule, the two steps of random split variable 
-selection and split point selection are repeated. With the definition of $r=1$ and the 
+selection and split point selection are repeated. With the definition of $r=1$ (the root node) and the 
 selection of $v_1$ and $s_1$, the initialization is complete and each individual of
@@
 The split point is changed to the next larger, or the next lower, unique value of $X_{v_r}$.
-\item {$X_{v_r}$ is nominal and has at least $20$ categories:}  at least one and at most $10\%$ of the variable's
+\item {$X_{v_r}$ is nominal and has at least $20$ categories:} At least one and at most $10\%$ of the variable's
 categories are changed.
@@
 constrain the solution to a minimum number of observations in each terminal
-node, a minimum number of observation in each internal node and a maximum tree
+node, a minimum number of observations in each internal node, and a maximum tree
 depth. Note that the memory requirements increase by the square of the maximum
 tree depth. Parameter \code{alpha} regulates the complexity parameter
-$\alpha$ in Equation~\ref{eq:classification} and~\ref{eq:regression}, respectively.
+$\alpha$ in Equations~\ref{eq:classification} and~\ref{eq:regression}, respectively.
 \code{niterations} and \code{ntrees} specify the maximum number of iterations
@@
 Additionally, a significance level of $1\%$ is employed in the construction of
-conditional inference trees which is more appropriate than the default 5\%~level for
+conditional inference trees, which is more appropriate than the default 5\%~level for
 $1,300$~observations. To provide uniform visualizations and predictions of the
-fitted models, `\code{party}' objects are used to represent all trees. For, `\code{rpart}'
-trees \pkg{partykit} provides a suitable \code{as.party()} method while a
+fitted models, `\code{party}' objects are used to represent all trees. For `\code{rpart}'
+trees, \pkg{partykit} provides a suitable \code{as.party()} method while a
 reimplementation of \code{ctree()} is provided in \pkg{partykit} (as opposed to
@@
   The variables used for splitting are the number of art books purchased previously (\code{art}), the number
-  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club
-  (\code{freq}) and the customer's \code{gender}.}
+  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club (\code{freq}), and the customer's \code{gender}.}
 \end{figure}
@@
    The target variable is the customer's choice of buying the book. The variables used for splitting are
-   the number of art books purchased previously (\code{art}) and the number of months since the first purchase
-    \code{first}.}
+   the number of art books purchased previously (\code{art}), and the number of months since the first purchase
+    (\code{first}).}
 \end{figure}
@@
 of months since the first purchase (\code{first}), the frequency of previous purchases
-(\code{freq}) and the customer's \code{gender}. Interestingly, though, the forward-search trees select
+(\code{freq}), and the customer's \code{gender}. Interestingly, though, the forward-search trees select
 the arguably most important variable in the first split while the evolutionary tree uses \code{first}
@@
 first book in the last year as opposed to older customers.
-While the former are predicted to be buyers if they previously bought
+While the former are predicted to be buyers if they had previously bought
 at least one art book, the latter are predicted to purchase the advertised art book only if they
-previously bought at least two other art books. Certainly, this classification is easy to understand
+had previously bought at least two other art books. Certainly, this classification is easy to understand
 and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.
@@
 %
-In this case, \code{evtree} classifies less
+In this case, \code{evtree} classifies fewer
 customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$). However,
@@
 All models are constrained to a minimum number of 7~observations per terminal node,
-20~observations per internal node and a maximum tree depth of~9. Apart from
-that, the default settings of the algorithms are used. For assessment
+20~observations per internal node, and a maximum tree depth of~9. Apart from
+that, the default settings of the algorithms are used. For the assessment
 of significant differences in predictive accuracy and complexity, respectively,
@@
 As missing values are currently not supported by \code{evtree} (e.g., by surrogate splits),
-the 16~missing values in the \textit{Breast Cancer Database} -- the only dataset in the
+the 16~missing values in the \textit{Breast cancer database} -- the only dataset in the
 study with missing values -- were removed before analysis.
@@
 It can be observed that on $12$ out of $17$ datasets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy.
-Only on the \textit{Contraceptive Method} dataset \code{evtree} performs slightly worse.
+Only on the \textit{Contraceptive method} dataset does \code{evtree} perform slightly worse.
 In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ datasets.
@@
 \caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{evtree} vs.\ \code{ctree} (lower panels).
-  Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean-squared error.
+  Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean squared error.
   The complexity (right panels) is compared by the relative difference of the number of terminal nodes.}
@@
 Another important issue to be considered is the random nature of evolutionary algorithms.
-For larger datasets, frequently, considerable different solutions exist that yield a similar or even the same evaluation function value.
+For larger datasets, frequently, considerably different solutions exist that yield a similar or even the same evaluation function value.
 Therefore, subsequent runs of \code{evtree} can result in very different tree structures.
@@
 The results are summarized in Table~\ref{tab:resultsChessboard}.
-It can be seen that, in the absence of noise, \code{rpart} models on average classify $69.1\%$ of the data points correctly and had $16.6$ terminal nodes.
-An average \code{ctree} model only has $1.1$ terminal nodes and a classification accuracy of $49.9\%$.
+It can be seen that, in the absence of noise, \code{rpart} models on average classify $69.1\%$ of the data points correctly and has $16.6$ terminal nodes.
+An average \code{ctree} model has only $1.1$ terminal nodes and a classification accuracy of $49.9\%$.
 In contrast, \code{evtree} classifies $93.2\%$ of the instances correctly and requires $14.4$ terminal nodes.
@@
 No obvious relationship between the data characteristics and \code{evtree}'s
-relative performance could be observed. However, one reason is that for some datasets,
-the recursive partitioning trees are already (nearly) optimal. For other datasets, 
- \code{evtree}'s global search strategy can find more complex interactions, 
-resulting in better predictive accuracy and/or sparser trees.
+relative performance could be observed. Probably, one explanation is that for some datasets,
+the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
+improvements. For other datasets, \code{evtree}'s global search strategy can find more
+complex interactions, resulting in better predictive accuracy and/or sparser trees.
 

### 2012-03-30 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex
continued revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@287 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \setkomavar{subject}[]{Author's response to the review of ``evtree: 
-Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''}
+Evolutionary Learning of Globally Optimal Classification and Regression Trees in R'' (JSS 817)}
 
@@
 
-We added the following paragraph to the ``Conclusions'', as an explanation as to when and why evtree outperforms its competitors:
+We added the following paragraph to the ``5.1 Benchmark and real-world problemss'', as an explanation as to when and why evtree outperforms its competitors:
 

### 2012-04-03 pkg/evtree/inst/doc/evtree.Rnw
continued revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@287 255f4249-2263-4417-b1ea-8498779e27a7
small improvements in notation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@288 255f4249-2263-4417-b1ea-8498779e27a7

@@
 %
-where $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
-\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
+where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
+associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
 Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
@@
 @
-\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
+\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}.
    The target variable is the customer's choice of buying the book. The variables used for splitting are
@@
 
+No obvious relationship between the data characteristics and \code{evtree}'s relative
+performance could be observed. Probably, one explanation is that for some datasets,
+the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
+improvements. For other datasets, \code{evtree}'s global search strategy can find more
+complex interactions, resulting in better predictive accuracy and/or sparser trees.
+
 Disadvantages of the \code{evtree} algorithm are computation time and memory requirements.
@@
 
-No obvious relationship between the data characteristics and \code{evtree}'s
-relative performance could be observed. Probably, one explanation is that for some datasets,
-the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
-improvements. For other datasets, \code{evtree}'s global search strategy can find more
-complex interactions, resulting in better predictive accuracy and/or sparser trees.
-
 However, the goal of \code{evtree} is not to replace the well-established algorithms like

### 2012-04-06 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex
Some touch-ups, rephrasings, and modifications. In partiular, the letter contains some more introductory remarks/comments and streamlines some of the straightforward changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@289 255f4249-2263-4417-b1ea-8498779e27a7

--- 2012-03-30 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex
@@
                fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}
-
 \usepackage[latin1]{inputenc}
-
 \usepackage{graphicx}
@@
 \setkomavar{fromaddress}{Sch\"opfstrasse 41/1, 6020 Innsbruck, Austria}
-\setkomavar{backaddress}{Dept. of Med. Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
+\setkomavar{backaddress}{Dept.~of Med.~Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
 \setkomavar{fromphone}{+43 512 9003 70906}
-\setkomavar{fromemail}{thomas.grubinger@i-med.ac.at}
-\setkomavar{signature}{Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer}
+\setkomavar{fromemail}{Thomas.Grubinger@i-med.ac.at}
+\setkomavar{signature}{Thomas Grubinger, Achim Zeileis, Karl-Peter Pfeiffer}
 \setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
-\setkomavar{subject}[]{Author's response to the review of ``evtree: 
-Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''}
+\setkomavar{subject}[]{Revision of  JSS 817}
 
@@
 
-\begin{letter}{The Editor \\ Journal of Statistical Software}
+\begin{letter}{
+  Jan de Leeuw\\
+  Department of Statistics\\
+  University of California, Los Angeles\\
+  United States of America
+}
 
-\opening{Dear Editor}
+\opening{Dear Prof.~de Leeuw,}
 
-we would like to thank the reviewer for the careful review and the constructive comments on our paper. We have revised our manuscript based on the reviewer's comments and respond on them below. 
-\\ \\
+please find attached the revised version of our manuscript
+``evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''
+(JSS 817). We would like to thank the reviewer for the careful review and the constructive
+comments. We have incorporated all of the reviewer's recommendations, in particular by
+cleaning up the notation and providing some more discussion of the illustrations. We
+hope to have addressed all concerns in our revision and provide a detailed
+point-to-point reply below.
+
+Sincerely,\\[-2cm]
+\closing{}
+
+\end{letter}
+
+\newpage
+
+\textbf{\large Point-to-Point Reply}
+
+\bigskip
+
 \textbf{1. Overall Thoughts}
 
-\textit{The paper did a good job of introducing a new tree-based algorithm overall.
-A lot of details pertaining to traditional tree models and the concept of evolutionary algorithms were covered, some of which could be clarified further.
-A clean-up in notation and wording would be greatly beneficial.
-The code appears easy-to-read and uses parameters consistent with other tree packages.}
+\textit{The paper did a good job of introducing a new tree-based algorithm
+overall. A lot of details pertaining to traditional tree models and the concept
+of evolutionary algorithms were covered, some of which could be clarified
+further. A clean-up in notation and wording would be greatly beneficial. The
+code appears easy-to-read and uses parameters consistent with other tree
+packages.}
 
-We cleaned-up the notation and wording according to the reviewer's recommendations.
-\\ \\
+Thanks for pointing these issues out. In fact, providing a unifying but still
+simple and easily accessible notation for presenting the framework was one
+important objective of the manuscript. While preparing the manuscript, we had
+revised the notation a couple of times and apparently did not clean it up
+carefully enough. Hence, we appreciate the thorough review and followed
+all of its recommendations in this revision. Details of the revision are
+provided below.
 
-\textit{In the end, the authors gave honest comparisons of their methods to the traditional rpart and ctree methods, but it is unclear as to when and why evtree outperforms its competitors apart from the specific chessboard example given in Section 5.2.}
+\medskip
 
-We added the following paragraph to the ``Conclusions'', as an explanation as to when and why evtree outperforms its competitors:
+\textit{In the end, the authors gave honest comparisons of their methods to the
+traditional rpart and ctree methods, but it is unclear as to when and why evtree
+outperforms its competitors apart from the specific chessboard example given in
+Section~5.2.}
 
-No obvious relationship between the data characteristics and evtree's
-relative performance could be observed. Probably, one explanation is that for some datasets,
-the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
-improvements. For other datasets, evtree's global search strategy can find more
-complex interactions, resulting in better predictive accuracy and/or sparser trees.
-\\ \\
-\textbf{2 Spelling/Grammatical Errors}
+As the forward-search algorithms just optimze over a subset of conceivable
+trees in each step compared to global-search algorithms, the latter are in principle
+able to outperform the former when the optimal solution contains more complex
+interaction patterns with low main effects. This is now discussed explicitly
+in the following new paragraph in Section~5.1:
 
-\textit{
-\begin{enumerate}
-\item (p.3) Above equation (2), ``predicition" should be ``prediction".
-\item (p.7) First paragraph, midway, ``spit points" should be ``split points"
-\item (p.8) Minor split rule mutation paragraph: ``less then" should be ``less than", and ``the split
-point is change to to..." should be ``the split point is changed to".
-\item (p.8) Crossover paragraph, last sentence: ``nods" should be ``nodes".
-\item (p.8) Evaluation function first paragraph: should be ``The evaluation function represents the
-requirements to which the population should adapt."
-\item (p.14) Middle of first pararaph: ``bookclub" should be ``book club".
-\end{enumerate}
-}
+``From these results, it is not obvious which characteristics drive \texttt{evtree}'s
+relative performance. Presumably, for some datasets the forward-search algorithms already
+yield trees that are close to optimal, thus leaving little room for further
+improvements. In contrast, for other datasets with more complex interaction patterns
+(and possibly low main effects) \texttt{evtree}'s global-search strategy is probably
+able to provide better predictive accuracy and/or sparser trees.''
 
-We changed all spelling- and grammatical errors as suggested.
-\\  \\
-\textbf{3 Notation}
+\bigskip
 
-\textit{In Section 2, the authors introduce some notation for the tree setup, but the notation seems inconsistent in the paper.
-The authors write $\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in \{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where $r=1,\dots,M-1$.
-Further, in Section 2.2, the authors talk about $s_r$ and $v_r$ where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the indexing given
-in equation (3). The authors should use a consistent notation for indexing the $v$ and $s$.}
+\textbf{2. Spelling/Grammatical Errors}
 
-We cleaned up the notation and made it consistent.
-A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$, were $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
-Now, in order to simplify the notation, the positions of the internal nodes are no longer made explicit.
-\\  \\
-\textbf{4 Case Study}
+Thanks for catching these. We have corrected all of them and ran again a spell
+checker on the manuscript.
 
-\textit{The authors have provided very minimal information on the dataset. (p.14) The authors write
-``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers." The text does not explain which variable corresponds to ``join date", but based on Figure 3, one can indirectly infer that this variable is first.
-The authors have described te variable first as the number of months since the first purchase, but do not say that members join the book club upon making their first purchase.}
+\bigskip
 
-We added a table (Table 2), which shortly describes the variables of the Bookbinder's Book Club dataset.
-We assumed the join date to be equal to the data of the first purchase.
-To avoid confusions we changed the sentence:
+\textbf{3. Notation}
 
-``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers."
+\textit{In Section 2, the authors introduce some notation for the tree setup,
+but the notation seems inconsistent in the paper. The authors write
+$\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in
+\{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where
+$r=1,\dots,M-1$. Further, in Section 2.2, the authors talk about $s_r$ and $v_r$
+where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the
+indexing given in equation (3). The authors should use a consistent notation for
+indexing the $v$ and $s$.}
 
-to:
+We cleaned up the notation and a binary tree is now consistently denoted
+as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$, 
+where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
+associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
+Thus, in order to keep the notation as simple as possible, the internal nodes of the
+tree are just labeled with their number $r \in \{1,...,M-1\}$.
 
-``Thus, the evolutionary tree uses a different cutoff in art for book club members that bought there first book in the last year as opposed to older customers.".
-\\ \\
+\bigskip
 
-\textit{The table comparing the misclassification and evaluation function values for the different methods identify the methods as evtree, rpart, ctree, rpart2, and ctree2.
-The paragraph which describes this table instead identifies the methods as ev, rp, ct, rp2, and ct2.
-It would be better to make sure the two are consistent.}
+\textbf{4. Case Study}
 
-We changed the identifiers in the describing paragraph to evtree, rpart, ctree, rpart2, and ctree2. The notation is now consistent.
-\\  \\
-\textbf{5 Miscellaneous}
+\textit{The authors have provided very minimal information on the dataset.
+(p.~14) The authors write ``Thus, the evolutionary tree uses a different cutoff
+in art for book club members that joined in the last year as opposed to older
+customers." The text does not explain which variable corresponds to ``join
+date", but based on Figure 3, one can indirectly infer that this variable is
+first. The authors have described te variable first as the number of months
+since the first purchase, but do not say that members join the book club upon
+making their first purchase.}
 
-In addition to the above, I have the following suggestions.
+We added a table (Table 2), which describes the variables of the
+Bookbinder's Book Club dataset. Also, we avoid the formulation ``join date''
+in the text and use ``first purchase'' instead.
 
-\begin{enumerate}
+\medskip
 
-\item \textit{(p.7) Section 3.1, ``the assignment of one category is flipped to the other terminal node" might be adjusted to ``one of the c categories is allocated to the other terminal node, to have the effect of ensuring both terminal nodes are nonempty".}
+\textit{The table comparing the misclassification and evaluation function values
+for the different methods identify the methods as evtree, rpart, ctree, rpart2,
+and ctree2. The paragraph which describes this table instead identifies the
+methods as ev, rp, ct, rp2, and ct2. It would be better to make sure the two are
+consistent.}
 
-We changed the sentence according to the recommendations.
+We changed the identifiers in the describing paragraph to also use \texttt{evtree},
+\texttt{rpart}, \texttt{ctree}, \texttt{rpart2}, and \texttt{ctree2} instead of the
+object names.
 
-\textit{\item (p.7) Split operator paragraphs use r again in 2 different ways: ``internal node r" vs.
-``internal node at position r".
-As with the remainder of the paper, r is the internal node, not the position.}
+\bigskip
 
-We changed the notation accordingly.
+\textbf{5. Miscellaneous}
 
-\textit{\item (p.7) Prune paragraph: We might append to the end of the sentence for clarity: ``...and no pruning occurs."}
+Thanks for pointing all of these out. All of the reviewer's recommendations
+have been incorporated into the revised version of the manuscript. The only
+point that needs additional explanation is the following:
 
-We changed the sentence according to the recommendations.
+\medskip
 
-\textit{\item (p.8) Minor split rule mutation paragraph: The description of the mutation rules is rather complex, with really 4 different cases to consider (nominal/$<$ 20 values, nominal/$\ge$ 20 values, continuous/$<$ 20 values, continuous/$\ge$ 20 values).
-It might be good to present this set of rules in a tabular format.}
+\textit{(p.~10) Third paragraph: Ambiguous ``they" in ``However, as they
+represent one of the best..., they are both...". Does ``they" refer to the
+parent and offspring trees? Does this mean that despite the $(1+1)$ rule, both
+the parent and offspring are both selected to the next generation? This seems
+unclear.}
 
-The rules are now presented in an easy to read tabular format.
+Yes, ``they'' refer to the parent and the offspring. However, the sentence
+explains the strong selective pressure of a $(\mu + \lambda)$ strategy with
+$\mu>1$ and $\lambda>1$. With a $(1+1)$ strategy, the parent and the offspring
+can never be selected for the next round (as it is stated in the last sentence
+of the paragraph).
 
-\textit{\item (p.8) Crossover paragraph, last sentence: ``these are omitted" might be better written as ``they are omitted" to be clear that ``these" refers to the invalid nodes.}
+For better clarification, we replaced ``they'' with ``the parent and the
+offspring" and changed the sentence:
 
-We changed the sentence according to the recommendations.
-
-\textit{\item Evaluation function paragraph: ``A suitable evaluation function... minimizes the models' accuracy...and the models' complexity." It appears that ``minimizes" should be ``optimizes".}
-
-We change the sentence to: ``A suitable evaluation function for classification and regression trees maximizes the models' accuracy on the training data, and minimizes the models' complexity."
-
-\textit{\item (p.8) Classification paragraph: ``misclassification MC" should be ``misclassification rate MC". }
-
-We changed the sentence according to the recommendations.
-
-\textit{\item (p.9) Survivor selection paragraph, last sentence on the page: This sentence is worded awkwardly.
-Perhaps the authors meant ``In the case of a mutation operator, either the solution before modification, $\theta_i$, or after modification, $\theta_{i+1}$, is kept in memory." Without the commas, the sentence is unclear.}
-
-We changed the sentence according to the recommendations.
-
-\textit{\item (p.10) Third paragraph: Ambiguous ``they" in ``However, as they represent one of the best..., they are both...".
-Does ``they" refer to the parent and offspring trees?
-Does this mean that despite the $(1+1)$ rule, both the parent and offspring are both selected to the next generation?
-This seems unclear.}
-
-Yes, ``they'' refer to the parent and the offspring.
-However, the sentence explains the strong selective pressure of a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$.
-With a $(1+1)$ strategy, the parent and the offspring can never be selected for the next round (as it is stated in the last sentence of the paragraph).
-
-For better clarification, we replaced ``they'' with ``the parent and the offspring" and changed the sentence:
-
-"Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in situations as follows."
+``Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in
+situations as follows."
 
@@
 
-"Using a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$, a strong selective pressure can occur in situations as follows.".
-
-\textit{\item (p.17) First paragraph: The authors talked about the demanding memory requirements, for which ``400 Mbit" seems very low. Perhaps the authors meant ``400 megabytes"?}
-
-It should be ``400 MB''. We used the wrong abbreviation.
-\end{enumerate}
-\vspace{0.5cm}
-
-Sincerely,
-
-\vspace{0.75cm}
-Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer
-
-\end{letter}
+``Using a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$, a strong
+selective pressure can occur in situations as follows.".
 

### 2012-04-06 pkg/evtree/inst/doc/evtree.Rnw
Some touch-ups, rephrasings, and modifications. In partiular, the letter contains some more introductory remarks/comments and streamlines some of the straightforward changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@289 255f4249-2263-4417-b1ea-8498779e27a7

@@
 data("BBBClub", package = "evtree")
-cache <- FALSE
+cache <- TRUE
 @
@@
 operators can be applied to modify the tree structure (e.g., number of splits, splitting
-variables, and corresponding splitpoints etc.) in order to optimize a
+variables, and corresponding split points etc.) in order to optimize a
 fitness functions such as the misclassificaion or error rate penalized by
@@
 To show why forward-search recursive partitioning algorithms typically
-lead to globablly suboptimal solutions, their parameter spaces and optimization
+lead to globally suboptimal solutions, their parameter spaces and optimization
 problems are presented and contrasted in a unified notation. Although all
@@
 $\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
-by exluding elements $\theta$ resulting in too small subsamples etc.).
+by excluding elements $\theta$ resulting in too small subsamples etc.).
 
@@
 \caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
-%%\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
 \end{figure}
@@
 or ${\theta}_{i+1}^2$, they are omitted.
+
+\pagebreak %% FIXME
 
@@
 in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
-If set to \code{-1L}, the seed is initializied by the system time.
+If set to \code{-1L}, the seed is initialized by the system time.
 
@@
 \toprule	
-     \textbf{Variable} & Description\\
+     Variable & Description\\
 \midrule	
-     \textbf{amount} & total amount of money spent at the BBB Club. \\
-     \textbf{art} & number of art books purchased. \\
-     \textbf{child} & number of children's books purchased. \\
-     \textbf{cook} & number of cookbooks purchased. \\
-     \textbf{diy} & number of do-it-yourself books purchased. \\    
-     \textbf{first} & number of months since the first purchase. \\   
-     \textbf{freq} & number of books purchased at the BBB Club. \\          
-     \textbf{gender} & factor indicating gender. \\
-     \textbf{last} & number of months since the last purchase. \\         
-     \textbf{youth} & number of youth books purchased. \\
+     \code{choice} & Did the customer buy the advertised book? \\
+     \code{amount} & Total amount of money spent at the book Club. \\
+     \code{art}    & Number of art books purchased. \\
+     \code{child}  & Number of children's books purchased. \\
+     \code{cook}   & Number of cookbooks purchased. \\
+     \code{diy}    & Number of do-it-yourself books purchased. \\    
+     \code{first}  & Number of months since the first purchase. \\   
+     \code{freq}   & Number of books purchased at the book Club. \\          
+     \code{gender} & Factor indicating gender. \\
+     \code{last}   & Number of months since the last purchase. \\         
+     \code{youth}  & Number of youth books purchased. \\
 \bottomrule
 \end{tabular}
-\caption{\label{tab:BBBClubDescription} This table describes the predictor variables of the \textit{Bookbinder's Book Club} dataset.} 
+\caption{\label{tab:BBBClubDescription} Variables of the Bookbinder's Book Club data.} 
 \end{table}
@@
 in the first split and \code{art} in both second splits.
-Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that bought there
-first book in the last year as opposed to older customers.
+Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that made their
+first purchase during the last year as opposed to older customers.
 While the former are predicted to be buyers if they had previously bought
@@
 For 15 out of 17 datasets \code{evtree} shows a better predictive performance.
-The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squarred error of \code{ctree} is $5.6\%$ lower.
+The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squared error of \code{ctree} is $5.6\%$ lower.
 However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models.
@@
 
-No obvious relationship between the data characteristics and \code{evtree}'s relative
-performance could be observed. Probably, one explanation is that for some datasets,
-the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
-improvements. For other datasets, \code{evtree}'s global search strategy can find more
-complex interactions, resulting in better predictive accuracy and/or sparser trees.
+From these results, it is not obvious which characteristics drive \code{evtree}'s
+relative performance. Presumably, for some datasets the forward-search algorithms already
+yield trees that are close to optimal, thus leaving little room for further
+improvements. In contrast, for other datasets with more complex interaction patterns
+(and possibly low main effects) \code{evtree}'s global-search strategy is probably
+able to provide better predictive accuracy and/or sparser trees.
 
@@
 However, if the primary goal is to interpret relationships in the data, based on the selected splits, the random nature of the algorithm has to be considered.
-
-%%\pagebreak %%FIXME
 

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_J48.R
moved plot functionality from benchmark_evtree.R and benchmark_J48.R to benchmark_plots.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@311 255f4249-2263-4417-b1ea-8498779e27a7

@@ -294,113 +294,3 @@ save(rchessboard44_52, file="results_j48/chessboard44_5_j48.RData")
 rchessboard44_102 <- benchchessboard44_j48(noise = 10, seed = 1000, nrealizations = 250)
 apply(apply(rchessboard44_102, 2, as.numeric),2,mean)
 save(rchessboard44_102, file="results_j48/chessboard44_10_j48.RData")
-
-
-###################################################
-### Benchmark plots
-###################################################
-
-### under construction 
-
-## load results
-for(i in Sys.glob("results/*.RData")) load(i)
-for(i in Sys.glob("results_j48/*.RData")) load(i)
-
-## preprocess for reference evtree
-preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
-    if(isclassification) d[, 1:3] <- 1 - d[ ,1:3]	
-    
-	# calculate total number of nodes of evtree trees
-	d[,4] <- d[,4]*2-1
-    d <- as.data.frame(d)
-    colnames(d) <- c("evtree", "J48", "J48_up","evtree", "J48", "J48_up")
-    for(i in 3:1) d[, i] <- d[, i] / d[, 1] * 100
-    for(i in 6:4) d[, i] <- d[, i] / d[, 4] * 100
-    x <- d[, 1:3]
-    y <- d[, 4:6]
-    rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
-    names(rval)[2] <- "accuracy"
-    rval$complexity <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")[,2]
-    rval$alg <- factor(rval$alg, levels = c("evtree", "J48", "J48_up"))
-    rval$ds <- dname
-    rval
-}
-
-## collect results for all datasets
-r <- rbind(
-preprocess(d = cbind(rglass[,1], rglass2[,c(1,3)], rglass[,4], rglass2[,c(2,4)]), dname = "Glass identification", isclassification = TRUE),
[…]

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_evtree.R

@@ -359,110 +359,3 @@ save(rchessboard44_5, file="results/chessboard44_5.RData")
 rchessboard44_10 <- benchchessboard44(noise = 10, seed = 1000, nrealizations = 250)
 apply(apply(rchessboard44_10, 2, as.numeric),2,mean)
 save(rchessboard44_10, file="results/chessboard44_10.RData")
-
-
-###################################################
-### Benchmark plots
-###################################################
-
-## load results
-for(i in Sys.glob("results/*.RData")) load(i)
-
-## preprocess for reference evtree
-preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
-    if(isclassification) d[, 1:3] <- 1 - d[ ,1:3]
-    d <- as.data.frame(d)
-    colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")
-    for(i in 3:1) d[, i] <- d[, i] / d[, 1] * 100
-    for(i in 6:4) d[, i] <- d[, i] / d[, 4] * 100
-    x <- d[, 1:3]
-    y <- d[, 4:6]
-    rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
-    names(rval)[2] <- "accuracy"
-    rval$complexity <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")[,2]
-    rval$alg <- factor(rval$alg, levels = c("evtree", "ctree", "rpart"))
-    rval$ds <- dname
-    rval
-}
-
-## collect results for all datasets
-r <- rbind(
-  preprocess(d = rglass, dname = "Glass identification", isclassification = TRUE),
-  preprocess(d = rheart, dname = "Statlog heart", isclassification = TRUE),
-  preprocess(d = rionosphere, dname = "Ionosphere", isclassification = TRUE),
-  preprocess(d = rmusk, dname = "Musk", isclassification = TRUE),
-  preprocess(d = rbreastcancer, dname = "Breast cancer database", isclassification = TRUE),
-  preprocess(d = rpima, dname = "Pima Indians diabetes", isclassification = TRUE),
[…]

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_evtree_parameters.R

@@ -379,11 +379,8 @@ library("plotrix")
 library("lme4")
 library("multcomp")
 
-load("results_parameter/heart_parameter.RData")
-load("results_parameter/credit_parameter.RData")
-load("results_parameter/spam_parameter.RData")
-load("results_parameter/chessboard44_5_parameter.RData")
-
+## load results
+for(i in Sys.glob("results_parameter/*.RData")) load(i)
 
 panel.mean <- function(x,y,...){
 	x <- as.numeric(x)
@@ -439,7 +436,7 @@ r2 <- sort_op(r2)
 pdf(file= "~/Desktop/evtree_op.pdf", width=10.3, height=20.45)
 b1 <- bwplot (value ~ factor(operatorprob)| nIter+ds , data= as.data.frame(r2), 
 horizontal = FALSE,  
-ylab= list("Accuracy [%]", cex=1.1),
+ylab= list("Accuracy (%)", cex=1.1),
 pch= '|',
 layout = c(3,4),
 ylim=as.data.frame(matrix(c(
@@ -448,13 +445,10 @@ rep(c(85,94),3),
 rep(c(60,80),3),
 rep(c(58,90),3)
 ), nrow=2)),
-#scales= list(x= list(rot=60), alternating = F), 
 scales= list(x= list(rot=60), y=list(relation="free"), alternating = F), 
-#scales=list(y=list(relation="free")),
 panel=function(x,y,...) {
-#panel.abline(h=c(60, 70, 80, 90, 100), lwd = 1, lty = 2, col = "gray27")
-	panel.bwplot(x,y,...)
-	panel.mean(x,y,...)
+panel.bwplot(x,y,...)
+panel.mean(x,y,...)
 }
 )
 b1
@@ -495,7 +489,7 @@ par.settings = list(cex=1.2)
 pdf(file= "~/Desktop/evtree_ntrees.pdf", width=10.3, height=5.6)
 b1 <- bwplot (value ~ ntrees | ds, data= as.data.frame(r), 
 horizontal = FALSE,  
-ylab= list("Accuracy [%]", cex=1.1),
+ylab= list("Accuracy (%)", cex=1.1),
 pch= '|',
 ylim=as.data.frame(matrix(c(
 c(60,90),
@@ -503,12 +497,9 @@ c(65,80),
 c(89,94),
 c(60,95)
 ), nrow=2)),
-#scales= list(x= list(rot=60), alternating = F), 
[…]

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_plots.R

@@ -1,0 +1,143 @@
+
+
+###################################################
+### Benchmark plots
+###################################################
+
+## load results
+rm(list = ls())
+for(i in Sys.glob("results/*.RData")) load(i)
+for(i in Sys.glob("results_j48/*.RData")) load(i)
+
+## preprocess for reference evtree
+preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
+    if(isclassification){
+        colnames(d) <- c("evtree", "rpart", "ctree", "J48","evtree", "rpart", "ctree", "J48")
+		d[, 1:4] <- 1 - d[ ,1:4]
+		# evtree complexity was recorded by the number of terminal nodes. J48 by the number of total nodes 
+		# (As J48 allows multiway splits)
+		# evtree and rpart/ctree are compared by the number of terminals
+		# evtree and J48 are compared on the basis of the total number of nodes.  	
+		# For binary trees: "total number of nodes" = "#terminal nodes" * 2 + 1 
+		# As only the ratio of the complexit comp(evtree)/comp(J48) is used we can divide
+		# the results of J48 by the factor(2+1)
+		d[, 8]  <-  d[, 8]/(2+1)    
+    }else{
+    	colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")    	
+   	}
+    d <- as.data.frame(d)
+	nAlgorithms = dim(d)[2]/2
+    for(i in nAlgorithms:1) d[, i] <- d[, i] / d[, 1] * 100
+    for(i in (nAlgorithms*2):(nAlgorithms+1)) d[, i] <- d[, i] / d[, nAlgorithms+1] * 100
+    x <- d[, 1:nAlgorithms]
+    y <- d[, (nAlgorithms+1):(nAlgorithms*2)]
+    rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
+    names(rval)[2] <- "accuracy"
+    rval$complexity <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")[,2]
+    if(isclassification) 
[…]

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_J48.R
corrected wrong file name in bechmark experiments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@313 255f4249-2263-4417-b1ea-8498779e27a7

@@ -199,7 +199,7 @@ save(rionosphere2, file="results_j48/ionosphere_j48.RData")
 data(musk)
 rmusk2 <- bootstrap_j48(musk, nboots=250)
 apply(apply(rmusk2, 2, as.numeric),2,mean)
-save(rmusk2, file="results_j48/musk.RData_j48")
+save(rmusk2, file="results_j48/musk_j48.RData")
 
 ### 5. BreastCancer ###
 data(BreastCancer)

### 2013-05-20 pkg/evtree/Submission_JSS/benchmark_plots.R
fixed problem with plot. current working version of the manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@314 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,5 +1,3 @@
-
-
 ###################################################
 ### Benchmark plots
 ###################################################
@@ -14,21 +12,18 @@ preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
     if(isclassification){
         colnames(d) <- c("evtree", "rpart", "ctree", "J48","evtree", "rpart", "ctree", "J48")
 		d[, 1:4] <- 1 - d[ ,1:4]
-		# evtree complexity was recorded by the number of terminal nodes. J48 by the number of total nodes 
-		# (As J48 allows multiway splits)
-		# evtree and rpart/ctree are compared by the number of terminals
-		# evtree and J48 are compared on the basis of the total number of nodes.  	
-		# For binary trees: "total number of nodes" = "#terminal nodes" * 2 + 1 
-		# As only the ratio of the complexit comp(evtree)/comp(J48) is used we can divide
-		# the results of J48 by the factor(2+1)
-		d[, 8]  <-  d[, 8]/(2+1)    
     }else{
     	colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")    	
    	}
     d <- as.data.frame(d)
 	nAlgorithms = dim(d)[2]/2
     for(i in nAlgorithms:1) d[, i] <- d[, i] / d[, 1] * 100
-    for(i in (nAlgorithms*2):(nAlgorithms+1)) d[, i] <- d[, i] / d[, nAlgorithms+1] * 100
+    if(isclassification)  # for J48 the total number of nodes is used
+	    d[, nAlgorithms*2] <- d[, nAlgorithms*2] / (d[, nAlgorithms+1]*2+1) * 100
+	else
+		d[, nAlgorithms*2] <- d[, nAlgorithms*2] / d[, nAlgorithms+1] * 100
+    for(i in (nAlgorithms*2-1):(nAlgorithms+1)) d[, i] <- d[, i] / d[, nAlgorithms+1] * 100
+    
     x <- d[, 1:nAlgorithms]
     y <- d[, (nAlgorithms+1):(nAlgorithms*2)]
     rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
@@ -71,7 +66,9 @@ r$dssamp <- r$ds:r$samp
 library("multcomp")
[…]

### 2013-06-09 pkg/evtree/Submission_JSS/Revisions/response_revision_09062013.tex
added response letter; updated the vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@317 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[DIN, pagenumber=false, parskip=half,% 
               fromalign=left, fromphone=true,%  
               fromemail=true, fromurl=false, %
               fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}
\usepackage[latin1]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}

\setkomavar{fromname}{Thomas Grubinger}
\setkomavar{fromaddress}{Kampm\"ullerweg 22, 4040 Linz, Austria}
\setkomavar{fromphone}{+43 660 3747597}
\setkomavar{fromemail}{Thomas.Grubinger@utanet.at}
\setkomavar{signature}{Thomas Grubinger, Achim Zeileis, Karl-Peter Pfeiffer}
%\setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
\setkomavar{subject}[]{Revision of  JSS 817}

\begin{document}

\begin{letter}{
  Jan de Leeuw\\
  Department of Statistics\\
  University of California, Los Angeles\\
  United States of America
}

\opening{Dear Prof.~de Leeuw,}

please find attached the revised version of our manuscript
``evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''
(JSS 817). We would like to thank the reviewer for the careful review and the constructive
comments. We have incorporated all of the reviewer's recommendations,
in particular by including a new section that provides guidance on the parameter selection and
giving a better description of the simulation setup, which led to misunderstandings.
We hope to have addressed all concerns in our revision and provide a detailed
point-to-point reply below.

Sincerely,\\[-2cm]
\closing{}

\end{letter}

\newpage

\textbf{\large Point-to-Point Reply}

\bigskip

\textbf{Originality / Previous research}

\textit{The porting of the an EA to the R package is novel.  There is
little or no novelty in the EA itself, as the authors fairly acknowledge. 
I would like to suggest that they could include the following
reviews or give their readers an up to date survey of the work 
in the field and current research issues.}

Thank you for your suggestion on the two surveys. We think they useful
for the reader and referenced them in the paper.

\bigskip

\textbf{Structure}

\textit{It is in general technically 
correct,  except the description of selection, which is definitely 
not a (1+1) strategy, since that would imply a population size of 1.
It looks to me as if what you actually implement is a steady state 
EA with  deterministic crowding.}

We agree that the description of our algorithm as (1+1) strategy was not correct.
We fixed the issue according to your recommendations and describe the algorithm now
as a steady state algorithm with deterministic crowding.

\textbf{Structure}

\textit{1 As an introduction to potential users,  it provides little 
guidance on how to set appropriate parameters
such as population size,  crossover probabilities etc.,  and what 
their effect might be on the results they obtain.  It also does not 
talk at all about the major issue with all  stochastic algorithms, 
which is that even on the same data set the results might different 
greatly from run to run. This is confounded by the fact that the 
variability observed will to no small effect depend on the choice 
of selection and variation operators.  There is a chapter all about 
the issue of how to use and compare evolutionary algorithms in 
the Eiben and Smith book the authors cite.}

We agree that some guidance on the parameter selection can be very useful to the user.
Therefore, we added a new section (6.~Parameter optimization) to the paper, where \textit{evtree} is simulated
with different operator probabilities, number of iterations, and population
sizes. For each setting, \textit{evtree} results on 250~different bootstrap samples are visualized via
boxplots, which should give a good estimate of the underlying variability.

\textit{2.  the ev-tree is a stochastic algorithm,  so the trees evolved 
will vary from run to run,  and will depend on the parameters.  
This is not addressed in section of the results where the performance 
of ev-tree is compared to CARt and rtree over 14 data sets and some 
simulated ones.  Thus this whole section suffers from a  severely 
flawed methodology.   There should be multiple runs and some 
indications of the variability in observed performance should be 
provided. There should be far better discussion of why the suite 
of datsets was chosen, what sort of characteristics they might, or 
might not, have.  Ideally appropriate statistical hypothesis tests 
should be used to see whether the observed differences are 
statistically significant.  This could then lead to some ideas of 
the type of problems on which ev-tree does or does not work well.
}

We fully understand your concerns on this point. However, we think
hat a poor description of the simulation setup led to misunderstandings, but 
the simulation as such was done correctly.

The results presented in the \textit{Performance  comparison} section are based
on 250~repetition for each algorithm and dataset. Here, we used the common approach of
bootstrapping to build 250 different version of the original dataset. The confidence intervals
in Figure~4 are adjusted multiple testing. Confidence intervals that do not cross the zero-line are 
indicate statistical significance at the $5\%$ level. To avoid misunderstandings, we improved 
the description in the paper as follows:

"The analysis is based on the evaluation of 250~bootstrap
samples for each of the 20~datasets. The misclassification rate on the
\textit{out-of-bag} (Hothorn~2005) samples is used as a measure of
predictive accuracy. Furthermore, the complexity is estimated by the number of
terminal nodes. The results are summarized by the mean differences of the 250 runs -- each corresponding to
one of the 250~different bootstrap samples. For the assessment of significant differences 
in predictive accuracy and complexity, respectively, 
Dunnett's correction from \textit{R}~package \textit{multcomp} (Hothorn, Bretz, and Westfall 2008) was used
for calculating simultaneous $95\%$ confidence intervals on the individual datasets. Confidence intervals
that do not cross the zero-line indicate significant differences at the $5\%$ level."

We also added following paragraph which describes our thoughts about the dataset selection:

"The datasets were chosen from different domains and cover a wide range of dataset characteristics and complexities. The sample sizes of the selected datasets range from 214 instances (\textit{Glass identification} data) to 19020 instances (\emph{MAGIC gamma telescope}). The number of attributes vary between 4 (\textit{Servo}) and 180 (\emph{DNA}). The types of attributes vary among datasets, and include datasets which have both categorical and numerical variables or just one of them. The number of classes for the classification task vary between 2 and 11 classes. "

\textit{3. From the perspective of  a piece of scientific work introducing 
and benchmarking a new  different algorithm, then results should 
ideally be provided for some "state of the art algorithms" - at 
least for reference.  If there is, as suggested, a relatively 
simpler way of calling weka libraries,  then why is this not 
done?  Why not compare against J48,  and against more modern 
classifiers like SVM etc?}

We included J48 into the simulations. We are aware that there are better methods in terms of predictive accuracy. However, these methods do not offer the simple interpretability of classification and regression trees. We added following paragraph which acknowledges this issue and narrows the scope of the paper.

"Classification and regression tree models are widely used and is especially attractive for many applications, as tree-structured
models offer a compact and intuitive representation. The goal of the \textit{evtree} was to maintain this simple tree structure and offer better performance (in terms of predictive accuracy and/or complexity) than commonly used recursive partitioning algorithms. However, in cases where the interpretation of the model is not important, black-box methods including \textit{support vector machines} (SVM; Vapnik~2005) and tree ensemble methods like \textit{Random Forests} (Breiman~2001) offer better predictive performance (Caruana and Niculescu-Mizil~2006;  Hastie, Tibshirani, and Friedman~2009) ."

\end{document}

### 2013-06-11 pkg/evtree/Submission_JSS/Revisions/response_revision_09062013.tex
some improvements in revised text and corresponding response
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@318 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Thank you for your suggestion on the two surveys. We think they useful
+Thank you for your suggestion on the two surveys. We concur that these are useful
 for the reader and referenced them in the paper.
@@
 \textit{2.  the ev-tree is a stochastic algorithm,  so the trees evolved 
-will vary from run to run,  and will depend on the parameters.  
-This is not addressed in section of the results where the performance 
+will vary from run to run,  and will depend on the parameters.  This is not addressed in section of the results where the performance 
 of ev-tree is compared to CARt and rtree over 14 data sets and some 
@@
 
-We fully understand your concerns on this point. However, we think
-hat a poor description of the simulation setup led to misunderstandings, but 
-the simulation as such was done correctly.
+We fully understand your concerns on this point. However, we feel
+that our poor description of the simulation setup led to misunderstandings, while
+the simulation as such already addressed your concerns.
 
 The results presented in the \textit{Performance  comparison} section are based
-on 250~repetition for each algorithm and dataset. Here, we used the common approach of
-bootstrapping to build 250 different version of the original dataset. The confidence intervals
-in Figure~4 are adjusted multiple testing. Confidence intervals that do not cross the zero-line are 
-indicate statistical significance at the $5\%$ level. To avoid misunderstandings, we improved 
-the description in the paper as follows:
+on 250~repetition for each algorithm and datasets. Here, we used the common approach of
+bootstrapping to build 250 different versions of the original dataset. Following the
+approach of Hothorn et~al.\ (2005), the confidence intervals in Figure~4 are adjusted
+multiple testing. Thus, confidence intervals that do not encompass the zero-line
+indicate statistical significance at the $5\%$ level.
 
-"The analysis is based on the evaluation of 250~bootstrap
+To avoid misunderstandings, we improved the description in the paper as follows:
+
+``The analysis is based on the evaluation of 250~bootstrap
 samples for each of the 20~datasets. The misclassification rate on the
-\textit{out-of-bag} (Hothorn~2005) samples is used as a measure of
+\textit{out-of-bag} (Hothorn et~al.~2005) samples is used as a measure of
 predictive accuracy. Furthermore, the complexity is estimated by the number of
@@
 for calculating simultaneous $95\%$ confidence intervals on the individual datasets. Confidence intervals
-that do not cross the zero-line indicate significant differences at the $5\%$ level."
+that do not encompass the zero-line indicate significant differences at the $5\%$ level."
 
-We also added following paragraph which describes our thoughts about the dataset selection:
+Furthermore, we added the following paragraph describing the motivation for the dataset selection:
 
-"The datasets were chosen from different domains and cover a wide range of dataset characteristics and complexities.
-The sample sizes of the selected datasets range from 214 instances (\textit{Glass identification} data) to 19020 instances (\emph{MAGIC gamma telescope}).
-The number of attributes vary between 4 (\textit{Servo}) and 180 (\emph{DNA}).
-The types of attributes vary among datasets, and include datasets which have both categorical and numerical variables or just one of them.
-The number of classes for the classification task vary between 2 and 11 classes.
-"
+``The datasets were chosen from different domains and cover a wide range of
+dataset characteristics and complexities. The sample sizes of the selected
+datasets range from 214 instances (\textit{Glass identification} data) to 19020
+instances (\emph{MAGIC gamma telescope}). The number of attributes varies between
+4 (\textit{Servo}) and 180 (\emph{DNA}). The types of attributes vary among
+datasets, and include datasets which have both categorical and numerical
+variables or just one of them. The number of classes for the classification task
+vary between 2 and 11 classes."
 
@@
 and benchmarking a new  different algorithm, then results should 
-ideally be provided for some "state of the art algorithms" - at 
+ideally be provided for some ``state of the art algorithms" - at 
 least for reference.  If there is, as suggested, a relatively 
@@
 
-We included J48 into the simulations.
-We are aware that there are better methods in terms of predictive accuracy.
-However, these methods do not offer the simple interpretability of classification and regression trees.
-We added following paragraph which acknowledges this issue and narrows the scope of the paper.
+We included J48 into the simulations showing that similar to rpart and ctree it
+is often outperformed with respect to prediction performance. Only for the ``Vowel''
+data there is a significant (albeit small) advantage of J48. With respect to complexity,
+J48 typically grows much larger trees (also compared to rpart and ctree).
 
-"Classification and regression tree models are widely used and is especially attractive for many applications, as tree-structured
-models offer a compact and intuitive representation.
-The goal of the \textit{evtree} was to maintain this simple tree structure and offer better performance (in terms of predictive accuracy and/or complexity) than commonly used recursive partitioning algorithms.
-However, in cases where the interpretation of the model is not important, black-box methods including \textit{support vector machines} (SVM; Vapnik~2005) and tree ensemble methods like \textit{Random Forests} (Breiman~2001) offer better predictive performance (Caruana and Niculescu-Mizil~2006;  Hastie, Tibshirani, and Friedman~2009) ."
+We are aware that there are better methods in terms of predictive accuracy. However, these methods do not offer the simple
+interpretability of classification and regression trees. We added following
+paragraph which acknowledges this issue and narrows the scope of the paper.
+
+Classification and regression tree models are widely used and are especially
+attractive for many applications, as tree-structured models offer a compact and
+intuitive representation. The goal of \code{evtree} is to maintain this
+simple tree structure and offer better performance (in terms of predictive
+accuracy and/or complexity) than commonly-used recursive partitioning
+algorithms. However, in cases where the interpretation of the model is not
+important, other ``black-box'' methods including \textit{support vector machines}
+(SVM; Vapnik~2005) and tree ensemble methods like \textit{random
+forests} (Breiman~2001) typically offer better predictive performance
+(Caruana and Niculescu-Mizil~2006;  Hastie, Tibshirani, and Friedman~2009).
 

### 2013-06-16 pkg/evtree/Submission_JSS/benchmark_evtree_parameters.R
moved plots from the benchmark files into benchmark_plots.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@322 255f4249-2263-4417-b1ea-8498779e27a7

@@ -371,145 +371,3 @@ save(rspam, file="results_parameter/spam_parameter.RData")
 rchessboard44_5 <- bootstrap_evtree_parameter(kdata = 1, nboots = 250)
 apply(apply(rchessboard44_5, 2, as.numeric),2,mean)
 save(rchessboard44_5, file="results_parameter/chessboard44_5_parameter.RData")
-
-###################################################
-### Visualisation of datasets
-###################################################
-library("plotrix")
-library("lme4")
-library("multcomp")
-
-## load results
-for(i in Sys.glob("results_parameter/*.RData")) load(i)
-
-panel.mean <- function(x,y,...){
-	x <- as.numeric(x)
-	x.unique <- unique(x)
-	for(X in x.unique) {
-		Y <- y[x == X]
-		if (!length(Y)) next
-		mean.value <- list(y = mean(Y), x = X)
-		do.call("lpoints", c(mean.value, pch = 20, col= "red"))
-	}
-}
-
-# functions for the visualisation of different operatorprobabilities 
-
-preprocess_op <- function(d, dname = "datasetname"){
-	d <- as.data.frame(d)
-    colnames(d) <- c("c0m50sp50", "c20m40sp40","c40m30sp30","c20m20sp60","c20m60sp20")
-    x <- d*100
-    x <- reshape(x, idvar="samp", times=names(x), timevar = "operatorprob",varying= list(names(x)), direction="long")
-	names(x)[[2]] <- "value"
-	x$ds <- dname
-    return(x)
-}
-
-preprocess_op_comp <- function(ntrees, colum){	
-	rt <- preprocess_op(d = rheart[,colum], dname = "Statlog heart")
-	rt <- rbind(rt, preprocess_op(d = rcredit[,colum], dname = "Statlog German credit"))
-	rt <- rbind(rt, preprocess_op(d = rspam[,colum], dname = "Spam"))
-	rt <- rbind(rt, preprocess_op(d = rchessboard44_5[,colum], dname = "Chessboard 4x4 with 5% noise"))
-	rt <- cbind(rt, rep( ntrees, dim(rt)[1]))
-	colnames(rt)[5] <- "nIter"
-	rt
-}
-
[…]

### 2013-06-16 pkg/evtree/Submission_JSS/benchmark_plots.R

@@ -1,5 +1,5 @@
 ###################################################
-### Benchmark plots
+### Benchmark plots for Section: 5. Performance comparisons 
 ###################################################
 
 ## load results
@@ -137,3 +137,151 @@ ciplot(acc_J48, xlim = xlim1, main = "J48", ylab = TRUE,
   xlab = "relative difference in predictive accuracy (%)")
 ciplot(com_J48, xlim = xlim2, main = "",      ylab = FALSE,
   xlab = "relative difference in complexity (%)")
+
+
+
+###################################################
+### Benchmark plots for Section: 6. Parameter optimization 
+###################################################
+library("plotrix")
+library("lme4")
+library("multcomp")
+
+## load results
+for(i in Sys.glob("results_parameter/*.RData")) load(i)
+
+panel.mean <- function(x,y,...){
+	x <- as.numeric(x)
+	x.unique <- unique(x)
+	for(X in x.unique) {
+		Y <- y[x == X]
+		if (!length(Y)) next
+		mean.value <- list(y = mean(Y), x = X)
+		do.call("lpoints", c(mean.value, pch = 20, col= "red"))
+	}
+}
+
+# functions for the visualisation of different operatorprobabilities 
+
+preprocess_op <- function(d, dname = "datasetname"){
+	d <- as.data.frame(d)
+    colnames(d) <- c("c0m50sp50", "c20m40sp40","c40m30sp30","c20m20sp60","c20m60sp20")
+    x <- d*100
+    x <- reshape(x, idvar="samp", times=names(x), timevar = "operatorprob",varying= list(names(x)), direction="long")
+	names(x)[[2]] <- "value"
+	x$ds <- dname
+    return(x)
+}
+
+preprocess_op_comp <- function(ntrees, colum){	
+	rt <- preprocess_op(d = rheart[,colum], dname = "Statlog heart")
+	rt <- rbind(rt, preprocess_op(d = rcredit[,colum], dname = "Statlog German credit"))
+	rt <- rbind(rt, preprocess_op(d = rspam[,colum], dname = "Spam"))
[…]

### 2013-09-03 pkg/partykitPatch/R/as.party.R
patch version 1.0-6: ipred -> TH.data; fix NAMESPACE issues
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@338 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,176 @@
+as.party <- function(obj, ...)
+    UseMethod("as.party")
+
+as.party.rpart <- function(obj, ...) {
+
+    ff <- obj$frame
+    n  <- nrow(ff)
+    if (n==1) return(partynode(as.integer(1)))  # special case of no splits
+
+    is.leaf <- (ff$var == "<leaf>")
+    vnames <- ff$var[!is.leaf]  #the variable names for the primary splits
+
+    index <- cumsum(c(1, ff$ncompete + ff$nsurrogate + 1*(!is.leaf)))
+    splitindex <- list()
+    splitindex$primary <- numeric(n)
+    splitindex$primary[!is.leaf] <- index[c(!is.leaf, FALSE)]
+    splitindex$surrogate <- lapply(1:n, function(i) {
+        prim <- splitindex$primary[i]
+        if (prim < 1 || ff[i, "nsurrogate"] == 0) return(NULL)
+        else return(prim + ff[i, "ncompete"] + 1:ff[i, "nsurrogate"])
+    })
+    
+    mf <- model.frame(obj)
+
+    rpart_fitted <- function() {
+	ret <- as.data.frame(matrix(nrow = NROW(mf), ncol = 0))
+	ret[["(fitted)"]] <- obj$where
+        ret[["(response)"]] <- model.response(mf)        
+        ret[["(weights)"]] <- model.weights(mf)
+        ret
+    }
+    fitted <- rpart_fitted()
+
+    rpart_kids <- function(i) {
+        if (is.leaf[i]) return(NULL)
+        else return(c(i + 1, 
+            which((cumsum(!is.leaf[-(1:i)]) + 1) == cumsum(is.leaf[-(1:i)]))[1] + 1 + i))
+    }
+
+    rpart_onesplit <- function(j) {
+        if (j < 1) return(NULL)
+        ### numeric
+        if (abs(obj$split[j, "ncat"]) == 1) {
+            ret <- partysplit(varid = which(rownames(obj$split)[j] == names(mf)),
+                      breaks = as.double(obj$split[j, "index"]),
+                      right = FALSE,
+                      index = if(obj$split[j, "ncat"] > 0) 2:1)
+        } else {
+            index <- obj$csplit[obj$split[j, "index"],]
+            ### csplit has columns 1:max(nlevels) for all factors
+            index <- index[1:nlevels(mf[, rownames(obj$split)[j]])]
+            index[index == 2] <- NA ### level not present in split
[…]

### 2013-09-03 pkg/partykitPatch/R/ctree.R

@@ -1,0 +1,346 @@
+
+### calculate quad p-values
+.MPinv <- function (X, tol = sqrt(.Machine$double.eps)) 
+{
+    if (length(dim(X)) > 2 || !(is.numeric(X) || is.complex(X))) 
+        stop("'X' must be a numeric or complex matrix")
+    if (!is.matrix(X)) 
+        X <- as.matrix(X)
+    Xsvd <- svd(X)
+    if (is.complex(X)) 
+        Xsvd$u <- Conj(Xsvd$u)
+    Positive <- Xsvd$d > max(tol * Xsvd$d[1], 0)
+    Xplus <- if (all(Positive)) 
+        Xsvd$v %*% (1/Xsvd$d * t(Xsvd$u))
+    else if (!any(Positive)) 
+        array(0, dim(X)[2:1])
+    else Xsvd$v[, Positive, drop = FALSE] %*% ((1/Xsvd$d[Positive]) * 
+        t(Xsvd$u[, Positive, drop = FALSE]))
+    list(Xplus = Xplus, rank = sum(Positive))
+}
+
+.pX2 <- function(lin, exp, cov, pval = TRUE) {
+    if (length(lin) == 1) {
+        if (cov < .Machine$double.eps) return(c(-Inf, -Inf))
+        X2 <- ((lin - exp)^2) / cov
+        df <- 1
+    } else {
+        tmp <- matrix(lin - exp, ncol = 1)
+        Xplus <- .MPinv(matrix(cov, ncol = length(lin)))
+        X2 <- crossprod(tmp, Xplus$Xplus) %*% tmp
+        df <- Xplus$rank
+    }
+    
+    X2 <- pmax(0, X2) ## Z: X2 may be slightly negative...
+    
+    if (pval)
+        return(c(log(X2), pchisq(X2, df = df, lower.tail = TRUE, 
+                                 log.p = TRUE)))
+    return(c(log(X2), NA))
+}
+
+### calculate max-T p-value
+.pmaxT <- function(lin, exp, cov, pval = TRUE) {
+
+    if (length(lin) == 1) {
+        if (cov < .Machine$double.eps) return(c(-Inf, -Inf))
+        maxT <- abs(lin - exp) / sqrt(cov)
+        v <- 1
+        V <- matrix(1)
+    } else {
+        v <- diag(V <- matrix(cov, ncol = length(lin)))
+        lin <- as.vector(lin)[v > 0]
[…]

### 2013-09-03 pkg/partykitPatch/R/party.R

@@ -1,0 +1,506 @@
+## FIXME: data in party
+##   - currently assumed to be a data.frame
+##   - potentially empty
+##   - the following are all assumed to work:
+##     dim(data), names(data)
+##     sapply(data, class), lapply(data, levels)
+##   - potentially these need to be modified if data/terms
+##     should be able to deal with data bases
+
+party <- function(node, data, fitted = NULL, terms = NULL, names = NULL, info = NULL) {
+
+    stopifnot(inherits(node, "partynode"))
+    stopifnot(inherits(data, "data.frame"))
+    ### make sure all split variables are there 
+    ids <- nodeids(node)[!nodeids(node) %in% nodeids(node, terminal = TRUE)]
+    varids <- unique(unlist(nodeapply(node, ids = ids, FUN = function(x) 
+        varid_split(split_node(x)))))
+    stopifnot(varids %in% 1:ncol(data))
+    
+    if(!is.null(fitted)) {
+        stopifnot(inherits(fitted, "data.frame"))
+        stopifnot("(fitted)" == names(fitted)[1])
+        stopifnot(nrow(data) == 0 | nrow(data) == nrow(fitted))
+
+        nt <- nodeids(node, terminal = TRUE)
+        stopifnot(all(fitted[["(fitted)"]] %in% nt))
+
+        node <- as.partynode(node, from = 1L)
+        nt2 <- nodeids(node, terminal = TRUE)
+        fitted[["(fitted)"]] <- nt2[match(fitted[["(fitted)"]], nt)]
+    } else {
+        node <- as.partynode(node, from = 1L)
+    }
+    
+    party <- list(node = node, data = data, fitted = fitted, 
+                  terms = NULL, names = NULL, info = info)
+    class(party) <- "party"
+
+    if(!is.null(terms)) {
+        stopifnot(inherits(terms, "terms"))
+	party$terms <- terms
+    }
+
+    if (!is.null(names)) {
+        n <- length(nodeids(party, terminal = FALSE))
+        if (length(names) != n)
+            stop("invalid", " ", sQuote("names"), " ", "argument")
+        party$names <- names
+    }
+
+    party
+}
+
+length.party <- function(x)
+    length(nodeids(x))
+
+names.party <- function(x)
+    .names_party(x)
+
+"names<-.party" <- function(x, value) {
+     n <- length(nodeids(x, terminal = FALSE))
+     if (!is.null(value) && length(value) != n)
[…]

### 2013-09-03 pkg/partykitPatch/R/plot.R

@@ -1,0 +1,706 @@
+.nobs_party <- function(party, id = 1) {
+  dat <- data_party(party, id = id)
+  if("(weights)" %in% names(dat)) sum(dat[["(weights)"]]) else NROW(dat)
+}
+
+node_inner <- function(obj, id = TRUE, abbreviate = FALSE, fill = "white", gp = gpar())
+{
+  meta <- obj$data
+  nam <- names(obj)
+
+  extract_label <- function(node) {
+    if(is.terminal(node)) return(rep.int("", 2))
+
+    varlab <- character_split(split_node(node), meta)$name
+    if(abbreviate > 0) varlab <- abbreviate(varlab, as.numeric(abbreviate))
+
+    ## FIXME: info processing
+    plab <- ""
+    return(c(varlab, plab))
+  }
+
+  maxstr <- function(node) {
+      lab <- extract_label(node)
+      klab <- if(is.terminal(node)) "" else unlist(lapply(kids_node(node), maxstr))
+      lab <- c(lab, klab)
+      lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
+      return(lab[which.max(nchar(lab))])
+  }
+
+  nstr <- maxstr(node_party(obj))
+  if(nchar(nstr) < 6) nstr <- "aAAAAa"
+
+  ### panel function for the inner nodes
+  rval <- function(node) {  
+    node_vp <- viewport(
+      x = unit(0.5, "npc"),
+      y = unit(0.5, "npc"),
+      width = unit(1, "strwidth", nstr) * 1.3, 
+      height = unit(3, "lines"),
+      name = paste("node_inner", id_node(node), sep = ""),
+      gp = gp
+    )
+    pushViewport(node_vp)
+
+    xell <- c(seq(0, 0.2, by = 0.01),
+  	      seq(0.2, 0.8, by = 0.05),
+  	      seq(0.8, 1, by = 0.01))
+    yell <- sqrt(xell * (1-xell))
+
+    lab <- extract_label(node)
+    fill <- rep(fill, length.out = 2)
+
+    grid.polygon(x = unit(c(xell, rev(xell)), "npc"),
+        	 y = unit(c(yell, -yell)+0.5, "npc"),
+        	 gp = gpar(fill = fill[1]))
+
+    ## FIXME: something instead of pval ?
+    grid.text(lab[1], y = unit(1.5 + 0.5 * FALSE, "lines"))
[…]

### 2013-09-03 pkg/partykitPatch/R/pmmlTreeModel.R

@@ -1,0 +1,272 @@
+pmmlTreeModel <- function(file, ...) {
+  stopifnot(require("XML"))
+  as.party(xmlRoot(xmlTreeParse(file)))
+}
+
+as.party.XMLNode <- function(obj, ...) {
+
+  stopifnot(require("XML"))
+  ## check whether XML specifies a TreeModel
+  stopifnot(c("DataDictionary", "TreeModel") %in% names(obj))
+  if(any(c("MiningBuildTask", "TransformationDictionary", "Extension") %in% names(obj)))
+    warning("not yet implemented")
+  
+  ## needed? obj[["Header"]]
+  
+  ## parse data dictionary
+  extract_empty_model_frame <- function(x) {
+
+    ## extract DataDictionary
+    dd <- x[["DataDictionary"]]
+
+    ## currently we can only look at DataField  
+    if(!all(names(dd) == "DataField")) warning("not yet implemented")
+  
+    ## check columns
+    nc <- as.numeric(xmlAttrs(dd)["numberOfFields"])
+    if(!is.na(nc)) stopifnot(nc == length(dd))
+
+    ## set up data frame (only numeric variables)
+    mf <- as.data.frame(rep(list(1), nc))[0,]
+    names(mf) <- xmlSApply(dd, function(x) xmlAttrs(x)["name"])
+
+    ## modify class if necessary
+    for(i in 1:nc) {
+      optype <- xmlAttrs(dd[[i]])["optype"]
+      switch(optype,
+        "categorical" = {
+           mf[[i]] <- factor(integer(0),
+	     levels = xmlSApply(dd[[i]], function(x) xmlAttrs(x)["value"]))
+        },
+        "ordinal" = {
+           mf[[i]] <- factor(integer(0), ordered = TRUE,
+	     levels = xmlSApply(dd[[i]], function(x) xmlAttrs(x)["value"]))
+        },
+        "continuous" = {
+          dataType <- xmlAttrs(dd[[i]])["dataType"]
+          if(dataType == "integer") mf[[i]] <- integer(0)
+        }
+      )
+    }
+    
+    return(mf)
+  }
+  mf <- extract_empty_model_frame(obj)
+  mf_names <- names(mf)
+  mf_levels <- lapply(mf, levels)
+
+  ## parse MiningSchema
+  extract_terms <- function(x) {
+
+    ## extract MiningSchema
+    ms <- x[["TreeModel"]]
[…]

### 2013-09-03 pkg/partykitPatch/R/simpleparty.R

@@ -1,0 +1,200 @@
+.make_formatinfo_simpleparty <- function(x, digits = getOption("digits") - 4, sep = "")
+{
+  ## digit processing
+  digits <- max(c(0, digits))
+  digits2 <- max(c(0, digits - 2))
+
+  ## type of predictions
+  y <- node_party(x)$info$prediction
+  yclass <- class(y)[1]
+  if(yclass == "ordered") yclass <- "factor"
+  if(!(yclass %in% c("survfit", "factor"))) yclass <- "numeric"
+
+  ## type of weights
+  n <- node_party(x)$info$n
+  if(is.null(names(n))) {
+    wdigits <- 0
+    wsym <- "n"
+  } else {
+    if(names(n) == "w") {
+      wdigits <- max(c(0, digits - 2))
+      wsym <- "w"
+    } else {
+      wdigits <- 0
+      wsym <- "n"
+    }
+  }
+
+  ## compute terminal node labels
+  FUN <- function(info) {
+    yhat <- info$prediction
+    if (yclass == "survfit") {
+        yhat <- .median_survival_time(yhat)
+        yclass <- "numeric"
+    }
+    if(yclass == "numeric") yhat <- format(round(yhat, digits = digits), nsmall = digits)
+    w <- info$n
+    yerr <- if(is.null(info$error)) "" else paste(", err = ",
+      format(round(info$error, digits = digits2), nsmall = digits2),
+      names(info$error), sep = "")
+    rval <- paste(yhat, sep,
+      " (", wsym, " = ", format(round(w, digits = wdigits), nsmall = wdigits),
+      yerr, ")", sep = "")
+    unlist(strsplit(rval, "\n"))
+  }
+  return(FUN)
+}
+
+plot.simpleparty <- function(x, digits = getOption("digits") - 4, tp_args = NULL, ...) {
+  if(is.null(tp_args)) tp_args <- list(FUN = .make_formatinfo_simpleparty(x, digits = digits, sep = "\n"))
+  plot.party(x, tp_args = tp_args, ...)
+}
+
+print.simpleparty <- function(x, digits = getOption("digits") - 4,
+  header = NULL, footer = TRUE, ...)
+{
+  ## header panel
+  if(is.null(header)) header <- !is.null(terms(x))
+  header_panel <- if(header) function(party) {
+    c("", "Model formula:", deparse(formula(terms(party))), "", "Fitted party:", "")
+  } else function(party) ""
+  
+  ## footer panel
[…]

### 2013-09-03 pkg/partykitPatch/src/ctreecall.c

@@ -1,0 +1,62 @@
+
+#include "ctree.h"
+
+SEXP R_split (const SEXP x, const SEXP y, const SEXP weights, const SEXP minbucket) {
+
+    SEXP ans, cx;
+    int n, q, sw = 0;
+    int *thisweights;
+    double *expinf, *covinf;
+
+    n = NROW(x);
+    q = NCOL(y);     
+
+    expinf = Calloc(q, double);
+    covinf = Calloc(q * q, double);
+    thisweights = Calloc(n, int);
+    
+    if (isReal(x)) {
+        PROTECT(ans = allocVector(REALSXP, 1));
+        NA_weights_double(INTEGER(weights), REAL(x), n, thisweights, &sw);
+        C_split_numeric(x, y, thisweights, INTEGER(minbucket)[0], expinf, covinf, REAL(ans));
+    }
+    if (isFactor(x) && !isOrdered(x)) {
+        PROTECT(ans = allocVector(INTSXP, C_nlevels(x)));
+        NA_weights_factor(INTEGER(weights), INTEGER(x), n, thisweights, &sw);
+        C_split_factor(x, y, thisweights, INTEGER(minbucket)[0], expinf, covinf, INTEGER(ans));
+    }
+    if (isInteger(x) || isOrdered(x)) {
+        PROTECT(ans = allocVector(REALSXP, 1));
+        PROTECT(cx = coerceVector(x, REALSXP));
+        NA_weights_double(INTEGER(weights), REAL(cx), n, thisweights, &sw);
+        C_split_numeric(cx, y, thisweights, INTEGER(minbucket)[0], expinf, covinf, REAL(ans));
+        UNPROTECT(1);
+    }
+
+    Free(thisweights);
+    Free(expinf);
+    Free(covinf);
+
+    UNPROTECT(1);
+    return(ans);
+}
+
+SEXP R_LinstatExpCov (const SEXP data, const SEXP inputs, 
+                      const SEXP y, const SEXP weights) {
+
+    SEXP ans, p;
+    int i, *iinputs, *thisweights;
+       
+    thisweights = Calloc(LENGTH(weights), int);
+    iinputs = LOGICAL(inputs);
+    PROTECT(ans = allocVector(VECSXP, LENGTH(data)));
+    for (i = 0; i < LENGTH(data); i++) {
+        if (iinputs[i]) {
+            SET_VECTOR_ELT(ans, i, p = allocVector(VECSXP, 4));
[…]

### 2013-09-03 pkg/partykitPatch/src/ctreelinstat.c

@@ -1,0 +1,349 @@
+
+#include "ctree.h"
+
+/**
+    Linear statistics for conditional inference based on Strasser & Weber (1999)
+    *\file Linstat.c
+    *\author $Author: hothorn $
+    *\date $Date: 2006-08-25 10:53:10 +0200 (Fri, 25 Aug 2006) $
+*/
+
+/**
+    Computes the linear statistic, formula (1) in the paper\n
+    *\param x values of the transformation
+    *\param p dimension of the transformation
+    *\param y values of the influence function
+    *\param q dimension of the influence function
+    *\param weights case weights
+    *\param n number of observations
+    *\param ans return value; a pointer to a REALSXP-vector of length pq
+*/
+  
+void C_Linstat_numeric (const double *x, const double *y, 
+                        const int q, const int *weights, const int n,
+                        double *ans) {
+              
+    int i, k;
+    double tmp;
+
+    for (k = 0; k < q; k++)
+        ans[k] = 0.0;
+            
+    for (i = 0; i < n; i++) {
+                
+        /* optimization: weights are often zero */
+        if (weights[i] == 0) continue;
+ 
+        tmp = x[i] * weights[i];
+        for (k = 0; k < q; k++)  
+            ans[k] += tmp * y[k * n + i];
+    }
+}
+
+
+void C_Linstat_factor (const int *x, const int p,
+                       const double *y, const int q,
+                       const int *weights, const int n,
+                       double *ans) {
+              
+    int i, k, kp, kn, pq;
+    double tmp;
+
+
+    pq = p * q;
+    for (k = 0; k < pq; k++) ans[k] = 0.0;
+
+    for (k = 0; k < q; k++) {
+
+        kn = k * n;
+        kp = k * p;
+            
+        for (i = 0; i < n; i++) {
+                
+            /* optimization: weights are often zero, treatment contrasts */
+            if (weights[i] == 0) continue;
+                
+            /* FIXME: treatment contrasts? */
+            tmp = y[kn + i] * weights[i];
+            ans[kp + (x[i] - 1)] += tmp;
+        }
+    }
+}
+
+/**
+    Conditional expectation and covariance of the influence function\n
[…]

### 2013-09-03 pkg/partykitPatch/src/ctreesplit.c

@@ -1,0 +1,203 @@
+
+#include "ctree.h"
+
+void C_split_numeric (SEXP x, SEXP y, int *iweights, int minbucket, 
+                      double *expinf, double *covinf, double *breaks) {
+
+     int *orderx, i, j, k, n, q, sw = 0, count;
+     double dsw = 0.0, *dx, lastx, thisx, xmax;
+     double cstat = 0.0, laststat = 0.0;
+     double *linstat, *explinstat, *covlinstat, *dy;
+
+     n = NROW(x);
+     if (NCOL(x) != 1 || !isReal(x))
+         error("");
+
+     /* orderx = order(x) */
+     orderx = Calloc(n, int);
+     for (i = 0; i < n; i++) orderx[i] = i;
+     dx = Calloc(n, double);
+     for (i = 0; i < n; i++) dx[i] = REAL(x)[i];
+     rsort_with_index(dx, orderx, n);
+     Free(dx);
+     dx = REAL(x);
+     xmax = dx[orderx[n-1]];
+
+     q = NCOL(y);
+     dy = REAL(y); 
+     
+     linstat = Calloc(q, double);
+     explinstat = Calloc(q, double);
+     covlinstat = Calloc(q * q, double);
+     
+     for (j = 0; j < q; j++) {
+         linstat[j] = 0.0;
+         explinstat[j] = 0.0;
+         for (k = 0; k < q; k++) 
+             covlinstat[j * q + k] = 0.0;
+     }
+
+     dsw = 0;
+     lastx = NA_REAL;
+     breaks[0] = NA_REAL;
+     
+     for (i = 0; i < n; i++) {
+         if (iweights[i] == 0 || ISNA(dx[i])) continue;
+         sw += iweights[i];
+     }
+
+     C_ExpInf(dy, q, iweights, sw, n, expinf);
+     C_CovInf(dy, q, iweights, sw, n, expinf, covinf);
+     
+      /* check sample size contraint */
+     if (sw >= minbucket) {
+     for (i = 0; i < (n - 1); i++) {
+
+         /* value of the orderx[i]th largest observation */
+         thisx = dx[orderx[i]];
+         
+         /* nothing do to for zero weights or NAs */
+         if (iweights[orderx[i]] == 0 || ISNA(thisx)) continue;
+
+         /* sum of the weights of all x <= thisx */
+         dsw += (double) iweights[orderx[i]];
+
+         /* linear statistic for y ~ (x <= thisx) */
[…]

### 2013-09-03 pkg/partykitPatch/src/partysplit.c

@@ -1,0 +1,161 @@
+
+/**
+    infrastructure for splits
+    *\file utils.c
+    *\author $Author$
+    *\date $Date$
+*/
+
+#include "party.h"
+#include "partysplit.h"
+
+
+void init_partysplit(SEXP varid, SEXP breaks, SEXP index, SEXP right,
+             SEXP prob, SEXP info, SEXP split) {
+             
+    if (LENGTH(split) != LENGTH_SPLIT)
+        error("split is not a list with %d elements", LENGTH_SPLIT);
+        
+    if (isInteger(varid))
+        SET_VECTOR_ELT(split, VARID_SPLIT, varid);
+
+    /* FIXME: sort first? */
+    SET_VECTOR_ELT(split, BREAKS_SPLIT, coerceVector(breaks, REALSXP));
+
+    SET_VECTOR_ELT(split, INDEX_SPLIT, index);
+    SET_VECTOR_ELT(split, RIGHT_SPLIT, right);
+    SET_VECTOR_ELT(split, PROB_SPLIT, prob);
+    SET_VECTOR_ELT(split, INFO_SPLIT, info);
+}
+
+int varid_split(SEXP split) {
+
+    return(INTEGER(VECTOR_ELT(split, VARID_SPLIT))[0]);
+}
+
+SEXP breaks_split(SEXP split) {
+
+    return(VECTOR_ELT(split, BREAKS_SPLIT));
+}
+
+SEXP index_split(SEXP split) {
+
+    return(VECTOR_ELT(split, INDEX_SPLIT));
+}
+
+int right_split(SEXP split) {
+
+    return(LOGICAL(VECTOR_ELT(split, RIGHT_SPLIT))[0]);
+}
+
+/**
+    extract or setup a probability distribution on the kids 
+    for random splitting (in case of missing values)
+    *\param split a split object
+*/
+
+SEXP prob_split(SEXP split) {
+
+    SEXP prob, index, breaks;
+    double sum = 0.0;
+    int i;
+
+    /* someone already specified the distribution */
+    prob = VECTOR_ELT(split, PROB_SPLIT);
+    if (prob != R_NilValue)
+        return(prob);
+        
+    /* set it up from index */
+    index = index_split(split);
+    
+    /* index not explicitly given */
+    if (index == R_NilValue) {
+        /* index = 1:(length(breaks) + 1) */
+        breaks = breaks_split(split);
+        if (breaks == R_NilValue)
[…]

### 2013-09-03 pkg/partykitPatch/src/partyutils.c

@@ -1,0 +1,59 @@
+
+/**
+    utility functions.
+    *\file utils.c
+    *\author $Author$
+    *\date $Date$
+*/
+
+#include "party.h"
+
+/**
+    determine in which of the n intervals given by  
+    (-Inf, breaks[0]], (breaks[0], breaks[1]], ..., (breaks[n - 1], +Inf)
+    the observation x falls (intervals are numbered 0, ..., n).
+    *\param x observation
+    *\param breaks ordered cutpoints
+    *\param n length(breaks)
+    *\param right logical: TRUE means intervals are closed on the right, 
+                  FALSE closed on the left
+*/
+
+int cut(double x, double *breaks, int n, int right) {
+
+    int ret, i;
+
+    ret = NA_INTEGER;
+    if (ISNA(x)) return(ret);
+
+    /* x falls in last intervall */
+    if (x > breaks[n - 1]) {
+        ret = n;
+    } else {
+        for (i = 0; i < n; i++) {
+            if (x <= breaks[i]) {
+                ret = i;
+                break;
+            }
+        }
+        /* intervals are closed on the left */
+        if (!right)
+            if (x == breaks[ret]) ret++;
+    }
+    /* ret is in 0, ..., n */
+    return(ret);
+}
+
+void C_SampleNoReplace(int *x, int m, int k, int *ans) {
+     
+     
+    int i, j, n = m;
+         
+    for (i = 0; i < m; i++)
+        x[i] = i;
+    for (i = 0; i < k; i++) {
+        j = floor((double) n * unif_rand());
+        ans[i] = x[j];
+        x[j] = x[--n];
+    }
+}

### 2013-09-03 pkg/partykitPatch/tests/ctree-ctree.R

@@ -1,0 +1,104 @@
+
+### compare partykit::ctree and party::ctree wrt predictions
+### and computing time. But not during R CMD check
+
+library("partykit")
+library("party")
+library("multicore")
+
+set.seed(29)
+
+n <- 200
+dgp <- function(n, response = c("numeric", "ordered", "binary", "factor"), na = FALSE) {
+
+    response <- match.arg(response)
+    x <- runif(n * 10)
+    x <- matrix(x, ncol = 10)
+    x <- cbind(x, x[,1:2] + cbind(rnorm(n, sd = 0.1), rnorm(n, sd = 0.1)))
+    x1 <- sample(gl(4, n / 4))
+    x2 <- sample(ordered(gl(4, n / 4)))
+    y <- as.integer(interaction(x[,1] > 0.5, x[,2] > 0.25))
+    y <- rnorm(n, mean = y)
+    if (na) x[sample(1:(n * 10), n)] <- NA
+    if (response == "numeric")
+        return(data.frame(x, x1, x2, y = y))
+    if (response == "binary") {
+        y <- factor(y < median(y))
+        return(data.frame(x, x1, x2, y = y))
+    } else {
+        y <- cut(y, 
+            breaks = c(-Inf, quantile(y, c(0.25, 0.5, 0.75)), Inf))
+    }
+    if (response == "ordered")
+        return(data.frame(x, x1, x2, y = ordered(y)))
+    if (response == "factor")
+        return(data.frame(x, x1, x2, y = y))
+}
+
+args <- list(response = c("numeric", "ordered", "binary", "factor"),
+             na = c(FALSE, TRUE),
+             teststat = c("quad", "max"),
+             testtype = c("Bonferroni", "Univariate", "Teststatistic"),
+             mincriterion = c(0.8, 0.9, 0.95),
+             maxsurrogate = 0:2, maxdepth = 3)
+
+gr <- do.call("expand.grid", args)
+tgr <- gr
+gr[["response"]] <- as.character(gr[["response"]])
+gr[["teststat"]] <- as.character(gr[["teststat"]])
+gr[["testtype"]] <- as.character(gr[["testtype"]])
+
+
+fun <- function(args) {
+
+    learn <- dgp(n, response = args$response, na = args$na)
+    test <- dgp(n, response = args$response, na = args$na)
+
[…]

### 2013-09-03 pkg/partykitPatch/tests/regtest-party.R

@@ -1,0 +1,62 @@
+## load package and fix seed
+library("partykit")
+library("survival")
+set.seed(1)
+
+## rpart: kyphosis data
+library("rpart")
+data("kyphosis", package = "rpart")
+fit <- rpart(Kyphosis ~ Age + Number + Start, data = kyphosis)
+pfit <- as.party(fit)
+all(predict(pfit, newdata = kyphosis, type = "node") == fit$where)
+
+## J48: iris data
+library("RWeka")
+data("iris", package = "datasets")
+itree <- J48(Species ~ ., data = iris)
+pitree <- as.party(itree)
+stopifnot(all(predict(pitree) == predict(pitree, newdata = iris[, 3:4])))
+
+all.equal(predict(itree, type = "prob", newdata = iris),
+          predict(pitree, type = "prob", newdata = iris))
+all.equal(predict(itree,  newdata = iris),
+          predict(pitree, newdata = iris))
+
+## rpart/J48: GlaucomaM data
+data("GlaucomaM", package = "TH.data")
+w <- runif(nrow(GlaucomaM))
+fit <- rpart(Class ~ ., data = GlaucomaM, weights = w)
+pfit <- as.party(fit)
+all(predict(pfit, type = "node") == fit$where)
+tmp <- GlaucomaM[sample(1:nrow(GlaucomaM), 100),]
+all.equal(predict(fit, type = "prob", newdata = tmp), predict(pfit, type = "prob", newdata = tmp))
+all.equal(predict(fit, type = "class", newdata = tmp), predict(pfit, newdata = tmp))
+itree <- J48(Class ~ ., data = GlaucomaM)
+pitree <- as.party(itree)
+all.equal(predict(itree, newdata = tmp, type = "prob"), predict(pitree, newdata = tmp, type = "prob"))
+
+## rpart: airquality data
+data("airquality")
+aq <- subset(airquality, !is.na(Ozone))
+w <- runif(nrow(aq), max = 3)
+aqr <- rpart(Ozone ~ ., data = aq, weights = w)
+aqp <- as.party(aqr)
+tmp <- subset(airquality, is.na(Ozone))
+all.equal(predict(aqr, newdata = tmp), predict(aqp, newdata = tmp))
+
+## rpart: GBSG2 data
+data("GBSG2", package = "TH.data")
+library("survival")
+fit <- rpart(Surv(time, cens) ~ ., data = GBSG2)
[…]

### 2013-10-05 pkg/partykit/inst/mloss/mloss.Rnw
added old MLOSS paper (with some small touch-ups)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@351 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[a4paper]{article}
\usepackage{jmlr2e}
\usepackage{color,hyperref}

%% software markup
\let\proglang=\textsf
\let\pkg=\textit
\let\code=\texttt

%% \usepackage{Sweave} is essentially
\RequirePackage[T1]{fontenc}
\RequirePackage{ae,fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\setkeys{Gin}{width=.7\textwidth}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, results=hide}
<<options, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", digits = 4)
@

%% header declarations
\author{Torsten Hothorn \and Achim Zeileis \and Kurt Hornik \\
        \addr{\proglang{R} Project for Statistical Computing and Graphics} \\
	\addr{E-mail: Firstname.Lastname@R-project.org}}
\title{Let's Have a \pkg{party}!\\An Open-Source Toolbox for Recursive Partytioning}
\editor{(initial submission for MLOSS track)}

% hyperref setup
\definecolor{Red}{rgb}{0.5,0,0}
\definecolor{Blue}{rgb}{0,0,0.5}
\hypersetup{%
  pdftitle = {Let's Have a party! An Open-Source Toolbox for Recursive Partytioning},
  pdfsubject = {submitted to JMLR-MLOSS},
  pdfkeywords = {open-source software, recursive partitioning, decision trees, statistical learning},
  pdfauthor = {Torsten Hothorn, Achim Zeileis, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

\begin{document}

\maketitle

\begin{abstract}
Package \pkg{party}, implemented in the \proglang{R} system for statistical
computing, provides basic classes and methods for recursive partitioning
along with reference implementations for three recently suggested tree-based
learners: conditional inference trees and forests, and model-based
recursive partitioning.
\end{abstract}

\begin{keywords}
open-source software, recursive partitioning, decision trees, statistical learning. 
\end{keywords}

\section{Overview}

Tree-based learners, also known as recursive partitioning techniques, belong to the
core tools for classification and regression, both in the machine
learning and statistical learning communities. Among the most popular representatives
are C4.5 \citep{mloss:Quinlan:1993} and CART \citep{mloss:Breiman+Friedman+Olshen:1984},
respectively, which were both accompanied by non-free implementations of the algorithms.
Following the success of these seminal algorithms, many modifications, extensions,
and enhancements of the basic recursive partitioning idea have been proposed in 
the literature---more often than not with proprietary or closed-source implementations or
no software at all. A notable exception is the open-source \proglang{Java} package \pkg{Weka}
\citep{mloss:Witten+Frank:2005} that provides free de facto reference implementations
of tree-based learners (including C4.5) or original implementations of novel
algorithms (such as LMT). In statistical learning, the ``lingua franca'' is
\proglang{R} \citep{mloss:R:2007}, an open-source programming environment that 
also provides a reference implementation for CART in package \pkg{rpart}.

Here, we present the \proglang{R} package \pkg{party}, a toolbox for recursive
partytioning. Unlike the implementations above, it does not only implement
specific algorithms, but aims to provide basic infrastructure for tree-based
learners along with flexible and adjustable learning algorithms built on top.
The basic data structures encompass a general high-level class \code{BinaryTree} 
defined in \proglang{R} for
representing various types of binary trees where arbitrary types of information 
can be attached to the nodes. It comes with generic functionality for summarizing
such trees, creating visualizations and computing predictions. This generic
infrastructure has been employed for implementing three flexible recursive
partitioning algorithms. The first is the conditional inference tree framework of
\cite{mloss:Hothorn+Hornik+Zeileis:2006} which learns classical trees with
constant fits in the terminal nodes, based on permutation tests for split selection.
Its user interface is the function \code{ctree} that can be applied out of the box
to regression and classification problems where both inputs and outputs can be
categorical, numerical, censored, or multivariate. Re-using this basic learner,
the package also provides an ensemble method for fitting forests of trees
in the function \code{cforest}. A key advantage of this random forest type
algorithm is that it can produce unbiased variable importance measures
\citep{mloss:Strobl+Boulesteix+Zeileis:2007}. Finally, function \code{mob}
provides a model-based recursive partitioning learner \citep{mloss:Zeileis+Hothorn+Hornik:2008}
that allows to attach arbitrary parametric models
(such as, e.g., linear or logistic regression, multivariate or survival models)
to the nodes in the tree and uses structural change tests for split selection.

The package \pkg{party} and all the components it depends on---including packages
with tools for model construction, with visualization infrastructure and with
inference techniques---are available under the General
Public License (GPL) from the Comprehensive \proglang{R} Archive Network (CRAN) at
\url{http://CRAN.R-project.org/}. The package contains \proglang{R} and \proglang{C}
code and has been successfully built on a multitude of platforms including Windows,
Mac~OS and various flavors of Linux.  It can be installed automatically from within
\proglang{R} via the command \code{install.packages("party", dependencies = TRUE)}
(which also installs all packages needed for examples); to get started, see the
online reference manual, e.g., via \code{help(package = "party")}. More advanced
user- and developer-level documentation is available in the package vignettes (see
\code{vignette(package = "party")}) and the \proglang{C} source code \pkg{doxygen}
documentation (in the \code{inst/documentation} sub-directory of the package sources).

In the following, we give a brief tour of the package by applying the three core
algorithms to classification problems before discussing some details of the developer-level
features.

\section{User interface}

<<GlaucomaM, echo=FALSE>>=
library("partykit")
data("GlaucomaM", package = "TH.data")
set.seed(10131218)
idx <- sample(1:nrow(GlaucomaM), 150)
GM1 <- GlaucomaM[idx,]
GM2 <- GlaucomaM[-idx,]
@
%% geeky detail: random seed was chosen as
%% str2num <- function(x) as.numeric(paste(gsub(" ", "0",
%%   format(c(10, match(tolower(strsplit(x, "")[[1]]), letters)))[-1]), collapse = ""))
%% str2num("JMLR")

To demonstrate usage of the high-level algorithms, we present some short examples
for fitting single trees, random forests and model-based trees in \pkg{party}
for two different binary classification tasks.

\paragraph{Trees and forests.} The \code{GlaucomaM} data set contains \Sexpr{nrow(GlaucomaM)}
observations of glaucoma classification in human eyes (affected vs.\ healthy) 
along with \Sexpr{ncol(GlaucomaM)-1} features derived from confocal laser scanning images
of the optic nerve head. In the \proglang{R} session below, we first load
package and data and then split it randomly into training and test sets of
150 and \Sexpr{nrow(GlaucomaM) - 150} observations, respectively.
<<GlaucomaM1, eval=FALSE>>=
<<GlaucomaM>>
@
A \code{ctree} for predicting the \code{Class} variable, as explained by (\code{\~})
all other variables (\code{.}) contained in the learning set \code{GM1}, can be fitted via
<<ctree>>=
ct <- ctree(Class ~ ., data = GM1)
@
creating a fitted \code{BinaryTree} object \code{ct}. For obtaining a short summary
of the tree it can be either printed by typing \code{ct} at the command line (or more
explicitly: \code{print(ct)}) or visualized by \code{plot(ct)}. The resulting plot
(depicted in the left panel of Figure~\ref{fig}) shows that the glaucoma probability 
increases clearly for lower volumes of the optic nerve head 
(as captured in the different volume measures
\code{vari} and \code{vasg}). After three splits, no more significant splits were found.
To evaluate the performance of this fitted tree on the test data \code{GM2}, we can
compute the confusion matrix of true and predicted class memberships
<<ctree-confusion, results=verbatim>>=
table(true = GM2$Class, pred = predict(ct, newdata = GM2))
@
corresponding to a misclassification rate of
\Sexpr{100 * round(1 - sum(diag(table(true = GM2$Class, pred = predict(ct, newdata = GM2))))/nrow(GM2), digits = 3)}\%.
Of course, this is only a rough ad hoc estimate of the generalization error, for a more
thorough evaluation and comparison with other tree algorithms see the original publications
mentioned above.

\paragraph{Model trees.} For illustrating the \code{mob} algorithm, we partition a logistic
regression for \code{diabetes} explained by (\code{\~}) \code{glucose} in the
well-known Pima Indians Diabetes data set, using five remaining variables for partitioning.
<<PimaIndiansDiabetes>>=
data("PimaIndiansDiabetes", package = "mlbench")
PID <- subset(PimaIndiansDiabetes, glucose > 0 & pressure > 0 & mass > 0)
@
This loads the data set and selects a subset \code{PID} of \Sexpr{nrow(PID)} observations
excluding physically impossible zero measurements in some of the variables. The \code{mob}
based on a logistic regression (i.e., a generalized linear model with binomial family)
can be fitted via
<<mob>>=
mb <- glmtree(diabetes ~ glucose | pregnant + pressure + mass + pedigree + age,
  data = PID, family = binomial)
@
Its visualization produced by \code{plot(mb)} is depicted in the right panel of
Figure~\ref{fig}. It shows the empirical \code{diabetes} proportions over \code{glucose}
groups along with the predictions from the logistic regression, conveying that the overall
level (regression intercept) increases with body mass index and age while the glucose
effect (regression slope) decreases. For a benchmark comparison of this algorithm with
other trees (including \pkg{Weka}'s LMT and J4.8 implementation of C4.5) see
\cite{mloss:Zeileis+Hothorn+Hornik:2008}.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<mob-plot, echo=FALSE, fig=TRUE, height=6, width=14>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
plot(ct, newpage = FALSE, pop = TRUE,
  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.8))))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
plot(mb, newpage = FALSE, pop = TRUE, tp_args = list(ylines = 1.3,
  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.8))),
  linecol = "black"))
popViewport(2)
@
\caption{\label{fig} Tree visualizations: distribution of glaucoma
for fitted \code{ctree} (left) and distribution of diabetes over glucose
for logistic-regression \code{mob} (right).}
\end{center}
\end{figure}

%% This very brief tour of the features in \pkg{party} only hinted at the
%% flexibility of the tools, some more information is given below. Detailed
%% information on arguments and control options is available in the
%% manual pages (\code{help("ctree", package = "party")} etc.) and
%% the package vignettes.

\section{Internal structure}

The \pkg{party} package contains \proglang{R} code defining classes and 
methods along with high-level algorithms built on top of these, as well as
an additional program layer in \proglang{C} that implements time-critical operations
by directly modifying previously created \proglang{R} objects.
In particular, the \proglang{R} building blocks encompass
classes and methods for \code{LearningSample} and \code{BinaryTree} objects
which can be re-used in high-level algorithms (such as
\code{ctree}, \code{cforest} and \code{mob}) or directly at \proglang{C} level.
\code{LearningSample} is a data set---created from external sources or conversion
from other objects---which is enhanced by meta-information such as variable
orderings that can speed up split searches. \code{BinaryTree} is
a tree representation based on a recursive list structure
containing node information (e.g., primary splits, surrogate splits, summary statistics,
predictions, etc.) as well as left and right daughter nodes. Recursive methods
for predicting future cases, textual and graphical representations etc., 
are defined for objects of class \code{BinaryTree}.

The tree-growing can be controlled in various ways: starting from simple 
hyper-parameters, such as minimal leaf size, significance level or 
maximal depth, to very complex arguments (e.g., parametric model classes 
for \code{mob} or visualization functions for terminal nodes). At the \proglang{R}
level, users and developers can make use of the documented API for
inspecting and manipulating \code{BinaryTree} objects and, in addition,
are free to extend currently implemented methodology.

Unit testing of \pkg{party} (as for more than 1,200 other \proglang{R} packages on
CRAN) is performed automatically on a daily basis for various platforms and
versions of \proglang{R}.

\section{Conclusions}

The package \pkg{party} extends the set of machine learning techniques
available in the \proglang{R} system for statistical computing
(see the corresponding CRAN task view at \url{http://CRAN.R-project.org/src/contrib/Views/})
with a toolbox for recursive partitioning. The package provides basic
infrastructure for binary trees along with reference implementations of 
three recently-suggested tree-based learners: conditional inference trees and forests,
and model-based recursive partitioning. Party on!

\bibliography{mloss}

\end{document}

### 2013-11-03 pkg/partykit/R/modelparty.R
improved default printing for model parties
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@360 255f4249-2263-4417-b1ea-8498779e27a7

@@ -716,11 +716,12 @@ sctest.modelparty <- function(object, node = NULL, ...)
 
 print.modelparty <- function(x, node = NULL,
   FUN = NULL, digits = getOption("digits") - 4L,
-  header = TRUE, footer = TRUE,
-  title = "Model-based recursive partitioning", objfun = "", ...)
+  header = TRUE, footer = TRUE, title = NULL, objfun = "", ...)
 {
   digits <- max(c(0, digits))
   if(objfun != "") objfun <- paste(" (", objfun, ")", sep = "")
+  if(is.null(title)) title <- sprintf("Model-based recursive partitioning (%s)",
+    deparse(x$info$call$fit))
 
   if(is.null(node)) {
     header_panel <- if(header) function(party) {      

### 2013-11-11 pkg/partykit/R/mob-plot.R
small bug fix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@375 255f4249-2263-4417-b1ea-8498779e27a7

@@ -214,7 +214,7 @@ node_bivplot <- function(mobobj, which = NULL, id = TRUE, pop = TRUE,
       yi <- y[o]
       xi <- xi[o]
       yfit <- if(is.null(node$info$object)) {
-        fitted(refit.modelparty(mobobj, node = nid))
+        fitted(refit.modelparty(mobobj, node = nid))[o]
       } else {
         fitted(node$info$object)[o]
       }

### 2013-12-13 pkg/partykit/R/ctree.R
deal with empty levels in quad teststat; reported by Wei
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@391 255f4249-2263-4417-b1ea-8498779e27a7

@@ -25,8 +25,15 @@
         X2 <- ((lin - exp)^2) / cov
         df <- 1
     } else {
+        v <- diag(V <- matrix(cov, ncol = length(lin)))
+        ### remove elements with zero variance first
+        lin <- as.vector(lin)[v > 0]
+        exp <- as.vector(exp)[v > 0]
+        V <- V[v > 0, v > 0, drop = FALSE]
+        v <- v[v > 0]
+        if (length(v) == 0) return(c(-Inf, -Inf))
         tmp <- matrix(lin - exp, ncol = 1)
-        Xplus <- .MPinv(matrix(cov, ncol = length(lin)))
+        Xplus <- .MPinv(V)
         X2 <- crossprod(tmp, Xplus$Xplus) %*% tmp
         df <- Xplus$rank
     }

### 2013-12-13 pkg/partykit/tests/bugfixes.R

@@ -337,3 +337,386 @@ data("Vowel")
 ct <- ctree(V2 ~ V1, data = Vowel[1:200,]) ### only levels 1:4 in V1
 p1 <- predict(ct, newdata = Vowel) ### 14 levels in V1
 
+### deal with empty levels for teststat = "quad" by
+### removing elements of the teststatistic with zero variance
+### reported by Wei-Yin Loh <loh@stat.wisc.edu>
+tdata <-
+structure(list(ytrain = structure(c(3L, 7L, 3L, 2L, 1L, 6L, 2L, 
+1L, 1L, 2L, 1L, 2L, 3L, 3L, 2L, 1L, 2L, 6L, 2L, 4L, 6L, 1L, 2L, 
+3L, 7L, 6L, 4L, 6L, 2L, 2L, 1L, 2L, 6L, 1L, 7L, 1L, 3L, 6L, 2L, 
+1L, 7L, 2L, 7L, 2L, 3L, 2L, 1L, 1L, 3L, 1L, 6L, 2L, 2L, 2L, 2L, 
+2L, 1L, 1L, 6L, 6L, 7L, 2L, 2L, 2L, 2L, 2L, 1L, 3L, 6L, 5L, 1L, 
+1L, 4L, 7L, 2L, 3L, 3L, 3L, 1L, 8L, 1L, 6L, 2L, 8L, 3L, 4L, 6L, 
+2L, 7L, 3L, 6L, 6L, 1L, 1L, 2L, 6L, 3L, 3L, 1L, 2L, 3L, 1L, 2L, 
+7L, 2L, 3L, 6L, 2L, 5L, 2L, 2L, 2L, 1L, 3L, 3L, 7L, 3L, 2L, 3L, 
+3L, 1L, 6L, 1L, 1L, 1L, 7L, 1L, 3L, 7L, 6L, 1L, 3L, 3L, 6L, 4L, 
+2L, 3L, 2L, 8L, 3L, 4L, 2L, 2L, 2L, 3L, 2L, 2L, 2L, 3L, 4L, 6L, 
[…]

### 2014-01-06 pkg/partykit/R/as.party.R
avoid importing RWeka -> instead add a require("RWeka") in as.party.J48
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@393 255f4249-2263-4417-b1ea-8498779e27a7

@@ -111,13 +111,16 @@ model.frame.rpart <- function(formula, ...) {
 
 as.party.J48 <- function(obj, ...) {
 
+  ## needs RWeka and rJava
+  stopifnot(require("RWeka"), require("rJava"))
+
   ## construct metadata
   mf <- model.frame(obj)
   mf_class <- sapply(mf, function(x) class(x)[1L])
   mf_levels <- lapply(mf, levels)
 
   x <- rJava::.jcall(obj$classifier, "S", "graph")
-  x <- parse_Weka_digraph(x, plainleaf = TRUE)
+  x <- RWeka::parse_Weka_digraph(x, plainleaf = TRUE)
   nodes <- x$nodes
   edges <- x$edges
   is.leaf <- x$nodes[, "splitvar"] == ""

### 2014-01-06 pkg/partykit/R/as.party.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@394 255f4249-2263-4417-b1ea-8498779e27a7

@@ -112,7 +112,7 @@ model.frame.rpart <- function(formula, ...) {
 as.party.J48 <- function(obj, ...) {
 
   ## needs RWeka and rJava
-  stopifnot(require("RWeka"), require("rJava"))
+  stopifnot(require("RWeka"))
 
   ## construct metadata
   mf <- model.frame(obj)

### 2014-01-06 pkg/partykit/inst/DSC-2009/abstract.tex
re-organized slides directory -> split up into individual conference dirs the references are always taken from partykit/vignettes/party.bib (have to be suitably linked to make available)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@395 255f4249-2263-4417-b1ea-8498779e27a7

--- 2009-02-25 pkg/inst/slides/DSC-2009_abstract.tex
@@
-\documentclass[10pt,a4paper]{article}
+\documentclass[10pt]{article}
 
-\usepackage{url,a4wide}
+\usepackage{url}
 \usepackage[authoryear,round]{natbib}
-\topmargin-1.5cm
-\textheight28cm
 
 \renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
-\renewcommand{\author}[1]{\begin{center}{\bf #1}\end{center}}
-\newenvironment{affiliations}{\footnotesize \begin{enumerate}}{\end{enumerate}}
-\newenvironment{keywords}{\paragraph{Keywords:}}{\bigskip}
+\newcommand{\affiliations}{\footnotesize}
+\newcommand{\keywords}{\paragraph{Keywords:}}
+
+\setlength{\oddsidemargin}{0cm} \setlength{\evensidemargin}{0cm}
+\setlength{\textwidth}{16.5cm} \setlength{\topmargin}{-1cm}
+\setlength{\textheight}{24.5cm}
 
@@
 \title{\pkg{partykit}: A Toolbox for Recursive Partytioning}
-\author{Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}
+
+\begin{center}
+  {\bf Torsten Hothorn$^{1}$, Achim Zeileis$^{2}$}
+\end{center}
 
 \begin{affiliations}
+\begin{enumerate}
   \item Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen, Germany\\
@@
         \texttt{Achim.Zeileis@R-project.org}
+\end{enumerate}
 \end{affiliations}
 
-\begin{keywords}
-Regression Trees, Visualization, CTree, CHAID, MOB, Weka, PMML.
-\end{keywords}
+\keywords Regression Trees, Visualization, CTree, CHAID, MOB, Weka, PMML.
+
+\vskip 0.8cm
 
@@
 available to the \proglang{R} user, e.g., through packages
-\pkg{rpart} \citep{partykit:rpart},
-\pkg{RWeka} \citep{partykit:RWeka},
-\pkg{party} \citep{partykit:party}, or
-\pkg{mvpart} \citep{partykit:mvpart}.
+\pkg{rpart} \citep{rpart},
+\pkg{RWeka} \citep{RWeka},
+\pkg{party} \citep{party}, or
+\pkg{mvpart} \citep{mvpart}.
 
@@
 
-To overcome these difficulties, the \pkg{partykit} package \citep{partykit:partykit} offers a 
+To overcome these difficulties, the \pkg{partykit} package \citep{partykit} offers a 
 unified representation of tree objects along with \code{predict()}, 
@@
 trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
-as well as a re-implementation of conditional inference trees \citep{partykit:Hothorn+Hornik+Zeileis:2006}.
+as well as a re-implementation of conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}.
 
@@
 such as CHAID or re-implement techniques such as model-based recursive partitioning
-\citep[MOB,][]{partykit:Zeileis+Hothorn+Hornik:2008}.
+\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}.
 
 \bibliographystyle{jss}
-\bibliography{partykit}
+\bibliography{party}
 

### 2014-01-06 pkg/partykit/inst/useR-2009/abstract.tex
re-organized slides directory -> split up into individual conference dirs the references are always taken from partykit/vignettes/party.bib (have to be suitably linked to make available)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@395 255f4249-2263-4417-b1ea-8498779e27a7

--- 2009-02-25 pkg/inst/slides/partykit_useR2009.tex
@@
-\documentclass[10pt,a4paper]{article}
+\documentclass[10pt]{article}
 
-\usepackage{url,a4wide}
+\usepackage{url}
 \usepackage[authoryear,round]{natbib}
@@
 \renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
-\renewcommand{\author}[1]{\begin{center}{\bf #1}\end{center}}
-\newenvironment{affiliations}{\footnotesize \begin{enumerate}}{\end{enumerate}}
-\newenvironment{keywords}{\paragraph{Keywords:}}{\bigskip}
+\newcommand{\affiliations}{\footnotesize}
+\newcommand{\keywords}{\paragraph{Keywords:}}
+
+\setlength{\oddsidemargin}{0cm} \setlength{\evensidemargin}{0cm}
+\setlength{\textwidth}{16.5cm} \setlength{\topmargin}{-1cm}
+\setlength{\textheight}{24.5cm}
 
@@
         \texttt{Achim.Zeileis@R-project.org}
+\end{enumerate}
 \end{affiliations}
 
-\begin{keywords}
-Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
-\end{keywords}
+\keywords Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
+
+\vskip 0.8cm
 
@@
 available to the \proglang{R} user, e.g., through packages
-\pkg{rpart} \citep{PKG:rpart},
-\pkg{RWeka} \citep{PKG:RWeka},
-\pkg{party} \citep{PKG:party}, or
-\pkg{mvpart} \citep{PKG:mvpart}.
+\pkg{rpart} \citep{rpart},
+\pkg{RWeka} \citep{RWeka},
+\pkg{party} \citep{party}, or
+\pkg{mvpart} \citep{mvpart}.
 
@@
 
-To overcome these difficulties, the \pkg{partykit} package \citep{PKG:partykit} offers a 
+To overcome these difficulties, the \pkg{partykit} package \citep{partykit} offers a 
 unified representation of tree objects along with \code{predict()}, 
@@
 trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
-as well as a re-implementation of conditional inference trees \citep{Hothorn:2006:JCGS}.
+as well as a re-implementation of conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}.
 
@@
 \bibliographystyle{jss}
-\bibliography{partykit}
+\bibliography{party}
 

### 2014-01-06 pkg/partykit/inst/useR-2009/slides.Rnw
re-organized slides directory -> split up into individual conference dirs the references are always taken from partykit/vignettes/party.bib (have to be suitably linked to make available)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@395 255f4249-2263-4417-b1ea-8498779e27a7

--- 2009-07-02 pkg/inst/slides/useR-2009_slides.Rnw
@@
 <<ex-rpart, echo = TRUE>>=
-data("GlaucomaM", package = "ipred")
+data("GlaucomaM", package = "TH.data")
 library("rpart")
@@
 
-My students implemented CHAID as a software project last winter. I wanted
-them to focus on the algorithm, not on technical details.
+My students implemented CHAID as a software project last winter. \\
+I wanted them to focus on the algorithm, not on technical details.
 
-We only had to implement variable selection, splitting and stopping rules,
+They only had to implement variable selection, splitting and stopping rules,
 \pkg{partykit} takes care of the rest.

### 2014-01-06 pkg/partykit/inst/ERCIM-2011/slides.Rnw
abstract and slides from ERCIM 2011 (slightly updated)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@396 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[11pt,compress,t]{beamer}
\usetheme{Z}
\usepackage{amsfonts,amstext,amsmath}
%% need no \usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}

%% commands
\newcommand{\ui}{\underline{i}}
\newcommand{\oi}{\overline{\imath}}
\newcommand{\darrow}{\stackrel{\mbox{\tiny \textnormal{d}}}{\longrightarrow}}
\newcommand{\parrow}{\stackrel{\mbox{\tiny \textnormal{p}}}{\longrightarrow}}
\newcommand{\dotequals}{\stackrel{\cdot}{=}}
\newcommand{\efp}{\mathrm{\it efp}}
\newcommand{\given}{\, | \,}
\newcommand{\ltime}{\lambda_\mathrm{time}}
\newcommand{\lcomp}{\lambda_\mathrm{comp}}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<packages>>=
library("partykit")
library("rpart")
library("RWeka")
library("evtree")
@

<<data>>=
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
ttnc <- transform(ttnc, Treatment = factor(
  Gender == "Female" | Age == "Child",
  levels = c(FALSE, TRUE),
  labels = c("Normal\n(Male&Adult)", "Preferential\n(Female|Child)")
))
@

<<rpart, eval=FALSE>>=
rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
@

<<ctree, eval=FALSE>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
  control = ctree_control(mincrit = 1 - 0.01, teststat = "max"))
@

<<evtree, eval=FALSE>>=
set.seed(1071)
ev <- evtree(Survived ~ Gender + Age + Class, data = ttnc,
  maxdepth = 3)
@

<<pmml, eval=FALSE>>=
pm <- pmmlTreeModel("ttnc.xml")
@

<<mob, eval=FALSE>>=
mb <- glmtree(Survived ~ Treatment | Age + Gender + Class,
  data = ttnc, family = binomial, alpha = 0.01)
@

<<cache>>=
if(file.exists("slides-trees.rda")) load("slides-trees.rda") else {
<<rpart>>
<<ctree>>
<<evtree>>
<<pmml>>
<<mob>>
save(rp, ct, ev, pm, mb, file = "slides-trees.rda")
}
@

<<J48>>=
j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
@

\begin{document}

\title{partykit: A Toolkit for Recursive Partytioning}
\author{Achim Zeileis, Torsten Hothorn}
\URL{http://eeecon.uibk.ac.at/~zeileis/}
\lecture{partykit: A Toolkit for Recursive Partytioning}

\subsection{Overview}

\begin{frame}
\frametitle{Overview}

\begin{itemize}
  \item Status quo: R software for tree models
  \item New package: partykit
  \begin{itemize}
    \item Unified infrastructure for recursive partytioning
    \item Classes and methods
    \item Interfaces to \code{rpart}, \code{J48}, \dots
    \item Illustrations
  \end{itemize}
  \item Future: Next steps
\end{itemize}

\end{frame}

\subsection{Status quo: R software for tree models}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Status quo:}
The CRAN task view on ``Machine Learning'' at \url{http://CRAN.R-project.org/view=MachineLearning}
lists numerous packages for tree-based modeling and recursive partitioning, including
%
\begin{itemize}
  \item rpart (CART),
  \item tree (CART),
  \item mvpart (multivariate CART),
  \item RWeka (J4.8, M5', LMT),
  \item party (CTree, MOB),
  \item and many more (C50, quint, stima, \dots).
\end{itemize}

\medskip

\textbf{Related:} Packages for tree-based ensemble methods
such as random forests or boosting, e.g., randomForest, gbm, mboost, etc.

\end{frame}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Furthermore:} Tree algorithms/software without R interface, e.g.,
%
\begin{itemize}
  \item QUEST,
  \item GUIDE,
  \item LOTUS,
  \item CRUISE,
  \item \dots
\end{itemize}

\medskip

\textbf{Currently:}
All algorithms/software have to deal with similar problems \emph{but}
provide different solutions without reusing code.

\end{frame}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Challenge:} For implementing new algorithms in R, code
is required not only for fitting the tree model on the learning data
but also
\begin{itemize}
  \item representing fitted trees,
  \item printing trees,
  \item plotting trees,
  \item computing predictions from trees.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Question:} Wouldn't it be nice if there were an R package
that provided code for
\begin{itemize}
  \item representing fitted trees,
  \item printing trees,
  \item plotting trees,
  \item computing predictions from trees?
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{R software for tree models}

\textbf{Answer:} The R package \emph{partykit} provides unified
infrastructure for recursive partytioning, especially
\begin{itemize}
  \item representing fitted trees,
  \item printing trees,
  \item plotting trees,
  \item computing predictions from trees!
\end{itemize}

\end{frame}

\subsection{New package: partykit}

\begin{frame}
\frametitle{partykit: Unified infrastructure}

\textbf{Design principles:} Toolkit for recursive partytitioning.
\begin{itemize}
  \item One `agnostic' base class which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits in each terminal node.
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing objects (rpart, J48, etc.) to new class.
  \item Use simple/fast S3 classes and methods.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Base classes}

\textbf{Class constructors:} Generate basic building blocks.
\begin{itemize}
  \item \code{partysplit(varid, breaks = NULL, index = NULL, ...)}

    where \code{breaks} provides the breakpoints wrt variable \code{varid};
    \code{index} determines to which kid node observations are assigned.

  \item \code{partynode(id, split = NULL, kids = NULL, ...)}

    where \code{split} is a \class{partysplit} and \code{kids} a list of \class{partynode}s.

  \item \code{party(node, data, fitted = NULL, ...)}

    where \code{node} is a \class{partynode} and \code{data} the
    corresponding (learning) data (optionally without any rows) and
    \code{fitted} the corresponding fitted nodes.
\end{itemize}

\medskip

\textbf{Additionally:} All three objects have an \code{info} slot where
optionally arbitrary information can be stored.

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Further classes and methods}

\textbf{Further classes:} For trees with constant fits in each terminal node,
both inheriting from \class{party}.
\begin{itemize}
  \item \class{constparty}: Stores full observed response and fitted terminal nodes
    in \code{fitted}; predictions are computed from empirical distribution of
    the response.
  \item \class{simpleparty}: Stores only one predicted response value along with
    some summary details (such as error and sample size) for each terminal node
    in the corresponding \code{info}.
\end{itemize}

\medskip

\textbf{Methods:}
\begin{itemize}
  \item Display: \fct{print}, \fct{plot}, \fct{predict}.
  \item Query: \fct{length}, \fct{width}, \fct{depth}, \fct{names}, \fct{nodeids}.
  \item Extract: \code{[}, \code{[[}, \fct{nodeapply}.
  \item Coercion: \fct{as.party}. 
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{partykit: Illustration}

\textbf{Intention:}
\begin{itemize}
  \item Illustrate several trees using the same data.
  \item Not use the iris data (or something from mlbench).
\end{itemize}

\medskip

\textbf{Solution:}
\begin{itemize}
  \item Use Titanic survival data (oh well\dots).
  \item In case you are not familiar with it: Survival status, gender, age (child/adult), and
    class (1st, 2nd, 3rd, crew) for the \Sexpr{nrow(ttnc)}~persons on the ill-fated
    maiden voyage of the Titanic.
\end{itemize}

\medskip

\textbf{Question:} Who survived? Or how does the probability of survival vary across
the covariates?

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Interface to \texttt{rpart}}

\textbf{CART:} Apply \fct{rpart} to preprocessed \code{ttnc} data (see also below).

<<rpart-show, echo=TRUE, eval=FALSE>>=
<<rpart>>
@

\medskip

Standard plot:

<<rpart-plot, echo=TRUE, eval=FALSE>>=
plot(rp)
text(rp)
@

\medskip

Visualization via partykit:

<<rpart-party-plot, echo=TRUE, eval=FALSE>>=
plot(as.party(rp))
@

\end{frame}

\begin{frame}
\frametitle{partykit: Interface to \texttt{rpart}}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-rpart-visualization1, fig=TRUE, height=6, width=9>>=
<<rpart-plot>>
@

\end{frame}

\begin{frame}
\frametitle{partykit: Interface to \texttt{rpart}}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-rpart-visualization2, fig=TRUE, height=8, width=12>>=
<<rpart-party-plot>>
@

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Interface to \texttt{rpart}}

<<rpart-print, echo=TRUE, results=verbatim>>=
rp
@

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Interface to \texttt{rpart}}

<<rpart-print, echo=TRUE, results=verbatim>>=
as.party(rp)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Interface to \texttt{rpart}}

\textbf{Prediction:} Compare rpart's C code and partykit's R code
for (artificially) large data set.

<<rpart-predict, echo=TRUE, results=verbatim>>=
nd <- ttnc[rep(1:nrow(ttnc), 100), ]
system.time(p1 <- predict(rp, newdata = nd, type = "class"))
system.time(p2 <- predict(as.party(rp), newdata = nd))
table(rpart = p1, party = p2)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Interface to \texttt{J48}}

\textbf{J4.8:} Open-source implementation of C4.5 in RWeka.

<<J48-show, echo=TRUE, eval=FALSE>>=
<<J48>>
@

\medskip

Results in a tree with multi-way splits which previously
could only be displayed via Weka itself or Graphviz but not in R
directly. Now:

<<J48-plot, echo=TRUE, eval=FALSE>>=
j48p <- as.party(j48)
plot(j48p)
@

\medskip

Or just a subtree:

<<J48-subtree, echo=TRUE, eval=FALSE>>=
plot(j48p[11])
@

\end{frame}

\begin{frame}
\frametitle{partykit: Interface to \texttt{J48}}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-J48-visualization, fig=TRUE, height=8, width=12>>=
plot(as.party(j48), tp_args = list(id = FALSE))
@

\end{frame}

\begin{frame}
\frametitle{partykit: Interface to \texttt{J48}}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-J48-visualization2, fig=TRUE, height=6, width=9>>=
plot(as.party(j48)[11], tnex = 1)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{partykit: Further interfaces}

\textbf{PMML:} Predictive Model Markup Language. XML-based format
exported by various software packages including SAS, SPSS, R/pmml.
Here, reimport CART tree.

<<pmml-show, echo=TRUE, eval=FALSE>>=
<<pmml>>
@

\medskip

\textbf{evtree:} Evolutionary learning of globally optimal trees, directly
using partykit.

<<evtree-show, echo=TRUE, eval=FALSE>>=
<<evtree>>
@

\medskip

\textbf{CTree:} Conditional inference trees \fct{ctree} are reimplemented
more efficiently within partykit.

\medskip

\textbf{CHAID:} R package on R-Forge, directly using partykit.
(Alternatively, use SPSS and export via PMML.)

\end{frame}

\begin{frame}
\frametitle{partykit: PMML}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-pmml-visualization, fig=TRUE, height=8, width=12>>=
plot(pm)
@

\end{frame}

\begin{frame}
\frametitle{partykit: evtree}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-evtree-visualization, fig=TRUE, height=8, width=12>>=
plot(ev)
@

\end{frame}

\subsection{Future: Next steps}

\begin{frame}
\frametitle{Next steps}

\textbf{To do:} The current code is already fairly well-tested and mostly
stable. However, there are some important items on the task list.
\begin{itemize}
  \item Extend/smooth package vignette.
  \item Add \code{xtrafo}/\code{ytrafo} to \fct{ctree} reimplementation.
  \item Switch \fct{mob} to new \class{party} class.
  \item Create new subclass \class{modelparty} that facilitates
    handling of formulas, terms, model frames, etc.\ for model-based trees.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Next steps}

\textbf{Model-based recursive partitioning:} Trees with parametric models in each node
(e.g., based on least squares or maximum likelihood). Splitting based on parameter
instability tests.

\medskip

\textbf{Illustration:} Logistic regression (or logit model), assessing differences in
the effect of ``preferential treatment'' for women or children.

<<mob-show, echo=TRUE, eval=FALSE>>=
<<mob>>
@

Odds ratio of survival given treatment differs across subsets (slope), as does
the survival probability of male adults (intercept).

<<mob-coef, echo=TRUE, results=verbatim>>=
coef(mb)
@

\end{frame}

\begin{frame}
\frametitle{Next steps}

\vspace*{-0.3cm}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-mob-visualization, fig=TRUE, height=7, width=12>>=
plot(mb, tp_args = list(margins = c(1.5, 2, 1.5, 2.5)))
@

\end{frame}

\subsection{Computational details}

\begin{frame}[fragile]
\frametitle{Computational details}

\textbf{Software:} All examples have been produced with
R~\Sexpr{with(R.Version(), paste(major, minor, sep = "."))} and packages
partykit~\Sexpr{packageDescription("partykit")$Version},
rpart~\Sexpr{packageDescription("rpart")$Version},
RWeka~\Sexpr{packageDescription("RWeka")$Version},
evtree~\Sexpr{packageDescription("evtree")$Version}. All packages are
freely available under the GPL from \url{http://CRAN.R-project.org/}.

\bigskip

\textbf{Data:} The contingency table \code{Titanic} is transformed
to the data frame \code{ttnc} using the following code. 
%
<<data1, echo=TRUE, eval=FALSE>>=
<<data>>
@
%
The last transformation also adds the treatment variable for the model-based recursive
partitioning analysis.

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\small

Hothorn T, Zeileis A (2013).
 \textit{partykit: A Toolkit for Recursive Partytioning.}
 R~package vignette version~0.1-6.
 URL~\url{http://CRAN.R-project.org/package=partykit}

\bigskip

Hothorn T, Hornik K, Zeileis A (2006).
 \dquote{Unbiased Recursive Partitioning: A Conditional Inference Framework.}
 \textit{Journal of Computational and Graphical Statistics},
 \textbf{15}(3), 651--674.
 \doi{10.1198/106186006X133933}

\bigskip

Grubinger T, Zeileis A, Pfeiffer KP (2011).
 \dquote{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R.}
 \textit{Working Paper~2011-20}, Working Papers in Economics and Statistics,
 Research Platform Empirical and Experimental Economics, Universit\"at Innsbruck.
 URL~\url{http://EconPapers.RePEc.org/RePEc:inn:wpaper:2011-20}

\bigskip

Zeileis A, Hothorn T, Hornik K (2008).
 \dquote{Model-Based Recursive Partitioning.}
  \textit{Journal of Computational and Graphical Statistics},
  \textbf{17}(2), 492--514.
  \doi{10.1198/106186008X319331}

\end{frame}

\end{document}

### 2014-01-06 pkg/partykit/inst/ERCIM-2011/titanic.R

@@ -1,0 +1,60 @@
+## packages
+library("partykit")
+library("rpart")
+library("RWeka")
+library("CHAID")
+library("evtree")
+
+## data
+data("Titanic", package = "datasets")
+ttnc <- as.data.frame(Titanic)
+ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
+names(ttnc)[2] <- "Gender"
+ttnc <- transform(ttnc, First = factor(Gender == "Female" | Age == "Child",
+  levels = c(FALSE, TRUE), labels = c("Male&Adult", "Female|Child")))
+
+## rpart
+rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
+plot(as.party(rp))
+
+## ctree
+ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
+  control = ctree_control(mincrit = 0.99, teststat = "max"))
+plot(ct)
+
+## J4.8
+j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
+plot(as.party(j48))
+
+## CHAID
+chd <- chaid(Survived ~ Gender + Age + Class, data = ttnc)
+plot(chd)
+
+## evtree
+set.seed(1071)
+ev <- evtree(Survived ~ Gender + Age + Class, data = ttnc, maxdepth = 3)
+plot(ev)
+
+## performance
+mc <- function(obj) 1 - mean(predict(obj) == ttnc$Survived)
+evalfun <- function(obj) 2 * nrow(ttnc) * mc(obj) + width(obj) * log(nrow(ttnc))
+trees <- list(evtree = ev, rpart = as.party(rp), ctree = ct, J48 = as.party(j48))
+round(sapply(trees, function(obj) c(
+  "complexity" = width(obj),
+  "misclassification" = mc(obj),
+  "evaluation function" = evalfun(obj))), digits = 3)
+
+## compare predictions
+nd <- ttnc[rep(1:nrow(ttnc), 100), ]
+system.time(p1_rp <- predict(rp, newdata = nd, type = "class"))
+system.time(p2_rp <- predict(as.party(rp), newdata = nd))
+table(p1_rp, p2_rp)
+
+system.time(p1_j48 <- predict(j48, newdata = nd))
+system.time(p2_j48 <- predict(as.party(j48), newdata = nd))
+table(p1_j48, p2_j48)
+
+## mob
[…]

### 2014-01-16 pkg/partykit/R/modelparty.R
enabled multiway splits for categorical partitioning variables in mob()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@415 255f4249-2263-4417-b1ea-8498779e27a7

@@ -326,6 +326,10 @@ mob <- function(formula, data, subset, na.action, weights, offset,
 
     } else {
 
+      if(!is.ordered(zselect) & control$catsplit == "multiway") {
+        return(list(breaks = NULL, index = seq_along(levels(zselect))))      
+      }
+
       ## for categorical variables
       al <- mob_grow_getlevels(zselect)
       dev <- apply(al, 1L, function(w) {
@@ -537,7 +541,8 @@ mob_grow_getlevels <- function(z) {
 mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsplit = NULL, trim = 0.1,
   breakties = FALSE, parm = NULL, verbose = FALSE, caseweights = TRUE,
   ytype = "vector", xtype = "matrix", terminal = "object", inner = terminal,
-  model = TRUE, numsplit = "left", vcov = "opg", ordinal = "chisq", nrep = 10000)
+  model = TRUE, numsplit = "left", catsplit = "binary", vcov = "opg",
+  ordinal = "chisq", nrep = 10000)
 {
   ytype <- match.arg(ytype, c("vector", "data.frame", "matrix"))
   xtype <- match.arg(xtype, c("data.frame", "matrix"))
@@ -547,6 +552,7 @@ mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsplit = NULL, trim =
 
   numsplit <- match.arg(tolower(numsplit), c("left", "center", "centre"))
   if(numsplit == "centre") numsplit <- "center"
+  catsplit <- match.arg(tolower(catsplit), c("binary", "multiway"))
   vcov <- match.arg(tolower(vcov), c("opg", "info"))
   ordinal <- match.arg(tolower(ordinal), c("l2", "max", "chisq"))
 
@@ -555,7 +561,8 @@ mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsplit = NULL, trim =
     breakties = breakties, parm = parm, verbose = verbose,
     caseweights = caseweights, ytype = ytype, xtype = xtype,
     terminal = terminal, inner = inner, model = model,
-    numsplit = numsplit, vcov = vcov, ordinal = ordinal, nrep = nrep)
+    numsplit = numsplit, catsplit = catsplit, vcov = vcov,
+    ordinal = ordinal, nrep = nrep)
   return(rval)
 }
 

### 2014-01-16 pkg/partykit/inst/ERCIM-2011/titanic.R

@@ -55,6 +55,10 @@ system.time(p2_j48 <- predict(as.party(j48), newdata = nd))
 table(p1_j48, p2_j48)
 
 ## mob
-mb <- glmtree(Survived ~ First | Class, data = ttnc, family = binomial, alpha = 0.005)
+mb <- glmtree(Survived ~ First | Class + Gender + Age, data = ttnc,
+  family = binomial, alpha = 0.01)
 plot(mb)
 
+mb2 <- glmtree(Survived ~ First | Class + Gender + Age, data = ttnc,
+  family = binomial, alpha = 0.01, catsplit = "multiway")
+plot(mb2)

### 2014-01-20 pkg/partykit/R/party.R
check factor levels of newdata
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@423 255f4249-2263-4417-b1ea-8498779e27a7

@@ -21,9 +21,10 @@ party <- function(node, data, fitted = NULL, terms = NULL, names = NULL, info = 
         stopifnot(inherits(fitted, "data.frame"))
         stopifnot(nrow(data) == 0L | nrow(data) == nrow(fitted))
         
-	#Z# try to provide default (fitted) ?
+	# try to provide default variable "(fitted)"
 	if(nrow(data) > 0L) {
-          if(!("(fitted)" %in% names(fitted))) fitted[["(fitted)"]] <- fitted_node(node, data = data)
+          if(!("(fitted)" %in% names(fitted))) 
+            fitted[["(fitted)"]] <- fitted_node(node, data = data)
 	} else {
 	  stopifnot("(fitted)" == names(fitted)[1L])
 	}
@@ -36,9 +37,10 @@ party <- function(node, data, fitted = NULL, terms = NULL, names = NULL, info = 
         fitted[["(fitted)"]] <- nt2[match(fitted[["(fitted)"]], nt)]
     } else {
         node <- as.partynode(node, from = 1L)
-	#Z# new default (fitted) ?
-	if(nrow(data) > 0L & missing(fitted)) fitted <- data.frame(
-	  "(fitted)" = fitted_node(node, data = data), check.names = FALSE)
+	# default "(fitted)"
+	if(nrow(data) > 0L & missing(fitted)) 
+          fitted <- data.frame("(fitted)" = fitted_node(node, 
+            data = data), check.names = FALSE)
     }
     
     party <- list(node = node, data = data, fitted = fitted, 
@@ -264,8 +266,13 @@ predict.party <- function(object, newdata = NULL, ...)
         vclass <- structure(lapply(object$data, class), .Names = vnames)
         ndnames <- names(newdata)
         ndclass <- structure(lapply(newdata, class), .Names = ndnames)
-        if(all(unames %in% ndnames) &&
-           all(unlist(lapply(unames, function(x) vclass[[x]] == ndclass[[x]])))) {
+        checkclass <- all(sapply(unames, function(x) 
+          isTRUE(all.equal(vclass[[x]], ndclass[[x]]))))
+        factors <- sapply(unames, function(x) inherits(object$data[[x]], "factor"))
+        checkfactors <- all(sapply(unames[factors], function(x) 
+          isTRUE(all.equal(levels(object$data[[x]]), levels(newdata[[x]])))))
+        ## FIXME: inform about wrong classes / factor levels?
[…]

### 2014-01-21 pkg/partykit/R/party.R
type = prob returns list of ecdf objects for numeric responses; new node_ecdf panel function
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@429 255f4249-2263-4417-b1ea-8498779e27a7

@@ -355,7 +355,7 @@ predict_party.constparty <- function(party, id, newdata = NULL,
 
 .pred_Surv_response <- function(y, w)
     .median_survival_time(.pred_Surv(y, w))
-                    
+ 
 .pred_factor <- function(y, w) {
     lev <- levels(y)
     sumw <- tapply(w, y, sum)
@@ -373,7 +373,17 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     return(prob) 
 }
                     
-.pred_numeric <- function(y, w) weighted.mean(y, w, na.rm = TRUE)
+.pred_numeric_response <- function(y, w) weighted.mean(y, w, na.rm = TRUE)
+
+.pred_ecdf <- function(y, w) {
+    iw <- as.integer(round(w))
+    if (max(abs(w - iw)) < sqrt(.Machine$double.eps)) {
+        y <- rep(y, w)
+        return(ecdf(y))
+    } else {
+        stop("cannot compute empirical distribution function with non-integer weights")
+    }
+}
 
 ### workhorse: compute predictions based on fitted / response data
 .predict_party_constparty <- function(node, fitted, response, weights,
@@ -388,11 +398,7 @@ predict_party.constparty <- function(party, id, newdata = NULL,
         FUN <- switch(rtype,
             "Surv" = if (type == "response") .pred_Surv_response else .pred_Surv,
             "factor" = if (type == "response") .pred_factor_response else .pred_factor,
-            "numeric" = {
-                if (type == "prob")
-                    stop(sQuote("type = \"prob\""), " ", "is not available")
-                .pred_numeric
-           })
+            "numeric" = if (type == "response") .pred_numeric_response else .pred_ecdf)
     }
       
     ## empirical distribution in each leaf

### 2014-01-21 pkg/partykit/R/plot.R

@@ -753,6 +753,113 @@ node_surv <- function(obj, col = "black", ylines = 2,
 }
 class(node_surv) <- "grapcon_generator"
 
+node_ecdf <- function(obj, col = "black", ylines = 2,
+		      id = TRUE, mainlab = NULL, gp = gpar(), ...)
+{
+    ## extract response
+    y <- obj$fitted[["(response)"]]
+    stopifnot(inherits(y, "numeric") || inherits(y, "integer"))
+
+    dostep <- function(f) {
+        x <- knots(f)
+        y <- f(x)
+        ### create a step function based on x, y coordinates
+        ### modified from `survival:print.survfit'
+        if (is.na(x[1] + y[1])) {
+            x <- x[-1]
+            y <- y[-1]
+        }
+        n <- length(x)
+        if (n > 2) {  
+            # replace verbose horizonal sequences like
+            # (1, .2), (1.4, .2), (1.8, .2), (2.3, .2), (2.9, .2), (3, .1)
+            # with (1, .2), (3, .1).  They are slow, and can smear the looks
+            # of the line type.
+            dupy <- c(TRUE, diff(y[-n]) !=0, TRUE)
+            n2 <- sum(dupy)
+
+            #create a step function
+            xrep <- rep(x[dupy], c(1, rep(2, n2-1)))
+            yrep <- rep(y[dupy], c(rep(2, n2-1), 1))
+            RET <- list(x = xrep, y = yrep)
+        } else {
+            if (n == 1) {
+                RET <- list(x = x, y = y)
+            } else {
+                RET <- list(x = x[c(1,2,2)], y = y[c(1,1,2)])
+            }
+        }
+        return(RET)
+    }
+
+    ### panel function for ecdf in nodes
+    rval <- function(node) {
+
+        ## extract data
+	nid <- id_node(node)
+	dat <- data_party(obj, nid)
+	yn <- dat[["(response)"]]
+	wn <- dat[["(weights)"]]
+	if(is.null(wn)) wn <- rep(1, NROW(yn))
+
+        ## get ecdf in node
+        f <- .pred_ecdf(yn, wn)
+        a <- dostep(f)
+
+        ## set up plot
+        yscale <- c(0, 1)
+        xscale <- range(y)
+        a$x <- c(xscale[1], a$x[1], a$x, xscale[2])
+        a$x <- a$x - min(a$x)
+        a$x <- a$x / max(a$x)
+        a$y <- c(0, 0, a$y, 1)
+
[…]

### 2014-01-21 pkg/partykit/R/print.R

@@ -111,7 +111,7 @@ print.constparty <- function(x,
   
   if(is.null(FUN)) FUN <- switch(yclass,
     "numeric" = function(y, w, digits) {
-      yhat <- .pred_numeric(y, w)
+      yhat <- .pred_numeric_response(y, w)
       yerr <- sum(w * (y - yhat)^2)
       digits2 <- max(c(0, digits - 2))
       paste(format(round(yhat, digits = digits), nsmall = digits),

### 2014-01-21 pkg/partykit/R/simpleparty.R

@@ -144,7 +144,7 @@ as.simpleparty.constparty <- function(obj, ...) {
     
     switch(rtype,
       "numeric" = {
-        yhat <- .pred_numeric(y, w)
+        yhat <- .pred_numeric_response(y, w)
         list(prediction = yhat, n = structure(sum(w), .Names = wnam),
 	  error = sum(w * (y - yhat)^2), distribution = NULL)
       },

### 2014-01-21 pkg/partykit/R/party.R
cosmetics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@430 255f4249-2263-4417-b1ea-8498779e27a7

@@ -373,7 +373,8 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     return(prob) 
 }
                     
-.pred_numeric_response <- function(y, w) weighted.mean(y, w, na.rm = TRUE)
+.pred_numeric_response <- function(y, w) 
+    weighted.mean(y, w, na.rm = TRUE)
 
 .pred_ecdf <- function(y, w) {
     iw <- as.integer(round(w))

### 2014-01-23 pkg/partykit/inst/ERCIM-2011/titanic.R
added PMML example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@440 255f4249-2263-4417-b1ea-8498779e27a7

@@ -26,6 +26,10 @@ plot(ct)
 j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
 plot(as.party(j48))
 
+## QUEST
+qu <- pmmlTreeModel(file.path(system.file("pmml", package = "partykit"), "ttnc.pmml"))
+plot(qu)
+
 ## CHAID
 chd <- chaid(Survived ~ Gender + Age + Class, data = ttnc)
 plot(chd)

### 2014-02-02 pkg/partykit/R/modelparty.R
added maxdepth argument to mob_control and mob renamed minsplit to minsize because it really corresponds more to minbucket but is somewhat different...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@442 255f4249-2263-4417-b1ea-8498779e27a7

@@ -5,8 +5,8 @@ mob <- function(formula, data, subset, na.action, weights, offset,
   stopifnot(require("Formula"))
   
   ## control parameters (used repeatedly)
-  minsplit <- control$minsplit
-  if(!is.null(minsplit) && !is.integer(minsplit)) minsplit <- as.integer(minsplit)
+  minsize <- control$minsize
+  if(!is.null(minsize) && !is.integer(minsize)) minsize <- as.integer(minsize)
   verbose <- control$verbose
   ytype <- control$ytype
   xtype <- control$xtype
@@ -163,7 +163,7 @@ mob <- function(formula, data, subset, na.action, weights, offset,
 
     ## get critical values for supLM statistic
     from <- if(control$trim > 1) control$trim else ceiling(n * control$trim)
-    from <- max(from, minsplit)
+    from <- max(from, minsize)
     to <- n - from
     lambda <- ((n - from) * to)/(from * (n - to))
 
@@ -280,9 +280,9 @@ mob <- function(formula, data, subset, na.action, weights, offset,
   ### split in variable zselect, either ordered (numeric or ordinal) or nominal
   mob_grow_findsplit <- function(y, x, zselect, weights, offset, ...)
   {
-    ## process minsplit (to minimal number of observations)
-    if(minsplit > 0.5 & minsplit < 1) minsplit <- 1 - minsplit
-    if(minsplit < 0.5) minsplit <- ceiling(w2n(weights) * minsplit)
+    ## process minsize (to minimal number of observations)
+    if(minsize > 0.5 & minsize < 1) minsize <- 1 - minsize
+    if(minsize < 0.5) minsize <- ceiling(w2n(weights) * minsize)
   
     if(is.numeric(zselect)) {
     ## for numerical variables
@@ -294,7 +294,7 @@ mob <- function(formula, data, subset, na.action, weights, offset,
         zs <- zselect <= uz[i]
         start_left <- NULL
         start_right <- NULL
-        if(w2n(weights[zs]) < minsplit || w2n(weights[!zs]) < minsplit) {
+        if(w2n(weights[zs]) < minsize || w2n(weights[!zs]) < minsize) {
           dev[i] <- Inf
         } else {
           fit_left <- afit(y = suby(y, zs), x = subx(x, zs), start = start_left,
[…]

### 2014-02-02 pkg/partykit/R/lmtree.R
AIC/BIC-based post-pruning for mob()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@444 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,11 +1,31 @@
 ## simple wrapper function to specify fitter and return class
 lmtree <- function(formula, data, subset, na.action, weights, offset, ...)
 {
+  ## TODO: variance as model parameter
+
   ## use dots for setting up mob_control
   control <- mob_control(...)
   if(control$vcov == "info") {
     warning('vcov = "info" not supported in lmtree')
     control$vcov <- "opg"
+  }
+  if(!is.null(control$prune)) {
+    if(is.character(control$prune)) {
+      control$prune <- tolower(control$prune)
+      control$prune <- match.arg(control$prune, c("aic", "bic", "none"))
+      control$prune <- switch(control$prune,
+        "aic" = {
+	  function(objfun, df, nobs) (nobs[1L] * log(objfun[1L]) + 2 * df[1L]) < (nobs[1L] * log(objfun[2L]) + 2 * df[2L])
+	}, "bic" = {
+	  function(objfun, df, nobs) (nobs[1L] * log(objfun[1L]) + log(nobs[2L]) * df[1L]) < (nobs[1L] * log(objfun[2L]) + log(nobs[2L]) * df[2L])
+	}, "none" = {
+	  NULL
+	})      
+    }
+    if(!is.function(control$prune)) {
+      warning("unknown specification of 'prune'")
+      control$prune <- NULL
+    }
   }
 
   ## keep call

### 2014-02-02 pkg/partykit/R/modelparty.R

@@ -128,6 +128,26 @@ mob <- function(formula, data, subset, na.action, weights, offset,
     }
   }
 
+  ## process pruning options
+  if(!is.null(control$prune)) {
+    if(is.character(control$prune)) {
+      control$prune <- tolower(control$prune)
+      control$prune <- match.arg(control$prune, c("aic", "bic", "none"))
+      control$prune <- switch(control$prune,
+        "aic" = {
+	  function(objfun, df, nobs) (2 * objfun[1L] + 2 * df[1L]) < (2 * objfun[2L] + 2 * df[2L])
+	}, "bic" = {
+	  function(objfun, df, nobs) (2 * objfun[1L] + log(n) * df[1L]) < (2 * objfun[2L] + log(n) * df[2L])
+	}, "none" = {
+	  NULL
+	})      
+    }
+    if(!is.function(control$prune)) {
+      warning("unknown specification of 'prune'")
+      control$prune <- NULL
+    }
+  }
+
   ## core mob_grow_* functions
 
   ## variable selection: given model scores, conduct
@@ -398,12 +418,10 @@ mob <- function(formula, data, subset, na.action, weights, offset,
       TERMINAL <- TRUE
     }
     if(depth >= control$maxdepth) {
-      if(verbose) cat(sprintf("Maximum deptch reached, stop splitting (maxdepth = %i)\n\n", maxdepth))
+      if(verbose) cat(sprintf("Maximum depth reached, stop splitting (maxdepth = %i)\n\n", control$maxdepth))
       TERMINAL <- TRUE
     }
     if(TERMINAL) {
-      if(!terminal$estfun) mod$estfun <- NULL
-      if(!terminal$object) mod$object <- NULL
       return(partynode(id = id, info = mod))
     }
 
@@ -434,8 +452,6 @@ mob <- function(formula, data, subset, na.action, weights, offset,
     }
     
     ## update model information
-    if(!terminal$estfun) mod$estfun <- NULL
-    if(!terminal$object) mod$object <- NULL
     mod$test <- rbind("statistic" = test$stat, "p.value" = test$pval)
 
     if(TERMINAL) {
@@ -485,9 +501,77 @@ mob <- function(formula, data, subset, na.action, weights, offset,
     return(partynode(id = id, split = sp, kids = kids, info = mod))
   }
 
+  mob_prune <- function(node)
+  {
[…]

### 2014-02-03 pkg/partykit/R/modelparty.R
renamed "splits" argument to "dfsplit" because this is clearer and now integer arguments (e.g., dfsplit = 3) are also supported
added rather detailed description of pruning strategy in mob.Rnw



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@446 255f4249-2263-4417-b1ea-8498779e27a7

@@ -530,7 +530,7 @@ mob <- function(formula, data, subset, na.action, weights, offset,
       objfun <- sapply(nd, function(x) x$info$objfun)
       pok <- sapply(pnode, function(i) control$prune(
         objfun = c(objfun[i], sum(objfun[kids[[i]]])),
-	df = c(length(nd[[1]]$info$coefficients), length(kids[[i]]) * length(nd[[1]]$info$coefficients) + as.integer(control$splits)),
+	df = c(length(nd[[1]]$info$coefficients), length(kids[[i]]) * length(nd[[1]]$info$coefficients) + as.integer(control$dfsplit)),
         nobs = c(nd[[i]]$info$nobs, n)
       ))
 
@@ -639,7 +639,7 @@ mob_grow_getlevels <- function(z) {
 
 ## control splitting parameters
 mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsize = NULL, maxdepth = Inf,
-  trim = 0.1, breakties = FALSE, parm = NULL, splits = TRUE, prune = NULL, verbose = FALSE,
+  trim = 0.1, breakties = FALSE, parm = NULL, dfsplit = TRUE, prune = NULL, verbose = FALSE,
   caseweights = TRUE, ytype = "vector", xtype = "matrix", terminal = "object", inner = terminal,
   model = TRUE, numsplit = "left", catsplit = "binary", vcov = "opg",
   ordinal = "chisq", nrep = 10000)
@@ -658,7 +658,7 @@ mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsize = NULL, maxdept
 
   rval <- list(alpha = alpha, bonferroni = bonferroni, minsize = minsize, maxdepth = maxdepth,
     trim = ifelse(is.null(trim), minsize, trim),
-    breakties = breakties, parm = parm, splits = TRUE, prune = prune,
+    breakties = breakties, parm = parm, dfsplit = dfsplit, prune = prune,
     verbose = verbose, caseweights = caseweights, ytype = ytype, xtype = xtype,
     terminal = terminal, inner = inner, model = model,
     numsplit = numsplit, catsplit = catsplit, vcov = vcov,
@@ -785,15 +785,16 @@ apply_to_models <- function(object, node = NULL, FUN = NULL, drop = FALSE, ...) 
   return(rval)
 }
 
-logLik.modelparty <- function(object, splits = NULL, ...)
[…]

### 2014-02-07 pkg/partykit/inst/Psychoco-2014/slides.Rnw
slides for Psychoco 2014
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@451 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[11pt,compress,t]{beamer}
\usetheme{Z}
\usepackage{amsfonts,amstext,amsmath}
%% need no \usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}

\graphicspath{{images/}}

%% commands
\newcommand{\ui}{\underline{i}}
\newcommand{\oi}{\overline{\imath}}
\newcommand{\darrow}{\stackrel{\mbox{\tiny \textnormal{d}}}{\longrightarrow}}
\newcommand{\parrow}{\stackrel{\mbox{\tiny \textnormal{p}}}{\longrightarrow}}
\newcommand{\dotequals}{\stackrel{\cdot}{=}}
\newcommand{\efp}{\mathrm{\it efp}}
\newcommand{\given}{\, | \,}
\newcommand{\ltime}{\lambda_\mathrm{time}}
\newcommand{\lcomp}{\lambda_\mathrm{comp}}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}

<<packages, echo=FALSE, results=hide>>=
library("partykit")
library("psychotools")
data("Topmodel2007", package = "psychotree")
options(digits = 4)
@

<<btfit2, echo=FALSE, results=hide>>=
btfit2 <- function(y, x = NULL, start = NULL, weights = NULL,
  offset = NULL, ..., estfun = FALSE, object = FALSE) {
  rval <- btReg.fit(y, weights = weights, ...,
    estfun = estfun, vcov = object)
  list(
    coefficients = rval$coefficients,
    objfun = -rval$loglik,
    estfun = if(estfun) rval$estfun else NULL,
    object = if(object) rval else NULL
  )
}
@

<<estfun.btReg, echo=FALSE, results=hide>>=
estfun.btReg <- function(x, ...) x$estfun
@

<<panelfunction, echo=FALSE, results=hide>>=
## visualization function
node_btplot <- function(mobobj, id = TRUE,
  worth = TRUE, names = TRUE, abbreviate = TRUE, index = TRUE, ref = TRUE,
  col = "black", linecol = "lightgray", cex = 0.5, pch = 19, xscale = NULL, yscale = NULL, ylines = 1.5)
{
    ## node ids
    node <- nodeids(mobobj, terminal = FALSE)

    ## get all coefficients 
    cf <- partykit:::apply_to_models(mobobj, node, FUN = function(z)        
      if(worth) worth(z) else coef(z, all = FALSE, ref = TRUE))
    cf <- do.call("rbind", cf)
    rownames(cf) <- node

    ## get one full model
    mod <- partykit:::apply_to_models(mobobj, node = 1L, FUN = NULL)

    if(!worth) {
      if(is.character(ref) | is.numeric(ref)) {
        reflab <- ref
        ref <- TRUE
      } else {
        reflab <- mod$ref
      }
      if(is.character(reflab)) reflab <- match(reflab, mod$labels)
      cf <- cf - cf[,reflab]
    }

    ## reference
    if(worth) {
      cf_ref <- 1/ncol(cf)
    } else {
      cf_ref <- 0
    }

    ## labeling
    if(is.character(names)) {
      colnames(cf) <- names
      names <- TRUE
    }

    ## abbreviation
    if(is.logical(abbreviate)) {
      nlab <- max(nchar(colnames(cf)))
      abbreviate <- if(abbreviate) as.numeric(cut(nlab, c(-Inf, 1.5, 4.5, 7.5, Inf))) else nlab
    }
    colnames(cf) <- abbreviate(colnames(cf), abbreviate)

    if(index) {
      x <- 1:NCOL(cf)
      if(is.null(xscale)) xscale <- range(x) + c(-0.1, 0.1) * diff(range(x))
    } else {
      x <- rep(0, length(cf))
      if(is.null(xscale)) xscale <- c(-1, 1)      
    }
    if(is.null(yscale)) yscale <- range(cf) + c(-0.1, 0.1) * diff(range(cf))

    ## panel function for bt plots in nodes
    rval <- function(node) {

      ## node index
      id <- id_node(node)

      ## dependent variable setup
      cfi <- cf[id,]

      ## viewport setup
      top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3,
    			 widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")),  
        		 heights = unit(c(1, 1), c("lines", "null"))),
    			 width = unit(1, "npc"), 
    			 height = unit(1, "npc") - unit(2, "lines"),
        		 name = paste("node_btplot", id, sep = ""))
      pushViewport(top_vp)
      grid.rect(gp = gpar(fill = "white", col = 0))

      ## main title
      top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
      pushViewport(top)
      mainlab <- paste(ifelse(id, paste("Node", id, "(n = "), ""),
        	       info_node(node)$nobs, ifelse(id, ")", ""), sep = "")
      grid.text(mainlab)
      popViewport()

      ## actual plot  
      plot_vpi <- viewport(layout.pos.col = 2, layout.pos.row = 2,
        xscale = xscale, yscale = yscale, 
        name = paste("node_btplot", id, "plot", sep = ""))
      pushViewport(plot_vpi)

      grid.lines(xscale, c(cf_ref, cf_ref), gp = gpar(col = linecol), default.units = "native")
      if(index) {
        grid.lines(x, cfi, gp = gpar(col = col, lty = 2), default.units = "native")
        grid.points(x, cfi, gp = gpar(col = col, cex = cex), pch = pch, default.units = "native")
        grid.xaxis(at = x, label = if(names) names(cfi) else x)
      } else {  	
        if(names) grid.text(names(cfi), x = x, y = cfi, default.units = "native")
          else grid.points(x, cfi, gp = gpar(col = col, cex = cex), pch = pch, default.units = "native")
      }
      grid.yaxis(at = c(ceiling(yscale[1] * 100)/100, floor(yscale[2] * 100)/100))
      grid.rect(gp = gpar(fill = "transparent"))

      upViewport(2)
    }

    return(rval)
}
class(node_btplot) <- "grapcon_generator"
@

\begin{document}

\title{Parties, Models, Mobsters\newline
{\fontsize{11}{15}\selectfont A New Implementation of Model-Based Recursive Partitioning in R}}
\author{Achim Zeileis, Torsten Hothorn}
\URL{http://eeecon.uibk.ac.at/~zeileis/}
\lecture{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R}

\subsection{Overview}

\begin{frame}
\frametitle{Overview}

\begin{itemize}
  \item Model-based recursive partitioning
  \begin{itemize}
    \item A generic approach
    \item Example: Bradley-Terry trees
  \end{itemize}
  \item Implementation in R
  \begin{itemize}
    \item Building blocks: Parties, models, mobsters
    \item Old implementation in \emph{party}
    \item All new implementation in \emph{partykit}
  \end{itemize}
  \item Illustration
\end{itemize}

\end{frame}

\subsection{Model-based recursive partitioning}

\begin{frame}
\frametitle{Model-based recursive partitioning}

\textbf{Models:} Estimation of parametric models with
observations $y_i$ (and regressors $x_i$),
parameter vector $\theta$, and additive objective function $\Psi$.
\[ \widehat \theta \quad = \quad \argmin_{\theta}
   \sum_i \Psi(y_i, x_i, \theta). \]

\bigskip

\textbf{Recursive partitioning}:

\begin{enumerate}
  \item Fit the model in the current subsample.
  \item Assess the stability of $\theta$ across each partitioning variable $z_j$.
  \item Split sample along the $z_{j^*}$ with strongest association:
        Choose breakpoint with highest improvement of the model fit.
  \item Repeat steps 1--3 recursively in the subsamples
        until some stopping criterion is met.
\end{enumerate}

\end{frame}

\begin{frame}
\frametitle{Model-based recursive partitioning}

\textbf{Parameter instability tests:}
\begin{itemize}
  \item Based on empirical estimating functions (or score/gradient contributions):
    $\Psi'(y_i, x_i, \hat \theta)$.
  \item Under parameter stability: $\Psi'$ fluctuates randomly around its expectation zero.
  \item Under parameter instability: Systematic departures from zero in subsamples.
  \item Hence fluctuation can be captured across numeric partitioning variables
    or within levels of categorical partitioning variables.
  \item Bonferroni correction for testing across multiple partitioning variables.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Bradley-Terry trees}

\hspace*{0.2cm}%
\includegraphics[width=0.23\textwidth]{Barbara.jpg} \hspace*{0.1cm}
\includegraphics[width=0.23\textwidth]{Anni.jpg} \hspace*{0.1cm}
\includegraphics[width=0.23\textwidth]{Hana.jpg} \hspace*{0.1cm}
\includegraphics[width=0.23\textwidth]{Fiona.jpg}

\vspace*{0.1cm}
\hspace*{0.2cm}%
\includegraphics[width=0.23\textwidth]{Mandy.jpg} \hspace*{0.1cm}
\includegraphics[width=0.23\textwidth]{Anja.jpg} \hspace*{0.2cm}%
\begin{tabular}{p{0.48\textwidth}}
\textbf{Questions:} Which of these

women is more attractive?

How does the answer depend on

age, gender, and the familiarity

with the associated TV show

Germany's Next Topmodel?
\vspace*{3.4cm}
\end{tabular}

\end{frame}

\begin{frame}
\frametitle{Bradley-Terry trees}

<<topmodel-tree, echo=FALSE, results=hide>>=
bt0 <- mob(preference ~ 1 | gender + age + q1 + q2 + q3,
  data = Topmodel2007, fit = btfit2)
@

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.15\textwidth}
\hspace*{-0.7cm}%
<<topmodel-plot, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
plot(bt0, drop = TRUE, tnex = 2,
  terminal_panel = node_btplot(bt0, abbreviate = 1, yscale = c(0, 0.5)))
@

\end{frame}

\subsection{Implementation in R}

\begin{frame}
\frametitle{Implementation: Building blocks}

\textbf{Workhorse function:} \code{mob()} for
\begin{itemize}
  \item data handling,
  \item calling model fitters,
  \item carrying out parameter instability tests and
  \item recursive partitioning algorithm.
\end{itemize}

\bigskip

\textbf{Required functionality:}
\begin{itemize}
  \item \emph{Parties:} Class and methods for representing recursive partytions.
  \item \emph{Models:} Fitting functions for statistical models (optimizing suitable objective function).
  \item \emph{Mobsters:} High-level interfaces (\code{lmtree()}, \code{bttree()}, \dots)
    that call lower-level \code{mob()} with suitable options and methods.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Implementation: Old \texttt{mob()} in \emph{party}}

\textbf{Parties:} S4 class `\code{BinaryTree}'.
\begin{itemize}
  \item Originally developed only for \code{ctree()} and somewhat ``abused''.
  \item Rather rigid and hard to extend.
\end{itemize}

\medskip

\textbf{Models:} S4 `\code{StatModel}' objects.
\begin{itemize}
  \item Intended to conceptualize unfitted model objects.
  \item Required some ``glue code'' to accomodate non-standard
    interface for data handling and model fitting.
\end{itemize}

\medskip

\textbf{Mobsters:}
\begin{itemize}
  \item \code{mob()} already geared towards (generalized) linear models.
  \item Other interfaces in \emph{psychotree} and \emph{betareg}.
  \item Hard to do fine control due to adopted S4 classes: Many unnecessary
    computations and data storage.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Implementation: New \texttt{mob()} in \emph{partykit}}

\textbf{Parties:} S3 class `\code{modelparty}' built on `\code{party}'.
\begin{itemize}
  \item Separates data and tree structure.
  \item Inherits generic infrastructure for printing, predicting,
    plotting, \dots
\end{itemize}

\medskip

\textbf{Models:} Plain functions with input/output convention.
\begin{itemize}
  \item Basic and extended interface for rapid prototyping and for
    speeding up computings, respectively.
  \item Only minimial glue code required if models are well-designed.
\end{itemize}

\medskip

\textbf{Mobsters:}
\begin{itemize}
  \item \code{mob()} completely agnostic regarding models employed.
  \item Separate interfaces \code{lmtree()}, \code{glmtree()}, \dots
  \item New interfaces typically need to bring their model fitter and adapt the
    main methods \code{print()}, \code{plot()}, \code{predict()} etc.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Implementation: New \texttt{mob()} in \emph{partykit}}

\textbf{New inference options:} Not used by default by optionally available.
\begin{itemize}
  \item New parameter instability tests for ordinal partitioning variables.
    Alternative to unordered $\chi^2$ test but computationally intensive.
  \item Post-pruning based on information criteria (e.g., AIC or BIC), especially
    for very large datasets where traditional significance levels are not useful.
  \item Multiway splits for categorical partitioning variables.
  \item Treat weights as proportionality weights and not as case weights.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Implementation: Models}

\textbf{Input:} Basic interface.
\begin{Soutput}
  fit(y, x = NULL, start = NULL, weights = NULL,
    offset = NULL, ...)
\end{Soutput}
\code{y}, \code{x}, \code{weights}, \code{offset} are (the subset of)
the preprocessed data.

Starting values and further fitting arguments are in \code{start} and \code{...}.

\bigskip

\textbf{Output:} Fitted model object of class with suitable methods.
\begin{itemize}
  \item \code{coef()}: Estimated parameters $\hat \theta$.
  \item \code{logLik()}: Maximized log-likelihood function $-\sum_i \Psi(y_i, x_, \hat \theta)$.
  \item \code{estfun()}: Empirical estimating functions
    $\Psi'(y_i, x_i, \hat \theta)$.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Implementation: Models}

\textbf{Input:} Extended interface.
\begin{Soutput}
  fit(y, x = NULL, start = NULL, weights = NULL,
    offset = NULL, ..., estfun = FALSE, object = FALSE)
\end{Soutput}

\medskip

\textbf{Output:} List.
\begin{itemize}
  \item \code{coefficients}: Estimated parameters $\hat \theta$.
  \item \code{objfun}: Minimized objective function $\sum_i \Psi(y_i, x_, \hat \theta)$.
  \item \code{estfun}: Empirical estimating functions
    $\Psi'(y_i, x_i, \hat \theta)$. Only needed if \code{estfun = TRUE}, otherwise optionally \code{NULL}.
  \item \code{object}: A model object for which further methods could
    be available (e.g., \code{predict()}, or \code{fitted()}, etc.).
    Only needed if \code{object = TRUE}, otherwise optionally \code{NULL}.
\end{itemize}

\medskip

\textbf{Internally:} Extended interface constructed from basic interface if supplied.
Efficiency can be gained through extended approach.

\end{frame}

\subsection{Bradley-Terry trees}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

Data, packages, and \code{estfun()} method.
<<packages, eval=FALSE>>=
data("Topmodel2007", package = "psychotree")
library("partykit")
library("psychotools")
<<estfun.btReg>>
@

\medskip

Basic model fitting function:

<<btfit1>>=
btfit1 <- function(y, x = NULL, start = NULL, weights = NULL,
  offset = NULL, ...) btReg.fit(y, weights = weights, ...)
@

\medskip

Fit Bradley-Terry tree:

<<bt1>>=
system.time(bt1 <- mob(
  preference ~ 1 | gender + age + q1 + q2 + q3,
  data = Topmodel2007, fit = btfit1))
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

Extended model fitting function:

<<btfit2-2>>=
<<btfit2>>
@

\medskip

Fit Bradley-Terry tree again:

<<bt2>>=
system.time(bt2 <- mob(
  preference ~ 1 | gender + age + q1 + q2 + q3,
  data = Topmodel2007, fit = btfit2))
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

<<bt2-print, echo=FALSE>>=
out <- capture.output(bt2)
writeLines(out[1:21])
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

<<bt2-print, echo=FALSE>>=
out <- capture.output(bt2)
writeLines(out[-(1:22)])
@

\medskip

Standard methods readily available:
<<bt2-coef>>=
plot(bt2)
coef(bt2)
@

\medskip

Customization:
<<bt2-worthf, eval=FALSE>>=
worthf <- function(info) paste(info$object$labels,
  format(round(worth(info$object), digits = 2)), sep = ": ")
plot(bt2, FUN = worthf)
@

\code{}

\end{frame}

\begin{frame}
\frametitle{Illustration: Bradley-Terry trees}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.15\textwidth}
\hspace*{-0.7cm}%
<<bt2-plot, fig=TRUE, height=7.4, width=12, echo=FALSE>>=
plot(bt2)
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

\vspace*{-0.5cm}

\setkeys{Gin}{width=1.15\textwidth}
\hspace*{-0.7cm}%
<<bt2-plot2, fig=TRUE, height=7.4, width=12, echo=FALSE>>=
<<bt2-worthf>>
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

\vspace*{-0.5cm}

\setkeys{Gin}{width=\textwidth}
<<bt2-nodeapply, fig=TRUE, height=7, width=9, echo=FALSE, results=hide>>=
par(mfrow = c(2, 2))
nodeapply(bt2, ids = c(3, 5, 6, 7), FUN = function(n)
  plot(n$info$object, main = n$id, ylim = c(0, 0.4)))
@

\end{frame}

\begin{frame}[fragile]
\frametitle{Illustration: Bradley-Terry trees}

Apply plotting in all terminal nodes:
<<nodeapply, eval=FALSE>>=
<<bt2-nodeapply>>
@

\medskip

<<tm, echo=FALSE>>=
tm <- data.frame(age = c(60, 25, 35), gender = c("male", "female", "female"),
  q1 = "no", q2 = c("no", "no", "yes"), q3 = "no")
@

\medskip

Predicted nodes and ranking:
<<predict>>=
tm
predict(bt2, tm, type = "node")
predict(bt2, tm, type = function(object) t(rank(-worth(object))))
@
%predict(bt2, tm, type = function(object) t(worth(object)))

\end{frame}

\subsection{Summary}

\begin{frame}
\frametitle{Summary}

\begin{itemize}
  \item All new implementation of model-based recursive partitioning in \emph{partykit}.
  \item Enables more efficient computations, rapid prototyping, customization.
  \item Some new inference options.
\end{itemize}

\end{frame}

\subsection{References}

\begin{frame}
\frametitle{References}

\small

Hothorn T, Zeileis A (2014).
 \textit{partykit: A Toolkit for Recursive Partytioning.}
 R~package version~0.2-0.
 URL~\url{http://R-Forge.R-project.org/projects/partykit/}

\bigskip

Zeileis A, Hothorn T (2014).
 \textit{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R.}
 \code{vignette("mob", package = "partykit")}.

\bigskip

Strobl C, Wickelmaier F, Zeileis A (2011).
  \dquote{Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.}
  \textit{Journal of Educational and Behavioral Statistics}, \textbf{36}(2), 135--153.
  \doi{10.3102/1076998609359791}

\bigskip

Zeileis A, Hothorn T, Hornik K (2008).
 \dquote{Model-Based Recursive Partitioning.}
  \textit{Journal of Computational and Graphical Statistics},
  \textbf{17}(2), 492--514.
  \doi{10.1198/106186008X319331}

\end{frame}

\end{document}

### 2014-02-07 pkg/partykit/inst/Psychoco-2014/slides.Rnw
some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@453 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \begin{itemize}
-  \item \emph{Parties:} Class and methods for representing recursive partytions.
+  \item \emph{Parties:} Class and methods for recursive partytions.
   \item \emph{Models:} Fitting functions for statistical models (optimizing suitable objective function).
@@
   \item Hard to do fine control due to adopted S4 classes: Many unnecessary
-    computations and data storage.
+    computations and copies of data.
 \end{itemize}
@@
 
-Data, packages, and \code{estfun()} method.
+Data, packages, and \code{estfun()} method:
 <<packages, eval=FALSE>>=
@@
   \item All new implementation of model-based recursive partitioning in \emph{partykit}.
-  \item Enables more efficient computations, rapid prototyping, customization.
+  \item Enables more efficient computations, rapid prototyping, flexible customization.
   \item Some new inference options.

### 2014-02-10 pkg/partykit/inst/NUS-2014/constparty_TH_AZ.Rnw
start constant party talk
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@454 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-01-06 pkg/partykit/inst/ERCIM-2011/slides.Rnw
@@
 \documentclass[11pt,compress,t]{beamer}
-\usetheme{Z}
-\usepackage{amsfonts,amstext,amsmath}
-%% need no \usepackage{Sweave}
+
+\usetheme{uzhneu-en-informal}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{amsfonts,amstext,amsmath,amssymb,amsthm}
+
+%% need no 
+\usepackage{Sweave}
 \definecolor{InputColor}{rgb}{0,0,0.3}
@@
 %% commands
+\let\code\texttt
+\let\class\texttt
+\let\fct\texttt
+\let\doi\texttt
+\newcommand{\dquote}[1]{``#1''}
 \newcommand{\ui}{\underline{i}}
@@
 library("partykit")
+library("Formula")
 library("rpart")
@@
 <<rpart, eval=FALSE>>=
-rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
+rp <- rpart(Survived ~ Gender + Age + Class, 
+            data = ttnc)
 @
@@
 set.seed(1071)
-ev <- evtree(Survived ~ Gender + Age + Class, data = ttnc,
-  maxdepth = 3)
+ev <- evtree(Survived ~ Gender + Age + Class, 
+             data = ttnc, maxdepth = 3)
 @
@@
 <<pmml, eval=FALSE>>=
-pm <- pmmlTreeModel("ttnc.xml")
+pm <- pmmlTreeModel(file.path(system.file("pmml", 
+    package = "partykit"), "ttnc.pmml"))
 @
@@
 <<J48>>=
-j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
+j48 <- J48(Survived ~ Gender + Age + Class, 
+           data = ttnc)
 @
@@
 
-\title{partykit: A Toolkit for Recursive Partytioning}
-\author{Achim Zeileis, Torsten Hothorn}
-\URL{http://eeecon.uibk.ac.at/~zeileis/}
-\lecture{partykit: A Toolkit for Recursive Partytioning}
+\title[Constant Partying]{Constant Partying: \\ 
+      Introducing a Generic Toolkit for Recursive Partitioning in R}
+\author{Torsten Hothorn (Universit\"at Z\"urich) \\
+        Achim Zeileis (Universit\"at Innsbruck)}
+%\URL{http://user.math.uzh.ch/hothorn/}
+%\lecture{partykit: A Toolkit for Recursive Partytioning}
 
-\subsection{Overview}
+\maketitle
 
@@
   \end{itemize}
+  \item Here and now: Intro to partykit for trees with constant fits in each node.
   \item Future: Next steps
@@
 \end{frame}
-
-\subsection{Status quo: R software for tree models}
 
@@
 \end{frame}
-
-\subsection{New package: partykit}
 
@@
 \frametitle{partykit: Interface to \texttt{rpart}}
-
+\small{
 <<rpart-print, echo=TRUE, results=verbatim>>=
@@
 @
+}
 
@@
 
+\small{
 <<rpart-print, echo=TRUE, results=verbatim>>=
@@
 \textbf{Prediction:} Compare rpart's C code and partykit's R code
-for (artificially) large data set.
 
@@
 nd <- ttnc[rep(1:nrow(ttnc), 100), ]
-system.time(p1 <- predict(rp, newdata = nd, type = "class"))
+system.time(p1 <- predict(rp, newdata = nd, 
+                          type = "class"))
 system.time(p2 <- predict(as.party(rp), newdata = nd))
@@
 exported by various software packages including SAS, SPSS, R/pmml.
-Here, reimport CART tree.
+Here, import QUEST tree.
 
@@
 \begin{frame}
-\frametitle{partykit: PMML}
+\frametitle{partykit: QUEST via PMML}
 
@@
 
-\subsection{Future: Next steps}
-
 \begin{frame}
-\frametitle{Next steps}
-
-\textbf{To do:} The current code is already fairly well-tested and mostly
-stable. However, there are some important items on the task list.
-\begin{itemize}
-  \item Extend/smooth package vignette.
-  \item Add \code{xtrafo}/\code{ytrafo} to \fct{ctree} reimplementation.
-  \item Switch \fct{mob} to new \class{party} class.
-  \item Create new subclass \class{modelparty} that facilitates
-    handling of formulas, terms, model frames, etc.\ for model-based trees.
-\end{itemize}
+\frametitle{Your own tree}
 
@@
 
-\begin{frame}[fragile]
-\frametitle{Next steps}
+\begin{frame}
+\frametitle{What's next?}
 
@@
 
-\textbf{Illustration:} Logistic regression (or logit model), assessing differences in
-the effect of ``preferential treatment'' for women or children.
-
-<<mob-show, echo=TRUE, eval=FALSE>>=
-<<mob>>
-@
-
-Odds ratio of survival given treatment differs across subsets (slope), as does
-the survival probability of male adults (intercept).
-
-<<mob-coef, echo=TRUE, results=verbatim>>=
-coef(mb)
-@
+Talk by Achim; but I'm entitled to a coffee first...
 
 \end{frame}
-
-\begin{frame}
-\frametitle{Next steps}
-
-\vspace*{-0.3cm}
-
-\setkeys{Gin}{width=1.1\textwidth}
-\hspace*{-0.4cm}%
-<<Titanic-mob-visualization, fig=TRUE, height=7, width=12>>=
-plot(mb, tp_args = list(margins = c(1.5, 2, 1.5, 2.5)))
-@
-
-\end{frame}
-
-\subsection{Computational details}
 
@@
 \textbf{Data:} The contingency table \code{Titanic} is transformed
-to the data frame \code{ttnc} using the following code. 
-%
-<<data1, echo=TRUE, eval=FALSE>>=
-<<data>>
-@
-%
-The last transformation also adds the treatment variable for the model-based recursive
-partitioning analysis.
+as explained in the \code{constparty} package vignette.
 
 \end{frame}
-
-\subsection{References}
 

### 2014-02-10 pkg/partykit/inst/NUS-2014/constparty_TH_AZ.Rnw
show example tree algorithm
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@456 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+<<mytree, eval=FALSE>>=
+findsplit <- function(response, data, weights) {
+
+  ### extract response values from data
+  y <- data[[response]]
+
+  ### cycle through all features
+  logpmin <- 0; xselect <- NULL
+  for (i in which(names(data) != response)) {
+
+    ### expand data by case weights
+    x <- data[[i]]
+    xt <- rep(x, weights) 
+    yt <- rep(y, weights)
+
+    ### set-up all possible splits in two daughter nodes
+    lev <- levels(xt[drop = TRUE])
+    if (length(lev) == 1) next()
+    if (length(lev) == 2) comb <- list(lev[1])
+    if (length(lev) > 2) 
+        comb <- do.call("c", lapply(1:(length(lev) - 2),
+            function(x) combn(lev,x, simplify = FALSE)))
+
+    ### assess all potential splits by their chi^2 test log p-value
+    logp <- sapply(comb, function(q) {   
+      ct <- chisq.test(table(yt, xt %in% q))
+      pchisq(ct$statistic, ct$parameter, log = TRUE, lower.tail = FALSE)
+    })
+
+    ### if the best split in variable i significant AND
+    ### better than what we already had store this information
+    if (min(logp) < logpmin & min(logp) < log(0.01)) {
+      logpmin <- min(logp)
+      xselect <- i
+      splitpoint <- comb[[which.min(logp)]]
+    }
+  }
+
+  ### no significant split found, give up  
+  if (is.null(xselect)) return(NULL)
+
+  ### return split as partysplit object
+  return(partysplit(
+      varid = as.integer(xselect),         ### which variable?   
+      index = levels(data[[xselect]]) %in%
+              splitpoint + 1L,             ### which split point?
+      info = list(p.value = exp(logpmin)   ### save p-value in addition
+  )))
+}
+growtree <- function(id = 1L, response, data, weights) {
+
+  ### for less than 30 observations stop here
+  if (sum(weights) < 30) return(partynode(id = id))
+
+  ### find best split
+  sp <- findsplit(response, data, weights)
+  ### no split found, stop here
+  if (is.null(sp)) return(partynode(id = id))
+
+  ### actually split the data
+  kidids <- kidids_split(sp, data = data)
+
+  ### set-up all daugther nodes
+  kids <- vector(mode = "list", length = max(kidids))
+  for (kidid in 1:max(kidids)) {
+  ### select observations for current node
+  w <- weights
+  w[kidids != kidid] <- 0
+  ### get next node id
+  if (kidid > 1) {
+    myid <- max(nodeids(kids[[kidid - 1]]))
+  } else {
+    myid <- id
+  }
+  ### start recursion on this daugther node
+  kids[[kidid]] <- growtree(id = as.integer(myid + 1), response, data, w)
+  }
+
+  ### return nodes  
+  return(partynode(id = as.integer(id), split = sp, kids = kids))
+}
+mytree <- function(formula, data, weights = NULL) {
+
+  ### name of the response variable
+  response <- all.vars(formula)[1]
+  ### data without missing values, response comes last
+  data <- data[complete.cases(data), c(all.vars(formula)[-1], response)]
+  ### data is factors only
+  stopifnot(all(sapply(data, is.factor)))
+
+  if (is.null(weights)) weights <- rep(1L, nrow(data))
+  ### weights are case weights, i.e., integers
+  stopifnot(length(weights) == nrow(data) &
+    max(abs(weights - floor(weights))) < .Machine$double.eps)
+
+  ### grow tree
+  nodes <- growtree(id = 1L, response, data, weights)
+
+  ### compute terminal node number for each observation
+  fitted <- fitted_node(nodes, data = data)
+  ### return rich constparty object
+  ret <- party(nodes, 
+    data = data,
+    fitted = data.frame(
+      "(fitted)" = fitted,
+      "(response)" = data[[response]],
+      "(weights)" = weights,
+      check.names = FALSE),
+    terms = terms(formula))
+  as.constparty(ret)
+}  
+myttnc <- mytree(Survived ~ Class + Age + Gender, data = ttnc)
+@
+
 <<cache>>=
@@
 <<mob>>
-save(rp, ct, ev, pm, mb, file = "slides-trees.rda")
+<<mytree>>
+save(rp, ct, ev, pm, mb, myttnc, file = "slides-trees.rda")
 }
@@
 \begin{frame}
-\frametitle{Your own tree}
+\frametitle{Your first partykit tree}
+
+Finally, we want to set-up a little program that implements the
+following tree algorithm for the Titanic data
+\begin{itemize}
+\item In each node, do
+  \begin{itemize}
+    \item For each explanatory variable, do
+    \begin{itemize}
+      \item compute the minimal $\chi^2$ $p$-value over all
+            possible splits in two groups
+    \end{itemize}
+  \item Select the best variable/split combination
+        and implement the split
+  \end{itemize}
+\item Recurse until $p$-value > .01 or sample size too small.
+\end{itemize}
+
+We only need three little functions:
+\begin{itemize}
+\item \code{findsplit}
+\item \code{growtree}
+\item \code{mytree}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Your first partykit tree}
+
+\setkeys{Gin}{width=1.1\textwidth}
+\hspace*{-0.4cm}%
+<<Titanic-mytree-vis, fig = TRUE, height=8, width=12>>=
+plot(myttnc)
+@
 

### 2014-02-11 pkg/partykit/inst/NUS-2014/constparty_TH_AZ.Rnw
cosmetics; mostly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@457 255f4249-2263-4417-b1ea-8498779e27a7

@@
 ))
+set.seed(1071)
 @
@@
 <<evtree, eval=FALSE>>=
-set.seed(1071)
 ev <- evtree(Survived ~ Gender + Age + Class, 
@@
         Achim Zeileis (Universit\"at Innsbruck)}
+\date{NUS 2014}
 %\URL{http://user.math.uzh.ch/hothorn/}
@@
   \item Illustrate several trees using the same data.
-  \item Not use the iris data (or something from mlbench).
-\end{itemize}
-
-\medskip
-
-\textbf{Solution:}
-\begin{itemize}
-  \item Use Titanic survival data (oh well\dots).
+  \item Here: Titanic survival data.
   \item In case you are not familiar with it: Survival status, gender, age (child/adult), and
@@
 \hspace*{-0.4cm}%
-<<Titanic-pmml-visualization, fig=TRUE, height=8, width=12>>=
+<<Titanic-pmml-visualization, fig=TRUE, height=10, width=16>>=
 plot(pm)
@@
 \end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{findsplit}
+
+R code
+
+\end{frame}
+
+\begin{frame}
+\frametitle{growtree}
+
+R code
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{mytree}
+
+R code
+
+<<Titanic-mytree-vis, eval = FALSE, echo = TRUE>>=
+plot(mytree(Survived ~ Class + Age + Gender, 
+            data = ttnc))
+@
 
@@
 
-Hothorn T, Zeileis A (2013).
+Hothorn T, Zeileis A (2014).
  \textit{partykit: A Toolkit for Recursive Partytioning.}
- R~package vignette version~0.1-6.
+ R~package vignette version~0.2-0.
  URL~\url{http://CRAN.R-project.org/package=partykit}

### 2014-03-09 pkg/partykit/inst/NUS-2014/modelparty.Rnw
merged slides from JKU and Psychoco -> still needs a lot of work
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@465 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-02-07 pkg/partykit/inst/Psychoco-2014/slides.Rnw
@@
 
-\graphicspath{{images/}}
+\graphicspath{{../Psychoco-2014/images/}}
 
@@
 
-\subsection{Model-based recursive partitioning}
+\subsection{Motivation}
 
 \begin{frame}
-\frametitle{Model-based recursive partitioning}
+\frametitle{Motivation: Trees}
 
-\textbf{Models:} Estimation of parametric models with
-observations $y_i$ (and regressors $x_i$),
-parameter vector $\theta$, and additive objective function $\Psi$.
-\[ \widehat \theta \quad = \quad \argmin_{\theta}
-   \sum_i \Psi(y_i, x_i, \theta). \]
+Breiman (2001, \textit{Statistical Science}) distinguishes two cultures
+of statistical modeling.
+
+\medskip
+
+\begin{itemize}
+  \item \textbf{Data models:} Stochastic models, typically parametric.\\
+        $\rightarrow$ Classical strategy in statistics.
+	Regression models are still the workhorse for many empirical analyses.
+  \item \textbf{Algorithmic models:} Flexible models, data-generating process unknown.
+        $\rightarrow$ Still few applications in many fields, e.g., social sciences or economics.
+\end{itemize}
+
+\medskip
+
+\textbf{Examples:} Recursive partitioning, decision trees.
+Class of flexible methods for classification and regression.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation: Trees}
+
+\textbf{Formally}: Modeling of dependent variable $Y$ by ``learning''
+a recursive partition w.r.t\ explanatory variables $Z_1, \dots, Z_l$.
 
@@
 
-\textbf{Recursive partitioning}:
+\textbf{Key features}:
 
 \begin{enumerate}
-  \item Fit the model in the current subsample.
-  \item Assess the stability of $\theta$ across each partitioning variable $z_j$.
-  \item Split sample along the $z_{j^*}$ with strongest association:
+  \item Predictive power in nonlinear regression relationships.
+  \item Interpretability (enhanced by visualization), i.e., no ``black box'' methods.
+\end{enumerate}
+
+\bigskip
+
+\textbf{Examples}: CART and C4.5 in statistical and machine learning, respectively.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation: Leaves}
+
+\textbf{Typically:} Simple models for univeriate $Y$, e.g., mean.
+
+\bigskip
+
+\textbf{Idea:} More complex models for multivariate $Y$, e.g., multivariate
+normal model, regression models, etc.
+
+\bigskip
+
+\textbf{Here:} Synthesis of parametric data models and algorithmic tree models.
+
+\bigskip
+
+\textbf{Goal:} Fitting local models by partitioning of the sample space.
+
+\end{frame}
+
+\subsection{Recursive partitioning}
+
+\begin{frame}
+\frametitle{Recursive partitioning}
+
+\textbf{Base algorithm}:
+
+\begin{enumerate}
+  \item Fit model for $Y$.
+  \item Assess association of $Y$ and each $Z_j$.
+  \item Split sample along the $Z_{j^*}$ with strongest association:
         Choose breakpoint with highest improvement of the model fit.
@@
 
+%% \medskip
+%% 
+%% \textbf{Generally:} Tree algorithms differ w.r.t.\ choice of model (1),
+%% association measure~(2), split strategy (3) and stopping criterion (4).
+
+\bigskip
+
+\textbf{Here:} Segmentation~(3) of parametric models~(1) with 
+additive objective function using parameter instability tests~(2)
+and associated statistical significance~(4).
+
+\end{frame}
+
+\subsection{Model estimation}
+
+\begin{frame}
+\frametitle{1.~Model estimation}
+
+\textbf{Models:} $\mathcal{M}(Y, \theta)$ with (potentially) multivariate
+observations $Y \in \mathcal{Y}$ and $k$-dimensional parameter vector
+$\theta \in \Theta$.
+
+\bigskip
+
+\textbf{Parameter estimation:} $\widehat \theta$ by optimization of
+objective function $\Psi(Y, \theta)$ for $n$ observations $Y_i$ ($i = 1, \dots, n$):
+
+\[ \widehat \theta \quad = \quad \argmin_{\theta \in \Theta}
+   \sum_{i = 1}^n \Psi(Y_i, \theta). \]
+
+\bigskip
+
+\textbf{Special cases:} Maximum likelihood (ML), weighted and ordinary
+least squares (OLS and WLS), quasi-ML, and other M-estimators.
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{Model-based recursive partitioning}
+\frametitle{1.~Model estimation}
 
-\textbf{Parameter instability tests:}
-\begin{itemize}
-  \item Based on empirical estimating functions (or score/gradient contributions):
-    $\Psi'(y_i, x_i, \hat \theta)$.
-  \item Under parameter stability: $\Psi'$ fluctuates randomly around its expectation zero.
-  \item Under parameter instability: Systematic departures from zero in subsamples.
-  \item Hence fluctuation can be captured across numeric partitioning variables
-    or within levels of categorical partitioning variables.
-  \item Bonferroni correction for testing across multiple partitioning variables.
-\end{itemize}
+\textbf{Estimating function:} $\widehat{\theta}$ can also be defined in terms of
+  \[ \sum_{i = 1}^n \psi(Y_i, \widehat{\theta}) = 0, \]
+where $\psi(Y, \theta) = {\partial \Psi(Y, \theta)}/{\partial \theta}$.
+
+\bigskip
+
+\textbf{Central limit theorem:} If there is a true parameter $\theta_0$ and given
+certain weak regularity conditions:
+  \[ \sqrt{n} (\widehat{\theta} - \theta_0) \quad \darrow \quad \mathcal{N}(0, V(\theta_0)), \]
+where $V(\theta_0) = \{A(\theta_0)\}^{-1} B(\theta_0) \{A(\theta_0)\}^{-1}$.
+$A$ and $B$ are the expectation of the derivative of $\psi$ and
+the variance of $\psi$, respectively.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{1.~Model estimation}
+
+\textbf{Idea:}
+In many situations, a single global model 
+$\mathcal{M}(Y, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
+But it might be possible to find a partition w.r.t.\ the variables
+$Z = (Z_1, \dots, Z_l)$ so that a well-fitting model can be found locally
+in each cell of the partition.
+
+\bigskip
+
+\textbf{Tool:} Assess parameter instability w.r.t\ to partitioning variables
+$Z_j \in \mathcal{Z}_j \; (j = 1, \dots, l)$.
+
+\end{frame}
+
+\subsection{Parameter instability tests}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+Generalized M-fluctuation tests
+capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $Z_j$.
+
+\bigskip
+
+\textbf{Basis:} Empirical fluctuation process of cumulative deviations
+w.r.t.\ to an ordering $\sigma(Z_{ij})$.
+
+\[
+W_j(t, \widehat \theta)
+  \quad = \quad {\widehat B}^{-1/2} n^{-1/2}
+                \sum_{i = 1}^{\lfloor nt \rfloor}
+		\psi(Y_{\sigma(Z_{ij})}, \widehat \theta)
+\qquad (0 \le t \le 1)
+\]
+
+\bigskip
+
+\textbf{Functional central limit theorem:} Under parameter stability
+$W_j(\cdot) \, \darrow \, W^0(\cdot)$, where
+$W^0$ is a $k$-dimensional Brownian bridge.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+\setkeys{Gin}{width=1.08\textwidth}
+\hspace*{-0.7cm}%
+<<fig=TRUE, height=5, width=10, echo=FALSE, results=hide>>=
+## artificial data
+set.seed(42)
+a <- rnorm(41, mean = 0, sd = 0.12)
+b <- rnorm(41, mean = 1, sd = 0.12)
+b[40] <- b[40]-0.1
+b[41] <- b[41]-0.2
+a[1] <- a[1]+0.2
+a[2] <- a[2]+0.1
+x <- c(b, a)
+
+par(mar = c(5, 5, 3, 1), mfrow = c(1, 2))
+
+## data, model, and residuals
+plot(0:81, x, xlab = expression(Z[j]), ylab = "Y", ylim = c(-2, 3),
+  type = "l", xlim = c(1, 80), axes=FALSE)
+axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
+axis(2,at=seq(-2,3,1),labels=seq(200,1200,200),cex.axis=1.2, cex.lab=1.2)
+box()
+abline(h = mean(x), lty = 3)
+for(i in c(5, 15, 25, 35)) {
+  segments(x0 = i, x1 = i, y0 = x[i+1], y1 = mean(x), lty = 2)
+  i <- i + 40
+  segments(x0 = i, x1 = i, y0 = mean(x), y1 = x[i+1], lty = 2)
+}
+
+## CUSUM graph
+x_efp <- c(0, cumsum(x - mean(x)))/sqrt(length(x) * var(x))
+plot(-1:81, x_efp, xlab = expression(Z[j]), ylab = "Fluctuation Process", 
+  type = "l", xlim = c(1, 80), ylim = c(-5, 5), axes=FALSE)
+axis(1,at=seq(0,80,20),labels=seq(2004,2012,2),cex.axis=1.2, cex.lab=1.2)
+axis(2, cex.axis=1.2, cex.lab=1.2)
+box()
+abline(h = 0, lty = 3)
+for(i in seq(5, 75, by = 10)) segments(x0 = i, x1 = i, y0 = 0, y1 = x_efp[i+2], lty = 2)
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+\textbf{Test statistics:} Scalar functional $\lambda(W_j)$
+that captures deviations from zero.
+
+\bigskip
+
+\textbf{Null distribution:} Asymptotic distribution of $\lambda(W^0)$.
+
+\bigskip
+
+\textbf{Special cases:} Class of test encompasses many well-known
+tests for different classes of models.
+Certain functionals~$\lambda$ are particularly intuitive for numeric and
+categorical $Z_j$, respectively.
+
+\bigskip
+
+\textbf{Advantage:} Model $\mathcal{M}(Y, \widehat \theta)$ just has to be
+estimated once. Empirical estimating functions $\psi(Y_i, \widehat \theta)$
+just have to be re-ordered and aggregated for each $Z_j$.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+\textbf{Splitting numeric variables:} Assess instability using
+sup$\mathit{LM}$ statistics.
+
+\[
+\lambda_{\mbox{sup}\mathit{LM}} (W_j) \quad = \quad
+\max_{i = \ui, \dots, \oi}  \, \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
+\left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
+\]
+
+\bigskip
+
+\textbf{Interpretation:} Maximization of single shift $\mathit{LM}$ statistics
+for all conceivable breakpoints in $[\ui, \oi]$.
+
+\bigskip
+
+\textbf{Limiting distribution:} Supremum of a squared, $k$-dimensional tied-down
+Bessel process.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+\textbf{Splitting categorical variables:} Assess instability using
+$\chi^2$ statistics.
+
+\[ \lambda_{\chi^2} (W_j) \quad = \quad
+\sum_{c = 1}^C 
+\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2
+\]
+
+\bigskip
+
+\textbf{Feature:} Invariant for re-ordering of the $C$ categories and the
+observations within each category.
+
+\bigskip
+
+\textbf{Interpretation:} Captures instability for split-up into $C$ categories.
+
+\bigskip
+
+\textbf{Limiting distribution:} $\chi^2$ with $k \cdot (C-1)$ degrees of freedom.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{2.~Tests for parameter instability}
+
+\textbf{Splitting ordinal variables:} FIXME
+
+like categorical
+
+like numeric but invariant to reordering within ties
+
+\end{frame}
+
+\subsection{Segmentation}
+
+\begin{frame}
+\frametitle{3.~Segmentation}
+
+\textbf{Goal:} Split model into $b = 1, \dots, B$ segments along the
+partitioning variable $Z_j$ associated with the highest parameter instability.
+Local optimization of
+
+\[ \sum_b \sum_{i \in I_b} \Psi(Y_i, \theta_b). \]
+
+\bigskip
+
+$B = 2$: Exhaustive search of order $O(n)$.
+
+\bigskip
+
+$B > 2$: Exhaustive search is of order $O(n^{B-1})$, but can be replaced
+by dynamic programming of order $O(n^2)$. Different methods (e.g., information
+criteria) can choose $B$ adaptively.
+
+\bigskip
+
+\textbf{Here:} Binary partitioning.
+
+\end{frame}
+
+\subsection{Pruning}
+
+\begin{frame}
+\frametitle{4.~Pruning}
+
+\textbf{Pruning:} Avoid overfitting.
+
+\bigskip
+
+\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when
+there is no significant parameter instability based on
+Bonferroni-corrected $p$~values of the fluctuation tests.
+
+\bigskip
+
+\textbf{Post-pruning:} Grow large tree and prune splits that do not
+improve the model fit based on information criteria (e.g., AIC or BIC).
 

### 2014-03-11 pkg/partykit/inst/NUS-2014/modelparty.Rnw
improved theory section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@467 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\textbf{Examples:} Recursive partitioning, decision trees.
-Class of flexible methods for classification and regression.
+\textbf{Classical example}: Trees, i.e., modeling of dependent variable $y$ by ``learning''
+a recursive partition w.r.t\ explanatory variables $z_1, \dots, z_l$.
 
@@
 \begin{frame}
-\frametitle{Motivation: Trees}
-
-\textbf{Formally}: Modeling of dependent variable $Y$ by ``learning''
-a recursive partition w.r.t\ explanatory variables $Z_1, \dots, Z_l$.
-
-\bigskip
+\frametitle{Motivation: Leaves}
 
 \textbf{Key features}:
-
 \begin{enumerate}
@@
 
-\textbf{Examples}: CART and C4.5 in statistical and machine learning, respectively.
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Motivation: Leaves}
-
-\textbf{Typically:} Simple models for univeriate $Y$, e.g., mean.
+\textbf{Typically:} Simple models for univeriate $y$, e.g., mean.
 
@@
 
-\textbf{Idea:} More complex models for multivariate $Y$, e.g., multivariate
-normal model, regression models, etc.
+\textbf{Idea:} More complex models for more complex $y$, e.g.,
+regression models, multivariate normal model, item responses, etc.
 
@@
 
-\textbf{Base algorithm}:
+\textbf{Model-based algorithm}:
 
 \begin{enumerate}
-  \item Fit model for $Y$.
-  \item Assess association of $Y$ and each $Z_j$.
-  \item Split sample along the $Z_{j^*}$ with strongest association:
+  \item Fit the parametric model in the current subsample.
+  \item Assess the stability of the parameters across each partitioning variable $z_j$.
+  \item Split sample along the $z_{j^*}$ with strongest instability:
         Choose breakpoint with highest improvement of the model fit.
@@
 \end{enumerate}
-
-%% \medskip
-%% 
-%% \textbf{Generally:} Tree algorithms differ w.r.t.\ choice of model (1),
-%% association measure~(2), split strategy (3) and stopping criterion (4).
-
-\bigskip
-
-\textbf{Here:} Segmentation~(3) of parametric models~(1) with 
-additive objective function using parameter instability tests~(2)
-and associated statistical significance~(4).
 
@@
 
-\textbf{Models:} $\mathcal{M}(Y, \theta)$ with (potentially) multivariate
-observations $Y \in \mathcal{Y}$ and $k$-dimensional parameter vector
+\textbf{Models:} $\mathcal{M}(y, x, \theta)$ with (potentially multivariate)
+observations $y$, optionally regressors $x$, and $k$-dimensional parameter vector
 $\theta \in \Theta$.
@@
 \textbf{Parameter estimation:} $\widehat \theta$ by optimization of
-objective function $\Psi(Y, \theta)$ for $n$ observations $Y_i$ ($i = 1, \dots, n$):
+objective function $\Psi(y, x, \theta)$ for $n$ observations $y_i$ ($i = 1, \dots, n$):
 
 \[ \widehat \theta \quad = \quad \argmin_{\theta \in \Theta}
-   \sum_{i = 1}^n \Psi(Y_i, \theta). \]
+   \sum_{i = 1}^n \Psi(y_i, x_i, \theta). \]
 
@@
 \textbf{Estimating function:} $\widehat{\theta}$ can also be defined in terms of
-  \[ \sum_{i = 1}^n \psi(Y_i, \widehat{\theta}) = 0, \]
-where $\psi(Y, \theta) = {\partial \Psi(Y, \theta)}/{\partial \theta}$.
+  \[ \sum_{i = 1}^n \psi(y_i, x_i, \widehat{\theta}) = 0, \]
+where $\psi(y, x, \theta) = {\partial \Psi(y, x, \theta)}/{\partial \theta}$.
 
@@
 In many situations, a single global model 
-$\mathcal{M}(Y, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
+$\mathcal{M}(y, x, \theta)$ that fits \textbf{all} $n$~observations cannot be found.
 But it might be possible to find a partition w.r.t.\ the variables
-$Z = (Z_1, \dots, Z_l)$ so that a well-fitting model can be found locally
+$z_1, \dots, z_l$ so that a well-fitting model can be found locally
 in each cell of the partition.
@@
 
-\textbf{Tool:} Assess parameter instability w.r.t\ to partitioning variables
-$Z_j \in \mathcal{Z}_j \; (j = 1, \dots, l)$.
+\textbf{Tools:}
+\begin{itemize}
+  \item Assess parameter instability w.r.t\ to partitioning variables
+        $z_j \; (j = 1, \dots, l)$.
+  \item A general measure of deviation from the model is the estimating function $\psi(y, x, \theta)$.
+\end{itemize}
 
@@
 Generalized M-fluctuation tests
-capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $Z_j$.
+capture instabilities in $\widehat \theta$ for an ordering w.r.t\ $z_j$.
 
@@
 \textbf{Basis:} Empirical fluctuation process of cumulative deviations
-w.r.t.\ to an ordering $\sigma(Z_{ij})$.
+w.r.t.\ to an ordering $\sigma(z_{ij})$.
 
@@
                 \sum_{i = 1}^{\lfloor nt \rfloor}
-		\psi(Y_{\sigma(Z_{ij})}, \widehat \theta)
+		\psi(y_{\sigma(z_{ij})}, x_{\sigma(z_{ij})}, \widehat \theta)
 \qquad (0 \le t \le 1)
@@
 ## data, model, and residuals
-plot(0:81, x, xlab = expression(Z[j]), ylab = "Y", ylim = c(-2, 3),
+plot(0:81, x, xlab = expression(z[j]), ylab = "y", ylim = c(-2, 3),
   type = "l", xlim = c(1, 80), axes=FALSE)
@@
 x_efp <- c(0, cumsum(x - mean(x)))/sqrt(length(x) * var(x))
-plot(-1:81, x_efp, xlab = expression(Z[j]), ylab = "Fluctuation Process", 
+plot(-1:81, x_efp, xlab = expression(z[j]), ylab = "Fluctuation Process", 
   type = "l", xlim = c(1, 80), ylim = c(-5, 5), axes=FALSE)
@@
 Certain functionals~$\lambda$ are particularly intuitive for numeric and
-categorical $Z_j$, respectively.
+categorical $z_j$, respectively.
 
@@
 
-\textbf{Advantage:} Model $\mathcal{M}(Y, \widehat \theta)$ just has to be
-estimated once. Empirical estimating functions $\psi(Y_i, \widehat \theta)$
-just have to be re-ordered and aggregated for each $Z_j$.
+\textbf{Advantage:} Model $\mathcal{M}(y, x, \widehat \theta)$ just has to be
+estimated once. Empirical estimating functions $\psi(y_i, x_i, \widehat \theta)$
+just have to be re-ordered and aggregated for each $z_j$.
 
@@
 \[
-\lambda_{\mbox{sup}\mathit{LM}} (W_j) \quad = \quad
+\lambda_{\mathit{supLM}} (W_j) \quad = \quad
 \max_{i = \ui, \dots, \oi}  \, \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
@@
 
+\bigskip
+
+\textbf{Potential alternatives:} Many other parameter instability tests from the same
+class of tests, e.g., a Cram\'{e}r-von Mises test (or Nyblom-Hansen test),
+MOSUM tests, etc.
+
 \end{frame}
@@
 \sum_{c = 1}^C 
-\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2
+\frac{n}{|I_c|}  \left| \left| \Delta_{I_c} W_j \left( \frac{i}{n} \right) \right| \right|^2_2.
 \]
@@
 
-\textbf{Splitting ordinal variables:} FIXME
+\textbf{Splitting ordinal variables:} Several strategies conceivable.
+Assess instability either as for categorical variables (if $C$ is low),
+or as for numeric variables (if $C$ is high), or via a specialized test.
+\begin{eqnarray*}
+  \lambda_\mathit{maxLMo}(W_j) & = & \max_{i \in \{i_1, \dots, i_{C-1} \}} ~
+    \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1}
+    \left| \left| W_j \left( \frac{i}{n} \right) \right| \right|^2_2,\\
+  \lambda_\mathit{WDMo}(W_j) & = & \max_{i \in \{i_1, \dots, i_{C-1} \}} ~
+    \left(\frac{i}{n} \cdot \frac{n-i}{n} \right)^{-1/2}      
+    \left| \left| W_j\left( \frac{i}{n} \right) \right| \right|_{\infty}.
+\end{eqnarray*}
 
-like categorical
+\bigskip
 
-like numeric but invariant to reordering within ties
+\textbf{Interpretation:} Assess only the possible splitpoints $i_1, \dots, i_{C-1}$,
+based on $L_2$ or $L_\infty$ norm.
+
+\bigskip
+
+\textbf{Limiting distribution:} Maximum from selected points in a squared
+Bessel process or multivariate normal distribution, respectively.
 
@@
 \textbf{Goal:} Split model into $b = 1, \dots, B$ segments along the
-partitioning variable $Z_j$ associated with the highest parameter instability.
+partitioning variable $z_j$ associated with the highest parameter instability.
 Local optimization of
 
-\[ \sum_b \sum_{i \in I_b} \Psi(Y_i, \theta_b). \]
+\[ \sum_b \sum_{i \in I_b} \Psi(y_i, x_i, \theta_b). \]
 
@@
 
-\textbf{Here:} Binary partitioning.
+\textbf{Here:} Binary partitioning. Optionally, $B = C$ can be chosen (without search)
+for categorical variables.
 
@@
 
-\textbf{Pruning:} Avoid overfitting.
+\textbf{Goal:} Avoid overfitting.
 
@@
 
-\textbf{Pre-pruning:} Internal stopping criterium. Stop splitting when
-there is no significant parameter instability based on
-Bonferroni-corrected $p$~values of the fluctuation tests.
+\textbf{Pre-pruning:}
+\begin{itemize}
+  \item Internal stopping criterium.
+  \item Stop splitting when there is no significant parameter instability.
+  \item Based on Bonferroni-corrected $p$~values of the fluctuation tests.
+\end{itemize}
 
@@
 
-\textbf{Post-pruning:} Grow large tree and prune splits that do not
-improve the model fit based on information criteria (e.g., AIC or BIC).
+\textbf{Post-pruning:}
+\begin{itemize}
+  \item Grow large tree (e.g., with high significance level).
+  \item Prune splits that do not improve the model fit
+        based on information criteria (e.g., AIC or BIC).
+\end{itemize}
+
+\bigskip
+
+\textbf{Hyperparameters:} Significance level and information criterion penalty
+can be chosen manually (or possibly through cross-validation etc.).
 

### 2014-03-12 pkg/partykit/inst/NUS-2014/modelparty.Rnw
slides essentially ready except for class descriptions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@471 255f4249-2263-4417-b1ea-8498779e27a7

@@
 library("psychotools")
+options(digits = 4)
+@
+
+<<data, echo=FALSE, results=hide>>=
 data("Topmodel2007", package = "psychotree")
-options(digits = 4)
+data("Titanic", package = "datasets")
+ttnc <- as.data.frame(Titanic)
+ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
+names(ttnc)[2] <- "Gender"
+ttnc <- transform(ttnc, Treatment = factor(
+  Gender == "Female" | Age == "Child",
+  levels = c(FALSE, TRUE),
+  labels = c("Normal\n(Male&Adult)", "Preferential\n(Female|Child)")
+))
 @
@@
 \begin{itemize}
-  \item Model-based recursive partitioning
-  \begin{itemize}
-    \item A generic approach
-    \item Example: Bradley-Terry trees
+  \item Motivation: Trees and leaves
+  \item Model-based (MOB) recursive partitioning
+  \begin{itemize}
+    \item Model estimation
+    \item Tests for parameter instability
+    \item Segmentation
+    \item Pruning
+    \item Local models
   \end{itemize}
@@
   \end{itemize}
-  \item Illustration
+  \item Application
+  \begin{itemize}
+    \item Paired comparisons for Germany's Topmodel finalists
+    \item Bradley-Terry trees
+    \item Implementation from scratch
+  \end{itemize}
 \end{itemize}
@@
 
-\textbf{Model-based algorithm}:
+\textbf{Model-based (MOB) algorithm}:
 
@@
 \end{enumerate}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+\textbf{Example:} Logistic regression, assessing differences in
+the effect of ``preferential treatment'' (``women and children first''?).
+
+\bigskip
+
+\textbf{Result:} Log-odds ratio of survival given treatment differs across
+classes (slope), as does the survival probability of male adults (intercept).
+
+\bigskip
+
+\textbf{In R:} Generalized linear model tree with binomial family (and default logit link).
+
+<<ttnc, results=hide>>=
+mb <- glmtree(Survived ~ Treatment | Age + Gender + Class,
+  data = ttnc, family = binomial, alpha = 0.05, prune = "BIC")
+plot(mb)
+print(mb)
+@
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Recursive partitioning}
+
+\vspace*{-0.3cm}
+
+\setkeys{Gin}{width=1.1\textwidth}
+\hspace*{-0.4cm}%
+<<ttnc-plot, fig=TRUE, height=7, width=12, echo=FALSE, results=hide>>=
+plot(mb, tp_args = list(margins = c(1.5, 2, 1.5, 2.5)))
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Recursive partitioning}
+
+<<ttnc-print, echo=FALSE>>=
+ttnc$Treatment <- factor(ttnc$Treatment, labels = c("Normal", "Preferential"))
+mb0 <- glmtree(Survived ~ Treatment | Age + Gender + Class,
+  data = ttnc, family = binomial, alpha = 0.05, prune = "BIC")
+mb0
+@
 
@@
 
-\textbf{Interpretation:} Captures instability for split-up into $C$ categories.
+\textbf{Interpretation:} Capture instability for split-up into $C$ categories.
 
@@
 
-\begin{frame}
-\frametitle{Bradley-Terry trees}
-
-\hspace*{0.2cm}%
-\includegraphics[width=0.23\textwidth]{Barbara.jpg} \hspace*{0.1cm}
-\includegraphics[width=0.23\textwidth]{Anni.jpg} \hspace*{0.1cm}
-\includegraphics[width=0.23\textwidth]{Hana.jpg} \hspace*{0.1cm}
-\includegraphics[width=0.23\textwidth]{Fiona.jpg}
-
-\vspace*{0.1cm}
-\hspace*{0.2cm}%
-\includegraphics[width=0.23\textwidth]{Mandy.jpg} \hspace*{0.1cm}
-\includegraphics[width=0.23\textwidth]{Anja.jpg} \hspace*{0.2cm}%
-\begin{tabular}{p{0.48\textwidth}}
-\textbf{Questions:} Which of these
-
-women is more attractive?
-
-How does the answer depend on
-
-age, gender, and the familiarity
-
-with the associated TV show
-
-Germany's Next Topmodel?
-\vspace*{3.4cm}
-\end{tabular}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Bradley-Terry trees}
-
-<<topmodel-tree, echo=FALSE, results=hide>>=
-bt0 <- mob(preference ~ 1 | gender + age + q1 + q2 + q3,
-  data = Topmodel2007, fit = btfit2)
-@
-
-\vspace*{-0.5cm}
-
-\setkeys{Gin}{width=1.15\textwidth}
-\hspace*{-0.7cm}%
-<<topmodel-plot, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
-plot(bt0, drop = TRUE, tnex = 2,
-  terminal_panel = node_btplot(bt0, abbreviate = 1, yscale = c(0, 0.5)))
-@
+\subsection{Local models}
+
+\begin{frame}
+\frametitle{Local models}
+
+\textbf{Goals:}
+\begin{itemize}
+  \item Detection of interactions and nonlinearities in regressions.
+  \item Add explanatory variables for models without regressors.
+  \item Detect violations of parameter stability (measurement invariance)
+    across several variables adaptively.
+\end{itemize}
+
+\bigskip
+
+\textbf{Mobsters:}
+\begin{itemize}
+  \item Linear and generalized linear model trees (Zeileis \emph{et al.} 2008).
+  \item Censored survival regression trees: parametric proportional hazard
+    and accelerated failure time models (Zeileis \emph{et al.}~2008).
+  \item Beta regression trees (Gr\"un \emph{et al.}~2012).
+  \item Bradley-Terry trees for paired comparisons (Strobl \emph{et al.}~2011).
+  \item Item response theory (IRT) trees: Rasch, rating scale and partial credit model
+    (Strobl \emph{et al.}~2014, Abou El-Komboz \emph{et al.}~2014).
+\end{itemize}
 
@@
 
+\begin{frame}[fragile]
+\frametitle{Implementation: Class}
+
+\textbf{Class:} `\code{modelparty}'.
+
+FIXME
+
+\end{frame}
+
 \subsection{Bradley-Terry trees}
 
+\begin{frame}
+\frametitle{Bradley-Terry trees}
+
+\hspace*{0.2cm}%
+\includegraphics[width=0.23\textwidth]{Barbara.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.23\textwidth]{Anni.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.23\textwidth]{Hana.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.23\textwidth]{Fiona.jpg}
+
+\vspace*{0.1cm}
+\hspace*{0.2cm}%
+\includegraphics[width=0.23\textwidth]{Mandy.jpg} \hspace*{0.1cm}
+\includegraphics[width=0.23\textwidth]{Anja.jpg} \hspace*{0.2cm}%
+\begin{tabular}{p{0.48\textwidth}}
+\textbf{Questions:} Which of these
+
+women is more attractive?
+
+How does the answer depend on
+
+age, gender, and the familiarity
+
+with the associated TV show
+
+Germany's Next Topmodel?
+\vspace*{3.4cm}
+\end{tabular}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Bradley-Terry trees}
+
+\textbf{Task:} Preference scaling of attractiveness.
+
+\bigskip
+
+\textbf{Model:} Paired comparison via Bradley-Terry.
+\begin{itemize}
+  \item Paired comparisons of attractiveness for
+    \emph{Germany's Next Topmodel~2007} finalists:
+    Barbara, Anni, Hana, Fiona, Mandy, Anja.
+  \item Survey with 192~respondents at Universit{\"a}t T{\"u}bingen.
+  \item Available covariates: Gender, age, familiarty with the TV~show.
+  \item Familiarity assessed by yes/no questions:
+    (1)~Do you recognize the women?/Do you know the show?
+    (2)~Did you watch it regularly?
+    (3)~Did you watch the final show?/Do you know who won?
+\end{itemize}
+
+\bigskip
+
+\textbf{Mobster:} \code{bttree()} in \emph{psychotree} (Strobl \emph{et al.}~2011).
+
+\bigskip
+
+\textbf{Here:} Use \code{mob()} directly to build model from scratch.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Bradley-Terry trees}
+
+<<topmodel-tree, echo=FALSE, results=hide>>=
+bt0 <- mob(preference ~ 1 | gender + age + q1 + q2 + q3,
+  data = Topmodel2007, fit = btfit2)
+@
+
+\vspace*{-0.5cm}
+
+\setkeys{Gin}{width=1.15\textwidth}
+\hspace*{-0.7cm}%
+<<topmodel-plot, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
+plot(bt0, drop = TRUE, tnex = 2,
+  terminal_panel = node_btplot(bt0, abbreviate = 1, yscale = c(0, 0.5)))
+@
+
+\end{frame}
+
 \begin{frame}[fragile]
-\frametitle{Illustration: Bradley-Terry trees}
+\frametitle{Bradley-Terry trees}
 
@@
 \begin{frame}[fragile]
-\frametitle{Illustration: Bradley-Terry trees}
+\frametitle{Bradley-Terry trees}
 
@@
 \begin{frame}
-\frametitle{Illustration: Bradley-Terry trees}
+\frametitle{Bradley-Terry trees}
 
@@
 \begin{itemize}
-  \item All new implementation of model-based recursive partitioning in \emph{partykit}.
+  \item Synthesis of parametric data models and algorithmic tree models.
+  \item Based on modern class of parameter instability tests.
+  \item Aims to minimize clearly defined objective function by greedy forward search.
+  \item Can be applied general class of parametric models.
+  \item Alternative to traditional means of model specification,
+        especially for variables with unknown association.
+  \item All new implementation in \emph{partykit}.
   \item Enables more efficient computations, rapid prototyping, flexible customization.
-  \item Some new inference options.
 \end{itemize}
@@
 
-\small
+\footnotesize
+
+\textbf{Software:}
+
+\medskip
 
@@
 
-\bigskip
+\medskip
 
@@
  \textit{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R.}
- \code{vignette("mob", package = "partykit")}.
-
-\bigskip
-
-Strobl C, Wickelmaier F, Zeileis A (2011).
-  \dquote{Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.}
-  \textit{Journal of Educational and Behavioral Statistics}, \textbf{36}(2), 135--153.
-  \doi{10.3102/1076998609359791}
-
-\bigskip
+ \texttt{vignette("mob",~package~=~"partykit")}.
+
+\bigskip
+
+\textbf{Inference:}
+
+\medskip
+
+Zeileis A, Hornik K (2007).
+ \dquote{Generalized {M}-Fluctuation Tests for Parameter Instability.}
+ \textit{Statistica Neerlandica},
+ \textbf{61}(4), 488--508.
+ \doi{10.1111/j.1467-9574.2007.00371.x}
+
+\medskip
+
+Merkle EC, Zeileis A (2013).
+ \dquote{Tests of Measurement Invariance without Subgroups: A Generalization of Classical Methods.}
+ \textit{Psychometrika}, \textbf{78}(1), 59--82.
+ \doi{10.1007/S11336-012-9302-4}
+
+\medskip
+
+Merkle EC, Fan J, Zeileis A (2014).
+ \dquote{Testing for Measurement Invariance with Respect to an Ordinal Variable.}
+ \textit{Psychometrika}. Forthcoming.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{References}
+
+\footnotesize
+
+\textbf{Trees:}
+
+\medskip
 
@@
 
+\medskip
+
+Strobl C, Wickelmaier F, Zeileis A (2011).
+  \dquote{Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.}
+  \textit{Journal of Educational and Behavioral Statistics}, \textbf{36}(2), 135--153.
+  \doi{10.3102/1076998609359791}
+
+\medskip
+
+Gr\"un B, Kosmidis I, Zeileis A (2012).
+  \dquote{Extended Beta Regression in R: Shaken, Stirred, Mixed, and Partitioned.}
+  \textit{Journal of Statistical Software}, \textbf{48}(11), 1--25.
+  URL~\url{http://www.jstatsoft.org/v48/i11/}
+
+%% \medskip
+%% 
+%% Rusch T, Lee I, Hornik K, Jank W, Zeileis A (2013).
+%%   \dquote{Influencing Elections with Statistics: Targeting Voters with Logistic Regression Trees.}
+%%   \textit{The Annals of Applied Statistics}, \textbf{7}(3), 1612--1639.
+%%   \doi{10.1214/13-AOAS648}
+
+\medskip
+
+Strobl C, Kopf J, Zeileis A (2013).
+  \dquote{A New Method for Detecting Differential Item Functioning in the Rasch Model.}
+  \textit{Psychometrika}. Forthcoming.
+  \doi{10.1007/s11336-013-9388-3}
+
+\medskip
+
+Abou El-Komboz B, Zeileis A, Strobl C (2014).
+  \dquote{Detecting Differential Item and Step Functioning with Rating Scale and Partial Credit Trees.}
+  \textit{Technical Report 152}, Department of Statistics, Ludwig-Maximilians-Universit\"at M\"unchen.
+  URL~\url{http://epub.ub.uni-muenchen.de/17984/}
+
 \end{frame}

### 2014-03-12 pkg/partykit/R/party.R
implement and nodeprune
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@474 255f4249-2263-4417-b1ea-8498779e27a7

@@ -350,11 +350,15 @@ predict_party.constparty <- function(party, id, newdata = NULL,
 }
 
 ### functions for node prediction based on fitted / response
-.pred_Surv <- function(y, w)
+.pred_Surv <- function(y, w) {
+    if (length(y) == 0) return(NA)
     survfit(y ~ 1, weights = w, subset = w > 0)
+}
 
-.pred_Surv_response <- function(y, w)
+.pred_Surv_response <- function(y, w) {
+    if (length(y) == 0) return(NA)
     .median_survival_time(.pred_Surv(y, w))
+}
  
 .pred_factor <- function(y, w) {
     lev <- levels(y)
@@ -377,6 +381,7 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     weighted.mean(y, w, na.rm = TRUE)
 
 .pred_ecdf <- function(y, w) {
+    if (length(y) == 0) return(NA)
     iw <- as.integer(round(w))
     if (max(abs(w - iw)) < sqrt(.Machine$double.eps)) {
         y <- rep(y, w)
@@ -410,7 +415,10 @@ predict_party.constparty <- function(party, id, newdata = NULL,
         ### id may also refer to inner nodes
         tab <- as.array(lapply(sort(unique(id)), function(i) {
             index <- fitted %in% nodeids(node, i, terminal = TRUE)
-            FUN(response[index], weights[index])
+            ret <- FUN(response[index], weights[index])
+            ### no information about i in fitted
+            if (all(!index)) ret[1] <- NA
+            return(ret)
         }))
         names(tab) <- as.character(sort(unique(id)))
     }
@@ -482,6 +490,59 @@ width.party <- function(x, ...) {
 depth.party <- function(x, root = FALSE, ...) {
   depth(node_party(x), root = root, ...)
 }
+
+nodeprune <- function(x, ids, ...)
+    UseMethod("nodeprune")
+
+nodeprune.party <- function(x, ids, ...) {
+
+    ### map names to nodeids
+    if (!is.numeric(ids))
+        ids <- match(ids, names(x))
+    stopifnot(ids %in% nodeids(x))
+
+    ### compute indices path to each node
+    ### to be pruned off
+    idxs <- lapply(ids, .get_path, obj = x)
+
+    ### [[.party is NOT [[.list
+    cls <- class(x)
+    x <- unclass(x)
+
+    for (i in 1:length(idxs)) {
+    
+        idx <- idxs[[i]]
[…]

### 2014-03-12 pkg/partykit/R/utils.R

@@ -11,3 +11,55 @@
     med <- suppressWarnings(minmin(x$surv, x$time))
     return(med)
 }
+
+### get the recursive index
+### obj is of class "party"
+.get_path <- function(obj, i) {
+
+    idx <- which(names(unclass(obj)) == "node")
+    recFun <- function(node, i) {
+        if (id_node(node) == i) return(NULL)
+        idx <<- c(idx, which(names(unclass(node)) == "kids"))
+        kid <- sapply(kids_node(node), id_node)
+        nextid <- max(which(kid <= i))
+        idx <<- c(idx, nextid)
+        return(recFun(node[[nextid]], i))
+    }
+    out <- recFun(node_party(obj), i)
+    return(idx)
+}
+
+### <TH> shall we export this functionality?
+"nodeids<-" <- function(obj, value) UseMethod("nodeids<-")
+
+"nodeids<-.party" <- function(obj, value) {
+
+  id0 <- nodeids(obj)
+  id1 <- as.integer(value)
+  stopifnot(identical(id1, 1:length(id0)))
+
+  idxs <- lapply(id0, .get_path, obj = obj)
+  x <- unclass(obj)
+  nm <- x$names
+  for (i in 1:length(idxs))
+      x[[idxs[[i]]]]$id <- id1[i]
+  class(x) <- class(obj)
+  if (!is.null(nm)) 
+      names(x) <- nm[id0]
+
+  return(x)
+}
+
+"nodeids<-.constparty" <- function(obj, value) {
+
+  id0 <- nodeids(obj)
+  cls <- class(obj)
+  class(obj) <- "party"
+  nodeids(obj) <- value
+  id1 <- nodeids(obj)
+  obj$fitted[["(fitted)"]] <- 
+      id1[match(fitted(obj)[["(fitted)"]], id0)]
+  class(obj) <- cls
+  obj
+}
+### </TH>

### 2014-03-12 pkg/partykit/tests/regtest-party.R
nodeprune tests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@475 255f4249-2263-4417-b1ea-8498779e27a7

@@ -59,3 +59,50 @@ mp <- party(node_party(pfit), fitted = f, data = pfit$data)
 class(mp) <- c("constparty", "party")
 predict(mp, newdata = GBSG2[1:10,])
 
+### pruning
+## create party
+data("WeatherPlay", package = "partykit")
+py <- party(
+  partynode(1L,
+    split = partysplit(1L, index = 1:3),
+    kids = list(
+      partynode(2L,
+        split = partysplit(3L, breaks = 75),
+        kids = list(
+          partynode(3L, info = "yes"),
+          partynode(4L, info = "no"))),
+      partynode(5L, 
+        split = partysplit(3L, breaks = 20),
+        kids = list(
+          partynode(6L, info = "no"),
+          partynode(7L, info = "yes"))),
+      partynode(8L,
+        split = partysplit(4L, index = 1:2),
+        kids = list(
+          partynode(9L, info = "yes"),
+          partynode(10L, info = "no"))))),
+  WeatherPlay)
+names(py) <- LETTERS[nodeids(py)]
+
+## print
+print(py)
+(py5 <- nodeprune(py, 5))
+nodeids(py5)
+(pyH <- nodeprune(py5, "H"))
+nodeids(pyH)
+	
+ct <- ctree(Species ~ ., data = iris)
+(ctp <- nodeprune(ct, 4))
+
+names(ct) <- LETTERS[nodeids(ct)]
+(ctp <- nodeprune(ct, "D"))
+
+table(predict(ct, type = "node"), 
+      predict(ctp, type = "node"))
+
+(ct <- nodeprune(ct, names(ct)[names(ct) != "A"]))
+table(predict(ct, type = "node"))
+
+nodeprune(ct, "B")
+nodeprune(ct, "C")
+nodeprune(ct, "A")

### 2014-03-13 pkg/partykit/R/as.party.R
generalized as.party.J48 to as.party.Weka_tree ...a constparty is returned only for J48 though
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@481 255f4249-2263-4417-b1ea-8498779e27a7

@@ -109,10 +109,13 @@ model.frame.rpart <- function(formula, ...) {
   return(mf)
 }
 
-as.party.J48 <- function(obj, ...) {
+as.party.Weka_tree <- function(obj, ...) {
 
   ## needs RWeka and rJava
   stopifnot(require("RWeka"))
+
+  ## J48 tree? (can be transformed to "constparty")
+  j48 <- inherits(obj, "J48")
 
   ## construct metadata
   mf <- model.frame(obj)
@@ -120,21 +123,27 @@ as.party.J48 <- function(obj, ...) {
   mf_levels <- lapply(mf, levels)
 
   x <- rJava::.jcall(obj$classifier, "S", "graph")
+  if(j48) {
+    info <- NULL
+  } else {
+    info <- RWeka::parse_Weka_digraph(x, plainleaf = FALSE)$nodes[, 2L]
+    info <- strsplit(info, " (", fixed = TRUE)
+    info <- lapply(info, function(x) if(length(x) == 1L) x else c(x[1L], paste("(", x[-1L], sep = "")))
+  }
   x <- RWeka::parse_Weka_digraph(x, plainleaf = TRUE)
   nodes <- x$nodes
   edges <- x$edges
   is.leaf <- x$nodes[, "splitvar"] == ""
 
-  j48_kids <- function(i) {
+  weka_tree_kids <- function(i) {
     if (is.leaf[i]) return(NULL)
       else return(which(nodes[,"name"] %in% edges[nodes[i,"name"] == edges[,"from"], "to"]))
   }
 
-  j48_split <- function(i) {
+  weka_tree_split <- function(i) {
     if(is.leaf[i]) return(NULL)
     
     var_id <- which(nodes[i, "splitvar"] == names(mf))
-    ##
     edges <- edges[nodes[i,"name"] == edges[,"from"], "label"]
     split <- Map(c, sub("^([[:punct:]]+).*$", "\\1", edges), sub("^([[:punct:]]+) *", "", edges))
     ## ## for J48 the following suffices
@@ -160,21 +169,33 @@ as.party.J48 <- function(obj, ...) {
     return(split)
   }
 
-  j48_node <- function(i) {
-    if(is.null(j48_kids(i))) return(partynode(as.integer(i)))
-    partynode(as.integer(i), split = j48_split(i), kids = lapply(j48_kids(i), j48_node))
+  weka_tree_node <- function(i) {
[…]

### 2014-03-13 pkg/partykit/R/as.party.R
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@483 255f4249-2263-4417-b1ea-8498779e27a7

@@ -14,10 +14,10 @@ as.party.rpart <- function(obj, ...) {
     splitindex <- list()
     splitindex$primary <- numeric(n)
     splitindex$primary[!is.leaf] <- index[c(!is.leaf, FALSE)]
-    splitindex$surrogate <- lapply(1:n, function(i) {
+    splitindex$surrogate <- lapply(1L:n, function(i) {
         prim <- splitindex$primary[i]
         if (prim < 1 || ff[i, "nsurrogate"] == 0) return(NULL)
-        else return(prim + ff[i, "ncompete"] + 1:ff[i, "nsurrogate"])
+        else return(prim + ff[i, "ncompete"] + 1L:ff[i, "nsurrogate"])
     })
     
     mf <- model.frame(obj)
@@ -33,8 +33,8 @@ as.party.rpart <- function(obj, ...) {
 
     rpart_kids <- function(i) {
         if (is.leaf[i]) return(NULL)
-        else return(c(i + 1, 
-            which((cumsum(!is.leaf[-(1:i)]) + 1) == cumsum(is.leaf[-(1:i)]))[1] + 1 + i))
+        else return(c(i + 1L, 
+            which((cumsum(!is.leaf[-(1L:i)]) + 1L) == cumsum(is.leaf[-(1L:i)]))[1L] + 1L + i))
     }
 
     rpart_onesplit <- function(j) {
@@ -47,11 +47,11 @@ as.party.rpart <- function(obj, ...) {
                       index = if(obj$split[j, "ncat"] > 0) 2:1)
         } else {
             index <- obj$csplit[obj$split[j, "index"],]
-            ### csplit has columns 1:max(nlevels) for all factors
-            ### index <- index[1:obj$split[j, "ncat"]] ??? safer ???
-            index <- index[1:nlevels(mf[, rownames(obj$split)[j]])]
-            index[index == 2] <- NA ### level not present in split
-            index[index == 3] <- 2  ### 1..left, 3..right
+            ### csplit has columns 1L:max(nlevels) for all factors
+            ### index <- index[1L:obj$split[j, "ncat"]] ??? safer ???
+            index <- index[1L:nlevels(mf[, rownames(obj$split)[j]])]
+            index[index == 2L] <- NA ### level not present in split
+            index[index == 3L] <- 2L  ### 1..left, 3..right
             ret <- partysplit(varid = which(rownames(obj$split)[j] == names(mf)),
                       index = as.integer(index))
         }
[…]

### 2014-03-13 pkg/partykit/R/modelparty.R
assure non-negative nreg
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@485 255f4249-2263-4417-b1ea-8498779e27a7

@@ -614,7 +614,7 @@ mob <- function(formula, data, subset, na.action, weights, offset,
       fit = afit,
       control = control,
       dots = list(...),
-      nreg = as.integer(xreg) * (nyx - NCOL(Y)))
+      nreg = max(0L, as.integer(xreg) * (nyx - NCOL(Y))))
   )
   class(rval) <- c("modelparty", class(rval))
   return(rval)

### 2014-03-13 pkg/partykit/inst/NUS-2014/modelparty.Rnw
Formula refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@480 255f4249-2263-4417-b1ea-8498779e27a7
draft of class description
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@486 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \begin{frame}[fragile]
-\frametitle{Implementation: Class}
-
-\textbf{Class:} `\code{modelparty}'.
-
-FIXME
+\frametitle{Implementation: Parties}
+
+\textbf{Class:} `\code{modelparty}' inheriting from `\code{party}'.
+
+\medskip
+
+\textbf{Main addition:} Data handling for regressor and partitioning variables.
+\begin{itemize}
+  \item The \emph{Formula} package is used for Two-part formulas, e.g.,\\
+    \code{y ~ x1 + x2 | z1 + z2 + z3}.
+  \item The corresponding terms are stored for the combined
+    model and only for the partitioning variables.
+\end{itemize}
+
+\medskip
+
+\textbf{Reusability:} Could in principle be used for other model trees as well (inferred
+by other algorithms than MOB).
+
+\medskip
+
+\textbf{Additional information:} In \code{info} slots of `\code{party}' and `\code{partynode}'.
+\begin{itemize}
+  \item \code{call}, \code{formula}, \code{Formula}, \code{terms} (partitioning variables only),
+    \code{fit}, \code{control}, \code{dots}, \code{nreg}.
+  \item \code{coefficients}, \code{objfun}, \code{object}, \code{nobs}, \code{p.value}, \code{test}.
+\end{itemize}
 
@@
 
-\footnotesize
+\scriptsize
 
@@
  \textit{Parties, Models, Mobsters: A New Implementation of Model-Based Recursive Partitioning in R.}
- \texttt{vignette("mob",~package~=~"partykit")}.
+ \texttt{vignette("mob",~package~=~"partykit")}
+
+\medskip
+
+Zeileis A, Croissant Y (2010).
+  \dquote{Extended Model Formulas in R: Multiple Parts and Multiple Responses.}
+  \textit{Journal of Statistical Software}, \textbf{34}(1), 1--13. 
+  URL~\url{http://www.jstatsoft.org/v34/i01/}
 
@@
 
-%% \medskip
-%% 
-%% Rusch T, Lee I, Hornik K, Jank W, Zeileis A (2013).
-%%   \dquote{Influencing Elections with Statistics: Targeting Voters with Logistic Regression Trees.}
-%%   \textit{The Annals of Applied Statistics}, \textbf{7}(3), 1612--1639.
-%%   \doi{10.1214/13-AOAS648}
+\medskip
+
+Rusch T, Lee I, Hornik K, Jank W, Zeileis A (2013).
+  \dquote{Influencing Elections with Statistics: Targeting Voters with Logistic Regression Trees.}
+  \textit{The Annals of Applied Statistics}, \textbf{7}(3), 1612--1639.
+  \doi{10.1214/13-AOAS648}
 

### 2014-03-17 pkg/partykit/inst/NUS-2014/modelparty.Rnw
further improvements and touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@491 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \textbf{Example:} Logistic regression, assessing differences in
-the effect of ``preferential treatment'' (``women and children first''?).
-
-\bigskip
-
-\textbf{Result:} Log-odds ratio of survival given treatment differs across
-classes (slope), as does the survival probability of male adults (intercept).
+the effect of ``preferential treatment'' (``women and children first''?)
+in the Titanic survival data.
 
@@
 
+\bigskip
+
+\textbf{Result:} Log-odds ratio of survival given treatment differs across
+classes (slope), as does the survival probability of male adults (intercept).
+
 \end{frame}
@@
 \textbf{Parameter estimation:} $\widehat \theta$ by optimization of
-objective function $\Psi(y, x, \theta)$ for $n$ observations $y_i$ ($i = 1, \dots, n$):
+additive objective function $\Psi(y, x, \theta)$ for $n$ observations $y_i$ ($i = 1, \dots, n$):
 
@@
 w.r.t.\ to an ordering $\sigma(z_{ij})$.
-
 \[
@@
 
-\textbf{Model:} Paired comparison via Bradley-Terry.
-\begin{itemize}
-  \item Paired comparisons of attractiveness for
-    \emph{Germany's Next Topmodel~2007} finalists:
+\textbf{Data:} Paired comparisons of attractiveness.
+\begin{itemize}
+  \item \emph{Germany's Next Topmodel~2007} finalists:
     Barbara, Anni, Hana, Fiona, Mandy, Anja.
@@
 
+\end{frame}
+
+\begin{frame}
+\frametitle{Bradley-Terry trees}
+
+\textbf{Model:} Bradley-Terry (or Bradley-Terry-Luce) model.
+\begin{itemize}
+  \item Standard model for paired comparisons in social sciences.
+  \item Parametrizes probability $\pi_{ij}$ for preferring object $i$ over $j$
+    in terms of corresponding ``ability'' or ``worth'' parameters $\theta_i$.
+    \begin{eqnarray*}
+          \pi_{ij}\phantom{)}  & = & \frac{\theta_i}{\theta_i + \theta_j} \\
+      \mathsf{logit}(\pi_{ij}) & = & \log(\theta_i) - \log(\theta_j)
+    \end{eqnarray*}
+  \item Maximum likelihood as a logistic or log-linear GLM.
+\end{itemize}
+
 \bigskip
@@
 
-\textbf{Here:} Use \code{mob()} directly to build model from scratch.
+\textbf{Here:} Use \code{mob()} directly to build model from scratch using \code{btReg.fit()} from \emph{psychotree}.
 

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
rename
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@470 255f4249-2263-4417-b1ea-8498779e27a7
nodeprune example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@476 255f4249-2263-4417-b1ea-8498779e27a7
added better pictures from M&S 1963 added party tree for M&S 1963 further touch-ups and improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@494 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-02-11 pkg/partykit/inst/NUS-2014/constparty_TH_AZ.Rnw
@@
 
-%% need no 
-\usepackage{Sweave}
+%% instead of \usepackage{Sweave}
 \definecolor{InputColor}{rgb}{0,0,0.3}
 \definecolor{OutputColor}{rgb}{0.2,0.2,0.2}
+\setkeys{Gin}{width=0.8\textwidth}
+\RequirePackage{fancyvrb,upquote}
+\renewcommand{\ttdefault}{laett}
+\DefineVerbatimEnvironment{Scode}{Verbatim}{}
+\DefineVerbatimEnvironment{Sinput}{Verbatim}{formatcom=\color{InputColor}}
+\DefineVerbatimEnvironment{Soutput}{Verbatim}{formatcom=\color{OutputColor}}
+\newenvironment{Schunk}{\fontsize{9}{10}\selectfont}{}
 
 %% commands
-\let\code\texttt
-\let\class\texttt
-\let\fct\texttt
-\let\doi\texttt
+\newcommand{\code}[1]{{\tt \small #1}}
+\let\class\code
+\let\fct\code
+\let\doi\code
 \newcommand{\dquote}[1]{``#1''}
@@
 <<rpart, eval=FALSE>>=
-rp <- rpart(Survived ~ Gender + Age + Class, 
-            data = ttnc)
+rp <- rpart(Survived ~ Gender + Age + Class, data = ttnc)
 @
@@
 <<pmml, eval=FALSE>>=
-pm <- pmmlTreeModel(file.path(system.file("pmml", 
-    package = "partykit"), "ttnc.pmml"))
+pm <- pmmlTreeModel(system.file("pmml", "ttnc.pmml",
+    package = "partykit"))
 @
@@
 <<cache>>=
-if(file.exists("slides-trees.rda")) load("slides-trees.rda") else {
+if(file.exists("constparty-trees.rda")) load("constparty-trees.rda") else {
 <<rpart>>
@@
 <<mytree>>
-save(rp, ct, ev, pm, mb, myttnc, file = "slides-trees.rda")
+save(rp, ct, ev, pm, mb, myttnc, file = "constparty-trees.rda")
 }
@@
 <<J48>>=
-j48 <- J48(Survived ~ Gender + Age + Class, 
-           data = ttnc)
+j48 <- J48(Survived ~ Gender + Age + Class, data = ttnc)
+@
+
+<<Morgen+Sonquist-1963>>=
+ms63d <- data.frame(
+  earnings = numeric(),
+  ethnicity = factor(, levels = 1:2, labels = c("other", "afam")),
+  age = numeric(),
+  occupation = factor(, levels = 1:2, labels = c("other", "farmer")),
+  education = factor(, levels = 1:4, labels = c("elementary", "highschool", "college", "university"), ordered = TRUE)
+)
+
+ms63p <- partynode(1L,
+  split = partysplit(2L, index = 1:2),
+  kids = list(
+    partynode(2L,
+      split = partysplit(3L, breaks = 65, right = FALSE),
+      kids = list(
+        partynode(3L,
+	  split = partysplit(4L, index = 1:2),
+	  kids = list(
+	    partynode(4L,
+	      split = partysplit(5L, breaks = 3L, right = FALSE),
+	      kids = list(
+	        partynode(5L,
+		  split = partysplit(5L, breaks = 2L, right = FALSE),
+		  kids = list(
+		    partynode(6L, info = "$5000"),
+		    partynode(7L, info = "$6500")
+		  )
+		),
+		partynode(8L,
+		  split = partysplit(3L, breaks = 45, right = FALSE),
+		  kids = list(
+		    partynode(9L, info = "$7000"),
+		    partynode(10L, info = "$9000")
+		  )
+		)
+	      )
+	    ),
+	    partynode(11L, info = "$4500")
+	  )
+	),
+	partynode(12L,
+	  split = partysplit(5L, breaks = 2L, right = FALSE),
+	  kids = list(
+	    partynode(13L, info = "$2200"),
+	    partynode(14L, info = "$4500")
+	  )
+	)
+      ),
+      info = "$5500"
+    ),
+    partynode(15L,
+      split = partysplit(5L, breaks = 2L, right = FALSE),
+      kids = list(
+        partynode(16L, info = "$2500"),
+	partynode(17L, info = "$5000")
+      )
+    )
+  ),
+  info = "$5000"
+)
+ms63p <- party(ms63p, ms63d)
 @
@@
 \begin{frame}
+\frametitle{Celebrating 50th anniversary}
+
+\begin{center}
+\includegraphics{aid-abstract}
+\end{center}
+JASA $\bold{58}(302)$ June 1963 (well, there were no early online versions, so
+most people read the paper in 1964 for the first time). \\
+
+We need a \textbf{partykit}!
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Celebrating 50th anniversary}
+
+\vspace*{-0.5cm}
+\begin{center}
+\includegraphics[width=\textwidth]{aid-tree}
+\end{center}
+
+\end{frame}
+
+\begin{frame}
 \frametitle{Overview}
@@
   \item Here and now: Intro to partykit for trees with constant fits in each node.
-  \item Future: Next steps
 \end{itemize}
@@
 \textbf{Challenge:} For implementing new algorithms in R, code
-is required not only for fitting the tree model on the learning data
+is required not only for fitting the tree model
 but also
@@
         fits in each terminal node.
-  \item Nodes are recursive objects, i.e., a node can contain child nodes.
-  \item Keep (learning) data out of the recursive node and split structure.
-  \item Basic printing, plotting, and predicting for raw node structure.
-  \item Customization via suitable panel or panel-generating functions.
-  \item Coercion from existing objects (rpart, J48, etc.) to new class.
+  \item Nodes are recursive objects: nodes can contain child nodes.
+  \item Keep data out of the recursive node and split structure.
+  \item Basic print, plot, and predict for raw node structure.
+  \item Customization via panel or panel-generating functions.
+  \item Coercion from existing objects (e.g., rpart) to new class.
   \item Use simple/fast S3 classes and methods.
@@
     corresponding (learning) data (optionally without any rows) and
-    \code{fitted} the corresponding fitted nodes.
+    \code{fitted} the corresponding fitted node ids.
 \end{itemize}
@@
 optionally arbitrary information can be stored.
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{partykit: Base classes}
+
+<<ms63-party, echo=TRUE, results=verbatim>>=
+ms63d
+pn <- partynode(1L,
+  split = partysplit(2L, index = 1:2),
+  kids = list(
+    partynode(2L, info = "$5500"),
+    partynode(3L, info = "$3000")
+  )
+)
+py <- party(pn, ms63d)
+print(py)
+plot(py)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{partykit: Base classes}
+
+<<ms63-plot1, fig=TRUE>>=
+plot(py)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{partykit: Base classes}
+
+\setkeys{Gin}{width=1.15\textwidth}
+<<ms63-plot2, fig=TRUE, height=8, width=14>>=
+plot(ms63p, drop = TRUE)
+@
 
@@
 nd <- ttnc[rep(1:nrow(ttnc), 100), ]
-system.time(p1 <- predict(rp, newdata = nd, 
-                          type = "class"))
+system.time(p1 <- predict(rp, newdata = nd, type = "class"))
 system.time(p2 <- predict(as.party(rp), newdata = nd))
@@
 exported by various software packages including SAS, SPSS, R/pmml.
-Here, import QUEST tree.
+Here, import QUEST tree from SPSS.
 
@@
 
-\begin{frame}[fragile]
+\begin{frame}
 \frametitle{mytree}
@@
 
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{OK, fire away!}
+
 <<Titanic-mytree-vis, eval = FALSE, echo = TRUE>>=
-plot(mytree(Survived ~ Class + Age + Gender, 
-            data = ttnc))
+plot(mytree(Survived ~ Class + Age + Gender, data = ttnc))
 @
@@
 plot(myttnc)
+@
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Too big a tree? Prune!}
+
+\setkeys{Gin}{width=\textwidth}
+\hspace*{-0.4cm}%
+<<Titanic-mytree-prune, echo = TRUE, fig = TRUE, height=8, width=12>>=
+plot(nodeprune(myttnc, 10))
 @

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
touch-ups in graphics, added some FIXMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@495 255f4249-2263-4417-b1ea-8498779e27a7

@@
   earnings = numeric(),
-  ethnicity = factor(, levels = 1:2, labels = c("other", "afam")),
+  ethnicity = factor(, levels = 1:2, labels = c("cauc", "afam")),
   age = numeric(),
@@
 
+\setkeys{Gin}{width=0.65\textwidth}
+\begin{center}
 <<ms63-plot1, fig=TRUE>>=
@@
 @
+\end{center}
 
@@
 
-\setkeys{Gin}{width=1.15\textwidth}
+\setkeys{Gin}{width=1.2\textwidth}
+\hspace*{-1cm}%
 <<ms63-plot2, fig=TRUE, height=8, width=14>>=
@@
 
-\vspace*{-0.5cm}
+\vspace*{-0.2cm}
 
@@
 \hspace*{-0.4cm}%
-<<Titanic-pmml-visualization, fig=TRUE, height=10, width=16>>=
+<<Titanic-pmml-visualization, fig=TRUE, height=12, width=17>>=
 plot(pm)
@@
 
+FIXME: partykit still only on R-Forge
+
 \bigskip
@@
 as explained in the \code{constparty} package vignette.
+
+FIXME: Show code for completeness?
+
+FIXME: Change/update/enhance references?
 
@@
  \textit{partykit: A Toolkit for Recursive Partytioning.}
- R~package vignette version~0.2-0.
+ R~package vignette version~0.8-0.
  URL~\url{http://CRAN.R-project.org/package=partykit}

### 2014-03-18 pkg/partykit/R/simpleparty.R
correct and test simpleparty with empty nodes bug
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@496 255f4249-2263-4417-b1ea-8498779e27a7

@@ -133,6 +133,11 @@ as.simpleparty.constparty <- function(obj, ...) {
   FUN <- function(node, fitted) {
     fitted <- subset(fitted,
       fitted[["(fitted)"]] %in% nodeids(node, terminal = TRUE))
+
+    if (nrow(fitted) == 0)
+      return(list(prediction = NA, n = 0,
+                  error = NA, distribution = NULL))
+
     y <- fitted[["(response)"]]
     w <- fitted[["(weights)"]]
     if(is.null(w)) {
@@ -167,7 +172,8 @@ as.simpleparty.constparty <- function(obj, ...) {
     nlevels <- levels(fit)
     for(i in 1:length(idlist)) nlevels[match(idlist[[i]], levels(fit))] <- i
     levels(fit) <- nlevels
-    as.numeric(factor(fit))
+    ret <- factor(as.numeric(as.character(fit)), labels = 1:length(idlist), levels = 1:length(idlist))
+    ret
   }
 
   ## cycle through node
@@ -177,7 +183,7 @@ as.simpleparty.constparty <- function(obj, ...) {
       info = FUN(onode, fitted)))
     kids <- kids_node(onode)
     kids_tid <- lapply(kids, nodeids, terminal = TRUE)
-    kids_fitted <- base::split.data.frame(fitted, fit2id(fitted[["(fitted)"]], kids_tid))
+    kids_fitted <- base::split.data.frame(fitted, fit2id(fitted[["(fitted)"]], kids_tid), drop = FALSE)
     partynode(id = onode$id, split = onode$split,
       kids = lapply(1:length(kids), function(i) new_node(kids[[i]], kids_fitted[[i]])),
       surrogates = onode$surrogates,

### 2014-03-18 pkg/partykit/tests/bugfixes.R

@@ -720,3 +720,47 @@ p2 <- sapply(names(p), function(n)
     teststat = "quad")))
 stopifnot(max(abs(p - p2)) < sqrt(.Machine$double.eps))
 
+### check coersion of constparties to simpleparties
+### containing terminal nodes without corresponding observations
+## create party
+data("WeatherPlay", package = "partykit")
+py <- party(
+  partynode(1L,
+    split = partysplit(1L, index = 1:3),
+    kids = list(
+      partynode(2L,
+        split = partysplit(3L, breaks = 75),
+        kids = list(
+          partynode(3L, info = "yes"),
+          partynode(4L, info = "no"))),
+      partynode(5L, 
+        split = partysplit(3L, breaks = 20),
+        kids = list(
+          partynode(6L, info = "no"),
+          partynode(7L, info = "yes"))),
+      partynode(8L,
+        split = partysplit(4L, index = 1:2),
+        kids = list(
+          partynode(9L, info = "yes"),
+          partynode(10L, info = "no"))))),
+  WeatherPlay)
+names(py) <- LETTERS[nodeids(py)]
+
+pn <- node_party(py)
+cp <- party(pn,
+  data = WeatherPlay,
+  fitted = data.frame(
+    "(fitted)" = fitted_node(pn, data = WeatherPlay),
+    "(response)" = WeatherPlay$play,
+    check.names = FALSE),
+  terms = terms(play ~ ., data = WeatherPlay),
+)
+print(cp)
+cp <- as.constparty(cp)
+
+nd <- data.frame(outlook = factor("overcast", levels = levels(WeatherPlay$outlook)), 
+                 humidity = 10, temperature = 10, windy = "yes")
+predict(cp, type = "node", newdata = nd)
+predict(cp, type = "response", newdata = nd)
+as.simpleparty(cp)
+print(cp)

### 2014-03-18 pkg/partykit/R/plot.R
avoid warnings when there are no p-values
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@497 255f4249-2263-4417-b1ea-8498779e27a7

@@ -16,7 +16,7 @@ node_inner <- function(obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "
 
     ## FIXME: make more flexible rather than special-casing p-value
     if(pval) {
-      pval <- suppressWarnings(try(!is.null(info_node(node)$p.value)))
+      pval <- suppressWarnings(try(!is.null(info_node(node)$p.value), silent = TRUE))
       pval <- if(inherits(pval, "try-error")) FALSE else pval
     }
     if(pval) {

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
forgot one info slot in M&S tree
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@498 255f4249-2263-4417-b1ea-8498779e27a7

@@
 	partynode(17L, info = "$5000")
-      )
+      ),
+      info = "$3000"
     )

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
cosmetics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@499 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-We need a \textbf{partykit}!
+We need a partykit!
 
@@
 
-\textbf{Answer:} The R package \emph{partykit} provides unified
+\textbf{Answer:} The R package partykit provides unified
 infrastructure for recursive partytioning, especially
@@
 
-\textbf{CART:} Apply \fct{rpart} to preprocessed \code{ttnc} data (see also below).
+\textbf{CART:} Apply \fct{rpart} to preprocessed \code{ttnc} data (see \code{constparty} vignette in partykit).
 
@@
 
-\textbf{Software:} All examples have been produced with
+All examples have been produced with
 R~\Sexpr{with(R.Version(), paste(major, minor, sep = "."))} and packages
-partykit~\Sexpr{packageDescription("partykit")$Version},
-rpart~\Sexpr{packageDescription("rpart")$Version},
-RWeka~\Sexpr{packageDescription("RWeka")$Version},
-evtree~\Sexpr{packageDescription("evtree")$Version}. All packages are
+\begin{itemize}
+\item partykit~\Sexpr{packageDescription("partykit")$Version},
+\item rpart~\Sexpr{packageDescription("rpart")$Version},
+\item  RWeka~\Sexpr{packageDescription("RWeka")$Version},
+\item evtree~\Sexpr{packageDescription("evtree")$Version}. 
+\end{itemize}
+All packages are
 freely available under the GPL from \url{http://CRAN.R-project.org/}.
@@
 FIXME: partykit still only on R-Forge
-
-\bigskip
-
-\textbf{Data:} The contingency table \code{Titanic} is transformed
-as explained in the \code{constparty} package vignette.
-
-FIXME: Show code for completeness?
-
-FIXME: Change/update/enhance references?
 
@@
 
-\bigskip
-
-Zeileis A, Hothorn T, Hornik K (2008).
- \dquote{Model-Based Recursive Partitioning.}
-  \textit{Journal of Computational and Graphical Statistics},
-  \textbf{17}(2), 492--514.
-  \doi{10.1198/106186008X319331}
-
 \end{frame}

### 2014-03-18 pkg/partykit/R/node.R
be nice to our users (NHEMOtree this time): coerse non-integer id
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@500 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,8 +1,11 @@
 
 partynode <- function(id, split = NULL, kids = NULL, surrogates = NULL, info = NULL) {
 
-    if (!is.integer(id) || length(id) != 1)
-        stop(sQuote("id"), " ", "must be a single integer")
+    if (!is.integer(id) || length(id) != 1) {
+        id <- as.integer(id0 <- id)
+        if (any(is.na(id)) || !isTRUE(all.equal(id0, id)) || length(id) != 1)
+            stop(sQuote("id"), " ", "must be a single integer")
+    }
 
     if (is.null(split) != is.null(kids)) {
         stop(sQuote("split"), " ", "and", " ", sQuote("kids"), " ", 

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
improved M&S example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@501 255f4249-2263-4417-b1ea-8498779e27a7

@@
 <<ms63-party, echo=TRUE, results=verbatim>>=
-ms63d
+str(ms63d)
 pn <- partynode(1L,
@@
 <<ms63-plot2, fig=TRUE, height=8, width=14>>=
-plot(ms63p, drop = TRUE)
+plot(ms63p, drop = TRUE, ep_args = list(justmin = 10))
 @

### 2014-03-18 pkg/partykit/R/plot.R
added different types of justification to edge_simple()
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@502 255f4249-2263-4417-b1ea-8498779e27a7

@@ -166,20 +166,39 @@ node_terminal <- function(obj,
 }
 class(node_terminal) <- "grapcon_generator"
 
-edge_simple <- function(obj, digits = 3, abbreviate = FALSE)
+edge_simple <- function(obj, digits = 3, abbreviate = FALSE,
+  justmin = Inf, just = c("alternate", "increasing", "decreasing", "equal"))
 {
   meta <- obj$data
 
+  justfun <- function(i, split) {
+    myjust <- if(mean(nchar(split)) > justmin) {
+      match.arg(just, c("alternate", "increasing", "decreasing", "equal"))
+    } else {
+      "equal"
+    }
+    k <- length(split)
+    rval <- switch(myjust,
+      "equal" = rep.int(0, k),
+      "alternate" = rep(c(0.5, -0.5), length.out = k),
+      "increasing" = seq(from = -k/2, to =  k/2, by = 1),
+      "decreasing" = seq(from =  k/2, to = -k/2, by = -1)
+    )
+    unit(0.5, "npc") + unit(rval[i], "lines")
+  }
+
   ### panel function for simple edge labelling
   function(node, i) {
-    split <- character_split(split_node(node), meta)$levels[i]
+    split <- character_split(split_node(node), meta)$levels
+    y <- justfun(i, split)
+    split <- split[i]
     # try() because the following won't work for split = "< 10 Euro", for example.
     if(any(grep(">", split) > 0) | any(grep("<", split) > 0)) {
       tr <- suppressWarnings(try(parse(text = paste("phantom(0)", split)), silent = TRUE))
       if(!inherits(tr, "try-error")) split <- tr
     }
-    grid.rect(gp = gpar(fill = "white", col = 0), width = unit(1, "strwidth", split)) 
-    grid.text(split, just = "center")
+    grid.rect(y = y, gp = gpar(fill = "white", col = 0), width = unit(1, "strwidth", split))
+    grid.text(split, y = y, just = "center")
   }
 }
 class(edge_simple) <- "grapcon_generator"

### 2014-03-18 pkg/partykit/inst/NUS-2014/modelparty.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@492 255f4249-2263-4417-b1ea-8498779e27a7
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@505 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \begin{itemize}
-  \item The \emph{Formula} package is used for Two-part formulas, e.g.,\\
+  \item The \emph{Formula} package is used for two-part formulas, e.g.,\\
     \code{y ~ x1 + x2 | z1 + z2 + z3}.
@@
 
-\textbf{Reusability:} Could in principle be used for other model trees as well (inferred
-by other algorithms than MOB).
-
-\medskip
-
 \textbf{Additional information:} In \code{info} slots of `\code{party}' and `\code{partynode}'.
@@
 \end{itemize}
+
+\medskip
+
+\textbf{Reusability:} Could in principle be used for other model trees as well (inferred
+by other algorithms than MOB).
 
@@
  \textit{partykit: A Toolkit for Recursive Partytioning.}
- R~package version~0.2-0.
+ R~package version~0.8-0.
  URL~\url{http://R-Forge.R-project.org/projects/partykit/}

### 2014-03-19 pkg/partykit/inst/NUS-2014/constparty.Rnw
remove fixme
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@504 255f4249-2263-4417-b1ea-8498779e27a7
final
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@506 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\begin{frame}
+\begin{frame}[fragile]
 \frametitle{findsplit}
 
-R code
+\begin{Schunk}
+\begin{Sinput}
+findsplit <- function(response, data, weights) {
+  ### response: name of the response variable   
+  ### data: data.frame with all variables
+  ### weights: case weights for current node
+
+  ### ... some computations
+
+  ### return split as partysplit object
+  return(partysplit(
+      varid = as.integer(xselect),         ### which variable?
+      index = levels(data[[xselect]]) %in%
+              splitpoint + 1L,             ### which split point?
+      info = list(p.value = exp(logpmin)   ### save p-value
+  )))
+}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{frame}
+\begin{frame}[fragile]
 \frametitle{growtree}
 
-R code
+\begin{Schunk}
+\begin{Sinput}
+growtree <- function(id = 1L, response, data, weights) {
+  ### recursive function of
+  ### id: node identifier
+  ### response: name of the response variable 
+  ### data: data.frame with all variables 
+  ### weights: case weights for current node
+
+  ### ... some computations, call to findsplit()
+  ### and growtree()
+
+  ### return nodes  
+  return(partynode(id = as.integer(id), split = sp, kids = kids))
+}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{frame}
+\begin{frame}[fragile]
 \frametitle{mytree}
 
-R code
+\begin{Schunk}
+\begin{Sinput}
+mytree <- function(formula, data, weights = NULL) {
+  ### formula: a model formula
+  ### data: data.frame with all variables 
+  ### weights: case weights for root node
+
+  ### ... some computations
+
+  nodes <- growtree(id = 1L, response, data, weights)
+
+  ### ... some computations
+
+  ### return rich constparty object
+  ret <- party(nodes, data = data,
+    fitted = data.frame(
+      "(fitted)" = fitted,
+      "(response)" = data[[response]],
+      "(weights)" = weights,
+      check.names = FALSE), 
+    terms = terms(formula)) 
+  as.constparty(ret)
+}  
+\end{Sinput}
+\end{Schunk}
 
@@
 All packages are
-freely available under the GPL from \url{http://CRAN.R-project.org/}.
-
-FIXME: partykit still only on R-Forge
+freely available under the GPL from \url{http://R-forge.R-project.org/} (partykit~\Sexpr{packageDescription("partykit")$Version})
+or \url{http://CRAN.R-project.org/}.
 
@@
  R~package vignette version~0.8-0.
- URL~\url{http://CRAN.R-project.org/package=partykit}
+ URL~\url{https://r-forge.r-project.org/projects/partykit}
 

### 2014-03-19 pkg/partykit/inst/mloss/mloss2.Rnw
start assembling 4-page paper for MLOSS 2014
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@509 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[a4paper]{article}
\usepackage{jmlr2e}
\usepackage{color,hyperref}

%% software markup
\let\proglang=\textsf
\let\pkg=\textit
\let\code=\texttt
\newcommand{\squote}[1]{`{#1}'}
\newcommand{\dquote}[1]{``{#1}''}
\newcommand{\fct}[1]{\texttt{#1()}}
\newcommand{\class}[1]{\squote{\texttt{#1}}}

%% \usepackage{Sweave} is essentially
\RequirePackage[T1]{fontenc}
\RequirePackage{ae,fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\setkeys{Gin}{width=.7\textwidth}
\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE, echo=FALSE, results=hide}
<<options, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", digits = 4, width = 70)
library("partykit")
set.seed(290875)
@

%% header declarations
\author{Torsten Hothorn \\ \addr{Universit\"at Z\"urich} \\ \addr{E-mail: Torsten.Hothorn@R-project.org} \AND
        Achim Zeileis \\ \addr{Universit\"at Innsbruck} \\ \addr{E-mail: Achim.Zeileis@R-project.org}}
\title{\pkg{partykit}: An Open-Source Toolkit for Recursive Partytioning in \proglang{R}}
\editor{(submission for MLOSS track)}

% hyperref setup
\definecolor{Red}{rgb}{0.5,0,0}
\definecolor{Blue}{rgb}{0,0,0.5}
\hypersetup{%
  pdftitle = {partykit: An Open-Source Toolkit for Recursive Partytioning in R},
  pdfsubject = {submitted to JMLR-MLOSS},
  pdfkeywords = {open-source software, recursive partitioning, decision trees, statistical learning},
  pdfauthor = {Torsten Hothorn, Achim Zeileis},
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

\begin{document}

\maketitle

\begin{abstract}
  The \pkg{partykit} package provides a flexible toolkit with infrastructure for
  learning, representing, summarizing, and visualizing a wide range of tree-structured
  regression and classification models. The functionality encompasses: (a)~Basic
  infrastructure for \emph{representing} trees (inferred by any algorithm) so that
  unified \code{print}/\code{plot}/\code{predict} methods are available.
  (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
  along with suitable coercion functions to create such tree models (e.g., by
  \pkg{rpart}, \pkg{RWeka}, PMML). (c)~A reimplementation of \emph{conditional inference
  trees} (\code{ctree}, originally provided in the \pkg{party} package). (d)~An extended
  reimplementation of \emph{model-based recursive partitioning} (\code{mob}, also originally
  in \pkg{party}) along with dedicated methods for trees with parametric models in the
  leaves. Here, a brief overview of the package and its design is given while more detailed
  discussions of items~(a)--(d) are available in vignettes accompanying the package.
\end{abstract}

\begin{keywords}
open-source software, recursive partitioning, regression trees, classification trees, decision trees, statistical learning. 
\end{keywords}

\section{Overview}

In the more than fifty years since \cite{Morgan+Sonquist:1963}
published their seminal paper on ``automatic interaction detection'',
a wide range of methods has been suggested that is usually
termed ``recursive partitioning'' or ``decision trees'' or
``tree(-structured) models'' etc. Particularly influential were the
algorithms CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
and CTree \citep{Hothorn+Hornik+Zeileis:2006} among many others
\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
of conceptual algorithms, a wide range of computational implementations
in various software systems emerged: Typically the original authors of an
algorithms also provide accompanying software but many software systems,
e.g., including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
also provide collections of various types of trees.
Within \proglang{R} the list of prominent packages includes
\pkg{rpart} \citep[implementing the CART algorithm]{rpart},
\pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
\pkg{party} \citep[implementing CTree and MOB]{party} among
many others. See the CRAN task view ``Machine Learning''
\citep{ctv} for an overview.

All of these algorithms and software implementations have to deal with
very similar challenges. However, due to the fragmentation of the
communities in which the corresponding research is published -- ranging
from statistics over machine learning to various applied fields -- many
discussions of the algorithms do not reuse established theoretical results
and terminology. Similarly, there is no common ``language'' for the
software implementations and different solutions are provided by
different packages (even within \proglang{R}) with relatively little
reuse of code.

The \pkg{partykit} tries to address the latter point and improve the
computational situation by providing a common unified infrastructure
for recursive partytioning in the \proglang{R} system for statistical
computing. In particular, \pkg{partykit}
provides tools for representing fitted trees along with printing,
plotting, and computing predictions. The design principles are:
\begin{itemize}
  \item One `agnostic' base class (\class{party}) which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits (\class{constparty}) or with parametric models (\class{modelparty})
	in each terminal node (or leaf).
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing objects (\code{rpart}, \code{J48}, etc.) to the new class.
  \item Usage of simple/fast \proglang{S}3 classes and methods.
\end{itemize}
In addition to all of this generic infrastructure, two specific tree
algorithms are implemented in \pkg{partykit} as well: \fct{ctree}
for conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}
and \fct{mob} for model-based recursive partitioning \citep{Zeileis+Hothorn+Hornik:2008}.

\section{Installation and Documentation}

installation information

R-Forge

The vignette (\code{"partykit"}) introduces the basic \class{party} class
and associated infrastructure while three further vignettes discuss the tools
built on top of it: \code{"constparty"} covers the eponymous class for
constant-fit trees along with suitable coercion functions, and
\code{"ctree"} and \code{"mob"} discuss the new \fct{ctree} and
\fct{mob} implementations, respectively. Each of the vignettes can be viewed
within \proglang{R} via \code{vignette(}\emph{``name''}\code{, package = "partykit")}.

\section{User Interface}

formula interface

as.party

PMML

print, plot, predict

<<Titanic>>=
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
ttnc <- transform(ttnc, Treatment = factor(
  Gender == "Female" | Age == "Child", levels = c(FALSE, TRUE),
  labels = c("Normal\nMale&Adult", "Preferential\nFemale|Child")
))
@

<<rpart>>=
library("rpart")
ttnc_rp <- as.party(rpart(Survived ~ ., data = ttnc))
@

<<mob>>=
ttnc_mob <- glmtree(Survived ~ Treatment | Class + Gender + Age,
  data = ttnc, family = binomial, alpha = 0.05, prune = "BIC")
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\centering
<<mob-plot, echo=FALSE, fig=TRUE, height=7, width=18>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
plot(ttnc_rp, newpage = FALSE, pop = TRUE,
  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.8))))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
plot(ttnc_mob, newpage = FALSE, pop = TRUE, tp_args = list(ylines = 1.3,
  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.8))),
  linecol = "black"))
popViewport(2)
@
\caption{\label{fig} Tree visualizations of survival on Titanic:
\class{rpart} tree visualized by \pkg{partykit} (left) and
logistic-regression-based tree fitted by \fct{glmtree} (right).}
\end{figure}

\section{Internal Structure}

partysplit
partynode
party

constparty
modelparty

extractor functions

\section{Conclusions}

run-time, memory requirements, benchmarks?

\section*{Acknowledgments}

We are thankful to the organizers and participants of the ``Workshop on Classification and Regression Trees'' 2014,
sponsored by the Institute for Mathematical Sciences of the National University of Singapore,
for feedback and discussions.

\bibliography{../../vignettes/party.bib}

\end{document}

### 2014-03-23 pkg/partykit/R/party.R
add FUN argument to predict_party.default and use it for predict_party.simpleparty
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@523 255f4249-2263-4417-b1ea-8498779e27a7

@@ -291,7 +291,7 @@ predict_party <- function(party, id, newdata = NULL, ...)
     UseMethod("predict_party")
 
 ### do nothing expect returning the fitted ids
-predict_party.default <- function(party, id, newdata = NULL, ...) {
+predict_party.default <- function(party, id, newdata = NULL, FUN = NULL, ...) {
 
     if (length(list(...)) > 1) 
         warning("argument(s)", " ", sQuote(names(list(...))), " ", "have been ignored")
@@ -304,6 +304,10 @@ predict_party.default <- function(party, id, newdata = NULL, ...) {
       rownames(newdata)
     }
     if(length(nam) != length(id)) nam <- NULL
+
+    if (!is.null(FUN))
+        return(.simplify_pred(nodeapply(party, 
+            nodeids(party, terminal = TRUE), FUN, by_node = TRUE), id, nam))
 
     ## special case: fitted ids
     return(structure(id, .Names = nam))

### 2014-03-23 pkg/partykit/R/simpleparty.R

@@ -93,15 +93,13 @@ predict_party.simpleparty <- function(party, id, newdata = NULL,
 
   ## predictions
   if(type == "response") {
-    .simplify_pred(nodeapply(party, nodeids(party),
-      function(x) x$info$prediction, by_node = TRUE), id, nam)
+    FUN <- function(x) x$info$prediction
   } else {
     if(is.null(node_party(party)$info$distribution)) stop("probabilities not available")
     scale <- any(node_party(party)$info$distribution > 1)
-    .simplify_pred(nodeapply(party, nodeids(party),
-      function(x) if(scale) prop.table(x$info$distribution) else x$info$distribution,
-      by_node = TRUE), id, nam)
+    FUN <- function(x) if(scale) prop.table(x$info$distribution) else x$info$distribution
   }
+  predict_party.default(party, id, nam, FUN = FUN, ...)
 }
 
 as.simpleparty <- function(obj, ...) UseMethod("as.simpleparty")

### 2014-03-23 pkg/partykit/R/cforest.R
add quantile and density predictions for constparties
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@525 255f4249-2263-4417-b1ea-8498779e27a7

@@ -182,12 +182,6 @@ predict.cforest <- function(object, newdata = NULL, type = c("weights", "respons
     ret
 }
 
-.pred_quantile <- function(probs = c(0.1, 0.5, 0.9), ...)
-    function(y, w) quantile(rep(y, w), probs = probs, ...)
-
-.pred_density <- function(y, w)
-    density(y, weights = w / sum(w))
-
 if (FALSE) {
 
 library("partykit")
@@ -198,14 +192,18 @@ p <- predict(cf, newdata = cars, type = "response")
 plot(dist ~ speed, data = cars)
 lines(cars$speed, unlist(p))
 
-p <- predict(cf, newdata = cars, type = "response", FUN = .pred_quantile())
+myquantile <- function(y, w) quantile(rep(y, w), probs = c(0.1, 0.5, 0.9))
+
+p <- predict(cf, newdata = cars, type = "response", FUN = myquantile)
 
 plot(dist ~ speed, data = cars)
 lines(cars$speed, p[,1])
 lines(cars$speed, p[,2])
 lines(cars$speed, p[,3])
 
-p <- predict(cf, newdata = cars, type = "response", FUN = .pred_density)
+mydensity <- partykit:::.pred_density
+
+p <- predict(cf, newdata = cars, type = "response", FUN = mydensity)
 
 for (i in 1:length(p)) 
     plot(p[[i]], xlim = c(-10, 150), ylim = c(0, .025))

### 2014-03-23 pkg/partykit/R/party.R

@@ -314,7 +314,9 @@ predict_party.default <- function(party, id, newdata = NULL, FUN = NULL, ...) {
 }
 
 predict_party.constparty <- function(party, id, newdata = NULL,
-    type = c("response", "prob", "node"), FUN = NULL, simplify = TRUE, ...)
+    type = c("response", "prob", "quantile", "density", "node"),
+    at = if (type == "quantile") c(0.1, 0.5, 0.9),
+    FUN = NULL, simplify = TRUE, ...)
 {
     ## extract fitted information
     response <- party$fitted[["(response)"]]
@@ -338,7 +340,7 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     if (is.data.frame(response)) {
         ret <- lapply(response, function(r) {
             ret <- .predict_party_constparty(node_party(party), fitted = fitted, 
-                response = r, weights, id = id, type = type, FUN = FUN, ...)
+                response = r, weights, id = id, type = type, at = at, FUN = FUN, ...)
             if (simplify) .simplify_pred(ret, id, nam) else ret
         })
         if (all(sapply(ret, is.atomic)))
@@ -349,7 +351,7 @@ predict_party.constparty <- function(party, id, newdata = NULL,
 
     ### univariate response
     ret <- .predict_party_constparty(node_party(party), fitted = fitted, response = response, 
-        weights = weights, id = id, type = type, FUN = FUN, ...)
+        weights = weights, id = id, type = type, at = at, FUN = FUN, ...)
     if (simplify) .simplify_pred(ret, id, nam) else ret[as.character(id)]
 }
 
@@ -395,20 +397,38 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     }
 }
 
+.pred_quantile <- function(y, w) {
+    y <- rep(y, w)
+    function(p, ...) quantile(y, probs = p, ...)
+}
+
+.pred_density <- function(y, w)
+    density(y, weights = w / sum(w))
+
 ### workhorse: compute predictions based on fitted / response data
 .predict_party_constparty <- function(node, fitted, response, weights,
-    id = id, type = c("response", "prob"), FUN = NULL, ...) {
+    id = id, type = c("response", "prob", "quantile", "density"),
+    at = if (type == "quantile") c(0.1, 0.5, 0.9), FUN = NULL, ...) {
 
+    type <- match.arg(type)
     if (is.null(FUN)) {
 
         rtype <- class(response)[1]
         if (rtype == "ordered") rtype <- "factor"    
[…]

### 2014-03-23 pkg/partykit/tests/regtest-party.R

@@ -114,3 +114,15 @@ table(predict(ct, type = "node"))
 nodeprune(ct, "B")
 nodeprune(ct, "C")
 nodeprune(ct, "A")
+
+### check different predict flavours for numeric responses
+x <- runif(100)
+dd <- data.frame(y = rnorm(length(x), mean = 2 * (x < .5)), x = x)
+ct <- ctree(y ~ x, data = dd)
+nd <- data.frame(x = (1:9) / 10)
+predict(ct, newdata = nd, type = "node")
+predict(ct, newdata = nd, type = "response")
+predict(ct, newdata = nd, type = "prob")
+predict(ct, newdata = nd, type = "quantile")
+predict(ct, newdata = nd, type = "quantile", at = NULL)
+predict(ct, newdata = nd, type = "density")

### 2014-03-23 pkg/partykit/R/simpleparty.R
fixed typo and bug
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@528 255f4249-2263-4417-b1ea-8498779e27a7

@@ -107,11 +107,13 @@ as.simpleparty <- function(obj, ...) UseMethod("as.simpleparty")
 as.simpleparty.simpleparty <- function(obj, ...) obj
 
 as.simpleparty.party <- function(obj, ...) {
-  if (is.simpleparty(obj)) 
+  if (is.simpleparty(obj)) {
+      class(obj) <- unique(c("simpleparty", class(obj)))
       return(obj)
+  }
   if (is.constparty(obj)) 
       return(as.simpleparty(as.constparty(obj)))
-  stop("can't coerse objects of class ", sQuote(class(obj)), 
+  stop("cannot coerce objects of class ", sQuote(class(obj)), 
        " to class ", sQuote("simpleparty"))
 }
 

### 2014-03-23 pkg/partykit/R/party.R
as.constparty now coerces to party first (if necessary)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@530 255f4249-2263-4417-b1ea-8498779e27a7

@@ -96,6 +96,7 @@ is.constparty <- function(party) {
 }
 
 as.constparty <- function(obj, ...) {
+    if(!inherits(obj, "party")) obj <- as.party(obj)
     if (!is.constparty(obj)) {
         if(is.null(obj$fitted))
 	  obj$fitted <- data.frame("(fitted)" = predict(obj, type = "node"), check.names = FALSE)

### 2014-03-24 pkg/partykit/R/ctree.R
2 surrogate splits by default
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@541 255f4249-2263-4417-b1ea-8498779e27a7

@@ -228,7 +228,7 @@
 ctree_control <- function(teststat = c("quad", "max"),
     testtype = c("Bonferroni", "Univariate", "Teststatistic"),
     mincriterion = 0.95, minsplit = 20L, minbucket = 7L, minprob = 0.01,
-    stump = FALSE, maxsurrogate = 0L, mtry = Inf, maxdepth = Inf, 
+    stump = FALSE, maxsurrogate = 2L, mtry = Inf, maxdepth = Inf, 
     multiway = FALSE, splittry = 2L) {
 
     teststat <- match.arg(teststat)

### 2014-03-24 pkg/partykit/tests/ctree-ctree.R
remove meaningless configuration
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@542 255f4249-2263-4417-b1ea-8498779e27a7

@@ -4,7 +4,6 @@
 
 library("partykit")
 library("party")
-library("parallel")
 
 set.seed(29)
 
@@ -48,6 +47,9 @@ gr[["response"]] <- as.character(gr[["response"]])
 gr[["teststat"]] <- as.character(gr[["teststat"]])
 gr[["testtype"]] <- as.character(gr[["testtype"]])
 
+### random splits are done differently in party::ctree
+gr <- subset(gr, !na | (na & maxsurrogate > 0))
+
 
 fun <- function(args) {
 
@@ -70,7 +72,7 @@ fun <- function(args) {
     list(error = max(abs(unlist(old) - unlist(new))), time = c(ot, nt))
 }
 
-ret <- mclapply(1:nrow(gr), function(i) { print(i); fun(gr[i, ,drop = FALSE]); })
+ret <- lapply(1:nrow(gr), function(i) { print(i); fun(gr[i, ,drop = FALSE]); })
 
 tm <- t(sapply(ret, function(x) x$time))
 err <- sapply(ret, function(x) x$error)
@@ -151,3 +153,6 @@ tree2
 (w2 <- where(tree2))
 
 stopifnot(max(abs(w1 - w2)) == 0)
+
+gr <- cbind(gr, err = err)
+subset(gr, !na & err > 0.01)

### 2014-03-24 pkg/partykit/tests/ctree-ctree.R
comment
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@543 255f4249-2263-4417-b1ea-8498779e27a7

@@ -156,3 +156,7 @@ stopifnot(max(abs(w1 - w2)) == 0)
 
 gr <- cbind(gr, err = err)
 subset(gr, !na & err > 0.01)
+
+### most of the differences come from the fact that in the absence of
+### a surrogate splits observations are randomly assigned to daugther nodes
+### slightly differently in the two implementations

### 2014-03-24 pkg/partykit/R/ctree.R
implement majority, maxsurrogate = 0 again
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@544 255f4249-2263-4417-b1ea-8498779e27a7

@@ -200,6 +200,10 @@
     ret$info <- list(criterion = p, p.value = fmP(p)[iselp])
     thissurr <- NULL
     kidids <- kidids_node(ret, data)
+    prob <- prop.table(table(kidids))
+    if (ctrl$majority)  ### go with majority
+        prob <- numeric(0) + 1:length(prob) %in% which.max(prob)
+    ret$prob <- prob
 
     if (ctrl$maxsurrogate > 0) {
         inp <- inputs
@@ -228,8 +232,8 @@
 ctree_control <- function(teststat = c("quad", "max"),
     testtype = c("Bonferroni", "Univariate", "Teststatistic"),
     mincriterion = 0.95, minsplit = 20L, minbucket = 7L, minprob = 0.01,
-    stump = FALSE, maxsurrogate = 2L, mtry = Inf, maxdepth = Inf, 
-    multiway = FALSE, splittry = 2L) {
+    stump = FALSE, maxsurrogate = 0L, mtry = Inf, maxdepth = Inf, 
+    multiway = FALSE, splittry = 2L, majority = FALSE) {
 
     teststat <- match.arg(teststat)
     if (teststat == "max") stopifnot(require("mvtnorm"))
@@ -239,7 +243,7 @@ ctree_control <- function(teststat = c("quad", "max"),
          minsplit = minsplit, minbucket = minbucket, 
          minprob = minprob, stump = stump, mtry = mtry, 
          maxdepth = maxdepth, multiway = multiway, splittry = splittry,
-         maxsurrogate = maxsurrogate)
+         maxsurrogate = maxsurrogate, majority = majority)
 }
 
 ctree <- function(formula, data, weights, subset, na.action = na.pass, 

### 2014-03-24 pkg/partykit/R/party.R

@@ -403,8 +403,10 @@ predict_party.constparty <- function(party, id, newdata = NULL,
     function(p, ...) quantile(y, probs = p, ...)
 }
 
-.pred_density <- function(y, w)
-    density(y, weights = w / sum(w))
+.pred_density <- function(y, w) {
+    d <- density(y, weights = w / sum(w))
+    approxfun(d[1:2], rule = 2)
+}
 
 ### workhorse: compute predictions based on fitted / response data
 .predict_party_constparty <- function(node, fitted, response, weights,

### 2014-03-24 pkg/partykit/tests/ctree-ctree.R

@@ -62,6 +62,7 @@ fun <- function(args) {
     old <- treeresponse(oldmod, newdata = test)
     ct <- partykit:::ctree_control
     ctrl <- do.call("ct", args[-(1:2)])
+    ctrl$majority <- TRUE
     nt <- system.time(newmod <- try(partykit:::ctree(y ~ ., data = learn, control = ctrl)))[1]
     if (inherits(newmod, "try-error")) {
         new <- as.list(rep(1, length(old)))

### 2014-03-24 pkg/partykit/tests/regtest-party.R
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@545 255f4249-2263-4417-b1ea-8498779e27a7

@@ -126,3 +126,4 @@ predict(ct, newdata = nd, type = "prob")
 predict(ct, newdata = nd, type = "quantile")
 predict(ct, newdata = nd, type = "quantile", at = NULL)
 predict(ct, newdata = nd, type = "density")
+predict(ct, newdata = nd, type = "density", at = (1:9) / 10)

### 2014-03-24 pkg/partykit/inst/mloss/mloss2.Rnw
added R-Forge reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@538 255f4249-2263-4417-b1ea-8498779e27a7
various additions to MLOSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@547 255f4249-2263-4417-b1ea-8498779e27a7

@@
         Achim Zeileis \\ \addr{Universit\"at Innsbruck} \\ \addr{E-mail: Achim.Zeileis@R-project.org}}
-\title{\pkg{partykit}: An Open-Source Toolkit for Recursive Partytioning in \proglang{R}}
+\title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}
 \editor{(submission for MLOSS track)}
@@
 
-installation information
+The \pkg{partykit} package is an add-on package for the \proglang{R} system
+for statistical computing. It is available from the Comprehensive \proglang{R}
+Archive Network (CRAN) at \url{http://CRAN.R-project.org/package=partykit} and
+can be easily installed in the usual ways from within \proglang{R} using
+\fct{install.packages}. It depends on \proglang{R} (at least 2.15.0) as well
+as the base packages \pkg{graphics}, \pkg{grid}, \pkg{stats}, and the recommended
+\pkg{survival} package. Furthermore, various suggested packages are needed
+for certain special functionalities in the package. To install all of these
+required and suggested packages in one go, the command
+\code{install.packages("partykit", dependencies = TRUE)} can be used.
 
-R-Forge
+In addition to the stable release version of the package on CRAN, the
+current development release can be installed from \proglang{R}-Forge
+\citep{Theussl+Zeileis:2009}. In addition to source and binary packages
+of \pkg{partykit} the entire version history is available through
+\proglang{R}-Forge's \pkg{Subversion} source code management system.
 
-The vignette (\code{"partykit"}) introduces the basic \class{party} class
+Along with the package extensive documentation and examples are shipped.
+The manual pages provide basic technical information on all functions while
+much more detailed descriptions along with worked examples are provided in
+the four package vignettes. First, the vignette \code{"partykit"}
+introduces the basic \class{party} class
 and associated infrastructure while three further vignettes discuss the tools
-built on top of it: \code{"constparty"} covers the eponymous class for
-constant-fit trees along with suitable coercion functions, and
-\code{"ctree"} and \code{"mob"} discuss the new \fct{ctree} and
-\fct{mob} implementations, respectively. Each of the vignettes can be viewed
-within \proglang{R} via \code{vignette(}\emph{``name''}\code{, package = "partykit")}.
+built on top of it: \code{"constparty"} covers the eponymous class (as well as
+the simplified \class{simpleparty} class) for constant-fit trees along with
+suitable coercion functions, and \code{"ctree"} and \code{"mob"} discuss the
+new \fct{ctree} and \fct{mob} implementations, respectively. Each of the
+vignettes can be viewed within \proglang{R} via
+\code{vignette(}\emph{``name''}\code{, package = "partykit")} and the underlying
+source code (in \proglang{R} with {\LaTeX} text) is also available in the source
+package.
 
 \section{User Interface}
+
+two blocks: basic infrastructure (next section) and methods for learning trees,
+either directly or via coercion.
 
@@
 print, plot, predict
+
+nodeapply, nodeprune
+
+\begin{table}
+\centering
+\begin{tabular}{llll}
+\hline
+Algorithm  & Software                                                 & Object class       & Original reference \\ \hline
+CART/RPart & \code{rpart::rpart} + \code{as.party}                   & \code{constparty}  & \cite{Breiman+Friedman+Olshen:1984} \\
+C4.5/J4.8  & \pkg{Weka}/\code{RWeka::J48} + \code{as.party}          & \code{constparty}  & \cite{Quinlan:1993} \\
+QUEST      & \proglang{SPSS}/\pkg{AnswerTree} + \code{pmmlTreeModel} & \code{simpleparty} & \cite{Loh+Shih:1997} \\
+CTree      & \code{ctree}                                             & \code{constparty}  & \cite{Hothorn+Hornik+Zeileis:2006} \\
+MOB        & \code{mob}, \code{lmtree}, \code{glmtree}, \dots         & \code{modelparty}  & \cite{Zeileis+Hothorn+Hornik:2008} \\
+EvTree     & \code{evtree::evtree}                                    & \code{constparty}  & \cite{evtree} \\ \hline
+\end{tabular}
+\caption{\label{tab:overview} Selected tree algorithms than can be interfaced through \pkg{partykit}. The
+  software column lists external software, \proglang{R} functions from other packages (with \code{::} syntax)
+  and from \pkg{partykit}.}
+\end{table}
 
@@
 plot(ttnc_rp, newpage = FALSE, pop = TRUE,
-  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.8))))
+  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.85))))
 popViewport()
@@
 plot(ttnc_mob, newpage = FALSE, pop = TRUE, tp_args = list(ylines = 1.3,
-  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.8))),
+  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.85))),
   linecol = "black"))
@@
 \caption{\label{fig} Tree visualizations of survival on Titanic:
-\class{rpart} tree visualized by \pkg{partykit} (left) and
+\class{rpart} tree converted with \fct{as.party} and visualized by \pkg{partykit} (left); and
 logistic-regression-based tree fitted by \fct{glmtree} (right).}
@@
 \section{Internal Structure}
+
+especially for developers who either want to implement new tree algorithms
+or represent trees learning in other systems in \proglang{R} to facilitate
+predictions, comparisons, etc.
 
@@
 constparty
+simpleparty
 modelparty
@@
 sponsored by the Institute for Mathematical Sciences of the National University of Singapore,
-for feedback and discussions.
+for helpful feedback and stimulating discussions.
+
+\newpage
 

### 2014-03-25 pkg/partykit/inst/mloss/mloss2.Rnw
some more references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@549 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}
-\editor{(submission for MLOSS track)}
+\editor{(submission for MLOSS track, related to JMLR-07-240-1 that was handled by Cheng Soon Ong)}
 
@@
 
+\fct{rpart} from \pkg{rpart} \citep{rpart}
+CART \citep{Breiman+Friedman+Olshen:1984}
+
+C4.5 \citep{Quinlan:1993}
+\fct{J48} from \pkg{RWeka} \citep{RWeka} interfacing \pkg{Weka} \citep{Witten+Frank:2005}
+
+QUEST \citep{Loh+Shih:1997}
+
+CTree \citep{Hothorn+Hornik+Zeileis:2006}
+MOB \citep{Zeileis+Hothorn+Hornik:2008}
+
+ among many others
+\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
+of conceptual algorithms, a wide range of computational implementations
+in various software systems emerged: Typically the original authors of an
+algorithms also provide accompanying software but many software systems,
+e.g., including  or \proglang{R} \citep{R},
+also provide collections of various types of trees.
+Within \proglang{R} the list of prominent packages includes
+
 formula interface
@@
 
-\section{Internal Structure}
+\section{Developer Infrastructure}
 

### 2014-03-25 pkg/partykit/R/modelparty.R
added optional minbucket control argument -> mapped to minsize
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@550 255f4249-2263-4417-b1ea-8498779e27a7

@@ -667,9 +667,12 @@ mob_control <- function(alpha = 0.05, bonferroni = TRUE, minsize = NULL, maxdept
   verbose = FALSE, caseweights = TRUE, ytype = "vector", xtype = "matrix",
   terminal = "object", inner = terminal, model = TRUE,
   numsplit = "left", catsplit = "binary", vcov = "opg", ordinal = "chisq", nrep = 10000,
-  minsplit = minsize)
+  minsplit = minsize, minbucket = minsize)
 {
-  if(missing(minsize) & !missing(minsplit)) minsize <- minsplit
+  ## transform defaults
+  if(missing(minsize) & !missing(minsplit))  minsize <- minsplit
+  if(missing(minsize) & !missing(minbucket)) minsize <- minbucket
+  
   if(is.finite(mtry)) {
     mtry <- if(mtry < 1L) Inf else as.integer(mtry)
   }

### 2014-03-25 pkg/partykit/inst/mloss/mloss2.Rnw
first draft of user interface section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@551 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-03-24 pkg/partykit/inst/mloss/mloss2.Rnw
@@
 
-two blocks: basic infrastructure (next section) and methods for learning trees,
-either directly or via coercion.
-
-formula interface
-
-as.party
-
-PMML
-
-print, plot, predict
-
-nodeapply, nodeprune
+The \pkg{partykit} package provides functionality as different levels. First,
+there is basic infrastructure for representing, modifying, and displaying trees
+and recursive partitions and these tools are mostly intended for developers
+and described in the next section. Second, there are tools for inferring trees
+from data or for porting trees inferred by other software into \pkg{partykit}.
 
@@
 \end{table}
+
+While originally an important goal for the development of \pkg{partykit} was to
+provide infrastructure for the authors' own tree induction algorithms
+CTree \citep{Hothorn+Hornik+Zeileis:2006} and MOB \citep{Zeileis+Hothorn+Hornik:2008},
+the design was very careful to separate as much functionality as possible into
+more general classes that are useful for a far broader class of tree algorithms.
+Specifically, to be able to print/plot/predict different trees in a unified way,
+there are so-called coercion functions for transforming trees learned in other
+software packages (inside and outside of \proglang{R}) into the classes provided
+by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
+package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
+and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
+algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
+to the same object class \class{constparty}. This is a general class that can
+in principle represent all the major classical tree algorithms with constant
+fits in the terminal nodes. Also, the same class is employed for conditional
+inference trees (CTree) that can be learned with the \fct{ctree} function
+directly within \pkg{partykit} or evolutionary learned trees from \pkg{evtree}
+\citep{evtree}.
+
+But not only trees learned within \proglang{R} can be transformed to the
+proposed infrastructure but also trees from other software packages. Either
+a dedicated interface has to be created using the building blocks described
+in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or
+PMML (Predictive Model Markup Language) can be used as an intermediate
+exchange format. This is an XML standard created by an international
+consortium \citep{DMG:2014} that includes a \code{<TreeModel>} tag with
+support for constant-fit classification and regression trees. The function
+\fct{pmmlTreeModel} allows to read these files and represents them as
+\class{simpleparty} objects in \pkg{partykit}. The reason for not using the
+\class{constparty} class as above is that the PMML format only stores
+point predictions (e.g., a mean or proportion) and not all observations
+from the learning sample. So far, the PMML interface has been tested with
+output from the \proglang{R} package \pkg{pmml} and \proglang{SPSS}'s
+\pkg{AnswerTree} model. The latter includes an implementation of the
+QUEST algorithm \citep{Loh+Shih:1997}.
+
+Finally, the \pkg{partykit} function \fct{mob} implements model-based
+recursive partitioning (MOB) along with ``mobster'' interfaces for certain
+models (e.g., \fct{lmtree}, \fct{glmtree}). These return objects of class
+\class{modelparty} where nodes are associated with statistical models (as
+opposed to simple constant fits). In principle, this may also be adapted
+to other model trees (such as GUIDE, LMT, or M5') but no such interface
+is currently available.
+
+All of these different trees (see Table~\ref{tab:overview} for an overview)
+use the same infrastructure at the core but possibly with different options
+enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
+can be used to create textual and graphical displays of the tree and for
+computing predictions on new data, respectively. As an example for the
+visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
+to the well-known data on survival of passengers on the ill-fated maiden
+voyage of the MS~Titanic: The left panel shows a CART tree with constant fits
+learned by \pkg{rpart} and converted to \pkg{partykit}. The left panel
+shows a MOB tree learned within \pkg{partykit} with a logistic regression
+model for treatment effects in the terminal nodes. Additionally, the
+are further utility functions, e.g.,
+\fct{nodeapply} can be employed to access further information
+stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.
 
@@
 @
-\caption{\label{fig} Tree visualizations of survival on Titanic:
+\caption{\label{fig:ttnc} Tree visualizations of survival on Titanic:
 \class{rpart} tree converted with \fct{as.party} and visualized by \pkg{partykit} (left); and

### 2014-03-25 pkg/partykit/inst/mloss/mloss2.Rnw
touch-ups in sections 1-3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@553 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \newcommand{\dquote}[1]{``{#1}''}
-\newcommand{\fct}[1]{\texttt{#1()}}
+\newcommand{\fct}[1]{\texttt{#1}}
 \newcommand{\class}[1]{\squote{\texttt{#1}}}
@@
 \begin{abstract}
-  The \pkg{partykit} package provides a flexible toolkit with infrastructure for
+  The \proglang{R} package \pkg{partykit} provides a flexible toolkit for
   learning, representing, summarizing, and visualizing a wide range of tree-structured
@@
   (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
-  along with suitable coercion functions to create such tree models (e.g., by
+  along with suitable coercion functions to create such trees (e.g., by
   \pkg{rpart}, \pkg{RWeka}, PMML). (c)~A reimplementation of \emph{conditional inference
@@
 in various software systems emerged: Typically the original authors of an
-algorithms also provide accompanying software but many software systems,
-e.g., including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
+algorithm also provide accompanying software but many software systems,
+including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
 also provide collections of various types of trees.
 Within \proglang{R} the list of prominent packages includes
-\pkg{rpart} \citep[implementing the CART algorithm]{rpart},
+\pkg{rpart} \citep[implementing CART]{rpart},
 \pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
@@
 
-The \pkg{partykit} tries to address the latter point and improve the
-computational situation by providing a common unified infrastructure
+The \pkg{partykit} aims at mitigating the latter issue
+by providing a common unified infrastructure
 for recursive partytioning in the \proglang{R} system for statistical
@@
   \item Nodes are recursive objects, i.e., a node can contain child nodes.
-  \item Keep (learning) data out of the recursive node and split structure.
+  \item Keep the (learning) data out of the recursive node and split structure.
   \item Basic printing, plotting, and predicting for raw node structure.
   \item Customization via suitable panel or panel-generating functions.
-  \item Coercion from existing objects (\code{rpart}, \code{J48}, etc.) to the new class.
+  \item Coercion from existing object classes in \proglang{R} (\code{rpart}, \code{J48}, etc.) to the new class.
   \item Usage of simple/fast \proglang{S}3 classes and methods.
@@
 Archive Network (CRAN) at \url{http://CRAN.R-project.org/package=partykit} and
-can be easily installed in the usual ways from within \proglang{R} using
+can be easily installed in the usual ways from within \proglang{R}, e.g., using
 \fct{install.packages}. It depends on \proglang{R} (at least 2.15.0) as well
 as the base packages \pkg{graphics}, \pkg{grid}, \pkg{stats}, and the recommended
-\pkg{survival} package. Furthermore, various suggested packages are needed
+\pkg{survival}. Furthermore, various suggested packages are needed
 for certain special functionalities in the package. To install all of these
@@
 
-In addition to the stable release version of the package on CRAN, the
+In addition to the stable release version on CRAN, the
 current development release can be installed from \proglang{R}-Forge
 \citep{Theussl+Zeileis:2009}. In addition to source and binary packages
-of \pkg{partykit} the entire version history is available through
+the entire version history is available through
 \proglang{R}-Forge's \pkg{Subversion} source code management system.
 
-Along with the package extensive documentation and examples are shipped.
+Along with the package extensive documentation with examples is shipped.
 The manual pages provide basic technical information on all functions while
-much more detailed descriptions along with worked examples are provided in
+much more detailed descriptions along with hands-on examples are provided in
 the four package vignettes. First, the vignette \code{"partykit"}
@@
 
-The \pkg{partykit} package provides functionality as different levels. First,
+The \pkg{partykit} package provides functionality at different levels. First,
 there is basic infrastructure for representing, modifying, and displaying trees
-and recursive partitions and these tools are mostly intended for developers
+and recursive partitions -- these tools are mostly intended for developers
 and described in the next section. Second, there are tools for inferring trees
-from data or for porting trees inferred by other software into \pkg{partykit}.
+from data or for importing trees inferred by other software into \pkg{partykit}.
 
@@
 provide infrastructure for the authors' own tree induction algorithms
-CTree \citep{Hothorn+Hornik+Zeileis:2006} and MOB \citep{Zeileis+Hothorn+Hornik:2008},
+CTree and MOB,
 the design was very careful to separate as much functionality as possible into
 more general classes that are useful for a far broader class of tree algorithms.
-Specifically, to be able to print/plot/predict different trees in a unified way,
+In particular, to be able to print/plot/predict different trees in a unified way,
 there are so-called coercion functions for transforming trees learned in other
@@
 inference trees (CTree) that can be learned with the \fct{ctree} function
-directly within \pkg{partykit} or evolutionary learned trees from \pkg{evtree}
+directly within \pkg{partykit} or evolutionary trees from package \pkg{evtree}
 \citep{evtree}.
 
-But not only trees learned within \proglang{R} can be transformed to the
+Not only trees learned within \proglang{R} can be transformed to the
 proposed infrastructure but also trees from other software packages. Either
@@
 \class{constparty} class as above is that the PMML format only stores
-point predictions (e.g., a mean or proportion) and not all observations
+point predictions (e.g., a mean or proportion) rather than all observations
 from the learning sample. So far, the PMML interface has been tested with
@@
 to the well-known data on survival of passengers on the ill-fated maiden
-voyage of the MS~Titanic: The left panel shows a CART tree with constant fits
-learned by \pkg{rpart} and converted to \pkg{partykit}. The left panel
+voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
+learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
 shows a MOB tree learned within \pkg{partykit} with a logistic regression

### 2014-03-26 pkg/partykit/inst/mloss/mloss2.Rnw
first draft complete
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@555 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \begin{keywords}
-open-source software, recursive partitioning, regression trees, classification trees, decision trees, statistical learning. 
+recursive partitioning, regression trees, classification trees, statistical learning, \proglang{R}.
 \end{keywords}
@@
 termed ``recursive partitioning'' or ``decision trees'' or
-``tree(-structured) models'' etc. Particularly influential were the
-algorithms CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
+``tree(-structured) models'' etc. The particularly influential 
+algorithms include CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
 C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
@@
 \pkg{rpart} \citep[implementing CART]{rpart},
-\pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
+\pkg{RWeka} \citep[with interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
 \pkg{party} \citep[implementing CTree and MOB]{party} among
@@
 All of these algorithms and software implementations have to deal with
-very similar challenges. However, due to the fragmentation of the
-communities in which the corresponding research is published -- ranging
+similar challenges. However, due to the fragmentation of the
+communities in which they are published -- ranging
 from statistics over machine learning to various applied fields -- many
@@
 reuse of code.
-
+%
 The \pkg{partykit} aims at mitigating the latter issue
@@
 computing. In particular, \pkg{partykit}
-provides tools for representing fitted trees along with printing,
-plotting, and computing predictions. The design principles are:
+provides tools for representing, printing, plotting trees and
+computing predictions. The design principles are:
+
 \begin{itemize}
-  \item One `agnostic' base class (\class{party}) which can encompass an extremely wide
-        range of different types of trees.
+  \item One `agnostic' base class (\class{party}) encompassing a very wide
+        range of different tree types.
   \item Subclasses for important types of trees, e.g., trees with constant
@@
 the design was very careful to separate as much functionality as possible into
-more general classes that are useful for a far broader class of tree algorithms.
+more general classes that are useful for a far broader class of trees.
 In particular, to be able to print/plot/predict different trees in a unified way,
 there are so-called coercion functions for transforming trees learned in other
-software packages (inside and outside of \proglang{R}) into the classes provided
-by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
-package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
+software packages (inside and outside of \proglang{R}) to the classes provided
+by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart}
+(\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
 and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
@@
 learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
-shows a MOB tree learned within \pkg{partykit} with a logistic regression
-model for treatment effects in the terminal nodes. Additionally, the
+shows a MOB tree learned with \pkg{partykit} with a logistic regression
+for treatment effects in the terminal nodes. Additionally, the
 are further utility functions, e.g.,
@@
 \caption{\label{fig:ttnc} Tree visualizations of survival on Titanic:
-\class{rpart} tree converted with \fct{as.party} and visualized by \pkg{partykit} (left); and
-logistic-regression-based tree fitted by \fct{glmtree} (right).}
+\class{rpart} tree converted with \fct{as.party} and visualized by
+\pkg{partykit} (left); and logistic-regression-based tree fitted by
+\fct{glmtree} (right).}
 \end{figure}
@@
 
-especially for developers who either want to implement new tree algorithms
-or represent trees learning in other systems in \proglang{R} to facilitate
-predictions, comparisons, etc.
+The unified infrastructure at the core of \pkg{partykit} is especially
+appealing for developers who either want to implement new tree algorithms
+or represent trees learned in other systems.
 
-partysplit
-partynode
-party
+Here, we briefly outline
+the most important classes. More details can be found in the vignettes.
+\begin{description}
+  \item[\class{partysplit}:] Split specification with an integer ID for the
+    splitting variable, breakpoint(s), and indexes for the kids.
+  \item[\class{partynode}:] Node specification with an integer ID, a
+    \class{partysplit}, and a list of kids (if any) that are \class{partynode}
+    objects again.
+  \item[\class{party}:] Tree specification with a \class{partynode} tree
+    along with a \class{data.frame} (which optionally may be empty),
+    potentially plus further information about fitted values and so-called
+    \class{terms} that allow to preprocess new data for predictions.
+\end{description}
+All classes have an additional slot for storing arbitrary information at
+any level of the tree. This is exploited by \class{constparty}, \class{simpleparty},
+and \class{modelparty} which store the observed response, point predictions,
+and fitted parametrics models, respectively.
+% Extractor functions for the various components of splits, nodes, and trees are provided.
 
-constparty
-simpleparty
-modelparty
+\section{Discussion and Outlook}
 
-extractor functions
-
-\section{Conclusions}
-
-run-time, memory requirements, benchmarks?
+Package \pkg{partykit} provides a toolkit for trees in \proglang{R} that gives
+emphasis to flexibility and extensibility. The infrastructure is easily
+accessible and written in plain \proglang{R}. Nevertheless speed is not a critical
+concern because (a)~lean data structures are employed and (b)~the time-intensive
+part in tree computations is typically inferring the tree rather than
+representing/printing/plotting it. Hence, the two functions for inferring trees
+either have computationally intensive parts in \proglang{C} (\fct{ctree})
+or build on \proglang{R}'s fitting functions (\fct{mob}). For benchmarks of
+their predictive performance, see their original references.
+Regarding memory requirements, the \class{party} trees are very lean but can
+optionally be used to store any information. The recursive node structure
+itself is extremely lightweight and optionally zero rows from the learning data
+are sufficient, which future versions will exploit for storing forests.
 
@@
 
-We are thankful to the organizers and participants of the ``Workshop on Classification and Regression Trees'' 2014,
-sponsored by the Institute for Mathematical Sciences of the National University of Singapore,
-for helpful feedback and stimulating discussions.
-
-\newpage
+We are thankful to the organizers and participants of the
+``Workshop on Classification and Regression Trees'' (March 2014),
+sponsored by the Institute for Mathematical Sciences of the
+National University of Singapore, for helpful feedback and
+stimulating discussions.
 

### 2014-03-26 pkg/partykit/R/ctree.R
tie breaking for very rare cases
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@556 255f4249-2263-4417-b1ea-8498779e27a7

@@ -82,10 +82,13 @@
     storage.mode(response) <- "double"
 
     lin <- .Call("R_LinstatExpCov", data, inp, response, weights)
+    ### <FIXME> this is slow, chisq or even teststatistics might be better
     p <- sapply(lin[inp], function(x) do.call(".pX2", x[-1]))
+    ### </FIXME>
     colnames(p) <- colnames(data)[inp]
     rownames(p) <- c("teststat", "pval")
-    crit <- p["pval",]
+    ### <FIXME> break ties? </FIXME>
+    crit <- p["pval",,drop = TRUE]
 
     ret <- vector(mode = "list", length = min(sum(inp), ctrl$maxsurrogate))
 
@@ -139,7 +142,13 @@
     } 
     lin <- .Call("R_LinstatExpCov", data, inp, response, weights)
     p <- sapply(lin[inp], function(x) do.call(ctrl$cfun, x[-1]))
-    crit <- p[1,,drop = FALSE]
+    crit <- p[1,,drop = TRUE]
+    ### crit is maximised, but there might be ties
+    ties <- which(abs(crit - max(crit)) < .Machine$double.eps)
+    if (length(ties) > 1) {
+        ### add a small value (< 1/1000) to crit derived from order of teststats
+        crit[ties] <- crit[ties] + order(p["teststat", ties]) / (sum(ties) * 1000)
+    }
     p <- p[-1,,drop = FALSE]
     colnames(p) <- colnames(data)[inp]
 

### 2014-03-26 pkg/partykit/inst/mloss/mloss2.Rnw
cosmetics, mostly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@557 255f4249-2263-4417-b1ea-8498779e27a7

@@
   infrastructure for \emph{representing} trees (inferred by any algorithm) so that
-  unified \code{print}/\code{plot}/\code{predict} methods are available.
+  unified \code{print}/\code{plot}/\code{predict} methods are available;
   (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
   along with suitable coercion functions to create such trees (e.g., by
-  \pkg{rpart}, \pkg{RWeka}, PMML). (c)~A reimplementation of \emph{conditional inference
-  trees} (\code{ctree}, originally provided in the \pkg{party} package). (d)~An extended
+  \pkg{rpart}, \pkg{RWeka}, PMML); (c)~A reimplementation of \emph{conditional inference
+  trees} (\code{ctree}, originally provided in the \pkg{party} package); (d)~An extended
   reimplementation of \emph{model-based recursive partitioning} (\code{mob}, also originally
@@
 Archive Network (CRAN) at \url{http://CRAN.R-project.org/package=partykit} and
-can be easily installed in the usual ways from within \proglang{R}, e.g., using
+can be installed from within \proglang{R}, e.g., using
 \fct{install.packages}. It depends on \proglang{R} (at least 2.15.0) as well
@@
 a dedicated interface has to be created using the building blocks described
-in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or
-PMML (Predictive Model Markup Language) can be used as an intermediate
-exchange format. This is an XML standard created by an international
-consortium \citep{DMG:2014} that includes a \code{<TreeModel>} tag with
-support for constant-fit classification and regression trees. The function
+in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or PMML
+(Predictive Model Markup Language) can be used as an intermediate exchange
+format.  This is an XML standard created by an international consortium
+\citep{DMG:2014} that includes a \code{<TreeModel>} tag with support for
+constant-fit classification and regression trees.  The function
 \fct{pmmlTreeModel} allows to read these files and represents them as
 \class{simpleparty} objects in \pkg{partykit}.  The reason for not using the
-\class{constparty} class as above is that the PMML format only stores
-point predictions (e.g., a mean or proportion) rather than all observations
-from the learning sample. So far, the PMML interface has been tested with
-output from the \proglang{R} package \pkg{pmml} and \proglang{SPSS}'s
-\pkg{AnswerTree} model. The latter includes an implementation of the
-QUEST algorithm \citep{Loh+Shih:1997}.
+\class{constparty} class as above is that the PMML format only stores point
+predictions (e.g., a mean or proportion) rather than all observations of the
+response that was used for tree learning.  So far, the PMML interface has
+been tested with output from the \proglang{R} package \pkg{pmml} and
+\proglang{SPSS}'s \pkg{AnswerTree} model.  The latter includes an
+implementation of the QUEST algorithm \citep{Loh+Shih:1997}.
 
@@
 
+\vspace*{-.1cm}
+
 \section{Developer Infrastructure}
@@
 appealing for developers who either want to implement new tree algorithms
-or represent trees learned in other systems.
-
-Here, we briefly outline
-the most important classes. More details can be found in the vignettes.
+or represent trees learned in other systems. Here, we briefly outline
+the most important classes and refer to the vignettes for more details:
 \begin{description}

### 2014-03-26 pkg/partykit/inst/mloss/mloss2.Rnw
ready for submission
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@559 255f4249-2263-4417-b1ea-8498779e27a7

@@
   learning, representing, summarizing, and visualizing a wide range of tree-structured
-  regression and classification models. The functionality encompasses: (a)~Basic
+  regression and classification models. The functionality encompasses: (a)~basic
   infrastructure for \emph{representing} trees (inferred by any algorithm) so that
   unified \code{print}/\code{plot}/\code{predict} methods are available;
-  (b)~Dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
+  (b)~dedicated methods for trees with \emph{constant fits} in the leaves (or terminal nodes)
   along with suitable coercion functions to create such trees (e.g., by
-  \pkg{rpart}, \pkg{RWeka}, PMML); (c)~A reimplementation of \emph{conditional inference
-  trees} (\code{ctree}, originally provided in the \pkg{party} package); (d)~An extended
+  \pkg{rpart}, \pkg{RWeka}, PMML); (c)~a reimplementation of \emph{conditional inference
+  trees} (\code{ctree}, originally provided in the \pkg{party} package); (d)~an extended
   reimplementation of \emph{model-based recursive partitioning} (\code{mob}, also originally
@@
 \class{constparty} class as above is that the PMML format only stores point
-predictions (e.g., a mean or proportion) rather than all observations of the
-response that was used for tree learning.  So far, the PMML interface has
+predictions (e.g., a mean or proportion) rather than all observations
+from the learning sample.  So far, the PMML interface has
 been tested with output from the \proglang{R} package \pkg{pmml} and
@@
 
-\vspace*{-.1cm}
-
 \section{Developer Infrastructure}
@@
 appealing for developers who either want to implement new tree algorithms
-or represent trees learned in other systems. Here, we briefly outline
+or represent trees learned in other systems.
+
+Here, we briefly outline
 the most important classes and refer to the vignettes for more details:

### 2014-03-26 pkg/partykit/R/ctree.R
proposed to add sctest() method for "constparty" objects
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@562 255f4249-2263-4417-b1ea-8498779e27a7

@@ -409,3 +409,11 @@ ctree <- function(formula, data, weights, subset, na.action = na.pass,
     storage.mode(infl) <- "double"
     return(infl)
 }
+
+sctest.constparty <- function(object, node = NULL, ...)
+{
+  ids <- if(is.null(node)) nodeids(object, terminal = FALSE) else node
+  rval <- nodeapply(object, ids, function(n) info_node(n)$criterion)
+  names(rval) <- ids
+  if(length(ids) == 1L) rval[[1L]] else rval  
+}

### 2014-03-27 pkg/partykit/R/ctree.R
return NULL for terminal nodes in sctest
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@573 255f4249-2263-4417-b1ea-8498779e27a7

@@ -412,16 +412,22 @@ ctree <- function(formula, data, weights, subset, na.action = na.pass,
 
 sctest.constparty <- function(object, node = NULL, ...)
 {
-  ids <- if(is.null(node)) {
-      ### inner nodes only
-      n <- nodeids(object, terminal = FALSE) ### all nodes
-      n[!(n %in% nodeids(object, terminal = TRUE))] 
-  } else node
-  rval <- nodeapply(object, ids, function(n) {
-      ret <- exp(info_node(n)$criterion)
-      ret["p.value",] <- 1 - ret["p.value",]
-      ret
-  })
-  names(rval) <- ids
-  if(length(ids) == 1L) rval[[1L]] else rval  
+
+    if(is.null(node)) {
+        ids <- nodeids(object, terminal = FALSE) ### all nodes
+    } else {
+        ids <- node
+    }
+
+    rval <- nodeapply(object, ids, function(n) {
+        crit <- info_node(n)$criterion
+        if (is.null(crit)) return(NULL)
+        ret <- exp(crit)
+        ret["p.value",] <- 1 - ret["p.value",]
+        ret
+    })
+    names(rval) <- ids
+    if(length(ids) == 1L)
+        return(rval[[1L]])
+    return(rval)
 }

### 2014-03-27 pkg/partykit/inst/mloss/letter_mloss.tex
fix typos, in LaTeX
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@574 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[ngerman,uzh]{scrlttr2}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

%% ANPASSUNGSMÖGLICHKEITEN
\KOMAoptions{%
%backaddress=false%  Keine Rücksendeadresse im Adressfeld
%,foldmarks=false%  Faltmarken ausschalten
}

\usepackage{babel}

\begin{document}

%% VARIABLEN
% Falls Unterschrift vom definierten Namen abweicht:
\setkomavar{signature}{Torsten Hothorn}
\setkomavar{subject}{Betreff}
%\setkomavar{specialmail}{Interne Post}

\begin{letter}{
Drs.~Murphy and Schölkopf \\
Editors-in-chief \\
Journal of Machine Learning Research}

\subject{JMLR-MLOSS submission}

\opening{Dear Drs.~Murphy and Schölkopf,}

my co-author Achim Zeileis and I would like to submit the manuscript
``\textbf{partykit}: An Open-Source Toolkit for Recursive Partytioning in R'' to the
Machine Learning Open-Source Software track of the Journal of Machine
Learning Research.

The \textsf{R} add-on package \textbf{partykit} described in this submission implements a
flexible toolkit for learning, representing, summarizing, and visualizing a
wide range of tree-structured models.  The package offers a basic class
infrastructure for representing fitted trees, a unified set of methods for
printing and plotting of trees as well as for computing predictions for new
observations, tailored functionality for trees with constant fits and with
parametric models in each terminal node, and reimplementations of conditional
inference trees (CTree) and model-based recursive partitioning (MOB).

This manuscript is related to JMLR-07-240-1 ``Let's Have a party! An
Open-Source Toolbox for Recursive Partytioning'' describing the
\textbf{party} add-on
package.  This manuscript was rejected with encouragement for resubmission by
JMLR-MLOSS editor Cheng Soon Ong in November 2007.  The main concerns of the
reviewers were that it was not easy to extend the old \textbf{party} package by adding
new tree learners or representing existing trees in our package.  This was
a very important suggestion and part of the motivation for starting from
scratch and develop a new R package (\textbf{partykit}) that easily allows to
accommodate our own tree algorithms but makes the underlying infrastructure
easily accessible to others. Actually, the \textbf{partykit} package in its current
form goes well beyond what was suggested in the reviews of JMLR-07-240-1
and both manuscript and software were entirely rewritten. Hence, we felt it
to be more appropriate to submit this as a new manuscript rather than a
revision of JMLR-07-240-1.  However, if you would like to see a formal
point-to-point reply to the old submission, please let us know.

Sincerely yours, \\
\includegraphics[width = .3\textwidth]{TH_Unterschrift}

\end{letter}

\end{document}

### 2014-06-04 pkg/partykit/inst/mloss/review2/p2p.tex
start rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@585 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass{article}
\usepackage{natbib}

\title{Rebuttal JMLR 14-133}
\author{Torsten Hothorn and Achim Zeileis}

\begin{document}

\maketitle

\section*{AE}

\textit{
I regret to inform you that your submission ``partykit: A modular toolkit
for recursive partytioning in R'' has been rejected from publication at JMLR
Track for Machine Learning Open Source Software.  Two of the reviewers of
the previous version were enlisted for this version, and one new reviewer
was added.  While the reviewers voted for ``accept'', the actual reviews
were less positive and raised several serious concerns about the project. 
If the authors choose to resubmit the project, please include a description
addressing the reviewers' concerns.}

\section*{Referee 1}

\textit{
This is a clean and concise description of an interesting set of tools for
data analysis with recursive partitioning.  It is mainly a description or
announcement, and has minimal polemic or pedagogical value.  It will be a
useful reference for those a basic outline of the software's capabilities,
and provides a useful bibliography.}

Tja.

\section*{Referee 2}

\textit{
This paper describes work that, in my understanding, does two things:
\begin{enumerate}
\item It provides infrastructure that can in principle be used with any
      recursive partitioning implementation, and that at this point 
      interfaces to several different widely used implementations, 
      additional to those directly available in R's partykit package.
\item It provides implementations of two specialized tree-based methods.
\end{enumerate}
}

\textit{
The software has a useful role in providing commonality across diverse
recursive partitioning methods and implementations: for structuring printed
output, for generating predictions, and for plotting results.  Very
extensive documentation, some distributed with the package, is available
from references given in the paper.}

\textit{Quibbles and queries are:}
\begin{enumerate}
\item 
\textit{
The legend for Table 1 begins: ``Selected tree algorithms than can be
interfaced through partykit...''  Surely it is the output from selected
implementations that can be interfaced, not the algorithm used.  The word
‘algorithm’ seems to me anyway inappropriate in this context.  The same (or
very similar) algorithm may have different implementations.  Different
methods with their inevitably different algorithms (including, in partykit:
classical `constant’ trees, model-based trees, and conditional inference
trees) further wide widen the range of implementations.}

rephrase: interfaces data structures used by different implementations of different
tree growing methods / algorithms.

\item 
\textit{
The distinction between implementation (what the authors call `algorithm’)
and output is crucial also because as far as I am able to judge, partykit’s
abilities are limited to taking over and processing some part only of the
output from, e.g., R’s rpart package.  As I understand, there is no ability
to use or process, e.g., results from rpart’s on the fly cross-validation. 
Would it be straightforward to incorporate structure that is able to
accommodate such information?  See also point 3 that now follows.}

Hm, maybe store such information in \texttt{info} slots?

\item 
\textit{
I note the comment on the help page for the partykit function ctree() that
the inference process is based on permutation tests, with one or other type
of adjustment for multiple testing.  Then follows the strong claim that:
This statistical approach ensures that the right sized tree is grown and no
form of pruning or cross-validation or whatsoever is needed.  I think it
pertinent to ask what has been done to judge the effectiveness of this
approach, assessed using separate test data, against the approach used in,
e.g., rpart or CART?  This is an issue for the help page, rather than
perhaps for the paper.}

See \citep{Hothorn+Hornik+Zeileis:2006}.

\item 
\textit{
How well-adapted are the tools in partykit for implementing random forest
or other ensemble type approaches? I note that partykit does not (yet?) have
an equivalent of the cforest() function that is in the older party package.
}

Work in progress.

\end{enumerate}

\textit{
At the very least, I would like to see brief comments on where splitting and
accuracy assessment strategies, short of use of separate test data, fit in
or around the toolkit.}

infrastructure independent of such things.

\textit{To what extent are toolkit functions independent of
the methodology used for the partykit functions ctree() and mob()?}

completely.

\textit{If
necessary, some of the historical detail in the overview might be replaced
by references to discussion that appears elsewhere, in order to make way for
comment on these points.}

pfff

\section*{Referee 3}

\textit{
The paper describes an R library,}

an R package, actually.

\textit{ `partykit', that appears to be a unified
way of representing decision tree structures (or recursive partitioning
structures) in R and manipulating or visualizing them.  The authors note
that their library contains support for representing the output of trees
created with other packages, and the paper appears to function as a very
basic tutorial with links to very detailed R vignettes.  The package also
implements a few of the authors' own recursive partitioning algorithms.}

\textit{
The project seems to have fairly minor activity, and appears to have had a
Google Summer of Code project in previous years.  The subversion repository
seems to indicate that the code is still being (somewhat) actively developed
despite a relative lack of commits between 2009 and 2013.}

und?

\textit{There appears to
be no bug tracker or open mailing list.}

added

\textit{There is knowledge of the package
in the R ecosystem, but partykit does not seem to be as popular as other R
packages (admittedly, partykit is a niche product).}

Hard to judge, but reverse depends: NHEMOtree, quint. Number of citations of party paper.

\textit{
Support for trees produced by other libraries is provided, but I am not
certain how appealing the infrastructure will be to developers who want to
represent trees from other systems.  It seems like for non-PMML formatted
trees, implementing a conversion layer to partykit objects wouldn't be any
easier or harder than other packages.}

depends, eh?

\textit{
Clearly, for developing new tree algorithms, though, partykit would be a good choice.}

We actually have proof!

\textit{
Section 5 is very qualitative and vague.  The authors should definitely
improve this in future revisions.  It is written, "lean data structures are
employed", "the `party' structures are very lean", and also "The recursive
node structure itself is extremely lightweight" but without any sort of
quantitative details, it's difficult to believe those statements without
diving into the code.}

schwierig.

\textit{
In my opinion the authors could do one of these things to validate their
claims: provide proof that the data structures are minimal (with respect to
other implementations), show that very large tree structures can be easily
represented and handled by partykit, or simply leave out or significantly
shorten the discussion of speed and memory usage entirely.}

schwierig.

\textit{
Overall, partykit is a well-designed and extensively documented package, but
I have some reservations about the project's activity level and popularity
going into the future.  The developers should consider setting up a mailing
list or bug tracker to make the project more open to new contributors.}

done.

\bibliographystyle{plainnat}
\bibliography{../../../vignettes/party.bib}

\end{document}

### 2014-09-05 pkg/partykit/inst/mloss/review2/p2p.tex
added some comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@594 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+mention: previous round of reviews asked for this format of the paper
+
 \section*{Referee 2}
@@
 }
+
+yes
+
+maybe worth mentioning: due to the flexible infrastructure, the two
+specialized methods (ctree, mob) have become more flexible compared to
+earlier implementations (especially the mob framework)
 
@@
 implementations that can be interfaced, not the algorithm used.  The word
-‘algorithm’ seems to me anyway inappropriate in this context.  The same (or
+`algorithm' seems to me anyway inappropriate in this context.  The same (or
 very similar) algorithm may have different implementations.  Different
 methods with their inevitably different algorithms (including, in partykit:
-classical `constant’ trees, model-based trees, and conditional inference
+classical `constant' trees, model-based trees, and conditional inference
 trees) further wide widen the range of implementations.}
@@
 \textit{
-The distinction between implementation (what the authors call `algorithm’)
-and output is crucial also because as far as I am able to judge, partykit’s
+The distinction between implementation (what the authors call `algorithm')
+and output is crucial also because as far as I am able to judge, partykit's
 abilities are limited to taking over and processing some part only of the
-output from, e.g., R’s rpart package.  As I understand, there is no ability
-to use or process, e.g., results from rpart’s on the fly cross-validation. 
+output from, e.g., R's rpart package.  As I understand, there is no ability
+to use or process, e.g., results from rpart's on the fly cross-validation. 
 Would it be straightforward to incorporate structure that is able to
@@
 See \citep{Hothorn+Hornik+Zeileis:2006}.
+
+But probably also rephrase...
 
@@
 \textit{
-The paper describes an R library,}
-
-an R package, actually.
-
-\textit{ `partykit', that appears to be a unified
+The paper describes an R library, `partykit', that appears to be a unified
 way of representing decision tree structures (or recursive partitioning
@@
 
+mention: previous round of reviews asked for this format
+
 \textit{
@@
 
-und?
+not a big software project, just two main developers, hence sometimes longer breaks
+
+we feel that this is not an indicator of the quality of the code, though
 
@@
 
-added
+E-mails to the maintainer address were always possible and used fairly regularly
+by both end-users and other package developers. A mailing list was added now.
 
@@
 depends, eh?
+
+examples: CTree and display of p-values, J4.8 and multiway splits
 

### 2014-09-05 pkg/partykit/inst/mloss/review2/p2p.tex
elaborate based on discussion with AZ
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@597 255f4249-2263-4417-b1ea-8498779e27a7

@@
 data analysis with recursive partitioning.  It is mainly a description or
-announcement, and has minimal polemic or pedagogical value.  It will be a
+announcement, and has minimal polemic or pedagogical value.}
+
+When preparing this revision of our earlier version JMLR-07-240, we closely
+followed the suggestions given in the corresponding reviews. Our
+understanding of this remark of Referee 3 on JMLR-07-240
+\begin{quote} 
+Surely it is possible to give broadbrush information. There should be direct
+reference to web-based
+materials that give more detailed information and examples.
+\end{quote}
+was that the four page description should really focus on 
+broadbrush informtion, with all the interesting details being explained and
+discussed in the packages's documentation.
+
+\textit{It will be a
 useful reference for those a basic outline of the software's capabilities,
@@
 
-Tja.
-
-mention: previous round of reviews asked for this format of the paper
+Thank you.
 
@@
 
-yes
+Thank you for this nice and concise summary of the four page description.
 
-maybe worth mentioning: due to the flexible infrastructure, the two
-specialized methods (ctree, mob) have become more flexible compared to
-earlier implementations (especially the mob framework)
+When we started working on partykit, we redesigned all classes from the
+earlier party package from scretch and also reimplemented ctree and mob
+without much overlap between the party and partykit code bases. As a
+consequence of the much cleaner partykit design, the code base for both
+ctree and mob could be reduced considerably while at the same time making
+both implementations more powerful due to more flexible options.
 
@@
 
+Thank you.
+
 \textit{Quibbles and queries are:}
@@
 interfaced through partykit...''  Surely it is the output from selected
-implementations that can be interfaced, not the algorithm used.  The word
+implementations that can be interfaced, not the algorithm used.}
+
+Yes, thank you. We completely agree with you on the distinction between
+`algorithm' and it's `implementation' and changed the caption accordingly.
+
+\textit{The word
 `algorithm' seems to me anyway inappropriate in this context.  The same (or
@@
 classical `constant' trees, model-based trees, and conditional inference
-trees) further wide widen the range of implementations.}
+trees) further widen the range of implementations.}
 
-rephrase: interfaces data structures used by different implementations of different
-tree growing methods / algorithms.
+We completely agree and consistently use "algorithm" for the concept
+and "implementation" for the software throughout the paper now.
 
@@
 output from, e.g., R's rpart package.  As I understand, there is no ability
-to use or process, e.g., results from rpart's on the fly cross-validation. 
-Would it be straightforward to incorporate structure that is able to
+to use or process, e.g., results from rpart's on the fly cross-validation.}
+
+Yes, this is right. It is not possible to directly work on the rpart
+results.  The reason is that we coerce rpart objects to constparty objects
+and this class does not capture cost-complexity pruning since not all
+algorithms have a notion of a risk function and thus perform cost-complexity
+pruning.  The tools provided by partykit require that pruning is performed
+by rpart or J4.8 or by user-implemented add-ons.
+
+\textit{Would it be straightforward to incorporate structure that is able to
 accommodate such information?  See also point 3 that now follows.}
 
-Hm, maybe store such information in \texttt{info} slots?
+Yes, one `just' needs to define a new class that inherits from constparty,
+possibly containing a cost-complexity table in the info slot with
+appropriate methods for pruning (the nodeprune function can be used).  An
+example on the use of nodeprune is given in the constparty vignette,
+Section~3, and mob has a novel internal pruning functionality.
 
@@
 
-See \citep{Hothorn+Hornik+Zeileis:2006}.
-
-But probably also rephrase...
+This is a somewhat philosphical issue, discussed in more detail in
+Section~3.4 of \cite{Hothorn+Hornik+Zeileis:2006}. If one grows a tree
+on a very large data set with nominal level of 5\%, many statistically
+significant but practically irrelevant splits will be found. For small and
+medium sized problems, conditional inference trees with nominal level of 5\% 
+and appropriate adjustment for multiplicity were shown to produce trees
+of the correct size \citep[simulation results
+in][]{Hothorn+Hornik+Zeileis:2006}.
 
@@
 
-Work in progress.
+In principle it is straightforward to set-up a class for forests as a list
+of partynode objects. The advantage is that only the tree
+structures vary across the ensemble but not the data. However, we found such
+a class of rather limited use because we didn't see a unified way to produce
+meaningful plots (variable importances, partial dependencies etc) for
+ensembles obtained by different forest algorithms yet. Implementing a new
+class for forests (maybe called parties or orgie) is on our to do list.
 
@@
 
-infrastructure independent of such things.
+The general infrastructure provided by partykit covers the representation of
+trees but not tree induction itself. This makes sense because the details of
+how splits are searched for and selected differ considerably between
+algorithms. The same is true with respect to finding the right-sized tree.
+
+The implementing of tree algorithms is very much simplified by partykit
+because the programmer can concentrate on the technical details of learning
+the tree while using the partykit infrastructure for representing and
+further processing the resulting tree. The constparty vignette contains a
+very dense example of such an implementation (Section 3). Note that a
+non-trivial tree algorithm can be implemented by three small R functions
+only.
+
+We also added an example showing how the partykit infrastructure can be used
+to quickly implement model assessment strategies.
 
@@
 
-completely.
+Completely. <Z FIXME> mob.
 
-\textit{If
-necessary, some of the historical detail in the overview might be replaced
+\textit{If necessary, some of the historical detail in the overview might be replaced
 by references to discussion that appears elsewhere, in order to make way for
@@
 
-pfff
+see additions to constparty vignette
+<Z FIXME>
 
@@
 
-mention: previous round of reviews asked for this format
+Our understanding of the reviews on an earlier version of this paper was
+that the referees asked for this format, please see our answer to Referee 1.
 
@@
 
-not a big software project, just two main developers, hence sometimes longer breaks
-
-we feel that this is not an indicator of the quality of the code, though
+The package was written by only two developers, hence sometimes longer
+breaks to the new jobs or family businesses could not be avoided.
+It is true that there was not much development in partykit between 2009 and
+2013. During this time, two offspring packages (evtree and psychotree) where
+developed instead and partykit was actively maintained. Please note that the
+development of these two new packages did only bring some minor design or
+implementation deficits in partykit to light. This raised our confidence in
+partykit. For these reasons we feel that pure activity this is not an indicator of 
+the quality of the code.
 
@@
 
-E-mails to the maintainer address were always possible and used fairly regularly
-by both end-users and other package developers. A mailing list was added now.
+E-mails to the maintainer address were always possible and are in fact the standard
+way of communicating with CRAN package maintainers (and developers) and were 
+used fairly regularly by both end-users and other package developers. 
+Nevertheless, a mailing list was added now.
 
@@
 
-Hard to judge, but reverse depends: NHEMOtree, quint. Number of citations of party paper.
+One can argue about this statement being close to reality. As of now, the
+package website lists seven reverse dependencies on CRAN, we are not
+involved in three of them. Only 6\% of all CRAN packages have more 
+reverse dependencies than partykit. And we hope that some of the 24 
+reverse dependencies of party will switch to partykit
+(only 2\% have more reverse dependencies than party) now that the reference
+implementation of conditional inference trees is the one from partykit.
+The ctree-publication was cited 643 times and the mob-paper 103 times 
+(Google scholar), so we are confident that people will increasingly use
+either ctree or mob and also hopefully learn about the partykit
+infrastructure this way.
 
@@
 trees, implementing a conversion layer to partykit objects wouldn't be any
-easier or harder than other packages.}
+easier or harder than other packages. Clearly, for developing new tree
+algorithms, though, partykit would be a good choice.}
 
-depends, eh?
+Our experience while implementing the original code for ctree 10 years ago
+was that the infrastructure existing then had way too many
+algorithm-specific things hardwired than to be of any use. While rpart
+always allowed to change risk functions, doing something outside of the
+CART-way of finding splits was not possible. Also printing and plotting
+facilities were not able to, for example, put p-values onto inner nodes.
 
-examples: CTree and display of p-values, J4.8 and multiway splits
+Our party implementation of ctree suffered the same problems, we had to work
+extremely hard to somehow use the party plot methods to plot J4.8 multiway 
+trees fitted by the RWeka package.
 
-\textit{
-Clearly, for developing new tree algorithms, though, partykit would be a good choice.}
-
-We actually have proof!
+These experiences led to the partykit package. The reimplementation of
+ctree, mob and the new implementations of trees in evtree and psychotree
+turned out to be rather straightforward and we personally did benefit a lot
+from the partykit infrastructure already. We suppose this is also true for
+the developers of the NHEMOtree and quint packages.
 
@@
 
-schwierig.
+Z
 
@@
 
-schwierig.
+Z
+
+Total cost of ownership
+
+demo mit mlbench:Satellite und rpart und J4.8
 
@@
 
-done.
+Based on our experience with the party package for 10 years now, the fact
+that other developers started packages using partykit, and the feedback via
+email and on conferences, most important the ``Workshop on Classification
+and Regression Trees'' that took place in March 2014, we are more optimistic
+about the popularity of partykit.
 

### 2014-09-05 pkg/partykit/inst/mloss/review2/p2p.tex
note demo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@599 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-We also added an example showing how the partykit infrastructure can be used
-to quickly implement model assessment strategies.
+We added an example showing how the partykit infrastructure can be used
+to quickly implement model assessment strategies to the end of Section~3 in
+the constparty vignette.
 
@@
 
-see additions to constparty vignette
+Please see the new example in Section~3 of the constparty vignette.
 <Z FIXME>
@@
 
-demo mit mlbench:Satellite und rpart und J4.8
+We added a demo (executable via demo("memory-speed")) that compares object
+sizes and prediction speed for larger rpart and J48 trees.
 

### 2014-09-05 pkg/partykit/demo/memory-speed.R
some small memory consumption and speed comparisons
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@600 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,38 @@
+
+library("rpart")
+library("RWeka")
+library("partykit")
+
+data("LetterRecognition", package = "mlbench")
+
+### fit rpart and J48 trees
+rp <- rpart(lettr ~ ., data = LetterRecognition)
+j48 <- J48(lettr ~ ., data = LetterRecognition)
+
+### convert to party
+system.time(party_rp <- as.party(rp))
+system.time(party_j48 <- as.party(j48))
+
+### check depths
+depth(party_rp)
+depth(party_j48)
+
+### compare object sizes
+object.size(rp)
+object.size(party_rp)
+
+object.size(j48) ### tree stored in external pointer
+object.size(party_j48)
+
+### set-up large sample
+i <- sample(1:nrow(LetterRecognition), 1000000, replace = TRUE)
+x <- LetterRecognition[i,]
+
+### compare predictions (speed and accuracy)
+system.time(p_rp <- predict(rp, newdata = x, type = "prob"))
+system.time(p_party_rp <- predict(party_rp, newdata = x, type = "prob"))
+all.equal(p_rp, p_party_rp)
+
+system.time(p_j48 <- predict(j48, newdata = x))
+system.time(p_party_j48 <- predict(party_j48, newdata = x))
+all.equal(p_j48, p_party_j48, check.attributes = FALSE)

### 2014-09-05 pkg/partykit/demo/memory-speed.R
Shuttle instead of LetterRecognition
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@602 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,38 +1,42 @@
-
+### packages and data
 library("rpart")
 library("RWeka")
 library("partykit")
-
-data("LetterRecognition", package = "mlbench")
+data("Shuttle", package = "mlbench")
 
 ### fit rpart and J48 trees
-rp <- rpart(lettr ~ ., data = LetterRecognition)
-j48 <- J48(lettr ~ ., data = LetterRecognition)
+rp <- rpart(Class ~ ., data = Shuttle)
+j48 <- J48(Class ~ ., data = Shuttle)
 
 ### convert to party
 system.time(party_rp <- as.party(rp))
 system.time(party_j48 <- as.party(j48))
 
-### check depths
+### check depth/width
 depth(party_rp)
+width(party_rp)
 depth(party_j48)
+width(party_j48)
 
 ### compare object sizes
-object.size(rp)
-object.size(party_rp)
+osize <- function(x) print(object.size(x), units = "Kb")
+osize(rp)                    ## rpart representation
+osize(party_rp)              ## full party (with terms, fitted values)
+osize(node_party(party_rp))  ## only the raw partynode
+osize(j48)                   ## J48 tree in external Java pointer
+osize(party_j48)             ## full party (with terms, fitted values)
+osize(node_party(party_j48)) ## only the raw partynode
+osize(Shuttle)               ## learning data (not stored in any tree)
 
-object.size(j48) ### tree stored in external pointer
-object.size(party_j48)
-
-### set-up large sample
-i <- sample(1:nrow(LetterRecognition), 1000000, replace = TRUE)
-x <- LetterRecognition[i,]
+### set-up large prediction sample
+set.seed(1)
+nd <- Shuttle[sample(1:nrow(Shuttle), 1e6, replace = TRUE), ]
 
 ### compare predictions (speed and accuracy)
-system.time(p_rp <- predict(rp, newdata = x, type = "prob"))
-system.time(p_party_rp <- predict(party_rp, newdata = x, type = "prob"))
+system.time(p_rp <- predict(rp, newdata = nd, type = "prob"))
+system.time(p_party_rp <- predict(party_rp, newdata = nd, type = "prob"))
 all.equal(p_rp, p_party_rp)
 
-system.time(p_j48 <- predict(j48, newdata = x))
-system.time(p_party_j48 <- predict(party_j48, newdata = x))
+system.time(p_j48 <- predict(j48, newdata = nd))
[…]

### 2014-09-05 pkg/partykit/inst/mloss/mloss2.Rnw
rewritten discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@603 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}
-\editor{(submission for MLOSS track, related to JMLR-07-240-1 that was handled by Cheng Soon Ong)}
+\editor{Cheng Soon Ong (MLOSS track, JMLR-14-133)}
 
@@
 \hline
-Algorithm  & Software                                                & Object class       & Original reference \\ \hline
+Algorithm  & Software implementation                                 & Object class       & Original reference \\ \hline
 CART/RPart & \code{rpart::rpart} + \code{as.party}                   & \code{constparty}  & \cite{Breiman+Friedman+Olshen:1984} \\
@@
 \end{tabular}
-\caption{\label{tab:overview} Selected tree algorithms than can be interfaced through \pkg{partykit}. The
-  software column lists external software, \proglang{R} functions from other packages (with \code{::} syntax)
+\caption{\label{tab:overview} Selected implementations of tree algorithms that can be interfaced through \pkg{partykit}. The
+  second column lists external software, \proglang{R} functions from other packages (with \code{::} syntax)
   and from \pkg{partykit}.}
@@
 to the same object class \class{constparty}. This is a general class that can
-in principle represent all the major classical tree algorithms with constant
+in principle represent all the major classical tree types with constant
 fits in the terminal nodes. Also, the same class is employed for conditional
@@
 emphasis to flexibility and extensibility. The infrastructure is easily
-accessible and written in plain \proglang{R}. Nevertheless speed is not a critical
-concern because (a)~lean data structures are employed and (b)~the time-intensive
-part in tree computations is typically inferring the tree rather than
-representing/printing/plotting it. Hence, the two functions for inferring trees
-either have computationally intensive parts in \proglang{C} (\fct{ctree})
-or build on \proglang{R}'s fitting functions (\fct{mob}). For benchmarks of
-their predictive performance, see their original references.
-Regarding memory requirements, the \class{party} trees are very lean but can
-optionally be used to store any information. The recursive node structure
-itself is extremely lightweight and optionally zero rows from the learning data
-are sufficient, which future versions will exploit for storing forests.
+accessible and accompanied by detailed manual pages and package vignettes.
+The package facilitates the implementation of new algorithms or interfacing
+other software by providing common building blocks for computing on trees
+(representation, printing, plotting, predictions, etc.). Using these building
+blocks developers of tree software can focus on implementing the learning
+algorithm (selection of variables and split points, stopping criteria, pruning, etc.).
+The package also provides functions for inferring trees where the computationally
+intensive parts are either in \proglang{C} (\fct{ctree}) or employ \proglang{R}'s
+fitting functions (\fct{mob}). The simple and lean base classes that separate
+data and tree structure are also appealing for storing forests which are planned
+for future versions of the package. Users and developers that have questions or
+comments about the package can either contact the maintainers or use the forum
+on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/forum/forum.php?forum_id=852}.
 

### 2014-09-06 pkg/partykit/inst/mloss/review2/p2p.tex
p2p reply improved/extended
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@605 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-09-05 pkg/partykit/inst/mloss/review2/p2p.tex
@@
 \documentclass{article}
-\usepackage{natbib}
+\usepackage{a4wide,natbib}
 
-\title{Rebuttal JMLR 14-133}
+\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
+\setlength{\parindent}{0em}
+
+\title{Revision of JMLR-14-133: Point-to-Point Reply}
 \author{Torsten Hothorn and Achim Zeileis}
@@
 
-\textit{
+\textit{%
 I regret to inform you that your submission ``partykit: A modular toolkit
@@
 If the authors choose to resubmit the project, please include a description
-addressing the reviewers' concerns.}
+addressing the reviewers' concerns.
+}
 
-\section*{Referee 1}
+\smallskip
 
-\textit{
+Thank you for the constructive reviews and feedback and for giving us the
+chance to revise our manuscript/software and submit it to the MLOSS track
+again. Please find attached ...
+
+The most important changes ... discussion rewritten ...
+
+We hope that this is suitable for consideration in the MLOSS track of JMLR now.
+
+\section*{Reviewer 1}
+
+\textit{%
 This is a clean and concise description of an interesting set of tools for
 data analysis with recursive partitioning.  It is mainly a description or
-announcement, and has minimal polemic or pedagogical value.  It will be a
-useful reference for those a basic outline of the software's capabilities,
-and provides a useful bibliography.}
+announcement, and has minimal polemic or pedagogical value.
+}
 
-Tja.
+\smallskip
 
-mention: previous round of reviews asked for this format of the paper
+Yes, this follows advice that we were given in the rejection of an earlier
+submission of our \emph{party} package (JMLR-07-240). One of the reviewers
+requested that the four-page manuscript should focus on broadbrush
+information and guide the reader to more detailed documentation available
+along with the software:
+\begin{quote} 
+\emph{Surely it is possible to give broadbrush information. There should be direct
+reference to web-based materials that give more detailed information and examples.}
+\end{quote}
 
-\section*{Referee 2}
+\medskip
 
-\textit{
+\textit{%
+It will be a useful reference for those a basic outline of the software's
+capabilities, and provides a useful bibliography.
+}
+
+Thank you.
+
+\section*{Reviewer 2}
+
+\textit{%
 This paper describes work that, in my understanding, does two things:
@@
 
-yes
+\smallskip
 
-maybe worth mentioning: due to the flexible infrastructure, the two
-specialized methods (ctree, mob) have become more flexible compared to
-earlier implementations (especially the mob framework)
+Thank you for this nice and concise summary of the software and accompanying
+four-page manuscript.
 
-\textit{
+When we started working on \emph{partykit}, we redesigned all classes from the
+earlier \emph{party} package from scratch and also reimplemented \texttt{ctree} and \texttt{mob}
+without relatively little overlap between the old and new code bases. As a
+consequence of the much cleaner \emph{partykit} design, the code base for both
+\texttt{ctree} and \texttt{mob} could be reduced considerably while at the same time
+both implementations gained more flexibility and additional options by leveraging
+the new class structure.
+
+\medskip
+
+\textit{%
 The software has a useful role in providing commonality across diverse
@@
 extensive documentation, some distributed with the package, is available
-from references given in the paper.}
+from references given in the paper.
+}
+
+\smallskip
+
+Thank you.
+
+\medskip
 
@@
 \item 
-\textit{
+\textit{%
 The legend for Table 1 begins: ``Selected tree algorithms than can be
 interfaced through partykit...''  Surely it is the output from selected
-implementations that can be interfaced, not the algorithm used.  The word
-`algorithm' seems to me anyway inappropriate in this context.  The same (or
-very similar) algorithm may have different implementations.  Different
+implementations that can be interfaced, not the algorithm used.
+}
+
+\smallskip
+
+Yes, thank you. We completely agree with you on the distinction between
+`algorithm' and its `implementation' and improved the caption accordingly.
+
+\medskip
+
+\textit{%
+The word `algorithm' seems to me anyway inappropriate in this context. The
+same (or very similar) algorithm may have different implementations.  Different
 methods with their inevitably different algorithms (including, in partykit:
 classical `constant' trees, model-based trees, and conditional inference
-trees) further wide widen the range of implementations.}
+trees) further widen the range of implementations.
+}
 
-rephrase: interfaces data structures used by different implementations of different
-tree growing methods / algorithms.
+\smallskip
+
+We completely agree and try to use ``algorithm'' for the conceptual method
+``implementation'' for an accompanying software throughout the paper now.
 
 \item 
-\textit{
+\textit{%
 The distinction between implementation (what the authors call `algorithm')
@@
 to use or process, e.g., results from rpart's on the fly cross-validation.
+}
+
+\smallskip
+
+Yes, this is correct. It is not possible to directly work on the \texttt{rpart}
+results after coercing to the \texttt{constparty} class.  The reason is that not
+all constant-fit trees are based on a risk or loss function and hence there
+is no uniform approach for performing cost-complexity pruning.
+Hence, the current implementation assumes that the pruning has already been
+performed within \texttt{rpart} (or \texttt{J48}) prior to coercion.
+
+\medskip
+
+\textit{%
 Would it be straightforward to incorporate structure that is able to
-accommodate such information?  See also point 3 that now follows.}
+accommodate such information?  See also point 3 that now follows.
+}
 
-Hm, maybe store such information in \texttt{info} slots?
+\smallskip
+
+Yes, one `just' needs to define a new class that inherits from \texttt{constparty},
+possibly containing a cost-complexity table in the \texttt{info} slot with
+appropriate methods for determining the nodes to be pruned and then call
+the \texttt{nodeprune} function. An example for the use of \texttt{nodeprune} is
+given in Section~3 of the \texttt{constparty} vignette. Furthermore,
+this type of approach (based on information criteria rather than cost-complexity)
+has also been coded into the \texttt{mob} function (while not having been available
+in the old \emph{party} implementation).
 
 \item 
-\textit{
+\textit{%
 I note the comment on the help page for the partykit function ctree() that
@@
 e.g., rpart or CART?  This is an issue for the help page, rather than
-perhaps for the paper.}
+perhaps for the paper.
+}
 
-See \citep{Hothorn+Hornik+Zeileis:2006}.
+\smallskip
+As discussed in Section~3.4 of \cite{Hothorn+Hornik+Zeileis:2006}, two
+different views on this issue can be taken. Either a nominal level for
+the statistical tests can be selected that is useful for the sample size
+at hand or the significance level can be treated as a hyperparameter that
+can be tuned when learning the tree. The manual page has been changed
+to better reflect this.
 
-But probably also rephrase...
+For small to medium-sized data sets, conditional inference trees with
+nominal level of 5\% (and appropriate adjustment for multiplicity) were
+shown to produce trees of the correct size \citep[simulation results
+in][]{Hothorn+Hornik+Zeileis:2006}.
 
 \item 
-\textit{
+\textit{%
 How well-adapted are the tools in partykit for implementing random forest
@@
 
-Work in progress.
+\smallskip
+
+In principle, it would be straightforward to extend the idea of the \texttt{party}
+class. This is stored as the data information (including terms etc.) and a separate recursive
+\texttt{partynode} representing the tree. Hence, the natural class for forests
+(possibly named \texttt{parties} or \texttt{orgy}, maybe) would be the data
+information plus a list of \texttt{partynode}s. This exploits that only the tree
+structures vary across the ensemble but not the data. However, the class itself
+would only be the beginning and one would additionally need a unified way to produce
+meaningful plots (e.g., variable importances, partial dependencies), predictions,
+etc. These are less straightforward and need further time and thinking prior
+to implementation. This is on our to-do list, though.
 
@@
 
-\textit{
+\textit{%
 At the very least, I would like to see brief comments on where splitting and
 accuracy assessment strategies, short of use of separate test data, fit in
-or around the toolkit.}
+or around the toolkit.
+}
 
-infrastructure independent of such things.
+\smallskip
+
+The functionality for learning the tree and assessing its accuracy has to
+be provided by developers of new tree methods and is not readily provided
+by \emph{partykit} itself (although certain building blocks are available).
+This is pointed out more clearly in the rewritten discussion at the end of
+the manuscript.
+
+In our experience, this separation of the general infrastructure makes sense
+because the details of how variables and splits are selected and how trees are
+pruned differ considerably between algorithms. Our previous implementations
+in the \emph{party} package made the mistake to blend rather general and very
+specific features into the same tree classes. And then when we wanted to implement or
+interface a different tree model, we either needed to abuse some slots of existing
+classes or had to start writing a new class.
+
+The \texttt{constparty} vignette contains a worked example (Section 3) of
+implementing a non-trivial tree algorithm from scratch with only three
+relatively short \textsf{R} functions. This should highlight that the
+developer can concentrate on the technical details of learning
+the tree while using the \emph{partykit} infrastructure for representing and
+further processing the resulting tree. In the latest release of \emph{partykit}
+we enhanced this section in the vignette by illustrating how model
+assessment strategies could be implemented for such a tree.
+
+\medskip
 
@@
 
-completely.
+\smallskip
 
-\textit{If
-necessary, some of the historical detail in the overview might be replaced
+All the base classes (\texttt{party}, \texttt{partynode}, \texttt{partysplit}) are
+completely independent of \texttt{ctree} and \texttt{mob}. The classes
+\texttt{constparty} and \texttt{modelparty} are also agnostic and can represent
+trees learned by other software (as clearly demonstated for \texttt{constparty})
+but, of course, have been made flexible enough to accomodate all features
+of \texttt{ctree} and \texttt{mob}, respectively.
+
+\medskip
+
+\textit{If necessary, some of the historical detail in the overview might be replaced
 by references to discussion that appears elsewhere, in order to make way for
@@
 
-pfff
+\smallskip
 
-\section*{Referee 3}
+We have extended the discussion and outlook in the four-page manuscript. Provided
+that we may keep the acknowledgments along with the references on the fifth page,
+no further shortening was necessary. However, if this does not conform with the
+MLOSS requirements we will follow your advice and streamline the introduction to
+save some space.
 
-\textit{
+\section*{Reviewer 3}
+
+\textit{%
 The paper describes an R library, `partykit', that appears to be a unified
@@
 
-mention: previous round of reviews asked for this format
+\smallskip
 
-\textit{
+As already pointed out in our reply to Reviewer~1, this format of the
+four-page manuscript pointing to the vignettes and other documentation was
+requested by one of the reviewers of an earlier version of the submission.
+
+\medskip
+
+\textit{%
 The project seems to have fairly minor activity, and appears to have had a
@@
 seems to indicate that the code is still being (somewhat) actively developed
-despite a relative lack of commits between 2009 and 2013.}
+despite a relative lack of commits between 2009 and 2013.
+}
 
-not a big software project, just two main developers, hence sometimes longer breaks
+\smallskip
 
-we feel that this is not an indicator of the quality of the code, though
+The package is developed by only two researchers, hence there can sometimes
+be longer breaks in the commits, e.g., due to working on other projects,
+applying for professor positions and switching jobs, or starting families.
 
-\textit{There appears to
-be no bug tracker or open mailing list.}
+However, during 2009 and 2013 the package was actively maintained with six
+CRAN releases, e.g., fixing bugs or incorporating smaller user requests. Also
+during this time, two offspring packages (\emph{evtree} and \emph{psychotree}) were
+developed that leverage the \emph{partykit} infrastructure. These brought
+only some smaller design or implementation deficits to light that did not
+necessitate (re-)writing/modifying large parts of the code.
 
-E-mails to the maintainer address were always possible and used fairly regularly
-by both end-users and other package developers. A mailing list was added now.
+Therefore, we feel that \emph{partykit} has matured by now and pure activity
+is not necessarily an indicator of the quality of the code.
 
-\textit{There is knowledge of the package
-in the R ecosystem, but partykit does not seem to be as popular as other R
-packages (admittedly, partykit is a niche product).}
+\medskip
 
-Hard to judge, but reverse depends: NHEMOtree, quint. Number of citations of party paper.
+\textit{%
+There appears to be no bug tracker or open mailing list.
+}
 
-\textit{
+\smallskip
+
+E-mails to the maintainer address were always possible and this is still a very common
+way of communicating with CRAN package maintainers (and developers). In the past,
+this means of communication with us was used fairly regularly by both end-users
+and other package developers.
+
+As \textsf{R}-Forge easily allows for having mailing lists, forums, etc., we
+have added a discussion forum and also point the readers to this in the rewritten
+Section~5 of our manuscript.
+
+\medskip
+
+\textit{%
+There is knowledge of the package in the R ecosystem, but partykit does not seem
+to be as popular as other R packages (admittedly, partykit is a niche product).
+}
+
+\smallskip
+
+There are certainly more popular (but typically also more generally applicable) packages
+in the \textsf{R} world. However, we feel that \emph{partykit} is doing quite
+well.
+
+As of now, CRAN package web page lists seven reverse dependencies (in two of
+which we have no involvement at all and two others that are not maintained by us).
+Only 6\% of all CRAN packages have more reverse dependencies than that. And we hope
+that some of the 24~reverse dependencies of \emph{party} will switch to \emph{partykit}
+(only 2\% of all CRAN packages have more than 24 reverse dependencies) now that the
+recommended reference implementations of \texttt{ctree} and \texttt{mob} are in \emph{partykit}.
+Moreover, according to Google Scholar the \texttt{ctree} and \texttt{mob} papers
+published in \emph{Journal of Computational and Graphical Statistics} were cited
+643 and 103 times, respectively.
+
+Having said that, we nevertheless feel that such popularity measures not necessarily
+capture the quality and scientific value of a software project.
+
+\medskip
+
+\textit{%
 Support for trees produced by other libraries is provided, but I am not
@@
 trees, implementing a conversion layer to partykit objects wouldn't be any
-easier or harder than other packages.}
+easier or harder than other packages. Clearly, for developing new tree
+algorithms, though, partykit would be a good choice.
+}
 
-depends, eh?
+\smallskip
 
-examples: CTree and display of p-values, J4.8 and multiway splits
+Our experience is not that any tree model can be converted in the same easy
+or hard way to any other packages.
 
-\textit{
-Clearly, for developing new tree algorithms, though, partykit would be a good choice.}
+While implementing the original code for \texttt{ctree} 10~years ago, we could
+not easily leverage infrastructure that existed at the time because it had way too many
+algorithm-specific things hardwired into it. While \texttt{rpart}
+always allowed to change risk functions, doing something outside of the
+CART-way of finding splits was not really possible. Also printing and plotting
+facilities were somewhat limited and not directly able to directly include $p$-values
+in the inner nodes, for example.
+Our \emph{party} implementation of \texttt{ctree} then made the same mistake of
+wiring too many algorithm-specific requirements into the class. Hence, it really
+had to be abused to capture \texttt{mob} trees. The multi-way splits of
+\texttt{J48} from \emph{RWeka} could not be captured by either \emph{rpart}
+or \emph{party}.
 
-We actually have proof!
+These experiences led to the \emph{partykit} package. The reimplementation of
+\texttt{ctree}, \texttt{mob} and the new implementations of trees in \emph{evtree} and \emph{psychotree}
+turned out to be rather straightforward and we personally did benefit a lot
+from the \emph{partykit} infrastructure already. We suppose this is also true for
+the developers of the \emph{NHEMOtree} and \emph{quint} packages.
 
-\textit{
+\medskip
+
+\textit{%
 Section 5 is very qualitative and vague.  The authors should definitely
@@
 quantitative details, it's difficult to believe those statements without
-diving into the code.}
+diving into the code.
+}
 
-schwierig.
+\smallskip
 
-\textit{
+Yes, your are right, thank you for pointing this out. We have completely
+rewritten Section~5, refraining from going into details about memory and
+speed comparisons and highlighting the flexibility/extensibility of the
+package instead.
+
+\medskip
+
+\textit{%
 In my opinion the authors could do one of these things to validate their
@@
 represented and handled by partykit, or simply leave out or significantly
-shorten the discussion of speed and memory usage entirely.}
+shorten the discussion of speed and memory usage entirely.
+}
 
-schwierig.
+\smallskip
 
-\textit{
+We have added a demo to the package that briefly illustrates the
+memory and speed considerations that we alluded to in the initial submission.
+This can be run interactively in \textsf{R} via
+\texttt{demo("memory-speed", package = "partykit")}. However, it is not
+mentioned in the manuscript as this is not the main focus of the package.
+
+\medskip
+
+\textit{%
 Overall, partykit is a well-designed and extensively documented package, but
@@
 going into the future.  The developers should consider setting up a mailing
-list or bug tracker to make the project more open to new contributors.}
+list or bug tracker to make the project more open to new contributors.
+}
 
-done.
+\smallskip
+
+Based on our experience with the \emph{party} package for 10 years now, the fact
+that other developers started packages building on \emph{partykit}, and the feedback via
+e-mail and at conferences (most importantly the recent ``Workshop on Classification
+and Regression Trees'' at NUS in March 2014), we feel sufficiently confident
+about the popularity of \emph{partykit}.
+
+As for the openness to new contributors: Both end-users and package developers
+are invited to contact us personally, via e-mail or through the newly established
+forum. If (reasonable) changes in \emph{partykit} are necessary we try to accomodate them.
+More typically, however, new contributions based on \emph{partykit} can be more easily
+wrapped into new packages which may (as in the case of \emph{evtree}) or may
+not (as for \emph{quint} and \emph{NHEMOtree}) lead to co-development with us.
 
 \bibliographystyle{plainnat}
-\bibliography{../../../vignettes/party.bib}
+\bibliography{../ref.bib}
 

### 2014-09-07 pkg/partykit/inst/mloss/review2/p2p.tex
improvements and reply to AE
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@607 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \documentclass{article}
-\usepackage{a4wide,natbib}
+\usepackage{a4wide}
+\usepackage[authoryear,round]{natbib}
 
@@
 
-\textit{%
+\textbf{\textit{%
 I regret to inform you that your submission ``partykit: A modular toolkit
@@
 addressing the reviewers' concerns.
-}
+}}
 
@@
 chance to revise our manuscript/software and submit it to the MLOSS track
-again. Please find attached ...
+again. Please find attached the revision of our four-page manuscript
+and a new version (0.8-1) of the \emph{partykit} package which is already
+available from \texttt{http://CRAN.R-project.org/package=partykit} as well.
 
-The most important changes ... discussion rewritten ...
+The most important changes are a completely rewritten Section~5 in the four-page
+manuscript (Discussion and Outlook) and an extended Section~3 in the \texttt{constparty}
+vignette (showing how new tree algorithms can be implemented using \emph{partykit}).
+Moreover, an online support forum for \emph{partykit} has been established
+on \textsf{R}-Forge at \texttt{https://R-Forge.R-project.org/forum/forum.php?forum\_id=852}.
+Further details of the revision are discussed in the point-to-point reply
+to the reviewers.
+
+Note that the revised four-page manuscript now has the acknowledgments (four lines)
+on the fifth page along with the references. We are not sure whether this
+conforms with the MLOSS style guidelines: if not, we will, of course, streamline
+the main manuscript to place the acknowledgments on the first four pages as well.
 
@@
 
-\textit{%
+\textbf{\textit{%
 This is a clean and concise description of an interesting set of tools for
@@
 announcement, and has minimal polemic or pedagogical value.
-}
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 It will be a useful reference for those a basic outline of the software's
 capabilities, and provides a useful bibliography.
-}
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 This paper describes work that, in my understanding, does two things:
@@
 \end{enumerate}
-}
+}}
 
@@
 earlier \emph{party} package from scratch and also reimplemented \texttt{ctree} and \texttt{mob}
-without relatively little overlap between the old and new code bases. As a
+with relatively little overlap between the old and new code bases. As a
 consequence of the much cleaner \emph{partykit} design, the code base for both
@@
 
-\textit{%
+\textbf{\textit{%
 The software has a useful role in providing commonality across diverse
@@
 from references given in the paper.
-}
+}}
 
@@
 
-\textit{Quibbles and queries are:}
+\textbf{\textit{%
+Quibbles and queries are:
+}}
 \begin{enumerate}
 \item 
-\textit{%
+\textbf{\textit{%
 The legend for Table 1 begins: ``Selected tree algorithms than can be
@@
 implementations that can be interfaced, not the algorithm used.
-}
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 The word `algorithm' seems to me anyway inappropriate in this context. The
@@
 trees) further widen the range of implementations.
-}
+}}
 
@@
 \item 
-\textit{%
+\textbf{\textit{%
 The distinction between implementation (what the authors call `algorithm')
@@
 to use or process, e.g., results from rpart's on the fly cross-validation.
-}
+}}
 
@@
 Yes, this is correct. It is not possible to directly work on the \texttt{rpart}
-results after coercing to the \texttt{constparty} class.  The reason is that not
+results after coercion to the \texttt{constparty} class.  The reason is that not
 all constant-fit trees are based on a risk or loss function and hence there
 is no uniform approach for performing cost-complexity pruning.
-Hence, the current implementation assumes that the pruning has already been
+Therefore, the current implementation assumes that the pruning has already been
 performed within \texttt{rpart} (or \texttt{J48}) prior to coercion.
@@
 
-\textit{%
+\textbf{\textit{%
 Would it be straightforward to incorporate structure that is able to
 accommodate such information?  See also point 3 that now follows.
-}
+}}
 
@@
 \item 
-\textit{%
+\textbf{\textit{%
 I note the comment on the help page for the partykit function ctree() that
@@
 perhaps for the paper.
-}
+}}
 
@@
 \item 
-\textit{%
+\textbf{\textit{%
 How well-adapted are the tools in partykit for implementing random forest
@@
 an equivalent of the cforest() function that is in the older party package.
-}
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 At the very least, I would like to see brief comments on where splitting and
@@
 or around the toolkit.
-}
+}}
 
@@
 
-\textit{To what extent are toolkit functions independent of
-the methodology used for the partykit functions ctree() and mob()?}
+\textbf{\textit{%
+To what extent are toolkit functions independent of
+the methodology used for the partykit functions ctree() and mob()?
+}}
 
@@
 
-\textit{If necessary, some of the historical detail in the overview might be replaced
+\textbf{\textit{%
+If necessary, some of the historical detail in the overview might be replaced
 by references to discussion that appears elsewhere, in order to make way for
-comment on these points.}
+comment on these points.
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 The paper describes an R library, `partykit', that appears to be a unified
@@
 basic tutorial with links to very detailed R vignettes.  The package also
-implements a few of the authors' own recursive partitioning algorithms.}
+implements a few of the authors' own recursive partitioning algorithms.
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 The project seems to have fairly minor activity, and appears to have had a
@@
 despite a relative lack of commits between 2009 and 2013.
-}
+}}
 
@@
 only some smaller design or implementation deficits to light that did not
-necessitate (re-)writing/modifying large parts of the code.
+necessitate (re-)writing/modifying large parts of the \emph{partykit} code.
 
@@
 
-\textit{%
+\textbf{\textit{%
 There appears to be no bug tracker or open mailing list.
-}
+}}
 
@@
 
-E-mails to the maintainer address were always possible and this is still a very common
+E-mails to the maintainer address were always possible and this is a rather common
 way of communicating with CRAN package maintainers (and developers). In the past,
@@
 As \textsf{R}-Forge easily allows for having mailing lists, forums, etc., we
-have added a discussion forum and also point the readers to this in the rewritten
-Section~5 of our manuscript.
+have added a discussion forum (\texttt{https://R-Forge.R-project.org/forum/forum.php?forum\_id=852})
+and also point the readers to this in the rewritten Section~5 of our manuscript.
 
@@
 
-\textit{%
+\textbf{\textit{%
 There is knowledge of the package in the R ecosystem, but partykit does not seem
 to be as popular as other R packages (admittedly, partykit is a niche product).
-}
+}}
 
@@
 
-As of now, CRAN package web page lists seven reverse dependencies (in two of
+As of now, the CRAN package web page lists seven reverse dependencies (in two of
 which we have no involvement at all and two others that are not maintained by us).
@@
 
-\textit{%
+\textbf{\textit{%
 Support for trees produced by other libraries is provided, but I am not
@@
 algorithms, though, partykit would be a good choice.
-}
+}}
 
@@
 Our experience is not that any tree model can be converted in the same easy
-or hard way to any other packages.
+or hard way to any other package.
 
@@
 CART-way of finding splits was not really possible. Also printing and plotting
-facilities were somewhat limited and not directly able to directly include $p$-values
+facilities were somewhat limited and not able to directly include $p$-values
 in the inner nodes, for example.
@@
 wiring too many algorithm-specific requirements into the class. Hence, it really
-had to be abused to capture \texttt{mob} trees. The multi-way splits of
+had to be somewhat abused to capture \texttt{mob} trees. And the multi-way splits of
 \texttt{J48} from \emph{RWeka} could not be captured by either \emph{rpart}
@@
 
-\textit{%
+\textbf{\textit{%
 Section 5 is very qualitative and vague.  The authors should definitely
-improve this in future revisions.  It is written, "lean data structures are
-employed", "the `party' structures are very lean", and also "The recursive
+improve this in future revisions.  It is written, ``lean data structures are
+employed", ``the `party' structures are very lean", and also ``The recursive
 node structure itself is extremely lightweight" but without any sort of
@@
 diving into the code.
-}
+}}
 
@@
 
-\textit{%
+\textbf{\textit{%
 In my opinion the authors could do one of these things to validate their
@@
 shorten the discussion of speed and memory usage entirely.
-}
+}}
 
@@
 We have added a demo to the package that briefly illustrates the
-memory and speed considerations that we alluded to in the initial submission.
+memory and speed considerations that we alluded to in the initial submission
+by comparison to \texttt{rpart} and \texttt{J48}.
 This can be run interactively in \textsf{R} via
@@
 
-\textit{%
+\textbf{\textit{%
 Overall, partykit is a well-designed and extensively documented package, but
@@
 list or bug tracker to make the project more open to new contributors.
-}
+}}
 

### 2014-09-08 pkg/partykit/inst/mloss/letter_mloss2.tex
resubmission letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@609 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-03-27 pkg/partykit/inst/mloss/letter_mloss.tex
@@
 
-\subject{JMLR-MLOSS submission}
+\subject{JMLR-14-133 resubmission}
 
@@
 
-my co-author Achim Zeileis and I would like to submit the manuscript
+my co-author Achim Zeileis and I would like to resubmit the manuscript
 ``\textbf{partykit}: An Open-Source Toolkit for Recursive Partytioning in R'' to the
@@
 
-The \textsf{R} add-on package \textbf{partykit} described in this submission implements a
-flexible toolkit for learning, representing, summarizing, and visualizing a
-wide range of tree-structured models.  The package offers a basic class
-infrastructure for representing fitted trees, a unified set of methods for
-printing and plotting of trees as well as for computing predictions for new
-observations, tailored functionality for trees with constant fits and with
-parametric models in each terminal node, and reimplementations of conditional
-inference trees (CTree) and model-based recursive partitioning (MOB).
+We would like to thank the referees for their constructive reviews and
+action editor Cheng Soon Ong for overseeing the review process. While
+preparing this resubmission, we took all issues discussed by the referees
+into account and revised both the \emph{partykit} package and the four-page
+description. A detailed description of our revision can be found in the
+point to point reply.
 
-This manuscript is related to JMLR-07-240-1 ``Let's Have a party! An
-Open-Source Toolbox for Recursive Partytioning'' describing the
-\textbf{party} add-on
-package.  This manuscript was rejected with encouragement for resubmission by
-JMLR-MLOSS editor Cheng Soon Ong in November 2007.  The main concerns of the
-reviewers were that it was not easy to extend the old \textbf{party} package by adding
-new tree learners or representing existing trees in our package.  This was
-a very important suggestion and part of the motivation for starting from
-scratch and develop a new R package (\textbf{partykit}) that easily allows to
-accommodate our own tree algorithms but makes the underlying infrastructure
-easily accessible to others. Actually, the \textbf{partykit} package in its current
-form goes well beyond what was suggested in the reviews of JMLR-07-240-1
-and both manuscript and software were entirely rewritten. Hence, we felt it
-to be more appropriate to submit this as a new manuscript rather than a
-revision of JMLR-07-240-1.  However, if you would like to see a formal
-point-to-point reply to the old submission, please let us know.
+The package web page is \texttt{http://CRAN.R-project.org/package=partykit}
+and points to the revised version 0.8-1 distributed under the GPL (2 or 3).
+This version is also uploaded for review.
 
 Sincerely yours, \\
-\includegraphics[width = .3\textwidth]{TH_Unterschrift}
+\includegraphics[width = .25\textwidth]{TH_Unterschrift}
 

### 2014-09-08 pkg/partykit/inst/mloss/review2/p2p.tex
rephrase
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@608 255f4249-2263-4417-b1ea-8498779e27a7
resubmission
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@610 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\title{Revision of JMLR-14-133: Point-to-Point Reply}
+\title{Resubmission of JMLR-14-133: Point-to-Point Reply}
 \author{Torsten Hothorn and Achim Zeileis}
@@
 
-We hope that this is suitable for consideration in the MLOSS track of JMLR now.
+We hope that the \emph{partykit} package and the four-page description are
+suitable for consideration in the MLOSS track of JMLR now.
 

### 2014-09-12 pkg/partykit/inst/mloss/letter_mloss2.tex
correct title
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@611 255f4249-2263-4417-b1ea-8498779e27a7

@@
 my co-author Achim Zeileis and I would like to resubmit the manuscript
-``\textbf{partykit}: An Open-Source Toolkit for Recursive Partytioning in R'' to the
+``\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in R'' to the
 Machine Learning Open-Source Software track of the Journal of Machine
@@
 preparing this resubmission, we took all issues discussed by the referees
-into account and revised both the \emph{partykit} package and the four-page
+into account and revised both the \textbf{partykit} package and the four-page
 description. A detailed description of our revision can be found in the

### 2014-10-22 pkg/partykit/inst/mloss/letter_mloss2.tex
forgot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@625 255f4249-2263-4417-b1ea-8498779e27a7

@@
 The package web page is \texttt{http://CRAN.R-project.org/package=partykit}
-and points to the revised version 0.8-1 distributed under the GPL (2 or 3).
+and points to the revised version 0.8-2 distributed under the GPL (2 or 3).
 This version is also uploaded for review.

### 2014-10-22 pkg/partykit/inst/mloss/review3/p2p.tex
prepare revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@626 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass{article}
\usepackage{a4wide}
\usepackage[authoryear,round]{natbib}

\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
\setlength{\parindent}{0em}

\title{JMLR-14-133-3: Point-to-Point Reply}
\author{Torsten Hothorn and Achim Zeileis}

\begin{document}

\maketitle

\section*{AE}

\textbf{\textit{%
I am glad to inform you that your paper ``partykit: A Modular Toolkit for
Recursive Partytioning in R'' has been accepted for publication at JMLR Track for Machine Learning
Open Source Software. There are still some minor issues that should be fixed in the final version, which will
be sent out to Reviewer 3 for a quick confirmation.
As pointed out by Reviewer 3, partykit only provides a subset of the functionality of its
earlier cousin party. Porting the old functionality of party to the new
interface of partykit would improve its usefulness.}}

\textbf{\textit{%
After these changes have been completed, I would like to ask you to prepare
your paper for final submission following the instructions given at
http://www.jmlr.org/format/authors-guide.html.
Please fill out both the permission to publish form and the source code
release form.}}

\smallskip

\section*{Reviewer 3}

\textbf{\textit{%
The revised version of the paper is reasonable, especially the rewritten
Section 5.  I have no further complaints about the writing of the paper
itself.}}

\textbf{\textit{%
However, I still have issues with the project itself.  My primary concerns
from the previous review are listed below:
\begin{itemize}
 \item The project has little activity.
 \item There appears to be no bug tracker or open mailing list.
 \item The project itself lives in a small niche.
\end{itemize}
The point-by-point rebuttal addresses each of these issues.  The authors
point out that partykit is a mature package and at this point realistically
only requires maintenance.  The authors have also set up a forum (though no
questions have yet been asked).  To address the popularity of partykit, the
authors point out several reverse dependencies of partykit in CRAN, although
they feel that ``such popularity measures [do] not necessarily capture the
quality and scientific value of a software project'' (and this is a statement
I very much agree with).  The authors have pointed out that the algorithms
implemented by partykit are useful.}}

\textbf{\textit{%
I am satisfied with these rebuttals, but the underlying theme of those three
comments can be summarized as ``the project has limited scope''.  My concern
is motivated by criterion 2 of the "Review Criteria: 2. The novelty, breadth, and significance of the contribution (including
evidence of an active user community)''.}}

\textbf{\textit{%
Given that the package is primarily a reimplementation of the authors' own
algorithms "ctree" and "mob" plus a nice, unified API and wrappers for other
recursive partitioning algorithms, I am not convinced that the breadth of
this contribution is sufficient, given the much larger breadth of
contributions to JMLR-MLOSS such as Shogun, scikit-learn, Shark, mlpack,
dlib-ml, and other large-scale toolkits.}}

\textbf{\textit{%
Reviewer 2 from the last review cycle comments that the cforest()
functionality from party is not available in partykit, and today this is
still the case.  I would encourage the others to consider expanding the
breadth of partykit to at least entirely encompass the older party package,
and also to use their own clean and nice API for some implementations of
other recursive partitioning algorithms.  This should not be a problem,
given that the clean API allows easy implementation of other algorithms.
Increasing the breadth of the package should also increase its visibility,
usage, and popularity.
Overall, I do not want to give the authors the idea that I think their
package is not a good example of well-thought-out software engineering. 
partykit is well-designed and extensively documented.  But my primary
concern is a lack of breadth.
}}

\end{document}

### 2014-10-22 pkg/partykit/inst/mloss/letter_mloss3.tex
prepare p2p and letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@627 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[ngerman,uzh]{scrlttr2}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

%% ANPASSUNGSMÖGLICHKEITEN
\KOMAoptions{%
%backaddress=false%  Keine Rücksendeadresse im Adressfeld
%,foldmarks=false%  Faltmarken ausschalten
}

\usepackage{babel}

\begin{document}

%% VARIABLEN
% Falls Unterschrift vom definierten Namen abweicht:
\setkomavar{signature}{Torsten Hothorn}
\setkomavar{subject}{Betreff}
%\setkomavar{specialmail}{Interne Post}

\begin{letter}{
Dr.~Cheng Soon Ong \\
MLOSS Editor \\
Journal of Machine Learning Research}

\subject{JMLR-14-133-3 revision}

\opening{Dear Cheng Soon,}

I uploaded a rebuttal letter for our manuscript
``\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in R'' to the
Machine Learning Open-Source Software track of the Journal of Machine
Learning Research.

Referee 3 criticised the limited breath of our contribution and suggested to
copy the \texttt{cforest} function from package \emph{party} to package
\emph{partykit}, iterating an argument by Referee 2 from the previous
round. In the rebuttal letter, we explain the reason why it is currently not
helpful to move this function from one CRAN package to another one. In
addition, we took the liberty to discuss why we think that \emph{partykit}
is actually much more general than \emph{party} and, if one adopts a certain
point of view, also more general than most of the established machine
learning toolboxes. Since no changes in the manuscript were requested, we
did not touch the four page description.

Sincerely yours, \\
\includegraphics[width = .25\textwidth]{TH_Unterschrift}

\end{letter}

\end{document}

### 2014-10-22 pkg/partykit/inst/mloss/review3/p2p.tex
prepare p2p and letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@627 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\textbf{\textit{%
-After these changes have been completed, I would like to ask you to prepare
-your paper for final submission following the instructions given at
-http://www.jmlr.org/format/authors-guide.html.
-Please fill out both the permission to publish form and the source code
-release form.}}
+We don't see how moving functionality from one R package to another one
+increases the usefulness of the latter for the data analyst. As explained in our response
+to referee 2 and also below in our response to the new comments by referee
+3, a much improved implementation of \texttt{cforest} and its tools is on our
+research agenda but many of the details require further thinking.
 
@@
 only requires maintenance.  The authors have also set up a forum (though no
-questions have yet been asked).  To address the popularity of partykit, the
+questions have yet been asked).}}
+
+\smallskip
+
+As we already mentioned in our reply to this earlier comment, the use of forums
+is not very popular in the R community. In the time after we submitted our
+revised version, we received a number of comments and reports by different
+users that gave rise to changes in the \emph{partykit} package (svn
+revisions 612, 616 and 624). Our experience as maintainers of 15 (TH) and 20
+(AZ) CRAN packages shows that this corresponds to ``normal'' activity for
+comparable CRAN packages.
+
+\smallskip
+
+\textbf{\textit{%
+To address the popularity of partykit, the
 authors point out several reverse dependencies of partykit in CRAN, although
@@
 algorithms "ctree" and "mob" plus a nice, unified API and wrappers for other
-recursive partitioning algorithms, I am not convinced that the breadth of
+recursive partitioning algorithms,}}
+
+\smallskip
+
+In our opinion \emph{partykit} is not a reimplementation of \emph{party}.
+A submission describing package \emph{party} was evaluated by the MLOSS
+track of the JMLR (as JMLR-07-240) and rejected mainly because the referees
+thought (correctly, as we are convinced now) that the design and
+implementation of \texttt{ctree}, \texttt{mob} and \texttt{cforest} was not general
+enough. Package \emph{partykit} in fact generalises \emph{party} with only one exception
+(\texttt{cforest}) for very good reasons as explained below.
+
+\smallskip
+
+\textbf{\textit{%
+I am not convinced that the breadth of
 this contribution is sufficient, given the much larger breadth of
@@
 dlib-ml, and other large-scale toolkits.}}
+
+\smallskip
+
+It depends on how one measures ``breadth''. The mentioned toolkits provide a
+large set of methods for binary and multiclass classification, regression
+and clustering with a focus on SVMs.  \emph{Shark} and \texttt{mlpack} also
+contain implementations of tree algorithms for these situations.  Unlike
+\emph{partykit}, none of the toolkits provides methods for censored data (survival
+analysis) or ordered categorical regression.  With \texttt{mob},
+\emph{partykit} allows partitioning of a huge set of models, such as transformation models (Cox,
+Weibull), proportional odds models or IRT (item response theory) models
+(implemented in package \emph{psychotree} based on \emph{partykit} infrastructure).
+
+\smallskip
 
@@
 other recursive partitioning algorithms.  This should not be a problem,
-given that the clean API allows easy implementation of other algorithms.
+given that the clean API allows easy implementation of other algorithms.}}
+
+\smallskip
+
+In our response to Referee 2, we pointed out that a general implementation
+of forest-like procedures requires further research into its design. For
+example, there are multiple aggregation schemes that need to be taken into
+account as a simple majority vote or mean (the original proposal by Breiman)
+might be less attractive compared to the understanding of forests as
+estimators of conditional distribution functions \citep{Meinshausen_2006}. In
+addition, a pure forest is almost useless. Variable selection procedures
+(for example based on variable importances and stability selection) and visualisation techniques are
+the more interesting (and technically more challenging) things. Simply
+transforming currently available \texttt{cforest} functionality from \emph{party} to
+\emph{partykit}
+will not fix the currently present shortcomings in the design and
+implementation of these procedures. In addition, it will not increase the
+number of available procedures because the \emph{party} package is still maintained
+on CRAN and will be available for the forseeable future.
+
+\smallskip
+
+\textbf{\textit{%
 Increasing the breadth of the package should also increase its visibility,
@@
 
+\smallskip
+
+We are confident that the current \emph{partykit} infrastructure and API
+will motivate other researcher to implement novel tree-based procedures that
+in turn will increase the breath of recursive partitioning methods available
+to R users.  For example, package \emph{vcrpart}
+(\texttt{http://CRAN.R-project.org/package=vcrpart}) is a newly released
+package that implements varying coefficient models based on \emph{partykit}
+infrastructure.  The package was developled completely independent of the
+\emph{partykit} authors and, btw, the authors of \emph{vcrpart} never posted
+a question by email or in the forum which we tend to attribute to the
+extensive package documentation of \emph{partykit}.  Novel procedures such
+as the one implemented in \emph{vcrpart} is exactly the notion of ``breath''
+we had in mind when designing \emph{partykit}.  We expect other research
+groups to reuse \emph{partykit infrastructure} in the future, especially
+after the package and corresponding description was made popular by MLOSS.
+
+\bibliographystyle{plainnat}
+\bibliography{../ref.bib}
+
 \end{document}

### 2014-10-22 pkg/partykit/inst/mloss/review3/p2p.tex
revised p2p comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@628 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-10-22 pkg/partykit/inst/mloss/review3/p2p.tex
@@
 Recursive Partytioning in R'' has been accepted for publication at JMLR Track for Machine Learning
-Open Source Software. There are still some minor issues that should be fixed in the final version, which will
+Open Source Software.
+}}
+
+\smallskip
+
+Thank you for the good news, this is much appreciated.
+
+\smallskip
+
+\textbf{\textit{%
+There are still some minor issues that should be fixed in the final version, which will
 be sent out to Reviewer 3 for a quick confirmation.
@@
 
+\smallskip
+
+This suggestion was already brought up in the previous revision by Reviewer~2
+and answered to the satisfaction of the first two reviewers: While the basic idea for
+setting up a generic class for ``random forests'' is straightforward, many of the
+details in the implementation are not obvious at all and deserve much more attention.
+See also our full response included below.
+
+Of course, we can easily copy the code of the old \texttt{party::cforest} into the
+new \emph{partykit} package. However, we fail to see how this would increase usefulness
+for practitioners given that \emph{party} is easily available. More importantly, this
+would conflict with the view of Reviewer~2 who apparently agreed that more future
+work should be devoted to this important topic.
+
+The full question and response from the previous round of reviews was:
+%
+\begin{quote}
 \textbf{\textit{%
-After these changes have been completed, I would like to ask you to prepare
-your paper for final submission following the instructions given at
-http://www.jmlr.org/format/authors-guide.html.
-Please fill out both the permission to publish form and the source code
-release form.}}
+How well-adapted are the tools in partykit for implementing random forest
+or other ensemble type approaches? I note that partykit does not (yet?) have
+an equivalent of the cforest() function that is in the older party package.
+}}
+
+In principle, it would be straightforward to extend the idea of the \texttt{party}
+class. This is stored as the data information (including terms etc.) and a separate recursive
+\texttt{partynode} representing the tree. Hence, the natural class for forests
+(possibly named \texttt{parties} or \texttt{orgy}, maybe) would be the data
+information plus a list of \texttt{partynode}s. This exploits that only the tree
+structures vary across the ensemble but not the data. However, the class itself
+would only be the beginning and one would additionally need a unified way to produce
+meaningful plots (e.g., variable importances, partial dependencies), predictions,
+etc. These are less straightforward and need further time and thinking prior
+to implementation. This is on our to-do list, though.
+\end{quote}
 
@@
 itself.}}
+
+\smallskip
+
+Thank you, this is much appreciated.
+
+\smallskip
 
@@
 only requires maintenance.  The authors have also set up a forum (though no
-questions have yet been asked).  To address the popularity of partykit, the
+questions have yet been asked).}}
+
+\smallskip
+
+The lack of questions in the forum is due to the fact that the forum was not yet
+advertised. However, it will be once the manuscript is published in the MLOSS
+track and we expect that the forum will be used more actively then.
+
+So far, we (like many other package maintainers in the \textsf{R} community)
+have simply used e-mails for correspondence about problems, bugs, feature requests, etc.
+For example, in the one month since resubmission of the revised manuscript to the
+MLOSS track we received a number of comments and reports by different
+users that gave rise to several small changes in the \emph{partykit} package (SVN
+revisions 612, 616, and 624).
+
+\smallskip
+
+\textbf{\textit{%
+To address the popularity of partykit, the
 authors point out several reverse dependencies of partykit in CRAN, although
@@
 comments can be summarized as ``the project has limited scope''.  My concern
-is motivated by criterion 2 of the "Review Criteria: 2. The novelty, breadth, and significance of the contribution (including
-evidence of an active user community)''.}}
+is motivated by criterion 2 of the "Review Criteria: 2. The novelty, breadth,
+and significance of the contribution (including evidence of an active user
+community)''.}}
 
@@
 Given that the package is primarily a reimplementation of the authors' own
-algorithms "ctree" and "mob" plus a nice, unified API and wrappers for other
-recursive partitioning algorithms, I am not convinced that the breadth of
-this contribution is sufficient, given the much larger breadth of
-contributions to JMLR-MLOSS such as Shogun, scikit-learn, Shark, mlpack,
-dlib-ml, and other large-scale toolkits.}}
+algorithms \texttt{ctree} and \texttt{mob} plus a nice, unified API and wrappers for other
+recursive partitioning algorithms,}}
+
+\smallskip
+
+This appears to convey that the reimplementation of \texttt{ctree} and \texttt{mob}
+is not very substantial and only the new unified API is an original contribution.
+However, this is not correct.
+
+While the underlying conceptual algorithms were already published in two papers
+in the \emph{Journal of Computational and Graphical Statistics}, the software
+implementation was not yet published in a journal. Originally, we submitted the
+\emph{party} package to the MLOSS track of the JMLR as JMLR-07-240 which was rejected
+mainly because the reviewers felt (correctly, as we are convinced now) that the design and
+implementation was not general enough. This prompted a reimplementation in \emph{partykit},
+essentially from scratch, which was then resubmitted as JMLR-14-133.
+
+Therefore, the entire package along with its extensive documentation should be
+seen as an original contribution. It supersedes the \emph{party} implementation
+in all aspects with only one notable exception (\texttt{cforest}, for very good
+reasons as explained below).
+
+\smallskip
 
 \textbf{\textit{%
-Reviewer 2 from the last review cycle comments that the cforest()
+I am not convinced that the breadth of this contribution is sufficient, given the much larger breadth of
+contributions to JMLR-MLOSS such as Shogun, scikit-learn, Shark, mlpack, dlib-ml, and other large-scale toolkits.
+}}
+
+\smallskip
+
+This depends on the dimensions in which ``breadth'' is assessed. The above-mentioned
+toolkits provide a large set of methods for binary and multiclass classification, regression
+and clustering with a focus on SVMs.  \emph{Shark} and \emph{mlpack} also
+contain implementations of tree algorithms for these situations.  Unlike
+\emph{partykit}, none of the toolkits provides dedicated methods for targets
+with other structures, e.g., censored data (including survival analysis), ordered
+categorical regression (e.g., proportional odds models), item response theory (IRT),
+paired comparisons, multivariate data with mixed types (e.g., several endpoints in a clinical study
+with mixed numeric/categorical/censored measurement scales). Through the use
+of nonparametric transformations (\texttt{ctree}) or parametric statistical models (\texttt{mob})
+such cases can be addressed in \emph{partykit}.
+
+For example, this has enabled ourselves to successfully develop dedicated psychometric
+tree models in package \emph{psychotree} for IRT and paired comparison data -- but it also
+has enabled the authors of the \emph{quint} package to establish trees for detecting
+treatment-subgroup interactions in clinical data. For a third very recent example
+(package \emph{vcrpart}) see the comments below.
+
+\smallskip
+
+\textbf{\textit{%
+Reviewer 2 from the last review cycle comments that the \texttt{cforest()}
 functionality from party is not available in partykit, and today this is
@@
 given that the clean API allows easy implementation of other algorithms.
-Increasing the breadth of the package should also increase its visibility,
-usage, and popularity.
+}}
+
+\smallskip
+
+As pointed out in our response to Reviewer~2, the mere representation of the
+forest (= list of trees plus data and metainformation) is straightforward.
+The real challenge is to provide a \emph{unified and extensible} collection of methods
+for computations on these forests (as available for single trees): e.g.,
+aggregation of predictions across trees, assessement of variable importance,
+visualization of partial dependence. While Breiman's (and Cutler's) original
+proposal (and software) tackled these problems to some degree, their solutions
+have been either shown to perform poorly in certain situations or are not
+general enough in others.
+
+For example, the simple majority vote or mean aggregation
+might be less attractive compared to the understanding of forests as
+estimators of conditional distribution functions \citep{Meinshausen:2006},
+especially when applying them to more complex data (censored, ordered, IRT, \dots).
+Or as another example, the originally proposed variable importance measures are
+artificially biased when applied to predictors with different scales
+\citep{Strobl+Boulesteix+Kneib:2008} compromising their suitability in applied
+work (e.g., in bioinformatics).
+
+Of course, we can simply move the existing \texttt{cforest} code from \emph{party}
+to \emph{partykit} but this would not solve the problems discussed above. These
+deserve further thought and work, both conceptually and computationally.
+
+\smallskip
+
+\textbf{\textit{%
+Increasing the breadth of the package should also increase its visibility, usage, and popularity.
 Overall, I do not want to give the authors the idea that I think their
@@
 
+\smallskip
+
+We are confident that the current \emph{partykit} infrastructure and API
+will motivate other researchers to implement novel tree-based procedures that
+in turn will increase the breadth of recursive partitioning methods available
+in \textsf{R}.
+
+One example for such a ``success story'' that emerged while our revised manuscript was under review
+is the \emph{vcrpart} package (\texttt{http://CRAN.R-project.org/package=vcrpart}).
+This is a completely new package implementing varying coefficient trees for
+ordinal 2-stage linear mixed models based on \emph{partykit} infrastructure.
+By the way, apart from a few small questions about implementation
+details, the \emph{vcrpart} authors apparently did not need to be in contact with us and have
+been able to leverage the \emph{partykit} infrastructure based on its documentation.
+Moreover, the main author of \emph{vcrpart} submitted a few bug reports
+about smaller glitches in \emph{partykit} by e-mail (not the forum) that we have
+subsequently fixed.
+
+Novel procedures such as the one implemented in \emph{vcrpart} are exactly the
+notion of ``breadth'' we had in mind when designing \emph{partykit} and that
+the package aims to facilitate. We expect other research groups to pick up the
+\emph{partykit} infrastructure in the future, especially if it is given increased
+visibility in the MLOSS track of JMLR.
+
+\bibliographystyle{plainnat}
+\bibliography{../ref.bib}
+
 \end{document}

### 2014-10-22 pkg/partykit/inst/mloss/letter_mloss3.tex
added Strobl ref
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@629 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Referee 3 criticised the limited breath of our contribution and suggested to
+Reviewer 3 criticises the limited breath of our contribution and suggests to
 copy the \texttt{cforest} function from package \emph{party} to package
-\emph{partykit}, iterating an argument by Referee 2 from the previous
+\emph{partykit}, iterating an argument by Reviewer 2 from the previous
 round. In the rebuttal letter, we explain the reason why it is currently not

### 2014-10-23 pkg/partykit/inst/mloss/mloss2.Rnw
slightly tweaked references to fit on a single page
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@604 255f4249-2263-4417-b1ea-8498779e27a7
A bit of squeezing to optionally include the Acknowledgments on the first four pages:
- If Figure 1 is placed on page 3, then only the references are on page 5.
- If Figure 1 is placed on page 4, the acknowledgments end up on page 5
  as well (but the overall number of pages is not changed).



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@630 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Not only trees learned within \proglang{R} can be transformed to the
-proposed infrastructure but also trees from other software packages.  Either
-a dedicated interface has to be created using the building blocks described
-in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or PMML
-(Predictive Model Markup Language) can be used as an intermediate exchange
-format.  This is an XML standard created by an international consortium
-\citep{DMG:2014} that includes a \code{<TreeModel>} tag with support for
-constant-fit classification and regression trees.  The function
-\fct{pmmlTreeModel} allows to read these files and represents them as
-\class{simpleparty} objects in \pkg{partykit}.  The reason for not using the
-\class{constparty} class as above is that the PMML format only stores point
-predictions (e.g., a mean or proportion) rather than all observations
-from the learning sample.  So far, the PMML interface has
-been tested with output from the \proglang{R} package \pkg{pmml} and
-\proglang{SPSS}'s \pkg{AnswerTree} model.  The latter includes an
-implementation of the QUEST algorithm \citep{Loh+Shih:1997}.
-
-Finally, the \pkg{partykit} function \fct{mob} implements model-based
-recursive partitioning (MOB) along with ``mobster'' interfaces for certain
-models (e.g., \fct{lmtree}, \fct{glmtree}). These return objects of class
-\class{modelparty} where nodes are associated with statistical models (as
-opposed to simple constant fits). In principle, this may also be adapted
-to other model trees (such as GUIDE, LMT, or M5') but no such interface
-is currently available.
-
-All of these different trees (see Table~\ref{tab:overview} for an overview)
-use the same infrastructure at the core but possibly with different options
-enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
-can be used to create textual and graphical displays of the tree and for
-computing predictions on new data, respectively. As an example for the
-visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
-to the well-known data on survival of passengers on the ill-fated maiden
-voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
-learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
-shows a MOB tree learned with \pkg{partykit} with a logistic regression
-for treatment effects in the terminal nodes. Additionally, the
-are further utility functions, e.g.,
-\fct{nodeapply} can be employed to access further information
-stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.
-
 <<Titanic>>=
@@
 
+Not only trees learned within \proglang{R} can be transformed to the
+proposed infrastructure but also trees from other software packages.  Either
+a dedicated interface has to be created using the building blocks described
+in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or PMML
+(Predictive Model Markup Language) can be used as an intermediate exchange
+format.  This is an XML standard created by an international consortium
+\citep{DMG:2014} that includes a \code{<TreeModel>} tag with support for
+constant-fit classification and regression trees.  The function
+\fct{pmmlTreeModel} allows to read these files and represents them as
+\class{simpleparty} objects in \pkg{partykit}.  The reason for not using the
+\class{constparty} class as above is that the PMML format only stores point
+predictions (e.g., a mean or proportion) rather than all observations
+from the learning sample.  So far, the PMML interface has
+been tested with output from the \proglang{R} package \pkg{pmml} and
+\proglang{SPSS}'s \pkg{AnswerTree} model.  The latter includes an
+implementation of the QUEST algorithm \citep{Loh+Shih:1997}.
+
+Finally, the \pkg{partykit} function \fct{mob} implements model-based
+recursive partitioning (MOB) along with ``mobster'' interfaces for certain
+models (e.g., \fct{lmtree}, \fct{glmtree}). These return objects of class
+\class{modelparty} where nodes are associated with statistical models (as
+opposed to simple constant fits). In principle, this may also be adapted
+to other model trees (such as GUIDE, LMT, or M5') but no such interface
+is currently available.
+
+All of these different trees (see Table~\ref{tab:overview} for an overview)
+use the same infrastructure at the core but possibly with different options
+enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
+can be used to create textual and graphical displays of the tree and for
+computing predictions on new data, respectively. As an example for the
+visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
+to the well-known data on survival of passengers on the ill-fated maiden
+voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
+learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
+shows a MOB tree learned with \pkg{partykit} with a logistic regression
+for treatment effects in the terminal nodes. Additionally, the
+are further utility functions, e.g.,
+\fct{nodeapply} can be employed to access further information
+stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.
+
 \section{Developer Infrastructure}
@@
 \begin{description}
-  \item[\class{partysplit}:] Split specification with an integer ID for the
-    splitting variable, breakpoint(s), and indexes for the kids.
-  \item[\class{partynode}:] Node specification with an integer ID, a
+  \item[\class{partysplit}:] Split with integer ID for the
+    splitting variable, breakpoint(s), indexes for the kids.
+  \item[\class{partynode}:] Node specification with integer ID, a
     \class{partysplit}, and a list of kids (if any) that are \class{partynode}
     objects again.
-  \item[\class{party}:] Tree specification with a \class{partynode} tree
-    along with a \class{data.frame} (which optionally may be empty),
-    potentially plus further information about fitted values and so-called
-    \class{terms} that allow to preprocess new data for predictions.
+  \item[\class{party}:] Tree with a recursive \class{partynode}
+    and a \class{data.frame} (optionally empty),
+    potentially plus information about fitted values and
+    \class{terms} allowing to preprocess new data for predictions.
 \end{description}
@@
 
-\bibliography{../../vignettes/party.bib}
+\bibliography{ref.bib}
 

### 2014-10-28 pkg/partykit/inst/mloss/review3/p2p.tex
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@631 255f4249-2263-4417-b1ea-8498779e27a7

@@
 There are still some minor issues that should be fixed in the final version, which will
-be sent out to Reviewer 3 for a quick confirmation.
-As pointed out by Reviewer 3, partykit only provides a subset of the functionality of its
+be sent out to Reviewer~3 for a quick confirmation.
+As pointed out by Reviewer~3, partykit only provides a subset of the functionality of its
 earlier cousin party. Porting the old functionality of party to the new
@@
 Of course, we can easily copy the code of the old \texttt{party::cforest} into the
-new \emph{partykit} package. However, we fail to see how this would increase usefulness
-for practitioners given that \emph{party} is easily available. More importantly, this
-would conflict with the view of Reviewer~2 who apparently agreed that more future
-work should be devoted to this important topic.
+new \emph{partykit} package. However, the usefulness for the practitioner would not
+really be increased by this (given that \emph{party} is easily available). More
+importantly, if we take the recommendations of Reviewer~2 seriously (which we do),
+then we should not just copy existing code but build new extensible infrastructure
+for forests as well. If we only copy the old code now, future developments are likely
+to change at least parts of the interface and thus potentially breaking full
+backward compatibility. Therefore, we would much rather follow Reviewer~2 and devote
+more future work to this important topic.
 
@@
 seen as an original contribution. It supersedes the \emph{party} implementation
-in all aspects with only one notable exception (\texttt{cforest}, for very good
-reasons as explained below).
+in all aspects with only one notable exception (the \texttt{cforest} function,
+for very good reasons as explained below).
 
@@
 toolkits provide a large set of methods for binary and multiclass classification, regression
-and clustering with a focus on SVMs.  \emph{Shark} and \emph{mlpack} also
+and clustering with a focus on SVMs. Some of the packages above, e.g., \emph{Shark} and \emph{mlpack}, also
 contain implementations of tree algorithms for these situations.  Unlike
 \emph{partykit}, none of the toolkits provides dedicated methods for targets
-with other structures, e.g., censored data (including survival analysis), ordered
+with structures other than binary/multiclass/numeric. For example: censored data (including survival analysis), ordered
 categorical regression (e.g., proportional odds models), item response theory (IRT),
@@
 of nonparametric transformations (\texttt{ctree}) or parametric statistical models (\texttt{mob})
-such cases can be addressed in \emph{partykit}.
+all of the cases above can be addressed in \emph{partykit}.
 
@@
 One example for such a ``success story'' that emerged while our revised manuscript was under review
-is the \emph{vcrpart} package (\texttt{http://CRAN.R-project.org/package=vcrpart}).
+is the \emph{vcrpart} package (published on \texttt{http://CRAN.R-project.org/package=vcrpart}
+in September).
 This is a completely new package implementing varying coefficient trees for

### 2014-10-29 pkg/partykit/inst/mloss/review3/p2p.tex
typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@632 255f4249-2263-4417-b1ea-8498779e27a7

@@
 for forests as well. If we only copy the old code now, future developments are likely
-to change at least parts of the interface and thus potentially breaking full
+to change at least parts of the interface and thus potentially break full
 backward compatibility. Therefore, we would much rather follow Reviewer~2 and devote

### 2014-11-26 www_src_partykit/03_getting_started.Rmd
add citations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@638 255f4249-2263-4417-b1ea-8498779e27a7

---
layout: page
title: Getting Started
permalink: /getting_started/
---

Your first regression tree
========================================================

First, install and fire-up R on your computer. Within R, one needs to
install the party package by typing

```{r eval=FALSE}
install.packages("partykit")
```

and hitting the ENTER key. Once the package is installed, you can
load it using

```{r results: hide}
library("partykit")
```

Now all party functions are ready to be used, for example the ctree()
function for fitting a regression tree to the Ozone data (after removing observations
with missing response):

```{r}
### regression
airq <- subset(airquality, !is.na(Ozone))
airct <- ctree(Ozone ~ ., data = airq,   
               control = ctree_control(maxsurrogate = 3))
airct
```

The tree is represented by an object called airct which can be plotted

```{r fig.width=9, fig.height=6}
plot(airct)
```

or used for computing predictions

```{r}
summary(predict(airct))
```

which can be compared to the actual response values:

```{r fig.width=6, fig.height=6}
plot(airq$Ozone, predict(airct))
abline(a = 0, b = 1)
```

### 2014-11-26 www_src_partykit/04_publications.Rmd
add citations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@638 255f4249-2263-4417-b1ea-8498779e27a7

---
layout: page
title: Publications
permalink: /publications/
---

Scholary papers describing the methodology
========================================================

```{r setup, include = FALSE, cache = FALSE}
library("RefManageR")
bib <- ReadBib("TH.bib", check = FALSE)
BibOptions(check.entries = FALSE, style = "html", cite.style = "authoryear",
           bib.style = "numeric")
```
The partykit package implements conditional inference trees
`r AutoCite(bib, "Hothorn:2006:JCGS", before = "function ctree(), ")`,
and model-based recursive partitioning 
`r AutoCite(bib, "Zeileis+Hothorn+Hornik:2008", before = "function mob(), ")`.

**References**

```{r results = "asis", echo = FALSE}
PrintBibliography(bib, .opts = list(check.entries = FALSE, sorting = "ynt"))
```

### 2014-11-26 www_src_partykit/config.R

@@ -1,0 +1,7 @@
+
+pkg <- "partykit"
+dest <- "html"
+publish <- "../www/partykit"
+
+download.file("http://user.math.uzh.ch/hothorn/TH.bib", dest = "TH.bib")
+bib <- "TH.bib"

### 2014-11-26 www_src_partykit/publish.R

@@ -1,0 +1,16 @@
+
+source("config.R")
+
+x <- readLines("_config.yml")
+x[grep("^baseurl", x)] <- paste("baseurl: http://", pkg, ".r-forge.r-project.org/partykit", sep = "")
+
+writeLines(x, file.path(dest, "_config.yml"))
+
+wd <- setwd(dest)
+
+system("jekyll build")
+
+setwd(wd)
+
+system(paste("cp -ra", file.path(dest, "_site/*"), publish))
+

### 2014-11-26 www_src_partykit/setup.R

@@ -1,0 +1,47 @@
+
+library("pkg2html")
+library("markdown")
+
+source("config.R")
+
+if (!file.exists(dest))
+    dir.create(dest)
+
+stopifnot(file.exists(dest)) 
+
+template <- system.file("template", package = "pkg2html")
+
+system(paste("cp -ra", file.path(template, "*"), dest, sep = " "))
+
+wd <- setwd(file.path(dest, "_data"))
+R2yaml(pkg)
+writeLines(bib2yaml(file.path(wd, bib), 
+           c("Hothorn:2006:JCGS", "Zeileis+Hothorn+Hornik:2008")),
+           con = "cites.yml")
+
+setwd(wd)
+setwd(file.path(dest, "_posts"))
+NEWS2md(pkg)
+
+setwd(wd)
+
+Rmd <- list.files(pattern = "Rmd$")
+
+for (f in Rmd)
+    writeLines(Rmd2html(f), con = file.path(dest, gsub("Rmd$", "html", f)))
+
+x <- readLines(file.path(dest, "_data", "pkg.yml"))
+x <- x[x != ""]
+x <- c(x, paste("headpic: /img/", pkg, ".png", sep = ""))
+writeLines(x, con = file.path(dest, "_data", "pkg.yml"))
+
+file.copy("_config.yml", dest, overwrite = TRUE)
+
+yml <- list.files(pattern = "yml$")
+yml <- yml[-grep("^_", yml)]
+sapply(yml, function(f) file.copy(f, file.path(dest, "_data"), overwrite = TRUE))
+
+system(paste("cat site.css >> ", file.path(dest, "css", "main.css")))
+
+system(paste("cp ", file.path(publish, "img", "*"), file.path(dest, "img")))
+print(paste("cp ", file.path(publish, "img", "*"), file.path(dest, "img")))

### 2014-11-26 pkg/partykit/inst/mloss/letter_mloss3.tex
add URL
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@641 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-10-22 pkg/partykit/inst/mloss/letter_mloss3.tex
@@
 
-I uploaded a rebuttal letter for our manuscript
+I uploaded a revised version of our manuscript
 ``\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in R'' to the
@@
 
-Referee 3 criticised the limited breath of our contribution and suggested to
-copy the \texttt{cforest} function from package \emph{party} to package
-\emph{partykit}, iterating an argument by Referee 2 from the previous
-round. In the rebuttal letter, we explain the reason why it is currently not
-helpful to move this function from one CRAN package to another one. In
-addition, we took the liberty to discuss why we think that \emph{partykit}
-is actually much more general than \emph{party} and, if one adopts a certain
-point of view, also more general than most of the established machine
-learning toolboxes. Since no changes in the manuscript were requested, we
-did not touch the four page description.
+Reviewer~3 suggested to add forest functionality to the \emph{partykit}
+package to increase it breadth.
+This suggestion was already brought up in the previous revision by
+Reviewer~2 and answered to the satisfaction of the first two reviewers:
+While the basic idea for setting up a generic class for ``random forests''
+is straightforward, many of the details in the implementation are not
+obvious at all and deserve much more attention.  
+Of course, we could easily copy the code of the old \texttt{party::cforest}
+into the new \emph{partykit} package but this would not increase the
+usefulness for the practitioner because \emph{party} is easily available. We
+therefore decided not to follow the suggestion by reviewer~3 for the same
+reasons as explained in the previous round and, in more detail now, the
+rebuttal letter.
 
-Sincerely yours, \\
+We revised the manuscript following the Instructions for Authors closely and
+will send the copyright forms upon acceptance.
+
+Best regards, \\
 \includegraphics[width = .25\textwidth]{TH_Unterschrift}

### 2014-11-28 pkg/partykit/inst/mloss/review4/p2p.tex
4th round
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@643 255f4249-2263-4417-b1ea-8498779e27a7

The paper itself is fine.  Last review I had no further comments on the paper itself.  My review here then is restricted to the point-by-point reply.  This "discussion" is going downhill quickly and I will thus keep my comments short.

The authors have pointed out some measures of project activity, which are fine.  Of my concerns, the activity level and lack of open mailing list are less of a big deal, and my review stated as much.

The remaining sticking point is this:

"I am not convinced that the breadth of this contribution is sufficient." (from the previous review)

The author's replies have not convinced me otherwise despite their extensive length.  My suggestion was that support for forests be added, or some other support be added to increase the utility or breadth of partykit.  The responses given to this idea by the authors are concluded with remarks like the following:

"These [ideas] deserve further thought and work, both conceptually and computationally."

"Therefore, we would much rather follow Reviewer 2 and devote more future work to this important topic."

"This is on our to-do list, though."

If these things are on the authors' to-do lists, then they should do them and resubmit when they are done.

Personally I am not willing to consider acceptance as-is with the assumption that these features will be added later because the authors have stated elsewhere that partykit is mature and now only requires maintenance.

The package would be stronger and relevant to a wider audience, even if the added support was only preliminary.  And according to the authors themselves, this would be a significant contribution because flexible support for forests is a "real challenge".

### 2015-01-05 pkg/partykit/R/cforest.R
improve parallel predictions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@670 255f4249-2263-4417-b1ea-8498779e27a7

@@ -170,28 +170,23 @@ predict.cforest <- function(object, newdata = NULL, type = c("response", "prob",
     ### extract weights
     rw <- object$weights
 
-    w <- matrix(0L, nrow = NROW(responses), ncol = length(nam))
+    # w <- matrix(0L, nrow = NROW(responses), ncol = length(nam))
 
     applyfun <- lapply
     if (!is.null(object$info))
         applyfun <- object$info$control$applyfun
 
-    bids <- applyfun(1:length(forest), function(b) {
+    bw <- applyfun(1:length(forest), function(b) {
         ids <- nodeids(forest[[b]], terminal = TRUE)
         fnewdata <- fitted_node(forest[[b]], nd, vmatch = vmatch, ...)
         fdata <- fitted_node(forest[[b]], object$data, ...)
-        list(ids = ids, fnewdata = fnewdata, fdata = fdata)
-    })
-
-    for (b in 1:length(forest)) {
-        ids <- bids[[b]]$ids
-        fnewdata <- bids[[b]]$fnewdata
-        fdata <- bids[[b]]$fdata
         tw <- rw[[b]]
         if (OOB) tw <- as.integer(tw == 0)
         pw <- sapply(ids, function(i) tw * (fdata == i))
-        w <- w + pw[, match(fnewdata, ids)]
-    }
+        return(pw[, match(fnewdata, ids)])
+    })
+
+    w <- Reduce("+", bw)
 
     if (type == "weights") {
         ret <- w

### 2015-01-05 pkg/partykit/inst/mloss/review4/p2p.tex
start rebuttal
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@672 255f4249-2263-4417-b1ea-8498779e27a7

@@
-The paper itself is fine.
-Last review I had no further comments on the paper itself.
-My review here then is restricted to the point-by-point reply.
-This "discussion" is going downhill quickly and I will thus keep my comments short.
+\documentclass{article}
+\usepackage{a4wide}
+\usepackage[authoryear,round]{natbib}
 
-The authors have pointed out some measures of project activity, which are fine.  Of my concerns, the activity level and lack of open mailing list are less of a big deal, and my review stated as much.
+\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
+\setlength{\parindent}{0em}
 
+\title{JMLR-14-133-4: Point-to-Point Reply}
+\author{Torsten Hothorn and Achim Zeileis}
+
+\begin{document}
+
+\maketitle
+
+\section*{AE}
+
+\textbf{\textit{%
+I am concerned that your new revision is essentially a rebuttal of the
+request of the reviewer about adding random forest functionality (which
+existed in the previous version) into the current version.  As I said in my
+previous decision, porting functionality of party to partykit is important,
+and in my opinion a reasonable request.
+}}
+
+\smallskip
+
+We now provide a reimplementation of \texttt{party::cforest} in the new
+\textbf{partykit} framework.  From the reviews we were not entirely sure
+what your desired solution for the reimplementation was: We decided to write
+a new high-level implementation in R, demonstrating the breadth and
+usefulness of the \textbf{partykit} infrastructure also for ensemble
+learning.
+
+\section*{Referee 3}
+
+\textbf{\textit{%
+The paper itself is fine.  Last review I had no further comments on the
+paper itself.  My review here then is restricted to the point-by-point
+reply.  This "discussion" is going downhill quickly and I will thus keep my
+comments short.
+The authors have pointed out some measures of project activity, which are
+fine.  Of my concerns, the activity level and lack of open mailing list are
+less of a big deal, and my review stated as much.
 The remaining sticking point is this:
+"I am not convinced that the breadth of this contribution is sufficient." (from the previous review)
+The author's replies have not convinced me otherwise despite their extensive
+length.  My suggestion was that support for forests be added, or some other
+support be added to increase the utility or breadth of partykit.  The
+responses given to this idea by the authors are concluded with remarks like
+the following:
+\begin{itemize}
+\item "These [ideas] deserve further thought and work, both conceptually and computationally."
+\item "Therefore, we would much rather follow Reviewer 2 and devote more future work to this important topic."
+\item "This is on our to-do list, though."
+\end{itemize}
+If these things are on the authors' to-do lists, then they should do them
+and resubmit when they are done.}}
 
-"I am not convinced that the breadth of this contribution is sufficient." (from the previous review)
+\smallskip
 
-The author's replies have not convinced me otherwise despite their extensive length.
-My suggestion was that support for forests be added, or some other support be added to increase the utility or breadth of partykit.
-The responses given to this idea by the authors are concluded with remarks like the following:
+The new \texttt{partykit::cforest} shares the benefits of
+\texttt{party::cforest} over \texttt{randomForest::randomForest} (such as surrogate
+splits for dealing with missing values in the predictor variables or the
+ability of handling right-censored data) and even has a couple of
+advantages over the old \texttt{party::cforest}: First, it is very flexible
+and can exploit some features readily available in R, e.g., parallelization
+on multicore machines or clusters of workstations.  Second, it can be used
+as a blueprint for users/developers who wish to create custom forests with
+their desired tree learners.  However, as \texttt{partykit::cforest} is
+written entirely in R, it is slower than the old \texttt{party::cforest}
+when running on a single core only. Third, quantile regression forests
+(Meinshausen, JMLR 2006) are now part of the framework and even extensions
+to density estimation are available (see \texttt{?cforest} for an example).
 
-"These [ideas] deserve further thought and work, both conceptually and computationally."
+\textbf{\textit{%
+Personally I am not willing to consider acceptance as-is with the assumption
+that these features will be added later because the authors have stated
+elsewhere that partykit is mature and now only requires maintenance.
+The package would be stronger and relevant to a wider audience, even if the
+added support was only preliminary.  And according to the authors
+themselves, this would be a significant contribution because flexible
+support for forests is a "real challenge".
+}}
 
-"Therefore, we would much rather follow Reviewer 2 and devote more future work to this important topic."
+\smallskip
 
-"This is on our to-do list, though."
+The new \texttt{partykit::cforest} basically implements old functionality
+using the new framework. The code-base is much denser now and the two
+implementations were tested to return identical results (in situations where
+is can be expected). It also allows a little more flexibility with respect
+to probabilistic forecasts (via quantile regression forests) but is still
+limited to trees with constant fits in the nodes. Therefore, the challenge
+of adding flexible support for forests of model-based is still open and
+definitively a research project in its own right because also the underlying
+theory has to be developed.
 
-If these things are on the authors' to-do lists, then they should do them and resubmit when they are done.
-
-Personally I am not willing to consider acceptance as-is with the assumption that these features will be added later because the authors have stated elsewhere that partykit is mature and now only requires maintenance.
-
-The package would be stronger and relevant to a wider audience, even if the added support was only preliminary.
-And according to the authors themselves, this would be a significant contribution because flexible support for forests is a "real challenge".
+\end{document}

### 2015-01-05 pkg/partykit/inst/mloss/review4/p2p.tex
wrap-up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@673 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
+Certainly, much more work can be done on general software for random forests
+as well as improvements of the \texttt{cforest} function.  We intend to do
+so in the future by tackling some of the open conceptual issues described in
+our previous replies.  But, the current \texttt{partykit::cforest} function
+clearly shows that also implementations of random forests will benefit from
+the unified tree infrastructure established in the partykit package.
+
 \end{document}

### 2015-01-05 pkg/partykit/inst/mloss/review4/p2p.tex
small improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@674 255f4249-2263-4417-b1ea-8498779e27a7

@@
 The new \texttt{partykit::cforest} basically implements old functionality
-using the new framework. The code-base is much denser now and the two
+using the new framework. The code base is much denser now and the two
 implementations were tested to return identical results (in situations where
-is can be expected). It also allows a little more flexibility with respect
+this can be expected). It also allows a little more flexibility with respect
 to probabilistic forecasts (via quantile regression forests) but is still
 limited to trees with constant fits in the nodes. Therefore, the challenge
-of adding flexible support for forests of model-based is still open and
+of adding flexible support for forests of model-based trees is still open and
 definitively a research project in its own right because also the underlying
@@
 clearly shows that also implementations of random forests will benefit from
-the unified tree infrastructure established in the partykit package.
+the unified tree infrastructure established in the \textbf{partykit} package.
 

### 2015-01-06 pkg/partykit/inst/mloss/letter_mloss4.tex
add letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@678 255f4249-2263-4417-b1ea-8498779e27a7

--- 2014-10-22 pkg/partykit/inst/mloss/letter_mloss3.tex
@@
 
-\subject{JMLR-14-133-3 revision}
+\subject{JMLR-14-133-4 revision}
 
@@
 
-I uploaded a rebuttal letter for our manuscript
-``\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in R'' to the
-Machine Learning Open-Source Software track of the Journal of Machine
-Learning Research.
+I uploaded a revised version of the \textbf{partykit} package (version
+0.8-4, also the current CRAN version). Following the request by referee 3,
+we implemented \textbf{cforest()} in this version. The benefits and
+shortcomings of this implementation of random forests, compared to the
+implementations in the \textbf{randomForest} and \textbf{party} packages, are
+discussed in the rebuttal.
 
-Referee 3 criticised the limited breath of our contribution and suggested to
-copy the \texttt{cforest} function from package \emph{party} to package
-\emph{partykit}, iterating an argument by Referee 2 from the previous
-round. In the rebuttal letter, we explain the reason why it is currently not
-helpful to move this function from one CRAN package to another one. In
-addition, we took the liberty to discuss why we think that \emph{partykit}
-is actually much more general than \emph{party} and, if one adopts a certain
-point of view, also more general than most of the established machine
-learning toolboxes. Since no changes in the manuscript were requested, we
-did not touch the four page description.
+The manuscript was not changed and thus I did not upload a new version.
 
-Sincerely yours, \\
+Best regards, \\
 \includegraphics[width = .25\textwidth]{TH_Unterschrift}

### 2015-02-20 pkg/partykit/inst/mloss/mloss2.Rnw
added cforest comment in discussion, updated references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@687 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}
-\editor{Cheng Soon Ong (MLOSS track, JMLR-14-133)}
+\editor{Cheng Soon Ong}
 
@@
 fitting functions (\fct{mob}). The simple and lean base classes that separate
-data and tree structure are also appealing for storing forests which are planned
-for future versions of the package. Users and developers that have questions or
-comments about the package can either contact the maintainers or use the forum
+data and tree structure are also appealing for storing forests -- a first proof-of-concept
+reimplementation of \fct{cforest} is in the package with further extension planned.
+Users and developers that have questions or comments about the package can either
+contact the maintainers or use the forum
 on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/forum/forum.php?forum_id=852}.

### 2015-02-20 pkg/partykit/inst/mloss/mloss2.Rnw
more details in addr
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@688 255f4249-2263-4417-b1ea-8498779e27a7

@@
 %% header declarations
-\author{Torsten Hothorn \\ \addr{Universit\"at Z\"urich} \\ \addr{E-mail: Torsten.Hothorn@R-project.org} \AND
-        Achim Zeileis \\ \addr{Universit\"at Innsbruck} \\ \addr{E-mail: Achim.Zeileis@R-project.org}}
+\author{Torsten Hothorn \\ \addr{Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention, Universit\"at Z\"urich} \\ \email{E-mail: Torsten.Hothorn@R-project.org} \AND
+        Achim Zeileis \\ \addr{Department f\"ur Statistik, Universit\"at Innsbruck} \\ \email{E-mail: Achim.Zeileis@R-project.org}}
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}

### 2015-02-20 www_src_partykit/04_publications.Rmd
add MLOSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@693 255f4249-2263-4417-b1ea-8498779e27a7

@@
 ```
-The partykit package implements conditional inference trees
+The partykit package provides a flexible toolkit for learning, representing,
+summarizing, and visualizing a wide range of tree-structured regression
+and classification models `r AutoCite(bib, "Hothorn_Zeileis_2015")`.
+In addition, it implements conditional inference trees
 `r AutoCite(bib, "Hothorn:2006:JCGS", before = "function ctree(), ")`,

### 2015-02-21 pkg/partykit/inst/mloss/mloss2.Rnw
address
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@694 255f4249-2263-4417-b1ea-8498779e27a7

--- 2015-02-20 pkg/partykit/inst/mloss/mloss2.Rnw
@@
 %% header declarations
-\author{Torsten Hothorn \\ \addr{Universit\"at Z\"urich} \\ \addr{E-mail: Torsten.Hothorn@R-project.org} \AND
-        Achim Zeileis \\ \addr{Universit\"at Innsbruck} \\ \addr{E-mail: Achim.Zeileis@R-project.org}}
+\author{Torsten Hothorn \\ \addr{Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention, Universit\"at Z\"urich} \\ \email{E-mail: Torsten.Hothorn@R-project.org} \AND
+        Achim Zeileis \\ \addr{Institut f\"ur Statistik, Universit\"at Innsbruck} \\ \email{E-mail: Achim.Zeileis@R-project.org}}
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}

### 2015-02-21 www_src_partykit/04_publications.Rmd
1.0-0
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@695 255f4249-2263-4417-b1ea-8498779e27a7

@@
 summarizing, and visualizing a wide range of tree-structured regression
-and classification models `r AutoCite(bib, "Hothorn_Zeileis_2015")`.
+and classification models `r Citep(bib, "Hothorn_Zeileis_2015")`.
 In addition, it implements conditional inference trees

### 2015-03-12 pkg/glmertree/R/glmertree.R
took Marjolein's code for lmertree() and glmertree() and turned it into a rudimentary package so that it can be easily installed
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@701 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,104 @@
+lmertree <- function(lmtreeformula, randomformula, data, #subset = NULL, 
+  initialRandomEffects = NULL, 
+  ErrorTolerance = 0.001, MaxIterations = 1000, 
+  verbose = TRUE, plotting = TRUE,
+  lmer.control = lmerControl(), ...)
+{
+  iteration <- 0L
+  if (!is.null(initialRandomEffects)) { data$offset <- initialRandomEffects }
+  if (is.null(initialRandomEffects)) { data$offset <- rep(0, times=dim(data)[1L]) }
+  continuecondition <- TRUE
+  oldloglik <- -Inf
+  while (continuecondition) {
+    iteration <- iteration + 1L
+    tree <- lmtree(lmtreeformula, data, offset = offset, ...)
+    if(verbose){plot(tree)}
+    if(length(tree) > 1L) {
+      data$treeresponse <- treeresponse <- predict(tree, newdata = data)
+    }
+    if(length(tree) == 1L) { # if treedepth=1, predict.lmtree doesn't work, then use unpartitioned lm
+      lmformula <- formula(as.Formula(lmtreeformula), lhs = 1L, rhs = -2L)
+      data$treeresponse <- treeresponse <- predict(lm(lmformula, data = data), newdata = data)
+    }
+    lme <- lmer(randomformula, data = data, offset = treeresponse)
+    data$offset <- predict(lme, newdata = data)
+    newloglik <- logLik(lme)    
+    continuecondition <- (newloglik - oldloglik > ErrorTolerance) & (iteration < MaxIterations) 
+    oldloglik <- newloglik
+    if(verbose) {
+      print(newloglik)
+    }
+  } 
+  result <- list(
+    Tree = tree,
+    RandomEffectModel = lme,
+    RandomEffects = ranef(lme), 
+    BetweenMatrix = VarCorr(lme),
+    ErrorVariance = attr(VarCorr(lme),"sc")^2, 
+    data = data,
+    logLik = newloglik,
+    IterationsUsed = iteration, 
+    MaxIterations=MaxIterations,
+    initialRandomEffects = initialRandomEffects, 
+    TreeFormula = lmtreeformula,
+    RandomFormula = randomformula,
+    #Subset = subs, 
+    ErrorTolerance = ErrorTolerance,
+    #residuals = residuals, 
+    tree.control = list(...),
+    lmer.control = lmer.control
+  )
+}
+
[…]

### 2015-03-12 pkg/glmertree/inst/paper/paper.tex
inst folder, currently only with 'paper' folder all in .Rbuildignore
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@702 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,man]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{tikz}
\usepackage{lineno}
\pagewiselinenumbers

\graphicspath{{figures/}}

\usetikzlibrary{shapes,shadows,arrows,positioning}

\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets: Combining Model-Based Recursive Partitioning and Random-Effects Estimation}
\shorttitle{Detection of treatment-subgroup interactions in clustered datasets}
\author{M. Fokkema, N. Smits, H. Kelderman}
\affiliation{Vrije Universiteit, Amsterdam}
\abstract{
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers or persons. In the current paper we propose a new algorithm, called clusterMOB, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The clusterMOB algorithm uses model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) to detect treatment-subgroup interactions, and a linear mixed-effects model \cite{PinhyBate00} for estimation of random-effects parameters. In a simulation study, we evaluate the performance of clusterMOB, in terms of recovery of treatment-subgroup interactions and prediction of treatment differences, and compare it with that of MOB without random effects. The clusterMOB algorithm recovered the true treatment-subgroup interactions in 91.2\% of the simulated datasets, whereas MOB without random effects recovered the true interactions in 63.5\% of the simulated datasets. Also, clusterMOB predicted treatment outcome differences with an average accuracy of .94, whereas MOB predicted treatment differences with an average accuracy of .88. We conclude that clusterMOB is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss directions for future research.\\
}
%\acknowledgements{}
\rightheader{Detection of treatment-subgroup interactions in clustered datasets}
\leftheader{Detection of treatment-subgroup interactions in clustered datasets}

\begin{document}

\maketitle

\pagebreak

\section{Introduction}

In medicine-efficacy research, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Personalized medicine presents us with the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypothesis about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to detect treatment-subgroup interactions: they can handle many potential predictor variables at once, and can automatically detect (higher order) interactions between predictor variables. Several promising tree-based algorithms and software packages have been developed to assist in the detection of treatment-subgroup interactions (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) may be the most flexible method for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. Earlier, MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

Single randomized clinical trails may often be underpowered to detect treatment-subgroup interactions. Therefore, meta-analysis of individual-level patient data (IPD), in which datasets from several RCTs are pooled, is becoming increasingly popular \cite{KoopyHeij07}. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}. Likewise, analyses of datasets from single clinical trials conducted in multiple research centers, and longitudinal datasets may also require modeling of random effects.

However, none of the aforementioned tree-based algorithms allow for estimation of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92}. More importantly, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups.

In the current paper, we present a tree-based algorithm for treatment-subgroup interaction detection, which takes the clustered nature of datasets into account. The algorithm combines MOB with the estimation of random effects, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials).

In what follows, we will first discuss existing frameworks for estimating treatment effects: the linear fixed-effects model, model-based recursive partitioning of the linear fixed-effects model, and the linear mixed-effects model. Then, we present a new algorithm, which combines model-based recursive partitioning and random-effects estimation. In the methods and results section, we present a simulation study evaluating the comparative accuracy of the algorithm.

\section{General modeling framework}

\subsection{Linear fixed-effects model}

In a clinical trial, where the outcomes of two treatments are compared, an overall linear regression model may be used to estimate treatment effects. Let $N$ be the total number of observations in the dataset\footnote{In the Appendix, an overview of notation used in the current paper is provided}. The following linear regression model can be used to estimate the effects of the treatments:

\begin{equation}
\label{eq:fixedeffects}
    y = X \beta + \epsilon
\end{equation}

Where $y$ is an $N \times 1$ column vector of outcome variable values, $X$ is the $N \times 2$ matrix of predictor variable values (the first column consisting of a vector of ones, and the second column consisting of a dummy variable indicating treatment type); $\beta$ is the $2 \times 1$ column vector of regression coefficients (the first element of $\beta$ representing the intercept, or mean value of the outcome variable for treatment 1, and the second element of $\beta$ representing the difference in mean values of the outcome variables for treatments 1 and 2). $\epsilon$ is an $N \times 1$ column vector of residuals, which is assumed to be normally distributed with mean 0 and variance $\sigma^{2}_{\epsilon}$, and independent of all other variables in the model.

In Figure \ref{fig:fixed_effects}, a graphical representation of the model in Equation \ref{eq:fixedeffects}, estimated for a simulated dataset of 150 observations, is provided. The boxplots show the distribution of the outcome variable (posttreatment depression score) among 150 participants, who were randomly assigned to treatment 1 and treatment 2. Figure \ref{fig:fixed_effects} suggests little overall difference between the outcomes of treatment 1 and treatment 2, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by other variables. Conditional on other variables, the relationship between $X$ and $Y$ may be stronger (Simpson's paradox; \citeNP{Simp51}).

\subsection{Model-based recursive partitioning}

The rationale behind MOB is that a global model for all observations, like that in Equation \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates $U$ are available, it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, of which the distributions of the outcome variable are most different. Whereas CART trees have constant fits in the terminal nodes, MOB trees have parametric models with one or more predictor variables in the terminal nodes.

To find partitions and better-fitting local linear models, the MOB algorithm tests for parameter instability: differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more variables in $U$. To this end, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model $y=X \beta + \epsilon$ to the data set, (2) test for parameter instability over a set of partitioning variables $U_1$ through $U_r$, (3) if there is some overall parameter instability, split the data set with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subsamples.

More specifically, in step (2), to test for parameter instability, so-called \textit{scores} are computed, using the score function. The expected value of the scores over all observations in a data set is zero, by definition. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the value of a potential partitioning variable $U_k$. To statistically test whether there are systematic deviations of the scores from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used in MOB \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the data set is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). This split point for $U_{k*}$ is selected, that minimizes the sum of the RSSs in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partitions can be represented as a tree, with a local fixed effects linear regression model $y = X \beta + \epsilon$ in all $j = 1,...,t$ terminal nodes of the tree. As a result, the model in Equation \ref{eq:fixedeffects} keeps the same formulation in each endnode, yielding a $\beta$ that is a $2t \times 1$ column vector ${\left[ {\beta_1}^T,...,{\beta_j}^T,...,{\beta_t}^T \right]}^T$, where each $\beta_j$ relates to a subsample of the observations in terminal node $j$. $X$ becomes an $N \times 2t$ (block diagonal) design matrix, with the $N_{j} \times 2$ matrices of predictor variable values in terminal node $j$ on the diagonal. Note that when the null hypothesis of parameter stability can not be rejected in the first iteration, no partitions are created by MOB, and the resulting tree has a single node with the dimensions of $X$ remaining $N \times 2$ and the dimensions of $\beta$ remaining $2 \times 1$.

Figure \ref{fig:example_mobtree} provides an example of a graphical representation of a model-based partition, based on the same simulated data of Figure \ref{fig:fixed_effects}. By using several additional covariates (i.e., anxiety questionnaire score, duration of depressive symptoms at baseline, age), MOB separated the observations into four partitions, each with a different estimate for $\beta$ (i.e., different mean values for treatment 1 and/or 2). Figure \ref{fig:example_mobtree} shows a subgroup with low duration and low anxiety, for which treatment 1 is much more beneficial (i.e., lower posttreatment depression scores); a subgroup of patients with low duration and anxiety $\leq 12$, for whom treatment 1 is somewhat more beneficial; a subgroup of patients with low duration and anxiety $> 12$, for whom treatment 2 is slightly more beneficial; and a subgroup of patients with duration of depressive symptoms $> 6$ months, for whom treatment 2 is much more beneficial. Age did not have a main or moderating effect on treatment outcomes, and therefore does not appear as a splitting variable in the tree.

\subsection{Linear mixed-effects models}

When the dataset consists of observations from multiple clusters, the model in Equation \ref{eq:fixedeffects} can be extended to include cluster-specific, or random, effects. With the inclusion of random effects, the model becomes:

\begin{equation}
\label{eq:mixedeffects}
	y = X \beta + Z b + \epsilon
\end{equation}

where $X$ is the $N \times 2$ matrix of fixed-effects predictor variable values, of which the first column is a vector of ones, and the second column is a dummy indicator for treatment type. $\beta$ is the $2 \times 1$ column vector of fixed-effects coefficients. $Z$ is the $N \times n$ (block diagonal) design matrix of random-effects predictor variables, with $n$ being the number of clusters. $Z$ has $N_i \times 1$ vectors of ones on the diagonal, with $N_i$ being the number of observations in cluster $i$; $b$ is the $n \times 1$ column vector ${\left[ {b_1},...,{b_i},...,{b_n} \right]}^{T}$ of random-effects coefficients, and it is assumed that that $b_{i}$ is normally distributed with mean zero and (scalar) variance $\sigma^{2}_{b_i}$ (note that in the current paper, for simplicity, we incorporate random intercepts only, but random slopes can easily be included in the model). $\epsilon$ is the $N \times 1$ column vector of residuals, which are normally distributed with mean zero and variance $\sigma^{2}_{\epsilon}$. Further, it is assumed that $b_{i}$ and $\epsilon$ are independent from all other variables in the model. The parameters of the linear mixed-effects model $\beta$,  $\sigma^{2}_{\epsilon}$ and $\sigma^{2}_{b_i}$ can be estimated with, for example, maximum likelihood (ML) and restricted ML \cite{LairyWare82}.

\subsection{Combining model-based recursive partitioning and random-effects estimation}

As noted earlier, ordinary linear fixed-effects models (Equation \ref{eq:fixedeffects}) are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but it does not allow for estimation of random effects. Therefore, we propose an extension of the MOB algorithm, which incorporates estimation of random effects.

For this extensions, we replace the fixed-effects part of Equation \ref{eq:mixedeffects} by a MOB tree, so $\beta$ becomes the $2t \times 2$ column vector ${\left[{\beta_1}^T,...,{\beta_j}^T,...,{\beta_t}^T \right]}^T$, and $X$ becomes the $N \times 2t$ (block diagonal) design matrix, with the $N_{j} \times 2$ matrices of predictor variable values in terminal node $j$ on the diagonal. This tree is estimated as described above, in the subsection \textit{Model-based recursive partitioning}. The random-effects coefficients $b$ in Equation \ref{eq:mixedeffects} are estimated globally, that is, the random-effects model is invariant over the nodes of the MOB tree.

To estimate $\beta$ and $b$ for this model, we take an iterative approach, alternating between assuming random effects $b$ known, allowing for estimation of $\beta = {\left[ {\beta_l}^T,...,{\beta_j}^T,...,{\beta_t}^T \right]}^T$; and assuming $\beta$ known, allowing for estimation of the random-effects vector $b$. The algorithm initializes by setting $b$ to zero, since both the random and the fixed-effects parts are initially unknown. In every iteration, the MOB tree (i.e., fixed effects $\beta$) and random effects $b$ are re-estimated: the MOB tree is estimated, given the estimated value of $b$ from the last iteration, and $b$ is estimated, given the estimated value of $\beta$ from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.

A similar approach has been taken by \citeA{HajjyBell11} and \citeA{SelaySimo12}, who added random-effects estimation to CART trees with constant fits, instead of linear models, in the terminal nodes. Because the proposed algorithm allows for model-based recursive partitioning of linear fixed-effects models, as well as estimation of cluster-specific random effects, we will refer to it as clusterMOB.

In what follows, we present a simulation study in which we assess the performance of clusterMOB in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments. Furthermore, we will compare the performance of clusterMOB with that of MOB. We will generate a large number of datasets, varying seven parameters: total sample size, number of clusters, variance of cluster-specific random intercepts, number of potential partitioning variables, intercorrelations between potential partitioning variables, (in)dependence between cluster-specific random intercepts and potential partitioning variables, and magnitude of the mean difference in treatment outcomes.

For clusterMOB, we expect the accuracy of recovered trees and predictions to improve with increasing sample size and/or magnitude of the difference in treatment outcomes. For MOB, we have the same expectations, when cluster-specific random effects are absent in the dataset. When cluster-specific random effects are present, we expect clusterMOB to perform better than MOB, in terms of accuracy of recovered trees and predictions, especially when the variance of cluster-specific random intercepts and the dependence between cluster-specific random intercepts and potential partitioning variables increases, and/or when the magnitude of the mean difference in treatment outcomes decreases. In the absence of cluster-specific random effects, we expect clusterMOB not to perform better than MOB.

\section{Method}

\subsection{Software}

\verb|R| \cite{R10} was used for generation and analysis of all datasets. Default settings were used for all functions, with exception of the maximum tree depth, which was set to four (i.e., eight terminal nodes) for both clusterMOB and MOB.

\subsubsection{Estimation of MOB and clusterMOB trees}

For estimating MOB trees, the \verb|lmtree| function from the \verb|R| package \verb|partykit| \cite{HothyZeil14,ZeilyHoth08} was used. The \verb|lmtree| function builds a linear model tree: a model-based recursive partition based on least squares regression (Equation \ref{eq:fixedeffects}).

For building clusterMOB trees, we build a function following the design of the \verb|REEMtree| function from the \verb|R| package \verb|REEMtree| \cite{SelaySimo11}. The \verb|clusterMOB| function is used to estimate the partitions and parameters of the MOB-tree extended model (Equation \ref{eq:mixedeffects}, with $\beta = {\left[{\beta_1}^T,...,{\beta_j}^T,...,{\beta_t}^T \right]}^T$, and $X$ is the $N \times 2t$ block diagonal design matrix of predictor variable values). The \verb|clusterMOB| function makes use of the \verb|lmtree| function, and the \verb|lme| function from the \verb|R| package \verb|nlme| (linear and non-linear mixed-effects models; \citeNP{PinhyBate00}).

The \verb|clusterMOB| function iterates through the following two steps: (1) building a linear model tree using the \verb|lmtree| function, and (2) calculating the random-effects coefficient vector $b$ using the \verb|lme| function. More specifically, in step (1) of the first iteration, the random effects are assumed to be zero. In each iteration $>1$, the random effects as estimated by the \verb|lme| function in step (2) of the last iteration are included as an offset (a variable to be included in the model, with a known, pre-specified coefficient of one) in the \verb|lmtree| function. In step (2), indicators for terminal node of the linear model tree, and the interaction between terminal node and treatment type are included as fixed-effects factors, to account for the fixed effects in estimation of the random-effects parameters. Convergence of \verb|clusterMOB| is monitored using the log-likelihood value of the linear mixed-effects model estimated in step (2). When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|clusterMOB| has converged.

\subsection{Evaluation of performance}

\subsubsection{Tree size}

For each dataset, we calculate the tree size of the MOB and clusterMOB trees, and compare it with the true tree size. To detect predictors of tree size for MOB and clusterMOB, a linear model tree was built using the \verb|lmtree| function. The outcome variable in this tree is the number of nodes of the generated trees. The predictor variable for the linear model is algorithm type (MOB or clusterMOB), and the (potential) partitioning variables are the parameters of the data-generating process, which will be described in the \textit{Simulation design} subsection. The maximum depth will be set to four, as this yields a visual representation which is easy to interpret.

\subsubsection{Splitting points and variables}

We compared the performance of MOB and clusterMOB in recovering the true splitting variables and splitting values, for trees matching the true trees size and trees not matching the true tree sizes, separately.

\subsubsection{Predictive accuracy}

We evaluated the predictive accuracy of MOB and clusterMOB by calculating correlations between the predicted treatment differences and true treatment-effect differences ($\beta_1$, Figure \ref{fig:modelC}) of test observations. Evaluation using the same data for testing and for training model results in overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, so the MOB and clusterMOB trees were used for prediction of new observations in test datasets. These test datasets were generated from the same population as the training datasets. Because the cluster specific intercepts $b_i$ were randomly generated for training and testing, the observations in the test datasets were from 'new' clusters. As a result, only the linear model tree was used for prediction with clusterMOB.

For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment differences: one for clusterMOB, and one for MOB. To detect predictors of predictive accuracy of both algorithms, a linear model tree was built using the \verb|lmtree| function (with maximum tree depth set to four). The outcome variable in this tree is the correlation between the true and predicted treatment differences. The predictor variable for the linear model is algorithm type (MOB or clusterMOB), and the parameters of the data-generating process are the (potential) partitioning variables, which will be described next.

\subsection{Simulation design}

\subsubsection{Training datasets}
For generating treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure \ref{fig:modelC}. Figure \ref{fig:modelC} shows two subgroups with a treatment interaction, and two subgroups without a treatment interaction. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables in $U$ (e.g., $U_3$, $U_4$) are noise variables.

In generating the datasets, we varied seven parameters of the data-generating process. For each cell, 50 datasets were generated, resulting in a total of 50 x 3 x 2 x 2 x 3 x 3 x 3 x 2 = 32,400 training datasets:

\begin{enumerate}
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_r$: $r=5$, $r=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables and the other $U_k$s are not).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_r$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
\item Three levels for the number of clusters: $n=5$, $n=10$, $n=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b_i}=0$, $\sigma_{b_i}=5$, $\sigma_{b_i}=10$.
\item Three levels for the intercorrelations between $b_i$ and one of the $U_k$s: $b_i$ and $U_k$ uncorrelated, $b_i$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation between the correlated $U_k$ and $b_i$ of about 0.42), $b_i$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation between the correlated $U_k$ and $b_i$ of about 0.42). 
\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for treatment 1 ($X_{1}=0$) and treatment 2 ($X_{1}=1$). The levels for mean differences in subgroups with differential treatment effect were $|\beta_1| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_1| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

We expect the last four parameters to be most influential in determining the difference in performance between MOB and clusterMOB. We expect the first four parameters to have little effect on the difference in performance between MOB and clusterMOB.

As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{15}$ were drawn from a multivariate normal distribution with ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between the variables in $U$ vary according to the third facet of the simulation design described above.

To generate the random error term $\epsilon$, for every observation we drew a value from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.

To generate the cluster-specific intercepts $b_i$, we partitioned the sample into equally sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simualtion design. For each cluster we drew a single $b_i$ from a normal distribution with $\mu_{b_i} = 0$ and the value of $\sigma_{b_i}$ given by the fifth facet of the simulation design. When $b_i$ was correlated with one of the potential partitioning variables $U$, the partitioning or non-partitioning covariate correlated with $b_i$ was randomly selected.

To generate the node-specific fixed-effects, we partitioned the sample according to the endnodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with $p =.5$) as an indicator for treatment type (i.e., the second column of $X$).

Finally, outcome variable Y was calculated according to the model in Equation \ref{eq:fixedeffects}.

\section{Results}

\subsection{Tree structure recovery}

\subsection{Tree size}

Overall, smaller trees were created by clusterMOB than by MOB. The average tree size for clusterMOB, across the 32,400 training datasets, was 7.15 (SD=0.61). For MOB, average tree size was 8.11 (SD=2.05). The true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}). As Table \ref{tab:treesize} shows, 29,556 out of 32,400 (91.22\%) clusterMOB trees matched the true tree size, and 20,578 out of 32,400 (63.51\%) MOB trees matched the true tree size.

The linear model tree depicting the relationship between the various data-generating parameters and tree size is presented in Figure \ref{fig:treesizetree}. The endnotes of Figure \ref{fig:treesizetree} indicate that in many situations MOB and clusterMOB build about equally sized trees, with two notable exceptions: nodes 7 and 14.

Node 14 represents a large majority (59.26\%) of the total number of simulated datasets, and shows that clusterMOB recovered the true tree size in nearly all of these datasets, whereas MOB provided trees with more than 7 nodes in the majority of these datasets. The treesize distributions in node 14 indicate that the increased power provided by a larger sample size ($N=500$ or $N=1000$) results in spurious splits by MOB, when $\sigma_{b_i}$ is non-zero and when $b_i$ is correlated to one of the variables in $U$. Because clusterMOB can more adequately deal with the additional variance in $Y$ caused by non-zero values of $\sigma_{b_i}$, the size of clusterMOB trees seems not to be influenced by values of $\sigma_{b_i}$.

Node 7 indicates that with low sample size ($N=200$), when $b_i$ is not correlated to one of the $U$ variables, and the variance of $b_i$ is large, MOB tends to grow smaller trees. Because MOB cannot account for the variance in $Y$ due to cluster-specific effects, MOB has difficulty to detect partitions when $\sigma_{b_i}$ is large and sample size is low. At the same time, the upper whisker of the boxplot of MOB in node 7 reaches a value of 8, indicating that in many cases, MOB may still find spurious splits in small datasets, when $\sigma_{b_i}$ is large.

\subsubsection{Recovery of partitioning variables and values}

Overall, in recovering the true partitioning variables and values, clusterMOB performed better than MOB. However, for the first split, both MOB and clusterMOB always selected the true partitioning variable ($U_2$). The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94 for both clusterMOB and MOB. However, MOB showed higher variability in recovering the splitting value than clusterMOB (SD=0.155 and SD=0.126, respectively).

In Table \ref{tab:splitstats}, summary statistics for the splitting values recovered by MOB and clusterMOB are presented, for trees with the true tree size (i.e., total number of nodes $= 7$). As Table \ref{tab:splitstats} shows, for the right sized trees produced by clusterMOB, 22 out of 88,668 splits (0.025\%) involved the wrong variables. For the right sized trees produced by MOB, 361 out of 61,734 splits (0.585\%) involved the wrong variables. This makes the odds for selecting the wrong splitting variable 23.7 times higher for MOB than for clusterMOB. In recovering the true splitting values, the right-sized clusterMOB trees performed slightly better than the right-sized MOB trees: average splitting values were very close to the true splitting values for both algorithms, but standard deviations were 25 to 100\% higher for MOB, compared to clusterMOB (Table \ref{tab:splitstats}).

In Table \ref{tab:splitstats_wrongsizetrees}, summary statistics for the wrong-sized trees (i.e., total number of nodes $\neq 7$) are presented. For both MOB and clusterMOB, splitting values for the true partitioning variables were close to the true values, although the splitting values recovered by MOB had higher standard deviations. Out of a total number of 10,908 splits in wrong-sized trees created by clusterMOB, 1,522 (13.95\%) involved a wrong variable. For MOB, 11,582 (21.64\%) out of a total number of 53,524 splits in wrong-sized trees involved a wrong variable.

\subsection{Predictive accuracy on test data}

Overall, the treatment differences as predicted by clusterMOB were closer to the true differences than the predicted treatment differences of MOB. The average correlation between the true and predicted treatment differences over all 32,400 datasets was .88 (SD=0.20) for the trees generated by MOB, and .94 (SD=0.10) for the trees generated by clusterMOB.

To check whether correlations between true and predicted treatment differences are appropriate indicators of predictive accuracy, we compared the means and variances of the true and predicted treatment differences. The mean true treatment difference was 0 (Figure \ref{fig:modelC}). The mean treatment difference, averaged over all test datasets, as estimated by clusterMOB was -0.034 (SD=0.63), and as estimated by MOB was -0.031 (SD=0.95). This was only a small deviation from the expected value of 0, when compared to the spread of the distribution of treatment differences: The standard deviation of the true treatment differences, averaged over all test datasets, was 3.34 (SD=1.12). The average standard deviation of the predicted treatment differences for clusterMOB was 3.48 (SD=1.21) and 3.65 (SD=1.36) for MOB. These values indicate that true and predicted treatment differences had similar scales.

The linear model tree depicting the relationship between the various data-generating parameters and predictive accuracy of MOB and clusterMOB is presented in Figure \ref{fig:correlationtree}, showing that clusterMOB generally outperforms MOB. The dependent variable in this tree is the correlation between the true treatment differences, and the predictions of MOB and clusterMOB. Figure \ref{fig:correlationtree} shows clear main effects of sample size $N$, treatment difference size, and value of $\sigma_{b_i}$.

With larger sample sizes ($N =$ 500 or 1,000), both MOB and clusterMOB perform better than with smaller sample sizes ($N = 200$). With larger treatment differences ($d=5$ v.s. $d=2.5$), the difference between the performance of MOB and clusterMOB becomes smaller, and the performance of MOB shows less variation. When the random intercepts are sampled from a population distribution with large variance (i.e., $\sigma_{b_i} = 10$), the difference in accuracy between clusterMOB and MOB is more pronounced, than when there are no random intercept differences, or when the differences are small (i.e., $\sigma_{b_i} = $ 0 or 5).

As the boxplots in Figure \ref{fig:correlationtree} show, for some simulated datasets, correlations between predicted and true differences were obtained that were clear outliers. LIT should be noted that low or negative correlations between the true and predicted treatment differences were much more often found for MOB than for clusterMOB: a correlation $<.40$ was obtained in 930 out of 32,400 datasets for MOB, and in 175 out of 32,400 datasets for clusterMOB.

\subsection{Example trees}

To illustrate the results of the application of clusterMOB and MOB, we selected a dataset from the simulation where the results of MOB are confounded by random effects, whereas clusterMOB recovered the right tree. The total number of observations in the dataset was 1,000, the correlation between partitioning variables $U$ was 0.30, the number of potential partitioning variables in $U$ was 5, the treatment effect difference was 5, the value of the random intercept $b_i$ was correlated to the value of $U_4$, the number of clusters was 10, and the variance of the distribution from which the random intercepts were drawn ($\sigma_{b_{i}}$) was 10. The resulting clusterMOB and MOB trees are presented in Figure \ref{fig:exampleREEMob} and \ref{fig:exampleMob}, to illustrate the differences between the results of both algorithms.

Figure \ref{fig:exampleREEMob} shows that clusterMOB approximated the true tree structure as represented in Figure \ref{fig:modelC} quite well. The MOB tree in Figure \ref{fig:exampleMob} shows that the first partitioning variable ($U_2$) and splitting value ($\approx 30$) was correctly identified. However, after the first split, MOB incorrectly identified $U_4$ as a partitioning variable for observations with values $U_2 \leq 30$. Further, spurious partitions were made for these observations, involving variables $U_1$ and $U_4$. Although the splitting values were close to the true values for $U_1$ ($\approx 17$), node 5 wrongly indicates a lack of treatment effect differences, and node 10 wrongly indicates a treatment effect difference in favor of treatment 2. The right side of the tree also shows an additional spurious partition after node 11, involving $U_4$.

\section{Discussion}

The results of our simulation study show that clusterMOB performed very well in recovering treatment-subgroup interactions. clusterMOB recovered the right tree structure in 91.2\% of the simulated datasets. MOB performed less accurate than clusterMOB in recovering treatment-subgroup interactions, recovering the right tree structure in 63.5\% of the simulated datasets. The better performance of clusterMOB was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables. In these instances, the random effects gave rise to spurious subgroups (splits), in the trees resulting from application of MOB.

The accuracies of both MOB and clusterMOB in predicting the difference in treatment outcomes were high, with accuracies of .88 and .94, respectively. clusterMOB clearly outperformed MOB when random effects in the datasets were sizable, and random intercepts were correlated to potential partitioning variables.

As expected, when simulated datasets contained no random effects, MOB and clusterMOB performed equally well. This finding indicates that clusterMOB can be applied, whenever cluster-specific random effects are expected: In the absence of random effects, MOB and clusterMOB are expected to perform equally well, and in the presence of random effects, clusterMOB will outperform MOB. Not surprisingly, accuracy of predicted treatment of both algorithms deteriorated when sample size was low (i.e., $N=200$). Sample size influenced performance of MOB and clusterMOB similarly, suggesting that the larger number of estimated parameters for clusterMOB did adversely influence accuracy with low sample sizes. However, our simulation results do warrant caution for the detection of treatment-subgroup interactions or treatment moderators in small datasets (e.g., single RCTs), irrespective of the algorithm used.

Although these findings are encouraging for the use of clusterMOB in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations of our study and challenges for future research should be noted.

As noted in the Introduction, simulations in the current study were confined to random-intercept models. clusterMOB allows for the estimation of random slopes as well, but estimation of random treatment effects is currently not possible, as treatment effects are estimated with local linear fixed-effects models. Our simulations also did not include models with multiple fixed-effects predictor variables in $X$. Multiple fixed-effects predictor variables can be easily included in MOB and clusterMOB, but it should be noted that the parameters corresponding to these variables will then be included in tests for parameter instability as well, which may be undesirable. For example, in RCTs, an ANCOVA will often be used, to control for the linear effects of pretreatment values on the treatment outcome variable. Whether or not such parameters should be included in parameter stability tests, or should be allowed to vary over partitions, should be decided by the researcher.

One challenge for further research is the development of parameter stability tests for random-effects parameters. In the current study, random-effects parameters were estimated globally, using all observations in the dataset, and fixed-effects parameters were estimated locally, using the observations in a single node. This would allow, for example, for estimation of random treatment effects, instead of fixed treatment effects.

A second challenge is the development of more adequate ways to deal with missing data in treatment-subgroup interaction detection. MOB, like all tree-based algorithms for treatment-subgroup interaction detection, handles missing data by listwise deletion. However, missing data commonly occurs in clinical trails, and listwise deletion is obviously not the preferred method for dealing with missing data \cite<e.g., >{WoodyWhit04}.

In conclusion, clusterMOB provided highly accurate recovery of treatment-subgroup interactions and predictions of treatments differences in the presence and absence of cluster-specific random effects. Therefore, clusterMOB is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.

\bibliographystyle{apacite}
\bibliography{refs}

\section{Appendix: Notation}

\begin{table}
\begin{tabular}{ll}
$1,...,i,...,n$	&	denotes cluster number \\
$1,...,j,...,t$ &	denotes terminal node number in a tree \\
$1,...,k,...,r$ &   denotes partitioning variable number \\
$\beta$ 		&	$2t \times 1$ column vector ${\left[ {\beta_l}^T,...,{\beta_j}^T,...,{\beta_t}^T \right]}^T$ of fixed-effects coefficients \\
$\beta_{j}$		& 	$2 \times 1$ column vector of fixed-effects coefficients in terminal node $j$ \\
$b$ 			&	$n \times 1$ column vector ${\left[ {b_l},...,{b_i},...,{b_n} \right]}^T$ of random intercepts \\
$b_{i}$ 		&	random intercept in cluster $i$\\
$d$				&	$\beta_1 / \sigma_{\epsilon}$; effect size of expected difference between treatment 1 and 2 in outcome $Y$\\
				&	with $X_1 = 0$ and $X_1 = 1$\\
$\epsilon$ 		&	$N \times 1$ column vector of residuals \\
$N$ 			&	total number of observations \\
$N_{i}$ 		&	number of observations in cluster $i$ \\
$N_{j}$			&	number of observations in terminal node $j$ \\
$\sigma_{b_i}$	&	square root of variance of $b_i$\\
$\sigma_{\epsilon}$&square root of the variance of $\epsilon$\\
$U$ 			&	$N \times r$ matrix of (potential) partitioning variables \\
$X$ 			&	$N \times 2t$ (block diagonal) design matrix of treatment indicators\\
$y$ 			&	$N \times 1$ column vector of response variable values \\
$Z$ 			&	$N \times n$ (block diagonal) design matrix of random effects predictor variables, \\
				&	with $N_i \times 1$ column vectors of ones on the diagonal \\
\end{tabular}
\end{table}

\begin{figure}
    \includegraphics[width=8cm]{Interaction_ex;treedepth1.pdf}
	\caption{Example of a linear fixed-effects model for treatment outcomes (N=150). The dot for treatment 1 represents $\beta_0$, the dot for treatment 2 representes $\beta_1$.}
	\label{fig:fixed_effects}
\end{figure}

\begin{figure}
    \includegraphics[width=15cm]{Interaction_ex;treedepth3.pdf}
    \caption{Example of tree representation of model-based recursive partitions, based on the same data as Figure \ref{fig:fixed_effects}. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

\begin{figure}
    \tikzstyle{decision} = [diamond, draw]
    \tikzstyle{line} = [draw, -stealth, thick]
    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
        \node [elli] (node1) { };
        \node [elli, below of =node1, xshift=-9em] (node2) { };
        \node [elli, below of =node1, xshift=9em] (node3) { };
        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_0=17.5$\\$\beta_1=-5.0$\\$d=-1.0$};
        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_0=30.0$\\$\beta_1=0.0$\\$d=0.0$};
        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_0=30.0$\\$\beta_1=0.0$\\$d=0.0$};
        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_0=42.5$\\$\beta_1=5.0$\\$d=1.0$};
        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
    \end{tikzpicture}
	\caption{Data generating model for treatment-subgroup interactions. $d$ denotes the standardized mean difference between the outcomes of treatment 1 and 2 (i.e., $\beta_1 / \sigma_{\epsilon}$).}
	\label{fig:modelC}
\end{figure}

\begin{table}
\caption{Tree size distributions for MOB and clusterMOB.}
\small
\begin{tabular}{lrrrrrrrrr}
  \thickline
                    &       &   \multicolumn{7}{c}{MOB tree size}\\
                    \cline{3-9}
                    &       &   3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
  \hline
  clusterMOB  &   3   &   2   &   1   &    0  &   0   &   0  &    0   &    0  &    3  \\
  tree size                    &   5   &  13   & 228   &   33  &   3   &   0  &    0   &    0  &  277  \\
                    &   7   & 125   & 743  &19,936 &3,266  &3,731  &1,427   &  328 &29,556  \\
                    &   9   &   5   &  30   &  593 &1,161   & 404  &  220   &   59 & 2,472  \\
                    &  11   &   0   &   0   &   16  &  13   &  43  &   15   &    2  &   89  \\
                    &  13   &   0   &   0   &    1  &   0   &   0  &    3   &    0  &    3  \\
                    \cline{3-10}
                    & total &  145  &1002  &20,578	&4,443	&4,178 &1,665   &  389  &32,400  \\
  \hline
  \multicolumn{9}{l}{\textit{Note. }Tree sizes are expressed as the total number of nodes in the tree. A tree}\\
  \multicolumn{9}{l}{with a total of $k$ nodes has has $(k+1)/2$ terminal nodes.}\\
\end{tabular}
\label{tab:treesize}
\end{table}

\begin{sidewaysfigure}
    \includegraphics[width=20cm]{treesize_mob_maxdepth=4.pdf}
	\caption{Linear model tree of tree sizes for MOB and clusterMOB. The y-axes of the boxplots represent the total number of nodes in a tree (a tree with a total of $k$ nodes has $(k+1)/2$ terminal nodes). Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; corUbi = correlation between $b_i$ and one of the $U$ variables; sigmabi = $\sigma_{b_i}$; np = number of potential partitioning variables; rho = $\rho$; the correlation between the potential partitioning variables.}
	\label{fig:treesizetree}
\end{sidewaysfigure}

\begin{table}
\caption{Summary statistics for splitting values and variables of right-sized trees for MOB and clusterMOB}
\small
\begin{tabular}{cccrrrrrrr}
	\thickline
	 splitting & true && \multicolumn{3}{c}{MOB} && \multicolumn{3}{c}{clusterMOB}\\	
	\cline{4-6} \cline{8-10}
	variable &  value && \multicolumn{1}{c}{no. of splits} &  \multicolumn{1}{c}{M} & \multicolumn{1}{c}{SD} && \multicolumn{1}{c}{no. of splits}& \multicolumn{1}{c}{M} & \multicolumn{1}{c}{SD} \\
	\hline
	$U_1$ &  17 &  &  20,367&   16.49 & 1.07 & &  29,542& 16.57 & 0.84 \\
	$U_2$ &  30 &  &  20,602&   29.94 & 0.35 & &  29,562& 29.94 & 0.16 \\
	$U_3$-$U_4$& -&&    310&       - &   - & &      12&    - &  - \\
	$U_5$ &  63 &  &  20,404&   63.21 & 1.09 & &  29,542& 63.14 & 0.82 \\
	$U_6$-$U_{15}$&-&&   51& 	     - &  - & &	    10&     - &  - \\
	\cline{1-2} \cline{4-4} \cline{8-8} 
	\multicolumn{3}{l}{total no. of splits}& 61,734 & &&& 88,668 \\
\hline
\multicolumn{10}{l}{\textit{Note.} The total number of right-sized trees was 20,578 for MOB and 29,556 for}\\
\multicolumn{10}{l}{clusterMOB, out of a total of 32,400 datasets. The no. of splits refers to the number}\\
\multicolumn{10}{l}{of splits involving a variable; M and SD refer to the mean and the standard deviation}\\
\multicolumn{10}{l}{of the recovered splitting values for a variable. True splitting variables are $U_1$, $U_2$ and}\\
\multicolumn{10}{l}{$U_5$ (Figure \ref{fig:modelC}); therefore other variables do not have true splitting values. Note that}\\
\multicolumn{10}{l}{splitting variables may appear multiple times in the same tree, therefore a variable}\\
\multicolumn{10}{l}{may be involved in more splits than there are trees.}\\
\end{tabular}
\label{tab:splitstats}
\end{table}

\begin{table}
\caption{Summary statistics for splitting variables and values of wrong-sized trees for MOB and clusterMOB}
\small
\begin{tabular}{cccrrrrrrr}
	\thickline
	 & true && \multicolumn{3}{c}{MOB} && \multicolumn{3}{c}{clusterMOB}\\	
	\cline{4-6} \cline{8-10}
	variable &  value && \multicolumn{1}{c}{no. of splits} &  \multicolumn{1}{c}{M} & \multicolumn{1}{c}{SD} && \multicolumn{1}{c}{no. of splits}& \multicolumn{1}{c}{M} & \multicolumn{1}{c}{SD} \\
	\hline

	$U_1$ &  17 &   &  14,255&   15.00 & 4.57 & &    3,071&   16.05 & 3.43 \\
	$U_2$ &  30 &   &  13,592&   29.95 & 3.49 & &    3,243&   29.94 & 2.98 \\
	$U_3$-$U_4$&-& &   10,889&      - &   - & &     738&      -  &  - \\
	$U_5$ &  63 &   &  14,092&   64.77 & 4.53 & &    3,072&   63.81 & 3.32 \\
	$U_6$-$U_{15}$&-&&   696&	  - &   - & &      784	&   - &	 - \\
	\cline{1-2} \cline{4-4} \cline{8-8} 
	\multicolumn{3}{l}{total no. of splits} & 53,524 & &&& 10,908 \\
	\hline
\multicolumn{10}{l}{\textit{Note.} The total number of wrong-sized trees was 11,822 for MOB and 2,844 for}\\
\multicolumn{10}{l}{clusterMOB, out of a total of 32,400 datasets. The number of splits in a wrong-}\\ 
\multicolumn{10}{l}{sized tree is unequal to 3, and can be any other number from 1 through 7. The}\\
\multicolumn{10}{l}{no. of splits refers to the number of splits involving a variable; M and SD refer}\\
\multicolumn{10}{l}{to the mean and standard deviation of the recovered splitting value for a variable.}\\
\multicolumn{10}{l}{True splitting variables are $U_1$, $U_2$ and $U_5$ (Figure \ref{fig:modelC}); therefore other variables}\\
\multicolumn{10}{l}{do not have true splitting values. Note that splitting variables may appear multiple}\\ 
\multicolumn{10}{l}{times in the same tree, therefore a variable may be involved in more splits than}\\ 
\multicolumn{10}{l}{there are trees.}\\
\label{tab:splitstats_wrongsizetrees}
\end{tabular}
\end{table}

\begin{sidewaysfigure}
	\includegraphics[width=20cm]{correlation_mob_maxdepth=4.pdf}
	\caption{Linear model tree of correlations with true treatment differences for clusterMOB and MOB trees. Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; treatdiff = $\beta_1$, the unstandardized mean difference in treatment outcomes in subgroups with differential effects for treatment 1 and treatment 2; sigmabi = $\sigma_{b_i}$.}
	\label{fig:correlationtree}
\end{sidewaysfigure}

\begin{sidewaysfigure}
    \includegraphics[width=20cm]{Mobtree1218.pdf}
	\caption{Example tree build by MOB.}
	\label{fig:exampleMob}
\end{sidewaysfigure}

\begin{figure}
    \includegraphics[width=12cm]{REEMobtree1218.pdf}
	\caption{Example tree build by clusterMOB.}
	\label{fig:exampleREEMob}
\end{figure}

\end{document}

### 2015-07-13 pkg/partykit/inst/mloss/mloss2.Rnw
JMLR requests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@728 255f4249-2263-4417-b1ea-8498779e27a7

--- 2015-02-20 pkg/partykit/inst/mloss/mloss2.Rnw
@@
 
+\jmlrheading{17}{2015}{1-5}{4/14}{7/15}{Torsten Hothorn and Achim Zeileis}
+
+\ShortHeadings{A Modular Toolkit for Recursive Partytioning}{Hothorn and Zeileis}
+
 %% header declarations
-\author{Torsten Hothorn \\ \addr{Universit\"at Z\"urich} \\ \addr{E-mail: Torsten.Hothorn@R-project.org} \AND
-        Achim Zeileis \\ \addr{Universit\"at Innsbruck} \\ \addr{E-mail: Achim.Zeileis@R-project.org}}
+\author{\name Torsten Hothorn \email Torsten.Hothorn@R-project.org \\
+        \addr Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention, Universit\"at Z\"urich, Switzerland \\
+        \AND
+        \name Achim Zeileis \email Achim.Zeileis@R-project.org \\
+        \addr Institut f\"ur Statistik, Universit\"at Innsbruck, Austria} 
 \title{\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}}
@@
 
-\begin{abstract}
+\begin{abstract}%
   The \proglang{R} package \pkg{partykit} provides a flexible toolkit for
@@
 \begin{keywords}
-recursive partitioning, regression trees, classification trees, statistical learning, \proglang{R}.
+recursive partitioning, regression trees, classification trees, statistical learning, \proglang{R}
 \end{keywords}

### 2015-07-30 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@734 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{tikz}
\linespread{1.2}

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets: Combining Model-Based Recursive Partitioning and Random-Effects Estimation}
\shorttitle{Detection of treatment-subgroup interactions in clustered datasets}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

%\abstract{}

\begin{document}

\maketitle
\pagewiselinenumbers

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMMtree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMMtree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMMtree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMMtree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMMtree predicted treatment outcome differences more accurate than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMMtree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMMtree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss directions for future research.

\section{Introduction}

In medicine-efficacy research, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Personalized medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once, and can automatically detect (higher order) interactions between predictor variables. Several promising tree-based algorithms and software packages have been developed to assist in the detection of treatment-subgroup interactions (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Of these tree-based methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) may be the most flexible in detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, where datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}. When the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for treatment-subgroup interaction detection, which takes the clustered nature of datasets into account. The algorithm combines MOB with the estimation of random effects, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we first discuss the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning of the GLM, and the generalized linear mixed-effects model (GLMM). Then, we present a new algorithm, which combines model-based recursive partitioning and random-effects estimation: GLMMtree. In a simulation study, we evaluate the comparative accuracy of the new algorithm. Finally, we apply GLMMtree to an existing dataset on the effects of treatments for depression, to illustrate the application of the algorithm.

\section{General modeling framework}

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of treatment outcome $y$ for observation $i=1,\dots,N$ (an overview of notation used is provided in the Appendix):

\begin{equation}
\label{eq:fixedeffects}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta
\end{equation}

Where $y_i$ is the value of the linear predictor of the response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor and the mean of the response distribution function. In case of a continuous response variable, $g$ is the identity function. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type. $\beta$ is a vector of fixed-effects regression coefficients, containing the intercept, which is the mean value of the linear predictor in the first treatment group, and the slope, which is the difference in mean values of the linear predictor between the first and second treatment groups.

To keep notation and examples simple, we assume $x_{i}$ and $\beta$ to have length 2, That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the linear model. However, additional treatment conditions and covariates can easily be included.

An example of such a GLM for a continuous outcome variable is graphically represented in Figure \ref{fig:fixedeffects}. The example is based on simulated data. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the outcome variable (posttreatment depression score) among 150 participants, who were randomly assigned to treatments 1 and 2. Little overall difference between the outcomes of both treatments is suggested, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by other variables. Conditional on other variables (i.e., subgroup indicators), the relationship between $x$ and $y$ may vary in strength and/or direction (c.f., Simpson's paradox; \citeNP{Simp51}).

\begin{figure}[!htbp]
    \includegraphics[width=8cm]{Interaction_ex;treedepth1.pdf}
	\caption{Example of a linear fixed-effects model for treatment outcomes (N=150). The dot for treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
	\label{fig:fixedeffects}
\end{figure}

\subsection{Model-based recursive partitioning}

The rationale behind MOB is that a global model for all observations, like that in Equation \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. Whereas CART trees have constant fits in the terminal nodes, MOB trees have parametric models with one or more predictor variables in their terminal nodes.

To find partitions and better-fitting local linear models, the MOB algorithm tests for parameter instability. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more additional covariates. To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subsamples.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. The expected value of the scores over all observations in a dataset is zero, by definition. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$. To statistically test whether there are systematic deviations of the scores from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the residual sum of squares in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partitions can be represented as a tree. If the partitioning is based on the GLM, the result is a GLMtree, which has a local fixed-effects regression model in every $j$th (where $j = 1,...,J$) terminal node of the tree. As a result, in the GLMtree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ 'falls':

\begin{equation}
\label{eq:fixedeffects_MOB}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j}
\end{equation}

Figure \ref{fig:example_mobtree} provides an example of the GLMtree model in Equation \ref{eq:fixedeffects_MOB}, based on the same data as was used for Figure \ref{fig:fixedeffects}. By using four additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline, age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. The leftmost subgroup in Figure \ref{fig:example_mobtree} represents observations with low duration and low anxiety, for which treatment 1 is much more beneficial then treatment 2 (i.e., lower posttreatment depression scores). The rightmost subgroup represents observations with longer duration of depressive symptoms, for whom treatment 2 is much more beneficial than treatment 1. The two terminal nodes in the middle indicate that for patients with low duration, treatment 1 is only beneficial to those with moderate levels of anxiety (i.e., a values $>10$ and $\leq 12$). Age did not have an effect on treatment outcome, and therefore does not appear as a splitting variable in the tree.

\begin{figure}[!h]
    \includegraphics[width=15cm]{Interaction_ex;treedepth3.pdf}
    \caption{Example of tree representation of model-based recursive partitions, based on the same data as Figure \ref{fig:fixedeffects}. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

\subsection{GLMM}
When a dataset contains observations from multiple clusters (i.e., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} can be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

\begin{equation}
\label{eq:mixedeffects}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta + z_{i}b_{m}
\end{equation}

Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$, of which observation $i$ is part. Within the linear mixed-effects model, it is assumed that values of $b_{m}$ are normally distributed, with mean zero and variance $\sigma^{2}_{b_m}$. For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates can easily be included. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in for example \cite{BrykyRaud92}. Note that for prediction of the treatment outcome for a new observation from a new cluster, only the fixed-effects part of Equation \ref{eq:mixedeffects} would be used.

\subsection{Combining model-based recursive partitioning and random-effects estimation}

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMMtree, which combines the GLMM in Equation \ref{eq:mixedeffects} with the GLMtree in Equation \ref{eq:fixedeffects_MOB}:

\begin{equation}
\label{eq:glimmertree}
	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j} + z_{i}b_{m}
\end{equation}

To estimate the partitions, and values of $\beta_j$ and $b_m$ for this model, we take an iterative approach, alternating between (1) assuming the random-effects coefficients $b_m$ known, allowing for partitioning the datasets and estimating the corresponding $\beta_j$ values; and (2) assuming the partition and corresponding $\beta_j$ values known, allowing for estimation of the random-effects coefficients $b_m$.

In Figure \ref{fig:algorithm}, a schematic representation of the GLMMtree algorithm is presented. The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLMtree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLMtree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLMtree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}. A similar approach has been taken by \citeA{HajjyBell11} and \citeA{SelaySimo12}, who added random-effects estimation to CART trees with constant fits, instead of linear models, in the terminal nodes.

\begin{figure}[!htbp]
	\begin{tabular}{ll}
		\hline 
		\multicolumn{2}{l}{\textbf{Algorithm} GLMMtree}\\
		\hline
		&Step 0: Initialize by setting all values $b_m$ to $0$.\\

		&Step 1: Given the current $b_m$ values, partition dataset and estimate corresponding values of $\beta_j$.\\

		&Step 2: Given the current partition and corresponding $\beta_j$ values, estimate $b_m$ values. \\

		&Step 3: Repeat steps 1 and 2 until convergence.\\
		\hline
	\end{tabular}
	\caption{Description of the GLMMtree algorithm}
	\label{fig:algorithm}
\end{figure}

In what follows, we present a simulation study in which we assess the performance of GLMMtree in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments, for continuous outcomes. In addition, we will compare the performance of GLMMtree with that of GLMtree. In the simulation study, the main interest will be in the effects of sample size, and the presence and magnitude of treatment-subgroup interactions and random effects, but other parameters will be varied, as well.

For GLMMtree, we expect the accuracy of recovered trees and predictions to improve with increasing sample size, and magnitude of the differences in treatment outcomes. For GLMtree, we have the same expectation, when random effects are absent; that is, when the variance of the random coefficients is zero, we expect GLMtree and GLMMtree to perform equally well. When random effects are present, we expect GLMMtree to perform better than GLMtree, and more so when the variance of random-effects coefficients increases.

\section{Simulation: Method}

\subsection{Software}

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| \cite{HothyZeil14,ZeilyHoth08} for the estimation of GLMtrees, and \verb|lme4| \cite{BateyMeac12} for the estimation of random-effects coefficients. For all functions, default settings were used, with exception of maximum tree depth. Maximum tree depth was set to four (i.e., maximum of eight terminal nodes) for all trees, as this yields a model and graphical representation which is easy to interpret.

\subsubsection{Estimation of GLM- and GLMMtrees}

For estimating GLMtrees, the \verb|glmtree| function from the \verb|partykit| package was used. The \verb|glmtree| function builds a generalized linear model tree: a model-based recursive partition based on a GLM (Equation \ref{eq:fixedeffects_MOB}).

For estimating GLMMtrees, we implemented the algorithm as described in Figure \ref{fig:algorithm} in a function that iterates between (1) estimation of a linear model tree using the \verb|glmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|glmer| function. Convergence of \verb|GLMMtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2). When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|GLMMtree| has converged.

\subsection{Simulation design}

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure \ref{fig:modelC}. Figure \ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[!h]
    \tikzstyle{decision} = [diamond, draw]
    \tikzstyle{line} = [draw, -stealth, thick]
    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
        \node [elli] (node1) { };
        \node [elli, below of =node1, xshift=-9em] (node2) { };
        \node [elli, below of =node1, xshift=9em] (node3) { };
        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
    \end{tikzpicture}
	\caption{data-generating model for treatment-subgroup interactions. $d$ denotes the standardized mean difference between the outcomes of treatment 1 and 2 (i.e., $\beta_j1 / \sigma_{\epsilon}$).}
	\label{fig:modelC}
\end{figure}

\subsubsection{Datasets without treatment-subgroup interactions}
For generating datasets without treatment-subgroup interactions, we used a design in which there is only a main effect of treatment in the population. In these datasets, the mean of the outcome variable was 30, and $d$ and the first and second elements of $\beta$ have the same value for all observations (i.e., when $d = 1.0$, the first and second elements of $\beta$ are 37.5 and 42.5, respectively).

\subsubsection{Parameters of the data-generating process}

In generating the datasets, we varied seven parameters of the data-generating process:

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b_m}=0$, $\sigma_{b_m}=5$, $\sigma_{b_m}=10$.
\item Three levels for the intercorrelations between $b_m$ and one of the $U_k$ variables: $b_m$ and $U_k$ uncorrelated, $b_m$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b_m$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for treatment 1 ($X_{1}=0$) and treatment 2 ($X_{1}=1$). The levels for mean differences in subgroups with differential treatment effect were $|\beta_1| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_1| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in 50 x 3 x 2 x 2 x 3 x 3 x 3 x 2 = 32,400 training datasets. For the datasets with main treatment effect only, the 6th parameter of the data-generating process had only two levels ($b_m$ correlated with one of the $U_k$ variables, and $b_m$ not correlated with any of the $U_k$ variables). Therefore, 50 x 3 x 2 x 2 x 3 x 3 x 2 x 2 = 21,600 datasets without treatment-subgroup interactions were generated.

\subsubsection{Variable distributions}

As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{M}$ were drawn from a multivariate normal distribution with ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between the variables in $U$ vary according to the third facet of the simulation design described above.

To generate the random error term $\epsilon$, for every observation we drew a value from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster we drew a single $b_m$ from a normal distribution with $\mu_{b_m} = 0$ and the value of $\sigma_{b_m}$ given by the fifth facet of the simulation design. When $b_m$ was correlated with one of the potential partitioning variables, the partitioning or non-partitioning covariate correlated with $b_m$ was randomly selected.

To generate the node-specific fixed-effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with $p =.5$) as an indicator for treatment type.

Finally, outcome variable $y$ was calculated according to the model in Equation \ref{eq:fixedeffects}.

\subsection{Evaluation of performance}

\subsubsection{Tree size and accuracy}

For every dataset, the accuracy and size of the GLM- and GLMMtrees were evaluated. We calculated the total number of nodes in every tree, and compared it with the true tree size. For datasets with a main treatment effect only, this allowed us to assess the accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned. For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups.

For datasets with treatment-subgroup interactions, we assessed the accuracy of GLMtree and GLMMtree in recovering the true tree. An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., total number of nodes is 7), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation ($\sigma_{U_k}$) of the partitioning variable.

To detect predictors of the performance of both algorithms, GLMtrees were built using the \verb|glmtree| function. Two trees were built for predicting the number of nodes in the trees generated by both algorithms: one for datasets without, and one for datasets with treatment-subgroup interactions. In these trees, the outcome variable is the number of nodes in a tree, the predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process, described above. In addition, for datasets with treatment-subgroup interactions, a GLMtree was built for predicting the probability that the tree of treatment-subgroup interactions was accurately recovered. In this tree, the outcome variable is a binary indicator for whether the tree was accurately recovered (or not). Again, the predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process.

\subsubsection{Predictive accuracy}

We evaluated predictive accuracy of GLM- and GLMMtrees by calculating correlations between the predicted and true treatment-effect differences ($\beta_{j1}$, Figure \ref{fig:modelC}) for test observations. Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets with a main treatment effect only.

Using the same data for training and evaluation of a model results in overly optimistic estimates of predictive accuracy \cite{HastyTibs09}. Therefore, GLM- and GLMMtrees were used for prediction of new observations in test datasets. These test datasets were generated from the same population as the training datasets. Because the cluster-specific intercepts $b_m$ were randomly generated for training as well as test datasets, test observations were from 'new' clusters. As a result, a model without random effects was used for prediction with GLMMtree.

For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment differences: one for GLMtree, and one for GLMMtree. To detect predictors of predictive accuracy, a GLMtree was built. The outcome variable in this tree is the correlation between the true and predicted treatment differences. The predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process, described above.

\section{Simulation: Results}

\subsection{Tree size and accuracy in datasets with main treatment effect only}

In Table \ref{tab:treesize_main}, tree sizes for GLMtree and GLMMtree are presented for datasets with a main treatment effect only. Overall, smaller trees were created by GLMMtree: the average tree size was 1.09 (SD=0.44) for GLMMtree, and 2.02 (SD=1.68) for GLMtree. The estimated probability that a dataset was erroneously partitioned was very small for GLMMtree (.04; Table \ref{tab:treesize_main}), and much larger for GLMtree (.33; Table \ref{tab:treesize_main}).

\begin{table}[!htbp]
\caption{Tree size distributions for GLMtree and GLMMtree for datasets with a main treatment effect only.}
\small
\begin{tabular}{lrrrrrrr}
	\thickline
	&\multicolumn{7}{c}{tree size}\\
	\cline{2-7}
	&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
	\hline
	GLMMtree	& 20625 & 932 & 43 &0&0&0& 21,600\\
		  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   	GLMtree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.20) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
	\hline
  	\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  	\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $k$ nodes has has $(k+1)/2$ }\\
  	\multicolumn{8}{l}{terminal nodes ; the true tree size in datasets with a main treatment effect}\\	  	\multicolumn{8}{l}{only was 1.}\\
\end{tabular}
\label{tab:treesize_main}
\end{table}

A linear model tree (Figure \ref{fig:treesizetree_nointeract}) indicated that the main predictors of tree size in datasets without treatment-subgroup interactions were sample size and magnitude of $\sigma_{b_m}$. When random effects were absent (i.e., $\sigma_{b_m}=0$), both GLMtree and GLMMtree tend to create trees of size 1. In the presence of random effects, GLMMtree also tends to create trees of size 1, but GLMtree created larger trees. For GLMtree, tree size increased with both sample size, and magnitude of $\sigma_{b_m}$.

\begin{sidewaysfigure}[!htbp]
	\includegraphics[width=20cm]{treesize_mob_maxdepth=4_nointeract.pdf}
	\caption{Linear model tree of the size of GLMM- and GLMtrees in datasets without treatment-subgroup interactions. The y-axes of the terminal nodes represent tree size (total number of nodes in a tree). Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; sigmabm = $\sigma_{b_m}$.}
	\label{fig:treesizetree_nointeract}
\end{sidewaysfigure}

\subsection{Tree size in datasets with treatment-subgroup interactions}

In datasets with treatment-subgroup interactions, GLMMtrees were also smaller than GLMtrees. For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}). The sizes of the GLM- and GLMMtrees for datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}. The average size of GLMMtrees was 7.15 (SD=0.61), and the average size of GLMtrees was 8.11 (SD=2.05). The estimated probability that a datasets was erroneously not partitioned was 0, for both GLM- and GLMMtree. However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMMtrees matched the true tree size, whereas a proportion of only .64 of GLMtrees matched the true tree size (Table \ref{tab:treesize_interactions}).

\begin{table}[!htbp]
\caption{Tree size distributions for GLMtree and GLMMtree for datasets with treatment-subgroup interactions.}
\small
\begin{tabular}{lrrrrrrrr}
	\thickline
	&&\multicolumn{7}{c}{tree size}\\
	\cline{2-8}
	&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
	\hline
  	GLMMtree	& 3 & 227& 29556 & 2472 & 89 & 3 &0& 32,400 \\
				&  ($<.01$)  & ($<.01$)& (.91) & ($<.01$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

  	GLMtree		& 145 & 1002 & 20578 & 4443 & 4178 & 1665 & 389 & 32,400\\ 
 			 	& ($<.01$) & (.03) & (.64) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
	\hline
  	\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
  	\multicolumn{9}{l}{nodes in the tree. A tree with a total of $k$ nodes has has $(k+1)/2$ terminal nodes; the true}\\	
  	\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
\end{tabular}
\label{tab:treesize_interactions}
\end{table}

A linear model tree (Figure \ref{fig:treesizetree_interact}) indicated a three-way interaction between sample size, value of $\sigma_{b_m}$, and whether values of $b_m$ are correlated with a potential partitioning variable. Overall, both algorithms created trees of size $\approx 7$, with the following two exceptions:

\begin{sidewaysfigure}[!htbp]
	\includegraphics[width=20cm]{treesize_mob_maxdepth=4.pdf}
	\caption{Linear model tree of tree sizes of GLMM- and GLMtrees in datasets with treatment-subgroup interactions. The y-axes of the terminal nodes represent tree size (total number of nodes in a tree). Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; sigmabm = $\sigma_{b_m}$; corUbi = correlation between random intercepts and one of th e potential partitioning variables; np = number of potential partitioning variables.}
	\label{fig:treesizetree_interact}
\end{sidewaysfigure}

When sample size is small (i.e., $N=200$), the variance of $b_m$ is large (i.e., $\sigma_{b_m} = 10$), and values of $b_m$ are not correlated with one of the $U_k$ variables (Figure \ref{fig:treesizetree_interact}). This was the case for 2,400 datasets (7.41\%), in which mean tree size was 6.12 (SD=1.37) for GLMtree and 7.01 (SD=0.48) for GLMMtree. Because in these cases, GLMtree cannot account for the variance in $y$ due to cluster-specific effects, it has difficulty detecting partitions when $\sigma_{b_m}$ is large and sample size is low. However, the large variability for GLMtree indicates that it may still create spurious splits.

When sample size is larger (i.e., $N=500$ or $N=1000$), when the variance of $b_m$ is non-zero, and values of $b_m$ are not correlated with one of the $U_k$ variables (Figure \ref{fig:treesizetree_interact}). This was the case for 19,200 datasets (59.26\%), in which mean tree size was 10.53 (SD=2.04) for GLMtree, and 7.23 (SD=0.66) for GLMMtree. Obviously, in these cases, GLMtree is more likely to create spurious splits. Because GLMMtree can more adequately deal with the additional variance caused by non-zero values of $\sigma_{b_m}$, the size of GLMMtrees seems not to be influenced much by values of $\sigma_{b_m}$.

\subsection{Tree accuracy in datasets with treatment-subgroup interactions}

To assess accuracy, we inspected the partitioning variables and values selected in every dataset. For the first split, both GLMtree and GLMMtree always selected the true partitioning variable ($U_2$). The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94 for both GLMMtree and GMLtree. However, GLMtree showed somewhat higher variability in recovering the splitting value than GLMMtree (SD=0.155 and SD=0.126, respectively).

GLMMtree performed well in recovering treatment-subgroup interactions, accurately recovering the tree in 90.19\% of datasets. GLMtree accurately recovered the treatment-subgroup interactions in only 61.44\% of datasets. A generalized linear model tree (Figure \ref{fig:treeaccuracytree}) indicated that performance of both algorithms was predicted by a three-way interaction between sample size, value of $\sigma_{b_m}$, and whether values of $b_m$ are correlated with a potential partitioning variable. Overall, both algorithms recovered the true tree with about equal probabilities (i.e., all probabilities $>.85$), with three notable exceptions:

When $\sigma_{b_m} > 0$, and values of $b_m$ are correlated with a potential partitioning variable, GLMMtree clearly outperformed GLMtree. When sample size was low (i.e., $N=200$), the probability of accurately recovering the tree was .59 for GLMtree, and .91 for GLMMtree (Figure \ref{fig:treeaccuracytree}). With larger sample sizes (i.e., $N=500$ or $1,000$), the probability of accurately recovering the tree was .13 for GLMtree, and .89 for GLMMtree (Figure \ref{fig:treeaccuracytree}).

When $\sigma_{b_m}$ was non-zero, and values of $b_m$ are not correlated with a potential partitioning variable, GLMMtree also outperformed GLMtree, but only when sample size was small. In these datasets, the probability of accurately recovering the tree was .72 for GLMtree and .92 for GLMMtree (Figure \ref{fig:treeaccuracytree}). When sample size was larger (i.e., $N=500$ or $1,000$), the probability of accurately recovering the tree was .90 for both algorithms (Figure \ref{fig:treeaccuracytree}).

\begin{sidewaysfigure}[!htbp]
	\includegraphics[width=20cm]{truetree_mob_maxdepth=4.pdf}
	\caption{Generalized linear model tree of accurate recovery of treatment-subgroup interactions by GLMM- and GLMtree in datasets with treatment-subgroup interactions. Bar plots in the terminal nodes represent the proportion of datasets in which the treatment-subgroup interactions were accurately recovered. Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; sigmabm = $\sigma_{b_m}$; corUbi = correlation between random intercepts and one of th e potential partitioning variables; rho = correlation between potential partitioning variables; treatdiff = $\beta_{j1}$, the unstandardized mean difference in treatment outcomes in subgroups with differential effects for treatment 1 and treatment 2.}
	\label{fig:treeaccuracytree}
\end{sidewaysfigure}

\subsection{Predictive accuracy on test data}

Overall, the treatment differences predicted by GLMMtree were closer to the true differences than the treatment differences predicted by GLMtree. The average correlation between the true and predicted treatment differences over all 32,400 datasets was .88 (SD=0.20) for GLMtree, and .94 (SD=0.10) for GLMMtree.

The linear model tree depicting the relationship between the various data-generating parameters and predictive accuracy of GLMtree and GLMMtree, is presented in Figure \ref{fig:correlationtree}. Figure \ref{fig:correlationtree} shows clear main effects of sample size $N$, treatment difference size, and value of $\sigma_{b_m}$. With larger sample sizes ($N = 500$ or $1,000$), both GLMtree and GLMMtree perform better than with smaller sample sizes ($N = 200$). With larger treatment differences ($d=5$), the difference between the performance of GLMtree and GLMMtree becomes smaller, and the performance of GLMtree shows less variation. When the random intercepts are sampled from a population distribution with larger variance (i.e., $\sigma_{b_m} = 10$), the difference in accuracy between GLMMtree and GLMtree is more pronounced, than when there are no random intercept differences, or when these differences are small (i.e., $\sigma_{b_m} = $ 0 or 5).

As the boxplots in Figure \ref{fig:correlationtree} show, for some simulated datasets, correlations between predicted and true differences were obtained that were clear outliers. It should be noted that low or negative correlations between the true and predicted treatment differences were much more often found for GLMtree than for GLMMtree: a correlation $<.40$ was obtained in 930 out of 32,400 datasets for GLMtree, and in 175 out of 32,400 datasets for GLMMtree.

\begin{sidewaysfigure}[!htbp]
	\includegraphics[width=20cm]{correlation_mob_maxdepth=4.pdf}
	\caption{Linear model tree of correlations with true treatment differences for GLMMtrees and GLMtrees. The y-axes in the terminal nodes represent the correlation between the true and predicted treatment effect differences. Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$). N = total sample size; treatdiff = $\beta_{j1}$, the unstandardized mean difference in treatment outcomes in subgroups with differential effects for treatment 1 and treatment 2; sigmabm = $\sigma_{b_m}$.}
	\label{fig:correlationtree}
\end{sidewaysfigure}

\section{Application to real data}

\subsection{Method}
To illustrate the application and potential differences in the results of GLMtree and GLMMtree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}. This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavior therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender either predicted or moderated treatment outcomes. Further details on the dataset are provided in \citeA{CuijyWeit14}.

In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score. The linear predictor was treatment type (0=CBT and 1=PHA). An indicator for study was used as the cluster indicator.

In RCTs, treatment effects are often estimated after controlling posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure. Therefore, we included the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores, as an offset variable in all models. An offset variable is a linear predictor with an a-priori determined coefficient of one. Including the linear regression predictions as an offset has the same effect as statistically controlling for the linear effects of a of pretreatment scores, as is often done in ANCOVA.

The \verb|glmtree| function deals with missing data by listwise deletion. Therefore, we build the trees using data of a subset of 694 patients from 7 studies, as complete observations (i.e., observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score) for these patients were available. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \cite{CuijyWeit14}.

Predictive accuracy of GLMtree and GLMMtree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.

\subsection{Results}

We applied GLMtree and GLMMtree to the dataset with complete observations. The resulting trees are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}. Note that the GLMtree in Figure \ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMMtree algorithm.

\begin{figure}[!ht]
    \includegraphics[width=12cm]{app_lmtree.pdf}
	\caption{GLMtree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
	\label{fig:lmtree_C&W}
\end{figure}

The GLMtree (Figure \ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. Terminal node 2 of Figure \ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores. Terminal node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and antidepressant mediation. Terminal node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for pharmacotherapy.

\begin{figure}[!ht]
    \includegraphics[width=10cm]{app_lmertree.pdf}
	\caption{GLMMtree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
	\label{fig:lmertree_C&W}
\end{figure}

By taking into account the study-specific intercepts, the final GLMMtree (Figure \ref{fig:lmertree_C&W}) indicates that the first split made by GLMtree is a spurious split. The GLMMtree selected only presence of a comorbid anxiety disorder as a partitioning variable. The terminal nodes of Figure \ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores than CBT. The estimated variance of the random intercept term was 2.12, with an estimated intraclass correlation coefficient of .05.

Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMMtree had higher predictive accuracy than the GLMtree. The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .39 (SD=.20) for GLMMtree, and .31 (SD=.24) for GLMtree. This indicates that GLMMtree not only provided higher predictive accuracy, on average, but also had somewhat lower variability of predictive accuracy than GLMtree.

\section{Discussion}

The results of our simulation study show that GLMMtree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions. In the absence of treatment-subgroup interactions, GLMMtree erroneously detected subgroups in only 4\% of the datasets. GLMtree performed less accurate than GLMMtree: in datasets with treatment-subgroup interactions, GLMtree recovered the true tree structure in 61\% of the simulated datasets. In datasets without treatment-subgroup interactions, GLMtree erroneously detected subgroups in 33\% of the datasets.

The better performance of GLMMtree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables. In these instances, the random effects gave rise to spurious subgroup detection (spurious splits) by GLMtree, both in datasets with and without treatment-subgroup interactions.

Also, predictive accuracy of GLMMtree was higher than that of GLMtree. The average correlation between the true treatment differences and those predicted by GLMMtree was .94.  The average correlation between the true treatment differences and those predicted by GLMtree was .88. In terms of predictive accuracy, GLMMtree clearly outperformed GLMtree when random effects in the datasets were sizable, and random intercepts were correlated to potential partitioning variables.

As expected, when random effects were absent from the simulated datasets, GLMtree and GLMMtree showed high and equal predictive accuracy. This finding indicates that GLMMtree can be applied, whenever cluster-specific random effects are expected. In the absence of random effects, GLMtree and GLMMtree are expected to perform equally well, and in the presence of random effects, GLMMtree will outperform GLMtree. This is especially the case with large sample sizes ($N > 500$), as the increased power will likely cause GLMtree to create spurious splits in the presence of random effects.

Not surprisingly, for both algorithms, accuracy of predicted treatment differences was somewhat less when sample size was low (i.e., $N=200$). Sample size influenced performance of GLMtree and GLMMtree similarly, suggesting that the larger number of estimated parameters for GLMMtree did adversely influence accuracy with low sample sizes. However, our simulation results do warrant caution for the detection of treatment-subgroup interactions or treatment moderators in small datasets (e.g., single RCTs), irrespective of the algorithm used.

Although these findings are encouraging for the use of GLMMtree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations of our study and challenges for future research should be noted.

As noted earlier, simulations in the current study were confined to random-intercept models. GLMMtree allows for the estimation of random slopes as well, but estimation of random treatment effects is currently not possible, as treatment effects are estimated with local linear fixed-effects models. Our simulations also did not include models with multiple fixed-effects predictor variables in $X$. Multiple fixed-effects predictor variables can be easily included in GLMtree and GLMMtree, but it should be noted that the parameters corresponding to these variables will then be included in tests for parameter instability as well, which may be undesirable. For example, in RCTs, ANCOVA will often be used to control for the linear effects of pretreatment values on the treatment outcome variable. Whether or not such parameters should be included in parameter stability tests, or should be allowed to vary over partitions, should be decided by the researcher.

One challenge for further research is the development of parameter stability tests for random-effects parameters. In the current study, random-effects parameters were estimated globally, using all observations in the dataset, and fixed-effects parameters were estimated locally, using the observations in a single node. This would allow, for example, for estimation of random treatment effects, instead of fixed treatment effects.

A second challenge is the development of more adequate ways to deal with missing data in treatment-subgroup interaction detection. GLMtree, like most tree-based algorithms for treatment-subgroup interaction detection, handles missing data by listwise deletion. However, missing data commonly occurs in clinical trails, and listwise deletion is obviously not the preferred method for dealing with missing data \cite<e.g., >{WoodyWhit04}.

In conclusion, GLMMtree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatments differences in the presence and absence of cluster-specific random effects. Therefore, GLMMtree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.

\nolinenumbers
\bibliographystyle{apacite}
\bibliography{bib}

\section{Appendix: Notation}

\begin{table}[!h]
\begin{tabular}{ll}
$1,...,i,...,N$ &   observation number \\
$1,...,j,...,J$ &	terminal node number in a tree \\
$1,...,k,...,K$ &   partitioning variable number \\
$1,...,m,...,M$	&	cluster number \\
$\beta_{j}$		& 	column vector of fixed-effects coefficients in terminal node $j$ \\
$b_{m}$ 		&	column vector of random-effects coefficients in cluster $m$\\
$d_j$			&	$\beta_{j1} / \sigma_{\epsilon}$; effect size of predicted differences in treatment outcome $y$ between \\
				&	treatments 1 and 2, in terminal node $j$\\
$\epsilon$		& 	deviation of observed treatment outcome $y$ from its expected value\\
$\sigma_{b_m}$	&	square root of variance of $b_m$\\
$\sigma_{\epsilon}$&square root of the variance of $\epsilon$\\
$U$ 			&	(potential) partitioning variables \\
$U_k$			&	(potential) partitioning variable $k$ \\
$x_i$ 			&	row vector of fixed-effects predictor variable values for observation $i$\\
$y_i$ 			&	treatment outcome for observation $i$ \\
$z_i$ 			&	row vector of random-effects predictor variable values for observation $i$\\
\end{tabular}
\end{table}

\end{document}

### 2015-08-18 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@747 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \linespread{1.2}
+\usepackage{tabto}
 
 \usetikzlibrary{shapes,shadows,arrows,positioning}
-\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets: Combining Model-Based Recursive Partitioning and Random-Effects Estimation}
-\shorttitle{Detection of treatment-subgroup interactions in clustered datasets}
+\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model (GLMM) Trees}
+\shorttitle{GLMM Trees}
+\leftheader{GLMM Trees}
+\rightheader{GLMM Trees}
 \author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
@@
 In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons.
-In the current paper we propose a new algorithm, GLMMtree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
-The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a linear mixed-effects model for estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMMtree and compare it with that of MOB trees without random-effects estimation.
-In datasets without treatment-subgroup interactions, GLMMtree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively).
-Furthermore, in datasets with treatment-subgroup interactions, GLMMtree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively).
-Also, GLMMtree predicted treatment outcome differences more accurate than MOB trees without random effects (average accuracy of .94 and .88, respectively).
-We illustrate the application of GLMMtree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-We conclude that GLMMtree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss directions for future research.
+In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
+The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters.
+In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation.
+In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively).
+Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively).
+Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively).
+We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.
+
+\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees
+
+%\section{Citation stuff for journal publication}
+%
+%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
+%
+%''Do logistic regression and signal detection identify different subgroups at risk?
+Implications for the design of tailored interventions.
+Kiernan, Michaela; Kraemer, Helena C.; Winkleby, Marilyn A.; King, Abby C.; Taylor, C.
+Barr.
+Psychological Methods, Vol 6(1).
+''
+%
+%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. Psychological methods, 9(4), 426.''
+%
+%''Cooper, H., \& Patall, E.
+A.
+(2009).
+The relative benefits of meta-analysis conducted with individual participant data versus aggregated data.
+Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results.
+They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
+%
+%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''
+%
+%Statistics in Medicine references: \cite{DussyMech14,FostyTayl11,HiggyWhit01,LipkyDmit11}
 
@@
 
-In medicine-efficacy research, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important.
+In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important.
 Personalized medicine presents the challenge of finding which patients respond best to which treatments.
@@
 Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments.
-Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once, and can automatically detect (higher order) interactions between predictor variables.
-Several promising tree-based algorithms and software packages have been developed to assist in the detection of treatment-subgroup interactions (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
-Of these tree-based methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) may be the most flexible in detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates.
+Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables.
+Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
+Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates.
 The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM).
-Earlier, MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.
+Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.
 
@@
 In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}.
-For example, in individual-level patient data meta-analyses, where datasets of multiple trials evaluating the effects of the same treatments are pooled.
+For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled.
 In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}.
@@
 Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}.
-When the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
+More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
 
-In the current paper, we present a tree-based algorithm for treatment-subgroup interaction detection, which takes the clustered nature of datasets into account.
-The algorithm combines MOB with the estimation of random effects, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials).
-In what follows, we first discuss the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning of the GLM, and the generalized linear mixed-effects model (GLMM).
-Then, we present a new algorithm, which combines model-based recursive partitioning and random-effects estimation: GLMMtree.
-In a simulation study, we evaluate the comparative accuracy of the new algorithm.
-Finally, we apply GLMMtree to an existing dataset on the effects of treatments for depression, to illustrate the application of the algorithm.
+In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account.
+The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials).
+In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM).
+Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree.
+
+Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
+After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
+Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.
+
+\subsection{Artificial motivating dataset}
+
+To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 (78 observations) or Treatment 2 (72 observations).
+Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
+Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).
+
+The simulated dataset has 3 subgroups with treatment interactions.
+The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$.
+In this subgroup, the mean of the response variable for Treatment 1 is 7, and the mean for the response variable for Treatment 2 is 11.
+The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$.
+In this subgroup, the mean value of the response variable for Treatment 1 and 2 is 9.
+The third subgroup consists of observations with duration $>6$.
+In this subgroup, the mean value of the response variable for Treatment 1 is 12, and the mean of the response variable for Treatment 2 is 7.
+
+Observations were drawn from one of ten clusters, each with a different, cluster-specific (i..e, random) intercept.
+Data was generated such, that covariates and cluster-specific intercepts were uncorrelated.
+Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.
 
@@
 In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects.
-The goal is to estimate a model for predicting the value of treatment outcome $y$ for observation $i=1,\dots,N$ (an overview of notation used is provided in the Appendix):
+The goal is to estimate a model for predicting the value of a treatment outcome $Y$, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:
+
+\begin{equation}
+\label{eq:expected value}
+	E[y_i | x_i] = \mu_i 
+\end{equation}
 
@@
 \label{eq:fixedeffects}
-	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta
+	g(\mu_{i}) = x_{i}^{\top}\beta 
 \end{equation}
 
-Where $y_i$ is the value of the linear predictor of the response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor and the mean of the response distribution function.
-In case of a continuous response variable, $g$ is the identity function.
-Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type.
-$\beta$ is a vector of fixed-effects regression coefficients, containing the intercept, which is the mean value of the linear predictor in the first treatment group, and the slope, which is the difference in mean values of the linear predictor between the first and second treatment groups.
+Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function.
+In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}$.
+Further, $x_{i}^{\top}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type.
+$\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
 
-To keep notation and examples simple, we assume $x_{i}$ and $\beta$ to have length 2, That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the linear model.
+To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2.
+That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM.
 However, additional treatment conditions and covariates can easily be included.
 
-An example of such a GLM for a continuous outcome variable is graphically represented in Figure \ref{fig:fixedeffects}.
-The example is based on simulated data.
-The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the outcome variable (posttreatment depression score) among 150 participants, who were randomly assigned to treatments 1 and 2.
-Little overall difference between the outcomes of both treatments is suggested, as the slope of the regression line is nearly zero.
-We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by other variables.
-Conditional on other variables (i.e., subgroup indicators), the relationship between $x$ and $y$ may vary in strength and/or direction (c.f., Simpson's paradox; \citeNP{Simp51}).
+As an example, the estimated GLM of the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}.
+The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
+There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero.
+We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.
 
+\vspace{20pt}
 \begin{figure}[!htbp]
-    \includegraphics[width=8cm]{Interaction_ex;treedepth1.pdf}
-	\caption{Example of a linear fixed-effects model for treatment outcomes (N=150).
-The dot for treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
+    \includegraphics[width=6cm]{glm_example.pdf}
+	\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150).
+The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
 	\label{fig:fixedeffects}
@@
 
-The rationale behind MOB is that a global model for all observations, like that in Equation \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
+The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
 This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
-Whereas CART trees have constant fits in the terminal nodes, MOB trees have parametric models with one or more predictor variables in their terminal nodes.
+However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.
 
-To find partitions and better-fitting local linear models, the MOB algorithm tests for parameter instability.
-When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more additional covariates.
-To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subsamples.
+To find partitions and better-fitting local GLMs, the MOB algorithm tests for parameter instability.
+When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
+To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
 
 More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function.
-The expected value of the scores over all observations in a dataset is zero, by definition.
-Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$.
-To statistically test whether there are systematic deviations of the scores from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.
+By definition, the empirical scores of all observations in a dataset sum to zero, and when the model is correctly specified, the expected value of the score for each observation is also zero.
+Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}.
+To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.
 
@@
 In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2).
-The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the residual sum of squares in both partitions \cite{ZeilyHoth08}.
+The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
 In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.
 
-Due to the binary recursive nature of MOB, the resulting partitions can be represented as a tree.
-If the partitioning is based on the GLM, the result is a GLMtree, which has a local fixed-effects regression model in every $j$th (where $j = 1,...,J$) terminal node of the tree.
-As a result, in the GLMtree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ 'falls':
+Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree.
+If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,...,J$) terminal node of the tree.
+As a result, in the GLM-tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
 
@@
 \label{eq:fixedeffects_MOB}
-	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j}
+	g(\mu_{i}) = x_{i}^{\top}\beta_{j}
 \end{equation}
 
-Figure \ref{fig:example_mobtree} provides an example of the GLMtree model in Equation \ref{eq:fixedeffects_MOB}, based on the same data as was used for Figure \ref{fig:fixedeffects}.
-By using four additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline, age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
-The leftmost subgroup in Figure \ref{fig:example_mobtree} represents observations with low duration and low anxiety, for which treatment 1 is much more beneficial then treatment 2 (i.e., lower posttreatment depression scores).
-The rightmost subgroup represents observations with longer duration of depressive symptoms, for whom treatment 2 is much more beneficial than treatment 1.
-The two terminal nodes in the middle indicate that for patients with low duration, treatment 1 is only beneficial to those with moderate levels of anxiety (i.e., a values $>10$ and $\leq 12$).
-Age did not have an effect on treatment outcome, and therefore does not appear as a splitting variable in the tree.
+Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM.
+The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*\top}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
+$\beta^{* \top}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.
+
+Figure \ref{fig:example_mobtree} provides an example of the GLM-tree model in Equation \ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset.
+By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
+Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above.
+However, the two subgroups in the middle result from a spurious split.
 
 \begin{figure}[!h]
-    \includegraphics[width=15cm]{Interaction_ex;treedepth3.pdf}
-    \caption{Example of tree representation of model-based recursive partitions, based on the same data as Figure \ref{fig:fixedeffects}.
+    \includegraphics[width=15cm]{glmtree_example.pdf}
+    \caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset.
 Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
@@
 \subsection{GLMM}
-When a dataset contains observations from multiple clusters (i.e., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} can be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
+When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
 
@@
 \label{eq:mixedeffects}
-	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta + z_{i}b_{m}
+	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b_{m}
 \end{equation}
 
-Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$, of which observation $i$ is part.
-Within the linear mixed-effects model, it is assumed that values of $b_{m}$ are normally distributed, with mean zero and variance $\sigma^{2}_{b_m}$.
+Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$ ($m=1,...,M$), of which observation $i$ is part.
+Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
+The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.
+
 For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models.
-However, random-effects covariates can easily be included.
-The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in for example \cite{BrykyRaud92}.
-Note that for prediction of the treatment outcome for a new observation from a new cluster, only the fixed-effects part of Equation \ref{eq:mixedeffects} would be used.
+However, random-effects covariates and coefficients can easily be included.
+Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the linear predictor of a GLM.
 
-\subsection{Combining model-based recursive partitioning and random-effects estimation}
+\subsection{GLMM tree}
 
 As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects.
-Therefore, we propose the GLMMtree, which combines the GLMM in Equation \ref{eq:mixedeffects} with the GLMtree in Equation \ref{eq:fixedeffects_MOB}:
+Therefore, we propose the GLMM tree, which combines the GLMM from Equation \ref{eq:mixedeffects} with the tree from Equation \ref{eq:fixedeffects_MOB}:
 
@@
 \label{eq:glimmertree}
-	g(\mu_{i}) = g(E(y_{i})) = x_{i}\beta_{j} + z_{i}b_{m}
+	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b_{m}
 \end{equation}
 
-To estimate the partitions, and values of $\beta_j$ and $b_m$ for this model, we take an iterative approach, alternating between (1) assuming the random-effects coefficients $b_m$ known, allowing for partitioning the datasets and estimating the corresponding $\beta_j$ values; and (2) assuming the partition and corresponding $\beta_j$ values known, allowing for estimation of the random-effects coefficients $b_m$.
+To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
+\citeA{HajjyBell11} and \citeA{SelaySimo12} developed a method for estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees.
+In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
+To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.
 
-In Figure \ref{fig:algorithm}, a schematic representation of the GLMMtree algorithm is presented.
+For estimating GLMM trees, we take the MERT approach a step further, by using a GLM tree instead of a regression tree with constant fits.
+This allows not only for detection of differences in main effects,but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
+In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
+The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:
+
+\vspace{5pt}
+\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{m(r)}$ to 0.
+
+\vspace{5pt}
+\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using responses $y_i - z_{i}^{\top}\hat{b}_{m(r-1)}$.
+
+\vspace{5pt}
+\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{m(r)}$ using responses $y_i - x_{i}^{\top}\hat{\beta}_{j(r)}$.
+
+\vspace{5pt}
+\noindent Step 3: Repeat steps 1 and 2 until convergence.
+
+\vspace{5pt}
 The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
-In every iteration, the GLMtree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
-The GLMtree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLMtree from the current iteration.
+In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
+The GLM tree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLM tree from the current iteration.
 Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.
-A similar approach has been taken by \citeA{HajjyBell11} and \citeA{SelaySimo12}, who added random-effects estimation to CART trees with constant fits, instead of linear models, in the terminal nodes.
 
-\begin{figure}[!htbp]
-	\begin{tabular}{ll}
-		\hline 
-		\multicolumn{2}{l}{\textbf{Algorithm} GLMMtree}\\
-		\hline
-		&Step 0: Initialize by setting all values $b_m$ to $0$.\\
+In Figure \ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented.
+As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.
 
-		&Step 1: Given the current $b_m$ values, partition dataset and estimate corresponding values of $\beta_j$.\\
-
-		&Step 2: Given the current partition and corresponding $\beta_j$ values, estimate $b_m$ values. \\
-
-		&Step 3: Repeat steps 1 and 2 until convergence.\\
-		\hline
-	\end{tabular}
-	\caption{Description of the GLMMtree algorithm}
-	\label{fig:algorithm}
+\begin{figure}[!h]
+    \includegraphics[width=12cm]{glimmertree_example.pdf}
+    \caption{Generalized Linear Mixed Model tree of the motivating example dataset.
+Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
+    \label{fig:example_glimmertree}
 \end{figure}
 
-In what follows, we present a simulation study in which we assess the performance of GLMMtree in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments, for continuous outcomes.
-In addition, we will compare the performance of GLMMtree with that of GLMtree.
+\section{Simulation}
+
+We will asses the performance of GLMM tree in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments, in simulated datasets with continuous outcomes.
+In addition, we will compare the performance of GLMM tree with that of GLM tree.
 In the simulation study, the main interest will be in the effects of sample size, and the presence and magnitude of treatment-subgroup interactions and random effects, but other parameters will be varied, as well.
 
-For GLMMtree, we expect the accuracy of recovered trees and predictions to improve with increasing sample size, and magnitude of the differences in treatment outcomes.
-For GLMtree, we have the same expectation, when random effects are absent; that is, when the variance of the random coefficients is zero, we expect GLMtree and GLMMtree to perform equally well.
-When random effects are present, we expect GLMMtree to perform better than GLMtree, and more so when the variance of random-effects coefficients increases.
-
-\section{Simulation: Method}
-
-\subsection{Software}
-
-\verb|R| \cite{R14} was used for generation and analysis of all datasets.
-Two additional R packages were used: \verb|partykit| \cite{HothyZeil14,ZeilyHoth08} for the estimation of GLMtrees, and \verb|lme4| \cite{BateyMeac12} for the estimation of random-effects coefficients.
-For all functions, default settings were used, with exception of maximum tree depth.
-Maximum tree depth was set to four (i.e., maximum of eight terminal nodes) for all trees, as this yields a model and graphical representation which is easy to interpret.
-
-\subsubsection{Estimation of GLM- and GLMMtrees}
-
-For estimating GLMtrees, the \verb|glmtree| function from the \verb|partykit| package was used.
-The \verb|glmtree| function builds a generalized linear model tree: a model-based recursive partition based on a GLM (Equation \ref{eq:fixedeffects_MOB}).
-
-For estimating GLMMtrees, we implemented the algorithm as described in Figure \ref{fig:algorithm} in a function that iterates between (1) estimation of a linear model tree using the \verb|glmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|glmer| function.
-Convergence of \verb|GLMMtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2).
-When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|GLMMtree| has converged.
+For GLMM tree, we expect the accuracy of recovered trees and predictions to improve with increasing sample size, and magnitude of the differences in treatment outcomes.
+For GLM tree, we have the same expectation, when random effects are absent; that is, when the variance of the random coefficients is zero, we expect GLM tree and GLMM tree to perform equally well.
+When random effects are present, we expect GLMM tree to perform better than GLM tree, and more so when the variance of random-effects coefficients is larger.
 
@@
     \end{tikzpicture}
-	\caption{data-generating model for treatment-subgroup interactions. $d$ denotes the standardized mean difference between the outcomes of treatment 1 and 2 (i.e., $\beta_j1 / \sigma_{\epsilon}$).}
+	\caption{Data-generating model for treatment-subgroup interactions.
+Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
 	\label{fig:modelC}
@@
 For generating datasets without treatment-subgroup interactions, we used a design in which there is only a main effect of treatment in the population.
-In these datasets, the mean of the outcome variable was 30, and $d$ and the first and second elements of $\beta$ have the same value for all observations (i.e., when $d = 1.0$, the first and second elements of $\beta$ are 37.5 and 42.5, respectively).
+Put differently, the number of subgroups or terminal nodes in these datasets was $J=1$, and there was only a single value of $\beta_j = \beta$ in every dataset.
+The mean of the outcome variable in the datasets without treatment-subgroup interactions was 30, which is the same value as in the datasets with treatment-subgroup interactions.
+As a result, $\beta = (27.5, 32.5)$ for all observations when $d=1$.
 
@@
 
-In generating the datasets, we varied seven parameters of the data-generating process:
+In generating datasets, we varied seven parameters of the data-generating process:
 
@@
 \item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
-\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b_m}=0$, $\sigma_{b_m}=5$, $\sigma_{b_m}=10$.
-\item Three levels for the intercorrelations between $b_m$ and one of the $U_k$ variables: $b_m$ and $U_k$ uncorrelated, $b_m$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b_m$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42).
-\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for treatment 1 ($X_{1}=0$) and treatment 2 ($X_{1}=1$).
+\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
+\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42).
+\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for Treatment 1 ($x_{1}=0$) and Treatment 2 ($x_{1}=1$).
 The levels for mean differences in subgroups with differential treatment effect were $|\beta_1| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_1| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
@@
 For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in 50 x 3 x 2 x 2 x 3 x 3 x 3 x 2 = 32,400 training datasets.
-For the datasets with main treatment effect only, the 6th parameter of the data-generating process had only two levels ($b_m$ correlated with one of the $U_k$ variables, and $b_m$ not correlated with any of the $U_k$ variables).
+For the datasets without treatment-subgroup interactions, the 6th parameter of the data-generating process had only two levels ($b$ correlated with one of the $U_k$ variables, and $b$ not correlated with any of the $U_k$ variables).
 Therefore, 50 x 3 x 2 x 2 x 3 x 3 x 2 x 2 = 21,600 datasets without treatment-subgroup interactions were generated.
@@
 
-As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{M}$ were drawn from a multivariate normal distribution with ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively.
+As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively.
 The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
 All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$.
-Correlations between the variables in $U$ vary according to the third facet of the simulation design described above.
+Correlations between the $U_k$ variables vary according to the third facet of the simulation design described above.
 
@@
 
-To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
-For each cluster we drew a single $b_m$ from a normal distribution with $\mu_{b_m} = 0$ and the value of $\sigma_{b_m}$ given by the fifth facet of the simulation design.
-When $b_m$ was correlated with one of the potential partitioning variables, the partitioning or non-partitioning covariate correlated with $b_m$ was randomly selected.
+To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
+For each cluster we drew a single $b_m$ from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
+When $b$ was correlated with one of the potential partitioning variables, the correlated potential partitioning variable was randomly selected.
 
-To generate the node-specific fixed-effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3.
+To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3.
 In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$.
@@
 
-Finally, outcome variable $y$ was calculated according to the model in Equation \ref{eq:fixedeffects}.
+Finally, the response variable was calculated as the sum of the (node-specific) fixed effects, random effects and the error term: $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.
 
@@
 
-For every dataset, the accuracy and size of the GLM- and GLMMtrees were evaluated.
+For every dataset, accuracy and size of the GLM and GLMM tree was evaluated.
 We calculated the total number of nodes in every tree, and compared it with the true tree size.
-For datasets with a main treatment effect only, this allowed us to assess the accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned.
+For datasets without treatment-subgroup interactions, this allowed us to assess tree accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned.
 For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups.
 
-For datasets with treatment-subgroup interactions, we assessed the accuracy of GLMtree and GLMMtree in recovering the true tree.
-An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., total number of nodes is 7), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
-Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation ($\sigma_{U_k}$) of the partitioning variable.
+For datasets with treatment-subgroup interactions, we assessed the accuracy of the trees created by GLM and GLMM tree.
+An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., total number of nodes equals 7), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
+Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).
 
-To detect predictors of the performance of both algorithms, GLMtrees were built using the \verb|glmtree| function.
-Two trees were built for predicting the number of nodes in the trees generated by both algorithms: one for datasets without, and one for datasets with treatment-subgroup interactions.
-In these trees, the outcome variable is the number of nodes in a tree, the predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process, described above.
-In addition, for datasets with treatment-subgroup interactions, a GLMtree was built for predicting the probability that the tree of treatment-subgroup interactions was accurately recovered.
-In this tree, the outcome variable is a binary indicator for whether the tree was accurately recovered (or not).
-Again, the predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process.
+To detect predictors of tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
+The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.
+
+To detect predictors of tree accuracy for both algorithms in datasets with treatment-subgroup, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
+In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
+The effects of predictors with main and/or interaction effects with unstandardized coefficients $|\beta| > .2$ were further investigated using graphical displays.
 
@@
 
-We evaluated predictive accuracy of GLM- and GLMMtrees by calculating correlations between the predicted and true treatment-effect differences ($\beta_{j1}$, Figure \ref{fig:modelC}) for test observations.
-Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets with a main treatment effect only.
+We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure \ref{fig:modelC}) for test observations.
+Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
 
 Using the same data for training and evaluation of a model results in overly optimistic estimates of predictive accuracy \cite{HastyTibs09}.
-Therefore, GLM- and GLMMtrees were used for prediction of new observations in test datasets.
-These test datasets were generated from the same population as the training datasets.
+Therefore, GLM and GLMM trees were used for prediction of new observations from test datasets.
+Test datasets were generated from the same population as the training datasets.
 Because the cluster-specific intercepts $b_m$ were randomly generated for training as well as test datasets, test observations were from 'new' clusters.
-As a result, a model without random effects was used for prediction with GLMMtree.
+As a result, a model without random effects was used for prediction with GLMM tree.
 
-For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment differences: one for GLMtree, and one for GLMMtree.
-To detect predictors of predictive accuracy, a GLMtree was built.
-The outcome variable in this tree is the correlation between the true and predicted treatment differences.
-The predictor variable for the linear model is algorithm type (GLMtree or GLMMtree), and the (potential) partitioning variables are the parameters of the data-generating process, described above.
+For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment-effect differences: one for GLM tree, and one for GLMM tree.
+To detect predictors of predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
+The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.
 
-\section{Simulation: Results}
+\subsection{Software}
 
-\subsection{Tree size and accuracy in datasets with main treatment effect only}
+\verb|R| \cite{R14} was used for generation and analysis of all datasets.
+Two additional R packages were used: \verb|partykit| (version 0.8-3; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
 
-In Table \ref{tab:treesize_main}, tree sizes for GLMtree and GLMMtree are presented for datasets with a main treatment effect only.
-Overall, smaller trees were created by GLMMtree: the average tree size was 1.09 (SD=0.44) for GLMMtree, and 2.02 (SD=1.68) for GLMtree.
-The estimated probability that a dataset was erroneously partitioned was very small for GLMMtree (.04; Table \ref{tab:treesize_main}), and much larger for GLMtree (.33; Table \ref{tab:treesize_main}).
+\subsubsection{Estimation of GLM and GLMM trees}
+
+For estimating GLM trees, the \verb|lmtree| function from the \verb|partykit| package was used.
+The \verb|lmtree| function builds a linear model tree: a GLM-based recursive partition (Equation \ref{eq:fixedeffects_MOB}) for a real-valued response variable with normally distributed errors.
+Models in the nodes of the tree are estimated with ordinary least squares (OLS).
+The $\alpha$-level for assessing parameter instability was set to .05, a Bonferroni correction for multiple testing was applied, and the minimum number of observations in a node was set to 20.
+Maximum tree depth was set to four (i.e., maximum of eight terminal nodes), as this yields a model tree which is easy to interpret.
+
+\subsubsection{Estimation of GLMM trees}
+
+For estimating GLMM trees, we implemented the GLMM tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
+Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm.
+When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|glmmtree| has converged.
+
+For building the GLM trees in step (1) of the GLMM tree algorithm, the same settings as described above were used.
+That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.
+
+For estimating the $b_m$ values in step (2), the \verb|lmer| function from the \verb|lme4| package was used.
+This function estimates a linear mixed-effects model using maximum likelihood (ML) or restricted ML (REML).
+In the current paper, REML estimation was used.
+
+\subsection{Tree size and accuracy in datasets without treatment-subgroup interactions}
+
+In Table \ref{tab:treesize_main}, tree sizes for GLM and GLMM trees for datasets without treatment-subgroup interactions are presented.
+Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 (SD=0.44) for GLMM tree, and 2.02 (SD=1.68) for GLM tree.
+The estimated probability that a dataset was erroneously partitioned was very small for GLMM tree (.04; Table \ref{tab:treesize_main}), and much larger for GLM tree (.33; Table \ref{tab:treesize_main}).
 
 \begin{table}[!htbp]
-\caption{Tree size distributions for GLMtree and GLMMtree for datasets with a main treatment effect only.}
+\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
 \small
@@
 	\hline
-	GLMMtree	& 20625 & 932 & 43 &0&0&0& 21,600\\
+	GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
 		  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
-   	GLMtree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
+   	GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
     			& (.67)  & (.20) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
@@
   	\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
-  	\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $k$ nodes has has $(k+1)/2$ }\\
-  	\multicolumn{8}{l}{terminal nodes ; the true tree size in datasets with a main treatment effect}\\	  	\multicolumn{8}{l}{only was 1.}\\
+  	\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
+  	\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
 \end{tabular}
@@
 
-A linear model tree (Figure \ref{fig:treesizetree_nointeract}) indicated that the main predictors of tree size in datasets without treatment-subgroup interactions were sample size and magnitude of $\sigma_{b_m}$.
-When random effects were absent (i.e., $\sigma_{b_m}=0$), both GLMtree and GLMMtree tend to create trees of size 1.
-In the presence of random effects, GLMMtree also tends to create trees of size 1, but GLMtree created larger trees.
-For GLMtree, tree size increased with both sample size, and magnitude of $\sigma_{b_m}$.
+\begin{figure}[!htbp]
+	\includegraphics[width=12cm]{xy_treesizes_maineff.pdf}
+	\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions.
+Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.
+Reference line at $y=1$ represents the true tree size.}
+	\label{fig:xyplot_treesize_nointeract}
+\end{figure}
 
-\begin{sidewaysfigure}[!htbp]
-	\includegraphics[width=20cm]{treesize_mob_maxdepth=4_nointeract.pdf}
-	\caption{Linear model tree of the size of GLMM- and GLMtrees in datasets without treatment-subgroup interactions.
-The y-axes of the terminal nodes represent tree size (total number of nodes in a tree).
-Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$).
-N = total sample size; sigmabm = $\sigma_{b_m}$.}
-	\label{fig:treesizetree_nointeract}
-\end{sidewaysfigure}
+A graphical display was used to asses the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_nointeract}).
+When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1.
+In the presence of random effects, GLMM tree also tends to create trees of size 1, but GLM tree created much larger trees, when $b$ was correlated to one of the $U_k$ variables.
+This effect was stronger when sample size was larger.
 
@@
 
-In datasets with treatment-subgroup interactions, GLMMtrees were also smaller than GLMtrees.
+In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees.
 For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}).
-The sizes of the GLM- and GLMMtrees for datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}.
-The average size of GLMMtrees was 7.15 (SD=0.61), and the average size of GLMtrees was 8.11 (SD=2.05).
-The estimated probability that a datasets was erroneously not partitioned was 0, for both GLM- and GLMMtree.
-However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMMtrees matched the true tree size, whereas a proportion of only .64 of GLMtrees matched the true tree size (Table \ref{tab:treesize_interactions}).
+Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}.
+The average size of GLMM trees was 7.15 (SD=0.61), and the average size of GLM trees was 8.11 (SD=2.05).
+The estimated probability that a datasets was erroneously not partitioned was 0, for both GLM and GLMM tree.
+However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .64 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).
 
 \begin{table}[!htbp]
-\caption{Tree size distributions for GLMtree and GLMMtree for datasets with treatment-subgroup interactions.}
+\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
 \small
@@
 	\hline
-  	GLMMtree	& 3 & 227& 29556 & 2472 & 89 & 3 &0& 32,400 \\
+  	GLMM tree	& 3 & 227& 29556 & 2472 & 89 & 3 &0& 32,400 \\
 				&  ($<.01$)  & ($<.01$)& (.91) & ($<.01$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\
 
-  	GLMtree		& 145 & 1002 & 20578 & 4443 & 4178 & 1665 & 389 & 32,400\\ 
+  	GLM tree		& 145 & 1002 & 20578 & 4443 & 4178 & 1665 & 389 & 32,400\\ 
  			 	& ($<.01$) & (.03) & (.64) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
@@
   	\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
-  	\multicolumn{9}{l}{nodes in the tree. A tree with a total of $k$ nodes has has $(k+1)/2$ terminal nodes; the true}\\	
+  	\multicolumn{9}{l}{nodes in the tree. A tree with a total of $J$ nodes has $(J+1)/2$ terminal nodes; the true}\\	
   	\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
@@
 
-A linear model tree (Figure \ref{fig:treesizetree_interact}) indicated a three-way interaction between sample size, value of $\sigma_{b_m}$, and whether values of $b_m$ are correlated with a potential partitioning variable.
-Overall, both algorithms created trees of size $\approx 7$, with the following two exceptions:
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_interact}).
+When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees of size 7, on average.
 
-\begin{sidewaysfigure}[!htbp]
-	\includegraphics[width=20cm]{treesize_mob_maxdepth=4.pdf}
-	\caption{Linear model tree of tree sizes of GLMM- and GLMtrees in datasets with treatment-subgroup interactions.
-The y-axes of the terminal nodes represent tree size (total number of nodes in a tree).
-Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$).
-N = total sample size; sigmabm = $\sigma_{b_m}$; corUbi = correlation between random intercepts and one of th e potential partitioning variables; np = number of potential partitioning variables.}
-	\label{fig:treesizetree_interact}
-\end{sidewaysfigure}
+Clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$.
+When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average.
+When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$).
+When $b$ is correlated with one of the $U_k$ variables, GLM starts creating spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
 
-When sample size is small (i.e., $N=200$), the variance of $b_m$ is large (i.e., $\sigma_{b_m} = 10$), and values of $b_m$ are not correlated with one of the $U_k$ variables (Figure \ref{fig:treesizetree_interact}).
-This was the case for 2,400 datasets (7.41\%), in which mean tree size was 6.12 (SD=1.37) for GLMtree and 7.01 (SD=0.48) for GLMMtree.
-Because in these cases, GLMtree cannot account for the variance in $y$ due to cluster-specific effects, it has difficulty detecting partitions when $\sigma_{b_m}$ is large and sample size is low.
-However, the large variability for GLMtree indicates that it may still create spurious splits.
-
-When sample size is larger (i.e., $N=500$ or $N=1000$), when the variance of $b_m$ is non-zero, and values of $b_m$ are not correlated with one of the $U_k$ variables (Figure \ref{fig:treesizetree_interact}).
-This was the case for 19,200 datasets (59.26\%), in which mean tree size was 10.53 (SD=2.04) for GLMtree, and 7.23 (SD=0.66) for GLMMtree.
-Obviously, in these cases, GLMtree is more likely to create spurious splits.
-Because GLMMtree can more adequately deal with the additional variance caused by non-zero values of $\sigma_{b_m}$, the size of GLMMtrees seems not to be influenced much by values of $\sigma_{b_m}$.
+\begin{figure}[!htbp]
+	\includegraphics[width=12cm]{xy_treesizes_treatsubs.pdf}
+	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions.
+Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.
+Reference line at $y=7$ represents true tree size.}
+	\label{fig:xyplot_treesize_interact}
+\end{figure}
 
@@
 
-To assess accuracy, we inspected the partitioning variables and values selected in every dataset.
-For the first split, both GLMtree and GLMMtree always selected the true partitioning variable ($U_2$).
-The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94 for both GLMMtree and GMLtree.
-However, GLMtree showed somewhat higher variability in recovering the splitting value than GLMMtree (SD=0.155 and SD=0.126, respectively).
+To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset.
+For the first split, both GLM tree and GLMM tree always selected the true partitioning variable ($U_2$).
+The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94, for both GLM and GLMM tree.
+However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split, than did GLMM tree (SD=0.155 and SD=0.126, respectively).
 
-GLMMtree performed well in recovering treatment-subgroup interactions, accurately recovering the tree in 90.19\% of datasets.
-GLMtree accurately recovered the treatment-subgroup interactions in only 61.44\% of datasets.
-A generalized linear model tree (Figure \ref{fig:treeaccuracytree}) indicated that performance of both algorithms was predicted by a three-way interaction between sample size, value of $\sigma_{b_m}$, and whether values of $b_m$ are correlated with a potential partitioning variable.
-Overall, both algorithms recovered the true tree with about equal probabilities (i.e., all probabilities $>.85$), with three notable exceptions:
+Overall, GLMM tree performed well in recovering treatment-subgroup interactions, accurately recovering the tree in 90.19\% of datasets.
+GLM tree accurately recovered the treatment-subgroup interactions in only 61.44\% of datasets.
 
-When $\sigma_{b_m} > 0$, and values of $b_m$ are correlated with a potential partitioning variable, GLMMtree clearly outperformed GLMtree.
-When sample size was low (i.e., $N=200$), the probability of accurately recovering the tree was .59 for GLMtree, and .91 for GLMMtree (Figure \ref{fig:treeaccuracytree}).
-With larger sample sizes (i.e., $N=500$ or $1,000$), the probability of accurately recovering the tree was .13 for GLMtree, and .89 for GLMMtree (Figure \ref{fig:treeaccuracytree}).
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure \ref{fig:xyplot_treeaccuracy}).
+When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average.
+In the presence of random effects, GLM trees were much less accurate than GLMM trees.
+This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables.
+When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).
 
-When $\sigma_{b_m}$ was non-zero, and values of $b_m$ are not correlated with a potential partitioning variable, GLMMtree also outperformed GLMtree, but only when sample size was small.
-In these datasets, the probability of accurately recovering the tree was .72 for GLMtree and .92 for GLMMtree (Figure \ref{fig:treeaccuracytree}).
-When sample size was larger (i.e., $N=500$ or $1,000$), the probability of accurately recovering the tree was .90 for both algorithms (Figure \ref{fig:treeaccuracytree}).
-
-\begin{sidewaysfigure}[!htbp]
-	\includegraphics[width=20cm]{truetree_mob_maxdepth=4.pdf}
-	\caption{Generalized linear model tree of accurate recovery of treatment-subgroup interactions by GLMM- and GLMtree in datasets with treatment-subgroup interactions.
-Bar plots in the terminal nodes represent the proportion of datasets in which the treatment-subgroup interactions were accurately recovered.
-Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$).
-N = total sample size; sigmabm = $\sigma_{b_m}$; corUbi = correlation between random intercepts and one of th e potential partitioning variables; rho = correlation between potential partitioning variables; treatdiff = $\beta_{j1}$, the unstandardized mean difference in treatment outcomes in subgroups with differential effects for treatment 1 and treatment 2.}
-	\label{fig:treeaccuracytree}
-\end{sidewaysfigure}
+\begin{figure}[!htbp]
+	\includegraphics[width=12cm]{xy_accuracy.pdf}
+	\caption{Average accuracy of GLM and GLMM trees.
+Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered.
+Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
+	\label{fig:xyplot_treeaccuracy}
+\end{figure}
 
@@
 
-Overall, the treatment differences predicted by GLMMtree were closer to the true differences than the treatment differences predicted by GLMtree.
-The average correlation between the true and predicted treatment differences over all 32,400 datasets was .88 (SD=0.20) for GLMtree, and .94 (SD=0.10) for GLMMtree.
+To assess predictive accuracy of both algorithms, correlation between the true and predicted treatment-effect differences of both algorithms were calculated for every dataset.
+Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree.
+The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.20) for GLM tree, and .94 (SD=0.10) for GLMM tree.
 
-The linear model tree depicting the relationship between the various data-generating parameters and predictive accuracy of GLMtree and GLMMtree, is presented in Figure \ref{fig:correlationtree}.
-Figure \ref{fig:correlationtree} shows clear main effects of sample size $N$, treatment difference size, and value of $\sigma_{b_m}$.
-With larger sample sizes ($N = 500$ or $1,000$), both GLMtree and GLMMtree perform better than with smaller sample sizes ($N = 200$).
-With larger treatment differences ($d=5$), the difference between the performance of GLMtree and GLMMtree becomes smaller, and the performance of GLMtree shows less variation.
-When the random intercepts are sampled from a population distribution with larger variance (i.e., $\sigma_{b_m} = 10$), the difference in accuracy between GLMMtree and GLMtree is more pronounced, than when there are no random intercept differences, or when these differences are small (i.e., $\sigma_{b_m} = $ 0 or 5).
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the predictive accuracy of both algorithms (Figure \ref{fig:xyplot_correlations}).
+Both algorithms showed higher predictive accuracy when sample size was larger, and when treatment-effect differences were larger.
+When random effects were absent from the datasets (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were equally accurate.
+In the presence of random effects, GLM tree predictions were always much less accurate than those of GLMM tree.
+This effect was stronger when $\sigma_b$ was larger, sample size was larger, and/or treatment-effect differences were larger.
 
-As the boxplots in Figure \ref{fig:correlationtree} show, for some simulated datasets, correlations between predicted and true differences were obtained that were clear outliers.
-It should be noted that low or negative correlations between the true and predicted treatment differences were much more often found for GLMtree than for GLMMtree: a correlation $<.40$ was obtained in 930 out of 32,400 datasets for GLMtree, and in 175 out of 32,400 datasets for GLMMtree.
-
-\begin{sidewaysfigure}[!htbp]
-	\includegraphics[width=20cm]{correlation_mob_maxdepth=4.pdf}
-	\caption{Linear model tree of correlations with true treatment differences for GLMMtrees and GLMtrees.
-The y-axes in the terminal nodes represent the correlation between the true and predicted treatment effect differences.
-Circles represent outliers (values below $Q_1 - 1.5 \times IQR$ or above $Q3 + 1.5 \times IQR$).
-N = total sample size; treatdiff = $\beta_{j1}$, the unstandardized mean difference in treatment outcomes in subgroups with differential effects for treatment 1 and treatment 2; sigmabm = $\sigma_{b_m}$.}
-	\label{fig:correlationtree}
-\end{sidewaysfigure}
+\begin{figure}[!htbp]
+	\includegraphics[width=12cm]{xy_correlations.pdf}
+	\caption{Average predictive accuracy of GLM and GLMM trees.
+Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment 1 and 2.
+Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
+	\label{fig:xyplot_correlations}
+\end{figure}
 
@@
 \subsection{Method}
-To illustrate the application and potential differences in the results of GLMtree and GLMMtree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
-This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavior therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
+To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
+This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
 The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression.
@@
 In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
-The linear predictor was treatment type (0=CBT and 1=PHA).
+The predictor variable in the linear model was treatment type (0=CBT and 1=PHA).
 An indicator for study was used as the cluster indicator.
@@
 An offset variable is a linear predictor with an a-priori determined coefficient of one.
-Including the linear regression predictions as an offset has the same effect as statistically controlling for the linear effects of a of pretreatment scores, as is often done in ANCOVA.
+Including the linear regression predictions as an offset has the same effect as statistically controlling for the linear effects of pretreatment scores, as is often done in ANCOVA.
 
-The \verb|glmtree| function deals with missing data by listwise deletion.
-Therefore, we build the trees using data of a subset of 694 patients from 7 studies, as complete observations (i.e., observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score) for these patients were available.
+The \verb|lmtree| function deals with missing data by listwise deletion.
+Therefore, we build all trees using data of a subset of 694 patients from 7 studies, as complete observations (i.e., observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score) for these patients were available.
 Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \cite{CuijyWeit14}.
 
-Predictive accuracy of GLMtree and GLMMtree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
+Predictive accuracy of GLM and GLMM tree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
 
@@
 
-We applied GLMtree and GLMMtree to the dataset with complete observations.
-The resulting trees are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}.
-Note that the GLMtree in Figure \ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMMtree algorithm.
+The trees resulting from appplication of GLM and GLMM tree to the dataset are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}, respectively.
+Note that the GLM tree in Figure \ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMM-tree algorithm.
 
@@
     \includegraphics[width=12cm]{app_lmtree.pdf}
-	\caption{GLMtree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+	\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
 The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.
@@
 
-The GLMtree (Figure \ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-Terminal node 2 of Figure \ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
-Terminal node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and antidepressant mediation.
-Terminal node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for pharmacotherapy.
+The GLM tree (Figure \ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
+Node 2 of Figure \ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
+Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and antidepressant mediation.
+Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for pharmacotherapy.
 
@@
     \includegraphics[width=10cm]{app_lmertree.pdf}
-	\caption{GLMMtree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+	\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
 The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.
@@
 
-By taking into account the study-specific intercepts, the final GLMMtree (Figure \ref{fig:lmertree_C&W}) indicates that the first split made by GLMtree is a spurious split.
-The GLMMtree selected only presence of a comorbid anxiety disorder as a partitioning variable.
-The terminal nodes of Figure \ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores than CBT.
+By taking into account the study-specific intercepts, the final GLMM tree (Figure \ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree is a spurious split.
+The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable.
+The terminal nodes of Figure \ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores.
 The estimated variance of the random intercept term was 2.12, with an estimated intraclass correlation coefficient of .05.
 
-Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMMtree had higher predictive accuracy than the GLMtree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .39 (SD=.20) for GLMMtree, and .31 (SD=.24) for GLMtree.
-This indicates that GLMMtree not only provided higher predictive accuracy, on average, but also had somewhat lower variability of predictive accuracy than GLMtree.
+Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
+The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .39 (SD=.20) for GLMM tree, and .31 (SD=.24) for GLM tree.
+This indicates that GLMM tree not only provided higher predictive accuracy, on average, but also had somewhat lower variability of predictive accuracy than GLM tree.
 
@@
 
-The results of our simulation study show that GLMMtree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions.
-In the absence of treatment-subgroup interactions, GLMMtree erroneously detected subgroups in only 4\% of the datasets.
-GLMtree performed less accurate than GLMMtree: in datasets with treatment-subgroup interactions, GLMtree recovered the true tree structure in 61\% of the simulated datasets.
-In datasets without treatment-subgroup interactions, GLMtree erroneously detected subgroups in 33\% of the datasets.
+The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions.
+In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets.
+GLM tree performed less accurate than GLMM tree: in datasets with treatment-subgroup interactions, GLM tree recovered the true tree structure in 61\% of the simulated datasets.
+In datasets without treatment-subgroup interactions, GLM tree erroneously detected subgroups in 33\% of the datasets.
 
-The better performance of GLMMtree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables.
-In these instances, the random effects gave rise to spurious subgroup detection (spurious splits) by GLMtree, both in datasets with and without treatment-subgroup interactions.
+The better performance of GLMM tree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables.
+In these instances, the random effects gave rise to spurious subgroup detection (spurious splits) by GLM tree, both in datasets with and without treatment-subgroup interactions.
 
-Also, predictive accuracy of GLMMtree was higher than that of GLMtree.
-The average correlation between the true treatment differences and those predicted by GLMMtree was .94.
-The average correlation between the true treatment differences and those predicted by GLMtree was .88.
-In terms of predictive accuracy, GLMMtree clearly outperformed GLMtree when random effects in the datasets were sizable, and random intercepts were correlated to potential partitioning variables.
+Also, predictive accuracy of GLMM tree was higher than that of GLM tree.
+The average correlation between the true treatment differences and those predicted by GLMM tree was .94.
+The average correlation between the true treatment differences and those predicted by GLM tree was .88.
+In terms of predictive accuracy, GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, and the differences in treatment effects were relatively small (i.e., $d=.5$).
 
-As expected, when random effects were absent from the simulated datasets, GLMtree and GLMMtree showed high and equal predictive accuracy.
-This finding indicates that GLMMtree can be applied, whenever cluster-specific random effects are expected.
-In the absence of random effects, GLMtree and GLMMtree are expected to perform equally well, and in the presence of random effects, GLMMtree will outperform GLMtree.
-This is especially the case with large sample sizes ($N > 500$), as the increased power will likely cause GLMtree to create spurious splits in the presence of random effects.
+As expected, when random effects were absent from the simulated datasets, GLM tree and GLMM tree showed high and equal predictive accuracy.
+This finding indicates that GLMM tree can be applied, whenever cluster-specific random effects are expected.
+In the absence of random effects, GLM tree and GLMM tree are expected to perform equally well, and in the presence of random effects, GLMM tree will outperform GLM tree.
+This is especially the case with large sample sizes ($N > 200$), as the increased power will likely cause GLM tree to create spurious splits in the presence of random effects.
 
-Not surprisingly, for both algorithms, accuracy of predicted treatment differences was somewhat less when sample size was low (i.e., $N=200$).
-Sample size influenced performance of GLMtree and GLMMtree similarly, suggesting that the larger number of estimated parameters for GLMMtree did adversely influence accuracy with low sample sizes.
-However, our simulation results do warrant caution for the detection of treatment-subgroup interactions or treatment moderators in small datasets (e.g., single RCTs), irrespective of the algorithm used.
+Not surprisingly, for both algorithms, accuracy of predicted treatment differences was less when sample size was low (i.e., $N=200$).
+Sample size influenced performance of GLM tree and GLMM tree similarly, suggesting that a larger number of estimated parameters for GLMM tree does not adversely influences accuracy at low sample sizes.
+Our simulation results do warrant some caution for the detection of treatment-subgroup interactions or treatment moderators in small datasets (e.g., single RCTs), but irrespective of the algorithm used.
 
-Although these findings are encouraging for the use of GLMMtree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations of our study and challenges for future research should be noted.
+Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations and challenges for future research should be noted.
 
-As noted earlier, simulations in the current study were confined to random-intercept models.
-GLMMtree allows for the estimation of random slopes as well, but estimation of random treatment effects is currently not possible, as treatment effects are estimated with local linear fixed-effects models.
-Our simulations also did not include models with multiple fixed-effects predictor variables in $X$.
-Multiple fixed-effects predictor variables can be easily included in GLMtree and GLMMtree, but it should be noted that the parameters corresponding to these variables will then be included in tests for parameter instability as well, which may be undesirable.
-For example, in RCTs, ANCOVA will often be used to control for the linear effects of pretreatment values on the treatment outcome variable.
-Whether or not such parameters should be included in parameter stability tests, or should be allowed to vary over partitions, should be decided by the researcher.
+The simulations show that GLMM tree performs very well, if the model is correctly specified.
+That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then the algorithm will accurately recover those subgroups.
+However, misspecification of the model can reduce performance.
+One source of misspecification would be, when relevant variables are not included in the GLM or as partitioning variables.
+If there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using the partitioning variables that are available.
+Or, if the coefficients of other variables vary across subgroups, then those variables should also be included in the GLM.
+Another source of misspecification would be the inclusion of irrelevant variables in the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
+However, it should be note that in our simulations, the number of partitioning variables did not substantially influence performance of the algorithm.
 
-One challenge for further research is the development of parameter stability tests for random-effects parameters.
-In the current study, random-effects parameters were estimated globally, using all observations in the dataset, and fixed-effects parameters were estimated locally, using the observations in a single node.
-This would allow, for example, for estimation of random treatment effects, instead of fixed treatment effects.
+A challenge for future research is the development of more adequate ways to deal with missing data.
+GLM tree, and therefore also GLMM tree, handle missing data by listwise deletion, like most tree-based algorithms for treatment-subgroup interaction detection.
+However, missing data commonly occurs in clinical trails, and listwise deletion is not an optimal approach for dealing with missing data.
 
-A second challenge is the development of more adequate ways to deal with missing data in treatment-subgroup interaction detection.
-GLMtree, like most tree-based algorithms for treatment-subgroup interaction detection, handles missing data by listwise deletion.
-However, missing data commonly occurs in clinical trails, and listwise deletion is obviously not the preferred method for dealing with missing data \cite<e.g., >{WoodyWhit04}.
-
-In conclusion, GLMMtree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatments differences in the presence and absence of cluster-specific random effects.
-Therefore, GLMMtree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.
+In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects.
+Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.
 
@@
 
-\begin{table}[!h]
-\begin{tabular}{ll}
-$1,...,i,...,N$ &   observation number \\
-$1,...,j,...,J$ &	terminal node number in a tree \\
-$1,...,k,...,K$ &   partitioning variable number \\
-$1,...,m,...,M$	&	cluster number \\
-$\beta_{j}$		& 	column vector of fixed-effects coefficients in terminal node $j$ \\
-$b_{m}$ 		&	column vector of random-effects coefficients in cluster $m$\\
-$d_j$			&	$\beta_{j1} / \sigma_{\epsilon}$; effect size of predicted differences in treatment outcome $y$ between \\
-				&	treatments 1 and 2, in terminal node $j$\\
-$\epsilon$		& 	deviation of observed treatment outcome $y$ from its expected value\\
-$\sigma_{b_m}$	&	square root of variance of $b_m$\\
-$\sigma_{\epsilon}$&square root of the variance of $\epsilon$\\
-$U$ 			&	(potential) partitioning variables \\
-$U_k$			&	(potential) partitioning variable $k$ \\
-$x_i$ 			&	row vector of fixed-effects predictor variable values for observation $i$\\
-$y_i$ 			&	treatment outcome for observation $i$ \\
-$z_i$ 			&	row vector of random-effects predictor variable values for observation $i$\\
-\end{tabular}
-\end{table}
+\NumTabs{12}
+\noindent $1,...,i,...,N$ \tab{observation number}\\
+$1,...,j,...,J$ \tab{terminal node number in a tree}\\
+$1,...,k,...,K$ \tab{partitioning variable number}\\
+$1,...,m,...,M$	\tab{cluster number}\\
+$\beta_{j}$		\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
+$b_{m}$ 		\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
+$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment 1 and}\\
+\tab\tab\tab\tab{Treatment 2 in terminal node $j$}\\
+$\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
+$\sigma_{b}$	\tab\tab{square root of variance of $b$}\\
+$\sigma_{\epsilon}$\tab\tab{square root of the variance of $\epsilon$}\\
+$U_k$			\tab\tab{(potential) partitioning variable $k$}\\
+$x_i$ 			\tab\tab{column vector of fixed-effects predictor variable values for observation $i$}\\
+%$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
+%$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\
 

### 2015-08-18 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@748 255f4249-2263-4417-b1ea-8498779e27a7
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@749 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usetikzlibrary{shapes,shadows,arrows,positioning}
-\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model (GLMM) Trees}
-\shorttitle{GLMM Trees}
-\leftheader{GLMM Trees}
-\rightheader{GLMM Trees}
+\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM tree)}
+\shorttitle{GLMM Tree}
+\leftheader{GLMM Tree}
+\rightheader{GLMM Tree}
 \author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
@@
 
-%\abstract{}
+\acknowledgements{The authors would like to thank Prof.
+Pim Cuijpers, Prof.
+Jeanne Miranda, Dr.
+Boadie Dunlop, Prof.
+Rob DeRubeis, Prof.
+Zindel Segal, Dr.
+Sona Dimidjian, Prof.
+Steve Hollon and Erica Weitz for granting access to the dataset for the application.
+The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2013.
+The visit was supported by the Institute.}
 
@@
 We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.
-
+We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
+\\
 \textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees
@@
 %
-%''Do logistic regression and signal detection identify different subgroups at risk?
-Implications for the design of tailored interventions.
-Kiernan, Michaela; Kraemer, Helena C.; Winkleby, Marilyn A.; King, Abby C.; Taylor, C.
-Barr.
-Psychological Methods, Vol 6(1).
-''
-%
-%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. Psychological methods, 9(4), 426.''
+%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. 
 %
@@
 %''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''
-%
-%Statistics in Medicine references: \cite{DussyMech14,FostyTayl11,HiggyWhit01,LipkyDmit11}
 
@@
 Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments.
-Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables.
+Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
 Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
@@
 For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled.
-In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{Frie93,DersyLair86,HiggyWhit01}.
+In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
 Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects.
-Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,Hox98,NooryOpde05}.
+Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}.
 More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
@@
 In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects.
-The goal is to estimate a model for predicting the value of a treatment outcome $Y$, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:
+The goal is to estimate a model for predicting the value of a treatment outcome, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:
 
 \begin{equation}
-\label{eq:expected value}
+\label{eq:expected_value}
 	E[y_i | x_i] = \mu_i 
@@
 However, additional treatment conditions and covariates can easily be included.
+In addition, examples and datasets in the current paper will focus on continuous response variables with normally distributed errors, such as posttreatment severity of a disorder.
+But the models and algorithms to be discussed can also be applied with discrete outcomes, such as remission of a disorder (yes/no).
 
-As an example, the estimated GLM of the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}.
+To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}.
 The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
@@
 
+\newpage
 \section{Appendix: Notation}

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/Analysis of datasets.R
Added R simulation files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@753 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,133 @@
+library(lme4)
+library(partykit)
+source("lmertree.R")
+
+# Fit trees on training data
+for (c in 1:50) {
+  load(paste("datasets", c, sep="")) # load training data
+  load(paste("descriptions", c, sep="")) # load trainig data descriptions
+  REEMobtrees <- list()
+  Mobtrees <- list()
+  for (i in 1:length(datasets)) {
+    print(i)
+    if (descriptions[[i]][[3]] == "number of covariates = 5") {
+      formula <- Y ~ T | X1 + X2 + X3 + X4 + X5
+    }
+    if (descriptions[[i]][[3]] == "number of covariates = 15") {
+      formula <- Y ~ T | X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + 
+                  X11 + X12 + X13 + X14 + X15
+    }
+    REEMobtrees[[i]] <- lmertree(lmtreeformula = formula, data=datasets[[i]], 
+                                  randomformula = Y~(1|bi), verbose = F, plotting=F, 
+                                  ErrorTolerance = 0.001, maxdepth=4)
+    Mobtrees[[i]] <- lmtree(formula = formula, data=datasets[[i]], maxdepth=4)
+  }
+  save(REEMobtrees, file=paste("REEMobtrees", c, sep=""))
+  save(Mobtrees, file=paste("Mobtrees", c, sep=""))
+}
+
+# get tree characteristics
+for(c in 1:50){
+  print(c)
+  load(paste("Mobtrees",c,sep=""))
+  load(paste("REEMobtrees",c,sep=""))
+
+  # Summarize tree characteristics of REEMob trees
+  treesize.REEMobs <- list()
+  for (i in 1:length(datasets)) {
+    treesize.REEMobs[[i]] <- length(REEMobtrees[[i]]$Tree)
+  }
+  save(treesize.REEMobs, file=paste("treesize.REEMobs", c, sep=""))
+
+  breaks.REEMobs <- list()
+  varids.REEMobs <- list()
+  for (i in 1:length(datasets)) {
+    breaks.REEMobs[[i]] <- list()
+    varids.REEMobs[[i]] <- list()
+    for (j in 1:length(REEMobtrees[[i]][[1]])) {
[…]

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/Evaluation of results.R

@@ -1,0 +1,275 @@
+library(lme4)
+library(partykit)
+
+# Evaluate tree sizes
+REEMobtreesize <- vector()
+Mobtreesize <- vector()
+for (c in 1:50){
+  load(paste("treesize.REEMobs", c, sep=""))
+  REEMobtreesize <- c(REEMobtreesize, unlist(treesize.REEMobs))
+  load(paste("treesize.Mobs", c, sep=""))
+  Mobtreesize <- c(Mobtreesize, unlist(treesize.Mobs))  
+}
+treesizes <- data.frame(REEMobtreesize, Mobtreesize)
+
+load("testdescriptions1")
+for (i in 1:length(testdescriptions)) {
+  treesizes[i,"N"] <- substr(testdescriptions[[i]][[1]], 5, 8) # N
+  treesizes[i,"rho"] <- substr(testdescriptions[[i]][[2]], 7, 9) # is rho (correlation between covariates)
+  treesizes[i,"np"] <- substr(testdescriptions[[i]][[3]], 24, 25) # is number of covariates
+  treesizes[i,"treatdiff"] <- substr(testdescriptions[[i]][[4]], 31, 33) # is treatment effect difference
+  treesizes[i,"corUbi"] <- substr(testdescriptions[[i]][[5]], 32, 64) # is correlation between U and bi
+  treesizes[i,"numbclus"] <- substr(testdescriptions[[i]][[6]], 34, 35) # is number of random intercept values
+  treesizes[i,"sigmabi"] <- substr(testdescriptions[[i]][[7]], 12, 14) # max and -min random intercept value    
+}
+
+treesizes$N <- factor(treesizes$N)
+treesizes$rho <-  factor(treesizes$rho)            
+treesizes$np <- factor(treesizes$np)              
+treesizes$treatdiff <- factor(treesizes$treatdiff)          
+treesizes$corUbi <- factor(treesizes$corUbi)           
+treesizes$numbclus <- factor(treesizes$numbclus)        
+treesizes$sigmabi <- factor(treesizes$sigmabi)
+
+treesizes$N <- rep(treesizes$N[1:length(testdescriptions)], times=1)
+treesizes$rho <- rep(treesizes$rho[1:length(testdescriptions)], times=1)
[…]

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/generate test data.R

@@ -1,0 +1,140 @@
+N <- list() # different levels for N
+N[[1]] <- 200; N[[2]] <- 500; N[[3]] <- 1000 
+
+bi <- list() # different numbers and values for the study intercepts
+bi$numbclus <- c(5,10,25)
+bi$sigmas <- c(0, 5, 10)
+
+rho <- list() # levels of intercorrelations between the covariates
+rho[[1]] <- matrix(0, nrow=20, ncol=20)
+diag(rho[[1]]) <- 1
+rho[[2]] <- matrix(.3, nrow=20, ncol=20)
+diag(rho[[2]]) <- 1
+
+corUbi <- list()
+corUbi[[1]] <- "uncorrelated"
+corUbi[[2]] <- "bi and splitting U correlated"
+corUbi[[3]] <- "bi and non-splitting U correlated" 
+
+pXj <- list() # levels of number of covariates
+pXj[[1]] <- 5; pXj[[2]] <- 15 
+
+muYdiff <- list() # different levels for the difference in treatment outcome
+muYdiff[[1]] <- 2.5; muYdiff[[2]] <- 5
+
+testdata <- list()
+testdescriptions <- list()
+
+set.seed(38916032)
+
+for (c in 1:50) {
+counter <- 0
+load(paste("descriptions", c, sep="")) # loaded to make correlated covariate correspond between training and testing data
+for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b
+  for (e in 1:length(bi$numbclus)) {# e is counter for no. of clusters
+    numbclus <- bi$numbclus[e] # nRint is number of different random intercept values
+    for (f in 1:length(corUbi)) { # f is counter for correlation between U and b
+      for (g in 1:length(muYdiff)) { # g is counter for treatment effect differences
+        diff <- muYdiff[[g]]
+        for (h in 1:length(pXj)) { # h is counter for number of covariates
+          p <- pXj[[h]]
+          for (i in 1:length(N)) { # i is counter for sample size N
+            nobs <- N[[i]]
+            for (j in 1:length(rho)) { # j is counter for intercorrelations between covariates
+              L <- chol(rho[[j]]) # Cholesky decomposition of correlation matrices
+              nvars <- dim(rho[[j]])[1]
+              
+              # Random variables that follow an M correlation matrix
+              r <- t(L) %*% matrix(rnorm(nvars*nobs, sd=10), nrow=nvars, ncol=nobs)
+              r <- t(r)
+              
+              # create centered covariate datasets
[…]

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/generate training data.R

@@ -1,0 +1,130 @@
+N <- list() # different levels for N
+N[[1]] <- 200; N[[2]] <- 500; N[[3]] <- 1000 
+
+bi <- list() # different numbers and values for the study intercepts
+bi$numbclus <- c(5,10,25)
+bi$sigmas <- c(0, 5, 10)
+
+rho <- list() # levels of intercorrelations between the covariates
+rho[[1]] <- matrix(0, nrow=20, ncol=20)
+diag(rho[[1]]) <- 1
+rho[[2]] <- matrix(.3, nrow=20, ncol=20)
+diag(rho[[2]]) <- 1
+
+corUbi <- list()
+corUbi[[1]] <- "uncorrelated"
+corUbi[[2]] <- "bi and splitting U correlated"
+corUbi[[3]] <- "bi and non-splitting U correlated" 
+
+pXj <- list() # levels of number of covariates
+pXj[[1]] <- 5; pXj[[2]] <- 15 
+
+muYdiff <- list() # different levels for the difference in treatment outcome
+muYdiff[[1]] <- 2.5; muYdiff[[2]] <- 5
+
+datasets <- list()
+descriptions <- list()
+
+set.seed(23061983)
+
+for (c in 1:50) {  
+counter <- 0
+for (d in 1:length(bi$sigmas)) {# d is counter for sigma_b value
+  for (e in 1:length(bi$numbclus)) {# e is counter for no. of clusters
+    numbclus <- bi$numbclus[e] # numbclus is number of different random intercept values
+    for (f in 1:length(corUbi)) { # f is counter for correlation between U and b
+      for (g in 1:length(muYdiff)) { # g is counter for treatment effect differences
+        diff <- muYdiff[[g]]
+        for (h in 1:length(pXj)) { # h is counter for number of covariates
+          p <- pXj[[h]]
+          for (i in 1:length(N)) { # i is counter for sample size N
+            nobs <- N[[i]]
+            for (j in 1:length(rho)) { # j is counter for intercorrelations between covariates
+              L <- chol(rho[[j]]) # Cholesky decomposition of correlation matrices
+              nvars <- dim(rho[[j]])[1]
+              
+              # Random variables that follow an M correlation matrix
+              r <- t(L) %*% matrix(rnorm(nvars*nobs, sd=10), nrow=nvars, ncol=nobs)
+              r <- t(r)
+              
+              # create centered covariate datasets
+              r <- as.data.frame(r)
+              names(r) <- paste("X", 1:15, sep="")
+              
[…]

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/lmertree.R

@@ -1,0 +1,45 @@
+library(partykit)
+library(lme4)
+library(Formula)
+load("datasets1")
+load("descriptions1")
+
+lmertree <- function(lmtreeformula, randomformula, data, #subset = NULL, 
+                       initialRandomEffects = NULL, 
+                       ErrorTolerance = 0.001, MaxIterations = 1000, 
+                       verbose = T, plotting=T,
+                       lmer.control = lmerControl(), ...)
+{
+  iteration <- 0
+  if (!is.null(initialRandomEffects)) {data$offset <- initialRandomEffects}
+  if (is.null(initialRandomEffects)) {data$offset <- rep(0, times=dim(data)[1])}
+  continuecondition <- T
+  oldloglik <- -Inf
+  while (continuecondition) {
+    iteration <- iteration+1
+    tree <- lmtree(lmtreeformula, data, offset=offset, ...)
+    if(verbose){plot(tree)}
+    if(length(tree)>1) {
+      data$treeresponse <- predict(tree, newdata=data)
+    }
+    if(length(tree)==1) { # if treedepth=1, predict.lmtree doesn't work, then use unpartitioned lm
+      lmformula <- formula(as.Formula(lmtreeformula), lhs=1, rhs = -2)
+      data$treeresponse <- predict(lm(lmformula, data=data), newdata=data)
+    }
+    lme <- lmer(randomformula, data=data, offset=treeresponse)
+    data$offset <- predict(lme, newdata=data)
+    newloglik <- logLik(lme)    
+    continuecondition <- (newloglik - oldloglik > ErrorTolerance) & (iteration < MaxIterations) 
+    oldloglik <- newloglik
+    if(verbose){print(newloglik)}
+  } 
+  result <- list(Tree = tree, RandomEffectModel = lme, RandomEffects = ranef(lme), 
+                 BetweenMatrix = VarCorr(lme), ErrorVariance = attr(VarCorr(lme),"sc")^2, 
+                 data = data, logLik = newloglik, IterationsUsed = iteration, 
+                 MaxIterations=MaxIterations, initialRandomEffects = initialRandomEffects, 
+                 TreeFormula = lmtreeformula, RandomFormula = randomformula, #Subset = subs, 
+                 ErrorTolerance = ErrorTolerance, #residuals = residuals, 
+                 tree.control = list(...), lmer.control = lmer.control)
+}
+
+

### 2015-08-25 pkg/glmertree/paper/Simulation replication/Datasets without treat-sub interactions/generate training data.R

@@ -1,0 +1,120 @@
+N <- list() # different levels for N
+N[[1]] <- 200; N[[2]] <- 500; N[[3]] <- 1000 
+
+bi <- list() # different numbers and values for the study intercepts
+bi$numbclus <- c(5,10,25)
+bi$sigmas <- c(0, 5, 10)
+
+rho <- list() # levels of intercorrelations between the covariates
+rho[[1]] <- matrix(0, nrow=20, ncol=20)
+diag(rho[[1]]) <- 1
+rho[[2]] <- matrix(.3, nrow=20, ncol=20)
+diag(rho[[2]]) <- 1
+
+corUbi <- list()
+corUbi[[1]] <- "uncorrelated"
+corUbi[[2]] <- "bi and random U correlated"
+
+pXj <- list() # levels of number of covariates
+pXj[[1]] <- 5; pXj[[2]] <- 15 
+
+muYdiff <- list() # different levels for the difference in treatment outcome
+muYdiff[[1]] <- 2.5; muYdiff[[2]] <- 5
+
+datasets <- list()
+descriptions <- list()
+
+set.seed(24061983)
+
+for (c in 1:50) {  
+counter <- 0
+for (d in 1:length(bi$sigmas)) {# d is counter for sigma of b_i's
+  for (e in 1:length(bi$numbclus)) {# e is counter for number of clusters
+    numbclus <- bi$numbclus[e] # numbclus is number of different random intercept values
+    for (f in 1:length(corUbi)) { # f is counter for correlation between U and bi's
+      for (g in 1:length(muYdiff)) { # g is counter for treatment effect differences
+        diff <- muYdiff[[g]]
+        for (h in 1:length(pXj)) { # h is counter for number of covariates
+          p <- pXj[[h]]
+          for (i in 1:length(N)) { # i is counter for sample size N
+            nobs <- N[[i]]
+            for (j in 1:length(rho)) { # j is counter for intercorrelations between covariates
+              L <- chol(rho[[j]]) # Cholesky decomposition of correlation matrices
+              nvars <- dim(rho[[j]])[1]
+              
+              # Random variables that follow an M correlation matrix
+              r <- t(L) %*% matrix(rnorm(nvars*nobs, sd=10), nrow=nvars, ncol=nobs)
+              r <- t(r)
+              
+              # create centered covariate datasets
+              r <- as.data.frame(r)
+              names(r) <- paste("X", 1:15, sep="")
+              
+              # add population means to covariates
[…]

### 2015-08-25 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@755 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usetikzlibrary{shapes,shadows,arrows,positioning}
-\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM tree)}
+\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)}
 \shorttitle{GLMM Tree}
@@
 
-%\section{Citation stuff for journal publication}
-%
-%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
-%
-%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. 
-%
-%''Cooper, H., \& Patall, E.
-A.
-(2009).
-The relative benefits of meta-analysis conducted with individual participant data versus aggregated data.
-Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results.
-They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
-%
-%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''
-
 \section{Introduction}
 
-In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important.
-Personalized medicine presents the challenge of finding which patients respond best to which treatments.
+In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified medicine is becoming increasingly important.
+Stratified medicine presents the challenge of finding which patients respond best to which treatments.
 This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}.
@@
 Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
-After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
+After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
 Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.
@@
 
-To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 (78 observations) or Treatment 2 (72 observations).
+To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 or Treatment 2.
 Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
@@
 
-The simulated dataset has 3 subgroups with treatment interactions.
+The simulated dataset has 3 subgroups with different treatment effectiveness.
 The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$.
-In this subgroup, the mean of the response variable for Treatment 1 is 7, and the mean for the response variable for Treatment 2 is 11.
+In this subgroup, Treatment 1 is more effective than Treatment 2: the mean of the response variable is 7 for Treatment 1, and 11 for Treatment 2.
 The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$.
-In this subgroup, the mean value of the response variable for Treatment 1 and 2 is 9.
+In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment 1, and 9 for Treatment 2.
 The third subgroup consists of observations with duration $>6$.
-In this subgroup, the mean value of the response variable for Treatment 1 is 12, and the mean of the response variable for Treatment 2 is 7.
+In this subgroup, Treatment 2 is more effective than Treatment 1: the mean value of the response variable is 12 for Treatment 1, and 7 for Treatment 2.
 
@@
 Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function.
-In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}$.
-Further, $x_{i}^{\top}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type.
+In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$.
+Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (taking a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type).
 $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
 
-To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2.
+To keep notation and examples simple, we assume $x_{i}$ and $\beta$ to have length 2.
 That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM.
@@
 Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM.
-The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*\top}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
-$\beta^{* \top}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.
+The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
+$\beta^{*}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.
 
@@
 However, random-effects covariates and coefficients can easily be included.
-Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the linear predictor of a GLM.
+Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.
 
@@
 To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
-\citeA{HajjyBell11} and \citeA{SelaySimo12} developed a method for estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees.
+They developed a method for the estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees.
 In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
@@
 
-For estimating GLMM trees, we take the MERT approach a step further, by using a GLM tree instead of a regression tree with constant fits.
-This allows not only for detection of differences in main effects,but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
+For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM.
+This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
 In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
@@
 \item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42).
-\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for Treatment 1 ($x_{1}=0$) and Treatment 2 ($x_{1}=1$).
-The levels for mean differences in subgroups with differential treatment effect were $|\beta_1| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_1| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
+\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects.
+The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
 \end{enumerate}
@@
 Therefore, 50 x 3 x 2 x 2 x 3 x 3 x 2 x 2 = 21,600 datasets without treatment-subgroup interactions were generated.
+
+It should be noted that when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition.
+However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.
 
@@
 In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$.
-For every observation, we generated a binomial variable (with $p =.5$) as an indicator for treatment type.
+For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type.
 
@@
 
-To detect predictors of tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+To assess the impact of data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.
+The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.
 
-To detect predictors of tree accuracy for both algorithms in datasets with treatment-subgroup, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
+To assess the impact of data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The effects of predictors with main and/or interaction effects with unstandardized coefficients $|\beta| > .2$ were further investigated using graphical displays.
+The impact of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further investigated using graphical displays.
 
@@
 
-For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment-effect differences: one for GLM tree, and one for GLMM tree.
-To detect predictors of predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+For every dataset, correlation coefficients for each algorithm were calculated, representing the linear association between the true and predicted treatment-effect differences.
+To assess the impact of data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.
+The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.
 
@@
 
-\subsubsection{Estimation of GLM and GLMM trees}
+\subsubsection{Estimation of GLM trees}
 
@@
 
-For estimating GLMM trees, we implemented the GLMM tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
+For estimating GLMM trees, we implemented the GLMM-tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
 Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm.
@@
 
-For building the GLM trees in step (1) of the GLMM tree algorithm, the same settings as described above were used.
+For building the GLM trees in step (1) of the GLMM-tree algorithm, the same settings as described above were used.
 That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.
@@
 
-\subsection{Tree size and accuracy in datasets without treatment-subgroup interactions}
+\subsubsection{Tree size in datasets without treatment-subgroup interactions}
 
@@
    	GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
-    			& (.67)  & (.20) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
+    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
 	\hline
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=12cm]{xy_treesizes_maineff.pdf}
+	\includegraphics[width=15cm]{xy_treesizes_maineff.pdf}
 	\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions.
-Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.
+Values $200$, $500$ and $1000$ refer to sample size.
 Reference line at $y=1$ represents the true tree size.}
@@
 
-\subsection{Tree size in datasets with treatment-subgroup interactions}
+\subsubsection{Tree size in datasets with treatment-subgroup interactions}
 
@@
 Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}.
-The average size of GLMM trees was 7.15 (SD=0.61), and the average size of GLM trees was 8.11 (SD=2.05).
-The estimated probability that a datasets was erroneously not partitioned was 0, for both GLM and GLMM tree.
-However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .64 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).
+The average size of GLMM trees was 7.16 (SD=0.62), and the average size of GLM trees was 8.12 (SD=2.05).
+The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree.
+However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).
 
@@
 	\hline
-  	GLMM tree	& 3 & 227& 29556 & 2472 & 89 & 3 &0& 32,400 \\
-				&  ($<.01$)  & ($<.01$)& (.91) & ($<.01$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\
+  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
+				&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\
 
-  	GLM tree		& 145 & 1002 & 20578 & 4443 & 4178 & 1665 & 389 & 32,400\\ 
- 			 	& ($<.01$) & (.03) & (.64) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
+  	GLM tree		& 139 & 955 & 20530 & 4565 & 4164 & 1619 & 428 & 32,400\\ 
+ 			 	& ($<.01$) & (.03) & (.63) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
 	\hline
@@
 A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_interact}).
-When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees of size 7, on average.
+When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average.
+
+\begin{figure}[!htbp]
+	\includegraphics[width=15cm]{xy_treesizes_treatsubs.pdf}
+	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions.
+Values $200$, $500$ and $1000$ refer to sample size.
+Reference line at $y=7$ represents true tree size.}
+	\label{fig:xyplot_treesize_interact}
+\end{figure}
 
@@
 When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$).
-When $b$ is correlated with one of the $U_k$ variables, GLM starts creating spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
+When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
+This effect was stronger when $b$ was correlated to a non-splitting variable.
 
-\begin{figure}[!htbp]
-	\includegraphics[width=12cm]{xy_treesizes_treatsubs.pdf}
-	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions.
-Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.
-Reference line at $y=7$ represents true tree size.}
-	\label{fig:xyplot_treesize_interact}
-\end{figure}
-
-\subsection{Tree accuracy in datasets with treatment-subgroup interactions}
+\subsubsection{Tree accuracy in datasets with treatment-subgroup interactions}
 
 To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset.
-For the first split, both GLM tree and GLMM tree always selected the true partitioning variable ($U_2$).
-The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94, for both GLM and GLMM tree.
-However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split, than did GLMM tree (SD=0.155 and SD=0.126, respectively).
+For the first split, GLMM tree always selected the true partitioning variable ($U_2$).
+GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset.
+The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree.
+However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree (SD=0.154 and SD=0.127, respectively).
 
@@
 In the presence of random effects, GLM trees were much less accurate than GLMM trees.
-This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables.
+This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables, and the effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable.
 When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=12cm]{xy_accuracy.pdf}
+	\includegraphics[width=15cm]{xy_accuracy.pdf}
 	\caption{Average accuracy of GLM and GLMM trees.
 Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered.
-Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
+Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
 	\label{fig:xyplot_treeaccuracy}
@@
 Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree.
-The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.20) for GLM tree, and .94 (SD=0.10) for GLMM tree.
+The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.19) for GLM tree, and .94 (SD=0.11) for GLMM tree.
 
@@
 \subsection{Method}
+
 To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
@@
 
-Predictive accuracy of GLM and GLMM tree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
+The predictive accuracy of GLM and GLMM tree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
 
@@
 
-The trees resulting from appplication of GLM and GLMM tree to the dataset are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}, respectively.
+The trees resulting from application of GLM and GLMM tree to the dataset are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}, respectively.
 Note that the GLM tree in Figure \ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMM-tree algorithm.
@@
 
-By taking into account the study-specific intercepts, the final GLMM tree (Figure \ref{fig:lmertree_C&W}) indicates that the first split made by GLM tree is a spurious split.
+By taking into account the study-specific intercepts, the final GLMM tree (Figure \ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
 The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable.
 The terminal nodes of Figure \ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores.
-The estimated variance of the random intercept term was 2.12, with an estimated intraclass correlation coefficient of .05.
+The estimated intraclass correlation coefficient for the random intercepts was .05.
 
 Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .39 (SD=.20) for GLMM tree, and .31 (SD=.24) for GLM tree.
-This indicates that GLMM tree not only provided higher predictive accuracy, on average, but also had somewhat lower variability of predictive accuracy than GLM tree.
+The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 (var=.067) for GLMM tree, and .19 (var=.084) for GLM tree.
+This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.
 
@@
 
+In the current paper, we presented the GLMM-tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters.
 The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions.
-In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets.
-GLM tree performed less accurate than GLMM tree: in datasets with treatment-subgroup interactions, GLM tree recovered the true tree structure in 61\% of the simulated datasets.
-In datasets without treatment-subgroup interactions, GLM tree erroneously detected subgroups in 33\% of the datasets.
+In contrast, GLM tree recovered the true tree structure in only 61\% of the datasets with treatment-subgroup interactions.
+In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
 
@@
 If there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using the partitioning variables that are available.
-Or, if the coefficients of other variables vary across subgroups, then those variables should also be included in the GLM.
-Another source of misspecification would be the inclusion of irrelevant variables in the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
-However, it should be note that in our simulations, the number of partitioning variables did not substantially influence performance of the algorithm.
+Or, if the coefficients of other variables vary across subgroups, then those variables should also be included in the linear predictor of the GLM.
+Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
+However, it should be noted that in our simulations, the number of partitioning variables did not substantially influence performance of the algorithm(s).
 
@@
 $\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
+$r$				\tab\tab{iteration number}\\
 $\sigma_{b}$	\tab\tab{square root of variance of $b$}\\
@@
 $x_i$ 			\tab\tab{column vector of fixed-effects predictor variable values for observation $i$}\\
-%$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
-%$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\
+$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
+$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\
 

### 2015-08-25 pkg/glmertree/paper/treatment_interactions.Rnw
started Sweave version of the manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@756 255f4249-2263-4417-b1ea-8498779e27a7

--- 2015-08-25 pkg/glmertree/paper/treatment_interactions paper.tex
@@
 \usepackage{tabto}
+
+%% for \usepackage{Sweave}
+\SweaveOpts{echo=FALSE, results=hide}
+\setkeys{Gin}{width=0.8\textwidth}
+
+<<preliminaries>>=
+library("foreign")
+library("glmertree")
+library("lattice")
+@
 
@@
 \begin{figure}[!h]
-    \tikzstyle{decision} = [diamond, draw]
-    \tikzstyle{line} = [draw, -stealth, thick]
-    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
-   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
-    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
-        \node [elli] (node1) { };
-        \node [elli, below of =node1, xshift=-9em] (node2) { };
-        \node [elli, below of =node1, xshift=9em] (node3) { };
-        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
-        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
-        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
-        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
-        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
-        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
-        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
-        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
-        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
-        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
-    \end{tikzpicture}
+% \tikzstyle{decision} = [diamond, draw]
+% \tikzstyle{line} = [draw, -stealth, thick]
+% \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
+% \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
+% \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
+%     \node [elli] (node1) { };
+%     \node [elli, below of =node1, xshift=-9em] (node2) { };
+%     \node [elli, below of =node1, xshift=9em] (node3) { };
+%     \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
+%     \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
+%     \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
+%     \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
+%     \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
+%     \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
+%     \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
+%     \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
+%     \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
+%     \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
+% \end{tikzpicture}
+\setkeys{Gin}{width=0.8\textwidth}
+<<dgp-tree, fig=TRUE, height=5, width=7>>=
+fig4 <- party(
+  partynode(1L,
+    split = partysplit(2L, breaks = 30),
+    kids = list(
+      partynode(2L,
+        split = partysplit(1L, breaks = 17),
+        kids = list(
+          partynode(3L, info = c(
+	    expression(''),
+	    expression(beta[j0] == '17.5'),
+	    expression(''),
+	    expression(beta[j1] == -'5.0'),
+	    expression(''),
+	    expression(d[j] == -'1.0')
+	  )),
+          partynode(4L, info = c(
+	    expression(''),
+	    expression(beta[j0] == '30.0'),
+	    expression(''),
+	    expression(beta[j1] == '0.0'),
+	    expression(''),
+	    expression(d[j] == '0.0')
+	  )))),
+      partynode(5L,
+        split = partysplit(1L, breaks = 63),
+        kids = list(
+          partynode(6L, info = c(
+	    expression(''),
+	    expression(beta[j0] == '30.0'),
+	    expression(''),
+	    expression(beta[j1] == '0.0'),
+	    expression(''),
+	    expression(d[j] == '0.0')
+	  )),
+          partynode(7L, info = c(
+	    expression(''),
+	    expression(beta[j0] == '42.5'),
+	    expression(''),
+	    expression(beta[j1] == '5.0'),
+	    expression(''),
+	    expression(d[j] == '1.0')
+	  )))))),
+  data.frame(U1 = numeric(0), U2 = numeric(0))
+)
+plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
+@
 \caption{Data-generating model for treatment-subgroup interactions.
@@
 
+<<depresseion-data>>=
+metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
+metadata[metadata == 999] <- NA
+metadata[metadata == 888] <- NA
+vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
+factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
+metadata$education <- factor(metadata$education, ordered = TRUE)
+for (i in 1:length(factors)) {metadata[, factors[i]] <- factor(metadata[,factors[i]])}
+metadata <- metadata[vars] # select only relevant variables
+metadata <- metadata[complete.cases(metadata[, vars]), ] # select only complete data
+metadata <- metadata[!metadata$Tx_group == "placebo", ] # remove placebo observations
+metadata$Tx_group <- factor(metadata$Tx_group)
+@
+
 \subsection{Method}
@@
 
+<<depression-trees>>=
+formula <- HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0
+
+## GLM tree
+lmtree_app <- lmtree(formula, data = metadata,
+  offset = fitted(lm(HRSDt1 ~ HRSDt0), data = metadata))
+
+## GLMM tree
+lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
+  initialRandomEffects = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
+@
+
 \subsection{Results}
@@
 \begin{figure}[!ht]
-    \includegraphics[width=12cm]{app_lmtree.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<depression-glmtree, fig=TRUE, height=6, width=9>>=
+plot(lmtree_app)
+@
 \caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
@@
 \begin{figure}[!ht]
-    \includegraphics[width=10cm]{app_lmertree.pdf}
+\setkeys{Gin}{width=0.75\textwidth}
+<<depression-glmmtree, fig=TRUE, height=4, width=6>>=
+plot(lmertree_app$Tree)
+@
 \caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).

### 2015-08-27 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@757 255f4249-2263-4417-b1ea-8498779e27a7

--- 2015-08-25 pkg/glmertree/paper/treatment_interactions paper.tex
@@
 Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree.
+Subsequently, we evaluate the performances of the GLMM-tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
 
-Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
+Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
 After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
@@
 \subsection{Artificial motivating dataset}
+
+<<artificial-data>>=
+set.seed(123)
+treatment <- rbinom(n=150,size=1,prob=1/2)
+duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
+duration[duration<0] <- 1
+anxiety <- round(rnorm(150,mean=10,sd=3))
+age <- round(rnorm(150,mean=30,sd=10))
+age[age<18] <- 18
+age[age>75] <- 75
+cutscore1 <- 5+.44*3
+cutscore2 <- 10
+node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
+node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
+node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
+node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
+node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
+node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
+beta1 <- -2
+beta2 <- 0
+beta3 <- 2.5
+error <- rnorm(150,0,2)
+cluster <- error+rnorm(150, 0, 6)
+int_values <- rnorm(10, 0, 1)
+tmp <- rep(int_values,each=15)
+tmp <- tmp[order(tmp)]
+rand_int <- rep(NA, times=150)
+rand_int[order(cluster)] <- tmp
+cluster[order(cluster)] <- rep(1:10, each=15)
+cor(error,rand_int)
+true_error <- error - rand_int
+true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
+     + .4*treatment)
+outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
+                      + .4*treatment + true_error + rand_int)
+outcome[outcome<0] <- 1
+example <- data.frame(treatment,outcome,duration,anxiety,cluster, true_eff)
+example$treatment <- factor(example$treatment,labels=c("Treatment 1","Treatment 2"))
+#cor(outcome,rand_int)^2
+#cor(outcome, true_eff)^2
+#truetree <- lmtree(true_eff ~ treatment | age + duration + anxiety, data=example)
+@
 
@@
 
+<<example-trees>>=
+exampleglm <- lmtree(outcome ~ treatment | age, data=example)
+exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data=example)
+exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data=example)
+@
+
 \subsection{GLM}
@@
 In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$.
-Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (taking a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type).
+Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type).
 $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
@@
 
-\vspace{20pt}
-\begin{figure}[!htbp]
-    \includegraphics[width=6cm]{glm_example.pdf}
+\begin{figure}[!ht]
+\setkeys{Gin}{width=0.5\textwidth}
+<<exampleglm, fig=TRUE, height=5, width=5>>=
+plot(exampleglm)
+@
 	\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150).
@@
 
-The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
+The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
 This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
@@
 \begin{figure}[!h]
-    \includegraphics[width=15cm]{glmtree_example.pdf}
+\setkeys{Gin}{width=\textwidth}
+<<exampleglmtree, fig=TRUE, width=12, height=9>>=
+plot(exampleglmtree)
+@
     \caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset.
@@
 \label{eq:mixedeffects}
-	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b_{m}
+	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
 \end{equation}
 
-Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$ ($m=1,...,M$), of which observation $i$ is part.
+Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0, with $m$ ($m=1,...,M$) denotes the cluster of which observation $i$ is part.
+$b$ is random vector of length $M$, each element being the random intercept for $m$.
 Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
@@
 
-For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models.
+For simplicity, we assume that only cluster-specific intercepts are included in the models.
 However, random-effects covariates and coefficients can easily be included.
-Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.
+Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.
 
@@
 	\label{eq:glimmertree}
-	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b_{m}
+	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
 \end{equation}
@@
 \vspace{5pt}
-\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{m(r)}$ to 0.
+\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
 
 \vspace{5pt}
-\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using responses $y_i - z_{i}^{\top}\hat{b}_{m(r-1)}$.
+\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
 
 \vspace{5pt}
-\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{m(r)}$ using responses $y_i - x_{i}^{\top}\hat{\beta}_{j(r)}$.
+\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$ using $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
 
@@
 \vspace{5pt}
-The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
+The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
 In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
-The GLM tree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLM tree from the current iteration.
+The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
 Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.
 
-In Figure \ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented.
-As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.
-
 \begin{figure}[!h]
-    \includegraphics[width=12cm]{glimmertree_example.pdf}
+<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
+plot(exampleglmmtree$Tree)
+@
     \caption{Generalized Linear Mixed Model tree of the motivating example dataset.
@@
 
-\section{Simulation}
+In Figure \ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented.
+As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.
+
+\section{Emperical evaluation}
 
@@
 \begin{figure}[!h]
-    \tikzstyle{decision} = [diamond, draw]
-    \tikzstyle{line} = [draw, -stealth, thick]
-    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
-   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
-    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
-        \node [elli] (node1) { };
-        \node [elli, below of =node1, xshift=-9em] (node2) { };
-        \node [elli, below of =node1, xshift=9em] (node3) { };
-        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
-        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
-        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
-        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
-        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
-        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
-        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
-        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
-        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
-        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
-    \end{tikzpicture}
+\setkeys{Gin}{width=0.8\textwidth}
+<<dgp-tree, fig=TRUE, height=5, width=7>>=
+fig4 <- party(
+partynode(1L,
+ split = partysplit(2L, breaks = 30),
+ kids = list(
+   partynode(2L,
+     split = partysplit(1L, breaks = 17),
+     kids = list(
+       partynode(3L, info = c(
+  expression(''),
+  expression(beta[j0] == '17.5'),
+  expression(''),
+  expression(beta[j1] == -'5.0'),
+  expression(''),
+  expression(d[j] == -'1.0')
+ )),
+       partynode(4L, info = c(
+  expression(''),
+  expression(beta[j0] == '30.0'),
+  expression(''),
+  expression(beta[j1] == '0.0'),
+  expression(''),
+  expression(d[j] == '0.0')
+ )))),
+   partynode(5L,
+     split = partysplit(1L, breaks = 63),
+     kids = list(
+       partynode(6L, info = c(
+  expression(''),
+  expression(beta[j0] == '30.0'),
+  expression(''),
+  expression(beta[j1] == '0.0'),
+  expression(''),
+  expression(d[j] == '0.0')
+ )),
+       partynode(7L, info = c(
+  expression(''),
+  expression(beta[j0] == '42.5'),
+  expression(''),
+  expression(beta[j1] == '5.0'),
+  expression(''),
+  expression(d[j] == '1.0')
+ )))))),
+data.frame(U1 = numeric(0), U2 = numeric(0))
+)
+plot(fig4, tnex=1.5)
+@
 \caption{Data-generating model for treatment-subgroup interactions.
@@
 \item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
-\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
+\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
 \item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
@@
 \verb|R| \cite{R14} was used for generation and analysis of all datasets.
-Two additional R packages were used: \verb|partykit| (version 0.8-3; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
+Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
 
@@
 In the current paper, REML estimation was used.
+
+\subsection{Results}
+
+<<simulation-data>>=
+load("treespecs_long.dat")
+treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
+treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
+treespecs.long$treatdiff <- factor(as.numeric(as.character(treespecs.long$treatdiff)), ordered=T)
+treespecs.long$truetree.values <- as.numeric(treespecs.long$truetree.values)-1
+levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and splitting U correlated"] <- "b correlated with splitting U"
+levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
+levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="uncorrelated"] <- "b and U uncorrelated"
+@
 
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=15cm]{xy_treesizes_maineff.pdf}
+<<treesize-nointeract-xyplot, fig=TRUE, width=8, height=7>>=
+load("treesizes_long_nointeract.dat")
+treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
+treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
+aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
+print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, type="b", ylab="tree size", xlab="sigma_b", par.settings=standard.theme("pdf",color=F), abline=c(1,0), auto.key=list(space="top", columns=2, title="Algorithm type", cex.title=1,lines=T, points=T)))
+@
 \caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions.
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=15cm]{xy_treesizes_treatsubs.pdf}
+<<treesize-xyplot, fig=TRUE, width=9, height=9.5>>=
+aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
+print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+@
 \caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions.
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=15cm]{xy_accuracy.pdf}
+<<accuracy-xyplot, fig=TRUE, width=9, height=9.5>>=
+aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
+print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b",   ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+@
 \caption{Average accuracy of GLM and GLMM trees.
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=12cm]{xy_correlations.pdf}
+<<correlation-xyplot, fig=TRUE>>=
+aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
+print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+@
 \caption{Average predictive accuracy of GLM and GLMM trees.
@@
 
-\section{Application to real data}
+\section{Application to patient-level meta-analysis dataset on the effects of treatments for depression}
+
+<<depression-data>>=
+metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
+metadata[metadata == 999] <- NA
+metadata[metadata == 888] <- NA
+vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
+factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
+metadata$education <- factor(metadata$education, ordered = TRUE)
+for (i in 1:length(factors)) {metadata[, factors[i]] <- factor(metadata[,factors[i]])}
+metadata <- metadata[vars] # select only relevant variables
+metadata <- metadata[complete.cases(metadata[, vars]), ] # select only complete data
+metadata <- metadata[!metadata$Tx_group == "placebo", ] # remove placebo observations
+metadata$Tx_group <- factor(metadata$Tx_group)
+@
 
@@
 
+<<depression-trees>>=
+formula <- HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0
+
+## GLM tree
+lmtree_app <- lmtree(formula, data = metadata,
+ 		offset = fitted(lm(HRSDt1 ~ HRSDt0), data = metadata))
+
+## GLMM tree
+lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
+ 		initialRandomEffects = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
+@
+
 \subsection{Results}
@@
 \begin{figure}[!ht]
-    \includegraphics[width=12cm]{app_lmtree.pdf}
+\setkeys{Gin}{width=0.75\textwidth}
+<<depression-glmtree, fig=TRUE, height=6, width=8>>=
+plot(lmtree_app)
+@
 \caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
@@
 \begin{figure}[!ht]
-    \includegraphics[width=10cm]{app_lmertree.pdf}
+\setkeys{Gin}{width=0.75\textwidth}
+<<depression-glmmtree, fig=TRUE, height=4, width=6>>=
+	plot(lmertree_app$Tree)
+@
 \caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
@@
 In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
+In other words, the Type I error rate of GLMM tree very closely resembled the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.
 
@@
 
-Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations and challenges for future research should be noted.
+Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations should be noted.
 
@@
 
-A challenge for future research is the development of more adequate ways to deal with missing data.
-GLM tree, and therefore also GLMM tree, handle missing data by listwise deletion, like most tree-based algorithms for treatment-subgroup interaction detection.
-However, missing data commonly occurs in clinical trails, and listwise deletion is not an optimal approach for dealing with missing data.
-
 In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects.
@@
 
-\newpage
 \section{Appendix: Notation}

### 2015-08-28 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@760 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-The \verb|lmtree| function deals with missing data by listwise deletion.
-Therefore, we build all trees using data of a subset of 694 patients from 7 studies, as complete observations (i.e., observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score) for these patients were available.
+We build all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score.
+As a result, data from 694 patients from 7 studies were included for the analyses.
 Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \cite{CuijyWeit14}.

### 2015-08-28 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/Evaluation of results.R

@@ -10,58 +10,48 @@ for (c in 1:50){
   load(paste("treesize.Mobs", c, sep=""))
   Mobtreesize <- c(Mobtreesize, unlist(treesize.Mobs))  
 }
-treesizes <- data.frame(REEMobtreesize, Mobtreesize)
+treespecs <- data.frame(REEMobtreesize, Mobtreesize)
 
-load("testdescriptions1")
-for (i in 1:length(testdescriptions)) {
-  treesizes[i,"N"] <- substr(testdescriptions[[i]][[1]], 5, 8) # N
-  treesizes[i,"rho"] <- substr(testdescriptions[[i]][[2]], 7, 9) # is rho (correlation between covariates)
-  treesizes[i,"np"] <- substr(testdescriptions[[i]][[3]], 24, 25) # is number of covariates
-  treesizes[i,"treatdiff"] <- substr(testdescriptions[[i]][[4]], 31, 33) # is treatment effect difference
-  treesizes[i,"corUbi"] <- substr(testdescriptions[[i]][[5]], 32, 64) # is correlation between U and bi
-  treesizes[i,"numbclus"] <- substr(testdescriptions[[i]][[6]], 34, 35) # is number of random intercept values
-  treesizes[i,"sigmabi"] <- substr(testdescriptions[[i]][[7]], 12, 14) # max and -min random intercept value    
+load("descriptions1")
+for (i in 1:length(descriptions)) {
+  treespecs[i,"N"] <- substr(descriptions[[i]][[1]], 5, 8)
+  treespecs[i,"rho"] <- substr(descriptions[[i]][[2]], 7, 9)
+  treespecs[i,"np"] <- substr(descriptions[[i]][[3]], 24, 25)
+  treespecs[i,"treatdiff"] <- substr(descriptions[[i]][[4]], 31, 33)
+  treespecs[i,"corUbi"] <- substr(descriptions[[i]][[5]], 32, 64)
+  treespecs[i,"numbclus"] <- substr(descriptions[[i]][[6]], 34, 35)
+  treespecs[i,"sigmabi"] <- substr(descriptions[[i]][[7]], 12, 14)  
 }
 
-treesizes$N <- factor(treesizes$N)
-treesizes$rho <-  factor(treesizes$rho)            
-treesizes$np <- factor(treesizes$np)              
[…]

### 2015-08-28 pkg/glmertree/paper/Simulation replication/Datasets with treat-sub interactions/lmertree.R

@@ -1,45 +1,0 @@
-library(partykit)
-library(lme4)
-library(Formula)
-load("datasets1")
-load("descriptions1")
-
-lmertree <- function(lmtreeformula, randomformula, data, #subset = NULL, 
-                       initialRandomEffects = NULL, 
-                       ErrorTolerance = 0.001, MaxIterations = 1000, 
-                       verbose = T, plotting=T,
-                       lmer.control = lmerControl(), ...)
-{
-  iteration <- 0
-  if (!is.null(initialRandomEffects)) {data$offset <- initialRandomEffects}
-  if (is.null(initialRandomEffects)) {data$offset <- rep(0, times=dim(data)[1])}
-  continuecondition <- T
-  oldloglik <- -Inf
-  while (continuecondition) {
-    iteration <- iteration+1
-    tree <- lmtree(lmtreeformula, data, offset=offset, ...)
-    if(verbose){plot(tree)}
-    if(length(tree)>1) {
-      data$treeresponse <- predict(tree, newdata=data)
-    }
-    if(length(tree)==1) { # if treedepth=1, predict.lmtree doesn't work, then use unpartitioned lm
-      lmformula <- formula(as.Formula(lmtreeformula), lhs=1, rhs = -2)
-      data$treeresponse <- predict(lm(lmformula, data=data), newdata=data)
-    }
-    lme <- lmer(randomformula, data=data, offset=treeresponse)
-    data$offset <- predict(lme, newdata=data)
-    newloglik <- logLik(lme)    
-    continuecondition <- (newloglik - oldloglik > ErrorTolerance) & (iteration < MaxIterations) 
-    oldloglik <- newloglik
-    if(verbose){print(newloglik)}
-  } 
-  result <- list(Tree = tree, RandomEffectModel = lme, RandomEffects = ranef(lme), 
-                 BetweenMatrix = VarCorr(lme), ErrorVariance = attr(VarCorr(lme),"sc")^2, 
-                 data = data, logLik = newloglik, IterationsUsed = iteration, 
-                 MaxIterations=MaxIterations, initialRandomEffects = initialRandomEffects, 
-                 TreeFormula = lmtreeformula, RandomFormula = randomformula, #Subset = subs, 
-                 ErrorTolerance = ErrorTolerance, #residuals = residuals, 
-                 tree.control = list(...), lmer.control = lmer.control)
-}
-
-

### 2015-08-28 pkg/glmertree/paper/Simulation replication/Datasets without treat-sub interactions/Analys of datasets + evaluation.R

@@ -1,122 +1,0 @@
-library(lme4)
-library(partykit)
-source("lmertree.R")
-
-for (c in 9:50) {
-  load(paste("datasets", c, sep="")) # loads training data
-  load(paste("descriptions", c, sep="")) # loads trainig data descriptions
-
-  # Fit btrees on training data
-  REEMobtrees <- list()
-  Mobtrees <- list()
-  for (i in 1:length(datasets)) {
-  print(i)
-  if (descriptions[[i]][[3]] == "number of covariates = 5") {
-    formula <- Y ~ T | X1 + X2 + X3 + X4 + X5
-  }
-  if (descriptions[[i]][[3]] == "number of covariates = 15") {
-    formula <- Y ~ T | X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15
-  }
-  REEMobtrees[[i]] <- lmertree(lmtreeformula = formula, data=datasets[[i]], 
-                              randomformula = Y~(1|bi), verbose = F, plotting=F, 
-                              ErrorTolerance = 0.001, maxdepth=4)
-  Mobtrees[[i]] <- lmtree(formula = formula, data=datasets[[i]],
-                          maxdepth=4)
-  }
-  save(REEMobtrees, file=paste("REEMobtrees", c, sep=""))
-  save(Mobtrees, file=paste("Mobtrees", c, sep=""))
-}
-
-for(c in 1:50){
-  load(paste("Mobtrees",c,sep=""))
-  load(paste("REEMobtrees",c,sep=""))
-  # Summarize tree characteristics of REEMob trees
-  treesize.REEMobs <- list()
-  for (i in 1:length(datasets)) {
-    treesize.REEMobs[[i]] <- length(REEMobtrees[[i]]$Tree)
-  }
-  save(treesize.REEMobs, file=paste("treesize.REEMobs", c, sep=""))
-  # Summarize tree characteristics of Mob trees
-  treesize.Mobs <- list()
-  for (i in 1:length(datasets)) {
-    treesize.Mobs[[i]] <- length(Mobtrees[[i]])
-  }
-  save(treesize.Mobs, file=paste("treesize.Mobs", c, sep=""))
-}
-
-# Evaluate tree sizes
-REEMobtreesize <- vector()
-Mobtreesize <- vector()
-for (c in 1:50){
-  load(paste("treesize.REEMobs", c, sep=""))
[…]

### 2015-08-28 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@762 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations should be noted.
-
-The simulations show that GLMM tree performs very well, if the model is correctly specified.
+These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures.
+However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified.
 That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then the algorithm will accurately recover those subgroups.

### 2015-08-28 pkg/glmertree/paper/Rnw_2_tex_2_pdf.R

@@ -1,0 +1,2 @@
+Sweave("treatment_interactions.Rnw")
+tools::texi2dvi("treatment_interactions.tex", pdf=TRUE, clean=T)

### 2015-08-28 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@763 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \vspace{5pt}
-\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
+\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
 
 \vspace{5pt}
-\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$ using $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
+\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
 

### 2015-08-28 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@764 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0, with $m$ ($m=1,...,M$) denotes the cluster of which observation $i$ is part.
+Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,...,M$) denotes the cluster of which observation $i$ is part.
 $b$ is random vector of length $M$, each element being the random intercept for $m$.

### 2015-08-28 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@765 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \label{eq:fixedeffects_MOB}
-	g(\mu_{i}) = x_{i}^{\top}\beta_{j}
+	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
 \end{equation}
@@
 
-Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,...,M$) denotes the cluster of which observation $i$ is part.
-$b$ is random vector of length $M$, each element being the random intercept for $m$.
+Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,...,M$) denotes the cluster which observation $i$ is part of.
+Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$.
 Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.

### 2015-08-28 pkg/glmertree/paper/Rnw_2_tex_2_pdf.R

@@ -1,2 +1,0 @@
-Sweave("treatment_interactions.Rnw")
-tools::texi2dvi("treatment_interactions.tex", pdf=TRUE, clean=T)

### 2015-08-31 pkg/glmertree/paper/treatment_interactions.Rnw
Re-added height/width arguments in DGP tree plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@766 255f4249-2263-4417-b1ea-8498779e27a7

@@
 )
-plot(fig4, tnex=1.5)
+plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
 @
@@
 aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
-print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, type="b", ylab="tree size", xlab="sigma_b", par.settings=standard.theme("pdf",color=F), abline=c(1,0), auto.key=list(space="top", columns=2, title="Algorithm type", cex.title=1,lines=T, points=T)))
+print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
+	type="b", ylab="tree size", xlab="sigma_b", par.settings=standard.theme("pdf",color=F), 
+	abline=c(1,0), auto.key=list(space="top", columns=2, title="Algorithm type", cex.title=1, 
+	lines=T, points=T)))
 @
@@
 aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
-print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
+	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
+	abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
+)
 @
@@
 aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
-print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b",   ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
+	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
+	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
+)
 @
@@
 aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
-print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
+print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
+	type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
+	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
+)
 @

### 2015-09-03 pkg/glmertree/paper/treatment_interactions.Rnw
condensed title somewhat rephrased parts of intro (specficially mention advantages over existing methods) various cosmetic and style touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@767 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{lineno}
-\usepackage{tikz}
 \linespread{1.2}
@@
 
-\usetikzlibrary{shapes,shadows,arrows,positioning}
-\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)}
-\shorttitle{GLMM Tree}
-\leftheader{GLMM Tree}
-\rightheader{GLMM Tree}
+\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
+\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
+\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
+\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
 \author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
@@
 
-\acknowledgements{The authors would like to thank Prof.
-Pim Cuijpers, Prof.
-Jeanne Miranda, Dr.
-Boadie Dunlop, Prof.
-Rob DeRubeis, Prof.
-Zindel Segal, Dr.
-Sona Dimidjian, Prof.
-Steve Hollon and Erica Weitz for granting access to the dataset for the application.
-The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2013.
+\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
+The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.
 The visit was supported by the Institute.}
@@
 In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons.
-In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
-The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation.
-In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively).
-Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively).
-Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively).
+In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
+The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and GLMMs for estimation of random-effects parameters.
+In a simulation study, we evaluate the performance of GLMM trees and compare it with that of MOB trees without random-effects estimation.
+In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively).
+Furthermore, in datasets with treatment-subgroup interactions, GLMM trees recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively).
+Also, GLMM trees predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively).
 We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
+We conclude that GLMM trees are a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
 \\
@@
 
-Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments.
+Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments.
 Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
 Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
-Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates.
-The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM).
-Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.
+Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}.
+Specifically, this model class encompasses the generalized linear model (GLM) and GLM-based MOB trees have been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.
 
-However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets.
-In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}.
+However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of many datasets.
+In such cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}.
 For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled.
 In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
-Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects.
-Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}.
+Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects.
+Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}.
 More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
@@
 In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account.
-The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials).
-In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM).
-Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree.
-Subsequently, we evaluate the performances of the GLMM-tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
+The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
 
+In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM).
+Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees.
+Subsequently, we evaluate the performances of the GLMM tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
+%
 Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
-After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
+After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
 Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.
@@
 
-To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 or Treatment 2.
-Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
-Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).
+To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2.
+Every observation has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
+Further, all observations have values for three covariates: duration of depressive symptoms prior to treatment in months (range 0--15); age in years (range 18--75); anxiety inventory total score (range 3--18).
 
@@
 The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$.
-In this subgroup, Treatment 1 is more effective than Treatment 2: the mean of the response variable is 7 for Treatment 1, and 11 for Treatment 2.
+In this subgroup, Treatment~1 is more effective than Treatment~2: the mean of the response variable is 7 for Treatment~1, and 11 for Treatment~2.
 The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$.
-In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment 1, and 9 for Treatment 2.
+In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment~1, and 9 for Treatment~2.
 The third subgroup consists of observations with duration $>6$.
-In this subgroup, Treatment 2 is more effective than Treatment 1: the mean value of the response variable is 12 for Treatment 1, and 7 for Treatment 2.
+In this subgroup, Treatment~2 is more effective than Treatment~1: the mean value of the response variable is 12 for Treatment~1, and 7 for Treatment~2.
 
-Observations were drawn from one of ten clusters, each with a different, cluster-specific (i..e, random) intercept.
-Data was generated such, that covariates and cluster-specific intercepts were uncorrelated.
+Observations were drawn from one of ten clusters, each with a different, cluster-specific (i.e., random) intercept.
+Data was generated such that covariates and cluster-specific intercepts were uncorrelated.
 Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.
@@
 
-To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}.
-The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
+To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure~\ref{fig:fixedeffects}.
+The boxplots in Figure~\ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
 There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero.
@@
 @
-\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150).
-The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
+\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset ($N=150$).
+The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
 \label{fig:fixedeffects}
@@
 
-The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
+The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
 This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
@@
 Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree.
-If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,...,J$) terminal node of the tree.
-As a result, in the GLM-tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
+If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,\dots,J$) terminal node of the tree.
+As a result, in the GLM tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
 
@@
 
-Figure \ref{fig:example_mobtree} provides an example of the GLM-tree model in Equation \ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset.
+Figure~\ref{fig:example_mobtree} provides an example of the GLM tree model in Equation~\ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset.
 By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
@@
 \subsection{GLMM}
-When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
+When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
 
@@
 
-Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,...,M$) denotes the cluster which observation $i$ is part of.
+Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
 Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$.
@@
 As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects.
-Therefore, we propose the GLMM tree, which combines the GLMM from Equation \ref{eq:mixedeffects} with the tree from Equation \ref{eq:fixedeffects_MOB}:
+Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}:
 
@@
 In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
-The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:
+The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
 
@@
 \vspace{5pt}
-\noindent Step 3: Repeat steps 1 and 2 until convergence.
+\noindent Step 3: Repeat Steps~1 and~2 until convergence.
 
@@
 The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
-Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.
+Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.
 
@@
 @
-\caption{Generalized Linear Mixed Model tree of the motivating example dataset.
+\caption{GLMM tree of the motivating example dataset.
 Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
@@
 
-In Figure \ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented.
+In Figure~\ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented.
 As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.
@@
 \subsubsection{Datasets with treatment-subgroup interactions}
-For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure \ref{fig:modelC}.
-Figure \ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes.
+For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}.
+Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes.
 The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$.
@@
 \caption{Data-generating model for treatment-subgroup interactions.
-Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
+Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
 \label{fig:modelC}
@@
 
-For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in 50 x 3 x 2 x 2 x 3 x 3 x 3 x 2 = 32,400 training datasets.
+For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets.
 For the datasets without treatment-subgroup interactions, the 6th parameter of the data-generating process had only two levels ($b$ correlated with one of the $U_k$ variables, and $b$ not correlated with any of the $U_k$ variables).
-Therefore, 50 x 3 x 2 x 2 x 3 x 3 x 2 x 2 = 21,600 datasets without treatment-subgroup interactions were generated.
+Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.
 
@@
 
-As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively.
+As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, $-40$, and 70, respectively.
 The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
@@
 
-We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure \ref{fig:modelC}) for test observations.
+We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) for test observations.
 Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
@@
 \verb|R| \cite{R14} was used for generation and analysis of all datasets.
-Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
+Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
 
@@
 For estimating GLM trees, the \verb|lmtree| function from the \verb|partykit| package was used.
-The \verb|lmtree| function builds a linear model tree: a GLM-based recursive partition (Equation \ref{eq:fixedeffects_MOB}) for a real-valued response variable with normally distributed errors.
+The \verb|lmtree| function builds a linear model tree: a GLM-based recursive partition (Equation~\ref{eq:fixedeffects_MOB}) for a real-valued response variable with normally distributed errors.
 Models in the nodes of the tree are estimated with ordinary least squares (OLS).
@@
 
-For estimating GLMM trees, we implemented the GLMM-tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
+For estimating GLMM trees, we implemented the GLMM tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
 Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm.
@@
 
-For building the GLM trees in step (1) of the GLMM-tree algorithm, the same settings as described above were used.
+For building the GLM trees in step (1) of the GLMM tree algorithm, the same settings as described above were used.
 That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.
@@
 
-In Table \ref{tab:treesize_main}, tree sizes for GLM and GLMM trees for datasets without treatment-subgroup interactions are presented.
-Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 (SD=0.44) for GLMM tree, and 2.02 (SD=1.68) for GLM tree.
-The estimated probability that a dataset was erroneously partitioned was very small for GLMM tree (.04; Table \ref{tab:treesize_main}), and much larger for GLM tree (.33; Table \ref{tab:treesize_main}).
+In Table~\ref{tab:treesize_main}, tree sizes for GLM and GLMM trees for datasets without treatment-subgroup interactions are presented.
+Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
+The estimated probability that a dataset was erroneously partitioned was very small for GLMM tree (.04; Table~\ref{tab:treesize_main}), and much larger for GLM tree (.33; Table~\ref{tab:treesize_main}).
 
@@
 \begin{figure}[!htbp]
-<<treesize-nointeract-xyplot, fig=TRUE, width=8, height=7>>=
+<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
 load("treesizes_long_nointeract.dat")
@@
 print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
-	type="b", ylab="tree size", xlab="sigma_b", par.settings=standard.theme("pdf",color=F), 
-	abline=c(1,0), auto.key=list(space="top", columns=2, title="Algorithm type", cex.title=1, 
-	lines=T, points=T)))
+	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
+	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
+	lines=TRUE, points=TRUE)))
 @
@@
 
-A graphical display was used to asses the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_nointeract}).
+A graphical display was used to asses the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_nointeract}).
 When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1.
@@
 In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees.
-For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}).
-Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}.
-The average size of GLMM trees was 7.16 (SD=0.62), and the average size of GLM trees was 8.12 (SD=2.05).
+For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).
+Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table~\ref{tab:treesize_interactions}.
+The average size of GLMM trees was 7.16 ($\mathrm{SD}=0.62$), and the average size of GLM trees was 8.12 ($\mathrm{SD}=2.05$).
 The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree.
-However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).
+However, Table~\ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table~\ref{tab:treesize_interactions}).
 
@@
 
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_interact}).
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_interact}).
 When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average.
@@
 \begin{figure}[!htbp]
-<<treesize-xyplot, fig=TRUE, width=9, height=9.5>>=
+<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
 aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
@@
 GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset.
-The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree.
-However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree (SD=0.154 and SD=0.127, respectively).
+The true splitting value for $U_2$ was 30 (Figure~\ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree.
+However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).
 
@@
 
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure \ref{fig:xyplot_treeaccuracy}).
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure~\ref{fig:xyplot_treeaccuracy}).
 When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average.
@@
 \begin{figure}[!htbp]
-<<accuracy-xyplot, fig=TRUE, width=9, height=9.5>>=
+<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
 aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
 print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
-	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
-	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
+	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
+	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
 )
@@
 Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree.
-The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.19) for GLM tree, and .94 (SD=0.11) for GLMM tree.
+The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 ($\mathrm{SD}=0.19$) for GLM tree, and .94 ($\mathrm{SD}=0.11$) for GLMM tree.
 
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the predictive accuracy of both algorithms (Figure \ref{fig:xyplot_correlations}).
+A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the predictive accuracy of both algorithms (Figure~\ref{fig:xyplot_correlations}).
 Both algorithms showed higher predictive accuracy when sample size was larger, and when treatment-effect differences were larger.
@@
 \begin{figure}[!htbp]
-<<correlation-xyplot, fig=TRUE>>=
+<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
 aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
@@
 \caption{Average predictive accuracy of GLM and GLMM trees.
-Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment 1 and 2.
+Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2.
 Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
@@
 
-The trees resulting from application of GLM and GLMM tree to the dataset are presented in Figure \ref{fig:lmtree_C&W} and \ref{fig:lmertree_C&W}, respectively.
-Note that the GLM tree in Figure \ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMM-tree algorithm.
+The trees resulting from application of GLM and GLMM tree to the dataset are presented in Figure~\ref{fig:lmtree_C&W} and~\ref{fig:lmertree_C&W}, respectively.
+Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMM tree algorithm.
 
@@
 
-The GLM tree (Figure \ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-Node 2 of Figure \ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
+The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
+Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
 Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and antidepressant mediation.
@@
 
-By taking into account the study-specific intercepts, the final GLMM tree (Figure \ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
+By taking into account the study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
 The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable.
-The terminal nodes of Figure \ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores.
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores.
 The estimated intraclass correlation coefficient for the random intercepts was .05.
@@
 
-In the current paper, we presented the GLMM-tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters.
+In the current paper, we presented the GLMM tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters.
 The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions.
@@
 \NumTabs{12}
-\noindent $1,...,i,...,N$ \tab{observation number}\\
-$1,...,j,...,J$ \tab{terminal node number in a tree}\\
-$1,...,k,...,K$ \tab{partitioning variable number}\\
-$1,...,m,...,M$	\tab{cluster number}\\
+\noindent $1,\dots,i,\dots,N$ \tab{observation number}\\
+$1,\dots,j,\dots,J$ \tab{terminal node number in a tree}\\
+$1,\dots,k,\dots,K$ \tab{partitioning variable number}\\
+$1,\dots,m,\dots,M$	\tab{cluster number}\\
 $\beta_{j}$		\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
 $b_{m}$ 		\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
-$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment 1 and}\\
-\tab\tab\tab\tab{Treatment 2 in terminal node $j$}\\
+$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1 and}\\
+\tab\tab\tab\tab{Treatment~2 in terminal node $j$}\\
 $\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\

### 2015-09-04 pkg/glmertree/paper/treatment_interactions.Rnw
small touch-ups in modeling section -> more to do
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@773 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\begin{figure}[!ht]
+\begin{figure}[!th]
 \setkeys{Gin}{width=0.5\textwidth}
 <<exampleglm, fig=TRUE, height=5, width=5>>=
-plot(exampleglm)
+plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
 @
-\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset ($N=150$).
+\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$).
 The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
@@
 \verb|R| \cite{R14} was used for generation and analysis of all datasets.
-Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
+Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil15}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
 
@@
 
-\begin{table}[!htbp]
+\begin{table}[!thbp]
 	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!thbp]
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-\begin{table}[!htbp]
+\begin{table}[!thbp]
 \caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!thbp]
 <<treesize-xyplot, fig=TRUE, width=9, height=9>>=
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!thbp]
 <<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!thbp]
 <<correlation-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!th]
 \setkeys{Gin}{width=0.75\textwidth}
@@
 
-\begin{figure}[!ht]
-\setkeys{Gin}{width=0.75\textwidth}
-<<depression-glmmtree, fig=TRUE, height=4, width=6>>=
+\begin{figure}[!th]
+\setkeys{Gin}{width=0.6\textwidth}
+<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
 plot(lmertree_app$Tree)
@@
 Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 (var=.067) for GLMM tree, and .19 (var=.084) for GLM tree.
+The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathrm{var}=.067$) for GLMM tree, and .19 ($\mathrm{var}=.084$) for GLM tree.
 This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.
@@
 $r$			\tab\tab{iteration number}\\
-$\sigma_{b}$	\tab\tab{square root of variance of $b$}\\
-$\sigma_{\epsilon}$\tab\tab{square root of the variance of $\epsilon$}\\
+$\sigma_{b}$		\tab\tab{standard deviation of $b$}\\
+$\sigma_{\epsilon}$	\tab\tab{standard deviation of $\epsilon$}\\
 $U_k$			\tab\tab{(potential) partitioning variable $k$}\\

### 2015-09-05 pkg/glmertree/paper/treatment_interactions.Rnw
major revision of the glmertree package -> arguments and return values renamed to better match other functions -> streamlined code (e.g., special-casing of root-only tree should not    be necessary anymore) -> added S3 class with some basic methods -> collected example data in a single object with (somewhat) better name -> adapted paper accordingly
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@774 255f4249-2263-4417-b1ea-8498779e27a7

@@
 <<example-trees>>=
-exampleglm <- lmtree(outcome ~ treatment | age, data=example)
-exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data=example)
-exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data=example)
+exampleglm <- lmtree(outcome ~ treatment | age, data = example)
+exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
+exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data = example)
 @
@@
 <<exampleglmmtree, fig=TRUE, width=9, height=7>>=
-plot(exampleglmmtree$Tree)
+plot(exampleglmmtree)
 @
@@
 load("treesizes_long_nointeract.dat")
-treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
-treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
+treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
+treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
 aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
@@
 lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
- 		initialRandomEffects = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
+  ranefstart = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
 @
@@
 <<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
-plot(lmertree_app$Tree)
+plot(lmertree_app)
 @

### 2015-09-05 pkg/glmertree/R/glmertree.R

@@ -1,104 +1,137 @@
-lmertree <- function(lmtreeformula, randomformula, data, #subset = NULL, 
-  initialRandomEffects = NULL, 
-  ErrorTolerance = 0.001, MaxIterations = 1000, 
-  verbose = TRUE, plotting = FALSE,
-  lmer.control = lmerControl(), ...)
+lmertree <- function(formula, randomformula, data,
+  ranefstart = NULL, abstol = 0.001, maxit = 1000, 
+  verbose = FALSE, plot = FALSE, lmer.control = lmerControl(), ...)
 {
+  ## initialization
   iteration <- 0L
-  if (!is.null(initialRandomEffects)) { data$offset <- initialRandomEffects }
-  if (is.null(initialRandomEffects)) { data$offset <- rep(0, times=dim(data)[1L]) }
+  data$.ranef <- .ranefstart <- if (is.null(ranefstart)) {
+    rep(0, times = dim(data)[1L])
+  } else {
+    ranefstart  
+  }
   continuecondition <- TRUE
   oldloglik <- -Inf
+
+  ## iterate between lmer and lmtree estimation  
   while (continuecondition) {
     iteration <- iteration + 1L
-    tree <- lmtree(lmtreeformula, data, offset = offset, ...)
-    if(plotting){plot(tree)}
-    if(length(tree) > 1L) {
-      data$treeresponse <- treeresponse <- predict(tree, newdata = data)
-    }
-    if(length(tree) == 1L) { # if treedepth=1, predict.lmtree doesn't work, then use unpartitioned lm
-      lmformula <- formula(as.Formula(lmtreeformula), lhs = 1L, rhs = -2L)
-      data$treeresponse <- treeresponse <- predict(lm(lmformula, data = data), newdata = data)
-    }
-    lme <- lmer(randomformula, data = data, offset = treeresponse)
-    data$offset <- predict(lme, newdata = data)
+
+    ## lmtree
+    tree <- lmtree(formula, data, offset = .ranef, ...)
+    if(plot) plot(tree)
+    data$.treeresponse <- .treeresponse <- predict(tree, newdata = data)
+
+    ## lmer
+    lme <- lmer(randomformula, data = data, offset = .treeresponse)
+    data$.ranef <- .ranef <- predict(lme, newdata = data)
+
+    ## iteration information
     newloglik <- logLik(lme)    
-    continuecondition <- (newloglik - oldloglik > ErrorTolerance) & (iteration < MaxIterations) 
[…]

### 2015-09-06 pkg/glmertree/paper/treatment_interactions.Rnw
adapted R code to latest glmertree version -> still need to think about offset in empirical example
revised GLM-related descriptions somewhat



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@778 255f4249-2263-4417-b1ea-8498779e27a7

@@
 exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
-exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data = example)
+exampleglmmtree <- lmertree(outcome ~ treatment | cluster | age + duration + anxiety, data = example)
 @
@@
 \subsection{GLM}
-In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects.
-The goal is to estimate a model for predicting the value of a treatment outcome, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:
-
-\begin{equation}
+In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM is often used to estimate treatment effects.
+GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson - depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively.
+In all cases the expectation $\mu_i$ of the outcome variable $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and a suitable link function:
+%
+\begin{eqnarray}
 \label{eq:expected_value}
-	E[y_i | x_i] = \mu_i 
-\end{equation}
-
-\begin{equation}
+	E[y_i | x_i] & = & \mu_i, \\
 \label{eq:fixedeffects}
-	g(\mu_{i}) = x_{i}^{\top}\beta 
-\end{equation}
-
-Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function.
-In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$.
+	g(\mu_{i}) & = & x_{i}^{\top}\beta,
+\end{eqnarray}
+%
+where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function\footnote{An overview of notation used is provided in the appendix.}.
 Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type).
 $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.
+In case of a continuous response variable, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.
 
@@
 
-To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure~\ref{fig:fixedeffects}.
-The boxplots in Figure~\ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
-There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero.
-We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.
-
-\begin{figure}[!th]
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.5\textwidth}
@@
 \end{figure}
+
+To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure~\ref{fig:fixedeffects}.
+The boxplots in Figure~\ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups.
+There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero.
+We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.
 
@@
 Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree.
-If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,\dots,J$) terminal node of the tree.
+If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node of the tree.
 As a result, in the GLM tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
-
+%
 \begin{equation}
@@
 \end{equation}
-
-Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM.
-The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
+%
+Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the factor indicating the subgroup.
+Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$.
 $\beta^{*}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.
 
-Figure~\ref{fig:example_mobtree} provides an example of the GLM tree model in Equation~\ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset.
-By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
-Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above.
-However, the two subgroups in the middle result from a spurious split.
-
-\begin{figure}[!h]
+\begin{figure}[t!]
 \setkeys{Gin}{width=\textwidth}
@@
 
+Figure~\ref{fig:example_mobtree} provides an example of the GLM tree model in Equation~\ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset.
+By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$.
+Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above.
+However, the two subgroups in the middle result from a spurious split.
+
 \subsection{GLMM}
 When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
-
+%
 \begin{equation}
@@
 \end{equation}
-
-Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
+%
+Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of.
 Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$.
@@
 
-For simplicity, we assume that only cluster-specific intercepts are included in the models.
-However, random-effects covariates and coefficients can easily be included.
-Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.
+For simplicity, we assume that only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included.
+
+Note that, if the random-effects coefficients were known, the model could be estimated by a simple GLM as in Equation \ref{eq:fixedeffects} where $z_{i}^{\top}b$ would only be added as an offset (i.e., a variable with a fixed coefficient of 1) to the linear predictor.
 
@@
 Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}:
-
+%
 \begin{equation}
@@
 \end{equation}
-
-To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}.
-They developed a method for the estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees.
+%
+To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
 In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
@@
 
-For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM.
-This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
-In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
-The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
-
-\vspace{5pt}
-\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
-
-\vspace{5pt}
-\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
-
-\vspace{5pt}
-\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
-
-\vspace{5pt}
-\noindent Step 3: Repeat Steps~1 and~2 until convergence.
-
-\vspace{5pt}
-The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
-In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
-The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
-Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.
-
-\begin{figure}[!h]
+\begin{figure}[t!]
 <<exampleglmmtree, fig=TRUE, width=9, height=7>>=
@@
 \end{figure}
+
+For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM.
+This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes.
+In addition, GLMM trees can be estimated for continuous, as well as binary and count variables.
+The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
+
+\begin{description}
+  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
+
+  \item[Step 1:] Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.
+
+  \item[Step 2:] Estimate random effects in the mixed effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
+  % Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
+
+  \item[Step 3:] Repeat Steps~1 and~2 until convergence.
+\end{description}
+
+The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
+In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
+The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
+Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.8\textwidth}
@@
 
-\begin{table}[!thbp]
+\begin{table}[t!]
 	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
@@
 
-\begin{figure}[!thbp]
+\begin{figure}[t!]
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-\begin{table}[!thbp]
+\begin{table}[t!]
 \caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
@@
 
-\begin{figure}[!thbp]
+\begin{figure}[t!]
 <<treesize-xyplot, fig=TRUE, width=9, height=9>>=
@@
 
-\begin{figure}[!thbp]
+\begin{figure}[t!]
 <<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
@@
 
-\begin{figure}[!thbp]
+\begin{figure}[t!]
 <<correlation-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
+To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we apply both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
 This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
@@
 
-In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
-The predictor variable in the linear model was treatment type (0=CBT and 1=PHA).
+In our analyses, posttreatment HAM-D score is the outcome variable, and potential partitioning variables are age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
+The predictor variable in the linear model was treatment type (0 = CBT and 1 = PHA).
 An indicator for study was used as the cluster indicator.
@@
 As a result, data from 694 patients from 7 studies were included for the analyses.
-Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \cite{CuijyWeit14}.
+Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.
 
@@
 <<depression-trees>>=
-formula <- HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0
+hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
 
-## GLM tree
-lmtree_app <- lmtree(formula, data = metadata,
-  offset = fitted(lm(HRSDt1 ~ HRSDt0), data = metadata))
+## GLM tree (estimated via lmertree to determine offset iteratively)
+## lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+##   data = metadata, offset = hrs)
+lmtree_app <- lmertree(HRSDt1 ~ Tx_group | HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+  data = metadata, ranefstart = hrs, joint = FALSE)
 
-## GLMM tree
-lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
-  ranefstart = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
+## GLMM tree (estimated with joint = FALSE because fixed effect HRSDt0 is included in random formula)
+lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+  data = metadata, ranefstart = hrs, joint = FALSE)
 @
@@
 
-\begin{figure}[!th]
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.75\textwidth}
@@
 \caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.
-pharmacotherapy (PHA).}
+The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
 \label{fig:lmtree_C&W}
@@
 
-\begin{figure}[!th]
+\begin{figure}[t!]
 \setkeys{Gin}{width=0.6\textwidth}
@@
 \caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.
-pharmacotherapy (PHA).}
+The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
 \label{fig:lmertree_C&W}

### 2015-09-07 pkg/glmertree/paper/treatment_interactions.Rnw
added workaround for the situation where the random effects part also contains fixed effects: instead of omitting only the random effects from the offset prediction now the coefficients from the .tree are set to 0
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@780 255f4249-2263-4417-b1ea-8498779e27a7

@@
 lmtree_app <- lmertree(HRSDt1 ~ Tx_group | HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, ranefstart = hrs, joint = FALSE)
+  data = metadata, ranefstart = hrs)
 
@@
 lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, ranefstart = hrs, joint = FALSE)
+  data = metadata, ranefstart = hrs)
 @

### 2015-09-07 pkg/glmertree/R/glmertree.R

@@ -50,10 +50,15 @@ lmertree <- function(formula, data,
 
     ## lmer
     if(joint) {
+      ## estimate full lmer model but force all coefficients from the
+      ## .tree (and the overall intercept) to zero for the prediction
       lme <- lmer(rf, data = data)
-      data$.ranef <- predict(lme, newdata = data, re.form = NULL) -
-        predict(lme, newdata = data, re.form = NA)
+      b <- lme@beta
+      b[which(substr(names(coef(lme)[[1L]]), 1L, 5L) %in% c("(Inte", ".tree"))] <- 0
+      data$.ranef <- suppressWarnings(suppressMessages(predict(lme, newdata = data, newparams = list(beta = b))))
     } else {
+      ## estimate only a partial lmer model using the .tree fitted
+      ## values as an offset
       lme <- lmer(rf, data = data, offset = .tree)
       data$.ranef <- predict(lme, newdata = data)    
     }
@@ -140,10 +145,15 @@ glmertree <- function(formula, data, family = "binomial",
 
     ## glmer
     if(joint) {
+      ## estimate full glmer model but force all coefficients from the
+      ## .tree (and the overall intercept) to zero for the prediction
       glme <- glmer(rf, data = data, family = family)
-      data$.ranef <- predict(glme, newdata = data, type = "link", re.form = NULL) -
-        predict(glme, newdata = data, type = "link", re.form = NA)
+      b <- glme@beta
+      b[which(substr(names(coef(glme)[[1L]]), 1L, 5L) %in% c("(Inte", ".tree"))] <- 0
+      data$.ranef <- suppressWarnings(suppressMessages(predict(glme, newdata = data, type = "link", newparams = list(beta = b))))
     } else {
+      ## estimate only a partial glmer model using the .tree fitted
+      ## values as an offset
       glme <- glmer(rf, data = data, family = family, offset = .tree)
       data$.ranef <- predict(glme, newdata = data, type = "link")
     }

### 2015-09-07 pkg/glmertree/paper/treatment_interactions.Rnw
small improvement in joint GLMM estimation
added R-Forge glmertree reference to manuscript
-> updated the "Software" section somewhat



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@781 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-\verb|R| \cite{R14} was used for generation and analysis of all datasets.
-Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil15}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
+R \cite{R14} was used for generation and analysis of all datasets.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| would be available for GLM trees with other response distributions.
+For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed using restricted maximum likelihood (REML) estimation.
 
-\subsubsection{Estimation of GLM trees}
+For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
+This provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.
 
-For estimating GLM trees, the \verb|lmtree| function from the \verb|partykit| package was used.
-The \verb|lmtree| function builds a linear model tree: a GLM-based recursive partition (Equation~\ref{eq:fixedeffects_MOB}) for a real-valued response variable with normally distributed errors.
-Models in the nodes of the tree are estimated with ordinary least squares (OLS).
-The $\alpha$-level for assessing parameter instability was set to .05, a Bonferroni correction for multiple testing was applied, and the minimum number of observations in a node was set to 20.
-Maximum tree depth was set to four (i.e., maximum of eight terminal nodes), as this yields a model tree which is easy to interpret.
-
-\subsubsection{Estimation of GLMM trees}
-
-For estimating GLMM trees, we implemented the GLMM tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function.
-Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm.
-When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|glmmtree| has converged.
-
-For building the GLM trees in step (1) of the GLMM tree algorithm, the same settings as described above were used.
-That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.
-
-For estimating the $b_m$ values in step (2), the \verb|lmer| function from the \verb|lme4| package was used.
-This function estimates a linear mixed-effects model using maximum likelihood (ML) or restricted ML (REML).
-In the current paper, REML estimation was used.
+In all applications, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05 with a Bonferroni correction applied for multiple testing.
+The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.
+The iteration in the GLMM tree algorithm was repeated until the log-likelihood changes between two subsequent iterations fell below .001.
 

### 2015-09-07 pkg/glmertree/R/glmertree.R

@@ -53,8 +53,8 @@ lmertree <- function(formula, data,
       ## estimate full lmer model but force all coefficients from the
       ## .tree (and the overall intercept) to zero for the prediction
       lme <- lmer(rf, data = data)
-      b <- lme@beta
-      b[which(substr(names(coef(lme)[[1L]]), 1L, 5L) %in% c("(Inte", ".tree"))] <- 0
+      b <- structure(lme@beta, .Names = names(coef(lme)[[1L]]))
+      b[substr(names(b), 1L, 5L) %in% c("(Inte", ".tree")] <- 0
       data$.ranef <- suppressWarnings(suppressMessages(predict(lme, newdata = data, newparams = list(beta = b))))
     } else {
       ## estimate only a partial lmer model using the .tree fitted
@@ -148,8 +148,8 @@ glmertree <- function(formula, data, family = "binomial",
       ## estimate full glmer model but force all coefficients from the
       ## .tree (and the overall intercept) to zero for the prediction
       glme <- glmer(rf, data = data, family = family)
-      b <- glme@beta
-      b[which(substr(names(coef(glme)[[1L]]), 1L, 5L) %in% c("(Inte", ".tree"))] <- 0
+      b <- structure(glme@beta, .Names = names(coef(glme)[[1L]]))
+      b[substr(names(b), 1L, 5L) %in% c("(Inte", ".tree")] <- 0
       data$.ranef <- suppressWarnings(suppressMessages(predict(glme, newdata = data, type = "link", newparams = list(beta = b))))
     } else {
       ## estimate only a partial glmer model using the .tree fitted

### 2015-09-15 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@783 255f4249-2263-4417-b1ea-8498779e27a7

@@
 The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown.
-In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated.
+In every iteration, the GLM tree and random-effects coefficients $b$ are re-estimated.
 The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
@@
 Test datasets were generated from the same population as the training datasets.
-Because the cluster-specific intercepts $b_m$ were randomly generated for training as well as test datasets, test observations were from 'new' clusters.
+Because the cluster-specific intercepts $b$ were randomly generated for training as well as test datasets, test observations were from 'new' clusters.
 As a result, a model without random effects was used for prediction with GLMM tree.

### 2015-09-20 pkg/glmertree/paper/treatment_interactions.Rnw
- Removed frequency tables for tree sizes - Shortened some of the subsections somewhat
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@793 255f4249-2263-4417-b1ea-8498779e27a7

@@
 In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
-The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and GLMMs for estimation of random-effects parameters.
-In a simulation study, we evaluate the performance of GLMM trees and compare it with that of MOB trees without random-effects estimation.
-In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively).
-Furthermore, in datasets with treatment-subgroup interactions, GLMM trees recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively).
-Also, GLMM trees predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively).
+The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters.
+In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation.
+GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively).
+Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively).
+Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively).
 We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-We conclude that GLMM trees are a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
+We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
 \\
@@
 
-In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified medicine is becoming increasingly important.
+In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important.
 Stratified medicine presents the challenge of finding which patients respond best to which treatments.
@@
 Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
-Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}.
-Specifically, this model class encompasses the generalized linear model (GLM) and GLM-based MOB trees have been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.
+Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}.
+Specifically, this model class encompasses the generalized linear model (GLM).
+GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies.
+Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.
 
-However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of many datasets.
-In such cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}.
-For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled.
+In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure.
+However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset.
+For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
 In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
@@
 Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees.
-Subsequently, we evaluate the performances of the GLMM tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
-%
-Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated.
-After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy.
-Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.
+Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets.
+In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}.
+But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.
 
@@
 
-To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2.
+We created a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2.
 Every observation has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory.
@@
 The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration.
-Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.
+Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.
 
@@
 
-\section{Emperical evaluation}
+\section{Empirical evaluation}
 
@@
 In addition, we will compare the performance of GLMM tree with that of GLM tree.
-In the simulation study, the main interest will be in the effects of sample size, and the presence and magnitude of treatment-subgroup interactions and random effects, but other parameters will be varied, as well.
+In the simulation study, our main interest will be in the effects of sample size, the presence and magnitude of treatment-subgroup interactions, and the presence and magnitude of the random effects, but additonal parameters will be varied as well.
 
-For GLMM tree, we expect the accuracy of recovered trees and predictions to improve with increasing sample size, and magnitude of the differences in treatment outcomes.
-For GLM tree, we have the same expectation, when random effects are absent; that is, when the variance of the random coefficients is zero, we expect GLM tree and GLMM tree to perform equally well.
-When random effects are present, we expect GLMM tree to perform better than GLM tree, and more so when the variance of random-effects coefficients is larger.
+When random effects are absent from the datasets, we expect GLM and GLMM tree to perform equally well.
+In the presence of random effects, we expect GLMM tree to outperform GLM tree.
+We expect the difference in performance between both algorithms to increase, with increasing differences in treatment outcomes, increasing variance of random-effects coefficients, and/or increasing sample sizes.
 
@@
 The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$.
-In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.
+That is, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.
 
@@
 \item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
-\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42).
+\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition.
+However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}.
 \item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects.
@@
 Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.
-
-It should be noted that when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition.
-However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.
 
@@
 To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design.
-For each cluster we drew a single $b_m$ from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
+For each cluster, we drew a single value $b_m$ from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design.
 When $b$ was correlated with one of the potential partitioning variables, the correlated potential partitioning variable was randomly selected.
@@
 
-For every dataset, accuracy and size of the GLM and GLMM tree was evaluated.
-We calculated the total number of nodes in every tree, and compared it with the true tree size.
-For datasets without treatment-subgroup interactions, this allowed us to assess tree accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned.
-For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups.
+For every dataset, the total number of nodes in the resulting GLM and GLMM tree were calculated.
+For datasets without treatment-subgroup interactions, this allowed us to assess tree accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned (i.e., a tree of size $>1$ is created).
+For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups (i.e., a tree of size $> 7$ is created).
 
-For datasets with treatment-subgroup interactions, we assessed the accuracy of the trees created by GLM and GLMM tree.
-An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., total number of nodes equals 7), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
+For datasets with treatment-subgroup interactions, we assessed the accuracy of the GLM and GLMM trees.
+An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., tree size $= 7$), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$.
 Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).
 
-To assess the impact of data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+To assess the effects of the data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.
+The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance was further assessed using graphical displays.
 
-To assess the impact of data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
+To assess the effects of the data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The impact of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further investigated using graphical displays.
+The effects of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further assessed using graphical displays.
 
@@
 
-We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) for test observations.
+We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between the true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}).
 Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.
@@
 For every dataset, correlation coefficients for each algorithm were calculated, representing the linear association between the true and predicted treatment-effect differences.
-To assess the impact of data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
+To assess the effects of the data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables.
 In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables.
-The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.
+The effects of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.
 
@@
 R \cite{R14} was used for generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| would be available for GLM trees with other response distributions.
-For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed using restricted maximum likelihood (REML) estimation.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions.
+For other response distributions, the \verb|glmtree| function would be available.
+For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
 
@@
 
-In all applications, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05 with a Bonferroni correction applied for multiple testing.
+In all applications, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing.
 The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.
-The iteration in the GLMM tree algorithm was repeated until the log-likelihood changes between two subsequent iterations fell below .001.
+Iterations of the GLMM tree algorithm were repeated until the log-likelihood changes between two subsequent iterations fell below .001.
 
@@
 
-In Table~\ref{tab:treesize_main}, tree sizes for GLM and GLMM trees for datasets without treatment-subgroup interactions are presented.
 Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree.
-The estimated probability that a dataset was erroneously partitioned was very small for GLMM tree (.04; Table~\ref{tab:treesize_main}), and much larger for GLM tree (.33; Table~\ref{tab:treesize_main}).
+The estimated probability that a dataset was erroneously partitioned was .04 for GLMM tree, and .33 for GLM tree.
 
-\begin{table}[t!]
-	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
-	\small
-	\begin{tabular}{lrrrrrrr}
-		\thickline
-		&\multicolumn{7}{c}{tree size}\\
-		\cline{2-7}
-		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
-		\hline
-		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
-			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
-   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
-    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
-		\hline
-  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
-  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
-  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
-	\end{tabular}
-	\label{tab:treesize_main}
-\end{table}
+The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on tree size were assessed with a graphical display (Figure~\ref{fig:xyplot_treesize_nointeract}).
+When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tended to create trees of size 1.
+In the presence of random effects, GLMM tree also tended to create trees of size 1, but GLM tree created much larger trees, when $b$ was correlated with one of the $U_k$ variables.
+This effect was stronger when sample size was larger.
 
@@
 
-A graphical display was used to asses the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_nointeract}).
-When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1.
-In the presence of random effects, GLMM tree also tends to create trees of size 1, but GLM tree created much larger trees, when $b$ was correlated to one of the $U_k$ variables.
-This effect was stronger when sample size was larger.
-
 \subsubsection{Tree size in datasets with treatment-subgroup interactions}
@@
 For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).
-Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table~\ref{tab:treesize_interactions}.
 The average size of GLMM trees was 7.16 ($\mathrm{SD}=0.62$), and the average size of GLM trees was 8.12 ($\mathrm{SD}=2.05$).
 The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree.
-However, Table~\ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table~\ref{tab:treesize_interactions}).
+However, a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size.
 
-\begin{table}[t!]
-\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
-\small
-	\begin{tabular}{lrrrrrrrr}
-		\thickline
-		&&\multicolumn{7}{c}{tree size}\\
-		\cline{2-8}
-		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
-		\hline
-	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
-					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\
-
-	  	GLM tree	& 139 & 955 & 20530 & 4565 & 4164 & 1619 & 428 & 32,400\\ 
-	 			 	& ($<.01$) & (.03) & (.63) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
-		\hline
-  		\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
-  		\multicolumn{9}{l}{nodes in the tree. A tree with a total of $J$ nodes has $(J+1)/2$ terminal nodes; the true}\\	
-  		\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
-	\end{tabular}
-	\label{tab:treesize_interactions}
-\end{table}
-
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_interact}).
+The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on tree size were assessed with a graphical display (Figure~\ref{fig:xyplot_treesize_interact}).
 When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average.
+However, clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$.
+When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average.
+When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$).
+When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
+This effect was stronger when $b$ was correlated to a non-splitting variable.
 
@@
 
-Clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$.
-When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average.
-When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$).
-When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).
-This effect was stronger when $b$ was correlated to a non-splitting variable.
-
 \subsubsection{Tree accuracy in datasets with treatment-subgroup interactions}
@@
 GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset.
-The true splitting value for $U_2$ was 30 (Figure~\ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree.
+The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).
 However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).
 
-Overall, GLMM tree performed well in recovering treatment-subgroup interactions, accurately recovering the tree in 90.19\% of datasets.
-GLM tree accurately recovered the treatment-subgroup interactions in only 61.44\% of datasets.
-
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure~\ref{fig:xyplot_treeaccuracy}).
-When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average.
-In the presence of random effects, GLM trees were much less accurate than GLMM trees.
-This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables, and the effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable.
-When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).
+Overall, GLMM tree performed well in recovering treatment-subgroup interactions, by accurately recovering the tree in 90.19\% of datasets.
+GLM tree performed less accurate, by accurately recovering the tree in 61.44\% of datasets.
 
@@
 
-\subsection{Predictive accuracy on test data}
+The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on the probability of accurate tree recovery for GLM and GLMM tree were assessed with a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).
+When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average.
+In the presence of random effects, GLM trees were much less accurate than GLMM trees.
+This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables, and the effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable.
+When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).
 
-To assess predictive accuracy of both algorithms, correlation between the true and predicted treatment-effect differences of both algorithms were calculated for every dataset.
-Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree.
-The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 ($\mathrm{SD}=0.19$) for GLM tree, and .94 ($\mathrm{SD}=0.11$) for GLMM tree.
+\subsubsection{Predictive accuracy on test data}
 
-A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the predictive accuracy of both algorithms (Figure~\ref{fig:xyplot_correlations}).
-Both algorithms showed higher predictive accuracy when sample size was larger, and when treatment-effect differences were larger.
-When random effects were absent from the datasets (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were equally accurate.
-In the presence of random effects, GLM tree predictions were always much less accurate than those of GLMM tree.
-This effect was stronger when $\sigma_b$ was larger, sample size was larger, and/or treatment-effect differences were larger.
+To assess predictive accuracy, correlations between the true and predicted treatment-effect differences of both algorithms were calculated for every dataset.
+Overall, predicted treatment-effect differences of GLMM tree were closer to the true differences, than those of GLM tree.
+The average correlation between the true and predicted treatment-effect differences over all datasets with treatment-subgroup interactions was .94 ($\mathrm{SD}=0.11$) for GLMM tree, and .88 ($\mathrm{SD}=0.19$) for GLM tree.
 
@@
 \end{figure}
+
+The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on the predictive accuracy of both algorithms were assessed with a graphical display (Figure~\ref{fig:xyplot_correlations}).
+Both algorithms showed higher predictive accuracy when sample size was larger, and when treatment-effect differences were larger.
+When random effects were absent from the datasets (i.e., $\sigma_b = 0$), predictions of GLM and GLMM tree were equally accurate.
+In the presence of random effects, GLM tree predictions were always much less accurate than those of GLMM tree.
+This effect was stronger when $\sigma_b$ was larger, sample size was larger, and/or treatment-effect differences were larger.
 
@@
 
-To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we apply both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
-This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
+To illustrate the application of the GLM and GLMM tree algorithms and the differences in their results, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
+This meta-analysis was based on a dataset, consisting of observations for individual patients from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
 The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression.
@@
 
-In our analyses, posttreatment HAM-D score is the outcome variable, and potential partitioning variables are age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
+In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score.
 The predictor variable in the linear model was treatment type (0 = CBT and 1 = PHA).
@@
 In RCTs, treatment effects are often estimated after controlling posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure.
-Therefore, we included the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores, as an offset variable in all models.
-An offset variable is a linear predictor with an a-priori determined coefficient of one.
-Including the linear regression predictions as an offset has the same effect as statistically controlling for the linear effects of pretreatment scores, as is often done in ANCOVA.
+Therefore, we included the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset variable in growing the GLM and GLMM trees.
+An offset variable is a linear predictor with an a-priori specified coefficient of one.
+Including the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset has the same effect as statistically controlling for the linear effects of pretreatment scores, as is often done in ANCOVA.
 
-We build all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score.
+We built all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score.
 As a result, data from 694 patients from 7 studies were included for the analyses.
@@
 
-The predictive accuracy of GLM and GLMM tree was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
+Predictive accuracy of GLM and GLMM trees was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
 
@@
 
-## GLM tree (estimated via lmertree to determine offset iteratively)
-## lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-##   data = metadata, offset = hrs)
-lmtree_app <- lmertree(HRSDt1 ~ Tx_group | HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, ranefstart = hrs)
+## GLM tree
+lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+  data = metadata, offset = hrs)
 
 ## GLMM tree (estimated with joint = FALSE because fixed effect HRSDt0 is included in random formula)
-lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, ranefstart = hrs)
+lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(hrs) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+  data = metadata, ranefstart = hrs, joint = F)
 @
@@
 
-The trees resulting from application of GLM and GLMM tree to the dataset are presented in Figure~\ref{fig:lmtree_C&W} and~\ref{fig:lmertree_C&W}, respectively.
-Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree that is created in the first iteration of the GLMM tree algorithm.
+Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W}.
+Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree in the first iteration of the GLMM tree algorithm.
 
@@
 Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
-Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and antidepressant mediation.
-Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for pharmacotherapy.
+Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and PHA.
+Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA.
 
@@
 
-By taking into account the study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
+By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
 The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable.
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and antidepressant medication provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, antidepressant medication provides a greater reduction in HAM-D scores.
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores.
 The estimated intraclass correlation coefficient for the random intercepts was .05.
@@
 Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathrm{var}=.067$) for GLMM tree, and .19 ($\mathrm{var}=.084$) for GLM tree.
+The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathit{var}=.067$) for GLMM tree, and .19 ($\mathit{var}=.084$) for GLM tree.
 This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.
@@
 In the current paper, we presented the GLMM tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters.
-The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions.
-In contrast, GLM tree recovered the true tree structure in only 61\% of the datasets with treatment-subgroup interactions.
+Therefore, the GLMM tree algorithm seems preeminently suited for the detection of treatment-subgroup interactions in clustered datasets.
+In the empirical evaluation, we have assessed the performance of GLMM tree in a large number of simulated dataets, and compared it with GLM tree, an algorithm that creates GLM-based recursive partitions, without estimating random effects.
+In the application, we applied GLM and GLMM tree to an existing dataset of a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+The application showed that GLMM tree provides results that are easily interpretable, and also more accurate than a GLM tree without random effects.
+
+The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, as it accurately recovered the interactions in 90\% of the simulated datasets with treatment-subgroup interactions.
+In contrast, GLM tree accurately recovered the interactions in only 61\% of the datasets with treatment-subgroup interactions.
 In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
@@
 
-Also, predictive accuracy of GLMM tree was higher than that of GLM tree.
+Predictive accuracy of GLMM tree was also higher than that of GLM tree.
 The average correlation between the true treatment differences and those predicted by GLMM tree was .94.
@@
 In the absence of random effects, GLM tree and GLMM tree are expected to perform equally well, and in the presence of random effects, GLMM tree will outperform GLM tree.
-This is especially the case with large sample sizes ($N > 200$), as the increased power will likely cause GLM tree to create spurious splits in the presence of random effects.
+This may especially be the case with large sample sizes ($N > 200$), as the increased power will likely cause GLM tree to create spurious splits, when random effects are present in the data.
 
 Not surprisingly, for both algorithms, accuracy of predicted treatment differences was less when sample size was low (i.e., $N=200$).
-Sample size influenced performance of GLM tree and GLMM tree similarly, suggesting that a larger number of estimated parameters for GLMM tree does not adversely influences accuracy at low sample sizes.
+Sample size influenced performance of GLM and GLMM tree similarly, suggesting that a larger number of estimated parameters for GLMM tree does not adversely influences accuracy at low sample sizes.
 Our simulation results do warrant some caution for the detection of treatment-subgroup interactions or treatment moderators in small datasets (e.g., single RCTs), but irrespective of the algorithm used.
@@
 However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified.
-That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then the algorithm will accurately recover those subgroups.
+That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups.
 However, misspecification of the model can reduce performance.
@@
 \NumTabs{12}
-\noindent $1,\dots,i,\dots,N$ \tab{observation number}\\
-$1,\dots,j,\dots,J$	\tab{terminal node number in a tree}\\
+\noindent $1,\dots,i,\dots,N$ \tab\tab{observation number}\\
+$1,\dots,j,\dots,J$	\tab\tab{terminal node number in a tree}\\
 $1,\dots,k,\dots,K$	\tab{partitioning variable number}\\
 $1,\dots,m,\dots,M$	\tab{cluster number}\\
-$\beta_{j}$		\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
-$b_{m}$ 		\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
-$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1 and}\\
-			\tab\tab\tab\tab{Treatment~2 in terminal node $j$}\\
-$\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
-$r$			\tab\tab{iteration number}\\
-$\sigma_{b}$		\tab\tab{standard deviation of $b$}\\
-$\sigma_{\epsilon}$	\tab\tab{standard deviation of $\epsilon$}\\
-$U_k$			\tab\tab{(potential) partitioning variable $k$}\\
-$x_i$ 			\tab\tab{column vector of fixed-effects predictor variable values for observation $i$}\\
-$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
-$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\
+$\beta_{j}$		\tab\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
+$b_{m}$ 		\tab\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
+$d_{j}$			\tab\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1}\\
+\tab\tab\tab\tab{and Treatment~2 in terminal node $j$}\\
+$\epsilon$		\tab\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
+$r$			\tab\tab\tab{iteration number}\\
+$\sigma_{b}$		\tab\tab\tab{standard deviation of $b$}\\
+$\sigma_{\epsilon}$	\tab\tab\tab{standard deviation of $\epsilon$}\\
+$U_k$			\tab\tab\tab{(potential) partitioning variable $k$}\\
+$x_i$ 			\tab\tab\tab{column vector of fixed-effects predictor variable values for}\\
+			\tab\tab\tab\tab{observation $i$}\\
+$y_i$ 			\tab\tab\tab{treatment outcome for observation $i$}\\
+$z_i$ 			\tab\tab\tab{column vector of random-effects predictor variable values for }\\
+			\tab\tab\tab\tab{observation $i$}\\
 

### 2015-09-20 pkg/glmertree/paper/Application with CV.R
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@795 255f4249-2263-4417-b1ea-8498779e27a7

@@ -1,0 +1,104 @@
+library(foreign)
+library(glmertree)
+
+# DATA PREPARATION
+metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
+metadata[metadata == 999] <- NA
+metadata[metadata == 888] <- NA
+vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
+factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
+metadata$education <- factor(metadata$education, ordered = T)
+for (i in 1:length(factors)) {metadata[,factors[i]] <- factor(metadata[,factors[i]])}
+metadata <- metadata[vars] # select only relevant variables
+metadata <- metadata[complete.cases(metadata[,vars]),] # select only complete data
+metadata <- metadata[!metadata$Tx_group == "placebo",] # remove placebo observations
+metadata$Tx_group <- factor(metadata$Tx_group)
+summary(metadata)
+
+## calculate trees
+hrsd <- lm(HRSDt1 ~ HRSDt0, data = metadata) 
+metadata$HRSDfit <- fitted(hrsd) 
+lm_f <- lmtree(HRSDt1 ~ Tx_group + offset(HRSDfit) | Age + Gender +
+                 education + ComorbidAnxietyDisorder + HRSDt0, data = metadata)
+lm_o <- lmtree(HRSDt1 ~ Tx_group | Age + Gender +
+                 education + ComorbidAnxietyDisorder + HRSDt0, offset = HRSDfit, data = metadata)
+## when joint=TRUE, prediction for random effects becomes cumbersome
+#lmer_f <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) |
+#                     Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+#                   data = metadata, ranefstart = metadata$HRSDfit)
+lmer_f <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) |
+                     Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+                   data = metadata, ranefstart = metadata$HRSDfit, joint = F)
+plot(lm_f); print(lm_f)
+plot(lm_o); print(lm_o)
[…]

### 2015-09-20 pkg/glmertree/paper/report.Rnw
added working paper version (report.Rnw, ref.bib) using JSS style
small touch-ups in application R code, joint=FALSE should not be necessary
in the current glmertree version



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@796 255f4249-2263-4417-b1ea-8498779e27a7

--- 2015-09-20 pkg/glmertree/paper/treatment_interactions.Rnw
@@
-\documentclass[nobf,doc]{apa}
-\usepackage{amssymb}
-\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
-\usepackage{apacite}
-\usepackage[english]{babel}
-\usepackage{lscape,dcolumn,hhline}
-\usepackage{graphicx,epic,eepic,rotating}
-\usepackage[normalem]{ulem}
-\usepackage{fancyhdr}
-\usepackage{eurosym,bbding}
-\usepackage{verbatim}
-\usepackage{ctable}
-\usepackage{multirow}
-\usepackage{rotating}
-\usepackage{lineno}
-\linespread{1.2}
-\usepackage{tabto}
+\documentclass[nojss]{jss}
+\usepackage{amssymb,amsmath,amsthm}
+
+\usepackage{thumbpdf}
+\shortcites{CuijyWeit14,DrieySmit14}
 
@@
 \title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
-\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
-\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
-\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
-\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
-\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}
+\Shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
 
-\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
-The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.
-The visit was supported by the Institute.}
+\author{Marjolein Fokkema\\Universiteit Leiden \And
+        Niels Smits\\Universiteit van Amsterdam \And
+	Achim Zeileis\\Universit\"{a}t Innsbruck \AND
+	Torsten Hothorn\\Universit\"{a}t Z\"{u}rich \And
+	Henk Kelderman\\Universiteit Leiden\\Vrije Universiteit, Amsterdam}
+\Plainauthor{Marjolein Fokkema, Niels Smits, Achim Zeileis, Torsten Hothorn, Henk Kelderman}
 
-\begin{document}
-
-\maketitle
-\pagewiselinenumbers
-
-\section{Abstract}
+\Abstract{
   Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
@@
 We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
-\\
-\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees
+}
+\Keywords{model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees}
 
-\section{Introduction}
+\Address{
+  Marjolein Fokkema\\
+  Universiteit Leiden\\
+  Faculteit der Sociale Wetenschappen\\
+  Instituut Psychologie, Methodologie \& Statistiek\\
+  Pieter de la Court gebouw\\
+  Wassenaarseweg 52\\
+  2333 AK Leiden, The Netherlands\\
+  E-mail: \email{m.fokkema@fsw.leidenuniv.nl}\\
+  % URL: \url{http://www.socialsciences.leiden.edu/psychology/organisation/staff/fokkemam.html}	
+}
+
+\begin{document}
+
+\section{Introduction} \label{sec:intro}
 
@@
 Stratified medicine presents the challenge of finding which patients respond best to which treatments.
-This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}.
+This can be referred to as the detection of treatment-subgroup interactions \citep[e.g.,][]{DoovyDuss14}.
 In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori.
 This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions.
-As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.
+As noted by \citet{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.
 
 Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments.
-Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
-Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
-Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}.
+Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \citep{StroyMall09}.
+Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citealp{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citealp{DoovyDuss14} for an overview).
+Among these methods, model-based recursive partitioning (MOB; \citealp{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \citep{ZeilyHoth08}.
 Specifically, this model class encompasses the generalized linear model (GLM).
-GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies.
-Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.
+GLM-based MOB has been successfully applied by \citet{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies.
+Also, \citet{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.
 
@@
 However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset.
-For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
-In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
+For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \citep[e.g.,][]{KoopyHeij07}.
+In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \citep[e.g.,][]{CoopyPatt09,HiggyWhit01}.
 Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects.
-Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}.
-More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
+Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \citep[e.g.,][]{BrykyRaud92,NooryOpde05}.
+More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \citep[e.g.,][]{SelaySimo12}.
 
 In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account.
-The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
+The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \citep[which is not done in other tree-based treatment-subgroup interaction detection methods, e.g.,][]{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \citep[which is not available in previously suggested regression trees with random effects, e.g.,][]{HajjyBell11,SelaySimo12}.
 
-In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM).
+In Section~\ref{sec:framework}, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM).
 Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees.
-Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets.
-In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}.
+Subsequently, in the Section~\ref{sec:evaluation}, we evaluate the performance of the GLMM tree algorithm in simulated datasets.
+In the Section~\ref{sec:application}, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citet{CuijyWeit14}.
 But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.
 
-\subsection{Artificial motivating dataset}
+\subsection{Artificial motivating dataset} \label{sec:data}
 
@@
 
-\section{General modeling framework}
+\section{General modeling framework} \label{sec:framework}
 
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.5\textwidth}
@@
 
-The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}.
-This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
+The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \citep{ZeilyHoth08}.
+This is reminiscent of the classification and regression tree (CART) algorithm of \citet{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different.
 However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.
@@
 When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor.
-To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
+To find partitions, the MOB algorithm cycles iteratively through the following steps \citep{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.
 
@@
 By definition, the empirical scores of all observations in a dataset sum to zero, and when the model is correctly specified, the expected value of the score for each observation is also zero.
-Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}.
-To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.
+Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \citep[c.f.,][]{MerkyZeil13}.
+To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \citep{Zeil05,ZeilyHorn07}.
 
@@
 In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2).
-The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}.
+The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \citep{ZeilyHoth08}.
 In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=\textwidth}
@@
 Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$.
-The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.
+The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citet{BrykyRaud92}, for example.
 
@@
 %
-To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
+To estimate the parameters of this model, we take an approach similar to that of \citet{HajjyBell11} and \citet{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
 In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
@@
 \begin{figure}[t!]
+\centering
 <<exampleglmmtree, fig=TRUE, width=9, height=7>>=
@@
 
-\section{Empirical evaluation}
+\section{Empirical evaluation} \label{sec:evaluation}
 
@@
 \subsubsection{Datasets with treatment-subgroup interactions}
-For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}.
+For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citet{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}.
 Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes.
@@
 \begin{figure}[t!]
+\centering
 \setkeys{Gin}{width=0.8\textwidth}
@@
 \item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects.
-The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
+The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citealp{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
 \end{enumerate}
@@
 
-As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, $-40$, and 70, respectively.
+As in \citet{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, $-40$, and 70, respectively.
 The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$.
@@
 
-Using the same data for training and evaluation of a model results in overly optimistic estimates of predictive accuracy \cite{HastyTibs09}.
+Using the same data for training and evaluation of a model results in overly optimistic estimates of predictive accuracy \citep{HastyTibs09}.
 Therefore, GLM and GLMM trees were used for prediction of new observations from test datasets.
@@
 
-R \cite{R14} was used for generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions.
-For other response distributions, the \verb|glmtree| function would be available.
-For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
+\proglang{R} \citep{R14} was used for generation and analysis of all datasets.
+The \pkg{partykit} package (version 1.0-2; \citealp{HothyZeil15}) was employed for estimating GLM trees using the \code{lmtree} function for normal linear regressions.
+For other response distributions, the \code{glmtree} function would be available.
+For estimation of GLMMs the \code{lmer} (or \code{glmer}, respectively) from the \pkg{lme4} package (version 1.1-7; \citealp{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
 
-For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
-This provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.
+For the estimation of GLMM trees the former two packages were combined in a new package \pkg{glmertree} (version 0.1-0; \citealp{FokkyZeil15}; available from R-Forge).
+This provides functions \code{lmertree} and \code{glmertree} that iterate between estimation of the \code{lmtree}/\code{glmtree} model and the \code{lmer}/\code{glmer} model.
 
@@
 \begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.75\textwidth}
 <<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
@@
 \begin{figure}[t!]
+\centering
+\setkeys{Gin}{width=0.74\textwidth}
 <<treesize-xyplot, fig=TRUE, width=9, height=9>>=
@@
 \begin{figure}[t!]
+\centering
 <<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
@@
 \begin{figure}[t!]
+\centering
 <<correlation-xyplot, fig=TRUE, width=9, height=7>>=
@@
 
-\section{Application to patient-level meta-analysis dataset on the effects of treatments for depression}
+\section{Patient-level meta-analysis of depression treatments} \label{sec:application}
 
@@
 
-To illustrate the application of the GLM and GLMM tree algorithms and the differences in their results, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}.
+To illustrate the application of the GLM and GLMM tree algorithms and the differences in their results, we applied both algorithms to a dataset from a meta-analytic study of \citet{CuijyWeit14}.
 This meta-analysis was based on a dataset, consisting of observations for individual patients from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression.
-The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression.
-Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}).
-\citeA{CuijyWeit14} found no indication that gender either predicted or moderated treatment outcomes.
-Further details on the dataset are provided in \citeA{CuijyWeit14}.
+The study of \citet{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression.
+Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citealp{Hami60}).
+\citet{CuijyWeit14} found no indication that gender either predicted or moderated treatment outcomes.
+Further details on the dataset are provided in \citet{CuijyWeit14}.
 
@@
 As a result, data from 694 patients from 7 studies were included for the analyses.
-Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.
+Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citet{CuijyWeit14}.
 
@@
 <<depression-trees>>=
-hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
+## offset
+metadata$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
 
@@
 lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, offset = hrs)
+  data = metadata, offset = HRSDfit)
 
-## GLMM tree (estimated with joint = FALSE because fixed effect HRSDt0 is included in random formula)
-lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(hrs) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-  data = metadata, ranefstart = hrs, joint = F)
+## GLMM tree
+lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
+  data = metadata, ranefstart = metadata$HRSDfit)
 @
@@
 
-Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W}.
-Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree in the first iteration of the GLMM tree algorithm.
-
-\begin{figure}[t!]
-\setkeys{Gin}{width=0.75\textwidth}
+\begin{figure}[p!]
+\centering
+\setkeys{Gin}{width=0.78\textwidth}
 <<depression-glmtree, fig=TRUE, height=6, width=8>>=
@@
 \caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
-\label{fig:lmtree_C&W}
+The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
+\label{fig:lmtree_C&W}}
+
+\setkeys{Gin}{width=0.6\textwidth}
+<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
+plot(lmertree_app)
+@
+\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
+The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).
+\label{fig:lmertree_C&W}}
 \end{figure}
+
+Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W}.
+Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree in the first iteration of the GLMM tree algorithm.
 
@@
 Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA.
-
-\begin{figure}[t!]
-\setkeys{Gin}{width=0.6\textwidth}
-<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
-plot(lmertree_app)
-@
-\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D).
-The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs.\ pharmacotherapy (PHA).}
-\label{fig:lmertree_C&W}
-\end{figure}
 
@@
 
-\section{Discussion}
+\section{Discussion} \label{sec:discussion}
 
@@
 
-\nolinenumbers
-\bibliographystyle{apacite}
-\bibliography{bib}
+\section*{Acknowledgments}
 
-\section{Appendix: Notation}
+The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
+The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.
+The visit was supported by the Institute.
 
-\NumTabs{12}
-\noindent $1,\dots,i,\dots,N$ \tab\tab{observation number}\\
-$1,\dots,j,\dots,J$	\tab\tab{terminal node number in a tree}\\
-$1,\dots,k,\dots,K$	\tab{partitioning variable number}\\
-$1,\dots,m,\dots,M$	\tab{cluster number}\\
-$\beta_{j}$		\tab\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
-$b_{m}$ 		\tab\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
-$d_{j}$			\tab\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1}\\
-\tab\tab\tab\tab{and Treatment~2 in terminal node $j$}\\
-$\epsilon$		\tab\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
-$r$			\tab\tab\tab{iteration number}\\
-$\sigma_{b}$		\tab\tab\tab{standard deviation of $b$}\\
-$\sigma_{\epsilon}$	\tab\tab\tab{standard deviation of $\epsilon$}\\
-$U_k$			\tab\tab\tab{(potential) partitioning variable $k$}\\
-$x_i$ 			\tab\tab\tab{column vector of fixed-effects predictor variable values for}\\
-			\tab\tab\tab\tab{observation $i$}\\
-$y_i$ 			\tab\tab\tab{treatment outcome for observation $i$}\\
-$z_i$ 			\tab\tab\tab{column vector of random-effects predictor variable values for }\\
-			\tab\tab\tab\tab{observation $i$}\\
+\bibliography{ref}
+
+\begin{appendix}
+\section{Notation}
+
+\begin{tabular}{lp{11cm}}
+$1,\dots,i,\dots,N$ & {observation number}\\
+$1,\dots,j,\dots,J$ & {terminal node number in a tree}\\
+$1,\dots,k,\dots,K$ & {partitioning variable number}\\
+$1,\dots,m,\dots,M$ & {cluster number}\\
+$\beta_{j}$	 & {column vector of fixed-effects coefficients in terminal node $j$}\\
+$b_{m}$ 	 & {column vector of random-effects coefficients in cluster $m$}\\
+$d_{j}$		 & {$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1 and Treatment~2 in terminal node $j$}\\
+$\epsilon$	 & {deviation of observed treatment outcome $y$ from its expected value}\\
+$r$		 & {iteration number}\\
+$\sigma_{b}$	 & {standard deviation of $b$}\\
+$\sigma_{\epsilon}$ & {standard deviation of $\epsilon$}\\
+$U_k$		 & {(potential) partitioning variable $k$}\\
+$x_i$ 		 & {column vector of fixed-effects predictor variable values for observation~$i$}\\
+$y_i$ 		 & {treatment outcome for observation $i$}\\
+$z_i$ 		 & {column vector of random-effects predictor variable values for observation~$i$}\\
+\end{tabular}
+
+\end{appendix}
 

### 2015-09-21 pkg/glmertree/paper/report.Rnw
added title for EW in acknowledgements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@798 255f4249-2263-4417-b1ea-8498779e27a7

@@
 
-The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application.
+The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz,~MA, for granting access to the dataset for the application.
 The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.

### 2015-09-28 pkg/glmertree/paper/report.Rnw
Added thank you to ED
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@801 255f4249-2263-4417-b1ea-8498779e27a7
two small corrections: - added U5 as part var in fig 4 (dgp tree) - MERT abbreviation introduced in introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@806 255f4249-2263-4417-b1ea-8498779e27a7

@@
 %
-To estimate the parameters of this model, we take an approach similar to that of \citet{HajjyBell11} and \citet{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
+To estimate the parameters of this model, we take an approach similar to the mixed-effects regression tree (MERT) algorithm of \citet{HajjyBell11} and \citet{SelaySimo12}, but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
 In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
@@
       partynode(5L,
-        split = partysplit(1L, breaks = 63),
+        split = partysplit(3L, breaks = 63),
         kids = list(
@@
 	  )))))),
-  data.frame(U1 = numeric(0), U2 = numeric(0))
+  data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
 )
@@
 The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz,~MA, for granting access to the dataset for the application.
+The authors would like to thank Dr.
+Elise Dusseldorp for providing helpful comments on an earlier version of the manuscript.
 The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014.

### 2016-05-10 pkg/glmertree/paper/treatment_interactions.Rnw
Paper revised
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@847 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{lineno}
+\usepackage{afterpage}
 \linespread{1.2}
@@
 Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine.
-Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions.
-In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons.
-In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects.
+Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions in datasets from randomized clinical trials.
+In many instances, however, datasets may have a clustered structure, where observations are nested within, for example, research centers, studies or subjects.
+In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of random effects.
 The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters.
@@
 
-In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important.
+In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified (or personalized) medicine is becoming increasingly important.
 Stratified medicine presents the challenge of finding which patients respond best to which treatments.
 This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}.
-In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori.
-This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions.
-As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.
+Often, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized.
+This may hamper identification of which treatment works best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions.
+As noted by \citeA{KreayFran06}, there is a need for methods that generate instead of test such hypotheses.
 
@@
 Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
-Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview).
-Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}.
-Specifically, this model class encompasses the generalized linear model (GLM).
-GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies.
-Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.
+Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview).
+Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes.
+As such, the \citeA{ZhanyTsiaStat} method also provides a tree-based method for treatment-subgroup interaction detection.
 
-In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure.
-However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset.
-For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
-In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
-Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects.
-Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}.
-More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.
+Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model.
+For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}.
+In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}.
+In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}.
+In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
+However, none of the purely tree-based methods for treatment-subgroup interaction detection, while also taking account the clustered structure of a dataset.
+Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models.
 
-In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account.
-The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
+The new algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}).
+MOB offers a flexible framework for subgroup detection, and when the partitioning is based on a generalized linear model (GLM), it allows for detection of treatment-subgroup interactions.
+Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit14} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}.
+The new algorithm we present combines MOB with random-effects estimates and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
 
-In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM).
+In what follows, we will discuss existing frameworks the new algorithm builds on: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM).
 Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees.
 Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets.
-In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}.
-But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.
+In the Application section, we provide an illustration by applying the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}.
+But first, we introduce an artificial motivating data set with which will be used to illustrate the methods and algorithms.
 
@@
 In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM is often used to estimate treatment effects.
-GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson - depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively.
+GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson -- depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively.
 In all cases the expectation $\mu_i$ of the outcome variable $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and a suitable link function:
@@
 \subsection{GLMM}
-When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
+When a dataset contains observations from multiple clusters (e.g., trials or research centers), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
 %
@@
 
-In RCTs, treatment effects are often estimated after controlling posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure.
-Therefore, we included the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset variable in growing the GLM and GLMM trees.
-An offset variable is a linear predictor with an a-priori specified coefficient of one.
-Including the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset has the same effect as statistically controlling for the linear effects of pretreatment scores, as is often done in ANCOVA.
+In RCTs, ANCOVAs are often used to estimate average treatement effects.
+In these ANCOVA, treatment effects are often estimated after controlling the posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure.
+Therefore, we estimated the GLM and GLMM trees using posttreatment HAM-D scores, controlled for the linear effects of pretreatment HAM-D scores.
 
 We built all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score.
-As a result, data from 694 patients from 7 studies were included for the analyses.
+As a result, data from 694 patients from 7 studies were included in the analyses.
 Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.
 
-Predictive accuracy of GLM and GLMM trees was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.
+To provide a standardized estimate of the treatment effect differences in the final nodes of the trees, we calculated node-specific Cohen's $d$ values.
+Cohen's $d$ was calculated by dividing the node-specific predicted treatment outcome difference by the node-specific pooled standard deviation.
+Confidence intervals for Cohen's $d$ were not calculated, as these would not take into account the exploratory nature of our analyses (i.e., variable and split point selection).
+The predictive accuracy of GLM and GLMM trees was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.
 
 <<depression-trees>>=
-hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
+## offset
+metadata$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
 
@@
 
-Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W}.
-Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree in the first iteration of the GLMM tree algorithm.
+Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and Table~\ref{tab:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W} and Table~\ref{tab:lmertree_C&W}.
 
@@
 
-The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
-Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
-Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and PHA.
-Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA.
+\begin{table}[]
+	\caption{Node specific summary statistics and effect sizes of HAM-D scores for GLM tree}
+	\begin{tabular}{lcccc}
+		\thickline
+		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
+		\hline
+		2		&	13.25		& 10.29		& 7.52		& 0.39			\\
+		4		&	7.82		& 7.55		& 5.51		& 0.05			\\
+		5		&	10.35		& 7.69		& 6.64		& 0.40			\\
+		\hline
+		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
+		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
+		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
+		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
+	\end{tabular}
+	\label{tab:lmtree_C&W}
+\end{table}
 
@@
 
+\begin{table}[]
+	\caption{Node specific summary statistics and effect size for GLMM tree}
+	\label{tab:lmertree_C&W}
+	\begin{tabular}{lcccc}
+	\thickline
+		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
+		\hline
+		2		&	8.27		& 7.99		& 5.86		& 0.05			\\
+		3		&	10.84		& 8.21		& 6.80		& 0.39			\\
+		\hline
+		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
+		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
+		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
+		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
+	\end{tabular}
+\end{table}
+
 By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
-The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable.
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores.
-The estimated intraclass correlation coefficient for the random intercepts was .05.
+The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
+The estimated intraclass correlation coefficient for GLMM tree was .05.
+
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Table~\ref{tab:lmertree_C&W}).
+However, for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Table~\ref{tab:lmertree_C&W}).
 
@@
 In the current paper, we presented the GLMM tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters.
-Therefore, the GLMM tree algorithm seems preeminently suited for the detection of treatment-subgroup interactions in clustered datasets.
-In the empirical evaluation, we have assessed the performance of GLMM tree in a large number of simulated dataets, and compared it with GLM tree, an algorithm that creates GLM-based recursive partitions, without estimating random effects.
-In the application, we applied GLM and GLMM tree to an existing dataset of a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
-The application showed that GLMM tree provides results that are easily interpretable, and also more accurate than a GLM tree without random effects.
-
-The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, as it accurately recovered the interactions in 90\% of the simulated datasets with treatment-subgroup interactions.
+Therefore, the GLMM tree algorithm seems well suited for the detection of treatment-subgroup interactions in clustered datasets.
+This was confirmed by our simulation study, in which we compared the performance of GLMM trees with that of GLM trees without random effects.
+GLMM tree performed very well in recovering treatment-subgroup interactions, accurately recovering the interactions in 90\% of the simulated datasets with treatment-subgroup interactions.
 In contrast, GLM tree accurately recovered the interactions in only 61\% of the datasets with treatment-subgroup interactions.
 In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.
-In other words, the Type I error rate of GLMM tree very closely resembled the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.
+In other words, the Type I error rate of GLMM tree very closely approximated the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.
 
@@
 As expected, when random effects were absent from the simulated datasets, GLM tree and GLMM tree showed high and equal predictive accuracy.
-This finding indicates that GLMM tree can be applied, whenever cluster-specific random effects are expected.
+This is an important finding, as it indicates that GLMM tree can be applied whenever cluster-specific random effects are expected.
 In the absence of random effects, GLM tree and GLMM tree are expected to perform equally well, and in the presence of random effects, GLMM tree will outperform GLM tree.
-This may especially be the case with large sample sizes ($N > 200$), as the increased power will likely cause GLM tree to create spurious splits, when random effects are present in the data.
+This may especially be the case with large sample sizes ($N > 200$), as increased power will more likely result in spurious splits, when random effects are not taken into account.
 
@@
 
+In the application, we applied GLM and GLMM tree to an existing dataset of a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression.
+This showed that GLMM tree provides results that are easily interpretable, and also more accurate than a GLM tree without random effects.
+We have shown how node-specific means and variances can be used to calculate effect sizes, to judge clinical relevance of the findings, as would often be done in RCTs or meta-analysis.
+We have limited ourselves to calculating Cohen's $d$.
+Although equivalent values of the success rate difference or the number needed to treat can be calculated, it would involve additional distributional assumptions \cite{KraeyKupf06}.
+The node-specific effect sizes can also be used to prune the tree, when a researcher prefers to have a final tree which is not only based on statistical, but also on clinical relevance.
+A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process, but this is currently not possible.
+
 These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures.
@@
 However, misspecification of the model can reduce performance.
-One source of misspecification would be, when relevant variables are not included in the GLM or as partitioning variables.
-If there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using the partitioning variables that are available.
-Or, if the coefficients of other variables vary across subgroups, then those variables should also be included in the linear predictor of the GLM.
+One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables.
+For example, if there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using variables that are available for partitioning.
 Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
-However, it should be noted that in our simulations, the number of partitioning variables did not substantially influence performance of the algorithm(s).
+It should be noted, however, that the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.
 
 In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects.
-Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.
+Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials or individual-level patient data meta-analyses.
 

### 2016-05-10 pkg/glmertree/paper/Application with CV.R

@@ -28,14 +28,14 @@ lm_o <- lmtree(HRSDt1 ~ Tx_group | Age + Gender +
 #                   data = metadata, ranefstart = metadata$HRSDfit)
 lmer_f <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) |
                      Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
-                   data = metadata, ranefstart = metadata$HRSDfit, joint = F)
+                   data = metadata, ranefstart = metadata$HRSDfit)
 plot(lm_f); print(lm_f)
 plot(lm_o); print(lm_o)
 plot(lmer_f); print(lmer_f)
 
 ## calculate predictions
 metadata$lmtreepred <- predict(lm_f, newdata = metadata)
-metadata$lmertreepred <- predict(lmer_f$tree, newdata = metadata) + predict(lmer_f$lmer, newdata = metadata)
+metadata$lmertreepred <- predict(lmer_f, newdata = metadata) 
 mean(metadata$HRSDt1); hist(metadata$HRSDt1)
 mean(metadata$lmtreepred); hist(metadata$lmtreepred)
 mean(metadata$lmertreepred); hist(metadata$lmertreepred)
@@ -49,10 +49,28 @@ var(metadata$HRSDt1) # total variance
 lmer_f$varcor[[1]][[1]] / var(metadata$HRSDt1) # variance explained by random effects
 var(metadata$lmertreepred) / var(metadata$HRSDt1) # Total variance explained by lmertree
 var(metadata$lmtreepred) / var(metadata$HRSDt1) # variance explained by the lmtree
-var(metadata$hrs) / var(metadata$HRSDt1) # variance explained by HRSDt0
+var(metadata$HRSDt0) / var(metadata$HRSDt1) # variance explained by HRSDt0
 cor(metadata$HRSDt1, metadata$lmertreepred)^2 # lmertree explains 11% of variance in outcome variable
 cor(metadata$HRSDt1, metadata$lmtreepred)^2 # lmtree explains 9% of variance outcome variable
 
+
+## Calculate effect sizes: Cohen's d = x1 - x2 / pooled_sd
+mu_mu <- predict(lm(HRSDt1 ~ HRSDt0, data = metadata), newdata = list(HRSDt0 = mean(metadata$HRSDt0)))
[…]

### 2016-05-10 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@848 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{ctable}
-\usepackage{multirow}
 \usepackage{rotating}
 \usepackage{lineno}
-\usepackage{afterpage}
+\usepackage{placeins}
 \linespread{1.2}
@@
 Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes.
-As such, the \citeA{ZhanyTsiaStat} method also provides a tree-based method for treatment-subgroup interaction detection.
+As such, the \citeA{ZhanyTsia12Stat} method also provides a tree-based method for treatment-subgroup interaction detection.
 
@@
 
+The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education.
+Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores.
+Table~\ref{tab:lmtree_C&W} shows that the treatment effect difference in Node 2 amounts to a small to medium effect size (Cohen's $d = 0.39$).
+Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and PHA.
+Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA.
+In Node 5, a small to medium effect size was also observed (Cohen's $d = 0.40$; Table~\ref{tab:lmtree_C&W}).
+
+By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
+The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
+The estimated intraclass correlation coefficient for GLMM tree was .05.
+
+The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Table~\ref{tab:lmertree_C&W}).
+However, for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Table~\ref{tab:lmertree_C&W}).
+
+Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
+The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathit{var}=.067$) for GLMM tree, and .19 ($\mathit{var}=.084$) for GLM tree.
+This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.
+
+\clearpage
+
 \begin{figure}[t!]
@@
 
-\begin{table}[]
+\begin{table}[!b]
 	\caption{Node specific summary statistics and effect sizes of HAM-D scores for GLM tree}
@@
 
-\begin{table}[]
+\begin{table}[!b]
 	\caption{Node specific summary statistics and effect size for GLMM tree}
@@
 
-By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split.
-The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable.
-The estimated intraclass correlation coefficient for GLMM tree was .05.
-
-The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Table~\ref{tab:lmertree_C&W}).
-However, for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Table~\ref{tab:lmertree_C&W}).
-
-Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree.
-The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathit{var}=.067$) for GLMM tree, and .19 ($\mathit{var}=.084$) for GLM tree.
-This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.
-
+\FloatBarrier
 \section{Discussion}
@@
 Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups.
-It should be noted, however, that the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.
+However, the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.
+
+As discussed in the Introduction, treatment-subgroup interactions have gained interest from several areas, and several different tree-based methods for their detection have been developed.
+In the current study, we have limited our comparison to the performance of GLM and GLMM-based MOB.
+Obviously, further research on what the different methods for treatment-subgroup interaction detection have in common and what sets them apart is needed.
 

### 2016-05-19 pkg/glmertree/paper/treatment_interactions.Rnw
- Included discussion of Sies & Van Mechelen paper in paper - Corrected some errors in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@855 255f4249-2263-4417-b1ea-8498779e27a7

@@
 Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}.
-Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview).
-Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes.
-As such, the \citeA{ZhanyTsia12Stat} method also provides a tree-based method for treatment-subgroup interaction detection.
+Several tree-based algorithms to the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview).
+Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, which allows users to select from a range of statistical methods, including trees.
+As such, the \citeA{ZhanyTsia12Stat} method can also be referred to as a tree-based method.
 
@@
 In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}.
-However, none of the purely tree-based methods for treatment-subgroup interaction detection, while also taking account the clustered structure of a dataset.
+However, none of the purely tree-based methods for treatment-subgroup interaction detection, also take into account the clustered structure of a dataset.
 Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models.
@@
 MOB offers a flexible framework for subgroup detection, and when the partitioning is based on a generalized linear model (GLM), it allows for detection of treatment-subgroup interactions.
-Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit14} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}.
-The new algorithm we present combines MOB with random-effects estimates and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
+Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit16} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}.
+The new algorithm we present combines MOB with random-effects estimation and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.
 
@@
 In the Application section, we provide an illustration by applying the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}.
-But first, we introduce an artificial motivating data set with which will be used to illustrate the methods and algorithms.
+But first, we introduce an artificial motivating data set which will be used to illustrate the methods and algorithms.
 
@@
 %
-To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
+To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression trees (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}, but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm.
 In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual.
@@
 
-As discussed in the Introduction, treatment-subgroup interactions have gained interest from several areas, and several different tree-based methods for their detection have been developed.
-In the current study, we have limited our comparison to the performance of GLM and GLMM-based MOB.
-Obviously, further research on what the different methods for treatment-subgroup interaction detection have in common and what sets them apart is needed.
+As discussed in the Introduction, several tree-based methods for treatment-subgroup interaction detection are available.
+These methods have different objectives, and there is not yet an agreed-upon single best method.
+In a simulation study, \citeA{SiesyVanM16} found the method of \cite{ZhanyTsia12Stat} to perform best, followed by model-based recursive partitioning.
+However, the method of Zhang et al.
+performed worst under some conditions of the simulation study in terms of the Type I error rate.
+Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations were limited to GLM and GLMM-based MOB.
 

### 2016-05-20 pkg/glmertree/paper/treatment_interactions.Rnw
Minor improvements in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@858 255f4249-2263-4417-b1ea-8498779e27a7

@@
 \usepackage{verbatim}
+\usepackage{bigfoot}
 \usepackage{ctable}
@@
 outcome[outcome<0] <- 1
-example <- data.frame(treatment,outcome,duration,anxiety,cluster, true_eff)
-example$treatment <- factor(example$treatment,labels=c("Treatment 1","Treatment 2"))
-#cor(outcome,rand_int)^2
-#cor(outcome, true_eff)^2
-#truetree <- lmtree(true_eff ~ treatment | age + duration + anxiety, data=example)
+example <- data.frame(treatment, outcome, duration, anxiety, cluster, age)
+example$treatment <- factor(example$treatment, labels = c("Treatment 1", "Treatment 2"))
 @
@@
 Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.
+
+The \verb|R| code for running the illustrative examples is provided in the Appendix.
 
@@
 
-  \item[Step 2:] Estimate random effects in the mixed effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
+  \item[Step 2:] Estimate random effects in the mixed-effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
   % Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.
@@
 R \cite{R14} was used for generation and analysis of all datasets.
-The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions.
+The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions.
 For other response distributions, the \verb|glmtree| function would be available.
@@
 
-For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge).
+For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge.
 This provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.
